### 2018-07-17 notes/nbeSum.tex
ITC lecture: reflect/reify for implication

\nonstopmode
\documentclass[a4paper]{article}
\usepackage{natbib}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{amsmath}
%\usepackage{dsfont}

\input{macros}

\newtheorem{theorem}{Theorem}
%\newtheorem*{theorem*}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}

%\newcommand{\qed}{\hfill\ensuremath{\Box}}
% from llncs.cls
%\def\squareforqed{\hbox{\rlap{$\sqcap$}$\sqcup$}}
\def\squareforqed{\ensuremath{\Box}}
\def\qed{\ifmmode\squareforqed\else{\unskip\nobreak\hfil
\penalty50\hskip1em\null\nobreak\hfil\squareforqed
\parfillskip=0pt\finalhyphendemerits=0\endgraf}\fi}

\newenvironment{proof}[1][]{\noindent\ifthenelse{\equal{#1}{}}{{\it
      Proof.}}{{\it Proof #1.}}\hspace{2ex}}{\qed\bigskip}
\newenvironment{proof*}[1][]{\noindent\ifthenelse{\equal{#1}{}}{{\it
      Proof.}}{{\it Proof #1.}}\hspace{2ex}}{\bigskip}

\newcommand{\wk}{\mathord{\uparrow}}
\newcommand{\WW}{\mathbb{W}}
\newcommand{\tAr}{\mathsf{Ar}}
\newcommand{\Ar}[1]{\tAr(#1)}

\title{Normalization by Evaluation for \\ Intuitionistic Propositional Logic}
\author{Andreas Abel}
\date{July 2018}

\begin{document}
\maketitle

\section{Intuitionistic Propositional Logic (IPL)}

\para{Formulas} and hypotheses lists (contexts).
\[
\begin{array}{lllrll}
  \Atom & \ni & P,Q & & & \mbox{atomic propositions} \\
  \Form & \ni & A,B,C,\dots
     & ::= & \top \mid A \land B \mid A \imp B
       & \mbox{negative formulas} \\
     &&& \mid & \bot \mid A \lor B \mid P
       & \mbox{positive formulas} \\
\\[1ex]
  \Cxt & \ni & \Gamma,\Delta & ::= & \cempty & \mbox{empty context}
\\ &&& \mid & \Gamma.A & \mbox{context extension}
\end{array}
\]
Indices $x$ into the context $\Gamma$ of hypothesis, locating hypothesis
$A$, are written $x : \Hyp[\Gamma]A$ and defined inductively by the
following constructors:
\[
  \ru{}{\ttop : \Hyp[\Gamma.A]A}
\qquad
  \ru{x : \Hyp[\Gamma]A
    }{\pop[B] x : \Hyp[\Gamma.B]A}
\]
Deriviations $t$ of the truth of a formula $A$ under assumptions
$\Gamma$ (judgement $\Gamma \der A$)
are written $t : \Tm[\Gamma]A$ and defined inductively as follows:

\paradot{Implication and hypotheses}
\begin{gather*}
  \ru{x : \Hyp[\Gamma]A}{\hyp x : \TmG A}
\qquad
  \ru{t : \Tm[\Gamma.A]B}{\impI t : \TmG{A \imp B}}
\qquad
  \ru{t : \TmG{A \imp B} \qquad
      u : \TmG A
    }{\impE t u : \TmG B}
\end{gather*}

\paradot{Truth and conjunction}
\begin{gather*}
  \ru{}{\trueI : \TmG\top}
\qquad
  \ru{t : \TmG A \qquad
      u : \TmG B
    }{\andI t u : \TmG{A \land B}
    }
\\[1ex]
  \ru{t : \TmG{A \land B}
    }{\andE1 t : \TmG A
    }
\qquad
  \ru{t : \TmG{A \land B}
    }{\andE2 t : \TmG B
    }
\end{gather*}

\paradot{Absurdity and disjunction}
\begin{gather*}
  \ru{t : \TmG A
    }{\orI1 t : \TmG {A \lor B}
    }
\qquad
  \ru{t : \TmG B
    }{\orI2 t : \TmG {A \lor B}
    }
\\[1ex]
  \ru{t : \TmG \bot
    }{\falseE t : \TmG C}
\qquad
  \ru{t : \TmG {A \lor B} \qquad
      u : \Tm[\Gamma.A]C  \qquad
      v : \Tm[\Gamma.B]C
    }{\orE t u v : \TmG C}
\end{gather*}

\section{Subformula property and normal derivations}

For every derivable judgement $\Gamma \der A$ there are infinitely
many derivations $t : \TmG A$, since arbitrary detours are allowed.
For instance to prove $\top$ we can also proceed very indirectly by
introducing a hypothesis $\top$ and then eliminating it by proof
$\trueI$:
\[\impEP{\impIP{\hyp{\ttop}}}{\trueI} : \Tm[\cempty]\top.\]
A sensible restriction is that the proof rules guarantee that only
\emph{subformulas} of $A$ and the hypotheses in $\Gamma$ are mentioned when
deriving $\Gamma \der A$.

The following indexed grammar for \emph{normal} derivations $\NfG A$ of
judgement $\Gamma \der A$ ensures the subformula property.  It is
defined mutually with a grammar $\NeG A$ of \emph{neutral} derivations.
Neutral derivations are ``straightline'' consequences of hypotheses
(without case distinction or absurdity elimination).

\paradot{Implication and hypotheses}
\begin{gather*}
  \ru{x : \Hyp[\Gamma]A}{\hyp x : \NeG A}
\qquad
  \ru{t : \Nf[\Gamma.A]B}{\impI t : \NfG{A \imp B}}
\qquad
  \ru{t : \NeG{A \imp B} \qquad
      u : \NfG A
    }{\impE t u : \NeG B}
\end{gather*}

\paradot{Truth and conjunction}
\begin{gather*}
  \ru{}{\trueI : \NfG\top}
\qquad
  \ru{t : \NfG A \qquad
      u : \NfG B
    }{\andI t u : \NfG{A \land B}
    }
\\[1ex]
  \ru{t : \NeG{A \land B}
    }{\andE1 t : \NeG A
    }
\qquad
  \ru{t : \NeG{A \land B}
    }{\andE2 t : \NeG B
    }
\end{gather*}

\paradot{Absurdity and disjunction}
\begin{gather*}
  \ru{t : \NfG A
    }{\orI1 t : \NfG {A \lor B}
    }
\qquad
  \ru{t : \NfG B
    }{\orI2 t : \NfG {A \lor B}
    }
\\[1ex]
  \ru{t : \NeG \bot
    }{\falseE t : \NfG C}
\qquad
  \ru{t : \NeG {A \lor B} \qquad
      u : \Nf[\Gamma.A]C  \qquad
      v : \Nf[\Gamma.B]C
    }{\orE t u v : \NfG C}
\end{gather*}

\paradot{Embedding}
\[
  \ru{t : \NeG P
    }{\neu t : \NfG P}
\]

The introduction rules generate normal forms ($\tNf$s) from $\tNf$s.  The elimination
rules for negative formulas generate neutrals ($\tNe$s) from $\tNe$s, where
$\timpE$ requires a side argument in normal form.

Some care is needed with  the elimination rules for
positive formulas which are $\tfalseE$ and $\torE$.
These prove an arbitrary formula $C$, which, for the sake of the
subformula property, should not be subject to
further elimination but directly prove the goal.
Thus, in addition to restriction the elimination to neutrals
(perform case distinction only on neutrals),
we require them to produce a normal form
in case of $\torE$ from normal branches.

Any neutral derivation at atomic proposition $P$ is considered normal.
The restriction on atoms is not needed for the subformula property,
but forces derivations to be $\eta$-long.
We could drop the restriction, but even with this restriction, the
calculus is complete, meaning that for every derivation ($\tTm$) there
exists a normal derivation ($\tNf$).  This statement is called
\emph{normalization} and we will prove it constructively in the
following.

\section{Categories and Presheaves}

We define order-preserving embeddings (OPEs) $\tau : (\Delta \leq \Gamma)$ of a
context $\Gamma$ into a larger context $\Delta$ inductively by the
following constructors.
\begin{gather*}
  \ru{}{\id[\Gamma] : \Gamma \leq \Gamma}
\qquad
  \ru{\tau : \Delta \leq \Gamma
    }{\weak[A] \tau : \Delta.A \leq \Gamma}
\qquad
  \ru{\tau : \Delta \leq \Gamma
    }{\lift[A] \tau : \Delta.A \leq \Gamma.A}
\end{gather*}
For example $\weak [D] (\lift[C] (\weak[B] \id[A])) : A.B.C.D \leq
A.C$.  OPEs let us introduce extra, unused hypotheses into the
context.  Derivations of $\Gamma \der A$ can be weakened to
derivations $\Delta \der A$ for $\Delta \leq \Gamma$.

OPEs are closed under composition:  If $\tau : \Delta \leq \Gamma$ and
$\tau' : \Phi \leq \Delta$ then $\tau \comp \tau' : \Phi \leq \Delta$.
Let $\OPE$ be the category of contexts (as objects) and OPEs (as morphisms).
\begin{exercise}
  Define composition of OPEs by induction and show that $\OPE$ is
  indeed a category.
\end{exercise}

The context-indexed sets $\Hyp A$, $\Tm A$, $\Nf A$, and $\Ne A$ are
closed under weakening. \Eg{} $\tHyp$: for each $\tau : \Delta \leq \Gamma$
we have a function
\[
  \Hyp[\tau]A : \Hyp[\Gamma] A \to \Hyp[\Delta] A,
\]
\ie, a morphism in the category $\SET$ of sets and functions.
The collection $\Hyp A : (\tau : \Delta \leq \Gamma) \to
(\Hyp[\Gamma]A \to \Hyp[\Delta]A)$ of these function constitutes a
contravariant functor $\Hyp A : \OPE \to \SET$ mapping OPEs to
functions.  In other words, $\Hyp A$ is a \emph{presheaf} over $\OPE$
for each $A$, and so are $\Tm A$, $\Nf A$, and $\Ne A$.

\begin{exercise}
  Prove that $\Hyp A$, $\Tm A$, $\Nf A$, and $\Ne A$ are presheaves
  over $\OPE$ by defining the map functions and proving the functor laws!
\end{exercise}

A function $f$ between presheaves $\A$ and $\B$ is defined pointwise;
we write $f : \A \todot \B$ for the $\Cxt$-indexed collection
\[
  f_\Gamma : \A_\Gamma \to \B_\Gamma
\]
of functions (morphisms in $\SET$). Presheaves and functions between
them make a category by virtue of pointwise identity and composition.
However, typically the functions we consider are \emph{natural} in the
context index, \ie, they are \emph{natural transformations} between
presheaves, which means that they commute with OPEs.  Given $\tau :
\Delta \leq \Gamma$ and $f : \A \todot \B$, naturality means
\[
  f_\Delta \comp \A_\tau = \B_\tau \comp f_\Gamma .
\]
In words, first weakening with $\tau$ and then applying $f$ has the
same effect as applying $f$ first and perform the weakening later.
This guarantees some form of \emph{parametricity} of $f$ in the
context, in particular, $f$ cannot make decisions based on the length
of the context or its precise contents.

Presheaves over $\OPE$ and natural transformations form a category
$\PSh$, traditionally called $\widehat{\OPE}$.

\begin{exercise}
  Prove that $\PSh$ is indeed a category.
\end{exercise}

\begin{exercise}
  Prove that the following functions are morphisms in $\PSh$:
\[
\begin{array}{lllll}
  \thyp & : & \Hyp A & \todot & \Ne A \\
  \tandE1 & : & \Ne{A \land B} & \todot & \Ne A \\
  \torI1  & : & \Nf A & \todot & \Nf{A \lor B} \\
  \tfalseE & : & \Ne \bot & \todot & \Nf C \\
\end{array}
\]
  Find other examples of presheaf morphisms.
\end{exercise}

\section{NbE for the negative fragment}

We wish to define a normalization function for derivations of any
formula $A$
\[
  \tnorm^A : \Tm A \todot \Nf A .
\]
Normalization by evaluation (NbE) achieves normalization by first
evaluating terms, obtaining values and functions that compute by means
of the computation in the meta language, i.e., in $\SET$.  As a second
step, these values are reified back into syntax, yielding a normal
form.  Technically, we define a suitable presheaf $\den A$ for each
formula $A$ and two functions
\[
\begin{array}{lllll}
  \teval^A & : & \Tm A  & \todot & \den A \\
  \treify^A  & : & \den A & \todot & \Nf A \\
\end{array}
\]
which compose to $\tnorm$.

\subsection{Semantics, reflection and reification}

For now, let us focus on the negative
formulas $\top$, $A \land B$, $A \imp B$ and atoms $P$.
\[
\begin{array}{lll@{\qquad}lll}
  \den P & = & \Nf P
    & \reify[\Gamma] P t & = & t \\
  \den \top & = & \hatone
    & \reify[\Gamma] \top \_ & = & \trueI \\
  \den{A \land B} & = & \den A \hattimes \den B
    & \reify[\Gamma]{A \land B} {(a,b)} & = & \andIPP {\reify[\Gamma] A a} {\reify[\Gamma] B b} \\
\end{array}
\]
Herein, we use the unit presheaf $\hatone_\Gamma = 1$ where $1$ is the
unit set, and the pointwise product of presheaves
$(\A \hattimes \B)_\Gamma = \A_\Gamma \times \B_\Gamma$.  We note that
by these definitions, the category $\PSh$ has products, with the
pointwise terminal morphisms:
\[
\begin{array}{lll@{\qquad}lll}
   ! & : & \A \todot \hatone &
     [\_,\_] & : & (f : \C \todot \A) (g : \C \todot \B) \to (\C
                   \todot \A \hattimes \B) \\
   !_\Gamma(a) & = & ()       &
     [f,g]_\Gamma(c) & = & (f(c),g(c)) \\
\end{array}
\]

The interpretation $\den{A \imp B}$ of implication will follow the
Brouwer-Heyting-Kolmogorov (BHK) interpretation of intuitionistic
logic:
\begin{center}\itshape
  A proof of $A \imp B$ is a method turning any proof of $A$ into a
  proof of $B$.
\end{center}
However, the direct lifting of the functions space $\den{A \imp
  B}_\Gamma = \den A_\Gamma \to \den B_\Gamma$ does not work.  It does
not give a presheaf because of the contravariant occurrence of
presheaf $\den A$.  Instead, we force monotonicity by quantifying over
all OPEs of $\Gamma$:
\[
  \den{A \imp B}_\Gamma = \prod_{\tau : \Delta \leq \Gamma} (\den A
  _\Delta \to \den B _\Delta)
\]
Reification is also challenging.  Let $f : \den{A \imp B}_\Gamma$ and
attempt
%$\tau : \Delta \leq \Gamma$
\[
  \reify[\Gamma]{A \imp B} f %\;\Delta\;\tau =
    = \impIP {\reify[\Gamma.A]B{(f\;(\weak[A]{\id[\Gamma]})}\;\mbox{?})}
\]
where hole ``?'' should be filled with an element of $\den B_{\Gamma.A}$
which intuitively should represent the new hypothesis $\top :
\Hyp[\Gamma.A]A$ introduced by $\timpI$.  Thus we need a natural
transformation $\Hyp A \todot \den A$ which \emph{reflects} variables
into the semantics.  In order to define it by
induction on $A$, we need to generalize it to
\[
  \treflect^A : \Ne A \todot \den A
  .
\]
Let us make sure we can define it at atoms and conjunctions:
\[
\begin{array}{lllll}
  \treflect^P & = & \tne & : & \Ne P \todot \Nf P \\
  \treflect^\top & = & !  & : & \Ne \top \todot \hatone \\
  \treflect^{A \land B} & = & [\, \treflect^A \comp \andE1  , \;
                             \treflect^B \comp \andE2 ]
    & : & \Ne{A \land B} \todot \den A \hattimes \den B \\
\end{array}
\]
Observe the eliminations $\andE i$ introduced to reflect conjunctive hypotheses!

Now, we can complete the story for implication.  Let $\tau : \Delta
\leq \Gamma$ and $a : \den A_\Delta$ and set:
\[
\begin{array}{lll}
  \treflect^{A \imp B} & : & \Ne{A \imp B} \todot \den{A \imp B} \\
  \reflect[\Gamma]{A \imp B} t\;\Delta\;\tau\;a & = &
    \reflect[\Delta]B{(\impEP {\Ne[\tau] B t} {(\reify[\Delta] A a)})}
\\[1ex]
  \treify^{A \imp B} & : & \den{A \imp B} \todot \Nf{A \imp B} \\
  \reify[\Gamma]{A \imp B} f & = &
    \impIP {\reify[\Gamma.A]B{(f\;(\weak[A]{\id[\Gamma]})}\;(\reflect[\Gamma.A]A{(\hyp\ttop)})}
\end{array}
\]

\begin{exercise}
  Let $A = ((P \imp Q) \land (Q \imp \top)) \imp Q$ for some atoms
  $P,Q$.  Compute $\reflect[A]A{(\hyp \ttop)}$.
\end{exercise}

\begin{exercise}
  Let $A = (P \imp Q) \imp (P \imp Q)$ for some atoms $P,Q$.
  Compute $\treify^A \comp \treflect^A$.
\end{exercise}

\begin{exercise}
  Using the constructions in this section, show that the category of
  presheaves and natural transformations is cartesian closed.
\end{exercise}

\subsection{Evaluation}

\bibliographystyle{abbrvnat}
\bibliography{auto-nbeSum}

\end{document}

### 2018-07-17 notes/nbeSum.tex
ITC: evaluation

@@
 \[
-  \den{A \imp B}_\Gamma = \prod_{\tau : \Delta \leq \Gamma} (\den A
+  \den{A \imp B}_\Gamma = \prod_{\Delta : \Cxt} \prod_{\tau : \Delta \leq \Gamma} (\den A
   _\Delta \to \den B _\Delta)
@@
 
+Evaluation is turning an expression $t : \TmG A$ into a value in
+$\den A$.  However, a direct definition $\teval^A : \Tm A \todot \den A$
+by induction on the expression fails in case $\timpI$:
+\[
+\begin{array}{lll}
+   \teval^{A \imp B}_\Gamma & : & \TmG A \to \den {A \imp B} _\Gamma \\
+   \teval^{A \imp B}\,(\impI t)\;\Delta\;a & = & ? \\
+\end{array}
+\]
+To be able to pass the argument $a$ to the function $\teval\,(\impI
+t)$, we have to generalize evaluation to take an valuation of the
+context into account.
+\[
+  \teval^{A \imp B}_\Gamma  : \den \Gamma \todot \den {A \imp B}
+\]
+Here, $\den \Gamma$ is the presheaf obtained as the product of the
+hypotheses contained in $\Gamma$:
+\[
+\begin{array}{lll}
+  \den{\cempty} & = & \hatone \\
+  \den{\Gamma.A} & = & \den\Gamma \hattimes \den A \\
+\end{array}
+\]
+Looking up values in the environment is the following family of
+natural transformations:
+\[
+\begin{array}{l@{~}lcl}
+  \multicolumn 2 l {\tlookup^A_\Gamma} & : & \HypG A \to (\den \Gamma \todot \den A) \\
+  \lookup[\Gamma.A]A \ttop    & \Delta\,(\gamma,a) & = & a \\
+  \lookup[\Gamma.B]A {\pop x} &\Delta\,(\gamma,a) & = & \lookup[\Gamma]A x\, \Delta\, \gamma \\
+\end{array}
+\]
+\begin{exercise}
+  Express $\tlookup$ via the projections
+  $\pi_i : \A_1 \hattimes \A_2 \todot \A_i$
+  out of the presheaf product.
+\end{exercise}
+We are ready to define evaluation:
+\[
+\begin{array}{lll}
+  \evalG A {\hyp x} & = & \lookupG A x \\
+  \evalG {A \imp B} {\impI t}\,\Delta\,\gamma\,\Delta'\,(\tau : \Delta' \leq \Delta)\,a
+    & = & \eval[\Gamma.A]B t\,\Delta'\,(\den\Gamma_\tau\gamma,a) \\
+  \evalG {B} {\impE t u}\,\Delta\,\gamma & = &
+    \evalG {A \imp B} t\,\Delta\,\gamma\,\Delta\,\id[\Delta]\,
+      (\evalG A u\,\Delta\,\gamma) \\
+\end{array}
+\]
+\begin{exercise}
+  Complete the definition of $\teval$ for conjunctions.
+\end{exercise}
+\begin{exercise}
+  Express $\teval$ in terms of the operations of the CCC of
+  presheaves!
+\end{exercise}
+
+\subsection{Digression: The CCC of reifyable presheaves}
+
+Instead of defining the semantics $\den A$ of formula $A$ by induction
+on $A$, we can consider a category whose objects are interpretations
+of formulas, and show that this category is cartesian closed.
+
+An object in this category would be a quadruple consisting of:
+\[
+\begin{array}{lll@{\qquad}l}
+  \A & : & \PSh  & \mbox{a presheaf} \\
+  A  & : & \Form & \mbox{a formula} \\
+  \treflect^A & : & \Ne A \todot \A
+    & \mbox{a method to reflect neutrals into $\A$} \\
+  \treify^A & : & \A \todot \Nf A
+    & \mbox{a method to reify values of $\A$} \\
+\end{array}
+\]
+\begin{exercise}
+  Complete this definition by a suitable notion of morphism.
+\end{exercise}
+\begin{exercise}
+  Show that this category is cartesian closed.
+\end{exercise}
+
+\section{Extension to the positive connectives}
+
 \bibliographystyle{abbrvnat}

### 2018-07-18 notes/nbeSum.tex
ITC: reify/reflect for disjunction and absurdity

@@
 \]
+We write just $A$ for the singleton context $\cempty.A$.
+
 Indices $x$ into the context $\Gamma$ of hypothesis, locating hypothesis
@@
   \treflect^\top & = & !  & : & \Ne \top \todot \hatone \\
-  \treflect^{A \land B} & = & [\, \treflect^A \comp \andE1  , \;
-                             \treflect^B \comp \andE2 ]
+  \treflect^{A \land B} & = & \langle\, \treflect^A \comp \andE1  , \;
+                             \treflect^B \comp \andE2 \rangle
     & : & \Ne{A \land B} \todot \den A \hattimes \den B \\
@@
 
-\subsection{Digression: The CCC of reifyable presheaves}
+\subsection{Digression: The CCC of reifiable presheaves}
 
@@
 
-\section{Extension to the positive connectives}
+%\section{Extension to the positive connectives}
+\section{Disjunction and absurdity}
+
+For sets $S_1,S_2 : \SET$ let $S_1 + S_2$ denote the disjoint union with
+injections $\inj{i} : S_i \to S_1 + S_2$ and case distinction
+distinction $[f_1,f_2] : S_1 + S_2 \to T$ for $f_i : S_i \to T$
+such that $[f_1,f_2] \comp \inj i = f_i$.
+
+Presheaf coproduct is given pointwise by $(\A \hatplus \B)_\Gamma = \A_\Gamma + \B_\Gamma$.
+\begin{exercise}
+  Show that $\hatplus$ is indeed a coproduct in $\PSh$.
+\end{exercise}
+
+\subsection{Semantics, reflection and reification}
+
+The first idea how to model disjunction,
+$\den{A \lor B}_\Gamma = \den{A}_\Gamma + \den{B}_\Gamma$, fails
+because we cannot reflect disjunctive hypotheses.  \Eg, an element of
+$\den{A \lor B}_{A \lor B}$ should be of the form $\inj1\,a$ or
+$\inj2\,b$, but the decision on first or second injection cannot be
+made here;  it depends on the hypothesis $A \lor B$.  The solution is
+to extend the semantics to allow case distinction on variables (and
+neutrals) before making the decisions needed for disjunctions.
+
+Given a presheaf $\A$, we define a new presheaf $\Cover \A$
+inductively by the following constructors:
+\begin{gather*}
+  \ru{a : \A_\Gamma}{\returnC a : \Cover\A_\Gamma}
+\qquad
+  \ru{t : \NeG \bot}{\falseC t : \Cover\A_\Gamma}
+\\[1ex]
+  \ru{t : \NeG{C \lor D} \qquad
+      c : \Cover\A_{\Gamma.C} \qquad
+      d : \Cover\A_{\Gamma.D}
+    }{\orC t c d : \Cover\A_\Gamma}
+\end{gather*}
+\begin{exercise}
+  Show that $\Cover\A$ is indeed a presheaf.
+\end{exercise}
+We can think of $c : \Cover\A_\Gamma$ as a decision tree with leaves in $\A$
+and nodes labelled by terms we case on.  Nodes of type $\tfalseC$ have
+no subtrees since \emph{ex falsum quod libet}, and $\torC$ nodes
+casing on $t : \NeG{C \lor D}$ have
+two subtrees, one having a new hypothesis $C$, and one having a new
+hypothesis $C$.
+
+We can replace all the leaves according to a function $f : \A \todot \B$,
+this makes $\tCover$ a \emph{functor} in the category of presheaves.  We can
+also replace all the leaves by new case trees via a natural transformation
+\[
+   \tjoinC ~:~ \Cover{\Cover\A} \todot \Cover\A
+\]
+which makes $\tCover$ a \emph{monad} in $\PSh$.
+\begin{exercise}
+  Define the functorial action $\Cover f$ and $\tjoinC$ and prove the
+  monad laws.
+\end{exercise}
+Further, if the leaves of a case tree are normal forms, we can turn it into one big
+normal form via:
+\[
+  \tpasteNf ~:~ \Cover{\Nf A} \todot \Nf A
+\]
+\begin{exercise}
+  Define $\tpasteNf$.
+\end{exercise}
+
+\noindent
+We have everything now to model disjunction:
+\[
+\begin{array}{lcl}
+\den{A \lor B} & = & \Cover{\den A \hatplus \den B}
+\\[2ex]
+\treflect^{A \lor B} & : & \Ne{A \lor B} \todot \den{A \lor B} \\
+\reflect[\Gamma]{A \lor B}t & = & \orC t
+  {(\returnCP {\injP1 {\reflect[\Gamma.A]A{(\hyp \ttop)}}})}
+  {\\ && \qquad\ (\returnCP {\injP2 {\reflect[\Gamma.B]B{(\hyp \ttop)}}})}
+\\[2ex]
+\treifyOr^{A \lor B} & : & \den{A} \hatplus \den{B} \todot \Nf{A \lor B} \\
+\reifyOr[\Gamma]{A \lor B}(\inj1\, a) & = & \orIP1 {\reify[\Gamma]A a} \\
+\reifyOr[\Gamma]{A \lor B}(\inj2\, b) & = & \orIP2 {\reify[\Gamma]B b}
+\\[1.5ex]
+\treify^{A \lor B} & : & \den{A \lor B} \todot \Nf{A \lor B} \\
+\treify^{A \lor B} & = & \tpasteNf \comp \Cover{\treifyOr^{A \lor B}} \\
+\end{array}
+\]
+Reification works by first reifying the leaves of the case tree using
+$\treifyOr$, and then turning the whole tree into a normal form using $\tpasteNf$.
+\begin{exercise}
+  Compute $\reify[P \lor Q]{P \lor Q} (\reflect[P \lor Q]{P \lor Q}
+  (\hyp \ttop))$.
+\end{exercise}
+For absurdity, we use the presheaf $\hat0_\Gamma = \{\}$ with initial
+morphism $\botelim : \hat0 \todot \C$.
+\[
+\begin{array}{lcl}
+\den{\bot} & = & \Cover{\hat0}
+\\[2ex]
+\treflect^{\bot} & : & \Ne{\bot} \todot \den{\bot} \\
+\reflect[\Gamma]{\bot}t & = & \falseC t
+\\[2ex]
+\treify^{\bot} & : & \den{\bot} \todot \Nf{\bot} \\
+\treify^{\bot} & = & \tpasteNf \comp \Cover{\botelim} \\
+\end{array}
+\]
+To reify a value $c : \den\bot$, observe that because $\hat0$ is
+empty, $c$ can only be a case tree without leaves, \ie, all branches
+end in $\tfalseC$ nodes.  Thus, the mapping
+$\Cover{\botelim} : \Cover{\hat0} \todot \Cover{\Nf\bot}$ is merely a
+type cast.  The subsequent $\tpasteNf$ will turn the leafless tree $c$
+into a normal form consisting entirely of splits $\torE$ and $\tfalseE$.
+
+\begin{exercise}
+Prove that $\treify^{A} \comp \treflect^{A}$ produces the long
+$\eta$-normal form if applied to variables (\ie, hypotheses).
+\end{exercise}
+
+\subsection{Evaluation}
+
+Evaluation poses one more challenge.
 

### 2018-07-18 notes/nbeSum.tex
ITC: evaluation for orE

@@
 \[
-\begin{array}{lll}
+\begin{array}{lcl}
   \evalG A {\hyp x} & = & \lookupG A x \\
@@
 \end{exercise}
+
+\subsection{Normalization}
+
+Normalization is reification after evaluation in the identity
+environment:
+\[
+\begin{array}{lcl}
+\tnorm^A & : & \Tm A \todot \Nf A \\
+\norm[\Gamma] A t & = & \reify[\Gamma]A {(\eval[\Gamma] A t \,\Gamma\, \env[\Gamma])}
+\end{array}
+\]
+The identity environment $\env[\Gamma]$ is defined by induction on
+$\Gamma$ as follows:
+\[
+\begin{array}{lcl}
+  \env[\Gamma] & : & \den \Gamma _\Gamma \\
+  \env[\cempty] & = & () \\
+  \env[\Gamma.A] & = & (\den\Gamma_{\weak[A] \id[\Gamma]}(\env[\Gamma]), \;
+                          \reflect[\Gamma.A] A {(\hyp \ttop)})
+\end{array}
+\]
 
@@
 
-Evaluation poses one more challenge.
+Evaluation poses one more challenge. Consider the term:
+\[
+\ru{x : \HypG {C \lor D} \qquad
+    f : \Tm[\Gamma.C]{A \imp B} \qquad
+    g : \Tm[\Gamma.D.A]{B} \qquad
+    a : \TmG A
+  }{\impEP{\orE {(\hyp x)} f {(\impI g)}} a : \TmG B}
+\]
+There is a redex $\impEP {\impI g} a$ separated by a $\torE$.  The latter
+would naturally evaluate to a case tree with functions in the
+leaves.  To apply it to the evaluation of $a$, we have to apply each
+leaf to that argument.  We facilitate this by implementing
+\[
+  \tpaste^{A \imp B} : \Cover{\den{A \imp B}} \todot \den{A \imp B}
+\]
+that turns a case tree of functions into a function, which expects an
+argument that will then be passed to all the functions in the case
+tree.  In fact, we need pasting at all formulas $A$:
+\[
+\begin{array}{lcl}
+\tpaste^A & : & \Cover{\den A} \todot \den A \\[1ex]
+\tpaste^P & = & \tpasteNf \\
+\tpaste^\bot & = & \tjoinC \\
+\tpaste^{A \lor B} & = & \tjoinC \\
+\tpaste^{A \land B} & = & \langle\,
+  \tpaste^A \comp \Cover{\pi_1} \, , \;
+  \tpaste^B \comp \Cover{\pi_2} \,\rangle \\[1ex]
+\multicolumn 3 {l} {\paste[\Gamma]{A \imp B} c \;
+  \Delta \; (\tau : \Delta{\leq}\Gamma) \; (a : \den A _\Delta) \, = \,
+  (\tpaste^B \comp \mapC \phi \comp \Cover{\den {A \imp B}_\tau}\, c)
+}
+\end{array}
+\]
+Here, we need a stronger version of functoriality of $\tCover$ which
+only requires an
+\[
+  \phi : \prod_{\Phi} \prod_{\delta : \Phi \leq \Delta} \A_\Phi
+             \to \B_\Phi
+\]
+to produce a $\tmapC_\Delta t\; \phi : \Cover\A_\Delta \to \Cover\B_\Delta$.
+In our case, $\A = \den{A \imp B}$ and $\B = \den{B}$ and
+\[
+  \phi\;\Phi\;\delta\;f = f\;\id[\Phi]\;(\den A _\delta\, a)
+  .
+\]
+Pasting allows us to complete the definition of evaluation.
+\[
+\begin{array}{lcl}
+  \teval_\Gamma^A & : & \TmG A \to \den \Gamma \todot \den A \\[1ex]
+  \evalG C {\orE t u v} \Delta \; \gamma & = &
+    \tpaste^C \;
+     (\mapC {(\lambda \Phi\, (\tau  : \Phi \leq \Delta) \to [f,g])}\;c) \\[1ex]
+  \multicolumn 1 r {\mbox{ where } c} & = &
+      \evalG{A \lor B} t\;\Delta\;\gamma \\
+  \multicolumn 1 r {f\;a} & = &
+    \eval[\Gamma.A] C u \;\Phi\;(\den\Gamma_\tau\,\gamma, a) \\
+  \multicolumn 1 r {g\;b} & = &
+    \eval[\Gamma.B] C v \;\Phi\;(\den\Gamma_\tau\,\gamma, b) \\
+  \\
+\end{array}
+\]
+\begin{exercise}
+  Really complete the definition of $\teval$ (all missing cases).
+\end{exercise}
+That's it!
+
+\section{Conclusion}
+
+\paradot{Acknowledgments}  The presentation of this NbE proof and its
+Agda formalization benefitted greately from discussions with Christian
+Sattler.  Thanks to Jesper Cockx for listening to a prerun of the talk.
 

### 2018-07-19 notes/nbeSum.tex
ITC: bibliography
ITC: spell-check

@@
 %\usepackage{dsfont}
+\usepackage{bibentry}
 
@@
 
+\nobibliography{auto-nbeSum}
+
 \section{Intuitionistic Propositional Logic (IPL)}
@@
 \]
-Deriviations $t$ of the truth of a formula $A$ under assumptions
+Derivations $t$ of the truth of a formula $A$ under assumptions
 $\Gamma$ (judgement $\Gamma \der A$)
@@
 defined mutually with a grammar $\NeG A$ of \emph{neutral} derivations.
-Neutral derivations are ``straightline'' consequences of hypotheses
+Neutral derivations are ``straight-line'' consequences of hypotheses
 (without case distinction or absurdity elimination).
@@
   Using the constructions in this section, show that the category of
-  presheaves and natural transformations is cartesian closed.
+  presheaves and natural transformations is Cartesian closed.
 \end{exercise}
@@
 on $A$, we can consider a category whose objects are interpretations
-of formulas, and show that this category is cartesian closed.
+of formulas, and show that this category is Cartesian closed.
 
@@
 \begin{exercise}
-  Show that this category is cartesian closed.
+  Show that this category is Cartesian closed.
 \end{exercise}
@@
 We can think of $c : \Cover\A_\Gamma$ as a decision tree with leaves in $\A$
-and nodes labelled by terms we case on.  Nodes of type $\tfalseC$ have
+and nodes labeled by terms we case on.  Nodes of type $\tfalseC$ have
 no subtrees since \emph{ex falsum quod libet}, and $\torC$ nodes
@@
 \begin{array}{lcl}
-\den{\bot} & = & \Cover{\hat0}
+\den{\bot} & = & \tCover\,{\hat0}
 \\[2ex]
@@
 \begin{array}{lcl}
-\tpaste^A & : & \Cover{\den A} \todot \den A \\[1ex]
+\tpaste^A & : & \tCover\,{\den A} \todot \den A \\[1ex]
 \tpaste^P & = & \tpasteNf \\
@@
 
+\newcommand{\citeentry}[1]{\item\nocite{#1}\bibentry{#1}}
+
+\section{Literature}
+
+Further reading on normalization by evaluation in general:
+\begin{enumerate}
+
+\citeentry{bergerSchwichtenberg:lics91}
+
+\citeentry{jungTiuryn:tlca93}
+
+\citeentry{altenkirchHofmannStreicher:ctcs95}
+
+\citeentry{abel:habil}
+
+\end{enumerate}
+
+% \begin{itemize}
+% \item \citet{bergerSchwichtenberg:lics91}
+% \item \citet{jungTiuryn:tlca93}
+% \item \citet{altenkirchHofmannStreicher:ctcs95}
+% \item \citet{abel:habil}
+% \end{itemize}
+\noindent
+Literature dealing in particular with the case of disjunction /
+disjoint sums:
+
+\begin{enumerate}
+
+\citeentry{danvy:popl96}
+
+\citeentry{fioreSimpson:tlca99}
+
+\citeentry{altenkirchDybjerHofmannScott:lics01}
+
+\citeentry{balatDiCosmoFiore:popl04}
+
+\citeentry{altenkirchUustalu:flops04}
+
+\citeentry{barral:PhD}
+
+\citeentry{scherer:popl17}
+
+\end{enumerate}
+
+% \begin{itemize}
+% \item \citet{danvy:popl96}
+% \item \citet{fioreSimpson:tlca99}
+% \item \citet{altenkirchDybjerHofmannScott:lics01}
+% \item \citet{balatDiCosmoFiore:popl04}
+% \item \citet{altenkirchUustalu:flops04}
+
+% \item \citet{barral:PhD}
+% \item \citet{scherer:popl17}
+% \end{itemize}
+
 \section{Conclusion}
@@
 \paradot{Acknowledgments}  The presentation of this NbE proof and its
-Agda formalization benefitted greately from discussions with Christian
+Agda formalization benefited greatly from discussions with Christian
 Sattler.  Thanks to Jesper Cockx for listening to a prerun of the talk.
@@
 \bibliographystyle{abbrvnat}
-\bibliography{auto-nbeSum}
+%\bibliography{auto-nbeSum}
 

### 2018-07-23 notes/nbeSum.tex
ITC: fixes; write context consequently as subscripts

@@
   \reify[\Gamma]{A \imp B} f %\;\Delta\;\tau =
-    = \impIP {\reify[\Gamma.A]B{(f\;(\weak[A]{\id[\Gamma]})}\;\mbox{?})}
+    = \impIP {\reify[\Gamma.A]B{(f_{\Gamma.A}\;(\weak[A]{\id[\Gamma]})}\;\mbox{?})}
 \]
@@
   \reify[\Gamma]{A \imp B} f & = &
-    \impIP {\reify[\Gamma.A]B{(f\;(\weak[A]{\id[\Gamma]})}\;(\reflect[\Gamma.A]A{(\hyp\ttop)})}
+    \impIP {\reify[\Gamma.A]B{(f_{\Gamma.A}\;(\weak[A]{\id[\Gamma]})}\;(\reflect[\Gamma.A]A{(\hyp\ttop)})}
 \end{array}
@@
 \[
-\begin{array}{lll}
+\begin{array}{lcl}
    \teval^{A \imp B}_\Gamma & : & \TmG A \to \den {A \imp B} _\Gamma \\
-   \teval^{A \imp B}\,(\impI t)\;\Delta\;a & = & ? \\
+   \teval^{A \imp B}\,(\impI t)\,_\Delta\,a & = & ? \\
 \end{array}
@@
 \[
-  \teval^{A \imp B}_\Gamma  : \den \Gamma \todot \den {A \imp B}
+  \evalG{A \imp B}t  ~:~ \den \Gamma \todot \den {A \imp B}
 \]
@@
   \multicolumn 2 l {\tlookup^A_\Gamma} & : & \HypG A \to (\den \Gamma \todot \den A) \\
-  \lookup[\Gamma.A]A \ttop    & \Delta\,(\gamma,a) & = & a \\
-  \lookup[\Gamma.B]A {\pop x} &\Delta\,(\gamma,a) & = & \lookup[\Gamma]A x\, \Delta\, \gamma \\
+  \lookup[\Gamma.A]A \ttop    &_\Delta\,(\gamma,a) & = & a \\
+  \lookup[\Gamma.B]A {\pop x} &_\Delta\,(\gamma,a) & = & \lookup[\Gamma]A x\,_\Delta\, \gamma \\
 \end{array}
@@
   \evalG A {\hyp x} & = & \lookupG A x \\
-  \evalG {A \imp B} {\impI t}\,\Delta\,\gamma\,\Delta'\,(\tau : \Delta' \leq \Delta)\,a
-    & = & \eval[\Gamma.A]B t\,\Delta'\,(\den\Gamma_\tau\gamma,a) \\
-  \evalG {B} {\impE t u}\,\Delta\,\gamma & = &
-    \evalG {A \imp B} t\,\Delta\,\gamma\,\Delta\,\id[\Delta]\,
-      (\evalG A u\,\Delta\,\gamma) \\
+  \evalG {A \imp B} {\impI t}\,_\Delta\,\gamma\,_{\Delta'}\,(\tau : \Delta' \leq \Delta)\,a
+    & = & \eval[\Gamma.A]B t\,_{\Delta'}\,(\den\Gamma_\tau\gamma,a) \\
+  \evalG {B} {\impE t u}\,_\Delta\,\gamma & = &
+    \evalG {A \imp B} t\,_\Delta\,\gamma\,_\Delta\,\id[\Delta]\,
+      (\evalG A u\,_\Delta\,\gamma) \\
 \end{array}
@@
 \tnorm^A & : & \Tm A \todot \Nf A \\
-\norm[\Gamma] A t & = & \reify[\Gamma]A {(\eval[\Gamma] A t \,\Gamma\, \env[\Gamma])}
+\norm[\Gamma] A t & = & \reify[\Gamma]A {(\eval[\Gamma] A t \,_\Gamma\, \env[\Gamma])}
 \end{array}
@@
 two subtrees, one having a new hypothesis $C$, and one having a new
-hypothesis $C$.
+hypothesis $D$.
 
@@
 \]
-There is a redex $\impEP {\impI g} a$ separated by a $\torE$.  The latter
+There is a redex $\impEP {\impI g} a$ separated by an $\torE$.
+The function term $\orE {(\hyp x)} f {(\impI g)}$
 would naturally evaluate to a case tree with functions in the
@@
 \]
-that turns a case tree of functions into a function, which expects an
+that turns a case tree of functions into a function which expects an
 argument that will then be passed to all the functions in the case
@@
 \multicolumn 3 {l} {\paste[\Gamma]{A \imp B} c \;
-  \Delta \; (\tau : \Delta{\leq}\Gamma) \; (a : \den A _\Delta) \, = \,
-  (\tpaste^B \comp \mapC \phi \comp \Cover{\den {A \imp B}_\tau}\, c)
+  _\Delta \; (\tau : \Delta{\leq}\Gamma) \; (a : \den A _\Delta) \, = \,
+  (\paste[\Delta]B \comp \mapC[\Delta] \varphi \comp \Cover{\den {A \imp B}_\tau})\, c
 }
@@
 \]
-Here, we need a stronger version of functoriality of $\tCover$ which
-only requires an
+In the case of implication $A \imp B$,
+we need a stronger version of functoriality of $\tCover$ which
+only requires a
 \[
-  \phi : \prod_{\Phi} \prod_{\delta : \Phi \leq \Delta} \A_\Phi
+  \varphi : \prod_{\Phi} \prod_{\delta : \Phi \leq \Delta} \A_\Phi
              \to \B_\Phi
 \]
-to produce a $\tmapC_\Delta t\; \phi : \Cover\A_\Delta \to \Cover\B_\Delta$.
+to produce a $\mapC[\Delta]\varphi : \Cover\A_\Delta \to \Cover\B_\Delta$.
 In our case, $\A = \den{A \imp B}$ and $\B = \den{B}$ and
 \[
-  \phi\;\Phi\;\delta\;f = f\;\id[\Phi]\;(\den A _\delta\, a)
+  \varphi\,_\Phi\;\delta\;f = f\;\id[\Phi]\;(\den A _\delta\, a)
   .
@@
   \teval_\Gamma^A & : & \TmG A \to \den \Gamma \todot \den A \\[1ex]
-  \evalG C {\orE t u v} \Delta \; \gamma & = &
-    \tpaste^C \;
-     (\mapC {(\lambda \Phi\, (\tau  : \Phi \leq \Delta) \to [f,g])}\;c) \\[1ex]
+  \evalG C {\orE t u v} _\Delta \; \gamma & = &
+    \paste[\Delta]C
+     {(\mapC[\Delta]{\lambda\, _\Phi\, (\tau  : \Phi \leq \Delta) \to [f,g]}\;c)} \\[1ex]
   \multicolumn 1 r {\mbox{ where } c} & = &
-      \evalG{A \lor B} t\;\Delta\;\gamma \\
+      \evalG{A \lor B} t\;_\Delta\;\gamma \\
   \multicolumn 1 r {f\;a} & = &
-    \eval[\Gamma.A] C u \;\Phi\;(\den\Gamma_\tau\,\gamma, a) \\
+    \eval[\Gamma.A] C u \;_\Phi\;(\den\Gamma_\tau\,\gamma, a) \\
   \multicolumn 1 r {g\;b} & = &
-    \eval[\Gamma.B] C v \;\Phi\;(\den\Gamma_\tau\,\gamma, b) \\
+    \eval[\Gamma.B] C v \;_\Phi\;(\den\Gamma_\tau\,\gamma, b) \\
   \\

### 2019-02-04 fscd19/fscd19.tex
[ fscd19 ] Stub.

\documentclass[a4paper,USenglish,cleveref, autoref]{lipics-v2019}
%This is a template for producing LIPIcs articles.
%See lipics-manual.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling cleveref support, use "autoref"

% Disallowed packages:
% - enumitem
% - paralist
% - natbib
% - wrapfig

% Disallowed:
% - conditional inclusion of content
% - splitting over several files

% LIPICS enumerations
% - itemize
% - enumerate 1.
% - alphaenumerate (a)
% - romanenumerate (i)
% - bracketenumerate (1)

% LIPICS environments
% - lstlisting
% - theorem
% - lemma
% - corollary
% - proposition
% - exercise
% - definition
% - example
% - note and note*
% - remark and remark*
% - claim and claim*
% - proof and claimproof

% Features
% - env: claim and claimproof inside proof

% References to lemmata
% - \cref and \autoref (includes name, like Corollary 3)

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\bibliographystyle{plainurl}% the mandatory bibstyle

\title{Normalization by Evaluation for Call-by-Push-Value and Focused Lambda-Calculus}

\titlerunning{NbE for CBPV and Focusing}%optional, please use if title is longer than one line

\author{Andreas Abel}
  {Department of Computer Science and Engineering,
    Chalmers and Gothenburg University,
    Sweden \and
    \url{www.cse.chalmers.se/~abela}
  }
  {andreas.abel@gu.se}
  {https://orcid.org/0000-0003-0420-4492}
  {VR Grant 2014-04864 \emph{Termination
      Certificates for Dependently-Typed Programs and Proofs via
      Refinement Types}
  }
% TODO mandatory, please use full name;
% only 1 author per \author macro;
% first two parameters are mandatory, other parameters can be empty.
% Please provide at least the name of the affiliation and the country.
% The full address is optional

\author{Frederik Folkmar Ramcke}
  {Department of Computer Science and Engineering,
    Chalmers and Gothenburg University,
    Sweden
  }
  {}%email
  {}%orcid
  {}%funding

\authorrunning{A. Abel and F. F. Ramcke}%TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Andreas Abel and Frederik F. Ramcke}%TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\ccsdesc[100]{General and reference~General literature}
\ccsdesc[100]{General and reference}%TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

\keywords{Evaluation, Intuitionistc Propositional Logic,
  Lambda-Calculus, Monad, Normalization,
  Polarized Logic, Semantics}%TODO mandatory; please add comma-separated list of keywords

\category{}%optional, e.g. invited paper

\relatedversion{}%optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversion{A full version of the paper is available at \url{...}.}

\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

\acknowledgements{ The presentation of this NbE proof and its
Agda formalization benefited greatly from discussions with Christian
Sattler.}%optional

%\nolinenumbers %uncomment to disable line numbering

%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{Jakob Rehof and XXX}
\EventNoEds{2}
\EventLongTitle{Formal Structures for Computation and Deduction}
\EventShortTitle{FSCD 2019}
\EventAcronym{FSCD}
\EventYear{2019}
\EventDate{June 24--30, 2019}
\EventLocation{Dortmund, Germany}
\EventLogo{}
\SeriesVolume{}
\ArticleNo{}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\maketitle

\begin{abstract}
  We extend normalization by evaluation (NbE) to call-by-push-value
  (CBPV) and focused intutitionistic propositional logic.
  The developments have been partially formalized in Agda.
\end{abstract}

\section{Introduction}
\label{sec:intro}

\nocite{bergerSchwichtenberg:lics91}

\nocite{jungTiuryn:tlca93}

\nocite{altenkirchHofmannStreicher:ctcs95}

\nocite{abel:habil}

\nocite{danvy:popl96}

\nocite{fioreSimpson:tlca99}

\nocite{altenkirchDybjerHofmannScott:lics01}

\nocite{balatDiCosmoFiore:popl04}

\nocite{altenkirchUustalu:flops04}

\nocite{barral:PhD}

\nocite{scherer:popl17}

\section{Normalization by Evaluation for the Simply-Typed Lambda
  Calculus with Sums}
\label{sec:lambda}

\section{Call-By-Push Value}
\label{sec:cbpv}

\section{Focused Intuitionistic Propositional Logic}
\label{sec:fipl}

\section{Conclusion}
\label{sec:concl}

\bibliography{auto-fscd19}

% \appendix

\end{document}

### 2019-02-04 fscd19/fscd19.tex
[ fscd19 ] AST of STLC

@@
 %for enabling cleveref support, use "autoref"
+
+% \usepackage{dsfont} % black-board bold numbers, package doublestroke
+% LACKS: \mathds{0}
 
@@
 %\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
+
+\input{macros}
 
@@
 
+In this section, we review the normalization by evaluation (NbE)
+argument for the simply-typed lambda calculus with weak sums, setting
+the stage for the later sections.
+%
+We work in a type-theoretic meta-language, with the basic judgement
+$t : T$ meaning that object $t$ is an inhabitant of type $T$.
+Meta-theoretic types can be thought of as sets, thus, the colon $:$
+replaces the standard notion of elementhood $\in$, and we are free to
+reuse $\in$ for other purposes.
+
+%\paradot{Contexts and indices}
+We adapt a categorical aka de Bruijn style for the abstract syntax of
+terms, which we conceive as intrinsically well-typed.  In de Bruijn
+style, a context $\Gamma$ is just a snoc list of simple types $A$,
+meaning we write context extension as $\cext \Gamma A$, and the empty
+context as $\cempty$.  Membership \fbox{$A \in \Gamma$} and sublist
+relations \fbox{$\Gamma \sublist \Delta$} are given inductively by the
+following rules:
+\begin{gather*}
+  \nru{\tzero}{}{A \in \cext \Gamma A}
+\qquad
+  \nru{\tsuc}{A \in \Gamma}{A \in \cext \Gamma B}
+\qquad
+  \nru{\tdone}{}{\cempty \sublist \cempty}
+\qquad
+  \nru{\tlift}
+    {\Gamma \sublist \Delta}
+    {\cext \Gamma A \sublist \cext \Delta A}
+\qquad
+  \nru{\tweak}
+    {\Gamma \sublist \Delta}
+    {\Gamma \sublist \cext \Delta A}
+\end{gather*}
+We consider the rules as introductions of the indexed types
+$\_{\in}\_$ and $\_{\sublist}\_$ and the rule names as constructors.
+For instance, $\suc \tzero : A \in \cext {\cext \Gamma A} B$ for any
+$\Gamma$, $A$, and $B$; and if we read $\tsuc^n\,\tzero$ as unary
+number $n$, then $x : A \in \Gamma$ is exactly the (de Bruijn) index
+of $A$ in $\Gamma$.
+
+We can define
+\fbox{$\tid : \Gamma \sublist \Gamma$} and \fbox{$\_{\comp}\_ : \Gamma
+\sublist \Delta \to \Delta \sublist \Phi \to \Gamma \sublist \Phi$}
+by recursion, meaning that the (proof-relevant)
+sublist relation is reflexive and transitive.
+% The sublist relation is reflexive and transitive, the witnesses
+% \fbox{$\tid : \Gamma \sublist \Gamma$} and \fbox{$\_{\comp}\_ : \Gamma
+% \sublist \Delta \to \Delta \sublist \Phi \to \Gamma \sublist \Phi$}
+% have easy recursive definitions.
+Thus, lists $\Gamma$ form a category
+$\Cxt$ with morphisms $\tau : \Gamma \sublist \Delta$, and the category laws
+hold propositionally, e.g., we have $\id \comp \tau \equiv \tau$ in
+propositional equality for all morphisms $\tau$.
+
+The category $\Cxt$ allows us to consider $A{\in}\_$ as a presheaf over
+$\Cxt\op$ for any $A$, witnessed by \fbox{$\treindex : \Gamma \sublist
+  \Delta \to A \in \Gamma \to A \in \Delta$}, which is the morphism
+part of functor from $\Cxt$ to $\Set$, mapping object $\Gamma$ to the
+set $A \in \Gamma$ of the indices of $A$ in $\Gamma$.
+
+Simple types shall be given by the grammar
+\fbox{$S,T ::= P \mid S \To T \mid S \times T \mid \bone \mid S + T
+  \mid \bzero$}.
+Intrinsically well-typed lambda-terms, in abstract syntax,
+are just inhabitants $t$ of the type $A \dashv \Gamma$,
+inductively defined by the following rules.
+\begin{gather*}
+  \nru{\tvar}{A \in \Gamma}{A \dashv \Gamma}
+\qquad
+  \nru{\tabs}{B \from \cext \Gamma A}{A \To B \from \Gamma}
+\qquad
+  \nru{\tapp}
+    {A \To B \from \Gamma \qquad A \from \Gamma}
+    {B \from \Gamma}
+\\[2ex]
+  \nru{\tunit}{}{\bone \from \Gamma}
+\qquad
+  \nru{\tpair}
+    {A_1 \from \Gamma \qquad A_2 \from \Gamma}
+    {A_1 \times A_2 \from \Gamma}
+\qquad
+  \nru{\tprj_i}
+    {A_1 \times A_2 \from \Gamma}
+    {A_i \from \Gamma}
+\\[2ex]
+  \nru{\tinj_i}
+    {A_i \from \Gamma}
+    {A_1 + A_2 \from \Gamma}
+\qquad
+  \nru{\tcase}
+    {A_1 + A_2 \from \Gamma
+      \qquad C \from \cext \Gamma A_1
+      \qquad C \from \cext \Gamma A_2
+    }{C \from \Gamma}
+\qquad
+  \nru{\tabort}
+    {\bzero \from \Gamma}
+    {C \from \Gamma}
+\end{gather*}
+The skilled eye of the reader will immediately recognize the proof
+rules of intuitionistic propositional logic (IPL) under the
+Curry-Howard isomorphism.  Terms of type $A$ form a presheaf
+$A{\from}\_$ as witnessed by the standard weakening operation%
+\footnote{Here, $\tren$ is short for \emph{renaming}, but in a nameless
+calculus we should better speak of \emph{reindexing}, which could, a
+bit clumsily, also abbreviated to $\tren$.}
+\fbox{$\tren : \Gamma \sublist \Delta \to A \from \Gamma \to A \from \Delta$}
+defined by recursion over $t : A \from \Gamma$.
+
 \section{Call-By-Push Value}

### 2019-02-04 fscd19/fscd19.tex
[ fscd19 ] normal forms

@@
 In this section, we review the normalization by evaluation (NbE)
-argument for the simply-typed lambda calculus with weak sums, setting
+argument for the simply-typed lambda calculus (STLC) with weak sums, setting
 the stage for the later sections.
 %
-We work in a type-theoretic meta-language, with the basic judgement
+We work in a constructive type-theoretic meta-language, with the basic judgement
 $t : T$ meaning that object $t$ is an inhabitant of type $T$.
@@
 set $A \in \Gamma$ of the indices of $A$ in $\Gamma$.
+The associated naturality laws \fbox{$\treindex\,\tid\,x \equiv x$}
+and \fbox{$\treindex\,\tau_2\,(\treindex\,(\tau_1\,x)) \equiv
+\treindex\,(\tau_1 \comp \tau_2)\,x$} hold propositionally.
 
@@
 Intrinsically well-typed lambda-terms, in abstract syntax,
-are just inhabitants $t$ of the type $A \dashv \Gamma$,
+are just inhabitants $t$ of the type \fbox{$A \from \Gamma$},
 inductively defined by the following rules.
 \begin{gather*}
-  \nru{\tvar}{A \in \Gamma}{A \dashv \Gamma}
+  \nru{\tvar}{A \in \Gamma}{A \from \Gamma}
 \qquad
@@
 rules of intuitionistic propositional logic (IPL) under the
-Curry-Howard isomorphism.  Terms of type $A$ form a presheaf
+Curry-Howard isomorphism.  A term such as
+\(
+  \tabs\,(\tabs\,(\tpair\,
+    (\tvar\,(\tsuc\,\tzero))\,
+    (\tabs\,(\tapp\,(\tvar\,(\tsuc\,\tzero))\,(\tvar\,\tzero)))))
+\)
+could in concrete syntax be rendered as
+\(
+  \lambda x.\,\lambda y.\,(x,\,\lambda z.\,y\,z)
+  .
+\)
+We leave the exact connection to a printable
+syntax of the STLC to imagination of the reader,
+as we shall not be concerned with considering
+concrete terms in this article.
+
+Terms of type $A$ form a presheaf
 $A{\from}\_$ as witnessed by the standard weakening operation%
@@
 \fbox{$\tren : \Gamma \sublist \Delta \to A \from \Gamma \to A \from \Delta$}
-defined by recursion over $t : A \from \Gamma$.
+defined by recursion over $t : A \from \Gamma$,
+and naturality laws for $\tren$ analogously to $\treindex$.
 
+Normal forms%
+\footnote{
+There is also a stronger
+notion of normal form, requiring that two \emph{extensionally equal}
+lambda-terms, \ie, those that denote the
+same set-theoretical function, have the same normal form
+\cite{mitchell:foundations,altenkirchDybjerHofmannScott:lics01,scherer:popl17}.
+Such normal forms do not have a simple inductive definition, and we
+shall not consider them in this article.
+}
+are logically characterized as those fulfilling the
+\emph{subformula} property \cite{prawitz:natded,matthes:shortproofs}.
+Normal forms \fbox{$t : \Nf A \Gamma$} are mutually defined with
+\emph{neutral} normal forms \fbox{$t : \Ne A \Gamma$}.
+In the following inductive definition, we reuse the rule names from the
+term constructors.
+\begin{gather*}
+  \nru{\tvar}{A \in \Gamma}{\Ne A \Gamma}
+\qquad
+  \nru{\tabs}{\NfC B {\cext \Gamma A}}{\NfP {A \To B} \Gamma}
+\qquad
+  \nru{\tapp}
+    {\NeP {A \To B} \Gamma \qquad \Nf A \Gamma}
+    {\Ne B \Gamma}
+\qquad
+  \nru{\tne}{\Ne A \Gamma}{\Nf A \Gamma}
+\\[2ex]
+  \nru{\tunit}{}{\Nf \bone \Gamma}
+\qquad
+  \nru{\tpair}
+    {\Nf {A_1} \Gamma \qquad \Nf {A_2} \Gamma}
+    {\NfP {A_1 \times A_2} \Gamma}
+\qquad
+  \nru{\tprj_i}
+    {\NeP {A_1 \times A_2} \Gamma}
+    {\Ne {A_i} \Gamma}
+\\[2ex]
+  \nru{\tinj_i}
+    {\Nf {A_i} \Gamma}
+    {\NfP {A_1 + A_2} \Gamma}
+\quad
+  \nru{\tcase}
+    {\NeP {A_1 + A_2} \Gamma
+      \quad \NfC C {\cext \Gamma A_1}
+      \quad \NfC C {\cext \Gamma A_2}
+    }{\Nf C \Gamma}
+\quad
+  \nru{\tabort}
+    {\Ne \bzero \Gamma}
+    {\Nf C \Gamma}
+\end{gather*}
+The new rule $\tne$ embeds $\tNe$ into $\tNf$,
+\ie, allows us to forget neutrality.  Needless to say, $\tNe\,A$ and
+$\tNf\,A$ are presheaves, \ie, support reindexing with $\tren$ just as terms do.
+From a normal form we can extract the term via a overloaded function
+\fbox{$\tm\_ : \Nf A \Gamma \to A \from \Gamma$} and
+\fbox{$\tm\_ : \Ne A \Gamma \to A \from \Gamma$}
+that discards constructor $\tne$ but keeps all other constructors.
+This erasure function naturally commutes with reindexing, making it a
+natural transformation between the presheaves $\tNf\,A$ ($\tNe\,A$,
+resp.) and $A{\from}\_$.  We shall simply write, for instance,
+\fbox{$\tNf\,A \todot A{\from}\_$} for such presheaf morphisms.
+
+\emph{Proof normalization} is concerned with finding a normal form
+$u : \Nf A \Gamma$ for each derivation $t : A \from \Gamma$.  However,
+\emph{term normalization} additionally requires that $\tm u \cong t$
+with respect to a equational theory $\cong$ on terms.  Defining
+\fbox{$\NF t = \exists u \of \tNf A\,\Gamma.\; \tm u \cong t$}
+% for $t : A \from \Gamma$,
+we aim to construct a normalization function
+\[
+  \tnorm : (t : A \from \Gamma) \to \NF t
+\]
+that gives us for each term $t$ a $\cong$-equal normal form $u$.
+%\footnote{
+A more accurate presentation of $\tNF$ would have been
+$\sum_{u : \tNf A\,\Gamma} (\tm u \cong t)$ to make absolutely clear that we
+want get hold of the witness $u$ of the existential statement;
+however, we shall only use constructive
+existentials\footnote{Constructive existence in
+  contrast to classical existence and
+  \emph{mere existence} from Homotopy Type Theory which
+  requires propositional truncation.}
+in this article and thus stick to the notation $\exists$.
+%}
 \section{Call-By-Push Value}

### 2019-02-04 fscd19/fscd19.tex
[ fscd19 ] first intro to NbE

@@
 
-%\paradot{Contexts and indices}
+\subsection{Contexts and indices}
+
 We adapt a categorical aka de Bruijn style for the abstract syntax of
@@
 We can define
-\fbox{$\tid : \Gamma \sublist \Gamma$} and \fbox{$\_{\comp}\_ : \Gamma
+\fbox{$\tid : \Gamma \sublist \Gamma$} and \fbox{$\_{\rcomp}\_ : \Gamma
 \sublist \Delta \to \Delta \sublist \Phi \to \Gamma \sublist \Phi$}
@@
 % The sublist relation is reflexive and transitive, the witnesses
-% \fbox{$\tid : \Gamma \sublist \Gamma$} and \fbox{$\_{\comp}\_ : \Gamma
+% \fbox{$\tid : \Gamma \sublist \Gamma$} and \fbox{$\_{\rcomp}\_ : \Gamma
 % \sublist \Delta \to \Delta \sublist \Phi \to \Gamma \sublist \Phi$}
@@
 $\Cxt$ with morphisms $\tau : \Gamma \sublist \Delta$, and the category laws
-hold propositionally, e.g., we have $\id \comp \tau \equiv \tau$ in
+hold propositionally, e.g., we have $\id \rcomp \tau \equiv \tau$ in
 propositional equality for all morphisms $\tau$.
@@
 and \fbox{$\treindex\,\tau_2\,(\treindex\,(\tau_1\,x)) \equiv
-\treindex\,(\tau_1 \comp \tau_2)\,x$} hold propositionally.
+\treindex\,(\tau_1 \rcomp \tau_2)\,x$} hold propositionally.
+
+\subsection{STLC and its normal forms}
 
 Simple types shall be given by the grammar
-\fbox{$S,T ::= P \mid S \To T \mid S \times T \mid \bone \mid S + T
-  \mid \bzero$}.
+\fbox{$S,T ::= P \mid S \To T \mid S \times T \mid \tyone \mid S + T
+  \mid \tyzero$}.
 Intrinsically well-typed lambda-terms, in abstract syntax,
@@
 \\[2ex]
-  \nru{\tunit}{}{\bone \from \Gamma}
+  \nru{\tunit}{}{\tyone \from \Gamma}
 \qquad
@@
   \nru{\tabort}
-    {\bzero \from \Gamma}
+    {\tyzero \from \Gamma}
     {C \from \Gamma}
@@
 Normal forms%
-\footnote{
+\footnote{%
 There is also a stronger
@@
 \\[2ex]
-  \nru{\tunit}{}{\Nf \bone \Gamma}
+  \nru{\tunit}{}{\Nf \tyone \Gamma}
 \qquad
@@
   \nru{\tabort}
-    {\Ne \bzero \Gamma}
+    {\Ne \tyzero \Gamma}
     {\Nf C \Gamma}
@@
 \emph{term normalization} additionally requires that $\tm u \cong t$
-with respect to a equational theory $\cong$ on terms.  Defining
+with respect to a equational theory $\cong$ on terms.
+
+\subsection{Proof normalization}
+
+For didactical purposes, let us first focus on the simpler task of
+constructing a normalization function
+\fbox{$\tnorm : A \from \Gamma \to \Nf A \Gamma$}
+that does not guarantee any relation between its input term $t$ and
+output normal form $u$.
+% \emph{Normalization by evaluation} (NbE) decomposes normalization as
+% \fbox{\(
+%   \norm[\Gamma] A (t : A \from \Gamma)
+%   = \down[\Gamma] A \dent t {\idenv\Gamma}
+% \)}
+% which is
+% \emph{evaluation}
+% \nofbox{$\dent \_ \_ : (t : A \from \Gamma) \to \Den \Gamma \Delta \to
+%   \Den A \Delta$}
+% in the identity environment \nofbox{$\idenv\Gamma : \Den \Gamma \Gamma$}
+% followed by
+% \emph{reification}
+% \nofbox{$\downG A : \Den A \Gamma \to \Nf A \Gamma$}.
+\emph{Normalization by evaluation} (NbE) decomposes normalization as
+\fbox{\(
+%  \norm[\Gamma] A (t : A \from \Gamma)
+  \tnorm\, (t : A \from \Gamma)
+  = \down A \dent t {\idenv\Gamma}
+\)}
+which is
+\emph{evaluation}
+\nofbox{$\dent \_ \_ : (t : A \from \Gamma) \to \den \Gamma \todot
+  \den A$}
+in the identity environment \nofbox{$\idenv\Gamma : \den \Gamma \Gamma$}
+followed by
+\emph{reification}
+\nofbox{$\down A : \den A \todot \tNf A$}.
+
+\clearpage
+
+\subsection{Term normalization}
+
+Defining%
+\footnote{%
+A more accurate presentation of $\tNF$ would be
+$\sum_{u : \tNf A\,\Gamma} (\tm u \cong t)$
+in order to make absolutely clear that we
+want get hold of the witness $u$ of the existential statement;
+however, we shall only use constructive
+existentials
+in this article and thus stick to the notation $\exists$.
+We mean \emph{constructive existence} in
+  contrast to classical existence and
+  \emph{mere existence} from Homotopy Type Theory which
+  requires propositional truncation.
+}
 \fbox{$\NF t = \exists u \of \tNf A\,\Gamma.\; \tm u \cong t$}
@@
 that gives us for each term $t$ a $\cong$-equal normal form $u$.
-%\footnote{
-A more accurate presentation of $\tNF$ would have been
-$\sum_{u : \tNf A\,\Gamma} (\tm u \cong t)$ to make absolutely clear that we
-want get hold of the witness $u$ of the existential statement;
-however, we shall only use constructive
-existentials\footnote{Constructive existence in
-  contrast to classical existence and
-  \emph{mere existence} from Homotopy Type Theory which
-  requires propositional truncation.}
-in this article and thus stick to the notation $\exists$.
-%}
+
 \section{Call-By-Push Value}

### 2019-02-05 fscd19/fscd19.tex
[ fscd16 ] Presheaf constructions
Today I did not make any progress, my thoughts do not flow, I feel like wading through sticky slime.

@@
 
+\usepackage{bm} % bold math
 % \usepackage{dsfont} % black-board bold numbers, package doublestroke
@@
 $t : T$ meaning that object $t$ is an inhabitant of type $T$.
-Meta-theoretic types can be thought of as sets, thus, the colon $:$
-replaces the standard notion of elementhood $\in$, and we are free to
-reuse $\in$ for other purposes.
+However, to avoid confusion with object-level types such as the simple
+types of lambda calculus, we will refer to meta-level types as
+\emph{sets}.  Consequently, the colon $:$
+takes the role of elementhood $\in$ in set-theory, and we are free to
+reuse the symbol $\in$ for other purposes.
 
@@
 
-Simple types shall be given by the grammar
-\fbox{$S,T ::= P \mid S \To T \mid S \times T \mid \tyone \mid S + T
-  \mid \tyzero$}.
+Simple types shall be distinguished into positive types $P$ and
+negative types $N$, depending on their root type former.  Function
+($\To$) and product types ($\times$ and $\tyone$) are negative,
+while base types ($\tybase$) and sum types ($+$ and $\tyzero$) are positive.
+\[
+\begin{array}{lrl@{\qquad}l}
+  A,B,C & ::= & P \mid N & \mbox{simple types} \\
+  P     & ::= & \tyzero \mid A + B \mid \tybase & \mbox{positive types} \\
+  N     & ::= & \tyone \mid A \times B \mid A \To B
+  % P     & ::= & a \mid A + B \mid \tyzero & \mbox{positive types} \\
+  % N     & ::= & A \To B \mid A \times B \mid \tyone
+    & \mbox{negative types} \\
+\end{array}
+\]
+% Simple types shall be given by the grammar
+% % \fbox{$S,T ::= P \mid S \To T \mid S \times T \mid \tyone \mid S + T
+% %   \mid \tyzero$}.
+% \fbox{$A,B,C ::= a \mid A \To B \mid A \times B \mid \tyone \mid A + B
+%   \mid \tyzero$}.
+% Herein, $a$ ranges over atomic types aka base types.
 Intrinsically well-typed lambda-terms, in abstract syntax,
@@
 \emph{subformula} property \cite{prawitz:natded,matthes:shortproofs}.
-Normal forms \fbox{$t : \Nf A \Gamma$} are mutually defined with
+Normal forms \fbox{$t : \tNf A\, \Gamma$} are mutually defined with
 \emph{neutral} normal forms \fbox{$t : \Ne A \Gamma$}.
@@
 \qquad
-  \nru{\tne}{\Ne A \Gamma}{\Nf A \Gamma}
+  \nru{\tne}{\Ne \tybase \Gamma}{\Nf \tybase \Gamma}
 \\[2ex]
@@
     {\NeP {A_1 + A_2} \Gamma
-      \quad \NfC C {\cext \Gamma A_1}
-      \quad \NfC C {\cext \Gamma A_2}
-    }{\Nf C \Gamma}
+      \quad \NfC P {\cext \Gamma A_1}
+      \quad \NfC P {\cext \Gamma A_2}
+    }{\Nf P \Gamma}
 \quad
@@
     {\Ne \tyzero \Gamma}
-    {\Nf C \Gamma}
+    {\Nf P \Gamma}
 \end{gather*}
+These rules only allow the elimination of neutrals; guarantees
+the subformula property and prevents any kind of computational
+($\beta$) redex.
 The new rule $\tne$ embeds $\tNe$ into $\tNf$,
-\ie, allows us to forget neutrality.  Needless to say, $\tNe\,A$ and
+but only at base types $\tybase$.
+Further, case distintion via $\tcase$ and $\tabort$
+is restricted to positive types $P$.
+As a consequence, our normal forms are
+\emph{$\eta$-long}, meaning that any normal inhabitant of a negative
+type is a respective introduction ($\tabs$, $\tunit$, or $\tpair$).
+This justifies the attribute \emph{negative} for these types:
+the construction of their inhabitants proceeds mechanically, without
+any choices.  In contrast, to construct an inhabitant of a
+\emph{positive} type choice is required: whether case distinction is
+required, and which introduction to pick in the end ($\tinj_1$ or $\tinj_2$).
+
+Needless to say, $\tNe\,A$ and
 $\tNf\,A$ are presheaves, \ie, support reindexing with $\tren$ just as terms do.
-From a normal form we can extract the term via a overloaded function
+From a normal form we can extract the term via an overloaded function
 \fbox{$\tm\_ : \Nf A \Gamma \to A \from \Gamma$} and
@@
 \fbox{$\tNf\,A \todot A{\from}\_$} for such presheaf morphisms.
+(The point on the arrow is mnemoic for \emph{pointwise}.)
 
@@
 \nofbox{$\down A : \den A \todot \tNf A$}.
+The role of evaluation is to produce from a term the corresponding
+semantic (\ie, meta-theoretic) function, which is finally reified to a normal form.
+Since we are evaluating open terms $t$, we need to supply an environment
+$\idenv\Gamma$ which will map the free indices of $t$ to corresponding
+\emph{unknowns}.  To accomodate unknowns in the semantics, types $A$
+are mapped to presheaves $\den A$ (rather than just sets), and in
+particular each base type $\tybase$ is mapped to the presheaf
+$\tNe\,\tybase$ with the intention that the neutrals take the role of
+the unknowns.  The mapping $\up A : \tNe\,A \todot \den A$ from
+neutrals to unknowns is called \emph{reflection}, and defined mutually
+with reification by induction on type $A$.
+
+At this point, let us fix some notation for sets to prepare for some
+constructions of presheaves.  Let $\bone$ denote the unit set and
+$()$ its unique inhabitant, $\bzero$ the empty set and
+$\tmagic : \bzero \to T$ the \emph{ex falsum quod libet} elimination
+into any set $T$.  Given sets $S_1$ and $S_2$, their Cartesian product
+is written $S_1 \btimes S_2$ with projections $\pi_i : S_1 \btimes S_2
+\to S_i$, and their disjoint sum $S_1 \bplus S_2$ with injections
+$\iota_i : S_i \to S_1 \bplus S_2$ and elimination $[f_1,f_2] : S_1 \bplus
+S_2 \to T$ for arbitrary $f_i : S_i \to T$.
+
+Presheafs (co)products $\hat0$, $\hat1$, $\hatplus$, and $\hattimes$ are
+constructed pointwise, \eg, $\hat0\,\Gamma = \bzero$, and given two
+presheaves $\A$ and $\B$,
+$(\A \hatplus \B)\,\Gamma = \A\,\Gamma \bplus \B\,\Gamma$.  For the
+exponential of preasheaves, however, we need the \emph{Kripke function
+  space}
+$(\A \hatto \B)\,\Gamma = \forall \Delta.\, \Gamma \sublist \Delta
+\to \A\,\Delta \to \B\,\Delta$.
 

### 2019-02-05 notes/nbeSum.tex
[ travis ] xcolor redundant package
ITC: typo
[ fscd16 ] Presheaf constructions
Today I did not make any progress, my thoughts do not flow, I feel like wading through sticky slime.

@@
 \usepackage{natbib}
-\usepackage[usenames,dvipsnames]{xcolor}
+%\usepackage[usenames,dvipsnames]{xcolor}
 \usepackage{amsmath}
@@
 \end{center}
-However, the direct lifting of the functions space $\den{A \imp
+However, the direct lifting of the function space $\den{A \imp
   B}_\Gamma = \den A_\Gamma \to \den B_\Gamma$ does not work.  It does
@@
 where hole ``?'' should be filled with an element of $\den B_{\Gamma.A}$
-which intuitively should represent the new hypothesis $\top :
+which intuitively should represent the new hypothesis $\ttop :
 \Hyp[\Gamma.A]A$ introduced by $\timpI$.  Thus we need a natural

### 2019-02-06 fscd19/fscd19.tex
[ fscd19 ] Semantics of negative types.

@@
 constructions of presheaves.  Let $\bone$ denote the unit set and
-$()$ its unique inhabitant, $\bzero$ the empty set and
+$\ttempty$ its unique inhabitant, $\bzero$ the empty set and
 $\tmagic : \bzero \to T$ the \emph{ex falsum quod libet} elimination
@@
 $(\A \hatplus \B)\,\Gamma = \A\,\Gamma \bplus \B\,\Gamma$.  For the
-exponential of preasheaves, however, we need the \emph{Kripke function
+exponential of presheaves, however, we need the \emph{Kripke function
   space}
@@
 \to \A\,\Delta \to \B\,\Delta$.
+
+We will interpret simple types $A$ as corresponding presheaves
+$\den A$.  Let us start with the negative types, defining
+reflection $\up A : \tNe\, A \todot \den A$
+and reification $\down A : \den A \todot \tNf A$ along the way.
+\begin{gather*}
+\begin{array}{lcl}
+  \den   \tyone          & = & \hatone \\
+  \upG   \tyone \; u     & = & \ttempty \\
+  \downG \tyone \ttempty & = & \tunit \\
+\end{array}
+\qquad\qquad\qquad
+\begin{array}{lcl}
+  \den   {A \times B}       & = & \den A \hattimes \den B \\
+  \upG   {A \times B} \; u  & = & (\upG A (\prj1 u),\;\upG B (\prj2 u)) \\
+  \downG {A \times B} (a,b) & = & \tpair\,(\downG A a)\,(\downG B b) \\
+\end{array}
+\\[2ex]
+\begin{array}{lcl}
+  \den {A \To B}    & = & \den A \hatto \den B \\
+  \upG   {A \To B}    \; u \; (\tau : \Gamma \sublist \Delta) \; (a : \den A \Delta)
+     & = & \up[\Delta] B (\tapp\;(\tren\,\tau\, u)\;(\down[\Delta] A a)) \\
+  \downG {A \To B}    f        & = & \tabs\,(\down[\cext \Gamma A] B
+    (f\;\twk^A_\Gamma\;\tfresh^A_\Gamma))
+\end{array}
+\end{gather*}
+% We will interpret simple types $A$ as corresponding presheaves
+% $\den A$.  Let us start with the negative types:
+% \[
+% \begin{array}{lcl}
+%   \den \tyone       & = & \hatone \\
+%   \den {A \times B} & = & \den A \hattimes \den B \\
+%   \den {A \To B}    & = & \den A \hatto \den B \\
+% \end{array}
+% \]
+% Reflection $\up A : \tNe\, A \todot \den A$
+% and reification $\down A : \den A \todot \tNf A$
+% are defined mutually by induction on type $A$.
+% \begin{gather*}
+% \begin{array}{lcl@{\qquad}lcl}
+%   \upG   \tyone       u        & = & \ttempty &
+%   \downG \tyone       \ttempty & = & \tunit \\
+%   \upG   {A \times B} u        & = & (\upG A (\prj1 u),\;\upG B (\prj2 u)) &
+%   \downG {A \times B} (a,b)    & = & \tpair\,(\downG A a)\,(\downG B b) \\
+% \end{array}
+% \\
+% \begin{array}{lcl@{\qquad}lcl}
+%   \upG   {A \To B}    u \; (\tau : \Gamma \sublist \Delta) \; (a : \den A \Delta)
+%      & = & \up[\Delta] B (\tapp\;(\tren\;\tau\; u)\;(\down[\Delta] A a)) \\
+%   \downG {A \To B}    f        & = & \tabs\,(\down[\cext \Gamma A] B
+%     (f\,(\twk^A)\,(\tfresh^A_\Gamma)))
+% \end{array}
+% \end{gather*}
+In the equation for $\down{A \To B}$,
+the renaming $\twk : \Gamma \sublist \cext \Gamma A$,
+defined by $\twk = \tweak\,\tid$, makes room
+for a new variable of type $A$, which is reflected into $\den A$ by
+\[
+\begin{array}{lcl@{\qquad}lcl}
+  \tfresh : \den A (\cext \Gamma A) \\
+  \tfresh = \up[\cext \Gamma A] A (\tvar\;\tzero)
+  . \\
+\end{array}
+\]
+The ability to introduce fresh variables into a context, and to use
+semantic objects such as $f : \den{A \To B}\Gamma$ in a such extended
+context, is the reason for utilizing presheaves instead of just sets.
+
+Note also that in the equation for $\up{A \To B}$,
+the neutral $u : \Ne A \Gamma$ is
+transported into $\Ne A \Gamma$ via reindexing with
+$\tau : \Gamma \sublist \Delta$, in order to be applicable to the normal form
+$\down[\Delta]A a$
+reified from the semantic value $a$.
+
+A direct extension of our presheaf semantics to positive types cannot work.
 

### 2019-02-06 fscd19/fscd19.tex
[ fscd19 ] Why we need the monad

@@
 
-A direct extension of our presheaf semantics to positive types cannot work.
+A direct extension of our presheaf semantics to positive types cannot
+work.  For instance, with $\den \tyzero = \hat0$, simply
+$\up[\tyzero] \tyzero (\tvar\,\tzero) : \bzero$ would give us an
+inhabitant of the empty set, which means that reflection at the empty
+type would not be definable.  Similarily, the setting
+$\den{A + B} = \den A \hatplus \den B$ is refuted by
+% $\up[\tyone + \tyone]^{\tyone + \tyone} (\tvar\,\tzero) : \bone \bplus \bone$
+$\up[A + B]{A + B} (\tvar\,\tzero) : \den A (A+B) \bplus \den B (A+B)$
+which would require us to make a decision of whether $A$ holds or $B$ holds
+while only be given a hypothesis of type $A + B$.
+Not even the usual interpretation of base types $\den o = \tNe\,o$
+works in the presence of sums, as we would not be able to interpret
+the term
+$\tabs\,(\tcase\,(\tvar\,\tzero)\,(\tvar\,\tzero)\,(\tvar\,\tzero)) :
+(o+o) \To o$ in our semantics, as $\Ne o (o+o)$ is empty.
+What is needed are case distinctions on neutrals in the semantics,
+allowing us the elimination of positive hypotheses before producing a
+semantic value,
+and we shall capture this capability in a monad $\Cov$.
 

### 2019-02-07 fscd19/fscd19.tex
[ fscd19 ] Fragments in the introduction

@@
 
-\authorrunning{A. Abel and F. F. Ramcke}%TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
+\author{Christian Sattler}
+  {Department of Computer Science and Engineering,
+    Chalmers and Gothenburg University,
+    Sweden
+  }
+  {}%email
+  {}%orcid
+  {}%funding
 
-\Copyright{Andreas Abel and Frederik F. Ramcke}%TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+\authorrunning{A. Abel and F. F. Ramcke and C. Sattler}%TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
+
+\Copyright{Andreas Abel and Frederik F. Ramcke and Christian Sattler}%TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
 
@@
 \label{sec:intro}
+
+Weak coproducts.
+Weak sheaf (non-unique pasting).
+
+Contributions:
+\begin{enumerate}
+\item Abstract cover monad.
+\item NbE for CBPV and polarized lambda-calculus.
+\end{enumerate}
+
+CBPV differs from PolLambda in that hypotheses can be arbitrary
+positive formulas in CBPV, but only positive atoms or negative
+formulas in PolLambda.  This means that in CBPV, there is more choice
+on when and how to decompose a hypothesis into ones of smaller
+positive type, whereas in PolLambda there is no freedom: mandatory
+maximal splitting is dictated by the left inversion phase.
+
+Normal forms of CBPV thus retain individual shallow pattern matching
+constructs, while PolLambda has only deep, maximal pattern matching.
+
+Agda formalization:
+Sized types, nested data types.
 
@@
 The associated naturality laws \fbox{$\treindex\,\tid\,x \equiv x$}
-and \fbox{$\treindex\,\tau_2\,(\treindex\,(\tau_1\,x)) \equiv
+and \fbox{$\treindex\,\tau_2\,(\treindex\,\tau_1\,x) \equiv
 \treindex\,(\tau_1 \rcomp \tau_2)\,x$} hold propositionally.
@@
 semantic value,
-and we shall capture this capability in a monad $\Cov$.
+and we shall capture this capability in a monad $\C$.
+
+To recapitulate, a monad $\C$ on presheaves is first an endofunctor,
+i.e., it maps any presheaf $\A$ to the presheaf $\C\,\A$ and any
+presheaf morphism $f : \A \todot \B$ to the morphism
+$\tmap[\C]f: \C\,A \todot \C\,\B$ satisfying the functor laws for
+identity and composition.  Then, there are natural transformations
+$\treturn[\C] : \A \todot \C\,\A$ (unit) and
+$\tjoin[\C] : \C\,(\C\,A) \todot \C\,\A$ (concatenation) satisfying
+the monad laws.
+
+We are looking for a monad $\C$ that
+offers us these services:
+% \begin{enumerate}
+% \item Elimination of absurd neutrals
+% $\tabort^\C : \tNe\, \tyzero \todot \C\,\A$.
+% \item Case distinction on neutrals of sum type
+% $\tcase^\C_\Gamma : \Ne {(A + B)} \Gamma
+%   \to \C\,\A\,(\cext \Gamma A)
+%   \to \C\,\A\,(\cext \Gamma B)
+%   \to \C\,\A\,\Gamma$.
+% \item Run the monad to produce a normal form
+% $\trun^\C : \C\,(\Nf A) \todot \Nf A$.
+% \end{enumerate}
+\[
+\begin{array}{lcl@{\quad}l}
+\tabort^\C & : & \tNe\, \tyzero \todot \C\,\A
+  & \mbox{case on absurd neutral} \\
+\tcase^\C_\Gamma & : & \Ne {(A + B)} \Gamma
+  \to \C\,\A\,(\cext \Gamma A)
+  \to \C\,\A\,(\cext \Gamma B)
+  \to \C\,\A\,\Gamma
+  & \mbox{case on neutral} \\
+\trunNf[\C] & : & \C\,(\tNf A) \todot \tNf A
+  & \mbox{run the monad ($\tNf$ only)} \\
+\end{array}
+\]
+Using these services, we can complete our semantics:
+%\begin{gather*}
+\[
+\begin{array}{lcl}
+  \den \tybase & = & \C\,(\tNe\,\tybase) \\
+  \upG \tybase \; u & = & \treturn[\C]\,u \\
+  \downG \tybase \; m & = & \trunNf[\C]\,(\tmap[\C]\tne\;m) \\
+\end{array}
+\qquad
+\begin{array}{lcl}
+  \den \tyzero & = & \C\,\hat0 \\
+  \upG \tyzero \; u & = & \tabort^\C\,u \\
+  \downG \tyzero \; m & = & \trunNf[\C]\,(\tmap[\C]\tmagic\;m) \\
+\end{array}
+\]
+\[
+\begin{array}{lcl}
+  \den {A+B} & = & \C\,(\den A \hatplus \den B) \\
+  \upG {A+B} \; u & = & \tcase^\C\,u
+    \,(\treturn[\C]\,(\iota_1\,\tfresh^A_\Gamma))
+    \,(\treturn[\C]\,(\iota_2\,\tfresh^B_\Gamma))
+    \\
+  \downG {A+B} \; m & = & \trunNf[\C]\,
+    (\tmap[\C] [\tinj_1 \comp \down A,\; \tinj_2 \comp \down B]\;m) \\
+\end{array}
+\]
+%\end{gather*}
 

### 2019-02-08 fscd19/fscd19.tex
[ fscd19 ] run/paste and the Cov monad

@@
 
-\acknowledgements{ The presentation of this NbE proof and its
-Agda formalization benefited greatly from discussions with Christian
-Sattler.}%optional
+\acknowledgements{The first author took inspiration from some
+  unpublished notes by Thorsten Altenkirch titled
+\emph{Another topological completeness proof for intuitionistic logic}
+received by email on 16th March 2000.
+Thorsten in turn credits his inspiration to an ALF proof by Thierry
+Coquand.}
+% \acknowledgements{ The presentation of this NbE proof and its
+% Agda formalization benefited greatly from discussions with Christian
+% Sattler.}%optional
 
@@
 
+While the coproduct eliminations $\tcase$ and $\tabort$ are limited to
+normal forms of positive types $P$, their extension to negative types
+is admissible.
+\[
+\begin{array}{l@{~}lcl}
+  \multicolumn 4 l {\tabort^B ~:~ \tNe\, 0 \todot \tNf\,B} \\
+  \tabort^{\tyone}     & u & = & \tunit \\
+  \tabort^{A \times B} & u & = & \tpair\,(\tabort^A\,u)\,(\tabort^B\,u) \\
+  \tabort^{A \To B}    & u & = & \tabs\,(\tabort^B\,(\tren\,\twk^A\,u)) \\
+  \tabort^P           & u & = & \tabort\;u \\
+\end{array}
+\]
+$\tcase$ generalizes analogously, with a bit of care when weakening
+the branches.
+
 \emph{Proof normalization} is concerned with finding a normal form
@@
 semantic value,
-and we shall capture this capability in a monad $\C$.
+and we shall capture this capability in a strong monad $\C$
+which can \emph{cover} the cases.
 
@@
 presheaf morphism $f : \A \todot \B$ to the morphism
-$\tmap[\C]f: \C\,A \todot \C\,\B$ satisfying the functor laws for
+$\tmap[\C]f: \C\,\A \todot \C\;\B$ satisfying the functor laws for
 identity and composition.  Then, there are natural transformations
 $\treturn[\C] : \A \todot \C\,\A$ (unit) and
-$\tjoin[\C] : \C\,(\C\,A) \todot \C\,\A$ (concatenation) satisfying
+$\tjoin[\C] : \C\,(\C\,\A) \todot \C\,\A$ (concatenation) satisfying
 the monad laws.
 
-We are looking for a monad $\C$ that
+We are looking for a cover monad $\C$ that
 offers us these services:
@@
 \]
-Using these services, we can complete our semantics:
+We shall immediatly construct an instance of such a cover monad,
+the free cover monad
+$\Cov$ defined inductively by $\treturn[\Cov]$, $\tabort^\Cov$, and
+$\tcase^\Cov$.  One can visualize an element $c : \Cov\,\A\,\Gamma$ as
+binary case tree whose inner nodes ($\tcase$) are labeled by a neutral term
+of sum type $A+B$ and its two branches by the context extensions $A$
+and $B$, resp.  Leaves are either ($\tabort$)
+labeled by a neutral term of empty type $\tyzero$,
+or ($\treturn$) by an element of $\A$.  Functoriality amounts to
+replacing the labels of the $\treturn$-leaves, and the monadic bind replaces
+these leaves by further case trees.  (The uninspiring $\tjoin[\Cov]$
+flattens a 2-level case tree into a single one.)
+Finally $\trunNf[\Cov]$ is a simple recursion, replacing $\tcase^\Cov$
+and $\tabort^\Cov$
+by the $\tcase$ and $\tabort$ constructions on normal forms, and
+$\treturn[\Cov]$ by the identity.
+
+% the following constructions:
+% \begin{gather*}
+%   \nru{\treturn[\Cov]}{\A\,\Gamma}{\Cov\,\A\,\Gamma}
+% \end{gather*}
+
+Using the services of a generic cover monad $\C$, we can complete our semantics:
 %\begin{gather*}
@@
   \den   \tybase & = & \C\,(\tNe\,\tybase) \\
-  \upG \tybase \; u & = & \treturn[\C]\,u \\
-  \downG \tybase \; m & = & \trunNf[\C]\,(\tmap[\C]\tne\;m) \\
+  \upG   \tybase & = & \treturn[\C] \\
+  \downG \tybase & = & \trunNf[\C] \comp \tmap[\C]\tne \\
 \end{array}
@@
   \den   \tyzero & = & \C\,\hat0 \\
-  \upG \tyzero \; u & = & \tabort^\C\,u \\
-  \downG \tyzero \; m & = & \trunNf[\C]\,(\tmap[\C]\tmagic\;m) \\
+  \upG   \tyzero & = & \tabort^\C \\
+  \downG \tyzero & = & \trunNf[\C] \comp \tmap[\C]\tmagic \\
 \end{array}
 \]
+% \[
+% \begin{array}{lcl}
+%   \den \tybase & = & \C\,(\tNe\,\tybase) \\
+%   \upG \tybase \; u & = & \treturn[\C]\,u \\
+%   \downG \tybase \; m & = & \trunNf[\C]\,(\tmap[\C]\tne\;m) \\
+% \end{array}
+% \qquad
+% \begin{array}{lcl}
+%   \den \tyzero & = & \C\,\hat0 \\
+%   \upG \tyzero \; u & = & \tabort^\C\,u \\
+%   \downG \tyzero \; m & = & \trunNf[\C]\,(\tmap[\C]\tmagic\;m) \\
+% \end{array}
+% \]
 \[
@@
     \\
-  \downG {A+B} \; m & = & \trunNf[\C]\,
-    (\tmap[\C] [\tinj_1 \comp \down A,\; \tinj_2 \comp \down B]\;m) \\
+  \downG {A+B} & = & \trunNf[\C] \comp
+    \tmap[\C] [\tinj_1 \comp \down A,\; \tinj_2 \comp \down B] \\
 \end{array}
 \]
+% \[
+% \begin{array}{lcl}
+%   \den {A+B} & = & \C\,(\den A \hatplus \den B) \\
+%   \upG {A+B} \; u & = & \tcase^\C\,u
+%     \,(\treturn[\C]\,(\iota_1\,\tfresh^A_\Gamma))
+%     \,(\treturn[\C]\,(\iota_2\,\tfresh^B_\Gamma))
+%     \\
+%   \downG {A+B} \; m & = & \trunNf[\C]\,
+%     (\tmap[\C] [\tinj_1 \comp \down A,\; \tinj_2 \comp \down B]\;m) \\
+% \end{array}
+% \]
 %\end{gather*}
+
+All semantic types fulfill the weak sheaf condition aka weak pasting,
+meaning there is
+a natural transformation $\trun^A : \C\,\den A \todot \den A$ for any simple
+type $A$.  In other words, we can \emph{run} the monad, pushing its
+effects into $\den A$.  We proceed by induction on $A$.
+Positive types $P$ are already monadic,
+and $\trun^P$ is simply the $\tjoin$ of monad $\C$.
+At negative types we can recurse pointwise at a smaller type,
+exploiting that values of negative types are essentially
+(finite or infinite) tuples.
+\[
+\begin{array}{l@{~}l@{~~}c@{~~}l@{\qquad}l@{~~}c@{~~}l}
+  \multicolumn 7 l {\trun^A ~:~ \C \den A \todot \den A} \\
+  \trun^{\tyone} & c & = & \ttempty
+    &
+  \trun^{\tyzero} & = & \tjoin[\C]
+    \\
+  \trun^{A \times B} & c & = &
+      (\trun^A\,(\tmap[\C]\;\pi_1\;c),\
+       \trun^B\,(\tmap[\C]\;\pi_2\;c))
+    &
+  \trun^{A + B}   & = & \tjoin[\C]
+  \\
+  \trun^{A \To B} & c \; \tau \; a & = & \trun^B\,
+      (\stmap[\C]\;(\lambda\;\tau'\,f.\;f\;\tid\;(\tren\;\tau'\;a))\;(\tren\;\tau\;c))
+    &
+  \trun^{o}      & = & \tjoin[\C]
+\\
+\end{array}
+% \begin{array}{lcl@{\quad}l@{~}lcl}
+%   \trun^A & : & \C \den A \todot \den A \\
+%   \trun^{\tyzero} & = & \tjoin[\C]
+%     & \trun^{\tyone} & c & = & \ttempty \\
+%   \trun^{A + B}   & = & \tjoin[\C]
+%     & \trun^{A \times B} & c & = &
+%       (\trun^A\,(\tmap[\C]\;\pi_1\;c),\
+%        \trun^B\,(\tmap[\C]\;\pi_2\;c))\\
+%   \trun^{o}      & = & \tjoin[\C]
+%     & \trun^{A \To B} & c \; \tau \; a & = & \trun^B\,
+%       (\stmap[\C]\;(\lambda\;\tau'\,f.\;f\;\tid\;(\tren\;\tau'\;a))\;(\tren\;\tau\;c))
+% \\
+% \end{array}
+\]
+For the case of function types $A \To B$, we require the monad $\C$ to be
+\emph{strong}, which amounts to having
+$\stmap[\C]_\Gamma\, \ell : \C\A\,\Gamma \to \C\,\B\,\Gamma$ already for
+a local presheaf morphism $\ell : (\A \hatto \B)\Gamma$.
+The typings are $c : \C \den{A \To B} \Gamma$, $\tau : \Gamma \sublist
+\Delta$, $a : \den A \Delta$, and now we want to apply every function
+$f : \den{A \To B}$ in the cover $c$ to argument $a$.  Clearly,
+$\tmap[\C]$ is not applicable since it would expect a
+global presheaf morphism
+$\den{A \to B} \todot \den B$, \ie, something that works in \emph{any}
+context.  However, applying to $a : \den A \Delta$ can only work in
+context $\Delta$ or any extension $\tau' : \Delta \sublist \Phi$,
+since we can transport $a$ to such a $\Phi$ via
+$a' := \tren\; \tau'\; a : \den A \Phi$ but not to a context unrelated to
+$\Delta$.  We obtain our input to $\trun^B$ of type $\C \den B \Gamma$
+as an instance of $\stmap[\C]_\Gamma$ applied to the local presheaf morphism
+$(\lambda \; \tau'\, f. \  f\;\tid\;a') :
+\Delta \sublist \Phi \to  \den{A \To B}\Phi \to \den B \Phi$
+% (\tau' : \Delta \sublist \Phi) (f : \den{A \To B}\Phi) \to \den B \Phi$
+and the transported cover $\tren\;\tau\;c : \C\den{A \To B}\Delta$.
 

### 2019-02-08 fscd19/fscd19.tex
[ fscd19 ] started on evaluation function

@@
 \]
-We shall immediatly construct an instance of such a cover monad,
+To make things concrete,
+we shall immediately construct an instance of such a cover monad:
 the free cover monad
@@
 
+We extend the type interpretation pointwise to contexts, \ie,
+$\den \cempty = \hatone$ and
+$\den{\cext \Gamma A} = \den \Gamma \hattimes \den A$
+and obtain a natural projection function from the semantic enironments:
+\[
+\begin{array}{lcl}
+  \tlookup\,(x : A \in \Gamma) & : &  \den \Gamma \todot \den A \\
+  \tlookup\;\tzero     & = & \pi_2 \\
+  \tlookup\;(\tsuc\,x) & = & \tlookup\;x \comp \pi_1 \\
+\end{array}
+% \begin{array}{lcl}
+%   \tlookup & : & A \in \Gamma \to \den \Gamma \todot \den A \\
+%   \tlookup\;\tzero     & = & \pi_2 \\
+%   \tlookup\;(\tsuc\,x) & = & \tlookup\;x \comp \pi_1 \\
+% \end{array}
+\]
+The evaluation function
+\fbox{$\dent {t : A \from \Gamma} \_ : \den \Gamma \todot \den A$}
+can now be defined by recursion on $t$.
+Herein the environment $\gamma$ lives in $\den \Gamma \Delta$, thus, $\dent t \gamma : \den A \Delta$.
+% Herein, $\gamma : \den \Gamma \Delta$, thus, $\dent t \gamma : \den A \Delta$.
+\[
+\begin{array}{lcl}
+  \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
+  \dent{\tabs\,(t : B \from \cext \Gamma A)}\gamma & = &
+     \lambda\, (\tau : \Delta \sublist \Phi)\, (a : \den A \Phi). \
+        \dent t {(\tren\,\tau\,\gamma,\ a)} \\
+  \dent{\tapp\,t\,u} \gamma & = &
+     \dent t \gamma\;\tid\;\dent u \gamma \\
+  % \dent{\tapp\,(t : A \To B \from \Gamma)\,u} \gamma & = &
+  %    \dent t \gamma\;\tid\;\dent u \gamma \\
+  \dent{\tunit}\gamma & = & \ttempty \\
+  \dent{\tpair\,t_1\,t_2}\gamma & = &
+    (\dent{t_1}\gamma,\ \dent{t_2}\gamma) \\
+  \dent{\prj i t}\gamma & = & \pi_i \dent t \gamma \\
+  \dent{\inj i t}\gamma & = & \iota_i \dent t \gamma \\
+  \dent{\tcase\,t\,u\,v}\gamma & = & \\
+  \dent{\tabort\,t}\gamma & = & \\
+\end{array}
+\]
+
 \clearpage

### 2019-02-08 fscd19/fscd19.tex
[ fscd19 ] finished proof normalization for IPL

@@
 
-\title{Normalization by Evaluation for Call-by-Push-Value and Focused Lambda-Calculus}
+\title{Normalization by Evaluation for Call-by-Push-Value and Polarized Lambda-Calculus}
 
@@
 
-\author{Frederik Folkmar Ramcke}
-  {Department of Computer Science and Engineering,
-    Chalmers and Gothenburg University,
-    Sweden
-  }
-  {}%email
-  {}%orcid
-  {}%funding
+% \author{Frederik Folkmar Ramcke}
+%   {Department of Computer Science and Engineering,
+%     Chalmers and Gothenburg University,
+%     Sweden
+%   }
+%   {}%email
+%   {}%orcid
+%   {}%funding
 
@@
 
-\authorrunning{A. Abel and F. F. Ramcke and C. Sattler}%TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
+\authorrunning{A. Abel %and F. F. Ramcke
+and C. Sattler}%TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
 
-\Copyright{Andreas Abel and Frederik F. Ramcke and Christian Sattler}%TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+\Copyright{Andreas Abel %and Frederik F. Ramcke
+ and Christian Sattler}%TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
 
@@
 
-\keywords{Evaluation, Intuitionistc Propositional Logic,
+\keywords{Evaluation, Intuitionistic Propositional Logic,
   Lambda-Calculus, Monad, Normalization,
@@
 \begin{abstract}
-  We extend normalization by evaluation (NbE) to call-by-push-value
-  (CBPV) and focused intutitionistic propositional logic.
-  The developments have been partially formalized in Agda.
+\input{abstract.txt}
 \end{abstract}
@@
 a local presheaf morphism $\ell : (\A \hatto \B)\Gamma$.
-The typings are $c : \C \den{A \To B} \Gamma$, $\tau : \Gamma \sublist
-\Delta$, $a : \den A \Delta$, and now we want to apply every function
+The typings are
+$c : \C \den{A \To B} \Gamma$ and
+$\tau : \Gamma \sublist \Delta$ and
+$a : \den A \Delta$, and now we want to apply every function
 $f : \den{A \To B}$ in the cover $c$ to argument $a$.  Clearly,
@@
 The evaluation function
-\fbox{$\dent {t : A \from \Gamma} \_ : \den \Gamma \todot \den A$}
+\fbox{$\denty {t : A \from \Gamma}  : \den \Gamma \todot \den A$}
+% \fbox{$\dent {t : A \from \Gamma} \_ : \den \Gamma \todot \den A$}
 can now be defined by recursion on $t$.
@@
   \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
-  \dent{\tabs\,(t : B \from \cext \Gamma A)}\gamma & = &
-     \lambda\, (\tau : \Delta \sublist \Phi)\, (a : \den A \Phi). \
+  \dent{\tabs\,(t : B \from \cext \Gamma A)}\gamma \;
+     (\tau : \Delta \sublist \Phi)\, (a : \den A \Phi) & = &
         \dent t {(\tren\,\tau\,\gamma,\ a)} \\
+  % \dent{\tabs\,(t : B \from \cext \Gamma A)}\gamma & = &
+  %    \lambda\, (\tau : \Delta \sublist \Phi)\, (a : \den A \Phi). \
+  %       \dent t {(\tren\,\tau\,\gamma,\ a)} \\
   \dent{\tapp\,t\,u} \gamma & = &
@@
   \dent{\inj i t}\gamma & = & \iota_i \dent t \gamma \\
-  \dent{\tcase\,t\,u\,v}\gamma & = & \\
-  \dent{\tabort\,t}\gamma & = & \\
+  \dent{\tabort\,u}\gamma & = & \denabort\;\dent u \gamma \\
+  \dent{\tcase\,(u : A_1 + A_2 \from \Gamma)\;t_1\;t_2}\gamma & = &
+     \dencase\;\dent u \gamma\;f_1\;f_2
+    \ \mbox{ \textbf{where}} \\
+    \multicolumn 1 r {f_i\; (\tau : \Delta \sublist \Phi)\,(a_i : \den {A_i} \Phi)}
+        & = &
+        \dent {t_i} {(\tren\;\tau\;\gamma,\ {a_i})}
+   % \
+   %  (\lambda\; (\tau : \Delta \sublist \Phi)\,(b : \den A \Phi).\
+   %      \dent v {(\tren\;\tau\;\gamma,\ b)})
+    \\
 \end{array}
 \]
+The coproduct eliminations $\denabort$ and $\dencase$
+targeting an arbitrary semantic type $\den B$ are definable thanks to
+the weak sheaf property, \ie, the presence of pasting via $\trun^B$
+for any type $B$, and strong functoriality of $\C$.
+\[
+\begin{array}{lcl}
+  \denabort^B & : & \den \tyzero \todot \den B \\
+  \denabort^B & = & \trun^B \comp \tmap[\C]\,\tmagic
+\\[2ex]
+  \dencase^B_\Gamma  & : & \den{A_1 + A_2}\Gamma
+    \to \den{A_1 \To B}\Gamma
+    \to \den{A_2 \To B}\Gamma
+    \to \den{B}\Gamma \\
+  \dencase^B_\Gamma\;c\;f_1\;f_2 & = &
+    \trun^B (\stmap[\C]\;(\lambda\,\tau.\ [f_1\,\tau,\ f_2\,\tau])\;c) \\
+\end{array}
+\]
+To complete the normalization function
+$\tnorm\,(t : A \from \Gamma) = \downG A \dent t {\idenv \Gamma}$
+we define the identity environment $\idenv \Gamma : \den \Gamma \Gamma$,
+which maps each free index to its corresponding unknown in the
+semantics, by recursion on $\Gamma$:
+\[
+\begin{array}{lcl}
+  \idenv \cempty & = & \ttempty \\
+  \idenv {\cext \Gamma A} & = & (\tren\;\twk^A\,\idenv\Gamma, \  \tfresh^A_\Gamma)\\
+\end{array}
+\]
+
+\subsection{Continuation monad}
 

### 2019-02-09 fscd19/fscd19.tex
[ fscd19 ] normal forms of polarized lambda

@@
 the free cover monad
-$\Cov$ defined inductively by $\treturn[\Cov]$, $\tabort^\Cov$, and
-$\tcase^\Cov$.  One can visualize an element $c : \Cov\,\A\,\Gamma$ as
+$\Cov$ defined inductively by $\treturn[\Cov]$, $\tabort^{\Cov}$, and
+$\tcase^{\Cov}$.  One can visualize an element $c : \Cov\,\A\,\Gamma$ as
 binary case tree whose inner nodes ($\tcase$) are labeled by a neutral term
@@
 flattens a 2-level case tree into a single one.)
-Finally $\trunNf[\Cov]$ is a simple recursion, replacing $\tcase^\Cov$
-and $\tabort^\Cov$
+Finally $\trunNf[\Cov]$ is a simple recursion, replacing $\tcase^{\Cov}$
+and $\tabort^{\Cov}$
 by the $\tcase$ and $\tabort$ constructions on normal forms, and
@@
 
+Polarized lambda calculus \cite{espiritoSanto:entcs17}
+divides types deeply into positive and negative.
+
+\[
+\begin{array}{lllrl@{\qquad}l}
+  \PTy & \ni & P,Q & ::= & \patom
+    \mid \tyzero \mid P_1 + P_2
+    \mid \tyone \mid P_1 \times P_2
+    \mid \Thunk N
+  & \mbox{positive type}
+\\
+  \NTy & \ni & N,M & ::= & \natom
+    \mid \top \mid N_1 \ntimes N_2
+    \mid P \To N
+    \mid \Comp P
+  & \mbox{negative type}
+\\
+  \HTy & \ni & X,Y & ::= & \patom \mid P
+  & \mbox{hypothesis type}
+\\
+\end{array}
+\]
+Espirito Santo's version actually does not contain the positive
+product $\times$, however, it appears in Zeilberger
+\cite{zeilberger:PhD}.
+
+Polarized lambda-calculus is a focused calculus, it eagerly employs
+invertible rules: the introduction rules for negative types and the
+elimination rules for positive types.  As a consequence of the latter,
+variables are either of atomic or negative type ($\HTy$).
+Contexts $\Gamma,\Delta$ are lists of $\HTy$s.
+%\fbox{$X ::= \patom \mid P$}.
+
+To add a variable of positive type $P$ to the context, we need to
+break it apart until only atoms and negative bits remain.  This is
+performed by maximal pattern matching, called the left-invertible
+phase of focalization.
+We express maximal pattern matching on $P$ as a strong functor
+$\Add P$ in the category of presheaves
+mapping a presheaf $\J$ (``judgement'') to $\Add P \J$
+and a presheaf morphism $f : (\J \hatto \K)\Gamma$ to
+$\stmap[\Add P]_\Gamma f : \AddHyp P \J \Gamma \to \AddHyp P \K \Gamma$.
+For arbitrary $\J$ and $\Gamma$, the set \fbox{$\AddHyp P \J \Gamma$} is
+inductively constructed by the following rules:
+\begin{gather*}
+  \nru{\phyp}
+      {\J\,(\cext \Gamma \patom)}
+      {\AddHyp {\patom} \J \Gamma}
+\qquad
+  \nru{\nbranch}
+      {}
+      {\AddHyp {\tyzero} \J \Gamma}
+\qquad
+  \nru{\bbranch}
+      {\AddHyp {P_1} \J \Gamma \qquad
+       \AddHyp {P_2} \J \Gamma }
+      {\AddHyp {P_1 + P_2} \J \Gamma}
+\\[2ex]
+  \nru{\nhyp}
+      {\J\,(\cext \Gamma N)}
+      {\AddHyp {\Thunk N} \J \Gamma}
+\qquad
+  \nru{\nsplit}
+      {\J\,\Gamma}
+      {\AddHyp {\tyone} \J \Gamma}
+\qquad
+  \nru{\bsplit}
+      {\AddHyp {P_1} {(\Add{P_2}\,\J)} \Gamma}
+      {\AddHyp {P_1 \times P_2} \J \Gamma}
+\end{gather*}
+Note the recursive occurrence of $\Add{P_2}$ as argument to
+$\Add{P_1}$ in $\bsplit$, which makes $\Add P$ a nested datatype
+\cite{bird:nested}.  Agda supports such nested inductive types; but
+note that $\Add{P}$ is uncontroversial, since it could also be defined
+by recursion on $P$.   It is tempting to name $\bsplit$
+``$\tjoin$'' and $\nsplit$ ``$\treturn$''
+since $\Add P$ is a graded monad on the monoid $(1,\times)$
+of product types;
+however, this coincidence shall not matter for our further considerations.
+
+Assuming a family $\T\,N\,\Gamma$ of terms of negative type $N$ in
+context $\Gamma$, values \fbox{$v : \PValT P \Gamma$}
+of positive type $P$ are constructed by the
+following rules:
+\begin{gather*}
+  \nru{\pvar}{P \in \Gamma}{\PValT P \Gamma}
+\qquad
+  \nru{\tthunk}{\T\,N\,\Gamma}{\PValTP{\Thunk\,N}\Gamma}
+\\[2ex]
+  \nru{\punit}{}{\PValT{\tyone}\Gamma}
+\qquad
+  \nru{\ppair}
+      {\PValT{P_1}\Gamma \qquad \PValT{P_2}\Gamma}
+      {\PValTP{P_1 \times P_2}\Gamma}
+\qquad
+  \nru{\tinj_i}{\PValT{P_i}\Gamma}{\PValTP{P_1 + P_2}\Gamma}
+\end{gather*}
+Negative hypotheses can be eliminated by projections and application
+to values, as described by the following inductive family
+\fbox{$\NeT N \Gamma$}:
+\begin{gather*}
+  \nru{\nvar}{N \in \Gamma}{\NeT N \Gamma}
+\qquad
+  \nru{\tprj_i}
+      {\NeTP {N_1 \ntimes N_2} \Gamma}
+      {\NeT {N_i} \Gamma}
+\qquad
+  \nru{\tapp}
+      {\NeTP {P \To N} \Gamma \qquad \PValT P \Gamma}
+      {\NeT N \Gamma}
+\end{gather*}
+% Negative hypotheses can be eliminated by projections and application
+% to values $v : \V\,P\,\Gamma$ which we leave abstract in the following
+% inductive definition of \fbox{$\Ne[\V] N \Gamma$}:
+% \begin{gather*}
+%   \nru{\nvar}{N \in \Gamma}{\Ne[\V] N \Gamma}
+% \qquad
+%   \nru{\tprj_i}
+%       {\NeP[\V] {N_1 \ntimes N_2} \Gamma}
+%       {\Ne[\V] {N_i} \Gamma}
+% \qquad
+%   \nru{\tapp}
+%       {\NeP[\V] {P \To N} \Gamma \qquad \V\,P\,\Gamma}
+%       {\Ne[\V] N \Gamma}
+% \end{gather*}
+
+% decomposing
+% decomposes
+% \emph{values}, meaning inhabitants of positive types.
+An elimination may either result in a negative atom $\natom$ or a
+positive type $P$, which can be further taken apart by pattern
+matching.  Pattern matching on neutrals is organized in the
+inductively defined cover monad \fbox{$\Cov[\T]$}.
+\begin{gather*}
+  \nru{\treturn}{\J\,\Gamma}{\CoverT \J \Gamma}
+\qquad
+  \nru{\tmatch}
+      {\NeTP {\Comp\,P} \Gamma \qquad \AddHypP P {\Cov[\T]\J} \Gamma}
+      {\CoverT \J \Gamma}
+\end{gather*}
+
+Finally, normal forms of negative types are defined as inductive
+family \fbox{$\Nf N \Gamma$}.  It relies on the previous definitions
+with instance $\T = \tNf$.
+\begin{gather*}
+  \nru{\tne}
+      {\Cov[\tNf](\tNe[\tNf]\,\natom)\,\Gamma}
+      {\Nf {\natom} \Gamma}
+\qquad
+  \nru{\tval}
+      {\Cov[\tNf](\tVal[\tNf]\,P)\,\Gamma}
+      {\NfP {\Comp\,P} \Gamma}
+\\[2ex]
+  \nru{\npair}
+      {\Nf {N_1} \Gamma \qquad \Nf {N_2} \Gamma}
+      {\NfP {N_1 \ntimes N_2} \Gamma}
+\qquad
+  \nru{\tabs}
+      {\AddHypP P {\tNf\, N} \Gamma}
+      {\NfP {P \To N} \Gamma}
+\end{gather*}
+
 \section{Conclusion}

### 2019-02-09 fscd19/fscd19.tex
[ fscd19 ] Mutual instead of parametrized def of focused syntax

@@
 
+In the following, we mutually define the inductive families of
+presheaves $\tVal\,P$,
+$\tNe\,N$, $\Cov\,\J$,
+and $\tNf\,N$.
+
+Values \fbox{$v : \PVal P \Gamma$}
+of positive type $P$ are constructed by the
+following rules:
+\begin{gather*}
+  \nru{\pvar}{P \in \Gamma}{\PVal P \Gamma}
+\qquad
+  \nru{\tthunk}{\Nf N \Gamma}{\PValP{\Thunk\,N}\Gamma}
+\\[2ex]
+  \nru{\punit}{}{\PVal{\tyone}\Gamma}
+\qquad
+  \nru{\ppair}
+      {\PVal{P_1}\Gamma \qquad \PVal{P_2}\Gamma}
+      {\PValP{P_1 \times P_2}\Gamma}
+\qquad
+  \nru{\tinj_i}{\PVal{P_i}\Gamma}{\PValP{P_1 + P_2}\Gamma}
+\end{gather*}
+Negative hypotheses can be eliminated by projections and application
+to values, as described by the following inductive family
+\fbox{$\Ne N \Gamma$}:
+\begin{gather*}
+  \nru{\nvar}{N \in \Gamma}{\Ne N \Gamma}
+\qquad
+  \nru{\tprj_i}
+      {\NeP {N_1 \ntimes N_2} \Gamma}
+      {\Ne {N_i} \Gamma}
+\qquad
+  \nru{\tapp}
+      {\NeP {P \To N} \Gamma \qquad \PVal P \Gamma}
+      {\Ne N \Gamma}
+\end{gather*}
+An elimination may either result in a negative atom $\natom$ or a
+positive type $P$, which can be further taken apart by pattern
+matching.  Pattern matching on neutrals is organized in the
+inductively defined strong monad \fbox{$\Cov$}.
+\begin{gather*}
+  \nru{\treturn}{\J\,\Gamma}{\Cover \J \Gamma}
+\qquad
+  \nru{\tmatch}
+      {\NeP {\Comp\,P} \Gamma \qquad \AddHypP P {\Cov\,\J} \Gamma}
+      {\Cover \J \Gamma}
+\end{gather*}
+Finally, normal forms of negative types are defined as inductive
+family \fbox{$\Nf N \Gamma$}.
+\begin{gather*}
+  \nru{\tne}
+      {\CoverP {\tNe\,\natom} \Gamma}
+      {\Nf {\natom} \Gamma}
+\quad
+  \nru{\tval}
+      {\CoverP {\tVal\,P}  \Gamma}
+      {\NfP {\Comp\,P} \Gamma}
+\quad
+  \nru{\npair}
+      {\Nf {N_1} \Gamma \qquad \Nf {N_2} \Gamma}
+      {\NfP {N_1 \ntimes N_2} \Gamma}
+\quad
+  \nru{\tabs}
+      {\AddHypP P {\tNf\, N} \Gamma}
+      {\NfP {P \To N} \Gamma}
+\end{gather*}
+
+\clearpage
+
+\subsection{Parametrized syntax (superseded)}
+
 Assuming a family $\T\,N\,\Gamma$ of terms of negative type $N$ in

### 2019-02-09 fscd19/fscd19.tex
[ fscd19 ] sized join for Cov monad

@@
 \end{gather*}
+All these inductive families are presheaves, due to factored
+presentation using $\Add P$ and $\Cov$ the proof is not a simple
+mutual induction.  Yet, in Agda, the generic proof goes through using a sized
+typing for these inductive families.  Similarily, defining the
+$\tjoin$ for monad $\Cov$ relies on sized typing.
+\[
+\begin{array}{lll}
+  \tjoin & : & \forall i.\
+    \Cov[i]\,(\Cov[\infty]\J) \todot \Cov[\infty]\J  \\
+  \tjoin[i+1]\,(\treturn[i]\;c) & = & c \\
+  \tjoin[i+1]\,(\tmatch^i\;t\;k)
+    & = & \tmatch^\infty\;
+     (t : \Ne P \Gamma)\;
+     (\stmap[\Add P]_\Gamma\;\tjoin[i]\;(k : \AddHypP P {\Cov[i]\J} \Gamma)) \\
+  % \tjoin^{i+1}\,(\tmatch\,(t : \Ne P \Gamma)\,
+  %   (k : \AddHypP P {\Cov^i\,\J} \Gamma))
+  %   & = & \tmatch\;t\;(\stmap[\Add P]_\Gamma\;\tjoin^i\;k) \\
+\end{array}
+\]
+Herein, we used the sized typing of the constructors of $\Cov$:
+\[
+\begin{array}{lll}
+  \treturn & : & \forall i.\ \J \todot \Cov[i+1] \J \\
+  \tmatch  & : & \forall i.\
+    \tNe\,(\Comp\,P) \todot \Add P\,(\Cov[i]\J) \todot \Cov[i+1]\J \\
+\end{array}
+\]
 

### 2019-02-10 fscd19/fscd19.tex
[ fscd19 ] n-ary presheaf morphisms
[ fscd19 ] proof reading Sunday morning

@@
 propositional equality for all morphisms $\tau$.
+The singleton weakening $\twk^A_\Gamma : \Gamma \sublist \cext \Gamma A$,
+also written $\twk^A$ or $\twk$, is defined by $\twk = \tweak\,\tid$.
 
@@
   \Delta \to A \in \Gamma \to A \in \Delta$}, which is the morphism
-part of functor from $\Cxt$ to $\Set$, mapping object $\Gamma$ to the
+part of functor $A{\in}\_$ from $\Cxt$ to $\Set$, mapping object $\Gamma$ to the
 set $A \in \Gamma$ of the indices of $A$ in $\Gamma$.
@@
 Intrinsically well-typed lambda-terms, in abstract syntax,
-are just inhabitants $t$ of the type \fbox{$A \from \Gamma$},
+are just inhabitants $t$ of the indexed set \fbox{$A \from \Gamma$},
 inductively defined by the following rules.
@@
     {A_1 + A_2 \from \Gamma
-      \qquad C \from \cext \Gamma A_1
-      \qquad C \from \cext \Gamma A_2
-    }{C \from \Gamma}
+      \qquad B \from \cext \Gamma A_1
+      \qquad B \from \cext \Gamma A_2
+    }{B \from \Gamma}
 \qquad
@@
     {\tyzero \from \Gamma}
-    {C \from \Gamma}
+    {B \from \Gamma}
 \end{gather*}
@@
 calculus we should better speak of \emph{reindexing}, which could, a
-bit clumsily, also abbreviated to $\tren$.}
+bit clumsily, be also abbreviated to $\tren$.}
 \fbox{$\tren : \Gamma \sublist \Delta \to A \from \Gamma \to A \from \Delta$}
@@
 \emph{subformula} property \cite{prawitz:natded,matthes:shortproofs}.
-Normal forms \fbox{$t : \tNf A\, \Gamma$} are mutually defined with
-\emph{neutral} normal forms \fbox{$t : \Ne A \Gamma$}.
+Normal forms \fbox{$n : \tNf A\, \Gamma$} are mutually defined with
+\emph{neutral} normal forms \fbox{$u : \Ne A \Gamma$}.
 In the following inductive definition, we reuse the rule names from the
@@
 \end{gather*}
-These rules only allow the elimination of neutrals; guarantees
+These rules only allow the elimination of neutrals; this guarantees
 the subformula property and prevents any kind of computational
@@
 (The point on the arrow is mnemoic for \emph{pointwise}.)
+Slightly abusive, we shall extend this notation to $n$-ary
+morphisms, \eg, write $\A \todot \B \todot \C$ for
+$\forall \Gamma.\ \A\,\Gamma \to (\B\,\Gamma \to \C\,\Gamma)$.
 
 While the coproduct eliminations $\tcase$ and $\tabort$ are limited to
-normal forms of positive types $P$, their extension to negative types
-is admissible.
+normal forms of positive types $P$, their extension $\tcase^B$ and
+$\tabort^B$ to negative types
+is admissible, for instance:
 \[
@@
 \emph{Proof normalization} is concerned with finding a normal form
-$u : \Nf A \Gamma$ for each derivation $t : A \from \Gamma$.  However,
-\emph{term normalization} additionally requires that $\tm u \cong t$
+$n : \Nf A \Gamma$ for each derivation $t : A \from \Gamma$.  However,
+\emph{term normalization} additionally requires \emph{soundness},
+\ie, that $\tm u \cong t$
 with respect to a equational theory $\cong$ on terms.
@@
 \fbox{$\tnorm : A \from \Gamma \to \Nf A \Gamma$}
-that does not guarantee any relation between its input term $t$ and
-output normal form $u$.
+without any soundness guarantee.
+% that does not guarantee any relation between its input term $t$ and
+% output normal form $u$.
+%
 % \emph{Normalization by evaluation} (NbE) decomposes normalization as
@@
 $\idenv\Gamma$ which will map the free indices of $t$ to corresponding
-\emph{unknowns}.  To accomodate unknowns in the semantics, types $A$
+\emph{unknowns}.  To accommodate unknowns in the semantics, types $A$
 are mapped to presheaves $\den A$ (rather than just sets), and in
@@
 % \end{gather*}
-In the equation for $\down{A \To B}$,
-the renaming $\twk : \Gamma \sublist \cext \Gamma A$,
-defined by $\twk = \tweak\,\tid$, makes room
+%In the equation for
+In the reification at function types
+$\down{A \To B}$,
+the renaming $\twk^A_\Gamma : \Gamma \sublist \cext \Gamma A$
+makes room
 for a new variable of type $A$, which is reflected into $\den A$ by
@@
 semantic objects such as $f : \den{A \To B}\Gamma$ in a such extended
-context, is the reason for utilizing presheaves instead of just sets.
+context, is the reason for utilizing presheaves instead of just sets
+as semantic types.
 
@@
 % $\up[\tyone + \tyone]^{\tyone + \tyone} (\tvar\,\tzero) : \bone \bplus \bone$
-$\up[A + B]{A + B} (\tvar\,\tzero) : \den A (A+B) \bplus \den B (A+B)$
+$\up[A + B]{A + B} (\tvar\,\tzero) : \den A (A{+}B) \bplus \den B (A{+}B)$
 which would require us to make a decision of whether $A$ holds or $B$ holds
@@
 \begin{array}{lcl@{\quad}l}
-\tabort^\C & : & \tNe\, \tyzero \todot \C\,\A
+\tabort^\C & : & \tNe\, \tyzero \todot \C\,\B
   & \mbox{case on absurd neutral} \\
-\tcase^\C_\Gamma & : & \Ne {(A + B)} \Gamma
-  \to \C\,\A\,(\cext \Gamma A)
-  \to \C\,\A\,(\cext \Gamma B)
-  \to \C\,\A\,\Gamma
+\tcase^\C_\Gamma & : & \Ne {(A_1 + A_2)} \Gamma
+  \to \C\,\B\,(\cext \Gamma {A_1})
+  \to \C\,\B\,(\cext \Gamma {A_2})
+  \to \C\,\B\,\Gamma
   & \mbox{case on neutral} \\
@@
 binary case tree whose inner nodes ($\tcase$) are labeled by a neutral term
-of sum type $A+B$ and its two branches by the context extensions $A$
-and $B$, resp.  Leaves are either ($\tabort$)
-labeled by a neutral term of empty type $\tyzero$,
-or ($\treturn$) by an element of $\A$.  Functoriality amounts to
+of sum type $A_1+A_2$ and its two branches by the context extensions $A_1$
+and $A_2$, resp.  Leaves are either
+labeled by a neutral term of empty type $\tyzero$ (see $\tabort$),
+or by an element of $\A$ (see $\treturn$).  Functoriality amounts to
 replacing the labels of the $\treturn$-leaves, and the monadic bind replaces
 these leaves by further case trees.  (The uninspiring $\tjoin[\Cov]$
-flattens a 2-level case tree into a single one.)
-Finally $\trunNf[\Cov]$ is a simple recursion, replacing $\tcase^{\Cov}$
+flattens a 2-level case tree, \ie, a case tree with case trees as
+leaves, into a single one.)
+Finally $\trunNf[\Cov]$ is a simple recursion on the tree, replacing $\tcase^{\Cov}$
 and $\tabort^{\Cov}$
@@
 Positive types $P$ are already monadic,
-and $\trun^P$ is simply the $\tjoin$ of monad $\C$.
+and $\trun^P$ is simply the $\tjoin$ of the monad $\C$.
 At negative types we can recurse pointwise at a smaller type,
@@
 \\[2ex]
-  \dencase^B_\Gamma  & : & \den{A_1 + A_2}\Gamma
-    \to \den{A_1 \To B}\Gamma
-    \to \den{A_2 \To B}\Gamma
-    \to \den{B}\Gamma \\
-  \dencase^B_\Gamma\;c\;f_1\;f_2 & = &
+  \dencase^B  & : & \den{A_1 + A_2}
+    \todot \den{A_1 \To B}
+    \todot \den{A_2 \To B}
+    \todot \den{B} \\
+  \dencase^B\;c\;f_1\;f_2 & = &
     \trun^B (\stmap[\C]\;(\lambda\,\tau.\ [f_1\,\tau,\ f_2\,\tau])\;c) \\
@@
 }
-\fbox{$\NF t = \exists u \of \tNf A\,\Gamma.\; \tm u \cong t$}
+\fbox{$\NF t = \exists n \of \tNf A\,\Gamma.\ \tm n \cong t$}
 % for $t : A \from \Gamma$,
@@
 \]
-that gives us for each term $t$ a $\cong$-equal normal form $u$.
+that gives us for each term $t$ a $\cong$-equal normal form $n$.
 
@@
 We express maximal pattern matching on $P$ as a strong functor
-$\Add P$ in the category of presheaves
+$\Add P$ in the category of presheaves,
 mapping a presheaf $\J$ (``judgement'') to $\Add P \J$
@@
   \nru{\bsplit}
-      {\AddHyp {P_1} {(\Add{P_2}\,\J)} \Gamma}
+      {\AddHyp {P_1} {\left(\Add{P_2}\,\J\right)} \Gamma}
       {\AddHyp {P_1 \times P_2} \J \Gamma}
@@
      (t : \Ne P \Gamma)\;
-     (\stmap[\Add P]_\Gamma\;\tjoin[i]\;(k : \AddHypP P {\Cov[i]\J} \Gamma)) \\
+     \left(\stmap[\Add P]_\Gamma\;\tjoin[i]\;(k : \AddHypP P {\Cov[i]\J} \Gamma)\right) \\
   % \tjoin^{i+1}\,(\tmatch\,(t : \Ne P \Gamma)\,

### 2019-02-10 fscd19/fscd19.tex
[ fscd19 ] Continuation monad

@@
      \dencase\;\dent u \gamma\;f_1\;f_2
-    \ \mbox{ \textbf{where}} \\
+    \ \bwhere \\
     \multicolumn 1 r {f_i\; (\tau : \Delta \sublist \Phi)\,(a_i : \den {A_i} \Phi)}
@@
 
-\clearpage
+An alternative implementation of $\C$ is the continuation monad $\CC$
+on presheaves as, for instance, employed by Ilik et al.\
+\cite{ilikLeeHerbelin:apal10}
+for Kripke models and NbE of classical logic.
+\[
+  \CC\;\J = \forall A.\ (\J \hatto \tNf\,A) \hatto \tNf\,A
+\]
+The answer type of this continuation monad is always $\tNF$, however,
+we are polymorphic in the simple type $A$ of normal forms we produce.
+
+Agda has been really helpful to produce the rather technical but
+straightforward evidence that $\CC$ is a strong monad.
+The method $\trunNf[\CC] : \CC\,(\tNf\,A) \todot \tNf\,A$
+exists by definition, using the identity continuation $\tNf\,A \hatto \tNf\,A$.
+In the following, we demonstrate that $\CC$
+enables matching on neutrals:
+\[
+\begin{array}{l}
+  \tabort^{\CC} \, (u : \Ne \tyzero \Gamma) ~:~  \CC\,\J\,\Gamma \\
+  \tabort^{\CC}
+    \; u
+    \; (\tau : \Gamma \sublist \Delta)
+    \; (k : (\J \hatto \tNf\,B)\Delta)
+     ~=~
+    \tabort\,(\tren\,\tau\,u)
+\\[2ex]
+  \tcase^{\CC}
+     \; (u : \NeP {A_1 + A_2} \Gamma)
+     \; (c_1 : \CC\,\J\,(\cext \Gamma {A_1}))
+     \; (c_2 : \CC\,\J\,(\cext \Gamma {A_2}))
+    ~:~ \CC\,\J\,\Gamma \\
+  \tcase^{\CC}
+    \; u
+    \; c_1
+    \; c_2
+    \; (\tau : \Gamma \sublist \Delta)
+    \; (k : (\J \hatto \tNf\,B)\Delta)
+     ~=~
+    \tcase\;(\tren\,\tau\,u)\;n_1\;n_2
+    \bwhere \\
+\qquad
+  \begin{array}{l@{~~}c@{~~}l}
+     n_i & : & \Nf B {(\cext \Delta {A_i})} \\
+     n_i & = & c_i
+      \; (\tlift^{A_i}\,\tau
+            : \cext \Gamma {A_i} \sublist \cext \Delta {A_i})
+      \; \left(\lambda
+            \; (\tau' : \cext \Delta {A_i} \sublist \Phi)
+            \; (j :  \J\,\Phi) .
+            \ k
+            \; (\twk^{A_i} \rcomp \tau')
+            \; j
+         \right)
+  \end{array}
+\end{array}
+\]
+The NbE algorithm using $\CC$ is comparable to Danvy's type-directed
+partial evaluation \cite[Figure~8]{danvy:popl96}.  However, he uses
+shift-reset style continutations which can be programmed in the
+continuation monad, and relies on Scheme's \texttt{gensym} to produce
+fresh variables names rather than using Kripke function space / presheaves.
 
@@
 
+\clearpage
+
 \section{Call-By-Push Value}
@@
 \\
-  \HTy & \ni & X,Y & ::= & \patom \mid P
+  \HTy & \ni & H & ::= & \patom \mid P
   & \mbox{hypothesis type}
@@
 \qquad
-  \nru{\tthunk}{\Nf N \Gamma}{\PValP{\Thunk\,N}\Gamma}
+  \nru{\tthunk}{\Nf N \Gamma}{\PValP{\Thunk N}\Gamma}
 \\[2ex]
@@
   \nru{\tmatch}
-      {\NeP {\Comp\,P} \Gamma \qquad \AddHypP P {\Cov\,\J} \Gamma}
+      {\NeP {\Comp P} \Gamma \qquad \AddHypP P {\Cov\,\J} \Gamma}
       {\Cover \J \Gamma}
@@
       {\CoverP {\tVal\,P}  \Gamma}
-      {\NfP {\Comp\,P} \Gamma}
+      {\NfP {\Comp P} \Gamma}
 \quad
@@
   \tmatch  & : & \forall i.\
-    \tNe\,(\Comp\,P) \todot \Add P\,(\Cov[i]\J) \todot \Cov[i+1]\J \\
+    \tNe\,(\Comp P) \todot \Add P\,(\Cov[i]\J) \todot \Cov[i+1]\J \\
 \end{array}
@@
 \qquad
-  \nru{\tthunk}{\T\,N\,\Gamma}{\PValTP{\Thunk\,N}\Gamma}
+  \nru{\tthunk}{\T\,N\,\Gamma}{\PValTP{\Thunk N}\Gamma}
 \\[2ex]
@@
   \nru{\tmatch}
-      {\NeTP {\Comp\,P} \Gamma \qquad \AddHypP P {\Cov[\T]\J} \Gamma}
+      {\NeTP {\Comp P} \Gamma \qquad \AddHypP P {\Cov[\T]\J} \Gamma}
       {\CoverT \J \Gamma}
@@
       {\Cov[\tNf](\tVal[\tNf]\,P)\,\Gamma}
-      {\NfP {\Comp\,P} \Gamma}
+      {\NfP {\Comp P} \Gamma}
 \\[2ex]

### 2019-02-10 fscd19/fscd19.tex
[ fscd19 ] equational theory for IPL

@@
 $\treturn[\C] : \A \todot \C\,\A$ (unit) and
-$\tjoin[\C] : \C\,(\C\,\A) \todot \C\,\A$ (concatenation) satisfying
+$\tjoin[\C] : \C\,(\C\,\A) \todot \C\,\A$ (multiplication)
+% (concatenation)
+satisfying
 the monad laws.
@@
 
+%Equational theory \fbox{$t \cong t'$} for $t,t' : A \from \Gamma$.
+Term equality\fbox{$t \cong t'$} for $t,t' : A \from \Gamma$ is the
+least congruence over the following axioms, which we have grouped into
+$\beta$ (computation), $\eta$ (extensionality), and $\pi$ (permutation)
+rules.  We may refer to subgroups, like $\eta^N$ for the negative
+$\eta$-rules ($\retato$, $\retatimes$, $\retaone$).  Note that all
+typing restrictions are implicit in the well-typedness assumption,
+\eg, $t \cong \tunit$ presupposes $t : \tyone \from \Gamma$
+since $\tunit : \tyone \from \Gamma$.
+\begin{gather*}
+  \nru{\rbetato}{}{\tapp\,(\tabs\,t)\,u \cong \subone t u}
+\qquad
+  \nru{\rbetatimes}{}{\prjP i {\tpair\,t_1\,t_2} \cong t_i}
+\qquad
+  \nru{\rbetaplus}{}{\tcase\,(\inj i t)\,t_1\,t_2 \cong t_i}
+\\[2ex]
+  \nru{\retato}{}{t \cong
+    \tabs\,(\tapp\,(\tren\,\twk\,t)\,(\tvar\,\tzero))}
+\quad
+  \nru{\retatimes}{}{t \cong \tpair\,(\prj 1 t)\,(\prj 2 t)}
+\quad
+  \nru{\retaone}{}{t \cong \tunit}
+\\[1ex]
+  \nru{\retaplus}{}{t \cong
+    \tcase\,t
+      \,(\injP 1 {\tvar\,\tzero})
+      \,(\injP 2 {\tvar\,\tzero})
+    }
+\qquad
+  \nru{\retazero}{}{t \cong \tabort\,t}
+\\[2ex]
+  \nru{\rpi{\To}{\tyempty}}
+    {}{\tapp\,(\tabort\,t)\,u \cong \tabort\,t}
+\qquad
+  \nru{\rpi{\times}{\tyempty}}
+    {}{\prjP i {\tabort\,t} \cong \tabort\,t}
+\\[1ex]
+  \nru{\rpi{+}{\tyempty}}
+    {}{\tcase\,(\tabort\,t)\,t_1\,t_2 \cong \tabort\,t}
+\qquad
+  \nru{\rpi{\tyempty}{\tyempty}}
+    {}{\tabort\,(\tabort\,t) \cong \tabort\,t}
+\\[2ex]
+  \nru{\rpi{\To}{+}}
+    {}{\tapp\,(\tcase\;t\;t_1\;t_2)\,u \cong \tcase\;t\;t_1'\;t_2'}
+    \
+    t_i' = \tapp\,t_i\,(\tren\,\twk\,u)
+  % \nru{\rpi{\To}{+}}
+  %   {}{\tapp\,(\tcase\;t\;t_1\;t_2)\,u \cong \tcase\,t
+  %        \, (\tapp\,t_1\,(\tren\,\twk\,u))
+  %        \, (\tapp\,t_2\,(\tren\,\twk\,u))
+  %   }
+\\
+  \nru{\rpi{\times}{+}}
+    {}{\prjP i {\tcase\;t\;t_1\;t_2} \cong
+         \tcase\,t\,(\prj i {t_1})\,(\prj i {t_2})}
+\\
+  \nru{\rpi{+}{+}}
+    {}{\tcase\,(\tcase\;t\;t_1\;t_2)\,u_1\,u_2 \cong
+        \tcase\;t\;t_1'\;t_2'}
+    \
+    t_i' = \tcase\;t_i\;u_1'\;u_2'
+    ; \quad
+    u_j' = \tren\,(\tlift\,\twk)\,u_j
+    % t_i' = \tcase\;t_i
+    %   \,(\tren\,(\tlift\,\twk)\,u_1)
+    %   \,(\tren\,(\tlift\,\twk)\,u_2)
+\\
+  \nru{\rpi{\tyempty}{+}}
+    {}{\tabort\,(\tcase\;t\;t_1\;t_2) \cong \tcase\,t\,(\tabort\,t_1)\,(\tabort\,t_2)}
+\end{gather*}
+The different law classes contribute to the shape and completeness of
+the normal forms in the following way:
+Thanks to the $\eta$-laws, a normal form of negative type is always an
+introduction.  A normal form of sum type ($+$, $\tyzero$) is always a
+case tree whose leaves are injections.  A normal form of atomic type
+($o$) is always a neutral.  The $\beta$ and $\pi$ laws determine the
+shape of neutrals.
+Thanks to the $\beta$-laws,
+the principal argument of normal eliminations cannot be an
+introduction.  Because of the negative permutation rules
+$\rpi N P$, it cannot be positive elimination
+($\tcase$, $\tabort$) either, which leaves only negative eliminations
+in neutrals ($\tapp$, $\tprj$).  Finally the positive permutation
+rules $\rpi P P$ guarantee that having neutral scrutinees in case
+trees is sufficient.
+
 \clearpage

### 2019-02-10 fscd19/fscd19.tex
[ fscd19 ] Terms of CBPV

--- 2019-02-10 fscd19/fscd19.tex
@@
 A more accurate presentation of $\tNF$ would be
-$\sum_{u : \tNf A\,\Gamma} (\tm u \cong t)$
+$\Sigma\,{u : \tNf A\,\Gamma}.\ (\tm u \cong t)$
 in order to make absolutely clear that we
@@
   \emph{mere existence} from Homotopy Type Theory which
-  requires propositional truncation.
+  is the propositional truncation of constructive existence.
 }
@@
 
+%Equational theory \fbox{$t \cong t'$} for $t,t' : A \from \Gamma$.
+Term equality\fbox{$t \cong t'$} for $t,t' : A \from \Gamma$ is the
+least congruence over the following axioms, which we have grouped into
+$\beta$ (computation), $\eta$ (extensionality), and $\pi$ (permutation)
+rules.  We may refer to subgroups, like $\eta^N$ for the negative
+$\eta$-rules ($\retato$, $\retatimes$, $\retaone$).  Note that all
+typing restrictions are implicit in the well-typedness assumption,
+\eg, $t \cong \tunit$ presupposes $t : \tyone \from \Gamma$
+since $\tunit : \tyone \from \Gamma$.
+\begin{gather*}
+%% BETA LAWS
+  \nru{\rbetato}{}{\tapp\,(\tabs\,t)\,u \cong \subone t u}
+\qquad
+  \nru{\rbetatimes}{}{\prjP i {\tpair\,t_1\,t_2} \cong t_i}
+\qquad
+  \nru{\rbetaplus}{}{\tcase\,(\inj i t)\,t_1\,t_2 \cong t_i}
+\\[2ex]
+%% ETA LAWS
+  \nru{\retato}{}{t \cong
+    \tabs\,(\tapp\,(\tren\,\twk\,t)\,(\tvar\,\tzero))}
+\quad
+  \nru{\retatimes}{}{t \cong \tpair\,(\prj 1 t)\,(\prj 2 t)}
+\quad
+  \nru{\retaone}{}{t \cong \tunit}
+\\[1ex]
+  \nru{\retaplus}{}{t \cong
+    \tcase\,t
+      \,(\injP 1 {\tvar\,\tzero})
+      \,(\injP 2 {\tvar\,\tzero})
+    }
+\qquad
+  \nru{\retazero}{}{t \cong \tabort\,t}
+\\[2ex]
+%% PERMUTATIONS
+  \nru{\rpi{\To}{\tyempty}}
+    {}{\tapp\,(\tabort\,t)\,u \cong \tabort\,t}
+\qquad
+  \nru{\rpi{\To}{+}}
+      {t_i' = \tapp\,t_i\,(\tren\,\twk\,u)}
+      {\tapp\,(\tcase\;t\;t_1\;t_2)\,u \cong \tcase\;t\;t_1'\;t_2'}
+\\[1ex]
+  \nru{\rpi{\times}{\tyempty}}
+    {}{\prjP i {\tabort\,t} \cong \tabort\,t}
+\qquad
+  \nru{\rpi{\times}{+}}
+    {}{\prjP i {\tcase\;t\;t_1\;t_2} \cong
+         \tcase\,t\,(\prj i {t_1})\,(\prj i {t_2})}
+\\[2ex]
+  \nru{\rpi{+}{\tyempty}}
+    {}{\tcase\,(\tabort\,t)\,t_1\,t_2 \cong \tabort\,t}
+\qquad
+  \nru{\rpi{+}{+}}
+      {t_i' = \tcase\;t_i\;u_1'\;u_2'
+       \quad
+       u_j' = \tren\,(\tlift\,\twk)\,u_j
+      }
+      {\tcase\,(\tcase\;t\;t_1\;t_2)\,u_1\,u_2 \cong
+        \tcase\;t\;t_1'\;t_2'}
+    \
+    % t_i' = \tcase\;t_i
+    %   \,(\tren\,(\tlift\,\twk)\,u_1)
+    %   \,(\tren\,(\tlift\,\twk)\,u_2)
+\\[1ex]
+  \nru{\rpi{\tyempty}{\tyempty}}
+    {}{\tabort\,(\tabort\,t) \cong \tabort\,t}
+\qquad
+  \nru{\rpi{\tyempty}{+}}
+    {}{\tabort\,(\tcase\;t\;t_1\;t_2) \cong
+      \tcase\,t\,(\tabort\,t_1)\,(\tabort\,t_2)}
+%% OLD:
+%   \nru{\rpi{\To}{\tyempty}}
+%     {}{\tapp\,(\tabort\,t)\,u \cong \tabort\,t}
+% \qquad
+%   \nru{\rpi{\times}{\tyempty}}
+%     {}{\prjP i {\tabort\,t} \cong \tabort\,t}
+% \\[1ex]
+%   \nru{\rpi{+}{\tyempty}}
+%     {}{\tcase\,(\tabort\,t)\,t_1\,t_2 \cong \tabort\,t}
+% \qquad
+%   \nru{\rpi{\tyempty}{\tyempty}}
+%     {}{\tabort\,(\tabort\,t) \cong \tabort\,t}
+% \\[2ex]
+%   \nru{\rpi{\To}{+}}
+%     {}{\tapp\,(\tcase\;t\;t_1\;t_2)\,u \cong \tcase\;t\;t_1'\;t_2'}
+%     \
+%     t_i' = \tapp\,t_i\,(\tren\,\twk\,u)
+%   % \nru{\rpi{\To}{+}}
+%   %   {}{\tapp\,(\tcase\;t\;t_1\;t_2)\,u \cong \tcase\,t
+%   %        \, (\tapp\,t_1\,(\tren\,\twk\,u))
+%   %        \, (\tapp\,t_2\,(\tren\,\twk\,u))
+%   %   }
+% \\
+%   \nru{\rpi{\times}{+}}
+%     {}{\prjP i {\tcase\;t\;t_1\;t_2} \cong
+%          \tcase\,t\,(\prj i {t_1})\,(\prj i {t_2})}
+% \\
+%   \nru{\rpi{+}{+}}
+%     {}{\tcase\,(\tcase\;t\;t_1\;t_2)\,u_1\,u_2 \cong
+%         \tcase\;t\;t_1'\;t_2'}
+%     \
+%     t_i' = \tcase\;t_i\;u_1'\;u_2'
+%     ; \quad
+%     u_j' = \tren\,(\tlift\,\twk)\,u_j
+%     % t_i' = \tcase\;t_i
+%     %   \,(\tren\,(\tlift\,\twk)\,u_1)
+%     %   \,(\tren\,(\tlift\,\twk)\,u_2)
+% \\
+%   \nru{\rpi{\tyempty}{+}}
+%     {}{\tabort\,(\tcase\;t\;t_1\;t_2) \cong \tcase\,t\,(\tabort\,t_1)\,(\tabort\,t_2)}
+\end{gather*}
+The different law classes contribute to the shape and completeness of
+the normal forms in the following way:
+Thanks to the $\eta$-laws, a normal form of negative type is always an
+introduction.  A normal form of sum type ($+$, $\tyzero$) is always a
+case tree whose leaves are injections.  A normal form of atomic type
+($o$) is always case tree whose leaves are neutral.
+The $\beta$ and $\pi$ laws determine the shape of neutrals.
+Thanks to the $\beta$-laws,
+the principal argument of normal eliminations cannot be an
+introduction.  Because of the negative permutation rules
+$\rpi N P$ it cannot be positive elimination
+($\tcase$, $\tabort$) either, which leaves only negative eliminations
+in neutrals ($\tapp$, $\tprj$).  Finally the positive permutation
+rules $\rpi P P$ guarantee that having neutral scrutinees in case
+trees is sufficient.
+
 \clearpage
@@
 \label{sec:cbpv}
-
-\section{Focused Intuitionistic Propositional Logic}
-\label{sec:fipl}
-
-Polarized lambda calculus \cite{espiritoSanto:entcs17}
-divides types deeply into positive and negative.
 
@@
 \]
+
+Pure CBPV adds right non-invertible focusing.
+
+Assuming a family $\T\,N\,\Gamma$ of terms of negative type $N$ in
+context $\Gamma$, values \fbox{$v : \PValT P \Gamma$}
+of positive type $P$ are constructed by the
+following rules:
+\begin{gather*}
+  \nru{\pvar}{P \in \Gamma}{\PValT P \Gamma}
+\qquad
+  \nru{\tthunk}{\T\,N\,\Gamma}{\PValTP{\Thunk N}\Gamma}
+\\[2ex]
+  \nru{\punit}{}{\PValT{\tyone}\Gamma}
+\qquad
+  \nru{\ppair}
+      {\PValT{P_1}\Gamma \qquad \PValT{P_2}\Gamma}
+      {\PValTP{P_1 \times P_2}\Gamma}
+\qquad
+  \nru{\tinj_i}{\PValT{P_i}\Gamma}{\PValTP{P_1 + P_2}\Gamma}
+\end{gather*}
+
+It relies on the previous definitions
+with instance $\T = \tNf$.
+\fbox{$\Tm N \Gamma$}:
+\begin{gather*}
+  \nru{\tret}
+      {\PVal P \Gamma}
+      {\TmP {\Comp P} \Gamma}
+\quad
+  \nru{\tabs}
+      {\Tm N {(\cext \Gamma P)}}
+      {\TmP {P \To N} \Gamma}
+\quad
+  \nru{\npair}
+      {\Tm {N_1} \Gamma \qquad \Tm {N_2} \Gamma}
+      {\TmP {N_1 \ntimes N_2} \Gamma}
+\quad
+  \nru{\nunit}
+      {}
+      {\Tm \top \Gamma}
+\\[2ex]
+  \nru{\tforce}
+      {\PValP {\Thunk N} \Gamma}
+      {\Tm N \Gamma}
+\qquad
+  \nru{\tapp}
+      {\TmP {P \To N} \Gamma \qquad \PVal P \Gamma}
+      {\Tm N \Gamma}
+\qquad
+  \nru{\tprj_i}
+      {\TmP {N_1 \ntimes N_2} \Gamma}
+      {\Tm {N_i} \Gamma}
+\\[2ex]
+% \end{gather*}
+% \begin{gather*}
+  \nru{\tbind}
+      {\TmP {\Comp P} \Gamma \qquad \Tm N {(\cext \Gamma P)}}
+      {\Tm N \Gamma}
+\qquad
+  \nru{\tsplit}
+    {\TmP {P_1 \times P_2} \Gamma
+      \qquad \Tm N {(\cext {\cext \Gamma P_1} P_2)}
+    }{\Tm N \Gamma}
+\\[2ex]
+  \nru{\tcase}
+    {\TmP {P_1 + P_2} \Gamma
+      \qquad \Tm N {(\cext \Gamma P_1)}
+      \qquad \Tm N {(\cext \Gamma P_2)}
+    }{\Tm N \Gamma}
+\qquad
+  \nru{\tabort}
+    {\Tm \tyzero \Gamma}
+    {\Tm N \Gamma}
+\end{gather*}
+
+\section{Focused Intuitionistic Propositional Logic}
+\label{sec:fipl}
+
+Polarized lambda calculus \cite{espiritoSanto:entcs17}
+divides types deeply into positive and negative.
+
+\[
+\begin{array}{lllrl@{\qquad}l}
+  \PTy & \ni & P,Q & ::= & \patom
+    \mid \tyzero \mid P_1 + P_2
+    \mid \tyone \mid P_1 \times P_2
+    \mid \Thunk N
+  & \mbox{positive type}
+\\
+  \NTy & \ni & N,M & ::= & \natom
+    \mid \top \mid N_1 \ntimes N_2
+    \mid P \To N
+    \mid \Comp P
+  & \mbox{negative type}
+\\
+  \HTy & \ni & H & ::= & \patom \mid P
+  & \mbox{hypothesis type}
+\\
+\end{array}
+\]
 Espirito Santo's version actually does not contain the positive

### 2019-02-11 fscd19/fscd19.tex
[ fscd19 ] Towards positive reflection

@@
 
-\section{Call-By-Push Value}
+\section{Normalization to Call-By-Push Value}
 \label{sec:cbpv}
+
+As remarked in the previous section, the placement of the monad $\C$
+in the type semantics is a bit wasteful:  Each positive type is
+prefixed by $\C$.  In our grammar of normal forms, this corresponds to
+the ability to perform case distinctions ($\tcase$, $\tabort$) at any
+positive type $P$.  In fact, our type interpretation $\den A$
+corresponds to the translation of call-by-name (CBN) lambda-calculus
+into Moggi's monadic meta-language \cite{moggi:infcomp91,levy:hosc05}.
+
+It would be sufficient, however, to perform all necessary
+case distinctions
+when transitioning from a negative type to a positive type.
+Introduction of the function type adds hypotheses to the context,
+providing material for case distinctions, but introduction of positive
+types does not add anything in that respect.  Thus, we could
+\emph{focus} on positive introductions until we transition back to a
+negative type.  Such focusing is present in the call-by-value (CBV)
+lambda-calculus, where positive introductions only operate on values,
+and variables stand only for values.  This structure is even more
+clearly spelled out in Levy's call-by-push-value (CBPV)
+\cite{levy:hosc05}, as it comes with a deep classification of types
+into positive and negative ones.  In the following, we shall utilize
+pure (\ie, effect-free) CBPV to achieve chaining of positive
+introductions.
+
+CBPV calls positive types $P$ \emph{value types} $A$ and negative
+types $N$
+\emph{computation types} $\underline B$,
+however, we shall stick to our terminology
+which is known from research on focalization.
+However, we shall use $\tThunk$ for switch $\downarrow$ and $\tComp$ for
+switch $\uparrow$.
+\[
+\begin{array}{lllrl@{\qquad}l}
+  \PTy & \ni & P,Q & ::= & \patom
+    \mid \tyzero \mid P_1 + P_2
+    \mid \tyone \mid P_1 \times P_2
+    \mid \Thunk N
+  & \mbox{positive type}
+\\
+  \NTy & \ni & N,M & ::= & \natom
+    \mid \top \mid N_1 \ntimes N_2
+    \mid P \To N
+    \mid \Comp P
+  & \mbox{negative type}
+\\
+\end{array}
+\]
+CBPV uses $U$ for $\tThunk$ and $F$ for $\tComp$, however, we find
+these names uninspiring unless you have good knowledge of the intended
+model.  Further, CBPV employs labeled sums $\Sigma_I (P_i)_{i : I}$
+and labeled records $\Pi_I (P_i)_{i : I}$ for up to countably infinite
+label sets $I$ while we only have finite sums $(\tyzero,+)$ and
+records $(\top,\ntimes)$.  However, this difference in not essential,
+our treatment extends directly to the infinite case, since we are
+working in type theory which allows infinitely branching inductive
+types.   As a last difference, CBPV does not consider base types; in
+anticipation of the next section, we add them as both positive atoms
+($\patom$) and negative atoms ($\natom$).
+
+Getting a bit ahead of ourselves, let us consider the
+mutually defined interpretations
+$\den P$ and $\den N$ of positive and negative types as presheaves.
+\[
+\begin{array}{lcl}
+  \den{\tyone} & = & \hatone \\
+  \den{P_1 \times P_2} & = & \den{P_1} \hattimes \den{P_2} \\
+  \den{\tyzero} & = & \hatzero \\
+  \den{P_1 + P_2} & = & \den{P_1} \hatplus \den{P_2} \\
+  \den{\Thunk N} & = & \den N \\
+  \den{\patom} & = & \patom{\in}\_ \\
+\end{array}
+\qquad
+\begin{array}{lcl}
+  \den{\top} & = & \hatone \\
+  \den{N_1 \ntimes N_2} & = & \den{N_1} \hattimes \den{N_2} \\
+  \\
+  \den{P \To N} & = & \den P \hatto \den N \\
+  \den{\Comp P} & = & \C \den P \\
+  \den{\natom}  & = & \C (\tNe\, \natom) \\
+\end{array}
+\]
+Semantically, we do not distinguish between positive and negative
+products.  Notably, sum types can now be interpreted as plain presheaf
+sums.  The $\tThunk$ marker is ignored, yet $\tComp$, marking the
+switch from the negative to the positive type interpretation, places
+the cover monad.  Positive atoms, standing for value types without
+constructors, are only inhabited by variables $x : \patom \in \Gamma$.
+Negative atoms stand for computation types without own eliminations,
+thus, their inhabitants stem only from eliminations of more complex
+types, made from positive eliminations captured in $\C$ and negative
+eliminations chained together as neutral $\tNe\,\natom$, which we
+shall define below.
+
+Assuming a family $\T\,N\,\Gamma$ of terms of negative type $N$ in
+context $\Gamma$, values \fbox{$v : \PValT P \Gamma$}
+of positive type $P$ are constructed by the
+following rules:
+\begin{gather*}
+  \nru{\pvar}{P \in \Gamma}{\PValT P \Gamma}
+\qquad
+  \nru{\tthunk}{\T\,N\,\Gamma}{\PValTP{\Thunk N}\Gamma}
+\\[2ex]
+  \nru{\punit}{}{\PValT{\tyone}\Gamma}
+\qquad
+  \nru{\ppair}
+      {\PValT{P_1}\Gamma \qquad \PValT{P_2}\Gamma}
+      {\PValTP{P_1 \times P_2}\Gamma}
+\qquad
+  \nru{\tinj_i}{\PValT{P_i}\Gamma}{\PValTP{P_1 + P_2}\Gamma}
+\end{gather*}
+
+It relies on the previous definitions
+with instance $\T = \tNf$.
+\fbox{$\Tm N \Gamma$}:
+\begin{gather*}
+  \nru{\tret}
+      {\PVal P \Gamma}
+      {\TmP {\Comp P} \Gamma}
+\quad
+  \nru{\tabs}
+      {\Tm N {(\cext \Gamma P)}}
+      {\TmP {P \To N} \Gamma}
+\quad
+  \nru{\npair}
+      {\Tm {N_1} \Gamma \qquad \Tm {N_2} \Gamma}
+      {\TmP {N_1 \ntimes N_2} \Gamma}
+\quad
+  \nru{\nunit}
+      {}
+      {\Tm \top \Gamma}
+\\[2ex]
+  \nru{\tforce}
+      {\PValP {\Thunk N} \Gamma}
+      {\Tm N \Gamma}
+\qquad
+  \nru{\tapp}
+      {\TmP {P \To N} \Gamma \qquad \PVal P \Gamma}
+      {\Tm N \Gamma}
+\qquad
+  \nru{\tprj_i}
+      {\TmP {N_1 \ntimes N_2} \Gamma}
+      {\Tm {N_i} \Gamma}
+\\[2ex]
+% \end{gather*}
+% \begin{gather*}
+  \nru{\tbind}
+      {\TmP {\Comp P} \Gamma \qquad \Tm N {(\cext \Gamma P)}}
+      {\Tm N \Gamma}
+\qquad
+  \nru{\tsplit}
+    {\TmP {P_1 \times P_2} \Gamma
+      \qquad \Tm N {(\cext {\cext \Gamma P_1} P_2)}
+    }{\Tm N \Gamma}
+\\[2ex]
+  \nru{\tcase}
+    {\TmP {P_1 + P_2} \Gamma
+      \qquad \Tm N {(\cext \Gamma P_1)}
+      \qquad \Tm N {(\cext \Gamma P_2)}
+    }{\Tm N \Gamma}
+\qquad
+  \nru{\tabort}
+    {\Tm \tyzero \Gamma}
+    {\Tm N \Gamma}
+\end{gather*}
+
+Normal forms:  $\PVal[\tNf]$.
+
+\fbox{$\Ne N \Gamma$}:
+\begin{gather*}
+  \nru{\nvar}{\Thunk N \in \Gamma}{\Ne N \Gamma}
+\qquad
+  \nru{\tprj_i}
+      {\NeP {N_1 \ntimes N_2} \Gamma}
+      {\Ne {N_i} \Gamma}
+\qquad
+  \nru{\tapp}
+      {\NeP {P \To N} \Gamma \qquad \PVal P \Gamma}
+      {\Ne N \Gamma}
+\end{gather*}
+Pattern matching on variables is organized in the
+inductively defined strong monad \fbox{$\Cov$}.
+\begin{gather*}
+  \nru{\treturn}{\J\,\Gamma}{\Cover \J \Gamma}
+\qquad
+  \nru{\tbind}
+      {\NeP {\Comp P} \Gamma \qquad \Cover \J (\cext \Gamma P)}
+      {\Cover \J \Gamma}
+\\[2ex]
+  \nru{\tsplit}
+      {P_1 \times P_2 \in \Gamma \qquad
+       \Cover \J (\cext {\cext \Gamma {P_1}} {P_2})
+      }
+      {\Cover \J \Gamma}
+\\[2ex]
+  \nru{\tcase}
+      {P_1 + P_2 \in \Gamma
+       \qquad \Cover \J (\cext \Gamma {P_1})
+       \qquad \Cover \J (\cext \Gamma {P_2})
+      }
+      {\Cover \J \Gamma}
+\qquad
+  \nru{\tabort}
+      {\tyempty \in \Gamma}
+      {\Cover \J \Gamma}
+\end{gather*}
+Finally, normal forms of negative types are defined as inductive
+family \fbox{$\Nf N \Gamma$}.
+\begin{gather*}
+  \nru{\tne}
+      {\CoverP {\tNe\,\natom} \Gamma}
+      {\Nf {\natom} \Gamma}
+\quad
+  \nru{\tret}
+      {\CoverP {\tVal\,P}  \Gamma}
+      {\NfP {\Comp P} \Gamma}
+\quad
+  \nru{\npair}
+      {\Nf {N_1} \Gamma \qquad \Nf {N_2} \Gamma}
+      {\NfP {N_1 \ntimes N_2} \Gamma}
+\quad
+  \nru{\tabs}
+      {\Nf N {\cext \Gamma P}}
+      {\NfP {P \To N} \Gamma}
+\end{gather*}
+We again can run the cover monad on normal forms, \ie, have
+$\trunNf : \Cov\,(\tNf N) \todot \tNf N$.
+
+Reification $\down P : \den P \todot \tVal\,P$ at positive types produces
+a normal value, and $\down N : \den P \todot \tNf\,N$ at negative types a
+normal term.  During reification of function types $P \To N$ in
+context $\Gamma$ we need to embed a fresh variable
+$x : P \in \cext \Gamma P$ into $\den P$, breaking down $P$ to
+positive atoms $\patom$ and negative remainders $\Thunk N$.
+However, in $\den P$ we do not have case analysis available, thus,
+positive reflection $\upG P : P \in \Gamma \to \Cover {\den P} \Gamma$
+needs to run in the monad.  Luckily, as $\down N$ produces a normal
+form, monadic intermediate computations are permitted under a final
+$\trunNf$.  Negative reflection $\up N : \tNe\,N \todot \den N$ is as
+before generalized from variables to neutrals, to handle the breaking
+down of $N$ via eliminations.
+\[
+\begin{array}{lcl}
+  \upG P & : & P \in \Gamma \to \den P \\
+  \upG {\patom} \; x & = & x \\
+  \upG {\tyone} \; x & = & \treturn\;\ttempty \\
+  \upG {P_1 \times P_2} \; x & = & \tsplit\;x\;\tzip
+\end{array}
+\]
+
+\section{Focused Intuitionistic Propositional Logic}
+\label{sec:fipl}
+
+Polarized lambda calculus \cite{espiritoSanto:entcs17}
+divides types deeply into positive and negative.
 
@@
 \]
-
-Pure CBPV adds right non-invertible focusing.
-
-Assuming a family $\T\,N\,\Gamma$ of terms of negative type $N$ in
-context $\Gamma$, values \fbox{$v : \PValT P \Gamma$}
-of positive type $P$ are constructed by the
-following rules:
-\begin{gather*}
-  \nru{\pvar}{P \in \Gamma}{\PValT P \Gamma}
-\qquad
-  \nru{\tthunk}{\T\,N\,\Gamma}{\PValTP{\Thunk N}\Gamma}
-\\[2ex]
-  \nru{\punit}{}{\PValT{\tyone}\Gamma}
-\qquad
-  \nru{\ppair}
-      {\PValT{P_1}\Gamma \qquad \PValT{P_2}\Gamma}
-      {\PValTP{P_1 \times P_2}\Gamma}
-\qquad
-  \nru{\tinj_i}{\PValT{P_i}\Gamma}{\PValTP{P_1 + P_2}\Gamma}
-\end{gather*}
-
-It relies on the previous definitions
-with instance $\T = \tNf$.
-\fbox{$\Tm N \Gamma$}:
-\begin{gather*}
-  \nru{\tret}
-      {\PVal P \Gamma}
-      {\TmP {\Comp P} \Gamma}
-\quad
-  \nru{\tabs}
-      {\Tm N {(\cext \Gamma P)}}
-      {\TmP {P \To N} \Gamma}
-\quad
-  \nru{\npair}
-      {\Tm {N_1} \Gamma \qquad \Tm {N_2} \Gamma}
-      {\TmP {N_1 \ntimes N_2} \Gamma}
-\quad
-  \nru{\nunit}
-      {}
-      {\Tm \top \Gamma}
-\\[2ex]
-  \nru{\tforce}
-      {\PValP {\Thunk N} \Gamma}
-      {\Tm N \Gamma}
-\qquad
-  \nru{\tapp}
-      {\TmP {P \To N} \Gamma \qquad \PVal P \Gamma}
-      {\Tm N \Gamma}
-\qquad
-  \nru{\tprj_i}
-      {\TmP {N_1 \ntimes N_2} \Gamma}
-      {\Tm {N_i} \Gamma}
-\\[2ex]
-% \end{gather*}
-% \begin{gather*}
-  \nru{\tbind}
-      {\TmP {\Comp P} \Gamma \qquad \Tm N {(\cext \Gamma P)}}
-      {\Tm N \Gamma}
-\qquad
-  \nru{\tsplit}
-    {\TmP {P_1 \times P_2} \Gamma
-      \qquad \Tm N {(\cext {\cext \Gamma P_1} P_2)}
-    }{\Tm N \Gamma}
-\\[2ex]
-  \nru{\tcase}
-    {\TmP {P_1 + P_2} \Gamma
-      \qquad \Tm N {(\cext \Gamma P_1)}
-      \qquad \Tm N {(\cext \Gamma P_2)}
-    }{\Tm N \Gamma}
-\qquad
-  \nru{\tabort}
-    {\Tm \tyzero \Gamma}
-    {\Tm N \Gamma}
-\end{gather*}
-
-\section{Focused Intuitionistic Propositional Logic}
-\label{sec:fipl}
-
-Polarized lambda calculus \cite{espiritoSanto:entcs17}
-divides types deeply into positive and negative.
-
-\[
-\begin{array}{lllrl@{\qquad}l}
-  \PTy & \ni & P,Q & ::= & \patom
-    \mid \tyzero \mid P_1 + P_2
-    \mid \tyone \mid P_1 \times P_2
-    \mid \Thunk N
-  & \mbox{positive type}
-\\
-  \NTy & \ni & N,M & ::= & \natom
-    \mid \top \mid N_1 \ntimes N_2
-    \mid P \To N
-    \mid \Comp P
-  & \mbox{negative type}
-\\
-  \HTy & \ni & H & ::= & \patom \mid P
-  & \mbox{hypothesis type}
-\\
-\end{array}
-\]
 Espirito Santo's version actually does not contain the positive

### 2019-02-11 fscd19/fscd19.tex
[ fscd19 ] reflection and reification for cbpv

@@
 rules of intuitionistic propositional logic (IPL) under the
-Curry-Howard isomorphism.  A term such as
+Curry-Howard isomorphism.  Using shorthand
+\fbox{$\vv n = \tvar\,(\tsuc^n\,\tzero)$} for the $n$th variable,
+a term such as
 \(
   \tabs\,(\tabs\,(\tpair\,
-    (\tvar\,(\tsuc\,\tzero))\,
-    (\tabs\,(\tapp\,(\tvar\,(\tsuc\,\tzero))\,(\tvar\,\tzero)))))
+    \vv1\,
+    (\tabs\,(\tapp\,\vv1\,\vv0))))
 \)
@@
 work.  For instance, with $\den \tyzero = \hat0$, simply
-$\up[\tyzero] \tyzero (\tvar\,\tzero) : \bzero$ would give us an
+$\up[\tyzero] \tyzero \vv0 : \bzero$ would give us an
 inhabitant of the empty set, which means that reflection at the empty
@@
 $\den{A + B} = \den A \hatplus \den B$ is refuted by
-% $\up[\tyone + \tyone]^{\tyone + \tyone} (\tvar\,\tzero) : \bone \bplus \bone$
-$\up[A + B]{A + B} (\tvar\,\tzero) : \den A (A{+}B) \bplus \den B (A{+}B)$
+% $\up[\tyone + \tyone]^{\tyone + \tyone} \vv0 : \bone \bplus \bone$
+$\up[A + B]{A + B} \vv0 : \den A (A{+}B) \bplus \den B (A{+}B)$
 which would require us to make a decision of whether $A$ holds or $B$ holds
@@
 the term
-$\tabs\,(\tcase\,(\tvar\,\tzero)\,(\tvar\,\tzero)\,(\tvar\,\tzero)) :
+$\tabs\,(\tcase\,\vv0\,\vv0\,\vv0) :
 (o+o) \To o$ in our semantics, as $\Ne o (o+o)$ is empty.
@@
   \nru{\retato}{}{t \cong
-    \tabs\,(\tapp\,(\tren\,\twk\,t)\,(\tvar\,\tzero))}
+    \tabs\,(\tapp\,(\tren\,\twk\,t)\,\vv0)}
 \quad
@@
     \tcase\,t
-      \,(\injP 1 {\tvar\,\tzero})
-      \,(\injP 2 {\tvar\,\tzero})
+      \,(\inj 1 {\vv0})
+      \,(\inj 2 {\vv0})
     }
@@
 \begin{gather*}
-  \nru{\pvar}{P \in \Gamma}{\PValT P \Gamma}
+  \nru{\tvar}{P \in \Gamma}{\PValT P \Gamma}
+  % \nru{\pvar}{P \in \Gamma}{\PValT P \Gamma}
 \qquad
@@
   \nru{\tsplit}
-    {\TmP {P_1 \times P_2} \Gamma
+    {\PValP {P_1 \times P_2} \Gamma
       \qquad \Tm N {(\cext {\cext \Gamma P_1} P_2)}
@@
   \nru{\tcase}
-    {\TmP {P_1 + P_2} \Gamma
+    {\PValP {P_1 + P_2} \Gamma
       \qquad \Tm N {(\cext \Gamma P_1)}
@@
   \nru{\tabort}
-    {\Tm \tyzero \Gamma}
+    {\PVal \tyzero \Gamma}
     {\Tm N \Gamma}
@@
 \begin{gather*}
-  \nru{\nvar}{\Thunk N \in \Gamma}{\Ne N \Gamma}
+  \nru{\tforce}{\Thunk N \in \Gamma}{\Ne N \Gamma}
+  % \nru{\nvar}{\Thunk N \in \Gamma}{\Ne N \Gamma}
 \qquad
@@
 down of $N$ via eliminations.
-\[
+%
+In the following definition of reflection we use the abbreviation
+\fbox{$\tfresh^P_\Gamma = \up[\cext \Gamma P] P \vv0 : \Cov\,\den P$}.
+\[
+\def\arraystretch{1.3}
+\begin{array}{l@{\,}l@{~~}c@{~~}l}
+  \multicolumn 2 l {\upG P} & : & P \in \Gamma \to \Cov\,\den P \, \Gamma\\
+  \upG {\patom}         & x & = & x \\
+  \upG {\tyone}         & x & = & \treturn\;\ttempty \\
+  \upG {P_1 \times P_2} & x & = & \tsplit\;x\,
+    \left(
+     (\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_1}\,\vv1)
+     \star
+     (\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_2}\,\vv0)
+    \right) \\
+    % (\tzip
+    % \,(\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_1}\,\vv1)
+    % \,(\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_2}\,\vv0)
+    % ) \\
+  \upG{\tyzero}         & x & = & \tabort\;x \\
+  \upG{P_1 + P_2}       & x & = & \tcase\;x
+    \; (\tmap\;\iota_1\;\tfresh^{P_1}_\Gamma)
+    \; (\tmap\;\iota_2\;\tfresh^{P_2}_\Gamma)
+    \\
+  \upG{\Thunk N} & x & = & \treturn\,(\upG N (\tforce\,x)) \\
+\end{array}
+\,
 \begin{array}{lcl}
-  \upG P & : & P \in \Gamma \to \den P \\
-  \upG {\patom} \; x & = & x \\
-  \upG {\tyone} \; x & = & \treturn\;\ttempty \\
-  \upG {P_1 \times P_2} \; x & = & \tsplit\;x\;\tzip
+  \down P & : & \den P \todot \tVal\,P \\
+  \down {\patom} & = & \tvar \\
+  \downG {\tyone} () & = & \punit \\
+  \downG {P_1 \times P_2}(a_1,a_2) & = & \ppair
+    \,(\downG {P_1} a_1)
+    \,(\downG {P_2} a_2)
+    \\
+  \down {\tyzero} & = & \tmagic \\
+  \down {P_1 + P_2} & = &
+    [  \tinj_1 \comp \down{P_1}
+    ,\ \tinj_2 \comp \down{P_2}
+    ] \\
+  \down {\Thunk N} & = & \tthunk \circ \down N \\
+\end{array}
+\]
+Reflection at positive pairs uses monoidal functoriality
+$\C\,\A \todot \C\,\B \todot \C\,(\A \hattimes \B)$
+called $\star$ by McBride and Paterson \cite[Section 7]{mcbridePaterson:applicative}
+which for monads $\C$ can be defined by
+$c \star d = \tjoin \; (\tmap\;(\lambda\,b.\ (c,\, \treturn\,b))\;d$.
+% $\tzip\;c = \tjoin \comp \tmap\,(\lambda\,a.\ (c,\, \treturn\,a)$.
+
+For negative product types, reflection and reification works as before:
+\[
+\def\arraystretch{1.3}
+\begin{array}{l@{\,}l@{~~}c@{~~}l}
+  \multicolumn 2 l {\up N} & : & \tNe\,N \todot \den N \\
+  \upG \natom & u & = & \treturn\,u \\
+  \upG \top   & u & = & \ttempty \\
+  \upG {N_1 \ntimes N_2} & u & = & \left(
+    \upG {N_1} \; (\prj 1 u)
+    ,\
+    \upG {N_2} \; (\prj 2 u)
+    \right) \\
+\end{array}
+\
+\begin{array}{l@{\,}l@{~~}c@{~~}l}
+  \multicolumn 2 l {\down N} & : & \den N \todot \tNf\;N \\
+  \downG \natom & c  & = & \tne\,c \\
+  \downG \top   & () & = & \nunit \\
+  \downG {N_1 \ntimes N_2} & (b_1,b_2) & = & \npair
+    \, (\downG {N_1} \; b_1)
+    \, (\downG {N_2} \; b_1)
+    \\
+\end{array}
+\]
+Reflection for function types is also unchanged, except that $\tapp$
+expects a value argument now.
+\[
+\def\arraystretch{1.3}
+\begin{array}{l@{\,}l@{~~}c@{~~}l}
+  \upG {P \To N} & u & = & \lambda
+    \;(\tau : \Gamma \sublist \Delta)
+    \;(a : \den P \Delta)
+    . \
+    \up[\Delta] N \left( \tapp\,(\tren\,\tau\,u)\,(\down[\Delta] P a) \right)
+    \\
+  \downG {P \To N} & f & = & \tabs \left( \trunNf \left(
+    \stmap
+      \left(
+        \lambda
+          \;(\tau : (\cext \Gamma P) \sublist \Delta)
+          \;(a : \den P \Delta)
+          . \
+          f\; (\twk^P \rcomp \tau)\; a
+        \right)
+      \,\tfresh^P_\Gamma
+    \right) \right)
+    \\
 \end{array}

### 2019-02-11 fscd19/fscd19.tex
[ fscd19 ] factor out fden from eval

@@
 can now be defined by recursion on $t$.
+% using an auxiliary definition
+% \fbox{$\fden{t : B \from \cext \Gamma A} : \den \Gamma \todot \den{A \To B}$}.
 Herein the environment $\gamma$ lives in $\den \Gamma \Delta$, thus, $\dent t \gamma : \den A \Delta$.
@@
 \begin{array}{lcl}
+  % \fdent{t : B \from \cext \Gamma A}\gamma & = &
+  %   \lambda
+  %     \,(\tau : \Delta \sublist \Phi)
+  %     \,(a : \den A \Phi)
+  %     .\
+  %     \dent t {(\tren\,\tau\,\gamma,\ a)} \\[1ex]
   \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
-  \dent{\tabs\,(t : B \from \cext \Gamma A)}\gamma \;
-     (\tau : \Delta \sublist \Phi)\, (a : \den A \Phi) & = &
-        \dent t {(\tren\,\tau\,\gamma,\ a)} \\
+  \dent{\tabs\,t}\gamma
+      & = & \fdent t \gamma \\
+  % \dent{\tabs\,(t : B \from \cext \Gamma A)}\gamma \;
+  %    (\tau : \Delta \sublist \Phi)\, (a : \den A \Phi) & = &
+  %       \dent t {(\tren\,\tau\,\gamma,\ a)} \\
   % \dent{\tabs\,(t : B \from \cext \Gamma A)}\gamma & = &
@@
   \dent{\tabort\,u}\gamma & = & \denabort\;\dent u \gamma \\
-  \dent{\tcase\,(u : A_1 + A_2 \from \Gamma)\;t_1\;t_2}\gamma & = &
-     \dencase\;\dent u \gamma\;f_1\;f_2
-    \ \bwhere \\
-    \multicolumn 1 r {f_i\; (\tau : \Delta \sublist \Phi)\,(a_i : \den {A_i} \Phi)}
-        & = &
-        \dent {t_i} {(\tren\;\tau\;\gamma,\ {a_i})}
+  \dent{\tcase\;u\;t_1\;t_2}\gamma & = &
+     \dencase\;\dent u \gamma\;\fdent{t_1}\gamma\;\fdent{t_2}\gamma \\
+  % \dent{\tcase\,(u : A_1 + A_2 \from \Gamma)\;t_1\;t_2}\gamma & = &
+    %  \dencase\;\dent u \gamma\;f_1\;f_2
+    % \ \bwhere \\
+    % \multicolumn 1 r {f_i\; (\tau : \Delta \sublist \Phi)\,(a_i : \den {A_i} \Phi)}
+    %     & = &
+    %     \dent {t_i} {(\tren\;\tau\;\gamma,\ {a_i})}
    % \
@@
     \\
+\end{array}
+\]
+For interpretation of the binders $\tabs$ and $\tcase$ we use the
+mutually defined $\fden{\_}$.
+\[
+\begin{array}{lcl}
+  \fden{t : B \from \cext \Gamma A}
+    & : & \den \Gamma \todot \den{A \To B} \\
+    & = &
+      \lambda\, (\gamma : \den \Gamma \Delta)
+      \,(\tau : \Delta \sublist \Phi)
+      \,(a : \den A \Phi)
+      .\ \,
+      \dent t {(\tren\,\tau\,\gamma,\ a)} \\
+  % \fden{\_} ~~:~~ B \from \cext \Gamma A} \to \den \Gamma \todot \den{A \To B}
+  % \\
+  % \fdent{t : B \from \cext \Gamma A}\gamma
+  %     \,(\tau : \Delta \sublist \Phi)
+  %     \,(a : \den A \Phi)
+  %     ~~=~~
+  %     \dent t {(\tren\,\tau\,\gamma,\ a)} \\
 \end{array}

### 2019-02-12 fscd19/fscd19.tex
[ fscd19 ] Evaluation for CBPV

@@
 Semantically, we do not distinguish between positive and negative
-products.  Notably, sum types can now be interpreted as plain presheaf
+products.  Notably, sum types can now be interpreted as plain
+(pointwise) presheaf
 sums.  The $\tThunk$ marker is ignored, yet $\tComp$, marking the
@@
 shall define below.
+The method $\trunNf[\C]$ of cover monad $\C$ can be extended to
+$\trun^N : \C \den N \to \den N$ for negative types $N$, by recursion
+on $N$.  Informally, speaking this makes all negative types \emph{monadic}.
+
+Contexts are lists of \emph{positive} types as in CBPV variables stand
+for values.  Interpretation of contexts $\den \Gamma$ is again defined
+pointwise $\den \cempty = \hatone$ and
+$\den{\cext \Gamma P} = \den \Gamma \hattimes \den P$.
 
@@
 context $\Gamma$, values \fbox{$v : \PValT P \Gamma$}
-of positive type $P$ are constructed by the
+of positive type $P$ shall be constructed by the
 following rules:
@@
 
-It relies on the previous definitions
-with instance $\T = \tNf$.
-\fbox{$\Tm N \Gamma$}:
+The terms of pure CBPV are given by the inductive family \fbox{$\Tm N \Gamma$}.
+It repeats the introductions and eliminations of negative types,
+except that application is restricted to values.  Values of type
+$\Thunk N$ are embedded via $\tforce$.
+Further, values of type $P$ can embedded
+via $\tret$, producing a term of type $\Comp P$.  Such terms are
+eliminated by $\tbind$ which is, unlike the usual monadic bind, not
+only available for $\tComp$-types but for arbitrary negative types
+$N$.  This is justified due to the monadic character of negative
+types, by virtue of $\trun^N$.  Finally, there are eliminators
+($\tsplit$, $\tcase$, $\tabort$) for values of positive product and
+sum types.
 \begin{gather*}
   \nru{\tret}
-      {\PVal P \Gamma}
+      {\PValTm P \Gamma}
       {\TmP {\Comp P} \Gamma}
@@
   \nru{\tforce}
-      {\PValP {\Thunk N} \Gamma}
+      {\PValTmP {\Thunk N} \Gamma}
       {\Tm N \Gamma}
@@
   \nru{\tapp}
-      {\TmP {P \To N} \Gamma \qquad \PVal P \Gamma}
+      {\TmP {P \To N} \Gamma \qquad \PValTm P \Gamma}
       {\Tm N \Gamma}
@@
   \nru{\tsplit}
-    {\PValP {P_1 \times P_2} \Gamma
+    {\PValTmP {P_1 \times P_2} \Gamma
       \qquad \Tm N {(\cext {\cext \Gamma P_1} P_2)}
@@
   \nru{\tcase}
-    {\PValP {P_1 + P_2} \Gamma
+    {\PValTmP {P_1 + P_2} \Gamma
       \qquad \Tm N {(\cext \Gamma P_1)}
@@
   \nru{\tabort}
-    {\PVal \tyzero \Gamma}
+    {\PValTm \tyzero \Gamma}
     {\Tm N \Gamma}
 \end{gather*}
+
+Interpretation of values
+$\dent{ v : \PValTm P \Gamma } : \den \Gamma \todot \den P$
+and terms
+$\dent{ t : \Tm N \Gamma } : \den \Gamma \todot \den N$
+is straightforward, thanks to the pioneering work of
+Moggi \cite{moggi:infcomp91} and
+Levy \cite{levy:hosc05}
+put into the design of CBPV.
+\[
+\begin{array}{lcl}
+  \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
+  \dent{\punit}\gamma   & = & \ttempty \\
+  \dent{\ppair\,v_1\,v_2}\gamma & = &
+    (\dent{v_1}\gamma,\ \dent{v_2}\gamma) \\
+  \dent{\inj i v}\gamma & = & \iota_i \dent v \gamma \\
+  \dent{\tthunk\,t}\gamma & = & \dent t \gamma \\
+\end{array}
+\]
+Since $\tThunk$ serves only as an embedding of negative into positive
+types and has no semantic effect,
+we interpret thunking and forcing by the identity.
+\[
+\begin{array}{lcl}
+  \dent{\tforce\,v}\gamma & = & \dent v \gamma \\
+  \dent{\tabs\,t}\gamma
+      & = & \fdent t \gamma \\
+  \dent{\tapp\,t\,v} \gamma & = &
+     \dent t \gamma\;\tid\;\dent v \gamma \\
+  \dent{\punit}\gamma & = & \ttempty \\
+  \dent{\ppair\,t_1\,t_2}\gamma & = &
+    (\dent{t_1}\gamma,\ \dent{t_2}\gamma) \\
+  \dent{\prj i t}\gamma & = & \pi_i \dent t \gamma \\
+\end{array}
+\]
+\[
+\begin{array}{lcl}
+  \dent{\tret\,v}\gamma & = & \treturn[\C] \; \dent v \gamma \\
+  \dent{\tbind\,u\,t}\gamma & = & \trun\,
+    (\stmap[\C] \; \fdent t \gamma\; \dent u \gamma) \\
+  \dent{\tsplit\,v\,t}\gamma & = & (\lambda\,(a_1,a_2).\
+    \dent{t}{(\gamma,a_1,a_2)} ) \; \dent v \gamma \\
+  \dent{\tcase\,v\,t_1\,t_2}\gamma & = &
+    [  \lambda a_1.\ \dent{t_1}{(\gamma,a_1)}
+    ,\ \lambda a_2.\ \dent{t_2}{(\gamma,a_2)}
+    ]\; \dent v \gamma \\
+  \dent{\tabort\,v}\gamma & = & \tmagic\,\dent v \gamma \\
+\end{array}
+\]
 

### 2019-02-12 fscd19/fscd19.tex
[ fscd19 ] normalization function for cbpv

@@
 corresponds to the translation of call-by-name (CBN) lambda-calculus
-into Moggi's monadic meta-language \cite{moggi:infcomp91,levy:hosc05}.
-
-It would be sufficient, however, to perform all necessary
+into Moggi's monadic meta-language \cite{moggi:infcomp91,levy:hosc06}.
+
+It would be sufficient %, however,
+to perform all necessary
 case distinctions
@@
 clearly spelled out in Levy's call-by-push-value (CBPV)
-\cite{levy:hosc05}, as it comes with a deep classification of types
+\cite{levy:hosc06}, as it comes with a deep classification of types
 into positive and negative ones.  In the following, we shall utilize
@@
 introductions.
+
+\subsection{Types and polarization}
 
@@
 $\den{\cext \Gamma P} = \den \Gamma \hattimes \den P$.
+
+\subsection{Terms and evaluation}
 
@@
 Interpretation of values
-$\dent{ v : \PValTm P \Gamma } : \den \Gamma \todot \den P$
+$\denty{ v : \PValTm P \Gamma } : \den \Gamma \todot \den P$
 and terms
-$\dent{ t : \Tm N \Gamma } : \den \Gamma \todot \den N$
+$\denty{ t : \Tm N \Gamma } : \den \Gamma \todot \den N$
 is straightforward, thanks to the pioneering work of
 Moggi \cite{moggi:infcomp91} and
-Levy \cite{levy:hosc05}
+Levy \cite{levy:hosc06}
 put into the design of CBPV.
@@
 \begin{array}{lcl}
+  \\
   \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
@@
 \end{array}
-\]
-Since $\tThunk$ serves only as an embedding of negative into positive
-types and has no semantic effect,
-we interpret thunking and forcing by the identity.
-\[
+\quad
 \begin{array}{lcl}
-  \dent{\tforce\,v}\gamma & = & \dent v \gamma \\
   \dent{\tabs\,t}\gamma
@@
      \dent t \gamma\;\tid\;\dent v \gamma \\
-  \dent{\punit}\gamma & = & \ttempty \\
-  \dent{\ppair\,t_1\,t_2}\gamma & = &
+  \dent{\nunit}\gamma & = & \ttempty \\
+  \dent{\npair\,t_1\,t_2}\gamma & = &
     (\dent{t_1}\gamma,\ \dent{t_2}\gamma) \\
   \dent{\prj i t}\gamma & = & \pi_i \dent t \gamma \\
-\end{array}
-\]
+  \dent{\tforce\,v}\gamma & = & \dent v \gamma \\
+\end{array}
+\]
+Since $\tThunk$ serves only as an embedding of negative into positive
+types and has no semantic effect,
+we interpret thunking and forcing by the identity.
+The eliminations for positive types deal now only with values, thus,
+need not reference the monad operations.
 \[
 \begin{array}{lcl}
-  \dent{\tret\,v}\gamma & = & \treturn[\C] \; \dent v \gamma \\
-  \dent{\tbind\,u\,t}\gamma & = & \trun\,
-    (\stmap[\C] \; \fdent t \gamma\; \dent u \gamma) \\
   \dent{\tsplit\,v\,t}\gamma & = & (\lambda\,(a_1,a_2).\
@@
 \]
-
-Normal forms:  $\PVal[\tNf]$.
-
-\fbox{$\Ne N \Gamma$}:
+The use of the monad is confined to $\tret$ and $\tbind$.
+Note the availability of $\trun^\C : \C \den N \to \den N$
+at any negative type $N$ for the interpretation of $\tbind$.
+\[
+\begin{array}{lcl}
+  \dent{\tret\,v}\gamma & = & \treturn[\C] \; \dent v \gamma \\
+  \dent{\tbind\,u\,t}\gamma & = & \trun^\C\,
+    (\stmap[\C] \; \fdent t \gamma\; \dent u \gamma) \\
+\end{array}
+\]
+
+\subsection{Normal forms and normalization}
+
+Positive normal forms are values $v : \PVal[\tNf]$ whose thunks only
+contain negative normal forms.
+
+Neutral normal forms \fbox{$\Ne N \Gamma$} are negative eliminations
+starting from a forced $\tThunk$ rather than from variables of negative
+types (as those do not exist in CBPV).  However, due to normality
+the $\tThunk$ cannot be a $\tthunk$, but only a variable
+$\Thunk N \in \Gamma$.
 \begin{gather*}
@@
   \nru{\tapp}
-      {\NeP {P \To N} \Gamma \qquad \PVal P \Gamma}
+      {\NeP {P \To N} \Gamma \qquad \PValNf P \Gamma}
       {\Ne N \Gamma}
 \end{gather*}
-Pattern matching on variables is organized in the
+Variables are originally introduced by either $\tabs$ or
+the $\tbind$ind of a neutral of type $\Comp P$ to a new variable of type
+$P$.  Variables of composite value type can be broken down by pattern
+matching, introducing variables of smaller type.  These positive
+eliminations plus $\tbind$ are organized in the
 inductively defined strong monad \fbox{$\Cov$}.
@@
 Finally, normal forms of negative types are defined as inductive
-family \fbox{$\Nf N \Gamma$}.
+family \fbox{$\Nf N \Gamma$}.  They are generated by maximal negative
+introduction ($\tabs$, $\npair$, $\nunit$) until a negative atom or
+$\Comp P$ is reached.  Then, elimination of neutrals and variables is
+possible through the $\Cov$ monad until an answer can be given in form
+of a base neutral ($\tNe\,\natom$) or a normal value.
 \begin{gather*}
@@
       {\Nf {\natom} \Gamma}
-\quad
+\qquad
   \nru{\tret}
-      {\CoverP {\tVal\,P}  \Gamma}
+      {\CoverP {\tVal[\tNf]\,P}  \Gamma}
       {\NfP {\Comp P} \Gamma}
-\quad
+\\[2ex]
+  \nru{\nunit}{}{\Nf \top \Gamma}
+\qquad
   \nru{\npair}
@@
       {\NfP {N_1 \ntimes N_2} \Gamma}
-\quad
+\qquad
   \nru{\tabs}
@@
 We again can run the cover monad on normal forms, \ie, have
-$\trunNf : \Cov\,(\tNf N) \todot \tNf N$.
-
-Reification $\down P : \den P \todot \tVal\,P$ at positive types produces
+$\trunNf : \Cov\,(\tNf N) \todot \tNf N$, which extends to
+negative semantic values
+$\trun : \Cov \den N \todot \den N$.
+
+Reification $\down P : \den P \todot \tVal[\tNf]\,P$ at positive types produces
 a normal value, and $\down N : \den P \todot \tNf\,N$ at negative types a
@@
 context $\Gamma$ we need to embed a fresh variable
-$x : P \in \cext \Gamma P$ into $\den P$, breaking down $P$ to
+$x : P \in (\cext \Gamma P)$ into $\den P$, breaking down $P$ to
 positive atoms $\patom$ and negative remainders $\Thunk N$.
@@
 In the following definition of reflection we use the abbreviation
-\fbox{$\tfresh^P_\Gamma = \up[\cext \Gamma P] P \vv0 : \Cov\,\den P$}.
+\fbox{\(
+  \tfresh^P_\Gamma
+    = \up[\cext \Gamma P] P \vv0
+    : \Cover{\den P}{(\cext \Gamma P)}
+\)}.
 \[
@@
 \begin{array}{lcl}
-  \down P & : & \den P \todot \tVal\,P \\
+  \down P & : & \den P \todot \tVal[\tNf]\,P \\
   \down {\patom} & = & \tvar \\
@@
 Reflection at positive pairs uses monoidal functoriality
-$\C\,\A \todot \C\,\B \todot \C\,(\A \hattimes \B)$
+$\C\,\A_1 \todot \C\,\A_2 \todot \C\,(\A_1 \hattimes \A_2)$
 called $\star$ by McBride and Paterson \cite[Section 7]{mcbridePaterson:applicative}
 which for monads $\C$ can be defined by
-$c \star d = \tjoin \; (\tmap\;(\lambda\,b.\ (c,\, \treturn\,b))\;d$.
+$c_1 \star c_2 =
+   \tjoin \; (\tmap\;(\lambda\,a_1.\
+   \tmap\;(\lambda\,a_2.\ (a_1,a_2))
+    \;c_2)
+    \;c_1)$.
+% $c_1 \star c_2 =
+%    \tjoin \; (\tmap\;(\lambda\,a_1.\
+%    \tjoin \; (\tmap\;(\lambda\,a_2.\ \treturn\,(a_1,a_2))
+%     \;c_2))
+%     \;c_1)$.
+% $c \star d = \tjoin \; (\tmap\;(\lambda\,b.\ (c,\, \treturn\,b))\;d$.
 % $\tzip\;c = \tjoin \comp \tmap\,(\lambda\,a.\ (c,\, \treturn\,a)$.
@@
 \]
+The identity environment $\idenv \Gamma : \Cover {\den \Gamma} \Gamma$
+can only be generated in the monad, due to monadic positive
+reflection.
+\[
+\begin{array}{lcl}
+  \idenv{\cempty} & = & \treturn\,() \\
+  \idenv{\cext \Gamma P} & = &
+    (\tren\,\twk^P\,\idenv\Gamma) \star \tfresh^P_\Gamma \\
+\end{array}
+\]
+Putting things together, we obtain the normalization function
+\[
+  \tnorm\,(t : \Tm N \Gamma) = \trunNf\,
+    \left(
+      \tmap\,\left(\down N \comp \denty t\right)\,\idenv\Gamma
+    \right)
+.
+\]
+
+Taking stock, we have arrived at normal forms that eagerly introduce
+($\tNf$) and eliminate ($\tNe$) negative types and also eagerly
+introduce positive types ($\tVal$).  However, the elimination of
+positive types is still rather non-deterministic.
+It is possible to only partially break up a composite positive type
+and leave smaller, but still composite positive types for later
+pattern matching.
+
+The placement of $\tsplit$, $\tcase$, and $\tabort$ is only
+constrained by the presence of the split variable
 

### 2019-02-12 fscd19/fscd19.tex
[ fscd19 ] reflection for polarized lambda

@@
 
-Assuming a family $\T\,N\,\Gamma$ of terms of negative type $N$ in
+Assuming a family $\Tm N \Gamma$ of terms of negative type $N$ in
 context $\Gamma$, values \fbox{$v : \PValT P \Gamma$}
@@
 \qquad
-  \nru{\tthunk}{\T\,N\,\Gamma}{\PValTP{\Thunk N}\Gamma}
+  \nru{\tthunk}{\Tm N \Gamma}{\PValTP{\Thunk N}\Gamma}
 \\[2ex]
@@
 
-Positive normal forms are values $v : \PVal[\tNf]$ whose thunks only
+Positive normal forms are values $v : \VNf P \Gamma$
+referring only to atomic variables and
+whose thunks only
 contain negative normal forms.
+% Positive normal forms are values $v : \PVal[\tNf]$ whose thunks only
+% contain negative normal forms.
+\begin{gather*}
+  \nru{\tvar}{\patom \in \Gamma}{\PValNf {\patom} \Gamma}
+  % \nru{\pvar}{P \in \Gamma}{\PValNf P \Gamma}
+\qquad
+  \nru{\tthunk}{\Nf N \Gamma}{\PValNfP{\Thunk N}\Gamma}
+\\[2ex]
+  \nru{\punit}{}{\PValNf{\tyone}\Gamma}
+\qquad
+  \nru{\ppair}
+      {\PValNf{P_1}\Gamma \qquad \PValNf{P_2}\Gamma}
+      {\PValNfP{P_1 \times P_2}\Gamma}
+\qquad
+  \nru{\tinj_i}{\PValNf{P_i}\Gamma}{\PValNfP{P_1 + P_2}\Gamma}
+\end{gather*}
 
@@
 Variables are originally introduced by either $\tabs$ or
-the $\tbind$ind of a neutral of type $\Comp P$ to a new variable of type
+the $\tbind$ing of a neutral of type $\Comp P$ to a new variable of type
 $P$.  Variables of composite value type can be broken down by pattern
@@
   \nru{\tret}
-      {\CoverP {\tVal[\tNf]\,P}  \Gamma}
+      {\CoverP {\tVNf\,P}  \Gamma}
       {\NfP {\Comp P} \Gamma}
@@
 
-Reification $\down P : \den P \todot \tVal[\tNf]\,P$ at positive types produces
+Reification $\down P : \den P \todot \tVNf\,P$ at positive types produces
 a normal value, and $\down N : \den P \todot \tNf\,N$ at negative types a
@@
 \begin{array}{lcl}
-  \down P & : & \den P \todot \tVal[\tNf]\,P \\
+  \down P & : & \den P \todot \tVNf\,P \\
   \down {\patom} & = & \tvar \\
@@
   \multicolumn 2 l {\up N} & : & \tNe\,N \todot \den N \\
+  \upG {\Comp P} & u & = & \tbind\,u\,\tfresh^P_\Gamma \\
   \upG \natom    & u & = & \treturn\,u \\
@@
   \multicolumn 2 l {\down N} & : & \den N \todot \tNf\;N \\
+  \downG {\Comp P} & c  & = & \tmap\;(\down P)\;c \\
   \downG \natom    & c  & = & \tne\,c \\
@@
     \, (\downG {N_1} \; b_1)
-    \, (\downG {N_2} \; b_1)
+    \, (\downG {N_2} \; b_2)
     \\
@@
 ($\tNf$) and eliminate ($\tNe$) negative types and also eagerly
-introduce positive types ($\tVal$).  However, the elimination of
+introduce positive types ($\tVnf$).  However, the elimination of
 positive types is still rather non-deterministic.
@@
 pattern matching.
-
-The placement of $\tsplit$, $\tcase$, and $\tabort$ is only
-constrained by the presence of the split variable
+The last refinement, chaining also the positive eliminations, will be
+discussed in the following section.
+% The placement of $\tsplit$, $\tcase$, and $\tabort$ is only
+% constrained by the presence of the split variable
 
@@
 
-Polarized lambda calculus \cite{espiritoSanto:entcs17}
-divides types deeply into positive and negative.
-
-\[
-\begin{array}{lllrl@{\qquad}l}
-  \PTy & \ni & P,Q & ::= & \patom
-    \mid \tyzero \mid P_1 + P_2
-    \mid \tyone \mid P_1 \times P_2
-    \mid \Thunk N
-  & \mbox{positive type}
-\\
-  \NTy & \ni & N,M & ::= & \natom
-    \mid \top \mid N_1 \ntimes N_2
-    \mid P \To N
-    \mid \Comp P
-  & \mbox{negative type}
-\\
-  \HTy & \ni & H & ::= & \patom \mid P
-  & \mbox{hypothesis type}
-\\
-\end{array}
-\]
-Espirito Santo's version actually does not contain the positive
-product $\times$, however, it appears in Zeilberger
-\cite{zeilberger:PhD}.
-
-Polarized lambda-calculus is a focused calculus, it eagerly employs
+% Polarized lambda calculus \cite{espiritoSanto:entcs17}
+% divides types deeply into positive and negative.
+
+% \[
+% \begin{array}{lllrl@{\qquad}l}
+%   \PTy & \ni & P,Q & ::= & \patom
+%     \mid \tyzero \mid P_1 + P_2
+%     \mid \tyone \mid P_1 \times P_2
+%     \mid \Thunk N
+%   & \mbox{positive type}
+% \\
+%   \NTy & \ni & N,M & ::= & \natom
+%     \mid \top \mid N_1 \ntimes N_2
+%     \mid P \To N
+%     \mid \Comp P
+%   & \mbox{negative type}
+% \\
+%   \HTy & \ni & H & ::= & \patom \mid N
+%   & \mbox{hypothesis type}
+% \\
+% \end{array}
+% \]
+% Espirito Santo's version actually does not contain the positive
+% product $\times$, however, it appears in Zeilberger
+% \cite{zeilberger:PhD}.
+
+Polarized lambda-calculus
+\cite{zeilberger:PhD,espiritoSanto:entcs17}
+is a focused calculus, it eagerly employs
 invertible rules: the introduction rules for negative types and the
 elimination rules for positive types.  As a consequence of the latter,
-variables are either of atomic or negative type ($\HTy$).
-Contexts $\Gamma,\Delta$ are lists of $\HTy$s.
+variables are either of atomic or negative type $H ::= \patom \mid N$.
+%($\HTy$).
+Contexts $\Gamma,\Delta$ are lists of $H$s.
+%$\HTy$s.
 %\fbox{$X ::= \patom \mid P$}.
@@
 however, this coincidence shall not matter for our further considerations.
+
+Focalization is a technique to remove don't-care non-determinism from
+proof search, and as such, polarized lambda calculus is foremost a
+calculus of normal forms.
+These normal forms are given by four mutually defined inductive
+families of presheaves $\tVnf\,P$,
+$\tNe\,N$, $\Cov\,\J$,
+and $\tNf\,N$.  As they are very similar to the CBPV normal forms
+given in the last section we only report the differences.
+
+Values \fbox{$v : \PVal P \Gamma$} are unchanged, they can refer to
+atomic positive hypotheses ($\pvar$) and normal $\tthunk$s.
+
+Neutrals \fbox{$\Ne N \Gamma$} start with a negative variable instead
+of with $\tforce$, as forcing thunks is already performed in $\nhyp$
+when adding analysing hypotheses of $\tThunk$ type.
+\[
+  \nru{\nvar}{N \in \Gamma}{\Ne N \Gamma}
+\]
+
+The Cover monad \fbox{$\Cover \J \Gamma$}
+lacks constructors $\tsplit$, $\tcase$ and $\tabort$
+since the pattern matching is taken care of by $\Add P$.
+\begin{gather*}
+  \nru{\treturn}{\J\,\Gamma}{\Cover \J \Gamma}
+\qquad
+  \nru{\tbind}
+      {\NeP {\Comp P} \Gamma \qquad \AddHypP P {\Cov\,\J} \Gamma}
+      {\Cover \J \Gamma}
+\end{gather*}
+
+The normal forms \fbox{$\Nf N \Gamma$} of negative type are unchanged
+with the exception that pattern matching happens eagerly in $\tabs$.
+\[
+  \nru{\tabs}
+      {\AddHypP P {\tNf\, N} \Gamma}
+      {\NfP {P \To N} \Gamma}
+\]
+
+All these inductive families are presheaves, due to factored
+presentation using $\Add P$ and $\Cov$ the proof is not a simple
+mutual induction.  Yet, in Agda, the generic proof goes through using a sized
+typing for these inductive families.  Similarily, defining the
+$\tjoin$ for monad $\Cov$ relies on sized typing.
+\[
+\begin{array}{lll}
+  \tjoin & : & \forall i.\
+    \Cov[i]\,(\Cov[\infty]\J) \todot \Cov[\infty]\J  \\
+  \tjoin[i+1]\,(\treturn[i]\;c) & = & c \\
+  \tjoin[i+1]\,(\tbind^i\;t\;k)
+    & = & \tbind^\infty\;
+     (t : \Ne P \Gamma)\;
+     \left(\stmap[\Add P]_\Gamma\;\tjoin[i]\;
+       \left(k : \AddHypP P {\Cov[i]\J} \Gamma
+       \right)
+     \right)
+\end{array}
+\]
+Herein, we used the sized typing of the constructors of $\Cov$:
+\[
+\begin{array}{lll}
+  \treturn & : & \forall i.\ \J \todot \Cov[i+1] \J \\
+  \tbind  & : & \forall i.\
+    \tNe\,(\Comp P) \todot \Add P\,(\Cov[i]\J) \todot \Cov[i+1]\J \\
+\end{array}
+\]
+
+Due to the eager splitting of positive hypotheses, reflection at type $P$
+now lives in the $\Add P$ graded monad rather than $\Cov$.  Further,
+as pattern matching may produce $n \geq 0$ cases, reflection cannot simply
+produce one positive semantic value; instead, one such value is needed
+for every branch.  We implement
+\fbox{$\reflect P : (\den P \hatto \J) \todot \Add P \J$}
+as a higher-order function expecting a continuation $k$ which is
+invoked for each generated branch with the semantic value of type $P$
+constructed for this branch.
+\[
+\begin{array}{l@{~}lcl}
+  \reflect {\patom} & k & = &
+    \phyp \left( k\;\twk^{\patom}\,(\pvar\,\tzero) \right) \\
+  \reflect {\Thunk N} & k & = &
+    \nhyp \left( k\;\twk^N \left( \up N (\nvar\,\tzero) \right) \right)\\
+  \reflect {\tyzero} & k & = & \nbranch \\
+  \reflect {P_1 + P_2} & k & = & \bbranch
+    \,\left( \treflect^{P_1}\,(\lambda\,\tau. \ k\,\tau \comp \iota_1) \right)
+    \,\left( \treflect^{P_2}\,(\lambda\,\tau. \ k\,\tau \comp \iota_2) \right)
+    \\
+  \reflect{\tyone} & k & = & \nsplit\;(k\;\tid\;\ttempty) \\
+  \reflect{P_1 \times P_2} & k & = & \bsplit
+    \,\left(
+      \treflect^{P_1} \left( \lambda\,\tau_1\,a_1. \
+      \treflect^{P_2} \left( \lambda\,\tau_2\,a_2. \
+        k\, (\tau_1 \rcomp \tau_2)\, (\tren\,\tau_2\,a_1, \ a_2)
+        \right)
+      \right)
+    \right)
+\end{array}
+\]
+Reflecting at a positive atomic type $\patom$ is the regular ending of a
+reflection pass: we call continuation $k$ with a fresh variable
+$\pvar\,\tzero$ of type $\patom$, making space for the variable
+using $\twk^{\patom}$.
+In case we end at type $\Thunk N$, we add a new variable
+$\nvar\,\tzero$ of type $N$ and pass it to $k$, after full
+$\eta$-expansion via $\up N$.  Two more endings are possible:
+At type
+$\tyzero$, we have reached an absurd case, meaning that no
+continuation is necessary since we can conclude with
+\emph{ex falsum quod libet}.
+At type $\tyone$, there is no need to add a new variable, as values
+of type $\tyone$ contain no information.  We simply pass the unit
+value $\ttempty$ to $k$ in this case.
+Reflecting at $P_1 + P_2$ generates two branches, which may result in
+several uses of the continuation $k$.
+In the first branch, we recursively reflect at $P_1$.  Its
+continuation will receive a semantic value in $\den{P_1}$ which we
+inject via $\iota_1$ into $\den{P_1 + P_2}$ to pass it to $k$.  The
+second branch proceeds analogously.
+Finally reflecting at $P_1 \times P_2$ means we first have to analyze
+$P_1$, and in each of the generated branches we continue to analyze
+$P_2$.  Thus $\reflect{P_2}$ is passed as a continuation to
+$\reflect{P_1}$.  Each reflection phase gives us a semantic value
+$a_i$ of type $P_i$, which we combine to a tuple before passing it to
+$k$.  Note also that the context extension $\tau_1$ created in the
+first phase needs to be composed with the context extension $\tau_2$
+of the second phase to transport $k$ into the final context.  Further,
+the value $a_1$ was constructed relative to the target of
+$\tau_1$ and still needs
+to be transported with $\tau_2$ before paired up with $a_2$.
+
+\clearpage
+
+\subsection{Old, explicit syntax}
 
@@
   \nru{\tval}
-      {\Cov[\tNf](\tVal[\tNf]\,P)\,\Gamma}
+      {\Cov[\tNf](\tVNf\,P)\,\Gamma}
       {\NfP {\Comp P} \Gamma}

### 2019-02-12 fscd19/fscd19.tex
[ fscd19 ] uses of reflect^P

@@
 Due to the eager splitting of positive hypotheses, reflection at type $P$
-now lives in the $\Add P$ graded monad rather than $\Cov$.  Further,
+now lives in the graded monad $\Add P$ rather than $\Cov$.  Further,
 as pattern matching may produce $n \geq 0$ cases, reflection cannot simply
@@
 
+The method $\reflect P$ replaces previous uses of $\tfresh^P$ in
+reflection and reification at negative types.
+% and identity environment creation.
+\[
+\begin{array}{l@{~}lcl}
+  \downG {P \To N} & f & = & \tabs
+    \left(
+      \reflect[\Gamma] P
+        \left(
+          \lambda \, (\tau : \Gamma \sublist \Delta) \; a .\
+          \down[\Delta] P (f \; \tau \; a)
+        \right)
+    \right)
+  \\
+  \upG {\Comp P} & u & = & \tbind \; u
+    \left(
+      \reflect[\Gamma] P
+        \left(
+          \lambda \; \tau \; a .\
+          \treturn\; a
+        \right)
+    \right)
+  % \\
+  % \idenv{\cext \Gamma P} & = &
+\end{array}
+\]
+% To create an identity environment using $\treflect$, let us first
+% extend $\Add\_$ to contexts, following the intuition that
+Due to the strong limitation on context entries,
+the identity environment $\idenv \Gamma$ can be built
+straightforwardly using negative reflection.
+\[
+\begin{array}{lcl}
+  \idenv{\cempty} & = & \ttempty \\
+  \idenv{\cext \Gamma {\patom}} & = &
+    \left(
+      \tren\;\twk^{\patom}\;\idenv\Gamma
+      , \
+      \pvar\,\tzero
+    \right) \\
+  \idenv{\cext \Gamma N} & = &
+    \left(
+      \tren\;\twk^N\;\idenv\Gamma
+      , \
+      \upG N (\nvar\,\tzero)
+    \right) \\
+\end{array}
+\]
+
 \clearpage

### 2019-02-13 fscd19/fscd19.tex
[ fscd19 ] Evaluation for polarized lambda

@@
 
+Structural focalization for sequent calculus \cite{simmons:tocl14}.
+
 \nocite{bergerSchwichtenberg:lics91}
@@
 \emph{sets}.  Consequently, the colon $:$
-takes the role of elementhood $\in$ in set-theory, and we are free to
+takes the role of elementhood $\in$ in set theory, and we are free to
 reuse the symbol $\in$ for other purposes.
@@
 
+The terms \fbox{$\Tm N \Gamma$} of the focused lambda calculus are the
+ones of CBPV minus the positive eliminations ($\tsplit$, $\tcase$,
+$\tabort$), the added negative variable rule ($\nvar$) and the
+necessary changes to the binders $\tabs$ and $\tbind$.
+\begin{gather*}
+  \nru{\nvar}{N \in \Gamma}{\Tm N \Gamma}
+\qquad
+  \nru{\tabs}
+      {\AddHypP P {\tTm\, N} \Gamma}
+      {\TmP {P \To N} \Gamma}
+\qquad
+  \nru{\tbind}
+      {\Tm {\Comp P} \Gamma \qquad \AddHypP P {\tTm\,N} \Gamma}
+      {\Tm N \Gamma}
+\end{gather*}
+Term interpretation
+\fbox{$\denty\_ : \Tm N \Gamma \to \den \Gamma \todot \den N$}
+shall be as for CBPV except that we need to exchange the
+interpretation function for binders \fbox{$\fden\_ : \Tm N {(\cext
+    \Gamma P)} \to \den \Gamma \todot \den{P \To N}$}.  Since a binder
+for $P$ performs a maximal splitting on $P$ and takes the form of a
+function defined by a case (and split) tree,
+applying it to a value $v$ of type $P$ amounts to a complete
+matching of $v$ against the case tree, and bind the remaining
+atomic and negative crumbs.  This matching can be defined for a
+generic evaluation function of type
+$\Evvv \J \Delta \Gamma = \den \Gamma \Delta \to \J\,\Delta$.
+\[
+\begin{array}{l@{~}l@{~}l@{~}lcl}
+  \multicolumn 6 {l} {\tmatch ~~:~~ \den P \Delta \to
+    \AddHyp P {\Evv \J \Delta} \todot \Evv \J \Delta }\\
+  \tmatch & x & (\phyp\,e) & \gamma & = & e\,(\gamma,\; \treturn\,x) \\
+  \tmatch & b & (\nhyp\,e) & \gamma & = & e\,(\gamma,\; b) \\
+  \tmatch & \ttempty & (\nsplit\,e) & \gamma & = & e\,\gamma \\
+  \tmatch & (a_1,a_2) & (\bsplit\,e) & \gamma & = &
+    \tmatch\;a_1\;(\tmap[\Add P]\,(\tmatch\;a_2)\;e)\;\gamma \\
+  \tmatch & a & \nbranch & \gamma & = & \tmagic\,a \\
+  \tmatch & (\iota_1\;a_1) & (\bbranch\;e_1\;e_2) & \gamma & = &
+    \tmatch\;a_1\;e_1\;\gamma \\
+  \tmatch & (\iota_2\;a_2) & (\bbranch\;e_1\;e_2) & \gamma & = &
+    \tmatch\;a_2\;e_2\;\gamma \\
+\end{array}
+\]
+With instantiations $\J = \den N$ and
+$\denty\_ : \Tm N \todot \Evv {\den N} \Delta$,
+the interpretation $\fden t$
+of binder $t : \AddHypP P {\tTm\,N} \Gamma$
+is defined as follows:
+\[
+  \fdent{t}
+    {(\gamma : \den \Gamma \Delta)}
+    \, (\tau : \Delta \sublist \Phi)
+    \, (a : \den P \Phi)
+    ~~=~~
+    \tmatch
+      \; a
+      \; (\tmap[\Add P]\;\denty\_\;t)
+      \; (\tren\,\tau\,\gamma)
+\]
+This completes the definition of the normalization function
+$\tnorm\, (t : \Tm N \Gamma)
+  = \down N \dent t {\idenv\Gamma}$.
+
+\section{Conclusion}
+\label{sec:concl}
+
+\bibliography{auto-fscd19}
+
+% \appendix
+
+\end{document}
+
 \clearpage
@@
 \end{gather*}
-
-\section{Conclusion}
-\label{sec:concl}
-
-\bibliography{auto-fscd19}
-
-% \appendix
-
-\end{document}

### 2019-02-13 fscd19/fscd19.tex
[ fscd19 ] Moved term equality to appendix

@@
 
-\emph{Proof normalization} is concerned with finding a normal form
-$n : \Nf A \Gamma$ for each derivation $t : A \from \Gamma$.  However,
-\emph{term normalization} additionally requires \emph{soundness},
-\ie, that $\tm u \cong t$
-with respect to a equational theory $\cong$ on terms.
-
-\subsection{Proof normalization}
-
-For didactical purposes, let us first focus on the simpler task of
-constructing a normalization function
+% \emph{Proof normalization} is concerned with finding a normal form
+% $n : \Nf A \Gamma$ for each derivation $t : A \from \Gamma$.  However,
+% \emph{term normalization} additionally requires \emph{soundness},
+% \ie, that $\tm u \cong t$
+% with respect to a equational theory $\cong$ on terms.
+
+% \subsection{Proof normalization}
+
+% For didactical purposes, let us first focus on the simpler task of
+% constructing a normalization function
+% \fbox{$\tnorm : A \from \Gamma \to \Nf A \Gamma$}
+% without any soundness guarantee.
+
+\subsection{Normalization}
+
+%\emph{Normalization}
+Normalization is concerned with finding a normal form
+$n : \Nf A \Gamma$ for each term $t : A \from \Gamma$.
+The normal form should be \emph{sound},
+\ie, $\tm n \cong t$
+with respect to a equational theory $\cong$ on terms
+(see \cref{sec:eq}).
+Further, normalization should decide $\cong$, \ie,
+terms $t,t'$ with $t \cong t'$ should have the same normal form $n$.
+In this article, we implement only the normalization function
 \fbox{$\tnorm : A \from \Gamma \to \Nf A \Gamma$}
-without any soundness guarantee.
+with proving its soundness and completeness.
+From a logical perspective, we will compute for each derivation of $A
+\from \Gamma$ a normal derivation $\Nf A \Gamma$.
+
+% \subsection{Proof normalization}
+
+% For didactical purposes, let us first focus on the simpler task of
+% constructing a normalization function
+% \fbox{$\tnorm : A \from \Gamma \to \Nf A \Gamma$}
+% without any soundness guarantee.
+
 % that does not guarantee any relation between its input term $t$ and
@@
 and reification $\down A : \den A \todot \tNf A$ along the way.
-\begin{gather*}
+\[
 \begin{array}{lcl}
@@
 \end{array}
-\\[2ex]
+\]
+\[
 \begin{array}{lcl}
@@
 \end{array}
-\end{gather*}
+\]
 % We will interpret simple types $A$ as corresponding presheaves
@@
 
-\subsection{Term normalization}
-
-Defining%
-\footnote{%
-A more accurate presentation of $\tNF$ would be
-$\Sigma\,{u : \tNf A\,\Gamma}.\ (\tm u \cong t)$
-in order to make absolutely clear that we
-want get hold of the witness $u$ of the existential statement;
-however, we shall only use constructive
-existentials
-in this article and thus stick to the notation $\exists$.
-We mean \emph{constructive existence} in
-  contrast to classical existence and
-  \emph{mere existence} from Homotopy Type Theory which
-  is the propositional truncation of constructive existence.
+% \subsection{Term normalization}
+
+% Defining%
+% \footnote{%
+% A more accurate presentation of $\tNF$ would be
+% $\Sigma\,{u : \tNf A\,\Gamma}.\ (\tm u \cong t)$
+% in order to make absolutely clear that we
+% want get hold of the witness $u$ of the existential statement;
+% however, we shall only use constructive
+% existentials
+% in this article and thus stick to the notation $\exists$.
+% We mean \emph{constructive existence} in
+%   contrast to classical existence and
+%   \emph{mere existence} from Homotopy Type Theory which
+%   is the propositional truncation of constructive existence.
+% }
+% \fbox{$\NF t = \exists n \of \tNf A\,\Gamma.\ \tm n \cong t$}
+% % for $t : A \from \Gamma$,
+% we aim to construct a normalization function
+% \[
+%   \tnorm : (t : A \from \Gamma) \to \NF t
+% \]
+% that gives us for each term $t$ a $\cong$-equal normal form $n$.
+
+\section{Normalization to Call-By-Push Value}
+\label{sec:cbpv}
+
+As remarked in the previous section, the placement of the monad $\C$
+in the type semantics is a bit wasteful:  Each positive type is
+prefixed by $\C$.  In our grammar of normal forms, this corresponds to
+the ability to perform case distinctions ($\tcase$, $\tabort$) at any
+positive type $P$.  In fact, our type interpretation $\den A$
+corresponds to the translation of call-by-name (CBN) lambda-calculus
+into Moggi's monadic meta-language \cite{moggi:infcomp91,levy:hosc06}.
+
+It would be sufficient %, however,
+to perform all necessary
+case distinctions
+when transitioning from a negative type to a positive type.
+Introduction of the function type adds hypotheses to the context,
+providing material for case distinctions, but introduction of positive
+types does not add anything in that respect.  Thus, we could
+\emph{focus} on positive introductions until we transition back to a
+negative type.  Such focusing is present in the call-by-value (CBV)
+lambda-calculus, where positive introductions only operate on values,
+and variables stand only for values.  This structure is even more
+clearly spelled out in Levy's call-by-push-value (CBPV)
+\cite{levy:hosc06}, as it comes with a deep classification of types
+into positive and negative ones.  In the following, we shall utilize
+pure (\ie, effect-free) CBPV to achieve chaining of positive
+introductions.
+
+\subsection{Types and polarization}
+
+CBPV calls positive types $P$ \emph{value types} $A$ and negative
+types $N$
+\emph{computation types} $\underline B$,
+however, we shall stick to our terminology
+which is known from research on focalization.
+However, we shall use $\tThunk$ for switch $\downarrow$ and $\tComp$ for
+switch $\uparrow$.
+\[
+\begin{array}{lllrl@{\qquad}l}
+  \PTy & \ni & P,Q & ::= & \patom
+    \mid \tyzero \mid P_1 + P_2
+    \mid \tyone \mid P_1 \times P_2
+    \mid \Thunk N
+  & \mbox{positive type}
+\\
+  \NTy & \ni & N,M & ::= & \natom
+    \mid \top \mid N_1 \ntimes N_2
+    \mid P \To N
+    \mid \Comp P
+  & \mbox{negative type}
+\\
+\end{array}
+\]
+CBPV uses $U$ for $\tThunk$ and $F$ for $\tComp$, however, we find
+these names uninspiring unless you have good knowledge of the intended
+model.  Further, CBPV employs labeled sums $\Sigma_I (P_i)_{i : I}$
+and labeled records $\Pi_I (P_i)_{i : I}$ for up to countably infinite
+label sets $I$ while we only have finite sums $(\tyzero,+)$ and
+records $(\top,\ntimes)$.  However, this difference in not essential,
+our treatment extends directly to the infinite case, since we are
+working in type theory which allows infinitely branching inductive
+types.   As a last difference, CBPV does not consider base types; in
+anticipation of the next section, we add them as both positive atoms
+($\patom$) and negative atoms ($\natom$).
+
+Getting a bit ahead of ourselves, let us consider the
+mutually defined interpretations
+$\den P$ and $\den N$ of positive and negative types as presheaves.
+\[
+\begin{array}{lcl}
+  \den{\tyone} & = & \hatone \\
+  \den{P_1 \times P_2} & = & \den{P_1} \hattimes \den{P_2} \\
+  \den{\tyzero} & = & \hatzero \\
+  \den{P_1 + P_2} & = & \den{P_1} \hatplus \den{P_2} \\
+  \den{\Thunk N} & = & \den N \\
+  \den{\patom} & = & \patom{\in}\_ \\
+\end{array}
+\qquad
+\begin{array}{lcl}
+  \den{\top} & = & \hatone \\
+  \den{N_1 \ntimes N_2} & = & \den{N_1} \hattimes \den{N_2} \\
+  \\
+  \den{P \To N} & = & \den P \hatto \den N \\
+  \den{\Comp P} & = & \C \den P \\
+  \den{\natom}  & = & \C (\tNe\, \natom) \\
+\end{array}
+\]
+Semantically, we do not distinguish between positive and negative
+products.  Notably, sum types can now be interpreted as plain
+(pointwise) presheaf
+sums.  The $\tThunk$ marker is ignored, yet $\tComp$, marking the
+switch from the negative to the positive type interpretation, places
+the cover monad.  Positive atoms, standing for value types without
+constructors, are only inhabited by variables $x : \patom \in \Gamma$.
+Negative atoms stand for computation types without own eliminations,
+thus, their inhabitants stem only from eliminations of more complex
+types, made from positive eliminations captured in $\C$ and negative
+eliminations chained together as neutral $\tNe\,\natom$, which we
+shall define below.
+The method $\trunNf[\C]$ of cover monad $\C$ can be extended to
+$\trun^N : \C \den N \to \den N$ for negative types $N$, by recursion
+on $N$.  Informally, speaking this makes all negative types \emph{monadic}.
+
+Contexts are lists of \emph{positive} types as in CBPV variables stand
+for values.  Interpretation of contexts $\den \Gamma$ is again defined
+pointwise $\den \cempty = \hatone$ and
+$\den{\cext \Gamma P} = \den \Gamma \hattimes \den P$.
+
+\subsection{Terms and evaluation}
+
+Assuming a family $\Tm N \Gamma$ of terms of negative type $N$ in
+context $\Gamma$, values \fbox{$v : \PValT P \Gamma$}
+of positive type $P$ shall be constructed by the
+following rules:
+\begin{gather*}
+  \nru{\tvar}{P \in \Gamma}{\PValT P \Gamma}
+  % \nru{\pvar}{P \in \Gamma}{\PValT P \Gamma}
+\qquad
+  \nru{\tthunk}{\Tm N \Gamma}{\PValTP{\Thunk N}\Gamma}
+\\[2ex]
+  \nru{\punit}{}{\PValT{\tyone}\Gamma}
+\qquad
+  \nru{\ppair}
+      {\PValT{P_1}\Gamma \qquad \PValT{P_2}\Gamma}
+      {\PValTP{P_1 \times P_2}\Gamma}
+\qquad
+  \nru{\tinj_i}{\PValT{P_i}\Gamma}{\PValTP{P_1 + P_2}\Gamma}
+\end{gather*}
+
+The terms of pure CBPV are given by the inductive family \fbox{$\Tm N \Gamma$}.
+It repeats the introductions and eliminations of negative types,
+except that application is restricted to values.  Values of type
+$\Thunk N$ are embedded via $\tforce$.
+Further, values of type $P$ can embedded
+via $\tret$, producing a term of type $\Comp P$.  Such terms are
+eliminated by $\tbind$ which is, unlike the usual monadic bind, not
+only available for $\tComp$-types but for arbitrary negative types
+$N$.  This is justified due to the monadic character of negative
+types, by virtue of $\trun^N$.  Finally, there are eliminators
+($\tsplit$, $\tcase$, $\tabort$) for values of positive product and
+sum types.
+\begin{gather*}
+  \nru{\tret}
+      {\PValTm P \Gamma}
+      {\TmP {\Comp P} \Gamma}
+\quad
+  \nru{\tabs}
+      {\Tm N {(\cext \Gamma P)}}
+      {\TmP {P \To N} \Gamma}
+\quad
+  \nru{\npair}
+      {\Tm {N_1} \Gamma \qquad \Tm {N_2} \Gamma}
+      {\TmP {N_1 \ntimes N_2} \Gamma}
+\quad
+  \nru{\nunit}
+      {}
+      {\Tm \top \Gamma}
+\\[2ex]
+  \nru{\tforce}
+      {\PValTmP {\Thunk N} \Gamma}
+      {\Tm N \Gamma}
+\qquad
+  \nru{\tapp}
+      {\TmP {P \To N} \Gamma \qquad \PValTm P \Gamma}
+      {\Tm N \Gamma}
+\qquad
+  \nru{\tprj_i}
+      {\TmP {N_1 \ntimes N_2} \Gamma}
+      {\Tm {N_i} \Gamma}
+\\[2ex]
+% \end{gather*}
+% \begin{gather*}
+  \nru{\tbind}
+      {\TmP {\Comp P} \Gamma \qquad \Tm N {(\cext \Gamma P)}}
+      {\Tm N \Gamma}
+\qquad
+  \nru{\tsplit}
+    {\PValTmP {P_1 \times P_2} \Gamma
+      \qquad \Tm N {(\cext {\cext \Gamma P_1} P_2)}
+    }{\Tm N \Gamma}
+\\[2ex]
+  \nru{\tcase}
+    {\PValTmP {P_1 + P_2} \Gamma
+      \qquad \Tm N {(\cext \Gamma P_1)}
+      \qquad \Tm N {(\cext \Gamma P_2)}
+    }{\Tm N \Gamma}
+\qquad
+  \nru{\tabort}
+    {\PValTm \tyzero \Gamma}
+    {\Tm N \Gamma}
+\end{gather*}
+
+Interpretation of values
+$\denty{ v : \PValTm P \Gamma } : \den \Gamma \todot \den P$
+and terms
+$\denty{ t : \Tm N \Gamma } : \den \Gamma \todot \den N$
+is straightforward, thanks to the pioneering work of
+Moggi \cite{moggi:infcomp91} and
+Levy \cite{levy:hosc06}
+put into the design of CBPV.
+\[
+\begin{array}{lcl}
+  \\
+  \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
+  \dent{\punit}\gamma   & = & \ttempty \\
+  \dent{\ppair\,v_1\,v_2}\gamma & = &
+    (\dent{v_1}\gamma,\ \dent{v_2}\gamma) \\
+  \dent{\inj i v}\gamma & = & \iota_i \dent v \gamma \\
+  \dent{\tthunk\,t}\gamma & = & \dent t \gamma \\
+\end{array}
+\quad
+\begin{array}{lcl}
+  \dent{\tabs\,t}\gamma
+      & = & \fdent t \gamma \\
+  \dent{\tapp\,t\,v} \gamma & = &
+     \dent t \gamma\;\tid\;\dent v \gamma \\
+  \dent{\nunit}\gamma & = & \ttempty \\
+  \dent{\npair\,t_1\,t_2}\gamma & = &
+    (\dent{t_1}\gamma,\ \dent{t_2}\gamma) \\
+  \dent{\prj i t}\gamma & = & \pi_i \dent t \gamma \\
+  \dent{\tforce\,v}\gamma & = & \dent v \gamma \\
+\end{array}
+\]
+Since $\tThunk$ serves only as an embedding of negative into positive
+types and has no semantic effect,
+we interpret thunking and forcing by the identity.
+The eliminations for positive types deal now only with values, thus,
+need not reference the monad operations.
+\[
+\begin{array}{lcl}
+  \dent{\tsplit\,v\,t}\gamma & = & (\lambda\,(a_1,a_2).\
+    \dent{t}{(\gamma,a_1,a_2)} ) \; \dent v \gamma \\
+  \dent{\tcase\,v\,t_1\,t_2}\gamma & = &
+    [  \lambda a_1.\ \dent{t_1}{(\gamma,a_1)}
+    ,\ \lambda a_2.\ \dent{t_2}{(\gamma,a_2)}
+    ]\; \dent v \gamma \\
+  \dent{\tabort\,v}\gamma & = & \tmagic\,\dent v \gamma \\
+\end{array}
+\]
+The use of the monad is confined to $\tret$ and $\tbind$.
+Note the availability of $\trun^\C : \C \den N \to \den N$
+at any negative type $N$ for the interpretation of $\tbind$.
+\[
+\begin{array}{lcl}
+  \dent{\tret\,v}\gamma & = & \treturn[\C] \; \dent v \gamma \\
+  \dent{\tbind\,u\,t}\gamma & = & \trun^\C\,
+    (\stmap[\C] \; \fdent t \gamma\; \dent u \gamma) \\
+\end{array}
+\]
+
+\subsection{Normal forms and normalization}
+
+Positive normal forms are values $v : \VNf P \Gamma$
+referring only to atomic variables and
+whose thunks only
+contain negative normal forms.
+% Positive normal forms are values $v : \PVal[\tNf]$ whose thunks only
+% contain negative normal forms.
+\begin{gather*}
+  \nru{\tvar}{\patom \in \Gamma}{\PValNf {\patom} \Gamma}
+  % \nru{\pvar}{P \in \Gamma}{\PValNf P \Gamma}
+\qquad
+  \nru{\tthunk}{\Nf N \Gamma}{\PValNfP{\Thunk N}\Gamma}
+\\[2ex]
+  \nru{\punit}{}{\PValNf{\tyone}\Gamma}
+\qquad
+  \nru{\ppair}
+      {\PValNf{P_1}\Gamma \qquad \PValNf{P_2}\Gamma}
+      {\PValNfP{P_1 \times P_2}\Gamma}
+\qquad
+  \nru{\tinj_i}{\PValNf{P_i}\Gamma}{\PValNfP{P_1 + P_2}\Gamma}
+\end{gather*}
+
+Neutral normal forms \fbox{$\Ne N \Gamma$} are negative eliminations
+starting from a forced $\tThunk$ rather than from variables of negative
+types (as those do not exist in CBPV).  However, due to normality
+the $\tThunk$ cannot be a $\tthunk$, but only a variable
+$\Thunk N \in \Gamma$.
+\begin{gather*}
+  \nru{\tforce}{\Thunk N \in \Gamma}{\Ne N \Gamma}
+  % \nru{\nvar}{\Thunk N \in \Gamma}{\Ne N \Gamma}
+\qquad
+  \nru{\tprj_i}
+      {\NeP {N_1 \ntimes N_2} \Gamma}
+      {\Ne {N_i} \Gamma}
+\qquad
+  \nru{\tapp}
+      {\NeP {P \To N} \Gamma \qquad \PValNf P \Gamma}
+      {\Ne N \Gamma}
+\end{gather*}
+Variables are originally introduced by either $\tabs$ or
+the $\tbind$ing of a neutral of type $\Comp P$ to a new variable of type
+$P$.  Variables of composite value type can be broken down by pattern
+matching, introducing variables of smaller type.  These positive
+eliminations plus $\tbind$ are organized in the
+inductively defined strong monad \fbox{$\Cov$}.
+\begin{gather*}
+  \nru{\treturn}{\J\,\Gamma}{\Cover \J \Gamma}
+\qquad
+  \nru{\tbind}
+      {\NeP {\Comp P} \Gamma \qquad \Cover \J (\cext \Gamma P)}
+      {\Cover \J \Gamma}
+\\[2ex]
+  \nru{\tsplit}
+      {P_1 \times P_2 \in \Gamma \qquad
+       \Cover \J (\cext {\cext \Gamma {P_1}} {P_2})
       }
-\fbox{$\NF t = \exists n \of \tNf A\,\Gamma.\ \tm n \cong t$}
-% for $t : A \from \Gamma$,
-we aim to construct a normalization function
-\[
-  \tnorm : (t : A \from \Gamma) \to \NF t
-\]
-that gives us for each term $t$ a $\cong$-equal normal form $n$.
+      {\Cover \J \Gamma}
+\\[2ex]
+  \nru{\tcase}
+      {P_1 + P_2 \in \Gamma
+       \qquad \Cover \J (\cext \Gamma {P_1})
+       \qquad \Cover \J (\cext \Gamma {P_2})
+      }
+      {\Cover \J \Gamma}
+\qquad
+  \nru{\tabort}
+      {\tyempty \in \Gamma}
+      {\Cover \J \Gamma}
+\end{gather*}
+Finally, normal forms of negative types are defined as inductive
+family \fbox{$\Nf N \Gamma$}.  They are generated by maximal negative
+introduction ($\tabs$, $\npair$, $\nunit$) until a negative atom or
+$\Comp P$ is reached.  Then, elimination of neutrals and variables is
+possible through the $\Cov$ monad until an answer can be given in form
+of a base neutral ($\tNe\,\natom$) or a normal value.
+\begin{gather*}
+  \nru{\tne}
+      {\CoverP {\tNe\,\natom} \Gamma}
+      {\Nf {\natom} \Gamma}
+\qquad
+  \nru{\tret}
+      {\CoverP {\tVNf\,P}  \Gamma}
+      {\NfP {\Comp P} \Gamma}
+\\[2ex]
+  \nru{\nunit}{}{\Nf \top \Gamma}
+\qquad
+  \nru{\npair}
+      {\Nf {N_1} \Gamma \qquad \Nf {N_2} \Gamma}
+      {\NfP {N_1 \ntimes N_2} \Gamma}
+\qquad
+  \nru{\tabs}
+      {\Nf N {\cext \Gamma P}}
+      {\NfP {P \To N} \Gamma}
+\end{gather*}
+We again can run the cover monad on normal forms, \ie, have
+$\trunNf : \Cov\,(\tNf N) \todot \tNf N$, which extends to
+negative semantic values
+$\trun : \Cov \den N \todot \den N$.
+
+Reification $\down P : \den P \todot \tVNf\,P$ at positive types produces
+a normal value, and $\down N : \den P \todot \tNf\,N$ at negative types a
+normal term.  During reification of function types $P \To N$ in
+context $\Gamma$ we need to embed a fresh variable
+$x : P \in (\cext \Gamma P)$ into $\den P$, breaking down $P$ to
+positive atoms $\patom$ and negative remainders $\Thunk N$.
+However, in $\den P$ we do not have case analysis available, thus,
+positive reflection $\upG P : P \in \Gamma \to \Cover {\den P} \Gamma$
+needs to run in the monad.  Luckily, as $\down N$ produces a normal
+form, monadic intermediate computations are permitted under a final
+$\trunNf$.  Negative reflection $\up N : \tNe\,N \todot \den N$ is as
+before generalized from variables to neutrals, to handle the breaking
+down of $N$ via eliminations.
+%
+In the following definition of reflection we use the abbreviation
+\fbox{\(
+  \tfresh^P_\Gamma
+    = \up[\cext \Gamma P] P \vv0
+    : \Cover{\den P}{(\cext \Gamma P)}
+\)}.
+\[
+\def\arraystretch{1.3}
+\begin{array}{l@{\,}l@{~~}c@{~~}l}
+  \multicolumn 2 l {\upG P} & : & P \in \Gamma \to \Cov\,\den P \, \Gamma\\
+  \upG {\patom}         & x & = & x \\
+  \upG {\tyone}         & x & = & \treturn\;\ttempty \\
+  \upG {P_1 \times P_2} & x & = & \tsplit\;x\,
+    \left(
+     (\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_1}\,\vv1)
+     \star
+     (\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_2}\,\vv0)
+    \right) \\
+    % (\tzip
+    % \,(\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_1}\,\vv1)
+    % \,(\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_2}\,\vv0)
+    % ) \\
+  \upG{\tyzero}         & x & = & \tabort\;x \\
+  \upG{P_1 + P_2}       & x & = & \tcase\;x
+    \; (\tmap\;\iota_1\;\tfresh^{P_1}_\Gamma)
+    \; (\tmap\;\iota_2\;\tfresh^{P_2}_\Gamma)
+    \\
+  \upG{\Thunk N} & x & = & \treturn\,(\upG N (\tforce\,x)) \\
+\end{array}
+\,
+\begin{array}{lcl}
+  \down P & : & \den P \todot \tVNf\,P \\
+  \down {\patom} & = & \tvar \\
+  \downG {\tyone} () & = & \punit \\
+  \downG {P_1 \times P_2}(a_1,a_2) & = & \ppair
+    \,(\downG {P_1} a_1)
+    \,(\downG {P_2} a_2)
+    \\
+  \down {\tyzero} & = & \tmagic \\
+  \down {P_1 + P_2} & = &
+    [  \tinj_1 \comp \down{P_1}
+    ,\ \tinj_2 \comp \down{P_2}
+    ] \\
+  \down {\Thunk N} & = & \tthunk \circ \down N \\
+\end{array}
+\]
+Reflection at positive pairs uses monoidal functoriality
+$\C\,\A_1 \todot \C\,\A_2 \todot \C\,(\A_1 \hattimes \A_2)$
+called $\star$ by McBride and Paterson \cite[Section 7]{mcbridePaterson:applicative}
+which for monads $\C$ can be defined by
+$c_1 \star c_2 =
+   \tjoin \; (\tmap\;(\lambda\,a_1.\
+   \tmap\;(\lambda\,a_2.\ (a_1,a_2))
+    \;c_2)
+    \;c_1)$.
+% $c_1 \star c_2 =
+%    \tjoin \; (\tmap\;(\lambda\,a_1.\
+%    \tjoin \; (\tmap\;(\lambda\,a_2.\ \treturn\,(a_1,a_2))
+%     \;c_2))
+%     \;c_1)$.
+% $c \star d = \tjoin \; (\tmap\;(\lambda\,b.\ (c,\, \treturn\,b))\;d$.
+% $\tzip\;c = \tjoin \comp \tmap\,(\lambda\,a.\ (c,\, \treturn\,a)$.
+
+For negative product types, reflection and reification works as before:
+\[
+\def\arraystretch{1.3}
+\begin{array}{l@{\,}l@{~~}c@{~~}l}
+  \multicolumn 2 l {\up N} & : & \tNe\,N \todot \den N \\
+  \upG {\Comp P} & u & = & \tbind\,u\,\tfresh^P_\Gamma \\
+  \upG \natom    & u & = & \treturn\,u \\
+  \upG \top      & u & = & \ttempty \\
+  \upG {N_1 \ntimes N_2} & u & = & \left(
+    \upG {N_1} \; (\prj 1 u)
+    ,\
+    \upG {N_2} \; (\prj 2 u)
+    \right) \\
+\end{array}
+\
+\begin{array}{l@{\,}l@{~~}c@{~~}l}
+  \multicolumn 2 l {\down N} & : & \den N \todot \tNf\;N \\
+  \downG {\Comp P} & c  & = & \tmap\;(\down P)\;c \\
+  \downG \natom    & c  & = & \tne\,c \\
+  \downG \top      & () & = & \nunit \\
+  \downG {N_1 \ntimes N_2} & (b_1,b_2) & = & \npair
+    \, (\downG {N_1} \; b_1)
+    \, (\downG {N_2} \; b_2)
+    \\
+\end{array}
+\]
+Reflection for function types is also unchanged, except that $\tapp$
+expects a value argument now.
+\[
+\def\arraystretch{1.3}
+\begin{array}{l@{\,}l@{~~}c@{~~}l}
+  \upG {P \To N} & u & = & \lambda
+    \;(\tau : \Gamma \sublist \Delta)
+    \;(a : \den P \Delta)
+    . \
+    \up[\Delta] N \left( \tapp\,(\tren\,\tau\,u)\,(\down[\Delta] P a) \right)
+    \\
+  \downG {P \To N} & f & = & \tabs \left( \trunNf \left(
+    \stmap
+      \left(
+        \lambda
+          \;(\tau : (\cext \Gamma P) \sublist \Delta)
+          \;(a : \den P \Delta)
+          . \
+          f\; (\twk^P \rcomp \tau)\; a
+        \right)
+      \,\tfresh^P_\Gamma
+    \right) \right)
+    \\
+\end{array}
+\]
+The identity environment $\idenv \Gamma : \Cover {\den \Gamma} \Gamma$
+can only be generated in the monad, due to monadic positive
+reflection.
+\[
+\begin{array}{lcl}
+  \idenv{\cempty} & = & \treturn\,() \\
+  \idenv{\cext \Gamma P} & = &
+    (\tren\,\twk^P\,\idenv\Gamma) \star \tfresh^P_\Gamma \\
+\end{array}
+\]
+Putting things together, we obtain the normalization function
+\[
+  \tnorm\,(t : \Tm N \Gamma) = \trunNf\,
+    \left(
+      \tmap\,\left(\down N \comp \denty t\right)\,\idenv\Gamma
+    \right)
+.
+\]
+
+Taking stock, we have arrived at normal forms that eagerly introduce
+($\tNf$) and eliminate ($\tNe$) negative types and also eagerly
+introduce positive types ($\tVnf$).  However, the elimination of
+positive types is still rather non-deterministic.
+It is possible to only partially break up a composite positive type
+and leave smaller, but still composite positive types for later
+pattern matching.
+The last refinement, chaining also the positive eliminations, will be
+discussed in the following section.
+% The placement of $\tsplit$, $\tcase$, and $\tabort$ is only
+% constrained by the presence of the split variable
+
+\section{Focused Intuitionistic Propositional Logic}
+\label{sec:fipl}
+
+% Polarized lambda calculus \cite{espiritoSanto:entcs17}
+% divides types deeply into positive and negative.
+
+% \[
+% \begin{array}{lllrl@{\qquad}l}
+%   \PTy & \ni & P,Q & ::= & \patom
+%     \mid \tyzero \mid P_1 + P_2
+%     \mid \tyone \mid P_1 \times P_2
+%     \mid \Thunk N
+%   & \mbox{positive type}
+% \\
+%   \NTy & \ni & N,M & ::= & \natom
+%     \mid \top \mid N_1 \ntimes N_2
+%     \mid P \To N
+%     \mid \Comp P
+%   & \mbox{negative type}
+% \\
+%   \HTy & \ni & H & ::= & \patom \mid N
+%   & \mbox{hypothesis type}
+% \\
+% \end{array}
+% \]
+% Espirito Santo's version actually does not contain the positive
+% product $\times$, however, it appears in Zeilberger
+% \cite{zeilberger:PhD}.
+
+Polarized lambda-calculus
+\cite{zeilberger:PhD,espiritoSanto:entcs17}
+is a focused calculus, it eagerly employs
+invertible rules: the introduction rules for negative types and the
+elimination rules for positive types.  As a consequence of the latter,
+variables are either of atomic or negative type $H ::= \patom \mid N$.
+%($\HTy$).
+Contexts $\Gamma,\Delta$ are lists of $H$s.
+%$\HTy$s.
+%\fbox{$X ::= \patom \mid P$}.
+
+To add a variable of positive type $P$ to the context, we need to
+break it apart until only atoms and negative bits remain.  This is
+performed by maximal pattern matching, called the left-invertible
+phase of focalization.
+We express maximal pattern matching on $P$ as a strong functor
+$\Add P$ in the category of presheaves,
+mapping a presheaf $\J$ (``judgement'') to $\Add P \J$
+and a presheaf morphism $f : (\J \hatto \K)\Gamma$ to
+$\stmap[\Add P]_\Gamma f : \AddHyp P \J \Gamma \to \AddHyp P \K \Gamma$.
+For arbitrary $\J$ and $\Gamma$, the set \fbox{$\AddHyp P \J \Gamma$} is
+inductively constructed by the following rules:
+\begin{gather*}
+  \nru{\phyp}
+      {\J\,(\cext \Gamma \patom)}
+      {\AddHyp {\patom} \J \Gamma}
+\qquad
+  \nru{\nbranch}
+      {}
+      {\AddHyp {\tyzero} \J \Gamma}
+\qquad
+  \nru{\bbranch}
+      {\AddHyp {P_1} \J \Gamma \qquad
+       \AddHyp {P_2} \J \Gamma }
+      {\AddHyp {P_1 + P_2} \J \Gamma}
+\\[2ex]
+  \nru{\nhyp}
+      {\J\,(\cext \Gamma N)}
+      {\AddHyp {\Thunk N} \J \Gamma}
+\qquad
+  \nru{\nsplit}
+      {\J\,\Gamma}
+      {\AddHyp {\tyone} \J \Gamma}
+\qquad
+  \nru{\bsplit}
+      {\AddHyp {P_1} {\left(\Add{P_2}\,\J\right)} \Gamma}
+      {\AddHyp {P_1 \times P_2} \J \Gamma}
+\end{gather*}
+Note the recursive occurrence of $\Add{P_2}$ as argument to
+$\Add{P_1}$ in $\bsplit$, which makes $\Add P$ a nested datatype
+\cite{bird:nested}.  Agda supports such nested inductive types; but
+note that $\Add{P}$ is uncontroversial, since it could also be defined
+by recursion on $P$.   It is tempting to name $\bsplit$
+``$\tjoin$'' and $\nsplit$ ``$\treturn$''
+since $\Add P$ is a graded monad on the monoid $(1,\times)$
+of product types;
+however, this coincidence shall not matter for our further considerations.
+
+Focalization is a technique to remove don't-care non-determinism from
+proof search, and as such, polarized lambda calculus is foremost a
+calculus of normal forms.
+These normal forms are given by four mutually defined inductive
+families of presheaves $\tVnf\,P$,
+$\tNe\,N$, $\Cov\,\J$,
+and $\tNf\,N$.  As they are very similar to the CBPV normal forms
+given in the last section we only report the differences.
+
+Values \fbox{$v : \PVal P \Gamma$} are unchanged, they can refer to
+atomic positive hypotheses ($\pvar$) and normal $\tthunk$s.
+
+Neutrals \fbox{$\Ne N \Gamma$} start with a negative variable instead
+of with $\tforce$, as forcing thunks is already performed in $\nhyp$
+when adding analysing hypotheses of $\tThunk$ type.
+\[
+  \nru{\nvar}{N \in \Gamma}{\Ne N \Gamma}
+\]
+
+The Cover monad \fbox{$\Cover \J \Gamma$}
+lacks constructors $\tsplit$, $\tcase$ and $\tabort$
+since the pattern matching is taken care of by $\Add P$.
+\begin{gather*}
+  \nru{\treturn}{\J\,\Gamma}{\Cover \J \Gamma}
+\qquad
+  \nru{\tbind}
+      {\NeP {\Comp P} \Gamma \qquad \AddHypP P {\Cov\,\J} \Gamma}
+      {\Cover \J \Gamma}
+\end{gather*}
+
+The normal forms \fbox{$\Nf N \Gamma$} of negative type are unchanged
+with the exception that pattern matching happens eagerly in $\tabs$.
+\[
+  \nru{\tabs}
+      {\AddHypP P {\tNf\, N} \Gamma}
+      {\NfP {P \To N} \Gamma}
+\]
+
+All these inductive families are presheaves, due to factored
+presentation using $\Add P$ and $\Cov$ the proof is not a simple
+mutual induction.  Yet, in Agda, the generic proof goes through using a sized
+typing for these inductive families.  Similarily, defining the
+$\tjoin$ for monad $\Cov$ relies on sized typing.
+\[
+\begin{array}{lll}
+  \tjoin & : & \forall i.\
+    \Cov[i]\,(\Cov[\infty]\J) \todot \Cov[\infty]\J  \\
+  \tjoin[i+1]\,(\treturn[i]\;c) & = & c \\
+  \tjoin[i+1]\,(\tbind^i\;t\;k)
+    & = & \tbind^\infty\;
+     (t : \Ne P \Gamma)\;
+     \left(\stmap[\Add P]_\Gamma\;\tjoin[i]\;
+       \left(k : \AddHypP P {\Cov[i]\J} \Gamma
+       \right)
+     \right)
+\end{array}
+\]
+Herein, we used the sized typing of the constructors of $\Cov$:
+\[
+\begin{array}{lll}
+  \treturn & : & \forall i.\ \J \todot \Cov[i+1] \J \\
+  \tbind  & : & \forall i.\
+    \tNe\,(\Comp P) \todot \Add P\,(\Cov[i]\J) \todot \Cov[i+1]\J \\
+\end{array}
+\]
+
+Due to the eager splitting of positive hypotheses, reflection at type $P$
+now lives in the graded monad $\Add P$ rather than $\Cov$.  Further,
+as pattern matching may produce $n \geq 0$ cases, reflection cannot simply
+produce one positive semantic value; instead, one such value is needed
+for every branch.  We implement
+\fbox{$\reflect P : (\den P \hatto \J) \todot \Add P \J$}
+as a higher-order function expecting a continuation $k$ which is
+invoked for each generated branch with the semantic value of type $P$
+constructed for this branch.
+\[
+\begin{array}{l@{~}lcl}
+  \reflect {\patom} & k & = &
+    \phyp \left( k\;\twk^{\patom}\,(\pvar\,\tzero) \right) \\
+  \reflect {\Thunk N} & k & = &
+    \nhyp \left( k\;\twk^N \left( \up N (\nvar\,\tzero) \right) \right)\\
+  \reflect {\tyzero} & k & = & \nbranch \\
+  \reflect {P_1 + P_2} & k & = & \bbranch
+    \,\left( \treflect^{P_1}\,(\lambda\,\tau. \ k\,\tau \comp \iota_1) \right)
+    \,\left( \treflect^{P_2}\,(\lambda\,\tau. \ k\,\tau \comp \iota_2) \right)
+    \\
+  \reflect{\tyone} & k & = & \nsplit\;(k\;\tid\;\ttempty) \\
+  \reflect{P_1 \times P_2} & k & = & \bsplit
+    \,\left(
+      \treflect^{P_1} \left( \lambda\,\tau_1\,a_1. \
+      \treflect^{P_2} \left( \lambda\,\tau_2\,a_2. \
+        k\, (\tau_1 \rcomp \tau_2)\, (\tren\,\tau_2\,a_1, \ a_2)
+        \right)
+      \right)
+    \right)
+\end{array}
+\]
+Reflecting at a positive atomic type $\patom$ is the regular ending of a
+reflection pass: we call continuation $k$ with a fresh variable
+$\pvar\,\tzero$ of type $\patom$, making space for the variable
+using $\twk^{\patom}$.
+In case we end at type $\Thunk N$, we add a new variable
+$\nvar\,\tzero$ of type $N$ and pass it to $k$, after full
+$\eta$-expansion via $\up N$.  Two more endings are possible:
+At type
+$\tyzero$, we have reached an absurd case, meaning that no
+continuation is necessary since we can conclude with
+\emph{ex falsum quod libet}.
+At type $\tyone$, there is no need to add a new variable, as values
+of type $\tyone$ contain no information.  We simply pass the unit
+value $\ttempty$ to $k$ in this case.
+Reflecting at $P_1 + P_2$ generates two branches, which may result in
+several uses of the continuation $k$.
+In the first branch, we recursively reflect at $P_1$.  Its
+continuation will receive a semantic value in $\den{P_1}$ which we
+inject via $\iota_1$ into $\den{P_1 + P_2}$ to pass it to $k$.  The
+second branch proceeds analogously.
+Finally reflecting at $P_1 \times P_2$ means we first have to analyze
+$P_1$, and in each of the generated branches we continue to analyze
+$P_2$.  Thus $\reflect{P_2}$ is passed as a continuation to
+$\reflect{P_1}$.  Each reflection phase gives us a semantic value
+$a_i$ of type $P_i$, which we combine to a tuple before passing it to
+$k$.  Note also that the context extension $\tau_1$ created in the
+first phase needs to be composed with the context extension $\tau_2$
+of the second phase to transport $k$ into the final context.  Further,
+the value $a_1$ was constructed relative to the target of
+$\tau_1$ and still needs
+to be transported with $\tau_2$ before paired up with $a_2$.
+
+The method $\reflect P$ replaces previous uses of $\tfresh^P$ in
+reflection and reification at negative types.
+% and identity environment creation.
+\[
+\begin{array}{l@{~}lcl}
+  \downG {P \To N} & f & = & \tabs
+    \left(
+      \reflect[\Gamma] P
+        \left(
+          \lambda \, (\tau : \Gamma \sublist \Delta) \; a .\
+          \down[\Delta] P (f \; \tau \; a)
+        \right)
+    \right)
+  \\
+  \upG {\Comp P} & u & = & \tbind \; u
+    \left(
+      \reflect[\Gamma] P
+        \left(
+          \lambda \; \tau \; a .\
+          \treturn\; a
+        \right)
+    \right)
+  % \\
+  % \idenv{\cext \Gamma P} & = &
+\end{array}
+\]
+% To create an identity environment using $\treflect$, let us first
+% extend $\Add\_$ to contexts, following the intuition that
+Due to the strong limitation on context entries,
+the identity environment $\idenv \Gamma$ can be built
+straightforwardly using negative reflection.
+\[
+\begin{array}{lcl}
+  \idenv{\cempty} & = & \ttempty \\
+  \idenv{\cext \Gamma {\patom}} & = &
+    \left(
+      \tren\;\twk^{\patom}\;\idenv\Gamma
+      , \
+      \pvar\,\tzero
+    \right) \\
+  \idenv{\cext \Gamma N} & = &
+    \left(
+      \tren\;\twk^N\;\idenv\Gamma
+      , \
+      \upG N (\nvar\,\tzero)
+    \right) \\
+\end{array}
+\]
+
+The terms \fbox{$\Tm N \Gamma$} of the focused lambda calculus are the
+ones of CBPV minus the positive eliminations ($\tsplit$, $\tcase$,
+$\tabort$), the added negative variable rule ($\nvar$) and the
+necessary changes to the binders $\tabs$ and $\tbind$.
+\begin{gather*}
+  \nru{\nvar}{N \in \Gamma}{\Tm N \Gamma}
+\qquad
+  \nru{\tabs}
+      {\AddHypP P {\tTm\, N} \Gamma}
+      {\TmP {P \To N} \Gamma}
+\qquad
+  \nru{\tbind}
+      {\Tm {\Comp P} \Gamma \qquad \AddHypP P {\tTm\,N} \Gamma}
+      {\Tm N \Gamma}
+\end{gather*}
+Term interpretation
+\fbox{$\denty\_ : \Tm N \Gamma \to \den \Gamma \todot \den N$}
+shall be as for CBPV except that we need to exchange the
+interpretation function for binders \fbox{$\fden\_ : \Tm N {(\cext
+    \Gamma P)} \to \den \Gamma \todot \den{P \To N}$}.  Since a binder
+for $P$ performs a maximal splitting on $P$ and takes the form of a
+function defined by a case (and split) tree,
+applying it to a value $v$ of type $P$ amounts to a complete
+matching of $v$ against the case tree, and bind the remaining
+atomic and negative crumbs.  This matching can be defined for a
+generic evaluation function of type
+$\Evvv \J \Delta \Gamma = \den \Gamma \Delta \to \J\,\Delta$.
+\[
+\begin{array}{l@{~}l@{~}l@{~}lcl}
+  \multicolumn 6 {l} {\tmatch ~~:~~ \den P \Delta \to
+    \AddHyp P {\Evv \J \Delta} \todot \Evv \J \Delta }\\
+  \tmatch & x & (\phyp\,e) & \gamma & = & e\,(\gamma,\; \treturn\,x) \\
+  \tmatch & b & (\nhyp\,e) & \gamma & = & e\,(\gamma,\; b) \\
+  \tmatch & \ttempty & (\nsplit\,e) & \gamma & = & e\,\gamma \\
+  \tmatch & (a_1,a_2) & (\bsplit\,e) & \gamma & = &
+    \tmatch\;a_1\;(\tmap[\Add P]\,(\tmatch\;a_2)\;e)\;\gamma \\
+  \tmatch & a & \nbranch & \gamma & = & \tmagic\,a \\
+  \tmatch & (\iota_1\;a_1) & (\bbranch\;e_1\;e_2) & \gamma & = &
+    \tmatch\;a_1\;e_1\;\gamma \\
+  \tmatch & (\iota_2\;a_2) & (\bbranch\;e_1\;e_2) & \gamma & = &
+    \tmatch\;a_2\;e_2\;\gamma \\
+\end{array}
+\]
+With instantiations $\J = \den N$ and
+$\denty\_ : \Tm N \todot \Evv {\den N} \Delta$,
+the interpretation $\fden t$
+of binder $t : \AddHypP P {\tTm\,N} \Gamma$
+is defined as follows:
+\[
+  \fdent{t}
+    {(\gamma : \den \Gamma \Delta)}
+    \, (\tau : \Delta \sublist \Phi)
+    \, (a : \den P \Phi)
+    ~~=~~
+    \tmatch
+      \; a
+      \; (\tmap[\Add P]\;\denty\_\;t)
+      \; (\tren\,\tau\,\gamma)
+\]
+This completes the definition of the normalization function
+$\tnorm\, (t : \Tm N \Gamma)
+  = \down N \dent t {\idenv\Gamma}$.
+
+\section{Conclusion}
+\label{sec:concl}
+
+\bibliography{auto-fscd19}
+
+\clearpage
+\appendix
+
+\section{Equational Theory for STLC with Weak Sums}
+\label{sec:eq}
 
@@
 
-\clearpage
-
-\section{Normalization to Call-By-Push Value}
-\label{sec:cbpv}
-
-As remarked in the previous section, the placement of the monad $\C$
-in the type semantics is a bit wasteful:  Each positive type is
-prefixed by $\C$.  In our grammar of normal forms, this corresponds to
-the ability to perform case distinctions ($\tcase$, $\tabort$) at any
-positive type $P$.  In fact, our type interpretation $\den A$
-corresponds to the translation of call-by-name (CBN) lambda-calculus
-into Moggi's monadic meta-language \cite{moggi:infcomp91,levy:hosc06}.
-
-It would be sufficient %, however,
-to perform all necessary
-case distinctions
-when transitioning from a negative type to a positive type.
-Introduction of the function type adds hypotheses to the context,
-providing material for case distinctions, but introduction of positive
-types does not add anything in that respect.  Thus, we could
-\emph{focus} on positive introductions until we transition back to a
-negative type.  Such focusing is present in the call-by-value (CBV)
-lambda-calculus, where positive introductions only operate on values,
-and variables stand only for values.  This structure is even more
-clearly spelled out in Levy's call-by-push-value (CBPV)
-\cite{levy:hosc06}, as it comes with a deep classification of types
-into positive and negative ones.  In the following, we shall utilize
-pure (\ie, effect-free) CBPV to achieve chaining of positive
-introductions.
-
-\subsection{Types and polarization}
-
-CBPV calls positive types $P$ \emph{value types} $A$ and negative
-types $N$
-\emph{computation types} $\underline B$,
-however, we shall stick to our terminology
-which is known from research on focalization.
-However, we shall use $\tThunk$ for switch $\downarrow$ and $\tComp$ for
-switch $\uparrow$.
-\[
-\begin{array}{lllrl@{\qquad}l}
-  \PTy & \ni & P,Q & ::= & \patom
-    \mid \tyzero \mid P_1 + P_2
-    \mid \tyone \mid P_1 \times P_2
-    \mid \Thunk N
-  & \mbox{positive type}
-\\
-  \NTy & \ni & N,M & ::= & \natom
-    \mid \top \mid N_1 \ntimes N_2
-    \mid P \To N
-    \mid \Comp P
-  & \mbox{negative type}
-\\
-\end{array}
-\]
-CBPV uses $U$ for $\tThunk$ and $F$ for $\tComp$, however, we find
-these names uninspiring unless you have good knowledge of the intended
-model.  Further, CBPV employs labeled sums $\Sigma_I (P_i)_{i : I}$
-and labeled records $\Pi_I (P_i)_{i : I}$ for up to countably infinite
-label sets $I$ while we only have finite sums $(\tyzero,+)$ and
-records $(\top,\ntimes)$.  However, this difference in not essential,
-our treatment extends directly to the infinite case, since we are
-working in type theory which allows infinitely branching inductive
-types.   As a last difference, CBPV does not consider base types; in
-anticipation of the next section, we add them as both positive atoms
-($\patom$) and negative atoms ($\natom$).
-
-Getting a bit ahead of ourselves, let us consider the
-mutually defined interpretations
-$\den P$ and $\den N$ of positive and negative types as presheaves.
-\[
-\begin{array}{lcl}
-  \den{\tyone} & = & \hatone \\
-  \den{P_1 \times P_2} & = & \den{P_1} \hattimes \den{P_2} \\
-  \den{\tyzero} & = & \hatzero \\
-  \den{P_1 + P_2} & = & \den{P_1} \hatplus \den{P_2} \\
-  \den{\Thunk N} & = & \den N \\
-  \den{\patom} & = & \patom{\in}\_ \\
-\end{array}
-\qquad
-\begin{array}{lcl}
-  \den{\top} & = & \hatone \\
-  \den{N_1 \ntimes N_2} & = & \den{N_1} \hattimes \den{N_2} \\
-  \\
-  \den{P \To N} & = & \den P \hatto \den N \\
-  \den{\Comp P} & = & \C \den P \\
-  \den{\natom}  & = & \C (\tNe\, \natom) \\
-\end{array}
-\]
-Semantically, we do not distinguish between positive and negative
-products.  Notably, sum types can now be interpreted as plain
-(pointwise) presheaf
-sums.  The $\tThunk$ marker is ignored, yet $\tComp$, marking the
-switch from the negative to the positive type interpretation, places
-the cover monad.  Positive atoms, standing for value types without
-constructors, are only inhabited by variables $x : \patom \in \Gamma$.
-Negative atoms stand for computation types without own eliminations,
-thus, their inhabitants stem only from eliminations of more complex
-types, made from positive eliminations captured in $\C$ and negative
-eliminations chained together as neutral $\tNe\,\natom$, which we
-shall define below.
-The method $\trunNf[\C]$ of cover monad $\C$ can be extended to
-$\trun^N : \C \den N \to \den N$ for negative types $N$, by recursion
-on $N$.  Informally, speaking this makes all negative types \emph{monadic}.
-
-Contexts are lists of \emph{positive} types as in CBPV variables stand
-for values.  Interpretation of contexts $\den \Gamma$ is again defined
-pointwise $\den \cempty = \hatone$ and
-$\den{\cext \Gamma P} = \den \Gamma \hattimes \den P$.
-
-\subsection{Terms and evaluation}
-
-Assuming a family $\Tm N \Gamma$ of terms of negative type $N$ in
-context $\Gamma$, values \fbox{$v : \PValT P \Gamma$}
-of positive type $P$ shall be constructed by the
-following rules:
-\begin{gather*}
-  \nru{\tvar}{P \in \Gamma}{\PValT P \Gamma}
-  % \nru{\pvar}{P \in \Gamma}{\PValT P \Gamma}
-\qquad
-  \nru{\tthunk}{\Tm N \Gamma}{\PValTP{\Thunk N}\Gamma}
-\\[2ex]
-  \nru{\punit}{}{\PValT{\tyone}\Gamma}
-\qquad
-  \nru{\ppair}
-      {\PValT{P_1}\Gamma \qquad \PValT{P_2}\Gamma}
-      {\PValTP{P_1 \times P_2}\Gamma}
-\qquad
-  \nru{\tinj_i}{\PValT{P_i}\Gamma}{\PValTP{P_1 + P_2}\Gamma}
-\end{gather*}
-
-The terms of pure CBPV are given by the inductive family \fbox{$\Tm N \Gamma$}.
-It repeats the introductions and eliminations of negative types,
-except that application is restricted to values.  Values of type
-$\Thunk N$ are embedded via $\tforce$.
-Further, values of type $P$ can embedded
-via $\tret$, producing a term of type $\Comp P$.  Such terms are
-eliminated by $\tbind$ which is, unlike the usual monadic bind, not
-only available for $\tComp$-types but for arbitrary negative types
-$N$.  This is justified due to the monadic character of negative
-types, by virtue of $\trun^N$.  Finally, there are eliminators
-($\tsplit$, $\tcase$, $\tabort$) for values of positive product and
-sum types.
-\begin{gather*}
-  \nru{\tret}
-      {\PValTm P \Gamma}
-      {\TmP {\Comp P} \Gamma}
-\quad
-  \nru{\tabs}
-      {\Tm N {(\cext \Gamma P)}}
-      {\TmP {P \To N} \Gamma}
-\quad
-  \nru{\npair}
-      {\Tm {N_1} \Gamma \qquad \Tm {N_2} \Gamma}
-      {\TmP {N_1 \ntimes N_2} \Gamma}
-\quad
-  \nru{\nunit}
-      {}
-      {\Tm \top \Gamma}
-\\[2ex]
-  \nru{\tforce}
-      {\PValTmP {\Thunk N} \Gamma}
-      {\Tm N \Gamma}
-\qquad
-  \nru{\tapp}
-      {\TmP {P \To N} \Gamma \qquad \PValTm P \Gamma}
-      {\Tm N \Gamma}
-\qquad
-  \nru{\tprj_i}
-      {\TmP {N_1 \ntimes N_2} \Gamma}
-      {\Tm {N_i} \Gamma}
-\\[2ex]
-% \end{gather*}
-% \begin{gather*}
-  \nru{\tbind}
-      {\TmP {\Comp P} \Gamma \qquad \Tm N {(\cext \Gamma P)}}
-      {\Tm N \Gamma}
-\qquad
-  \nru{\tsplit}
-    {\PValTmP {P_1 \times P_2} \Gamma
-      \qquad \Tm N {(\cext {\cext \Gamma P_1} P_2)}
-    }{\Tm N \Gamma}
-\\[2ex]
-  \nru{\tcase}
-    {\PValTmP {P_1 + P_2} \Gamma
-      \qquad \Tm N {(\cext \Gamma P_1)}
-      \qquad \Tm N {(\cext \Gamma P_2)}
-    }{\Tm N \Gamma}
-\qquad
-  \nru{\tabort}
-    {\PValTm \tyzero \Gamma}
-    {\Tm N \Gamma}
-\end{gather*}
-
-Interpretation of values
-$\denty{ v : \PValTm P \Gamma } : \den \Gamma \todot \den P$
-and terms
-$\denty{ t : \Tm N \Gamma } : \den \Gamma \todot \den N$
-is straightforward, thanks to the pioneering work of
-Moggi \cite{moggi:infcomp91} and
-Levy \cite{levy:hosc06}
-put into the design of CBPV.
-\[
-\begin{array}{lcl}
-  \\
-  \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
-  \dent{\punit}\gamma   & = & \ttempty \\
-  \dent{\ppair\,v_1\,v_2}\gamma & = &
-    (\dent{v_1}\gamma,\ \dent{v_2}\gamma) \\
-  \dent{\inj i v}\gamma & = & \iota_i \dent v \gamma \\
-  \dent{\tthunk\,t}\gamma & = & \dent t \gamma \\
-\end{array}
-\quad
-\begin{array}{lcl}
-  \dent{\tabs\,t}\gamma
-      & = & \fdent t \gamma \\
-  \dent{\tapp\,t\,v} \gamma & = &
-     \dent t \gamma\;\tid\;\dent v \gamma \\
-  \dent{\nunit}\gamma & = & \ttempty \\
-  \dent{\npair\,t_1\,t_2}\gamma & = &
-    (\dent{t_1}\gamma,\ \dent{t_2}\gamma) \\
-  \dent{\prj i t}\gamma & = & \pi_i \dent t \gamma \\
-  \dent{\tforce\,v}\gamma & = & \dent v \gamma \\
-\end{array}
-\]
-Since $\tThunk$ serves only as an embedding of negative into positive
-types and has no semantic effect,
-we interpret thunking and forcing by the identity.
-The eliminations for positive types deal now only with values, thus,
-need not reference the monad operations.
-\[
-\begin{array}{lcl}
-  \dent{\tsplit\,v\,t}\gamma & = & (\lambda\,(a_1,a_2).\
-    \dent{t}{(\gamma,a_1,a_2)} ) \; \dent v \gamma \\
-  \dent{\tcase\,v\,t_1\,t_2}\gamma & = &
-    [  \lambda a_1.\ \dent{t_1}{(\gamma,a_1)}
-    ,\ \lambda a_2.\ \dent{t_2}{(\gamma,a_2)}
-    ]\; \dent v \gamma \\
-  \dent{\tabort\,v}\gamma & = & \tmagic\,\dent v \gamma \\
-\end{array}
-\]
-The use of the monad is confined to $\tret$ and $\tbind$.
-Note the availability of $\trun^\C : \C \den N \to \den N$
-at any negative type $N$ for the interpretation of $\tbind$.
-\[
-\begin{array}{lcl}
-  \dent{\tret\,v}\gamma & = & \treturn[\C] \; \dent v \gamma \\
-  \dent{\tbind\,u\,t}\gamma & = & \trun^\C\,
-    (\stmap[\C] \; \fdent t \gamma\; \dent u \gamma) \\
-\end{array}
-\]
-
-\subsection{Normal forms and normalization}
-
-Positive normal forms are values $v : \VNf P \Gamma$
-referring only to atomic variables and
-whose thunks only
-contain negative normal forms.
-% Positive normal forms are values $v : \PVal[\tNf]$ whose thunks only
-% contain negative normal forms.
-\begin{gather*}
-  \nru{\tvar}{\patom \in \Gamma}{\PValNf {\patom} \Gamma}
-  % \nru{\pvar}{P \in \Gamma}{\PValNf P \Gamma}
-\qquad
-  \nru{\tthunk}{\Nf N \Gamma}{\PValNfP{\Thunk N}\Gamma}
-\\[2ex]
-  \nru{\punit}{}{\PValNf{\tyone}\Gamma}
-\qquad
-  \nru{\ppair}
-      {\PValNf{P_1}\Gamma \qquad \PValNf{P_2}\Gamma}
-      {\PValNfP{P_1 \times P_2}\Gamma}
-\qquad
-  \nru{\tinj_i}{\PValNf{P_i}\Gamma}{\PValNfP{P_1 + P_2}\Gamma}
-\end{gather*}
-
-Neutral normal forms \fbox{$\Ne N \Gamma$} are negative eliminations
-starting from a forced $\tThunk$ rather than from variables of negative
-types (as those do not exist in CBPV).  However, due to normality
-the $\tThunk$ cannot be a $\tthunk$, but only a variable
-$\Thunk N \in \Gamma$.
-\begin{gather*}
-  \nru{\tforce}{\Thunk N \in \Gamma}{\Ne N \Gamma}
-  % \nru{\nvar}{\Thunk N \in \Gamma}{\Ne N \Gamma}
-\qquad
-  \nru{\tprj_i}
-      {\NeP {N_1 \ntimes N_2} \Gamma}
-      {\Ne {N_i} \Gamma}
-\qquad
-  \nru{\tapp}
-      {\NeP {P \To N} \Gamma \qquad \PValNf P \Gamma}
-      {\Ne N \Gamma}
-\end{gather*}
-Variables are originally introduced by either $\tabs$ or
-the $\tbind$ing of a neutral of type $\Comp P$ to a new variable of type
-$P$.  Variables of composite value type can be broken down by pattern
-matching, introducing variables of smaller type.  These positive
-eliminations plus $\tbind$ are organized in the
-inductively defined strong monad \fbox{$\Cov$}.
-\begin{gather*}
-  \nru{\treturn}{\J\,\Gamma}{\Cover \J \Gamma}
-\qquad
-  \nru{\tbind}
-      {\NeP {\Comp P} \Gamma \qquad \Cover \J (\cext \Gamma P)}
-      {\Cover \J \Gamma}
-\\[2ex]
-  \nru{\tsplit}
-      {P_1 \times P_2 \in \Gamma \qquad
-       \Cover \J (\cext {\cext \Gamma {P_1}} {P_2})
-      }
-      {\Cover \J \Gamma}
-\\[2ex]
-  \nru{\tcase}
-      {P_1 + P_2 \in \Gamma
-       \qquad \Cover \J (\cext \Gamma {P_1})
-       \qquad \Cover \J (\cext \Gamma {P_2})
-      }
-      {\Cover \J \Gamma}
-\qquad
-  \nru{\tabort}
-      {\tyempty \in \Gamma}
-      {\Cover \J \Gamma}
-\end{gather*}
-Finally, normal forms of negative types are defined as inductive
-family \fbox{$\Nf N \Gamma$}.  They are generated by maximal negative
-introduction ($\tabs$, $\npair$, $\nunit$) until a negative atom or
-$\Comp P$ is reached.  Then, elimination of neutrals and variables is
-possible through the $\Cov$ monad until an answer can be given in form
-of a base neutral ($\tNe\,\natom$) or a normal value.
-\begin{gather*}
-  \nru{\tne}
-      {\CoverP {\tNe\,\natom} \Gamma}
-      {\Nf {\natom} \Gamma}
-\qquad
-  \nru{\tret}
-      {\CoverP {\tVNf\,P}  \Gamma}
-      {\NfP {\Comp P} \Gamma}
-\\[2ex]
-  \nru{\nunit}{}{\Nf \top \Gamma}
-\qquad
-  \nru{\npair}
-      {\Nf {N_1} \Gamma \qquad \Nf {N_2} \Gamma}
-      {\NfP {N_1 \ntimes N_2} \Gamma}
-\qquad
-  \nru{\tabs}
-      {\Nf N {\cext \Gamma P}}
-      {\NfP {P \To N} \Gamma}
-\end{gather*}
-We again can run the cover monad on normal forms, \ie, have
-$\trunNf : \Cov\,(\tNf N) \todot \tNf N$, which extends to
-negative semantic values
-$\trun : \Cov \den N \todot \den N$.
-
-Reification $\down P : \den P \todot \tVNf\,P$ at positive types produces
-a normal value, and $\down N : \den P \todot \tNf\,N$ at negative types a
-normal term.  During reification of function types $P \To N$ in
-context $\Gamma$ we need to embed a fresh variable
-$x : P \in (\cext \Gamma P)$ into $\den P$, breaking down $P$ to
-positive atoms $\patom$ and negative remainders $\Thunk N$.
-However, in $\den P$ we do not have case analysis available, thus,
-positive reflection $\upG P : P \in \Gamma \to \Cover {\den P} \Gamma$
-needs to run in the monad.  Luckily, as $\down N$ produces a normal
-form, monadic intermediate computations are permitted under a final
-$\trunNf$.  Negative reflection $\up N : \tNe\,N \todot \den N$ is as
-before generalized from variables to neutrals, to handle the breaking
-down of $N$ via eliminations.
-%
-In the following definition of reflection we use the abbreviation
-\fbox{\(
-  \tfresh^P_\Gamma
-    = \up[\cext \Gamma P] P \vv0
-    : \Cover{\den P}{(\cext \Gamma P)}
-\)}.
-\[
-\def\arraystretch{1.3}
-\begin{array}{l@{\,}l@{~~}c@{~~}l}
-  \multicolumn 2 l {\upG P} & : & P \in \Gamma \to \Cov\,\den P \, \Gamma\\
-  \upG {\patom}         & x & = & x \\
-  \upG {\tyone}         & x & = & \treturn\;\ttempty \\
-  \upG {P_1 \times P_2} & x & = & \tsplit\;x\,
-    \left(
-     (\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_1}\,\vv1)
-     \star
-     (\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_2}\,\vv0)
-    \right) \\
-    % (\tzip
-    % \,(\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_1}\,\vv1)
-    % \,(\up[\cext{\cext \Gamma {P_1}}{P_2}]{P_2}\,\vv0)
-    % ) \\
-  \upG{\tyzero}         & x & = & \tabort\;x \\
-  \upG{P_1 + P_2}       & x & = & \tcase\;x
-    \; (\tmap\;\iota_1\;\tfresh^{P_1}_\Gamma)
-    \; (\tmap\;\iota_2\;\tfresh^{P_2}_\Gamma)
-    \\
-  \upG{\Thunk N} & x & = & \treturn\,(\upG N (\tforce\,x)) \\
-\end{array}
-\,
-\begin{array}{lcl}
-  \down P & : & \den P \todot \tVNf\,P \\
-  \down {\patom} & = & \tvar \\
-  \downG {\tyone} () & = & \punit \\
-  \downG {P_1 \times P_2}(a_1,a_2) & = & \ppair
-    \,(\downG {P_1} a_1)
-    \,(\downG {P_2} a_2)
-    \\
-  \down {\tyzero} & = & \tmagic \\
-  \down {P_1 + P_2} & = &
-    [  \tinj_1 \comp \down{P_1}
-    ,\ \tinj_2 \comp \down{P_2}
-    ] \\
-  \down {\Thunk N} & = & \tthunk \circ \down N \\
-\end{array}
-\]
-Reflection at positive pairs uses monoidal functoriality
-$\C\,\A_1 \todot \C\,\A_2 \todot \C\,(\A_1 \hattimes \A_2)$
-called $\star$ by McBride and Paterson \cite[Section 7]{mcbridePaterson:applicative}
-which for monads $\C$ can be defined by
-$c_1 \star c_2 =
-   \tjoin \; (\tmap\;(\lambda\,a_1.\
-   \tmap\;(\lambda\,a_2.\ (a_1,a_2))
-    \;c_2)
-    \;c_1)$.
-% $c_1 \star c_2 =
-%    \tjoin \; (\tmap\;(\lambda\,a_1.\
-%    \tjoin \; (\tmap\;(\lambda\,a_2.\ \treturn\,(a_1,a_2))
-%     \;c_2))
-%     \;c_1)$.
-% $c \star d = \tjoin \; (\tmap\;(\lambda\,b.\ (c,\, \treturn\,b))\;d$.
-% $\tzip\;c = \tjoin \comp \tmap\,(\lambda\,a.\ (c,\, \treturn\,a)$.
-
-For negative product types, reflection and reification works as before:
-\[
-\def\arraystretch{1.3}
-\begin{array}{l@{\,}l@{~~}c@{~~}l}
-  \multicolumn 2 l {\up N} & : & \tNe\,N \todot \den N \\
-  \upG {\Comp P} & u & = & \tbind\,u\,\tfresh^P_\Gamma \\
-  \upG \natom    & u & = & \treturn\,u \\
-  \upG \top      & u & = & \ttempty \\
-  \upG {N_1 \ntimes N_2} & u & = & \left(
-    \upG {N_1} \; (\prj 1 u)
-    ,\
-    \upG {N_2} \; (\prj 2 u)
-    \right) \\
-\end{array}
-\
-\begin{array}{l@{\,}l@{~~}c@{~~}l}
-  \multicolumn 2 l {\down N} & : & \den N \todot \tNf\;N \\
-  \downG {\Comp P} & c  & = & \tmap\;(\down P)\;c \\
-  \downG \natom    & c  & = & \tne\,c \\
-  \downG \top      & () & = & \nunit \\
-  \downG {N_1 \ntimes N_2} & (b_1,b_2) & = & \npair
-    \, (\downG {N_1} \; b_1)
-    \, (\downG {N_2} \; b_2)
-    \\
-\end{array}
-\]
-Reflection for function types is also unchanged, except that $\tapp$
-expects a value argument now.
-\[
-\def\arraystretch{1.3}
-\begin{array}{l@{\,}l@{~~}c@{~~}l}
-  \upG {P \To N} & u & = & \lambda
-    \;(\tau : \Gamma \sublist \Delta)
-    \;(a : \den P \Delta)
-    . \
-    \up[\Delta] N \left( \tapp\,(\tren\,\tau\,u)\,(\down[\Delta] P a) \right)
-    \\
-  \downG {P \To N} & f & = & \tabs \left( \trunNf \left(
-    \stmap
-      \left(
-        \lambda
-          \;(\tau : (\cext \Gamma P) \sublist \Delta)
-          \;(a : \den P \Delta)
-          . \
-          f\; (\twk^P \rcomp \tau)\; a
-        \right)
-      \,\tfresh^P_\Gamma
-    \right) \right)
-    \\
-\end{array}
-\]
-The identity environment $\idenv \Gamma : \Cover {\den \Gamma} \Gamma$
-can only be generated in the monad, due to monadic positive
-reflection.
-\[
-\begin{array}{lcl}
-  \idenv{\cempty} & = & \treturn\,() \\
-  \idenv{\cext \Gamma P} & = &
-    (\tren\,\twk^P\,\idenv\Gamma) \star \tfresh^P_\Gamma \\
-\end{array}
-\]
-Putting things together, we obtain the normalization function
-\[
-  \tnorm\,(t : \Tm N \Gamma) = \trunNf\,
-    \left(
-      \tmap\,\left(\down N \comp \denty t\right)\,\idenv\Gamma
-    \right)
-.
-\]
-
-Taking stock, we have arrived at normal forms that eagerly introduce
-($\tNf$) and eliminate ($\tNe$) negative types and also eagerly
-introduce positive types ($\tVnf$).  However, the elimination of
-positive types is still rather non-deterministic.
-It is possible to only partially break up a composite positive type
-and leave smaller, but still composite positive types for later
-pattern matching.
-The last refinement, chaining also the positive eliminations, will be
-discussed in the following section.
-% The placement of $\tsplit$, $\tcase$, and $\tabort$ is only
-% constrained by the presence of the split variable
-
-\section{Focused Intuitionistic Propositional Logic}
-\label{sec:fipl}
-
-% Polarized lambda calculus \cite{espiritoSanto:entcs17}
-% divides types deeply into positive and negative.
-
-% \[
-% \begin{array}{lllrl@{\qquad}l}
-%   \PTy & \ni & P,Q & ::= & \patom
-%     \mid \tyzero \mid P_1 + P_2
-%     \mid \tyone \mid P_1 \times P_2
-%     \mid \Thunk N
-%   & \mbox{positive type}
-% \\
-%   \NTy & \ni & N,M & ::= & \natom
-%     \mid \top \mid N_1 \ntimes N_2
-%     \mid P \To N
-%     \mid \Comp P
-%   & \mbox{negative type}
-% \\
-%   \HTy & \ni & H & ::= & \patom \mid N
-%   & \mbox{hypothesis type}
-% \\
-% \end{array}
-% \]
-% Espirito Santo's version actually does not contain the positive
-% product $\times$, however, it appears in Zeilberger
-% \cite{zeilberger:PhD}.
-
-Polarized lambda-calculus
-\cite{zeilberger:PhD,espiritoSanto:entcs17}
-is a focused calculus, it eagerly employs
-invertible rules: the introduction rules for negative types and the
-elimination rules for positive types.  As a consequence of the latter,
-variables are either of atomic or negative type $H ::= \patom \mid N$.
-%($\HTy$).
-Contexts $\Gamma,\Delta$ are lists of $H$s.
-%$\HTy$s.
-%\fbox{$X ::= \patom \mid P$}.
-
-To add a variable of positive type $P$ to the context, we need to
-break it apart until only atoms and negative bits remain.  This is
-performed by maximal pattern matching, called the left-invertible
-phase of focalization.
-We express maximal pattern matching on $P$ as a strong functor
-$\Add P$ in the category of presheaves,
-mapping a presheaf $\J$ (``judgement'') to $\Add P \J$
-and a presheaf morphism $f : (\J \hatto \K)\Gamma$ to
-$\stmap[\Add P]_\Gamma f : \AddHyp P \J \Gamma \to \AddHyp P \K \Gamma$.
-For arbitrary $\J$ and $\Gamma$, the set \fbox{$\AddHyp P \J \Gamma$} is
-inductively constructed by the following rules:
-\begin{gather*}
-  \nru{\phyp}
-      {\J\,(\cext \Gamma \patom)}
-      {\AddHyp {\patom} \J \Gamma}
-\qquad
-  \nru{\nbranch}
-      {}
-      {\AddHyp {\tyzero} \J \Gamma}
-\qquad
-  \nru{\bbranch}
-      {\AddHyp {P_1} \J \Gamma \qquad
-       \AddHyp {P_2} \J \Gamma }
-      {\AddHyp {P_1 + P_2} \J \Gamma}
-\\[2ex]
-  \nru{\nhyp}
-      {\J\,(\cext \Gamma N)}
-      {\AddHyp {\Thunk N} \J \Gamma}
-\qquad
-  \nru{\nsplit}
-      {\J\,\Gamma}
-      {\AddHyp {\tyone} \J \Gamma}
-\qquad
-  \nru{\bsplit}
-      {\AddHyp {P_1} {\left(\Add{P_2}\,\J\right)} \Gamma}
-      {\AddHyp {P_1 \times P_2} \J \Gamma}
-\end{gather*}
-Note the recursive occurrence of $\Add{P_2}$ as argument to
-$\Add{P_1}$ in $\bsplit$, which makes $\Add P$ a nested datatype
-\cite{bird:nested}.  Agda supports such nested inductive types; but
-note that $\Add{P}$ is uncontroversial, since it could also be defined
-by recursion on $P$.   It is tempting to name $\bsplit$
-``$\tjoin$'' and $\nsplit$ ``$\treturn$''
-since $\Add P$ is a graded monad on the monoid $(1,\times)$
-of product types;
-however, this coincidence shall not matter for our further considerations.
-
-Focalization is a technique to remove don't-care non-determinism from
-proof search, and as such, polarized lambda calculus is foremost a
-calculus of normal forms.
-These normal forms are given by four mutually defined inductive
-families of presheaves $\tVnf\,P$,
-$\tNe\,N$, $\Cov\,\J$,
-and $\tNf\,N$.  As they are very similar to the CBPV normal forms
-given in the last section we only report the differences.
-
-Values \fbox{$v : \PVal P \Gamma$} are unchanged, they can refer to
-atomic positive hypotheses ($\pvar$) and normal $\tthunk$s.
-
-Neutrals \fbox{$\Ne N \Gamma$} start with a negative variable instead
-of with $\tforce$, as forcing thunks is already performed in $\nhyp$
-when adding analysing hypotheses of $\tThunk$ type.
-\[
-  \nru{\nvar}{N \in \Gamma}{\Ne N \Gamma}
-\]
-
-The Cover monad \fbox{$\Cover \J \Gamma$}
-lacks constructors $\tsplit$, $\tcase$ and $\tabort$
-since the pattern matching is taken care of by $\Add P$.
-\begin{gather*}
-  \nru{\treturn}{\J\,\Gamma}{\Cover \J \Gamma}
-\qquad
-  \nru{\tbind}
-      {\NeP {\Comp P} \Gamma \qquad \AddHypP P {\Cov\,\J} \Gamma}
-      {\Cover \J \Gamma}
-\end{gather*}
-
-The normal forms \fbox{$\Nf N \Gamma$} of negative type are unchanged
-with the exception that pattern matching happens eagerly in $\tabs$.
-\[
-  \nru{\tabs}
-      {\AddHypP P {\tNf\, N} \Gamma}
-      {\NfP {P \To N} \Gamma}
-\]
-
-All these inductive families are presheaves, due to factored
-presentation using $\Add P$ and $\Cov$ the proof is not a simple
-mutual induction.  Yet, in Agda, the generic proof goes through using a sized
-typing for these inductive families.  Similarily, defining the
-$\tjoin$ for monad $\Cov$ relies on sized typing.
-\[
-\begin{array}{lll}
-  \tjoin & : & \forall i.\
-    \Cov[i]\,(\Cov[\infty]\J) \todot \Cov[\infty]\J  \\
-  \tjoin[i+1]\,(\treturn[i]\;c) & = & c \\
-  \tjoin[i+1]\,(\tbind^i\;t\;k)
-    & = & \tbind^\infty\;
-     (t : \Ne P \Gamma)\;
-     \left(\stmap[\Add P]_\Gamma\;\tjoin[i]\;
-       \left(k : \AddHypP P {\Cov[i]\J} \Gamma
-       \right)
-     \right)
-\end{array}
-\]
-Herein, we used the sized typing of the constructors of $\Cov$:
-\[
-\begin{array}{lll}
-  \treturn & : & \forall i.\ \J \todot \Cov[i+1] \J \\
-  \tbind  & : & \forall i.\
-    \tNe\,(\Comp P) \todot \Add P\,(\Cov[i]\J) \todot \Cov[i+1]\J \\
-\end{array}
-\]
-
-Due to the eager splitting of positive hypotheses, reflection at type $P$
-now lives in the graded monad $\Add P$ rather than $\Cov$.  Further,
-as pattern matching may produce $n \geq 0$ cases, reflection cannot simply
-produce one positive semantic value; instead, one such value is needed
-for every branch.  We implement
-\fbox{$\reflect P : (\den P \hatto \J) \todot \Add P \J$}
-as a higher-order function expecting a continuation $k$ which is
-invoked for each generated branch with the semantic value of type $P$
-constructed for this branch.
-\[
-\begin{array}{l@{~}lcl}
-  \reflect {\patom} & k & = &
-    \phyp \left( k\;\twk^{\patom}\,(\pvar\,\tzero) \right) \\
-  \reflect {\Thunk N} & k & = &
-    \nhyp \left( k\;\twk^N \left( \up N (\nvar\,\tzero) \right) \right)\\
-  \reflect {\tyzero} & k & = & \nbranch \\
-  \reflect {P_1 + P_2} & k & = & \bbranch
-    \,\left( \treflect^{P_1}\,(\lambda\,\tau. \ k\,\tau \comp \iota_1) \right)
-    \,\left( \treflect^{P_2}\,(\lambda\,\tau. \ k\,\tau \comp \iota_2) \right)
-    \\
-  \reflect{\tyone} & k & = & \nsplit\;(k\;\tid\;\ttempty) \\
-  \reflect{P_1 \times P_2} & k & = & \bsplit
-    \,\left(
-      \treflect^{P_1} \left( \lambda\,\tau_1\,a_1. \
-      \treflect^{P_2} \left( \lambda\,\tau_2\,a_2. \
-        k\, (\tau_1 \rcomp \tau_2)\, (\tren\,\tau_2\,a_1, \ a_2)
-        \right)
-      \right)
-    \right)
-\end{array}
-\]
-Reflecting at a positive atomic type $\patom$ is the regular ending of a
-reflection pass: we call continuation $k$ with a fresh variable
-$\pvar\,\tzero$ of type $\patom$, making space for the variable
-using $\twk^{\patom}$.
-In case we end at type $\Thunk N$, we add a new variable
-$\nvar\,\tzero$ of type $N$ and pass it to $k$, after full
-$\eta$-expansion via $\up N$.  Two more endings are possible:
-At type
-$\tyzero$, we have reached an absurd case, meaning that no
-continuation is necessary since we can conclude with
-\emph{ex falsum quod libet}.
-At type $\tyone$, there is no need to add a new variable, as values
-of type $\tyone$ contain no information.  We simply pass the unit
-value $\ttempty$ to $k$ in this case.
-Reflecting at $P_1 + P_2$ generates two branches, which may result in
-several uses of the continuation $k$.
-In the first branch, we recursively reflect at $P_1$.  Its
-continuation will receive a semantic value in $\den{P_1}$ which we
-inject via $\iota_1$ into $\den{P_1 + P_2}$ to pass it to $k$.  The
-second branch proceeds analogously.
-Finally reflecting at $P_1 \times P_2$ means we first have to analyze
-$P_1$, and in each of the generated branches we continue to analyze
-$P_2$.  Thus $\reflect{P_2}$ is passed as a continuation to
-$\reflect{P_1}$.  Each reflection phase gives us a semantic value
-$a_i$ of type $P_i$, which we combine to a tuple before passing it to
-$k$.  Note also that the context extension $\tau_1$ created in the
-first phase needs to be composed with the context extension $\tau_2$
-of the second phase to transport $k$ into the final context.  Further,
-the value $a_1$ was constructed relative to the target of
-$\tau_1$ and still needs
-to be transported with $\tau_2$ before paired up with $a_2$.
-
-The method $\reflect P$ replaces previous uses of $\tfresh^P$ in
-reflection and reification at negative types.
-% and identity environment creation.
-\[
-\begin{array}{l@{~}lcl}
-  \downG {P \To N} & f & = & \tabs
-    \left(
-      \reflect[\Gamma] P
-        \left(
-          \lambda \, (\tau : \Gamma \sublist \Delta) \; a .\
-          \down[\Delta] P (f \; \tau \; a)
-        \right)
-    \right)
-  \\
-  \upG {\Comp P} & u & = & \tbind \; u
-    \left(
-      \reflect[\Gamma] P
-        \left(
-          \lambda \; \tau \; a .\
-          \treturn\; a
-        \right)
-    \right)
-  % \\
-  % \idenv{\cext \Gamma P} & = &
-\end{array}
-\]
-% To create an identity environment using $\treflect$, let us first
-% extend $\Add\_$ to contexts, following the intuition that
-Due to the strong limitation on context entries,
-the identity environment $\idenv \Gamma$ can be built
-straightforwardly using negative reflection.
-\[
-\begin{array}{lcl}
-  \idenv{\cempty} & = & \ttempty \\
-  \idenv{\cext \Gamma {\patom}} & = &
-    \left(
-      \tren\;\twk^{\patom}\;\idenv\Gamma
-      , \
-      \pvar\,\tzero
-    \right) \\
-  \idenv{\cext \Gamma N} & = &
-    \left(
-      \tren\;\twk^N\;\idenv\Gamma
-      , \
-      \upG N (\nvar\,\tzero)
-    \right) \\
-\end{array}
-\]
-
-The terms \fbox{$\Tm N \Gamma$} of the focused lambda calculus are the
-ones of CBPV minus the positive eliminations ($\tsplit$, $\tcase$,
-$\tabort$), the added negative variable rule ($\nvar$) and the
-necessary changes to the binders $\tabs$ and $\tbind$.
-\begin{gather*}
-  \nru{\nvar}{N \in \Gamma}{\Tm N \Gamma}
-\qquad
-  \nru{\tabs}
-      {\AddHypP P {\tTm\, N} \Gamma}
-      {\TmP {P \To N} \Gamma}
-\qquad
-  \nru{\tbind}
-      {\Tm {\Comp P} \Gamma \qquad \AddHypP P {\tTm\,N} \Gamma}
-      {\Tm N \Gamma}
-\end{gather*}
-Term interpretation
-\fbox{$\denty\_ : \Tm N \Gamma \to \den \Gamma \todot \den N$}
-shall be as for CBPV except that we need to exchange the
-interpretation function for binders \fbox{$\fden\_ : \Tm N {(\cext
-    \Gamma P)} \to \den \Gamma \todot \den{P \To N}$}.  Since a binder
-for $P$ performs a maximal splitting on $P$ and takes the form of a
-function defined by a case (and split) tree,
-applying it to a value $v$ of type $P$ amounts to a complete
-matching of $v$ against the case tree, and bind the remaining
-atomic and negative crumbs.  This matching can be defined for a
-generic evaluation function of type
-$\Evvv \J \Delta \Gamma = \den \Gamma \Delta \to \J\,\Delta$.
-\[
-\begin{array}{l@{~}l@{~}l@{~}lcl}
-  \multicolumn 6 {l} {\tmatch ~~:~~ \den P \Delta \to
-    \AddHyp P {\Evv \J \Delta} \todot \Evv \J \Delta }\\
-  \tmatch & x & (\phyp\,e) & \gamma & = & e\,(\gamma,\; \treturn\,x) \\
-  \tmatch & b & (\nhyp\,e) & \gamma & = & e\,(\gamma,\; b) \\
-  \tmatch & \ttempty & (\nsplit\,e) & \gamma & = & e\,\gamma \\
-  \tmatch & (a_1,a_2) & (\bsplit\,e) & \gamma & = &
-    \tmatch\;a_1\;(\tmap[\Add P]\,(\tmatch\;a_2)\;e)\;\gamma \\
-  \tmatch & a & \nbranch & \gamma & = & \tmagic\,a \\
-  \tmatch & (\iota_1\;a_1) & (\bbranch\;e_1\;e_2) & \gamma & = &
-    \tmatch\;a_1\;e_1\;\gamma \\
-  \tmatch & (\iota_2\;a_2) & (\bbranch\;e_1\;e_2) & \gamma & = &
-    \tmatch\;a_2\;e_2\;\gamma \\
-\end{array}
-\]
-With instantiations $\J = \den N$ and
-$\denty\_ : \Tm N \todot \Evv {\den N} \Delta$,
-the interpretation $\fden t$
-of binder $t : \AddHypP P {\tTm\,N} \Gamma$
-is defined as follows:
-\[
-  \fdent{t}
-    {(\gamma : \den \Gamma \Delta)}
-    \, (\tau : \Delta \sublist \Phi)
-    \, (a : \den P \Phi)
-    ~~=~~
-    \tmatch
-      \; a
-      \; (\tmap[\Add P]\;\denty\_\;t)
-      \; (\tren\,\tau\,\gamma)
-\]
-This completes the definition of the normalization function
-$\tnorm\, (t : \Tm N \Gamma)
-  = \down N \dent t {\idenv\Gamma}$.
-
-\section{Conclusion}
-\label{sec:concl}
-
-\bibliography{auto-fscd19}
-
-% \appendix
-
 \end{document}

### 2019-02-13 fscd19/fscd19.tex
[ fscd19 ] squeezing to make room for introduction

@@
   \tabort^{\tyone}     & u & = & \tunit \\
+  \tabort^P           & u & = & \tabort\;u \\
+\end{array}
+\qquad
+\begin{array}{l@{~}lcl}
+  \\
   \tabort^{A \times B} & u & = & \tpair\,(\tabort^A\,u)\,(\tabort^B\,u) \\
   \tabort^{A \To B}    & u & = & \tabs\,(\tabort^B\,(\tren\,\twk^A\,u)) \\
-  \tabort^P           & u & = & \tabort\;u \\
-\end{array}
+\end{array}
+% \begin{array}{l@{~}lcl}
+%   \multicolumn 4 l {\tabort^B ~:~ \tNe\, 0 \todot \tNf\,B} \\
+%   \tabort^{\tyone}     & u & = & \tunit \\
+%   \tabort^{A \times B} & u & = & \tpair\,(\tabort^A\,u)\,(\tabort^B\,u) \\
+%   \tabort^{A \To B}    & u & = & \tabs\,(\tabort^B\,(\tren\,\twk^A\,u)) \\
+%   \tabort^P           & u & = & \tabort\;u \\
+% \end{array}
 \]
@@
 for a new variable of type $A$, which is reflected into $\den A$ by
-\[
-\begin{array}{lcl@{\qquad}lcl}
-  \tfresh : \den A (\cext \Gamma A) \\
-  \tfresh = \up[\cext \Gamma A] A (\tvar\;\tzero)
-  . \\
-\end{array}
-\]
+\fbox{$\tfresh^A_\Gamma = \up[\cext \Gamma A] A {\vv0} : \den A (\cext \Gamma A)$}.
+% \[
+% \begin{array}{lcl@{\qquad}lcl}
+%   \tfresh : \den A (\cext \Gamma A) \\
+%   \tfresh = \up[\cext \Gamma A] A {\vv0}
+%   . \\
+% \end{array}
+% \]
 The ability to introduce fresh variables into a context, and to use
@@
 % \fbox{$\fden{t : B \from \cext \Gamma A} : \den \Gamma \todot \den{A \To B}$}.
-Herein the environment $\gamma$ lives in $\den \Gamma \Delta$, thus, $\dent t \gamma : \den A \Delta$.
+Herein, the environment $\gamma$ lives in $\den \Gamma \Delta$, thus, $\dent t \gamma : \den A \Delta$.
 % Herein, $\gamma : \den \Gamma \Delta$, thus, $\dent t \gamma : \den A \Delta$.
@@
 \begin{array}{lcl}
-  % \fdent{t : B \from \cext \Gamma A}\gamma & = &
-  %   \lambda
-  %     \,(\tau : \Delta \sublist \Phi)
-  %     \,(a : \den A \Phi)
-  %     .\
-  %     \dent t {(\tren\,\tau\,\gamma,\ a)} \\[1ex]
-  \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
-  \dent{\tabs\,t}\gamma
-      & = & \fdent t \gamma \\
-  % \dent{\tabs\,(t : B \from \cext \Gamma A)}\gamma \;
-  %    (\tau : \Delta \sublist \Phi)\, (a : \den A \Phi) & = &
-  %       \dent t {(\tren\,\tau\,\gamma,\ a)} \\
-  % \dent{\tabs\,(t : B \from \cext \Gamma A)}\gamma & = &
-  %    \lambda\, (\tau : \Delta \sublist \Phi)\, (a : \den A \Phi). \
-  %       \dent t {(\tren\,\tau\,\gamma,\ a)} \\
-  \dent{\tapp\,t\,u} \gamma & = &
-     \dent t \gamma\;\tid\;\dent u \gamma \\
-  % \dent{\tapp\,(t : A \To B \from \Gamma)\,u} \gamma & = &
-  %    \dent t \gamma\;\tid\;\dent u \gamma \\
   \dent{\tunit}\gamma & = & \ttempty \\
@@
     (\dent{t_1}\gamma,\ \dent{t_2}\gamma) \\
+  \dent{\tabs\,t}\gamma & = & \fdent t \gamma \\
+  \dent{\inj i t}\gamma & = & \iota_i \dent t \gamma \\
+  \\
+\end{array}
+\quad
+\begin{array}{lcl}
+  \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
   \dent{\prj i t}\gamma & = & \pi_i \dent t \gamma \\
-  \dent{\inj i t}\gamma & = & \iota_i \dent t \gamma \\
-  \dent{\tabort\,u}\gamma & = & \denabort\;\dent u \gamma \\
+  \dent{\tapp\,t\,u} \gamma & = &
+     \dent t \gamma\;\tid\;\dent u \gamma \\
   \dent{\tcase\;u\;t_1\;t_2}\gamma & = &
      \dencase\;\dent u \gamma\;\fdent{t_1}\gamma\;\fdent{t_2}\gamma \\
-  % \dent{\tcase\,(u : A_1 + A_2 \from \Gamma)\;t_1\;t_2}\gamma & = &
-    %  \dencase\;\dent u \gamma\;f_1\;f_2
-    % \ \bwhere \\
-    % \multicolumn 1 r {f_i\; (\tau : \Delta \sublist \Phi)\,(a_i : \den {A_i} \Phi)}
-    %     & = &
-    %     \dent {t_i} {(\tren\;\tau\;\gamma,\ {a_i})}
-   % \
-   %  (\lambda\; (\tau : \Delta \sublist \Phi)\,(b : \den A \Phi).\
-   %      \dent v {(\tren\;\tau\;\gamma,\ b)})
-    \\
-\end{array}
-\]
+  \dent{\tabort\,u}\gamma & = & \denabort\;\dent u \gamma \\
+\end{array}
+% \begin{array}{lcl}
+%   \\
+%   \dent{\tabs\,t}\gamma & = & \fdent t \gamma \\
+%   \dent{\tunit}\gamma & = & \ttempty \\
+%   \dent{\tpair\,t_1\,t_2}\gamma & = &
+%     (\dent{t_1}\gamma,\ \dent{t_2}\gamma) \\
+%   \dent{\inj i t}\gamma & = & \iota_i \dent t \gamma \\
+%   \\
+% \end{array}
+% \begin{array}{lcl}
+%   \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
+%   \dent{\tapp\,t\,u} \gamma & = &
+%      \dent t \gamma\;\tid\;\dent u \gamma \\
+%   \\
+%   \dent{\prj i t}\gamma & = & \pi_i \dent t \gamma \\
+%   \dent{\tcase\;u\;t_1\;t_2}\gamma & = &
+%      \dencase\;\dent u \gamma\;\fdent{t_1}\gamma\;\fdent{t_2}\gamma \\
+%   \dent{\tabort\,u}\gamma & = & \denabort\;\dent u \gamma \\
+% \end{array}
+% \begin{array}{lcl}
+%   \\
+%   \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
+%   \dent{\tabort\,u}\gamma & = & \denabort\;\dent u \gamma \\
+%   \dent{\tcase\;u\;t_1\;t_2}\gamma & = &
+%      \dencase\;\dent u \gamma\;\fdent{t_1}\gamma\;\fdent{t_2}\gamma \\
+%   \dent{\inj i t}\gamma & = & \iota_i \dent t \gamma \\
+% \end{array}
+% \begin{array}{lcl}
+%   \dent{\tabs\,t}\gamma & = & \fdent t \gamma \\
+%   \dent{\tapp\,t\,u} \gamma & = &
+%      \dent t \gamma\;\tid\;\dent u \gamma \\
+%   \dent{\tunit}\gamma & = & \ttempty \\
+%   \dent{\tpair\,t_1\,t_2}\gamma & = &
+%     (\dent{t_1}\gamma,\ \dent{t_2}\gamma) \\
+%   \dent{\prj i t}\gamma & = & \pi_i \dent t \gamma \\
+% \end{array}
+\]
+% \[
+% \begin{array}{lcl}
+%   % \fdent{t : B \from \cext \Gamma A}\gamma & = &
+%   %   \lambda
+%   %     \,(\tau : \Delta \sublist \Phi)
+%   %     \,(a : \den A \Phi)
+%   %     .\
+%   %     \dent t {(\tren\,\tau\,\gamma,\ a)} \\[1ex]
+%   \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
+%   \dent{\tabs\,t}\gamma
+%       & = & \fdent t \gamma \\
+%   % \dent{\tabs\,(t : B \from \cext \Gamma A)}\gamma \;
+%   %    (\tau : \Delta \sublist \Phi)\, (a : \den A \Phi) & = &
+%   %       \dent t {(\tren\,\tau\,\gamma,\ a)} \\
+%   % \dent{\tabs\,(t : B \from \cext \Gamma A)}\gamma & = &
+%   %    \lambda\, (\tau : \Delta \sublist \Phi)\, (a : \den A \Phi). \
+%   %       \dent t {(\tren\,\tau\,\gamma,\ a)} \\
+%   \dent{\tapp\,t\,u} \gamma & = &
+%      \dent t \gamma\;\tid\;\dent u \gamma \\
+%   % \dent{\tapp\,(t : A \To B \from \Gamma)\,u} \gamma & = &
+%   %    \dent t \gamma\;\tid\;\dent u \gamma \\
+%   \dent{\tunit}\gamma & = & \ttempty \\
+%   \dent{\tpair\,t_1\,t_2}\gamma & = &
+%     (\dent{t_1}\gamma,\ \dent{t_2}\gamma) \\
+%   \dent{\prj i t}\gamma & = & \pi_i \dent t \gamma \\
+%   \dent{\inj i t}\gamma & = & \iota_i \dent t \gamma \\
+%   \dent{\tabort\,u}\gamma & = & \denabort\;\dent u \gamma \\
+%   \dent{\tcase\;u\;t_1\;t_2}\gamma & = &
+%      \dencase\;\dent u \gamma\;\fdent{t_1}\gamma\;\fdent{t_2}\gamma \\
+%   % \dent{\tcase\,(u : A_1 + A_2 \from \Gamma)\;t_1\;t_2}\gamma & = &
+%     %  \dencase\;\dent u \gamma\;f_1\;f_2
+%     % \ \bwhere \\
+%     % \multicolumn 1 r {f_i\; (\tau : \Delta \sublist \Phi)\,(a_i : \den {A_i} \Phi)}
+%     %     & = &
+%     %     \dent {t_i} {(\tren\;\tau\;\gamma,\ {a_i})}
+%    % \
+%    %  (\lambda\; (\tau : \Delta \sublist \Phi)\,(b : \den A \Phi).\
+%    %      \dent v {(\tren\;\tau\;\gamma,\ b)})
+%     \\
+% \end{array}
+% \]
 For interpretation of the binders $\tabs$ and $\tcase$ we use the
@@
 and $\tNf\,N$.  As they are very similar to the CBPV normal forms
-given in the last section we only report the differences.
-
+given in the last section, we only report the differences.
+%
 Values \fbox{$v : \PVal P \Gamma$} are unchanged, they can refer to
 atomic positive hypotheses ($\pvar$) and normal $\tthunk$s.
-
+%
 Neutrals \fbox{$\Ne N \Gamma$} start with a negative variable instead
@@
 when adding analysing hypotheses of $\tThunk$ type.
+% \[
+%   \nru{\nvar}{N \in \Gamma}{\Ne N \Gamma}
+% \]
+%
+The normal forms \fbox{$\Nf N \Gamma$} of negative type are unchanged
+with the exception that pattern matching happens eagerly in $\tabs$.
 \[
   \nru{\nvar}{N \in \Gamma}{\Ne N \Gamma}
-\]
-
+\qquad
+\qquad
+  \nru{\tabs}
+      {\AddHypP P {\tNf\, N} \Gamma}
+      {\NfP {P \To N} \Gamma}
+\]
+%
 The Cover monad \fbox{$\Cover \J \Gamma$}
@@
 \end{gather*}
-
-The normal forms \fbox{$\Nf N \Gamma$} of negative type are unchanged
-with the exception that pattern matching happens eagerly in $\tabs$.
-\[
-  \nru{\tabs}
-      {\AddHypP P {\tNf\, N} \Gamma}
-      {\NfP {P \To N} \Gamma}
-\]
 
@@
   \idenv{\cext \Gamma {\patom}} & = &
-    \left(
+    (
       \tren\;\twk^{\patom}\;\idenv\Gamma
@@
       \pvar\,\tzero
-    \right) \\
+    ) \\
   \idenv{\cext \Gamma N} & = &
-    \left(
+    (
       \tren\;\twk^N\;\idenv\Gamma
@@
       \upG N (\nvar\,\tzero)
-    \right) \\
-\end{array}
+    ) \\
+\end{array}
+% \begin{array}{lcl}
+%   \idenv{\cempty} & = & \ttempty \\
+%   \idenv{\cext \Gamma {\patom}} & = &
+%     \left(
+%       \tren\;\twk^{\patom}\;\idenv\Gamma
+%       , \
+%       \pvar\,\tzero
+%     \right) \\
+%   \idenv{\cext \Gamma N} & = &
+%     \left(
+%       \tren\;\twk^N\;\idenv\Gamma
+%       , \
+%       \upG N (\nvar\,\tzero)
+%     \right) \\
+% \end{array}
 \]

### 2019-02-13 fscd19/fscd19.tex
[ fscd19 ] removed some fboxes

@@
 The category $\Cxt$ allows us to consider $A{\in}\_$ as a presheaf over
-$\Cxt\op$ for any $A$, witnessed by \fbox{$\treindex : \Gamma \sublist
-  \Delta \to A \in \Gamma \to A \in \Delta$}, which is the morphism
+$\Cxt\op$ for any $A$, witnessed by $\treindex : \Gamma \sublist
+  \Delta \to A \in \Gamma \to A \in \Delta$, which is the morphism
 part of functor $A{\in}\_$ from $\Cxt$ to $\Set$, mapping object $\Gamma$ to the
 set $A \in \Gamma$ of the indices of $A$ in $\Gamma$.
-The associated naturality laws \fbox{$\treindex\,\tid\,x \equiv x$}
-and \fbox{$\treindex\,\tau_2\,(\treindex\,\tau_1\,x) \equiv
-\treindex\,(\tau_1 \rcomp \tau_2)\,x$} hold propositionally.
+The associated naturality laws $\treindex\,\tid\,x \equiv x$
+and $\treindex\,\tau_2\,(\treindex\,\tau_1\,x) \equiv
+\treindex\,(\tau_1 \rcomp \tau_2)\,x$ hold propositionally.
 
@@
 % \nofbox{$\downG A : \Den A \Gamma \to \Nf A \Gamma$}.
-\emph{Normalization by evaluation} (NbE) decomposes normalization as
+\emph{Normalization by evaluation} (NbE)
 \fbox{\(
@@
 \)}
-which is
+decomposes normalization into
 \emph{evaluation}
-\nofbox{$\dent \_ \_ : (t : A \from \Gamma) \to \den \Gamma \todot
+\nofbox{$\denty\_ : (t : A \from \Gamma) \to \den \Gamma \todot
   \den A$}

### 2019-02-13 fscd19/fscd19.tex
[ fscd19 ] removed definition of lookup
[ fscd19 ] reference Filinski for continuation monad

@@
 \label{sec:intro}
+
+The idea behind \emph{normalization by evaluation} (NbE) is to utilize a
+standard interpreter, usually evaluating closed terms, to compute the
+normal form of an open term.
 
@@
 $\den{\cext \Gamma A} = \den \Gamma \hattimes \den A$
-and obtain a natural projection function from the semantic enironments:
-\[
-\begin{array}{lcl}
-  \tlookup\,(x : A \in \Gamma) & : &  \den \Gamma \todot \den A \\
-  \tlookup\;\tzero     & = & \pi_2 \\
-  \tlookup\;(\tsuc\,x) & = & \tlookup\;x \comp \pi_1 \\
-\end{array}
+and obtain a natural projection function
+$\tlookup\,(x : A \in \Gamma) : \den \Gamma \todot \den A$
+from the semantic enironments.
+% \[
 % \begin{array}{lcl}
-%   \tlookup & : & A \in \Gamma \to \den \Gamma \todot \den A \\
+%   \tlookup\,(x : A \in \Gamma) & : &  \den \Gamma \todot \den A \\
 %   \tlookup\;\tzero     & = & \pi_2 \\
@@
 % \end{array}
-\]
+% % \begin{array}{lcl}
+% %   \tlookup & : & A \in \Gamma \to \den \Gamma \todot \den A \\
+% %   \tlookup\;\tzero     & = & \pi_2 \\
+% %   \tlookup\;(\tsuc\,x) & = & \tlookup\;x \comp \pi_1 \\
+% % \end{array}
+% \]
 The evaluation function
@@
 
-An alternative implementation of $\C$ is the continuation monad $\CC$
-on presheaves as, for instance, employed by Ilik et al.\
-\cite{ilikLeeHerbelin:apal10}
-for Kripke models and NbE of classical logic.
+As already observed by Filinski
+\cite[Section~5.4]{filinski:semaccounttdpe}
+\cite[Section~3.2]{filinski:tlca01},
+normalization by evaluation can be carried out in the continuation
+monad.
+In our setting, we use a continuation monad $\CC$ on presheaves
+defined as
+% An alternative implementation of $\C$ is the continuation monad $\CC$
+% on presheaves as, for instance, employed by Ilik et al.\
+% \cite{ilikLeeHerbelin:apal10}
+% for Kripke models and NbE of classical logic.
 \[
   \CC\;\J = \forall A.\ (\J \hatto \tNf\,A) \hatto \tNf\,A
+  .
 \]
 The answer type of this continuation monad is always $\tNF$, however,
-we are polymorphic in the simple type $A$ of normal forms we produce.
+we are polymorphic in the simple type $A$ of normal forms we produce.%
 
@@
 performed by maximal pattern matching, called the left-invertible
-phase of focalization.
+phase of focalization.%
+\footnote{Filinski \cite[Section~4]{filinski:tlca01} achieves maximal pattern
+matching through an additional, ordered context $\Theta$ for positive
+variables which are eagerly split.}
 We express maximal pattern matching on $P$ as a strong functor

### 2019-02-14 fscd19/fscd19.tex
[ fscd19 ] start on introduction

@@
 normal form of an open term.
-
-Weak coproducts.
-Weak sheaf (non-unique pasting).
-
-Contributions:
-\begin{enumerate}
-\item Abstract cover monad.
-\item NbE for CBPV and polarized lambda-calculus.
-\end{enumerate}
-
-CBPV differs from PolLambda in that hypotheses can be arbitrary
-positive formulas in CBPV, but only positive atoms or negative
-formulas in PolLambda.  This means that in CBPV, there is more choice
-on when and how to decompose a hypothesis into ones of smaller
-positive type, whereas in PolLambda there is no freedom: mandatory
-maximal splitting is dictated by the left inversion phase.
-
-Normal forms of CBPV thus retain individual shallow pattern matching
-constructs, while PolLambda has only deep, maximal pattern matching.
-
-Agda formalization:
-Sized types, nested data types.
-
-Structural focalization for sequent calculus \cite{simmons:tocl14}.
-
-\nocite{bergerSchwichtenberg:lics91}
-
-\nocite{jungTiuryn:tlca93}
-
-\nocite{altenkirchHofmannStreicher:ctcs95}
-
-\nocite{abel:habil}
-
-\nocite{danvy:popl96}
-
-\nocite{fioreSimpson:tlca99}
-
-\nocite{altenkirchDybjerHofmannScott:lics01}
-
-\nocite{balatDiCosmoFiore:popl04}
-
-\nocite{altenkirchUustalu:flops04}
-
-\nocite{barral:PhD}
-
-\nocite{scherer:popl17}
+The normal form is achieved by a type-directed \emph{reification}
+procedure after evaluating the open term to a semantic value, mapping
+the free variables to corresponding \emph{unknowns} in the semantics.
+The literal use of a standard interpreter
+can be achieved for the pure simply-typed lambda-calculus
+\cite{bergerSchwichtenberg:lics91,filinski:semaccounttdpe}
+by modelling uninterpreted base types as sets of neutral (aka atomic)
+terms, or more precisely, as presheaves or sets of neutral term
+families, in order to facilitate fresh bound variable generation during
+reification of functions to lambdas.
+Thanks to $\eta$-equality at function types, free variables of
+function type can be reflected into the semantics as functions
+applying the variable to their reified argument,
+becoming a neutral.  This mechanism provides us with unknowns of
+function type which can be faithfully reified to normal forms.
+% Since neutrals are closed under application to normal arguments, they
+% can act as unknown functions, storing the reified semantic argument
+% they are invoked with (thus, these unknowns are aka accumulators
+% \cite{gregoireleroy:icfp02}).
+
+For the extension to sum types (logically disjunctions and categorically
+weak coproducts), this reflection trick does not work anymore.  A
+semantic value of binary sum type is either a left or a right
+injection, but the decision between left or right cannot be taken at
+reflection time, since a variable of sum type does not provide us with
+the information.  A literal standard interpreter for closed terms can
+thus no longer be used for NbE.  Instead, we can utilize a monadic
+interpreter.  When the interpreter attempts a case distinction on an
+unknown of sum type, it asks an oracle whether the unknown is a left
+or a right injection.  The oracle returns one of these alternatives,
+wrapping a new unknown in the injection.  The communication with the
+oracle can be modeled in a \emph{monad} $\C$,
+% with the service $\tNe\,(A + B) \to \C\,(\den A + \den B)$
+which records the questions asked and the continuation of the
+interpreter for each of the possible answers.
+A monadic semantic value is thus a \emph{case tree} where the leaves
+are occupied by non-monadic values.
+In this article we only consider weak sum types, producing non-unique
+normal forms, where it does not
+matter in which order the questions are asked (commuting case splits),
+and whether the same
+question is asked several times (redundant case splits).
+The model would need refinement for strong, extensional sums
+\cite{altenkirchDybjerHofmannScott:lics01,altenkirchUustalu:flops04,balatDiCosmoFiore:popl04,barral:PhD,scherer:popl17}.
+
+Filinski \cite{filinski:tlca01} studied NbE for Moggi's computional lambda
+calculus \cite{moggi:infcomp91} shedding light on the difference
+between call-by-name (CBN) and call-by-value (CBV) NbE, where Danvy's
+type-directed partial evaluation \cite{danvy:popl96} falls into the
+latter class.  The contribution of the computational lambda calculus
+is to make explicit where the monad is envoked during monadic
+evaluation, and this placement of the monad carries over to the NbE
+setting.  Moggi's studies were continued by Levy \cite{levy:hosc06}
+who designed the call-by-push-value (CBPV) lambda-calculus to embed
+both the CBN and CBV lambda calculus.
+
+In this work, we formulate NbE for CBPV
+(Section~\ref{sec:cbpv}), with the aim to investigate
+later whether CBN and CBV NbE can be recovered from CBPV NbE via the
+standard translations of the CBN and CBV calculi into CBPV.
+
+In contrast to the normal forms of CBN NbE, which is the algorithmic
+counterpart of the completeness proof for intuitionistic propositional
+logic (IPL) using Beth models, CBPV NbE gives us more restrained
+normal forms, where the production of a value via injections cannot be
+interrupted by more questions to the oracle.  In the research field of
+focalization \cite{andreoli:focusing,liangMiller:csl07} we speak of
+\emph{chaining non-invertible introductions}.
+Invertible introductions are already chained in NbE thanks to
+extensionality ($\eta$) for function (an more generally, negative)
+types.  Non-invertible eliminations are also happing in a chain,
+building neutrals.  What is missing from the picture is the chaining
+of invertible eliminations, i.e., case distinctions and, more
+generally, pattern matching.  The picture is completed by extending
+NbE to \emph{polarized lambda calculus}
+\cite{zeilberger:PhD,brockNannestadSchuermann:lpar10,espiritoSanto:entcs17}
+in Section~\ref{sec:fipl}.
+
+In our presentation of the various lambda calculi and NbE, we use
+presheaves for the abstract syntax and the semantics
+\cite{altenkirchHofmannStreicher:ctcs95}.
+
+In contrast to Zeilberger \cite{zeilberger:PhD}.
+
+% Weak coproducts.
+% Weak sheaf (non-unique pasting).
+
+% Contributions:
+% \begin{enumerate}
+% \item Abstract cover monad.
+% \item NbE for CBPV and polarized lambda-calculus.
+% \end{enumerate}
+
+% CBPV differs from PolLambda in that hypotheses can be arbitrary
+% positive formulas in CBPV, but only positive atoms or negative
+% formulas in PolLambda.  This means that in CBPV, there is more choice
+% on when and how to decompose a hypothesis into ones of smaller
+% positive type, whereas in PolLambda there is no freedom: mandatory
+% maximal splitting is dictated by the left inversion phase.
+
+% Normal forms of CBPV thus retain individual shallow pattern matching
+% constructs, while PolLambda has only deep, maximal pattern matching.
+
+% Agda formalization:
+% Sized types, nested data types.
+
+% Structural focalization for sequent calculus \cite{simmons:tocl14}.
+
+% \nocite{bergerSchwichtenberg:lics91}
+
+% \nocite{jungTiuryn:tlca93}
+
+% \nocite{altenkirchHofmannStreicher:ctcs95}
+
+% \nocite{abel:habil}
+
+% \nocite{danvy:popl96}
+
+% \nocite{fioreSimpson:tlca99}
+
+% \nocite{altenkirchDybjerHofmannScott:lics01}
+
+% \nocite{balatDiCosmoFiore:popl04}
+
+% \nocite{altenkirchUustalu:flops04}
+
+% \nocite{barral:PhD}
+
+% \nocite{scherer:popl17}
 
@@
 \label{sec:fipl}
+\label{sec:pol}
 

### 2019-02-14 fscd19/fscd19.tex
[ fscd19 ] an introduction and a conclusion

@@
 
-\ccsdesc[100]{General and reference~General literature}
-\ccsdesc[100]{General and reference}%TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
+% \begin{CCSXML}
+% <ccs2012>
+% <concept>
+% <concept_id>10003752.10003790.10011740</concept_id>
+% <concept_desc>Theory of computation~Type theory</concept_desc>
+% <concept_significance>300</concept_significance>
+% </concept>
+% <concept>
+% <concept_id>10003752.10010124.10010125.10010127</concept_id>
+% <concept_desc>Theory of computation~Functional constructs</concept_desc>
+% <concept_significance>300</concept_significance>
+% </concept>
+% <concept>
+% <concept_id>10003752.10010124.10010125.10010130</concept_id>
+% <concept_desc>Theory of computation~Type structures</concept_desc>
+% <concept_significance>300</concept_significance>
+% </concept>
+% <concept>
+% <concept_id>10003752.10010124.10010131.10010137</concept_id>
+% <concept_desc>Theory of computation~Categorical semantics</concept_desc>
+% <concept_significance>300</concept_significance>
+% </concept>
+% <concept>
+% <concept_id>10003752.10003790.10003792</concept_id>
+% <concept_desc>Theory of computation~Proof theory</concept_desc>
+% <concept_significance>100</concept_significance>
+% </concept>
+% </ccs2012>
+% \end{CCSXML}
+
+\ccsdesc[500]{Theory of computation~Type theory}
+\ccsdesc[500]{Theory of computation~Type structures}
+\ccsdesc[300]{Theory of computation~Functional constructs}
+\ccsdesc[300]{Theory of computation~Proof theory}
+\ccsdesc[300]{Theory of computation~Categorical semantics}
+\ccsdesc[100]{Theory of computation~Operational semantics}
+%TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
 
@@
 
-\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
+\supplement{\url{https://andreasabel.github.io/ipl/html/NfModelMonad.html}}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
 
@@
 procedure after evaluating the open term to a semantic value, mapping
+(\emph{reflecting})
 the free variables to corresponding \emph{unknowns} in the semantics.
@@
 applying the variable to their reified argument,
-becoming a neutral.  This mechanism provides us with unknowns of
+forming a neutral term.  This mechanism provides us with unknowns of
 function type which can be faithfully reified to normal forms.
@@
 
-For the extension to sum types (logically disjunctions and categorically
+For the extension to sum types (logically, disjunctions and categorically,
 weak coproducts), this reflection trick does not work anymore.  A
@@
 the information.  A literal standard interpreter for closed terms can
-thus no longer be used for NbE.  Instead, we can utilize a monadic
+thus no longer be used for NbE;  instead, we can utilize a monadic
 interpreter.  When the interpreter attempts a case distinction on an
@@
 A monadic semantic value is thus a \emph{case tree} where the leaves
-are occupied by non-monadic values.
+are occupied by non-monadic values \cite{altenkirchUustalu:flops04}.
 In this article we only consider weak sum types, producing non-unique
@@
 Filinski \cite{filinski:tlca01} studied NbE for Moggi's computional lambda
-calculus \cite{moggi:infcomp91} shedding light on the difference
+calculus \cite{moggi:infcomp91}, shedding light on the difference
 between call-by-name (CBN) and call-by-value (CBV) NbE, where Danvy's
@@
 latter class.  The contribution of the computational lambda calculus
-is to make explicit where the monad is envoked during monadic
+is to make explicit where the monad is invoked during monadic
 evaluation, and this placement of the monad carries over to the NbE
@@
 Invertible introductions are already chained in NbE thanks to
-extensionality ($\eta$) for function (an more generally, negative)
-types.  Non-invertible eliminations are also happing in a chain,
+extensionality ($\eta$) for function, and more generally, negative
+types.  Non-invertible eliminations are also happening in a chain,
 building neutrals.  What is missing from the picture is the chaining
@@
 
-In our presentation of the various lambda calculi and NbE, we use
-presheaves for the abstract syntax and the semantics
-\cite{altenkirchHofmannStreicher:ctcs95}.
-
-In contrast to Zeilberger \cite{zeilberger:PhD}.
+In our presentation of the various lambda calculi we ignore the
+concrete syntax, only consider the abstract syntax obtained by the
+Curry-Howard-Isomorphism.  A term is simply a derivation tree whose
+nodes are rule invokations.  Thus, a intrinsically typed,
+nameless syntax is most natural,
+and our syntactic classes are just presheaves over the category of
+typing contexts and renamings.  The use of presheaves then smoothly
+extents to the semantic constructions \cite{altenkirchHofmannStreicher:ctcs95}.
+
+Concerning the presentation of polarized lambda calculus, we depart from
+Zeilberger \cite{zeilberger:PhD} who employs \emph{a priori}
+infinitary syntax, modelling a case tree as a meta-level function mapping
+well-typed patterns to branches.  Instead, we use a graded monad
+representing complete pattern matching over a newly added hypothesis,
+which is in spirit akin to Filinski's \cite[Section~4]{filinski:tlca01}
+and Krishnaswami's \cite{krishnaswami:popl09} treatment of eager
+pattern matching using a separate context of variables to be matched
+on.
+
+Our design choices were guided by an Agda formalization of sections 2
+(complete) and 4 (partial), available at \url{https://github.com/andreasabel/ipl}.
+Agda was particularly helpful to correctly handle the renamings
+abundantly present when working with presheaves.
 
@@
 \label{sec:concl}
+
+We have implemented NbE for CBPV and polarized lambda calculus
+formulated with intrinsically well-typed syntax and presheaf
+semantics.  As a side result, we have proven semantically
+that the normal forms of both systems are logically complete, \ie,
+each derivable judgement $\Gamma \der N$ has a normal derivation.
+It remains to show that NbE for these calculi is also computationally
+sound and complete, i.e., the computational behavior of term and
+normal form should agree, and normalization should decide a suitable
+equational theory on terms.
+
+Additionally, a natural question to investigate is whether known CBN
+and CBV NbE algorithms can be obtained from our NbE algorithms by
+embedding simply-typed lambda calculus into our polarized calculi,
+using known CBN and CBV translations.  Further, we would like to study
+the NbE algorithm for STLC arising from the optimal translation, i.e.,
+the one inserting a minimal amount of $\tThunk$ and $\tComp$ transitions.
 

### 2019-02-15 fscd19/fscd19.tex
[ fscd19 ] add citation to Catarina Coquand csl93
[ fscd19 ] spell-checker run

@@
 
-Filinski \cite{filinski:tlca01} studied NbE for Moggi's computional lambda
+Filinski \cite{filinski:tlca01} studied NbE for Moggi's computational lambda
 calculus \cite{moggi:infcomp91}, shedding light on the difference
@@
 Curry-Howard-Isomorphism.  A term is simply a derivation tree whose
-nodes are rule invokations.  Thus, a intrinsically typed,
+nodes are rule invocations.  Thus, a intrinsically typed,
 nameless syntax is most natural,
@@
 typing contexts and renamings.  The use of presheaves then smoothly
-extents to the semantic constructions \cite{altenkirchHofmannStreicher:ctcs95}.
+extents to the semantic constructions
+\cite{catarina:csl93,altenkirchHofmannStreicher:ctcs95}.
 
@@
 but only at base types $\tybase$.
-Further, case distintion via $\tcase$ and $\tabort$
+Further, case distinction via $\tcase$ and $\tabort$
 is restricted to positive types $P$.
@@
 \fbox{$\tNf\,A \todot A{\from}\_$} for such presheaf morphisms.
-(The point on the arrow is mnemoic for \emph{pointwise}.)
+(The point on the arrow is mnemonic for \emph{pointwise}.)
 Slightly abusive, we shall extend this notation to $n$-ary
@@
 
-Presheafs (co)products $\hat0$, $\hat1$, $\hatplus$, and $\hattimes$ are
+Presheaves (co)products $\hat0$, $\hat1$, $\hatplus$, and $\hattimes$ are
 constructed pointwise, \eg, $\hat0\,\Gamma = \bzero$, and given two
@@
 inhabitant of the empty set, which means that reflection at the empty
-type would not be definable.  Similarily, the setting
+type would not be definable.  Similarly, the setting
 $\den{A + B} = \den A \hatplus \den B$ is refuted by
@@
 $\tlookup\,(x : A \in \Gamma) : \den \Gamma \todot \den A$
-from the semantic enironments.
+from the semantic environments.
 % \[
@@
 partial evaluation \cite[Figure~8]{danvy:popl96}.  However, he uses
-shift-reset style continutations which can be programmed in the
+shift-reset style continuations which can be programmed in the
 continuation monad, and relies on Scheme's \texttt{gensym} to produce
@@
 mutual induction.  Yet, in Agda, the generic proof goes through using a sized
-typing for these inductive families.  Similarily, defining the
+typing for these inductive families.  Similarly, defining the
 $\tjoin$ for monad $\Cov$ relies on sized typing.
@@
 
-\clearpage
+%\clearpage
 \appendix

### 2019-02-16 fscd19/fscd19.tex
[ fscd19 ] post-submission proof reading

@@
 normal form of an open term.
-The normal form is achieved by a type-directed \emph{reification}
+The normal form is obtained by a type-directed \emph{reification}
 procedure after evaluating the open term to a semantic value, mapping
@@
 
-For the extension to sum types (logically, disjunctions and categorically,
+For the extension to sum types (logically, disjunctions, and categorically,
 weak coproducts), this reflection trick does not work anymore.  A
@@
 reflection time, since a variable of sum type does not provide us with
-the information.  A literal standard interpreter for closed terms can
+such information.  A literal standard interpreter for closed terms can
 thus no longer be used for NbE;  instead, we can utilize a monadic
@@
 or a right injection.  The oracle returns one of these alternatives,
-wrapping a new unknown in the injection.  The communication with the
+wrapping a new unknown in the respective injection.  The communication with the
 oracle can be modeled in a \emph{monad} $\C$,
@@
 extensionality ($\eta$) for function, and more generally, negative
-types.  Non-invertible eliminations are also happening in a chain,
+types.  Non-invertible eliminations are also happening in a chain when
 building neutrals.  What is missing from the picture is the chaining
@@
 nameless syntax is most natural,
-and our syntactic classes are just presheaves over the category of
+and our syntactic classes are all presheaves over the category of
 typing contexts and renamings.  The use of presheaves then smoothly
@@
 set $A \in \Gamma$ of the indices of $A$ in $\Gamma$.
-The associated naturality laws $\treindex\,\tid\,x \equiv x$
+The associated functor laws $\treindex\,\tid\,x \equiv x$
 and $\treindex\,\tau_2\,(\treindex\,\tau_1\,x) \equiv
@@
 rules of intuitionistic propositional logic (IPL) under the
-Curry-Howard isomorphism.  Using shorthand
+Curry-Howard isomorphism, where $A \from \Gamma$ is to be read as
+``$A$ follows from $\Gamma$''.  Using shorthand
 \fbox{$\vv n = \tvar\,(\tsuc^n\,\tzero)$} for the $n$th variable,
@@
 We leave the exact connection to a printable
-syntax of the STLC to imagination of the reader,
+syntax of the STLC to the imagination of the reader,
 as we shall not be concerned with considering
@@
 defined by recursion over $t : A \from \Gamma$,
-and naturality laws for $\tren$ analogously to $\treindex$.
+and functor laws for $\tren$ analogously to $\treindex$.
 
@@
 \end{gather*}
-These rules only allow the elimination of neutrals; this guarantees
+These rules only allow the elimination of \emph{neutrals}; this
+restriction guarantees
 the subformula property and prevents any kind of computational
@@
 The new rule $\tne$ embeds $\tNe$ into $\tNf$,
-but only at base types $\tybase$.
+but only at base types $\tybase$ \cite[Section~3.3]{altenkirchHofmannStreicher:ctcs95}.
 Further, case distinction via $\tcase$ and $\tabort$
@@
 the construction of their inhabitants proceeds mechanically, without
-any choices.  In contrast, to construct an inhabitant of a
-\emph{positive} type choice is required: whether case distinction is
+any choices.  In contrast, constructing an inhabitant of a
+\emph{positive} type involves choice: whether case distinction is
 required, and which introduction to pick in the end ($\tinj_1$ or $\tinj_2$).
@@
 resp.) and $A{\from}\_$.  We shall simply write, for instance,
-\fbox{$\tNf\,A \todot A{\from}\_$} for such presheaf morphisms.
+$\tNf\,A \todot A{\from}\_$ for such presheaf morphisms.
 (The point on the arrow is mnemonic for \emph{pointwise}.)
@@
 
+\begin{remark*}
 While the coproduct eliminations $\tcase$ and $\tabort$ are limited to
@@
 the branches.
+\end{remark*}
 
@@
 \emph{reification}
-\nofbox{$\down A : \den A \todot \tNf A$}.
+\nofbox{$\down A : \den A \todot \tNf A$}
+(aka quoting).
 The role of evaluation is to produce from a term the corresponding
@@
 the unknowns.  The mapping $\up A : \tNe\,A \todot \den A$ from
-neutrals to unknowns is called \emph{reflection}, and defined mutually
+neutrals to unknowns is called \emph{reflection} (aka unquoting), and defined mutually
 with reification by induction on type $A$.
@@
   \upG   \tyone \; u        & = & \ttempty \\
-  \downG \tyone \ttempty & = & \tunit \\
+  \downG \tyone \; \ttempty & = & \tunit \\
 \end{array}
@@
 work.  For instance, with $\den \tyzero = \hat0$, simply
-$\up[\tyzero] \tyzero \vv0 : \bzero$ would give us an
+% $\up[\tyzero] \tyzero \vv0 : \bzero$
+$\tfresh^{\tyzero}_{\cempty} : \bzero$
+would give us an
 inhabitant of the empty set, which means that reflection at the empty
@@
 % $\up[\tyone + \tyone]^{\tyone + \tyone} \vv0 : \bone \bplus \bone$
-$\up[A + B]{A + B} \vv0 : \den A (A{+}B) \bplus \den B (A{+}B)$
+$\tfresh^{A + B}_{\cempty} : \den A (A{+}B) \bplus \den B (A{+}B)$
+% $\up[A + B]{A + B} \vv0 : \den A (A{+}B) \bplus \den B (A{+}B)$
 which would require us to make a decision of whether $A$ holds or $B$ holds
@@
 the free cover monad
-$\Cov$ defined inductively by $\treturn[\Cov]$, $\tabort^{\Cov}$, and
+$\Cov$ defined
+%inductively by
+as an inductive family with constructors
+$\treturn[\Cov]$, $\tabort^{\Cov}$, and
 $\tcase^{\Cov}$.  One can visualize an element $c : \Cov\,\A\,\Gamma$ as
@@
 or by an element of $\A$ (see $\treturn$).  Functoriality amounts to
-replacing the labels of the $\treturn$-leaves, and the monadic bind replaces
+replacing the labels of the $\treturn$-leaves, and the monadic bind
+(aka Kleisli extension) replaces
 these leaves by further case trees.  (The uninspiring $\tjoin[\Cov]$
@@
   \den   \tybase & = & \C\,(\tNe\,\tybase) \\
-  \upG   \tybase & = & \treturn[\C] \\
-  \downG \tybase & = & \trunNf[\C] \comp \tmap[\C]\tne \\
+  \up    \tybase & = & \treturn[\C] \\
+  \down  \tybase & = & \trunNf[\C] \comp \tmap[\C]\tne \\
 \end{array}
@@
   \den   \tyzero & = & \C\,\hat0 \\
-  \upG   \tyzero & = & \tabort^\C \\
-  \downG \tyzero & = & \trunNf[\C] \comp \tmap[\C]\tmagic \\
+  \up    \tyzero & = & \tabort^\C \\
+  \down  \tyzero & = & \trunNf[\C] \comp \tmap[\C]\tmagic \\
 \end{array}
@@
 $\stmap[\C]_\Gamma\, \ell : \C\A\,\Gamma \to \C\,\B\,\Gamma$ already for
-a local presheaf morphism $\ell : (\A \hatto \B)\Gamma$.
+a ``local'' presheaf morphism $\ell : (\A \hatto \B)\Gamma$.
 The typings are
@@
 % \]
-For interpretation of the binders $\tabs$ and $\tcase$ we use the
+For the interpretation of the binders $\tabs$ and $\tcase$ we use the
 mutually defined $\fden{\_}$.
@@
 \]
-The answer type of this continuation monad is always $\tNF$, however,
+The answer type of this continuation monad is always $\tNf$, however,
 we are polymorphic in the simple type $A$ of normal forms we produce.%
@@
 
-As remarked in the previous section, the placement of the monad $\C$
-in the type semantics is a bit wasteful:  Each positive type is
+The placement of the monad $\C$
+in the type semantics
+of the previous section
+is a bit wasteful:  Each positive type is
 prefixed by $\C$.  In our grammar of normal forms, this corresponds to
@@
 \emph{computation types} $\underline B$,
-however, we shall stick to our terminology
-which is known from research on focalization.
+yet we shall stick to our terminology
+which is common in publications on focalization.
 However, we shall use $\tThunk$ for switch $\downarrow$ and $\tComp$ for
@@
   \PTy & \ni & P,Q & ::= & \patom
+    \mid \tyone \mid P_1 \times P_2
     \mid \tyzero \mid P_1 + P_2
-    \mid \tyone \mid P_1 \times P_2
     \mid \Thunk N
@@
 $\trun^N : \C \den N \to \den N$ for negative types $N$, by recursion
-on $N$.  Informally, speaking this makes all negative types \emph{monadic}.
-
-Contexts are lists of \emph{positive} types as in CBPV variables stand
+on $N$.  Informally speaking, this makes all negative types \emph{monadic}.
+
+Contexts are lists of \emph{positive} types since in CBPV variables stand
 for values.  Interpretation of contexts $\den \Gamma$ is again defined
@@
 only available for $\tComp$-types but for arbitrary negative types
-$N$.  This is justified due to the monadic character of negative
+$N$.  This is justified by the monadic character of negative
 types, by virtue of $\trun^N$.  Finally, there are eliminators
@@
 
-Reification $\down P : \den P \todot \tVNf\,P$ at positive types produces
-a normal value, and $\down N : \den P \todot \tNf\,N$ at negative types a
+Reification $\down P : \den P \todot \tVNf\,P$ at positive types $P$ produces
+a normal value, and $\down N : \den N \todot \tNf\,N$ at negative
+types $N$ a
 normal term.  During reification of function types $P \To N$ in
@@
 
-For negative product types, reflection and reification works as before:
+For negative types, reflection and reification works as before:
 \[
@@
 is a focused calculus, it eagerly employs
-invertible rules: the introduction rules for negative types and the
+so-called \emph{invertible} rules: the introduction rules for negative types and the
 elimination rules for positive types.  As a consequence of the latter,
@@
 $\stmap[\Add P]_\Gamma f : \AddHyp P \J \Gamma \to \AddHyp P \K \Gamma$.
-For arbitrary $\J$ and $\Gamma$, the set \fbox{$\AddHyp P \J \Gamma$} is
+For arbitrary $\J$ and $\Gamma$, the family \fbox{$\AddHyp P \J \Gamma$} is
 inductively constructed by the following rules:
@@
 of with $\tforce$, as forcing thunks is already performed in $\nhyp$
-when adding analysing hypotheses of $\tThunk$ type.
+when adding hypotheses of $\tThunk$ type.
 % \[
@@
 The normal forms \fbox{$\Nf N \Gamma$} of negative type are unchanged
-with the exception that pattern matching happens eagerly in $\tabs$.
+with the exception that pattern matching happens eagerly in $\tabs$,
+by virtue of $\Add P$.
 \[
@@
 typing for these inductive families.  Similarly, defining the
-$\tjoin$ for monad $\Cov$ relies on sized typing.
+$\tjoin$ for monad $\Cov$ relies on sized typing \cite{abel:PhD}.
 \[
@@
 as pattern matching may produce $n \geq 0$ cases, reflection cannot simply
-produce one positive semantic value; instead, one such value is needed
+produce a single positive semantic value; instead, one such value is needed
 for every branch.  We implement
@@
 % extend $\Add\_$ to contexts, following the intuition that
-Due to the strong limitation on context entries,
+%Due to the strong limitation on context entries,
+Due to the absence of composite positive types in contexts,
 the identity environment $\idenv \Gamma$ can be built
@@
   \nru{\tbind}
-      {\Tm {\Comp P} \Gamma \qquad \AddHypP P {\tTm\,N} \Gamma}
+      {\TmP {\Comp P} \Gamma \qquad \AddHypP P {\tTm\,N} \Gamma}
       {\Tm N \Gamma}
@@
     \tmatch\;a_1\;(\tmap[\Add P]\,(\tmatch\;a_2)\;e)\;\gamma \\
-  \tmatch & a & \nbranch & \gamma & = & \tmagic\,a \\
+  \tmatch & a & \nbranch & \gamma & = & \tmagic\;a \\
   \tmatch & (\iota_1\;a_1) & (\bbranch\;e_1\;e_2) & \gamma & = &

### 2019-03-03 types19/types19.tex
[ types19 ] TYPES abstract

\nonstopmode
\RequirePackage[dvipsnames]{xcolor}

\documentclass[a4paper]{easychair}
\usepackage{amssymb}
%\usepackage{dsmath}
%\usepackage{bbm}
\usepackage{bbold}
\usepackage[utf8x]{inputenc}

\usepackage{natbib}
\usepackage{hyperref}
% \usepackage{agda} % colors

\definecolor{DeepJungleGreen}{rgb}{0.0, 0.29, 0.29}
\definecolor{DarkSpringGreen}{rgb}{0.09, 0.45, 0.27}
\definecolor{ForestGreenTraditional}{rgb}{0.0, 0.27, 0.13}
\definecolor{GreenPigment}{rgb}{0.0, 0.65, 0.31}
\definecolor{CalPolyPomonaGreen}{rgb}{0.12, 0.3, 0.17}
\definecolor{CSSGreen}{rgb}{0.0, 0.5, 0.0}
\definecolor{DarkCerulean}{rgb}{0.03, 0.27, 0.49}
\definecolor{DarkRaspberry}{rgb}{0.53, 0.15, 0.34}

\colorlet{DarkBlue}{MidnightBlue!80!black}
\definecolor{HanBlue}{rgb}{0.27, 0.42, 0.81}
\definecolor{FrenchBlue}{rgb}{0.0, 0.45, 0.73}
\definecolor{EgyptianBlue}{rgb}{0.06, 0.2, 0.65}

\hypersetup{colorlinks=true,
  linkcolor=Red,
  citecolor=DarkRaspberry,%DarkPurple,%DarkCerulean,%CSSGreen,%CalPolyPomonaGreen,%GreenPigment,%ForestGreenTraditional,%DarkSpringGreen,%DeepJungleGreen,%ForestGreen,
  urlcolor=MidnightBlue}

\makeatletter
\renewcommand\bibsection{%
  \section*{\large{\refname}}%
}%
\makeatother

%% Macros

\input{macros}

\renewcommand{\patom}{\ensuremath{o}}
\renewcommand{\tVnf}{\mathsf{Nf}}

\begin{document}

\title{Normalization by Evaluation for Call-by-Push-Value}
\titlerunning{Normalization by Evaluation for Call-by-Push-Value}
\authorrunning{A. Abel}
\author{Andreas Abel}
\institute{
  Department of Computer Science and Engineering,
  Gothenburg University
}

% Keywords:
% Modal typing, linear typing, strictness typing, positivity checking, guarded recursion

\maketitle

%\begin{abstract}
%\end{abstract}

\noindent

Normalization by evaluation (NbE) \citep{bergerSchwichtenberg:lics91}
is the interpretation of
an (open) term of type $A$ as value in a suitable model $\den A$,
followed by \emph{reification} of the value to a normal form of type
$A$.  Functions $f$ in $\den{A \To B}$ are reified as
$\lambda$-abstractions whose bodies are obtained
by \emph{reflecting} a fresh variable of type $A$ as value $a$ in
$\den{A}$ and reifying the application $f\,a$ at type $B$.
A suitable model that supports fresh variable generation
are \emph{presheaves} over the category of
typing contexts $\Gamma$ and order-preserving embeddings
$\Gamma \subseteq \Gamma'$, where a base type $o$ is interpreted as
the presheaf $\tNe\;o$ of neutral normal forms of type $o$, and
function types by the presheaf exponential aka Kripke function space
\citep{catarina:csl93,altenkirchHofmannStreicher:ctcs95}.

NbE for sum types requires a refinement of the model, as reflection of
a variable of type $A + B$ as a value in $\den{A + B}$ requires case
distinction in the model.  One such refinement are \emph{sheaves}
\citep{altenkirchDybjerHofmannScott:lics01}; another is the use of a
monad $\C$
\citep{filinski:semaccounttdpe,filinski:tlca01,barral:PhD}
in the category of presheaves
for the interpretation of sum types:
$\den{A + B} = \C(\den A + \den B)$.
The smallest such \emph{``cover'' monad} $\C$ are binary trees where leaves are
the monadic unit aka $\treturn$, and the nodes case distinctions
over neutrals $\tNe\,(A_1 + A_2)$ of sum type.  When leaves are normal
forms, the whole tree represents a normal form, thus,
$\trunNf : \C(\tNf\,A) \to \tNf\,A$ is trivial.
This \emph{running of the monad}
on normal forms represents the algorithmic part of the sheaf condition
on $\tNf\,A$ and extends as $\trun : \C \den A \to \den A$ to all
semantic types.

The given interpretation of sum types
$\den{A + B} = \C(\den A + \den B)$ corresponds to the call-by-name
(CBN) lambda calculus with lazy constructors.  NbE can also be
performed in call-by-value (CBV) style, then the monad is placed in
the codomain of function types: $\den{A \To B} = \den A \To \C \den B$
\citep{danvy:popl96}.  A systematic semantic analysis of CBN and CBV
lambda-calculi has been pioneered by \citet{moggi:infcomp91} through
translation into his computational lambda calculus;
\citet{filinski:tlca01} studied NbE for the latter calculus using the
continuation monad.  Moggi's work was continued and refined
by \citet{levy:hosc06} who subsumed CBV and CBN under his monadic
call-by-push-value (CBPV) calculus.  In this work, we study NbE for CBPV.

CBPV was designed to study lambda-calculus with effects.  It separates
types into \emph{value} types $P$ and \emph{computation} types $N$, which we, in
analogy to polarized lambda-calculus
\citep{zeilberger:PhD,brockNannestadSchuermann:lpar10,espiritoSanto:entcs17}
refer to as \emph{positive} and \emph{negative} types.
Variables stand for values, thus, have positive types.
The monad that models the effects is placed at the transition from values
to computations $\Comp P$, and computations can be embedded into
values by \emph{thunking} ($\Thunk N$).
\[
\begin{array}{lllrl@{\qquad}l}
  \PTy & \ni & P,Q & ::= & \patom
    % \mid \tyone \mid P_1 \times P_2
    % \mid \tyzero
    \mid P_1 + P_2
    \mid \Thunk N
  & \mbox{positive type / value type}
\\
  \NTy & \ni & N,M & ::= & % \natom
    % \mid \top \mid N_1 \ntimes N_2
    % \mid
    P \To N
    \mid \Comp P
  & \mbox{negative type / computation type}
\\
\end{array}
\]
We restrict to a fragment of \emph{pure} CBPV
with a single positive connective, sum
types $P_1 + P_2$, and a single negative connective, call-by-value
function types $P \To N$.  While we have no proper effects, the
evaluation of open terms requires the effect of case distinction over
neutrals, modeled by a cover monad $\C$.  In the following, we give
inductive definitions of the presheaves of
normal ($\tNf$) and neutral normal forms
($\tNe$) of our fragment of CBPV and a concrete, strong cover monad $\Cov$.
\begin{small}
\begin{gather*}
  \nru{\tvar}{\Var \patom \Gamma}{\PValNf {\patom} \Gamma}
\quad
  \nru{\tthunk}{\Nf N \Gamma}{\PValNfP{\Thunk N}\Gamma}
\quad
  \nru{\tinj_i}{\PValNf{P_i}\Gamma}{\PValNfP{P_1 + P_2}\Gamma}
%\\[2ex]
\quad
  \nru{\tret}
      {\CoverP {\tVNf\,P}  \Gamma}
      {\NfP {\Comp P} \Gamma}
\quad
  \nru{\tabs}
      {\Nf N {(\cext \Gamma P)}}
      {\NfP {P \To N} \Gamma}
\end{gather*}
\end{small}
\begin{small}
\begin{gather*}
  \nru{\tforce}{\VarP {\Thunk N} \Gamma}{\Ne N \Gamma}
\qquad
  \nru{\tapp}
      {\NeP {P \To N} \Gamma \qquad \PValNf P \Gamma}
      {\Ne N \Gamma}
\qquad
  \nru{\tbind}
      {\NeP {\Comp P} \Gamma \qquad \Cover \J (\cext \Gamma P)}
      {\Cover \J \Gamma}
\\[2ex]
  \nru{\treturn}{\J\,\Gamma}{\Cover \J \Gamma}
\qquad
  \nru{\tcase}
      {\VarP {P_1 + P_2} \Gamma
       \qquad \Cover \J (\cext \Gamma {P_1})
       \qquad \Cover \J (\cext \Gamma {P_2})
      }
      {\Cover \J \Gamma}
\end{gather*}
\end{small}
\noindent
($\J$ stands for an arbitrary presheaf in $\Cov\,\J$.)
Normal forms are introductions starting from a variable of base type,
except that the monad is placed at the
transition $\tret$ from positive to negative types ($\Comp P$).
Neutrals are eliminations of variables of type $\Thunk N$ into a
positive type $\Comp P$, and can then be bound (see $\tbind$)
to a variable of type $P$ to be used in a computation.  Variables of
sum type $P_1 + P_2$ can be utilized in computations through a
$\tcase$ split.
\emph{Terms} $\tTm$ of CBPV are obtained by blurring the
distinction between $\tNe$ and $\tNf$,
generalizing $\tbind$ and $\tcase$ from $\Cov\,\J$ to
computations $\tTm\,N$,
relaxing $\tvar$ to variables
of arbitrary type $P$, and $\tforce$ to arbitrary terms of type
$\Thunk N$.  Terms are evaluated in the following presheaf model,
which interprets $\tThunk$ as the identity and $\tComp$ as $\Cov$.
\[
\begin{array}[t]{lcl}
  % \den{\tyone} & = & \hatone \\
  % \den{P_1 \times P_2} & = & \den{P_1} \hattimes \den{P_2} \\
  % \den{\tyzero} & = & \hatzero \\
  \den{P_1 + P_2} & = & \den{P_1} \hatplus \den{P_2} \\
  \den{\Thunk N} & = & \den N \\
  \den{\patom} & = & \tVar\;\patom \\
\end{array}
\qquad
\begin{array}[t]{lcl}
  % \den{\top} & = & \hatone \\
  % \den{N_1 \ntimes N_2} & = & \den{N_1} \hattimes \den{N_2} \\
  % \\
  \den{P \To N} & = & \den P \hatto \den N \\
  \den{\Comp P} & = & \Cov \den P \\
  % \den{\natom}  & = & \C (\tNe\, \natom) \\
\end{array}
\]
The evaluation of $\tbind$-terms in $\Tm N$ relies on
$\trun : \Cov \den N \to \den N$, which makes any computation type
monadic.  Reflection $\uparrow$ and reification $\downarrow$ take the
usual form, only that reflection of positive variables is monadic, to
allow the complete splitting of sums via $\tcase$.
It is invoked by reification of functions $\down {P \To N}$ via $\trunNf$.
\[
\begin{array}[t]{lcl}
  \up P  & : & \tVar\, P \to \Cov\,\den P \\
  \up N & : & \tNe\,N \to \den N \\
\end{array}
\qquad
\begin{array}[t]{lcl}
  \down P & : & \den P \to \tVNf\,P \\
  \down N & : & \den N \to \tNf\,N \\
\end{array}
\]
The details of our construction, plus extension to product types and
polarized lambda calculus, can be found in the full version at
\url{https://arxiv.org/abs/1902.06097}.
A partial Agda formalization is available at
\url{https://github.com/andreasabel/ipl}.

% \[
% \begin{array}[t]{lcl}
%   \upG P  & : & \Var P \Gamma \to \Cov\,\den P \, \Gamma\\
%   \downG P & : & \den P \Gamma \to \VNf P \Gamma \\
% \end{array}
% \qquad
% \begin{array}[t]{l@{\,}l@{~~}c@{~~}l}
%   \multicolumn 2 l {\upG N} & : & \Ne N \Gamma \to \den N \Gamma \\
%   \multicolumn 2 l {\downG N} & : & \den N \Gamma \to \Nf N \Gamma \\
% \end{array}
% \]

% Positive \fbox{$\VNf P \Gamma$} and negative \fbox{$\Nf N \Gamma$}
% normal forms.
% \begin{small}
% \begin{gather*}
%   \nru{\tvar}{\Var \patom \Gamma}{\PValNf {\patom} \Gamma}
% \quad
%   \nru{\tthunk}{\Nf N \Gamma}{\PValNfP{\Thunk N}\Gamma}
% \quad
%   \nru{\tinj_i}{\PValNf{P_i}\Gamma}{\PValNfP{P_1 + P_2}\Gamma}
% %\\[2ex]
% \quad
%   \nru{\tret}
%       {\CoverP {\tVNf\,P}  \Gamma}
%       {\NfP {\Comp P} \Gamma}
% \quad
%   \nru{\tabs}
%       {\Nf N {\cext \Gamma P}}
%       {\NfP {P \To N} \Gamma}
% \end{gather*}
% \end{small}

% Neutral normal forms \fbox{$\Ne N \Gamma$}
% and the strong monad \fbox{$\Cov$}.
% \begin{small}
% \begin{gather*}
%   \nru{\tforce}{\VarP {\Thunk N} \Gamma}{\Ne N \Gamma}
% \qquad
%   \nru{\tapp}
%       {\NeP {P \To N} \Gamma \qquad \PValNf P \Gamma}
%       {\Ne N \Gamma}
% \qquad
%   \nru{\tbind}
%       {\NeP {\Comp P} \Gamma \qquad \Cover \J (\cext \Gamma P)}
%       {\Cover \J \Gamma}
% \\[2ex]
%   \nru{\treturn}{\J\,\Gamma}{\Cover \J \Gamma}
% \qquad
%   \nru{\tcase}
%       {\VarP {P_1 + P_2} \Gamma
%        \qquad \Cover \J (\cext \Gamma {P_1})
%        \qquad \Cover \J (\cext \Gamma {P_2})
%       }
%       {\Cover \J \Gamma}
% \end{gather*}
% \end{small}
% Finally, normal forms of negative types are defined as inductive
% family \fbox{$\Nf N \Gamma$}.  They are generated by maximal negative
% introduction ($\tabs$, $\npair$, $\nunit$) until a negative atom or
% $\Comp P$ is reached.  Then, elimination of neutrals and variables is
% possible through the $\Cov$ monad until an answer can be given in form
% of a base neutral ($\tNe\,\natom$) or a normal value.

% We again can run the cover monad on normal forms, \ie, have
% $\trunNf : \Cov\,(\tNf N) \todot \tNf N$, which extends to
% negative semantic values
% $\trun : \Cov \den N \todot \den N$.

\paragraph*{Acknowledgments.}
%Thanks to the anonymous referees, who helped improving the quality of
%this abstract through their feedback.
This work was supported by
Vetenskapsr\aa{}det under Grant
  No.~621-2014-4864 \emph{Termination Certificates
  for Dependently-Typed Programs and Proofs via Refinement Types}
and by the EU Cost Action CA15123 \emph{Types for programming and verification}.

\begin{footnotesize}

\bibliographystyle{abbrvnat}
\bibliography{auto-types19}

\end{footnotesize}
\end{document}

### 2019-03-04 types19/types19.tex
[ types19 ] Q and M not needed in type grammar
[ types19 ] Little polishing for submission

@@
 
-NbE for sum types requires a refinement of the model, as reflection of
+NbE for sum types requires a refinement of the model, since reflection of
 a variable of type $A + B$ as a value in $\den{A + B}$ requires case
@@
 \begin{array}{lllrl@{\qquad}l}
-  \PTy & \ni & P,Q & ::= & \patom
+  \PTy & \ni & P & ::= & \patom
     % \mid \tyone \mid P_1 \times P_2
@@
 \\
-  \NTy & \ni & N,M & ::= & % \natom
+  \NTy & \ni & N & ::= & % \natom
     % \mid \top \mid N_1 \ntimes N_2
@@
 ($\J$ stands for an arbitrary presheaf in $\Cov\,\J$.)
-Normal forms are introductions starting from a variable of base type,
-except that the monad is placed at the
+Normal forms start from a variable of base type and continue with
+introductions,
+except that the services of the monad can be used at the
 transition $\tret$ from positive to negative types ($\Comp P$).
 Neutrals are eliminations of variables of type $\Thunk N$ into a
-positive type $\Comp P$, and can then be bound (see $\tbind$)
-to a variable of type $P$ to be used in a computation.  Variables of
+positive type $\Comp P$, and can then be bound
+to a variable of type $P$ to be used in a computation (see $\tbind$).
+Variables of
 sum type $P_1 + P_2$ can be utilized in computations through a
 $\tcase$ split.
+
 \emph{Terms} $\tTm$ of CBPV are obtained by blurring the
@@
 computations $\tTm\,N$,
-relaxing $\tvar$ to variables
-of arbitrary type $P$, and $\tforce$ to arbitrary terms of type
+and relaxing $\tvar$ to variables
+of arbitrary type $P$ and $\tforce$ to arbitrary terms of type
 $\Thunk N$.  Terms are evaluated in the following presheaf model,
@@
 \]
-The evaluation of $\tbind$-terms in $\Tm N$ relies on
+The evaluation of $\tbind$ terms in $\Tm N$ relies on
 $\trun : \Cov \den N \to \den N$, which makes any computation type
-monadic.  Reflection $\uparrow$ and reification $\downarrow$ take the
+monadic.  Reflection $\uparrow$ and reification $\downarrow$ are
+defined mutually by induction on the type.  They take the
 usual form, only that reflection of positive variables is monadic, to
@@
 
-% \[
-% \begin{array}[t]{lcl}
-%   \upG P  & : & \Var P \Gamma \to \Cov\,\den P \, \Gamma\\
-%   \downG P & : & \den P \Gamma \to \VNf P \Gamma \\
-% \end{array}
-% \qquad
-% \begin{array}[t]{l@{\,}l@{~~}c@{~~}l}
-%   \multicolumn 2 l {\upG N} & : & \Ne N \Gamma \to \den N \Gamma \\
-%   \multicolumn 2 l {\downG N} & : & \den N \Gamma \to \Nf N \Gamma \\
-% \end{array}
-% \]
-
-% Positive \fbox{$\VNf P \Gamma$} and negative \fbox{$\Nf N \Gamma$}
-% normal forms.
-% \begin{small}
-% \begin{gather*}
-%   \nru{\tvar}{\Var \patom \Gamma}{\PValNf {\patom} \Gamma}
-% \quad
-%   \nru{\tthunk}{\Nf N \Gamma}{\PValNfP{\Thunk N}\Gamma}
-% \quad
-%   \nru{\tinj_i}{\PValNf{P_i}\Gamma}{\PValNfP{P_1 + P_2}\Gamma}
-% %\\[2ex]
-% \quad
-%   \nru{\tret}
-%       {\CoverP {\tVNf\,P}  \Gamma}
-%       {\NfP {\Comp P} \Gamma}
-% \quad
-%   \nru{\tabs}
-%       {\Nf N {\cext \Gamma P}}
-%       {\NfP {P \To N} \Gamma}
-% \end{gather*}
-% \end{small}
-
-% Neutral normal forms \fbox{$\Ne N \Gamma$}
-% and the strong monad \fbox{$\Cov$}.
-% \begin{small}
-% \begin{gather*}
-%   \nru{\tforce}{\VarP {\Thunk N} \Gamma}{\Ne N \Gamma}
-% \qquad
-%   \nru{\tapp}
-%       {\NeP {P \To N} \Gamma \qquad \PValNf P \Gamma}
-%       {\Ne N \Gamma}
-% \qquad
-%   \nru{\tbind}
-%       {\NeP {\Comp P} \Gamma \qquad \Cover \J (\cext \Gamma P)}
-%       {\Cover \J \Gamma}
-% \\[2ex]
-%   \nru{\treturn}{\J\,\Gamma}{\Cover \J \Gamma}
-% \qquad
-%   \nru{\tcase}
-%       {\VarP {P_1 + P_2} \Gamma
-%        \qquad \Cover \J (\cext \Gamma {P_1})
-%        \qquad \Cover \J (\cext \Gamma {P_2})
-%       }
-%       {\Cover \J \Gamma}
-% \end{gather*}
-% \end{small}
-% Finally, normal forms of negative types are defined as inductive
-% family \fbox{$\Nf N \Gamma$}.  They are generated by maximal negative
-% introduction ($\tabs$, $\npair$, $\nunit$) until a negative atom or
-% $\Comp P$ is reached.  Then, elimination of neutrals and variables is
-% possible through the $\Cov$ monad until an answer can be given in form
-% of a base neutral ($\tNe\,\natom$) or a normal value.
-
-% We again can run the cover monad on normal forms, \ie, have
-% $\trunNf : \Cov\,(\tNf N) \todot \tNf N$, which extends to
-% negative semantic values
-% $\trun : \Cov \den N \todot \den N$.
-
 \paragraph*{Acknowledgments.}

### 2019-03-26 fscd19/fscd19.tex
[ fscd19 ] arXiv version
[ fscd19 ] typos spotted by Hugo

@@
 \ccsdesc[100]{Theory of computation~Operational semantics}
-%TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
+%TODO mandatory: Please choose ACM 2012 classifications from
+%https://dl.acm.org/ccs/ccs_flat.cfm
+
+% ACM old
+% F.3.2 Semantics of Programming Languages (D.3.1)
+% F.4.1 Mathematical Logic (F.1.1, I.2.2, I.2.3, I.2.4)
+
+% MSC2010
+% 03F05 Cut-elimination and normal-form theorems
 
@@
 Thorsten in turn credits his inspiration to an ALF proof by Thierry
-Coquand.}
+Coquand.  Thanks to Hugo Herbelin for feedback and spotting some typos.}
 % \acknowledgements{ The presentation of this NbE proof and its
@@
 
-%\nolinenumbers %uncomment to disable line numbering
-
-%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
+\nolinenumbers %uncomment to disable line numbering
+
+\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
 
@@
 Curry-Howard-Isomorphism.  A term is simply a derivation tree whose
-nodes are rule invocations.  Thus, a intrinsically typed,
+nodes are rule invocations.  Thus, an intrinsically typed,
 nameless syntax is most natural,
@@
 \ie, $\tm n \cong t$
-with respect to a equational theory $\cong$ on terms
+with respect to an equational theory $\cong$ on terms
 (see \cref{sec:eq}).
@@
 
-All these inductive families are presheaves, due to factored
-presentation using $\Add P$ and $\Cov$ the proof is not a simple
+All these inductive families are presheaves, however, due to the factored
+presentation using $\Add P$ and $\Cov$, the proof is not a simple
 mutual induction.  Yet, in Agda, the generic proof goes through using a sized
@@
 case tree whose leaves are injections.  A normal form of atomic type
-($o$) is always case tree whose leaves are neutral.
+($o$) is always a case tree whose leaves are neutral.
 The $\beta$ and $\pi$ laws determine the shape of neutrals.

### 2019-04-01 fscd19/fscd19.tex
[ fscd19 ] first changes after review reports

@@
 
-\nolinenumbers %uncomment to disable line numbering
-
-\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
+% \nolinenumbers %uncomment to disable line numbering
+
+% \hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
 
@@
 the neutral $u : \Ne A \Gamma$ is
-transported into $\Ne A \Gamma$ via reindexing with
+transported into $\Ne A \Delta$ via reindexing with
 $\tau : \Gamma \sublist \Delta$, in order to be applicable to the normal form
@@
 
-To recapitulate, a monad $\C$ on presheaves is first an endofunctor,
+Recall that a monad $\C$ on presheaves is first an endofunctor,
 i.e., it maps any presheaf $\A$ to the presheaf $\C\,\A$ and any
@@
 (aka Kleisli extension) replaces
-these leaves by further case trees.  (The uninspiring $\tjoin[\Cov]$
-flattens a 2-level case tree, \ie, a case tree with case trees as
-leaves, into a single one.)
+these leaves by further case trees.  Bind is realized via $\tjoin[\Cov]$
+which flattens a 2-level case tree, \ie, a case tree with case trees as
+leaves, into a single one.
 Finally $\trunNf[\Cov]$ is a simple recursion on the tree, replacing $\tcase^{\Cov}$
@@
 
-All semantic types fulfill the weak sheaf condition aka weak pasting,
-meaning there is
-a natural transformation $\trun^A : \C\,\den A \todot \den A$ for any simple
-type $A$.  In other words, we can \emph{run} the monad, pushing its
-effects into $\den A$.  We proceed by induction on $A$.
-Positive types $P$ are already monadic,
+The unwinding of the monad $\C$ via $\trunNf$ can be extended to a
+natural transformation $\trun^A : \C\,\den A \todot \den A$ for any simple
+type $A$, meaning we can \emph{run} the monad,
+pushing its effects into $\den A$.%
+\footnote{In categorical terminology, the existence of
+$\trun^A$ means that
+all semantic types $\den{A}$ fulfill the \emph{weak sheaf condition},
+aka \emph{weak pasting}.}
+% All semantic types fulfill the weak sheaf condition aka weak pasting,
+% meaning there is
+% a natural transformation $\trun^A : \C\,\den A \todot \den A$ for any simple
+% type $A$.  In other words, we can \emph{run} the monad, pushing its
+% effects into $\den A$.
+We proceed by induction on $A$.
+Positive semantic types $\den P$ are already monadic,
 and $\trun^P$ is simply the $\tjoin$ of the monad $\C$.
@@
 
-As already observed by Filinski
+Besides the free cover monad $\Cov$, there is another canonical
+implementation of its interface, the continuation monad.
+Filinski
 \cite[Section~5.4]{filinski:semaccounttdpe}
-\cite[Section~3.2]{filinski:tlca01},
-normalization by evaluation can be carried out in the continuation
-monad.
+\cite[Section~3.2]{filinski:tlca01}
+already utilized the continuation monad for
+normalization by evaluation.
 In our setting, we use a continuation monad $\CC$ on presheaves
@@
 The method $\trunNf[\C]$ of cover monad $\C$ can be extended to
-$\trun^N : \C \den N \to \den N$ for negative types $N$, by recursion
-on $N$.  Informally speaking, this makes all negative types \emph{monadic}.
+$\trun^N : \C \den N \todot \den N$ for negative types $N$, by recursion
+on $N$.
+This makes the effects of $\C$ available at any negative type,
+in other words, makes all negative types \emph{monadic}.
 
@@
 inductively defined strong monad \fbox{$\Cov$}.
+In the following rules, parameter $\J$ (for ``judgement'') stands for an
+arbitrary presheaf.
 \begin{gather*}
@@
   \multicolumn 6 {l} {\tmatch ~~:~~ \den P \Delta \to
-    \AddHyp P {\Evv \J \Delta} \todot \Evv \J \Delta }\\
+    \AddHypP P {\Evv \J \Delta} \todot \Evv \J \Delta }\\
   \tmatch & x & (\phyp\,e) & \gamma & = & e\,(\gamma,\; \treturn\,x) \\

### 2019-04-01 fscd19/response-fscd19.txt
[ fscd19 ] Response first material

A big thanks to the reviewers who spent considerable time to read and
understand our paper and to provide valuable feedback and pointers to
relevant literature.  We will take all comments into account when
preparing the revision.

- Missing theorems (1,3)

- Incomplete formalization (1,2)

- The choice of abstract syntax (3)

- Relation to category theory (2,3)

- CBPV as logical system (2)

No NbE for effects.

  let x = launchNuclearMissile; 1
  in  x + 1

- Future work: relationship to CBN CBV (1,2)

> ----------------------- REVIEW 1 ---------------------
> PAPER: 50
> TITLE: Normalization by Evaluation for Call-by-Push-Value and Polarized Lambda-Calculus
> AUTHORS: Andreas Abel and Christian Sattler
>
> Overall evaluation: 0 (borderline paper)
>
> ----------- Overall evaluation -----------
>
> This paper present a definition of normalization by evaluation for
> call-by-push-value.  The authors first present NbE for a simply type
> lambda calculus with sums. For that, they need to introduce a
> (coverage) monad to be able to do backtracking or to build a case tree
> of possible normalization. A version of this monad is then used to
> interpret computations in CBPV and gives rise to an NbE algorithm for
> CBPV. Finally, the authors discuss modification for focused polarized
> logic, which is closed to CPBV.
>
> General Comment :
>
> This is an interesting work on how to extend NbE technique that are
> compatible with the introduction of effect in the calculus. The paper
> seems mathematically correct and the idea of using such a monad to
> extend NbE to potentially effectful terms is very natural.
>
> However, I have severe reserves about the paper. Therefore I cannot be
> positive about this paper.
>
> - The paper is missing theorem in two senses. First, it is missing a
>   clear statement of what is claimed to be achieved, namely “logical
>   consistency”. The expected theorem are not there.  I was expected at
>   least computational soundness. Also I found quite disappointing that
>   the the combination with the CBN and CBV translation to CBPV has not
>   been investigated, without even explaining why. No time, too
>   difficult, not sure is corresponds to usual NbE ?

To investigate the relationship to the CBN and CBV translations was an
idea we had during the writing, we did not have the time to
investigate it yet.

> - The formalization in Agda i just a link to a GitHub repo that is
>   very hard to dive into. There is no README to know what has been
>   done and where is has been done.  We don’t know what has been
>   formalized in Agda for section 4. It seems that nothing has been
>   formalized for section 3, why ?  The repo looks more like the link
>   to an ongoing project than a link to some documented and finish
>   material aimed at being understood by a reader.

The Agda formalization helped us getting the constructions correct,
but the paper is intended to be self-contained without the Agda
formalization.

> Minor remarks:
>
> The paper is longer than the 15 pages including biblio of the CFP, but
> I didn’t take this into account in my review.
>
> However, putting the equational theory in Appendix rather feels like
> the authors needed to save more space and decided quite arbitrarily to
> cut this.
>
> In the definition of ↑A⇒B ,  where does ∆ comes from ?

When quantifying over (τ : Γ ⊆ Δ) we implicitly quantify over Δ,
similarily as when writing ∀ Δ ⊇ Γ. P, which is a common shorthand for
∀ Δ. Γ ⊆ Δ ⇒ P.

> typos:
> - “a equational” page 5
> - I am not sure "ex falsum quod libet” is correct, one usually says  “ex falso quodlibet”
> - line 307: type of ↓N, P should be N

Thanks, FIXED.

> ----------------------- REVIEW 2 ---------------------
> PAPER: 50
> TITLE: Normalization by Evaluation for Call-by-Push-Value and Polarized Lambda-Calculus
> AUTHORS: Andreas Abel and Christian Sattler
>
> Overall evaluation: 1 (weak accept)
>
> ----------- Overall evaluation -----------
>
> The paper develops normalization by evaluation for simply-typed,
> call-by-push-value and polarized lambda-calculus, all of them
> featuring sums, in a category of presheaves equipped with a
> monad. Besides the fact that the paper seems to give the first
> definition of NbE for CBPV and polarized lambda-calculus, the main
> contribution is the observation how and where to place the monad,
> which simplifies the definitions considerably.
>
> I am not an expert in NbE so I especially enjoyed the very detailed
> and careful introduction of the idea spelled out for the simply-typed
> lambda-calculus. The paper uses categorical terms, but should be
> easily accessible without much categorical knowledge. Every new
> concept is introduced well and category theory is used more as a
> language to describe concrete definitions rather than as a foundation.
>
> NbE for CBPV and polarized lambda-calculus seem to be novel and in my
> belief deserve publication, because it's a standard technique carried
> out for (by now) standard systems in logics and programming languages
> theory. However, I feel that the work is lacking some embedding in
> related work w.r.t the calculi considered. Two or three sentences on
> standard normalization proofs and an explicit sentence that this is
> the first NbE proof for both systems (if it is) would be tremendously
> helpful to judge the value of the contribution.
>
> The treatment of CBPV is more from a logical perspective than as a
> computational system, which is left implicit in the paper. In my
> opinion, the two major aspects of CBPV are (1) the "A value is, a
> computation does"-mantra, which allows fine-grained control over
> reduction and thus enables embedding CBV and CBN and (2) the
> possibility to embed computational effects via the monad. Both aspects
> are not even mentioned, which seems strange to me. Looking at the
> equational theory and not at computation removes the "a value
> is"-aspect, but are there ideas how the NbE method could be adapted to
> also normalize w.r.t call-by-push-value reduction? Can effects be
> easily incorporated into the proof, or would that require completely
> new ideas? Considering that CBPV is in the title, I think this
> questions definitely have to be discussed in the paper.
>
> The paper comes with a partial Agda formalization. I believe the
> paper's claim that this helped in getting the renamings right, but in
> the current state I'm hesitant to count the formalization as valuable
> contribution. Most importantly, the paper needs to make precise what
> is formalized and what is not (possibly in an appendix), where it
> currently just claims a "partial formalization". And then I would hope
> for some explanation why the CBPV part is not formalized. Is it due to
> time pressure? Or would it just be more of the same? If the
> formalization is an explicit contribution I also would expect that
> related work in this area is covered. There must have been formalized
> NbE proofs for simply-typed lambda-calculus before (in Agda and in
> other systems), other formalizations of polarized lambda-calculus and
> of CBPV.
>
>
>
> Comments for the authors
>
> 12: The abstract drops several buzzwords (which I find good), but
> doesn't pick up on them later. For instance, "bi-cartesian" is only
> mentioned in the first sentence of the abstract and never again.
>
> 122: I feel like the choice why the presheaves are all defined on
> Cxt^{op} needs some explanation. The technique I am aware of to build
> categorical models of type theory is to take a category of contexts
> where morally Γ ⊆ Δ gives an arrow Δ → Γ. This confused me more than
> once, especially at points where you said "this is a presheaf"
> omitting the category

We are using *covariant* functors J : Cxt → Set from the category Cxt
of contexts and order-preserving embeddings Γ ⊆ Δ to the category Set
of sets and functions, which gives us reindexing morphisms J Γ → J Δ.
However, a presheaf is a *contravariant* functor J : Cᵒᵖ → Set by
terminology, thus, to match our use case, we have to choose C = Cxtᵒᵖ,
which means that we are dealing with presheaves over Cxtᵒᵖ.  The
double dualization (Cxtᵒᵖ)ᵒᵖ cancels to simply Cxt then.
TODO: explain in paper.

> 146: Maybe briefly say what a presheaf is to increase the
> accessibility of the paper even more?

TODO: explain in paper.

> 178: I think it should be stressed even more here that you are talking
> about normal forms up to an equational theory and not up to
> reduction. That's especially important for CBPV, where the
> normal-forms w.r.t. the equational theory completely ignore the
> reduction order induced by call-by-push-value.

Indeed, following the "reduction-free normalization" philosophy of NbE
[3], we are just looking for a canonical representative of the
equivalence classes on terms induced by the equational theory.
However, our normal forms are already fully characterized *logically*
via the bidirectional grammar for Ne/Nf, which emerges from the
introduction/elimination structure and the polarization of the
formulas.  The equational theory matching these normal forms can also
be conceived a *posteriori*, which is the approach we are taking in
this paper.

> 316: I agree that U and F are strange if you do not have lots of
> experience with CBPV. On the other hand, changing the names makes it
> hard for readers which do have experience with CBPV and makes it hard
> for inexperienced readers to match definitions with definitions in
> other papers.

You have a point here, but in any case, we want a common notation for
CBPV and polarized lambda calculus, and a bit longer names (such as
Comp/Thunk) provide more orientation than F/U and the up/down arrows
(which seem very arbitrary; I always confuse them).

> 319: I think talking about CBPV as a fixed definition (as defined in
> the initial paper) rather than a family of definitions is strange. The
> initial CBPV definition is one member of this family, but your
> finitary presentation is also a member. The contrast "CBPV epmloys
> ... while we have ..." is unnecessarily sharp. For instance, there are
> lots of presentation which where the language where a language with
> finitary sums and products is called CBPV, e.g. [1,2].

You are right.  TODO: fix in paper.

> 326: Normally U (F A) is morally the same as A, up to effects. A
> sentence explaining why this is the case or why you deviate here might
> be nice.

This is indeed the case, U.F amounts to the monad, which represents
the effects in our case.

> 361: At least mentioning the standard set-theoretic semantics for CBPV
> with algebras would be good here. There, computations and values map
> to semantically different entities. Do I understand correctly that you
> can use the same entity for both values and computations because
> you're looking at effect-free CBPV?

The difference between positive and negative product does not show up
in the type interpretation but, for instance, in reflection, where we
use `split` in the positive case and the projections in the negative
case.

> 372: Again, I feel some motivation is needed why this are the normal
> forms of CBPV up to the equational theory. When thinking about
> computation, values are always normal because they don't compute, and
> the whole idea of a thunk is that it is a normal form with no
> restrictions.

It is important to remember that we are dealing with open terms and we
are taking eta-equality into account.  Thus, in contrast to the
picture arising from computation (where we only deal with closed terms
and have no eta reduction/expansion), a variable is not always a
normal form, but only if it is of base type or negative type.

> 513: I'm a bit unhappy with the word "implemented" here. Sure, you
> have implemented on paper, but implemented for me means at least
> written down in some formal system/programming language (which is Agda
> here), which is missing for CBPV

Ok, but given the formality of our definitions, the Agda
implementation is a simple exercise.

> 519: What's hard about the agreement proof? Or do you just leave it
> out because you think it's out of scope for the paper?

We are already over the page limit...

> Minor comments
>
> 119: I would maybe write "x : (A ∈ Γ)" to make clear that "x : A ∈ Γ"
> is not a definition here, but you really mean an inhabitant x of a
> type. In named presentations it might be a short hand for "Γ x = A"
>
> 122: Is Cxt a typo? If not, a sentence why Ctx is not used might make
> sense. Is it to avoid confusion with the usual presentation of Ctx
> with arrows inverse to Cxt?
>
> 146: I would either say "all presheaves will be defined on this and
> this category" or always say which category the presheaf is defined
> on. This makes it way easier for readers to quickly calculate in the
> head that this is really a presheaf.
>
> 164: I feel "Needless to say" is unnecessarily snappy to readers with
> only very basic category theory knowledge. As mentioned, your paper is
> doing a really good job in making the presentation accessible, it's a
> good thing that it spells out the details, so I don't think that's
> needless to say at all.
>
> 178: "an equational theory"
>
> 201: The symbol for the Kripke function space looks like a hybrid
> between an arrow with a very strange, asymmetric tip and an arrow with
> a hat. I think the latter is intended, so maybe move the hat a little
> bit higher?
>
> 211: "such an extended context"
>
> 214: It should be "Ne A Δ"
>
> 217: I was confused by the 0 in the index for quite a while. This is a
> singleton context, isn't it? Then maybe write ε.0 or mention that
> singleton contexts look just like types
>
> 233: I feel repeating the category you build the monad on would
> help. It's on the category of presheaves over Cxt^{op}, right?
>
> 246: There are Γs missing twice on the reifications
>
> 253: the map should be indexed by Δ
>
> 261: again, the map needs a Δ
>
> 270: the Δ is the point your defining on pointwise, but that's too
> implicit currently in my opinion
>
> 315: Two sentences why products appear both negatively and positively
> might be nice for readers not familiar with polarized systems
>
> 319: "is not essential"
>
>
> [1] Rizkallah, Christine, Dmitri Garbuzov, and Steve Zdancewic.
> "A Formal Equational Theory for Call-By-Push-Value."
> International Conference on Interactive Theorem Proving. Springer, Cham, 2018.
>
> [2] Forster, Yannick, et al.
> "Call-by-push-value in Coq: operational, equational, and denotational theory."
> Proceedings of the 8th ACM SIGPLAN International Conference on Certified Programs and Proofs. ACM, 2019.
>
> ----------------------- REVIEW 3 ---------------------
> PAPER: 50
> TITLE: Normalization by Evaluation for Call-by-Push-Value and Polarized Lambda-Calculus
> AUTHORS: Andreas Abel and Christian Sattler
>
> Overall evaluation: 1 (weak accept)
>
> ----------- Overall evaluation -----------
>
> Normalization by Evaluation (NbE) is the (by now well-understood)
> beautiful idea that we can normalize terms in a typed lambda-calculus
> through mutually recursive functions for evaluation and reification,
> partially based on types.  NbE is one way to relate programming
> languages (where we don't evaluate under lambda-abstractions) and pure
> proof theory (where we fully normalize).
>
> Sums however present a difficulty.  In Section 2 the present paper
> adapts a prior idea on using monads to update the semantic domain
> needed for NbE to capture sums.  I find this idea compelling and
> NbE factors very cleanly for a lambda-calculus with products and
> sums.
>
> This, however, is characterized as "wasteful" (although I did not
> understand what the authors meant by this), so Section 3 polarizes
> the types and restructures NbE so that the use of the monad is
> confined to the transition from positive to negative types.  This
> seems like a natural idea, clarifying the approach of Section 2.
> The simplicity of the definitions at the end of this section is
> once again compelling, although some complexity is hidden in the
> definition of the semantic functions.
>
> Again, the authors express some dissatisfaction I couldn't quite
> follow and move on to a polarized and focused presentation.  This
> step makes perfect sense to me, although it was difficult for me
> to extract additional insight from this section.
>
> I am not familiar enough with the prior work to judge the degree
> of originality confidently, but the idea of using a monad to
> handle sums in NbE in this manner was new to me, as was the
> extension to the polarized language.  The results have been
> (mostly?) formalized, so I don't have any reason to doubt their
> correctness.  NbE is certainly interesting and I find the results
> to be significant in its further study, even if the open problems
> mentioned in the conclusion are left to future work.
>
> The presentation is idiosyncratic in multiple ways, which one may
> view as either charming or annoying.  Ultimately, I care if ideas
> are communicated well so they can be easily understood, and for me
> the paper receives mixed grades on that score.  I am going into
> some detail on issues of the presentation below.  Overall, I like
> the subject, the paper itself, and I believe I understood the main
> ideas, but I think the presentation can be improved to make it
> more understandable with less background.
>
> Presentation issues:
>
> Intrinsic typing.  Certainly an acceptable choice and seems to
> work well in the formalization.
>
> No syntax.  That presented a significant obstacle to me.  Why not
> use familiar syntax the the lambda-calculus instead of just the
> names of the rules?  Could be some abstract syntax, but even that
> would have helped.

The rules (e.g. for normal forms) are a precise (dependently-typed)
grammar for the abstract syntax (e.g. of normal forms).  The rules are
a pretty-printing of the actual Agda data types that make up the
various (intrinsically typed) syntactic classes.

Note also that restriction to abstract syntax allows for a more
factored presentation of syntax (e.g. Cov in Section 3) and syntax
constructions which are not easily rendered in concrete (linearized)
syntax (such as the nesting of <<P>> in Section 4).

> Presheaves.  In the early parts of the paper there is a lot of
> jargon regarding presheaves, which is, frankly, not necessary to
> understand the paper.

The jargon is indeed not necessary, but it connects to the categorical
notions.  The paper is intended as a bridge between conventional and
categorical meta-theory of lambda-calculus, and we want to be
accessible for both camps.

> No theorems.  The paper presents no formal definitions or theorems
> but simply develops a narrative.  I guess the existence of certain
> dependently typed functions takes the pace of any formal theorems,
> but for me it would still have been helpful to see standard
> theorems regarding the properties of various constructions.  It
> might also help to put some of the definitions into figures so
> they can be referenced in the text (even if the authors want to
> avoid explicit Definition 1, Definition 2, ...).

"No formal definitions" is probably not what was meant here, since, in
fact, *all* definitions in the paper are fully formal (with a few
definitions omitted).
Indeed, as you say, dependently typed total functions second as proofs
of the theorems their types express.  Since we work in Martin-Löf Type
Theory, all are constructions are proof-relevant, thus, there are no
theorems in the original sense.

All constructions we give are uniquely determined by their name, thus,
there is no secondary reference structure such as Definition nnn.  To
help the reader with locating the definition, we have boxed important
definitions which do not have their own disply.

> Other feedback:
>
> The introduction talks about "oracles", but not the body of the
> paper.  I didn't understand what the authors mean here and where
> an "oracle" was supposed to be invoked.

A path in a case tree corresponds to a sequence of questions asked
to the oracle ("is this neutral a left or a right injection?")
with the respective answers.  TODO: explain in paper.

> L.133.  I found it strange to present the types with a polarity
> distinction only at the top level.  P and N are later used (more
> appropriately) in the polarized form.

Strange but convenient. ;-)

> L.136-138.  Why use a backward turnstile?

We flip Γ ⊢ A to A ⊣ Γ so that we can uniformly write the index Γ last
in all presheaves we introduce in the paper.

> L.180-181.  It should be more clear right here what you are
> "proving" in this paper and what is left to future work.  I had
> difficulty interpreting this sentence.

TODO

> L.214. Should this be "Ne A Delta"?

Indeed. FIXED

> L.226. "recapitulate".  Since you do not introduce, discuss, or
> otherwise treat this topic before, "recapitulate" seems like the
> wrong term here.

FIXED. "Recall that"

> L.239.  Actually join seems to be used in a rather essential way
> shortly, so why "uninspiring"?

FIXED formulation.

> L.247.  Not sure what you are trying to say here regarding "pasting".

MOVED to footnote.  It is just to connect to category-theoretic concepts.

> L.249.  "Positive types P are already monadic".  Don't know what
> you mean by that.

[[P]] = C ...

> Section 2.4.  Don't understand the purpose of this section.  Please
> explain.

ADDED an introductory sentence.

> L.336.  "Informally speaking, this makes all negative type monadic".
> I don't understand this sentence.

FIXED.  It means that the effects of monad C are available at any negative type.

> L.384-387.  What is "J" in these rules?  Please explain.

ADDED: J stands for an arbitrary presheaf.

> L.397-405.  I could not understand this paragraph.  For example,
> you talk about "needs to run in a monad".  What does that mean?

It means that ↑ᴾ cannot be a pure function, but must be a monadic function
(see its definition).

> L.422.  "However, the elimination of positive types is still rather
> non-deterministic."  In what sense is it non-deterministic?  What do
> you mean by "It is possible..."  In what sense?

It is non-deterministic in the sense that it is not fully determined,
because "it is possible to only partially break up a composite
positive type and..."

### 2019-04-01 fscd19/response-fscd19.txt
[ fscd19 ] First draft of main rebuttal

@@
 
-- Missing theorems (1,3)
+- Contribution
 
-- Incomplete formalization (1,2)
+As the reviewers correctly understand, we are only proving logical
+consistency in this paper, via normalization.  Our main contribution
+isn't any theorem (consistency of IPL has been proven a gazillion
+times before), but the organization of normalization via a monad which
+we specify abstractly (cover monad) and for which we give two
+implementations (case trees and continuations).
 
-- The choice of abstract syntax (3)
+Reviewer 1 voices disappointment that computational soundness is not
+demonstrated.  Indeed, we wish we could have included it.  However, we
+already struggled to fit the current content into the page limit (as
+Reviewer 1 also noticed), and adding more technical content would have
+sabotaged the goal of reaching a wide audience through an accessible
+presentation (acknowledged by Reviewer 2).  The main idea, the use of
+a monad in NbE and its placement at the polarization shift can already
+be communicated through the definition of the normalization function
+alone, and very effectively, as we believe.
 
-- Relation to category theory (2,3)
+A normalization function with intrinsic soundness is quite a bit more
+complicated, see
 
-- CBPV as logical system (2)
+  https://github.com/andreasabel/ipl/blob/master/src/NbeModel.agda
 
-No NbE for effects.
+for the case of IPL, and the added technicalities may cloud the main
+idea.  Thus, we save it for a publication venue without page limit
+such as a journal.
 
-  let x = launchNuclearMissile; 1
-  in  x + 1
+Without computational soundness, there was no need to retain the
+equational theory of IPL, thus, we relegated it to the appendix.
+Indeed, our normal forms can be understood purely logically, based on
+the distinction of introductions from eliminations and negative
+(choiceless) types from positive (choiceful) types.
 
-- Future work: relationship to CBN CBV (1,2)
+- Missing theorems (reviewers 1,3)
+
+- Incomplete formalization (reviewers 1,2)
+
+Reviewers 2 and 3 note that the Agda formalization is incomplete and
+unpolished.  The Agda formalization helped us to get the constructions
+right.  However, we like to think of this paper as an implementation
+of Nbe in *Type Theory*.  Systems like Agda disappear (where is LEGO
+now?), but type theory is likely to stay with us for a while.  Thus,
+we do not consider the Agda code as a contribution of this paper.
+Instead, we have strived for a rigorous, self-contained mathematical
+presentation.
+
+If wished, we can remove the reference to the partial Agda
+formalization from the paper.
+
+- The choice of abstract syntax (Reviewer 3)
+
+Reviewer 3 asks why we do not use the "familiar syntax [of] the
+lambda-calculus instead of just the names of the rules?  Could
+be some abstract syntax".
+
+In fact we have given an abstract syntax since the rules correspond
+1-1 to dependently typed constructors, thus a judgement with rules
+corresponds 1-1 to a gramar for elements of indexed set.  Disposing of
+concrete syntax enables us to factor syntax conveniently, for instance
+by using Cov in the definition of Nf (line 392).  In case of split₂
+(line 438) it is not even immediate what the concrete-syntax analogy
+of nesting should be.
+
+- Relation to category theory (reviewers 2,3)
+
+As reviewers 2 and 3 point out, the paper could have been presented
+without any reference to category theory.  We use category theory just
+as a source of terminology to accurately name the concepts we define.
+Further, we like to include experts on category theory in our
+readership, who could be upset about not naming the obvious.  The
+paper is intended as a bridge between conventional approaches to
+lambda-calculus and category-theoretic ones, without demanding too
+much expertise.
+
+- CBPV as logical system (Reviewer 2)
+
+In the context of focused lambda-calculi, which we study to reduce the
+degrees of liberty in normal forms (in other words, removing
+don't-care non-determinism), CBPV arises naturally as calculus with
+focusing on positive introductions but not on positive eliminations.
+By "coincidence" (or rather not by coincidence, since there are no
+coincidences in mathematics) it is the same calculus as the one found
+in the search for a lambda-calculus with effects.
+
+We are indeed only interested in CBPV as logical system.  Normalizing
+effects is beyond the scope of this paper.  In any case, since effects
+do not follow the introduction/elimination scheme, additional
+considerations would be necessary for normalizing effects.
+
+- Future work: relationship to CBN and CBV translations (reviewers 1,2)
+
+We are sorry we could not satisfy the curiousity of our reviewers at
+this point.  The idea to investigate the relationship to the CBN/CBV
+translations came late in the process of writing, and there was
+neither time nor space to address it.
 

### 2019-04-01 fscd19/response-fscd19.txt
Fixed some typos and minor issues.

@@
 > type. In named presentations it might be a short hand for "Γ x = A"
+
+WON'T FIX.
+
 >
@@
 > with arrows inverse to Cxt?
->
+
+NO COMMENT.
+
 > 146: I would either say "all presheaves will be defined on this and
@@
 > head that this is really a presheaf.
->
+
+ADDED. We now say that we only consider presheaves over Ctx^{op}.
+
 > 164: I feel "Needless to say" is unnecessarily snappy to readers with
@@
 > needless to say at all.
->
+
+FIXED.
+
 > 178: "an equational theory"
->
+
+FIXED.
+
 > 201: The symbol for the Kripke function space looks like a hybrid
@@
 > bit higher?
->
+
+FIXED.
+
 > 211: "such an extended context"
->
+
+FIXED. But was also right before.
+
 > 214: It should be "Ne A Δ"
+
+FIXED. Also A should have been A ⇒ B.
+
 >
@@
 > singleton contexts look just like types
->
+
+FIXED.
+
 > 233: I feel repeating the category you build the monad on would

### 2019-04-01 fscd19/fscd19.tex
Fixed some typos and minor issues.

@@
 
-The category $\Cxt$ allows us to consider $A{\in}\_$ as a presheaf over
-$\Cxt\op$ for any $A$, witnessed by $\treindex : \Gamma \sublist
-  \Delta \to A \in \Gamma \to A \in \Delta$, which is the morphism
-part of functor $A{\in}\_$ from $\Cxt$ to $\Set$, mapping object $\Gamma$ to the
-set $A \in \Gamma$ of the indices of $A$ in $\Gamma$.
-The associated functor laws $\treindex\,\tid\,x \equiv x$
-and $\treindex\,\tau_2\,(\treindex\,\tau_1\,x) \equiv
-\treindex\,(\tau_1 \rcomp \tau_2)\,x$ hold propositionally.
+By a presheaf, we will always mean a presheaf over the category $\Cxt\op$, i.e.,
+a functor from $\Cxt$ to $\Set$.  For example, given any type $A$, we can
+consider $A{\in}\_$ as such a presheaf, with action on objects mapping $\Gamma$
+to the set $A \in \Gamma$ of indices of $A$ in $\Gamma$ and action on morphisms
+witnessed by $\treindex : \Gamma \sublist \Delta \to A \in \Gamma \to A \in
+\Delta$.  The associated functor laws $\treindex\,\tid\,x \equiv x$ and
+$\treindex\,\tau_2\,(\treindex\,\tau_1\,x) \equiv \treindex\,(\tau_1 \rcomp
+\tau_2)\,x$ hold propositionally.
 
@@
 
-Needless to say, $\tNe\,A$ and
-$\tNf\,A$ are presheaves, \ie, support reindexing with $\tren$ just as terms do.
+As expected, $\tNe\,A$ and $\tNf\,A$ are presheaves,
+\ie, support reindexing with $\tren$ just as terms do.
 From a normal form we can extract the term via an overloaded function
@@
 % \ie, that $\tm u \cong t$
-% with respect to a equational theory $\cong$ on terms.
+% with respect to an equational theory $\cong$ on terms.
 
@@
 The ability to introduce fresh variables into a context, and to use
-semantic objects such as $f : \den{A \To B}\Gamma$ in a such extended
+semantic objects such as $f : \den{A \To B}\Gamma$ in such an extended
 context, is the reason for utilizing presheaves instead of just sets
@@
 Note also that in the equation for $\up{A \To B}$,
-the neutral $u : \Ne A \Gamma$ is
-transported into $\Ne A \Delta$ via reindexing with
+the neutral $u : \Ne {(A \To B)} \Gamma$ is
+transported into $\Ne {(A \To B)} \Delta$ via reindexing with
 $\tau : \Gamma \sublist \Delta$, in order to be applicable to the normal form

### 2019-04-01 fscd19/response-fscd19.txt
Typos and rephrasement.

@@
 1-1 to dependently typed constructors, thus a judgement with rules
-corresponds 1-1 to a gramar for elements of indexed set.  Disposing of
+corresponds 1-1 to a grammar for elements of an indexed set.  Disposing of
 concrete syntax enables us to factor syntax conveniently, for instance
@@
 
-We are indeed only interested in CBPV as logical system.  Normalizing
-effects is beyond the scope of this paper.  In any case, since effects
-do not follow the introduction/elimination scheme, additional
-considerations would be necessary for normalizing effects.
+We are indeed only interested in CBPV as logical system.  Normalizing effects is
+beyond the scope of this paper.  In any case, since effects do not follow the
+introduction/elimination scheme, additional considerations would be necessary
+for normalizing effects and would like likely depend on the individual effects
+considered.
 

### 2019-04-01 fscd19/response-fscd19.txt
Minor changes.

@@
 
-We are using *covariant* functors J : Cxt → Set from the category Cxt
-of contexts and order-preserving embeddings Γ ⊆ Δ to the category Set
-of sets and functions, which gives us reindexing morphisms J Γ → J Δ.
-However, a presheaf is a *contravariant* functor J : Cᵒᵖ → Set by
-terminology, thus, to match our use case, we have to choose C = Cxtᵒᵖ,
-which means that we are dealing with presheaves over Cxtᵒᵖ.  The
-double dualization (Cxtᵒᵖ)ᵒᵖ cancels to simply Cxt then.
-TODO: explain in paper.
+In these techniques, one is always looking at a wide subcategory of the category of contexts and substitutions.
+An order-preserving embedding Γ ⊆ Δ gives rise to a substitution Δ → Γ in the opposite direction, that's why we have to take presheaves over Cxt^op.
+We agree that the direction of arrows in the category Cxt of contexts and order-preserving embeddings is non-standard from this perspective.
+However, it has the benefit that our presheaves are just *covariant* functor J : Cxt → Set.
+
+ADDED sentence explaining the relevance of Cxt^op.
 
@@
 
-TODO: explain in paper.
+FIXED. Made the introduction of presheaves more intuitive (thought still not
+giving a general definition).
 
@@
 > 319: "is not essential"
->
->
+
+FIXED.
+
 > [1] Rizkallah, Christine, Dmitri Garbuzov, and Steve Zdancewic.

### 2019-04-01 fscd19/fscd19.tex
Minor changes.

@@
 
-By a presheaf, we will always mean a presheaf over the category $\Cxt\op$, i.e.,
-a functor from $\Cxt$ to $\Set$.  For example, given any type $A$, we can
+By a \emph{presheaf} we will always mean a presheaf over the category $\Cxt\op$,
+i.e., a functor from $\Cxt$ to $\Set$.\footnote{With respect to the terms
+defined later, $\Cxt\op$ is to be seen as a category of contexts and those
+substitutions given by weakening.}  For example, given any type $A$, we can
 consider $A{\in}\_$ as such a presheaf, with action on objects mapping $\Gamma$
@@
 label sets $I$ while we only have finite sums $(\tyzero,+)$ and
-records $(\top,\ntimes)$.  However, this difference in not essential,
+records $(\top,\ntimes)$.  However, this difference is not essential,
 our treatment extends directly to the infinite case, since we are

### 2019-04-01 fscd19/response-fscd19.txt
[ fscd19 ] response final version

@@
 preparing the revision.
+
+In the following, we clarify the main questions of the reviewers:
 
@@
 right.  However, we like to think of this paper as an implementation
-of Nbe in *Type Theory*.  Systems like Agda disappear (where is LEGO
+of NbE in *Type Theory*.  Systems like Agda disappear (where is LEGO
 now?), but type theory is likely to stay with us for a while.  Thus,
 we do not consider the Agda code as a contribution of this paper.
-Instead, we have strived for a rigorous, self-contained mathematical
+Instead, we have striven for a rigorous, self-contained mathematical
 presentation.
@@
 1-1 to dependently typed constructors, thus a judgement with rules
-corresponds 1-1 to a grammar for elements of an indexed set.  Disposing of
-concrete syntax enables us to factor syntax conveniently, for instance
-by using Cov in the definition of Nf (line 392).  In case of split₂
-(line 438) it is not even immediate what the concrete-syntax analogy
-of nesting should be.
+corresponds 1-1 to a grammar for elements of an indexed set.
+Disposing of concrete syntax enables us to factor syntax conveniently,
+for instance by using Cov in the definition of Nf (line 392).  In case
+of split₂ (line 438) it is not even immediate what the concrete-syntax
+analogy of nesting should be.
 
@@
 
-We are indeed only interested in CBPV as logical system.  Normalizing effects is
-beyond the scope of this paper.  In any case, since effects do not follow the
-introduction/elimination scheme, additional considerations would be necessary
-for normalizing effects and would like likely depend on the individual effects
-considered.
+We are indeed only interested in CBPV as logical system.  Normalizing
+effects is beyond the scope of this paper.  In any case, since effects
+do not follow the introduction/elimination scheme, additional
+considerations would be necessary for normalizing effects and would
+like likely depend on the individual effects considered.
 
@@
 
-We are sorry we could not satisfy the curiousity of our reviewers at
+We are sorry we could not satisfy the curiosity of our reviewers at
 this point.  The idea to investigate the relationship to the CBN/CBV
@@
 neither time nor space to address it.
+
+We hope that we have addressed the major questions of the reviewers.
+Replies to many minor questions can be found in the longer version at
+
+  https://github.com/andreasabel/ipl/blob/master/fscd19/response-fscd19.txt
 
@@
 > CBPV. Finally, the authors discuss modification for focused polarized
-> logic, which is closed to CPBV.
+> logic, which is closed to CBPV.
 >
@@
 When quantifying over (τ : Γ ⊆ Δ) we implicitly quantify over Δ,
-similarily as when writing ∀ Δ ⊇ Γ. P, which is a common shorthand for
+similarly as when writing ∀ Δ ⊇ Γ. P, which is a common shorthand for
 ∀ Δ. Γ ⊆ Δ ⇒ P.
@@
 
+See 105-109 for explicit explanations of ":" and "∈".
+See also "suc zero : A ∈ Γ.A.B" in line 118.
 WON'T FIX.
 
->
 > 122: Is Cxt a typo? If not, a sentence why Ctx is not used might make
@@
 help the reader with locating the definition, we have boxed important
-definitions which do not have their own disply.
+definitions which do not have their own display.
 

### 2019-05-05 types19/types19.tex
[ TYPES 2019 ] formatted down to 2 pages (condensed bib/ackn)

@@
 \usepackage{hyperref}
+\urlstyle{sf}
 % \usepackage{agda} % colors
@@
 
+% less space between entries in bibliography
+\let\OLDthebibliography\thebibliography
+\renewcommand\thebibliography[1]{
+  \OLDthebibliography{#1}
+  \setlength{\parskip}{0pt}
+  \setlength{\itemsep}{0.2ex plus 0.3ex}
+}
+
 %% Macros
@@
 \authorrunning{A. Abel}
-\author{Andreas Abel}
+\author{Andreas Abel\thanks{Supported by VR grant 621-2014-4864 and EU
+  Cost Action CA15123.}}
 \institute{
@@
 monad $\C$
-\citep{filinski:semaccounttdpe,filinski:tlca01,barral:PhD}
+\citep{%filinski:semaccounttdpe,
+filinski:tlca01,barral:PhD}
 in the category of presheaves
@@
 analogy to polarized lambda-calculus
-\citep{zeilberger:PhD,brockNannestadSchuermann:lpar10,espiritoSanto:entcs17}
+\citep{zeilberger:PhD} %,brockNannestadSchuermann:lpar10,espiritoSanto:entcs17}
 refer to as \emph{positive} and \emph{negative} types.
@@
 
-\paragraph*{Acknowledgments.}
-%Thanks to the anonymous referees, who helped improving the quality of
-%this abstract through their feedback.
-This work was supported by
-Vetenskapsr\aa{}det under Grant
-  No.~621-2014-4864 \emph{Termination Certificates
-  for Dependently-Typed Programs and Proofs via Refinement Types}
-and by the EU Cost Action CA15123 \emph{Types for programming and verification}.
+% \paragraph*{Acknowledgments.}
+% %Thanks to the anonymous referees, who helped improving the quality of
+% %this abstract through their feedback.
+% This work was supported by
+% Vetenskapsr\aa{}det under Grant
+%   No.~621-2014-4864 \emph{Termination Certificates
+%   for Dependently-Typed Programs and Proofs via Refinement Types}
+% and by the EU Cost Action CA15123 \emph{Types for programming and verification}.
 
@@
 \bibliographystyle{abbrvnat}
+%\bibliographystyle{abbrvurl}
+%\bibliographystyle{plainurl} % no author-year cit.
 \bibliography{auto-types19}

### 2019-05-06 ppdp19/ppdp19.tex
[ ppdp19 ] started reformatting to 2-column style

--- 2019-04-01 fscd19/fscd19.tex
@@
-\documentclass[a4paper,USenglish,cleveref, autoref]{lipics-v2019}
-%This is a template for producing LIPIcs articles.
-%See lipics-manual.pdf for further information.
-%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
-%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
-%for section-numbered lemmas etc., use "numberwithinsect"
-%for enabling cleveref support, use "cleveref"
-%for enabling cleveref support, use "autoref"
-
+%
+% The first command in your LaTeX source must be the \documentclass command.
+\documentclass[sigplan,screen]{acmart}
+
+\newcommand\hmmax{0}
+\newcommand\bmmax{0}
 \usepackage{bm} % bold math
-% \usepackage{dsfont} % black-board bold numbers, package doublestroke
-% LACKS: \mathds{0}
-
-% Disallowed packages:
-% - enumitem
-% - paralist
-% - natbib
-% - wrapfig
-
-% Disallowed:
-% - conditional inclusion of content
-% - splitting over several files
-
-% LIPICS enumerations
-% - itemize
-% - enumerate 1.
-% - alphaenumerate (a)
-% - romanenumerate (i)
-% - bracketenumerate (1)
-
-% LIPICS environments
-% - lstlisting
-% - theorem
-% - lemma
-% - corollary
-% - proposition
-% - exercise
-% - definition
-% - example
-% - note and note*
-% - remark and remark*
-% - claim and claim*
-% - proof and claimproof
-
-% Features
-% - env: claim and claimproof inside proof
-
-% References to lemmata
-% - \cref and \autoref (includes name, like Corollary 3)
-
-%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
-
+
+\usepackage{cleveref}
+
+\hypersetup{draft}
+
+\newtheorem{remark}{Remark}
 \input{macros}
 
-\bibliographystyle{plainurl}% the mandatory bibstyle
-
-\title{Normalization by Evaluation for Call-by-Push-Value and Polarized Lambda-Calculus}
-
-\titlerunning{NbE for CBPV and Focusing}%optional, please use if title is longer than one line
-
+% %
+% % \BibTeX command to typeset BibTeX logo in the docs
+% \AtBeginDocument{%
+%   \providecommand\BibTeX{{%
+%     \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}
+
+% Rights management information.
+% This information is sent to you when you complete the rights form.
+% These commands have SAMPLE values in them; it is your responsibility as an author to replace
+% the commands and values with those provided to you when you complete the rights form.
+%
+% These commands are for a PROCEEDINGS abstract or paper.
+\copyrightyear{2019}
+\acmYear{2019}
+\setcopyright{acmlicensed}
+\acmConference[PPDP 2019]{PPDP 2019: 21st International Symposium on Principles
+  and Practice of Declarative Programming}{7-9 October 2019}{Porto, Portugal}
+\acmBooktitle{PPDP 2019: 21st International Symposium on Principles
+  and Practice of Declarative Programming, 7-9 October 2019, Porto, Portugal}
+\acmPrice{0.00}
+\acmDOI{10.1145/1122445.1122456}
+\acmISBN{978-1-4503-9999-9/18/06}
+
+%
+% These commands are for a JOURNAL article.
+%\setcopyright{acmcopyright}
+%\acmJournal{TOG}
+%\acmYear{2018}\acmVolume{37}\acmNumber{4}\acmArticle{111}\acmMonth{8}
+%\acmDOI{10.1145/1122445.1122456}
+
+%
+% Submission ID.
+% Use this when submitting an article to a sponsored event. You'll receive a unique submission ID from the organizers
+% of the event, and this ID should be used as the parameter to this command.
+%\acmSubmissionID{123-A56-BU3}
+
+%
+% The majority of ACM publications use numbered citations and references. If you are preparing content for an event
+% sponsored by ACM SIGGRAPH, you must use the "author year" style of citations and references. Uncommenting
+% the next command will enable that style.
+\citestyle{acmauthoryear}
+
+%
+% end of the preamble, start of the body of the document source.
+\begin{document}
+
+%
+% The "title" command has an optional parameter, allowing the author to define a "short title" to be used in page headers.
+\title[NbE for CBPV and Polarized Lambda Calculus]{Normalization by Evaluation for Call-By-Push-Value and
+  Polarized Lambda Calculus}
+
+%
+% The "author" command and its associated commands are used to define the authors and their affiliations.
+% Of note is the shared affiliation of the first two authors, and the "authornote" and "authornotemark" commands
+% used to denote shared contribution to the research.
 \author{Andreas Abel}
-  {Department of Computer Science and Engineering,
-    Chalmers and Gothenburg University,
-    Sweden \and
-    \url{www.cse.chalmers.se/~abela}
+%\authornote{Both authors contributed equally to this research.}
+\email{andreas.abel@gu.se}
+\orcid{0000-0003-0420-4492}
+%\authornotemark[1]
+\affiliation{%
+  \institution{Department of Computer Science and Engineering, Chalmers and Gothenburg University}
+  \streetaddress{Rännvägen 6b}
+  \city{Göteborg}
+  \state{Sweden}
+  \postcode{41296}
 }
-  {andreas.abel@gu.se}
-  {https://orcid.org/0000-0003-0420-4492}
-  {VR Grant 2014-04864 \emph{Termination
-      Certificates for Dependently-Typed Programs and Proofs via
-      Refinement Types}
+
+\author{Christian Sattler}
+\affiliation{%
+  \institution{Nottingham University}
+  \streetaddress{}
+  \city{Nottingham}
+  \country{UK}
 }
-% TODO mandatory, please use full name;
-% only 1 author per \author macro;
-% first two parameters are mandatory, other parameters can be empty.
-% Please provide at least the name of the affiliation and the country.
-% The full address is optional
-
-% \author{Frederik Folkmar Ramcke}
-%   {Department of Computer Science and Engineering,
-%     Chalmers and Gothenburg University,
-%     Sweden
-%   }
-%   {}%email
-%   {}%orcid
-%   {}%funding
-
-\author{Christian Sattler}
-  {Department of Computer Science and Engineering,
-    Chalmers and Gothenburg University,
-    Sweden
-  }
-  {}%email
-  {}%orcid
-  {}%funding
-
-\authorrunning{A. Abel %and F. F. Ramcke
-and C. Sattler}%TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
-
-\Copyright{Andreas Abel %and Frederik F. Ramcke
- and Christian Sattler}%TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
-
-% \begin{CCSXML}
-% <ccs2012>
-% <concept>
-% <concept_id>10003752.10003790.10011740</concept_id>
-% <concept_desc>Theory of computation~Type theory</concept_desc>
-% <concept_significance>300</concept_significance>
-% </concept>
-% <concept>
-% <concept_id>10003752.10010124.10010125.10010127</concept_id>
-% <concept_desc>Theory of computation~Functional constructs</concept_desc>
-% <concept_significance>300</concept_significance>
-% </concept>
-% <concept>
-% <concept_id>10003752.10010124.10010125.10010130</concept_id>
-% <concept_desc>Theory of computation~Type structures</concept_desc>
-% <concept_significance>300</concept_significance>
-% </concept>
-% <concept>
-% <concept_id>10003752.10010124.10010131.10010137</concept_id>
-% <concept_desc>Theory of computation~Categorical semantics</concept_desc>
-% <concept_significance>300</concept_significance>
-% </concept>
-% <concept>
-% <concept_id>10003752.10003790.10003792</concept_id>
-% <concept_desc>Theory of computation~Proof theory</concept_desc>
-% <concept_significance>100</concept_significance>
-% </concept>
-% </ccs2012>
-% \end{CCSXML}
+\email{christian.sattler@gmail.com}
+
+% By default, the full list of authors will be used in the page headers. Often, this list is too long, and will overlap
+% other information printed in the page headers. This command allows the author to define a more concise list
+% of authors' names for this purpose.
+%\renewcommand{\shortauthors}{A. Abel and C. Sattler}
+
+\begin{abstract}
+\input{abstract.txt}
+\end{abstract}
+
+%
+% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
+% Please copy and paste the code instead of the example below.
+%
+
+\begin{CCSXML}
+<ccs2012>
+<concept>
+<concept_id>10003752.10003790.10011740</concept_id>
+<concept_desc>Theory of computation~Type theory</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+<concept>
+<concept_id>10003752.10010124.10010125.10010127</concept_id>
+<concept_desc>Theory of computation~Functional constructs</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+<concept>
+<concept_id>10003752.10010124.10010125.10010130</concept_id>
+<concept_desc>Theory of computation~Type structures</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+<concept>
+<concept_id>10003752.10010124.10010131.10010137</concept_id>
+<concept_desc>Theory of computation~Categorical semantics</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+<concept>
+<concept_id>10003752.10003790.10003792</concept_id>
+<concept_desc>Theory of computation~Proof theory</concept_desc>
+<concept_significance>100</concept_significance>
+</concept>
+</ccs2012>
+\end{CCSXML}
 
@@
 \ccsdesc[100]{Theory of computation~Operational semantics}
-%TODO mandatory: Please choose ACM 2012 classifications from
-%https://dl.acm.org/ccs/ccs_flat.cfm
-
-% ACM old
-% F.3.2 Semantics of Programming Languages (D.3.1)
-% F.4.1 Mathematical Logic (F.1.1, I.2.2, I.2.3, I.2.4)
-
-% MSC2010
-% 03F05 Cut-elimination and normal-form theorems
-
+
+% Keywords. The author(s) should pick words that accurately describe the work being
+% presented. Separate the keywords with commas.
 \keywords{Evaluation, Intuitionistic Propositional Logic,
   Lambda-Calculus, Monad, Normalization,
-  Polarized Logic, Semantics}%TODO mandatory; please add comma-separated list of keywords
-
-\category{}%optional, e.g. invited paper
-
-\relatedversion{}%optional, e.g. full version hosted on arXiv, HAL, or other respository/website
-%\relatedversion{A full version of the paper is available at \url{...}.}
-
-\supplement{\url{https://andreasabel.github.io/ipl/html/NfModelMonad.html}}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
-
-%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors.
-Please enter author specific funding statements as fifth argument of the \author macro.
-
-\acknowledgements{The first author took inspiration from some
-  unpublished notes by Thorsten Altenkirch titled
-\emph{Another topological completeness proof for intuitionistic logic}
-received by email on 16th March 2000.
-Thorsten in turn credits his inspiration to an ALF proof by Thierry
-Coquand.  Thanks to Hugo Herbelin for feedback and spotting some typos.}
-% \acknowledgements{ The presentation of this NbE proof and its
-% Agda formalization benefited greatly from discussions with Christian
-% Sattler.}%optional
-
-% \nolinenumbers %uncomment to disable line numbering
-
-% \hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
-
-%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\EventEditors{Jakob Rehof and XXX}
-\EventNoEds{2}
-\EventLongTitle{Formal Structures for Computation and Deduction}
-\EventShortTitle{FSCD 2019}
-\EventAcronym{FSCD}
-\EventYear{2019}
-\EventDate{June 24--30, 2019}
-\EventLocation{Dortmund, Germany}
-\EventLogo{}
-\SeriesVolume{}
-\ArticleNo{}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{document}
+  Polarized Logic, Semantics}
+
+% % A "teaser" image appears between the author and affiliation information and the body
+% % of the document, and typically spans the page.
+% \begin{teaserfigure}
+%   \includegraphics[width=\textwidth]{acmart-master/samples/sampleteaser}
+%   \caption{Seattle Mariners at Spring Training, 2010.}
+%   %\Description{Enjoying the baseball game from the third-base seats. Ichiro Suzuki preparing to bat.}
+%   \label{fig:teaser}
+% \end{teaserfigure}
 
 \maketitle
-
-\begin{abstract}
-\input{abstract.txt}
-\end{abstract}
 
@@
   \nru{\tsuc}{A \in \Gamma}{A \in \cext \Gamma B}
-\qquad
+\\[2ex]
   \nru{\tdone}{}{\cempty \sublist \cempty}
@@
   \nru{\tvar}{A \in \Gamma}{A \from \Gamma}
-\qquad
+\quad
   \nru{\tabs}{B \from \cext \Gamma A}{A \To B \from \Gamma}
-\qquad
+\quad
   \nru{\tapp}
-    {A \To B \from \Gamma \qquad A \from \Gamma}
+    {A \To B \from \Gamma \quad A \from \Gamma}
     {B \from \Gamma}
@@
   \nru{\tunit}{}{\tyone \from \Gamma}
-\qquad
+\quad
   \nru{\tpair}
-    {A_1 \from \Gamma \qquad A_2 \from \Gamma}
+    {A_1 \from \Gamma \quad A_2 \from \Gamma}
     {A_1 \times A_2 \from \Gamma}
-\qquad
+\quad
   \nru{\tprj_i}
@@
     {A_1 + A_2 \from \Gamma}
-\qquad
+\quad
   \nru{\tcase}
     {A_1 + A_2 \from \Gamma
-      \qquad B \from \cext \Gamma A_1
-      \qquad B \from \cext \Gamma A_2
+      \quad B \from \cext \Gamma A_1
+      \quad B \from \cext \Gamma A_2
     }{B \from \Gamma}
-\qquad
+\\[2ex]
   \nru{\tabort}
@@
 a term such as
-\(
+\[
   \tabs\,(\tabs\,(\tpair\,
@@
     (\tabs\,(\tapp\,\vv1\,\vv0))))
-\)
+\]
 could in concrete syntax be rendered as
@@
 \qquad
-  \nru{\tabs}{\NfC B {\cext \Gamma A}}{\NfP {A \To B} \Gamma}
-\qquad
+  \nru{\tprj_i}
+    {\NeP {A_1 \times A_2} \Gamma}
+    {\Ne {A_i} \Gamma}
+\\[2ex]
   \nru{\tapp}
@@
     {\NfP {A_1 \times A_2} \Gamma}
-\qquad
-  \nru{\tprj_i}
-    {\NeP {A_1 \times A_2} \Gamma}
-    {\Ne {A_i} \Gamma}
-\\[2ex]
+\\[2ex]
+  \nru{\tabs}{\NfC B {\cext \Gamma A}}{\NfP {A \To B} \Gamma}
+\qquad
   \nru{\tinj_i}
@@
     {\NfP {A_1 + A_2} \Gamma}
-\quad
+\\[2ex]
   \nru{\tcase}
     {\NeP {A_1 + A_2} \Gamma
-      \quad \NfC P {\cext \Gamma A_1}
-      \quad \NfC P {\cext \Gamma A_2}
+      \qquad \NfC P {\cext \Gamma A_1}
+      \qquad \NfC P {\cext \Gamma A_2}
     }{\Nf P \Gamma}
-\quad
+\\[2ex]
   \nru{\tabort}
@@
 
-\begin{remark*}
+\begin{remark}
 While the coproduct eliminations $\tcase$ and $\tabort$ are limited to
@@
   \tabort^P           & u & = & \tabort\;u \\
-\end{array}
-\qquad
-\begin{array}{l@{~}lcl}
-  \\
+% \end{array}
+% \qquad
+% \begin{array}{l@{~}lcl}
+%   \\
   \tabort^{A \times B} & u & = & \tpair\,(\tabort^A\,u)\,(\tabort^B\,u) \\
@@
 the branches.
-\end{remark*}
+\end{remark}
 
@@
 \emph{Normalization by evaluation} (NbE)
-\fbox{\(
+%\fbox{\(
 %  \norm[\Gamma] A (t : A \from \Gamma)
+\[
   \tnorm\, (t : A \from \Gamma)
   = \down A \dent t {\idenv\Gamma}
-\)}
+\]
+%\)}
 decomposes normalization into
@@
 \[
+\def\arraystretch{1.3}
 \begin{array}{lcl}
@@
   \upG   \tyone \; u        & = & \ttempty \\
-  \downG \tyone \; \ttempty & = & \tunit \\
-\end{array}
-\qquad\qquad\qquad
-\begin{array}{lcl}
+  \downG \tyone \; \ttempty & = & \tunit \\[1ex]
+% \end{array}
+% \qquad\qquad\qquad
+% \begin{array}{lcl}
   \den   {A \times B}       & = & \den A \hattimes \den B \\
   \upG   {A \times B} \; u  & = & (\upG A (\prj1 u),\;\upG B (\prj2 u)) \\
-  \downG {A \times B} (a,b) & = & \tpair\,(\downG A a)\,(\downG B b) \\
-\end{array}
-\]
-\[
-\begin{array}{lcl}
+  \downG {A \times B} (a,b) & = & \tpair\,(\downG A a)\,(\downG B b) \\[1ex]
+% \end{array}
+% \]
+% \[
+% \begin{array}{lcl}
   \den {A \To B}    & = & \den A \hatto \den B \\
-  \upG   {A \To B}    \; u \; (\tau : \Gamma \sublist \Delta) \; (a : \den A \Delta)
-     & = & \up[\Delta] B (\tapp\;(\tren\,\tau\, u)\;(\down[\Delta] A a)) \\
+  \upG   {A \To B}\; u & = & \lambda \, (\tau : \Gamma \sublist \Delta) \; a %(a : \den A \Delta)
+     .\ \up[\Delta] B (\tapp\;(\tren\,\tau\, u)\;(\down[\Delta] A a)) \\
+  % \upG   {A \To B}    \; u \; (\tau : \Gamma \sublist \Delta) \; a %(a : \den A \Delta)
+  %    & = & \up[\Delta] B (\tapp\;(\tren\,\tau\, u)\;(\down[\Delta] A a)) \\
   \downG {A \To B}    f        & = & \tabs\,(\down[\cext \Gamma A] B
@@
 \[
-\begin{array}{lcl@{\quad}l}
+\begin{array}{l@{~}c@{~}l@{\quad}l}
 \tabort^\C & : & \tNe\, \tyzero \todot \C\,\B
+\\ &
   & \mbox{case on absurd neutral} \\
 \tcase^\C_\Gamma & : & \Ne {(A_1 + A_2)} \Gamma
-  \to \C\,\B\,(\cext \Gamma {A_1})
+\\
+  &\to& \C\,\B\,(\cext \Gamma {A_1})
   \to \C\,\B\,(\cext \Gamma {A_2})
   \to \C\,\B\,\Gamma
+\\ &
   & \mbox{case on neutral} \\
 \trunNf[\C] & : & \C\,(\tNf A) \todot \tNf A
+\\ &
   & \mbox{run the monad ($\tNf$ only)} \\
@@
   \up    \tybase & = & \treturn[\C] \\
-  \down  \tybase & = & \trunNf[\C] \comp \tmap[\C]\tne \\
-\end{array}
-\qquad
-\begin{array}{lcl}
-  \den   \tyzero & = & \C\,\hat0 \\
-  \up    \tyzero & = & \tabort^\C \\
-  \down  \tyzero & = & \trunNf[\C] \comp \tmap[\C]\tmagic \\
-\end{array}
-\]
-% \[
-% \begin{array}{lcl}
-%   \den \tybase & = & \C\,(\tNe\,\tybase) \\
-%   \upG \tybase \; u & = & \treturn[\C]\,u \\
-%   \downG \tybase \; m & = & \trunNf[\C]\,(\tmap[\C]\tne\;m) \\
+  \down  \tybase & = & \trunNf[\C] \comp \tmap[\C]\tne \\[1ex]
 % \end{array}
@@
 % \begin{array}{lcl}
-%   \den \tyzero & = & \C\,\hat0 \\
-%   \upG \tyzero \; u & = & \tabort^\C\,u \\
-%   \downG \tyzero \; m & = & \trunNf[\C]\,(\tmap[\C]\tmagic\;m) \\
+  \den   \tyzero & = & \C\,\hat0 \\
+  \up    \tyzero & = & \tabort^\C \\
+  \down  \tyzero & = & \trunNf[\C] \comp \tmap[\C]\tmagic \\[1ex]
 % \end{array}
 % \]
-\[
-\begin{array}{lcl}
+% \[
+% \begin{array}{lcl}
   \den {A+B} & = & \C\,(\den A \hatplus \den B) \\
   \upG {A+B} \; u & = & \tcase^\C\,u
-    \,(\treturn[\C]\,(\iota_1\,\tfresh^A_\Gamma))
+    \,(\treturn[\C]\,(\iota_1\,\tfresh^A_\Gamma)) \\ & & \qquad\quad
     \,(\treturn[\C]\,(\iota_2\,\tfresh^B_\Gamma))
     \\
-  \downG {A+B} & = & \trunNf[\C] \comp
+  \down {A+B} & = & \trunNf[\C] \comp
     \tmap[\C] [\tinj_1 \comp \down A,\; \tinj_2 \comp \down B] \\
@@
 \[
-\begin{array}{l@{~}l@{~~}c@{~~}l@{\qquad}l@{~~}c@{~~}l}
-  \multicolumn 7 l {\trun^A ~:~ \C \den A \todot \den A} \\
+\begin{array}{ll@{~~}cl}%{l@{~}l@{~~}c@{~~}l}
+  \multicolumn 4 l {\trun^A ~:~ \C \den A \todot \den A} \\
   \trun^{\tyone} & c & = & \ttempty
-    &
-  \trun^{\tyzero} & = & \tjoin[\C]
     \\
@@
        \trun^B\,(\tmap[\C]\;\pi_2\;c))
-    &
-  \trun^{A + B}   & = & \tjoin[\C]
   \\
-  \trun^{A \To B} & c \; \tau \; a & = & \trun^B\,
-      (\stmap[\C]\;(\lambda\;\tau'\,f.\;f\;\tid\;(\tren\;\tau'\;a))\;(\tren\;\tau\;c))
-    &
-  \trun^{o}      & = & \tjoin[\C]
+  \trun^{A \To B} & c  & = & \lambda \,\tau \; a. \ \trun^B\,
+      (\stmap[\C]\;(\lambda\;\tau'\,f.\;f\;\tid\;(\tren\;\tau'\;a))
+  \\ &&& \qquad \qquad (\tren\;\tau\;c))
   \\
-\end{array}
-% \begin{array}{lcl@{\quad}l@{~}lcl}
-%   \trun^A & : & \C \den A \todot \den A \\
-%   \trun^{\tyzero} & = & \tjoin[\C]
-%     & \trun^{\tyone} & c & = & \ttempty \\
-%   \trun^{A + B}   & = & \tjoin[\C]
-%     & \trun^{A \times B} & c & = &
-%       (\trun^A\,(\tmap[\C]\;\pi_1\;c),\
-%        \trun^B\,(\tmap[\C]\;\pi_2\;c))\\
-%   \trun^{o}      & = & \tjoin[\C]
-%     & \trun^{A \To B} & c \; \tau \; a & = & \trun^B\,
-%       (\stmap[\C]\;(\lambda\;\tau'\,f.\;f\;\tid\;(\tren\;\tau'\;a))\;(\tren\;\tau\;c))
-% \\
-% \end{array}
+  \trun^{o}      & & = & \tjoin[\C]
+  \\
+  \trun^{\tyzero} & & = & \tjoin[\C]
+  \\
+  \trun^{A + B}   & & = & \tjoin[\C]
+  \\
+\end{array}
 \]
@@
 \begin{array}{lcl}
+  \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
+  \dent{\tabs\,t}\gamma & = & \fdent t \gamma \\
+  \dent{\tapp\,t\,u} \gamma & = &
+     \dent t \gamma\;\tid\;\dent u \gamma \\
   \dent{\tunit}\gamma & = & \ttempty \\
@@
     (\dent{t_1}\gamma,\ \dent{t_2}\gamma) \\
-  \dent{\tabs\,t}\gamma & = & \fdent t \gamma \\
+  \dent{\prj i t}\gamma & = & \pi_i \dent t \gamma \\
   \dent{\inj i t}\gamma & = & \iota_i \dent t \gamma \\
-  \\
-\end{array}
-\quad
-\begin{array}{lcl}
-  \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
-  \dent{\prj i t}\gamma & = & \pi_i \dent t \gamma \\
-  \dent{\tapp\,t\,u} \gamma & = &
-     \dent t \gamma\;\tid\;\dent u \gamma \\
   \dent{\tcase\;u\;t_1\;t_2}\gamma & = &
@@
 \end{array}
-% \begin{array}{lcl}
-%   \\
-%   \dent{\tabs\,t}\gamma & = & \fdent t \gamma \\
-%   \dent{\tunit}\gamma & = & \ttempty \\
-%   \dent{\tpair\,t_1\,t_2}\gamma & = &
-%     (\dent{t_1}\gamma,\ \dent{t_2}\gamma) \\
-%   \dent{\inj i t}\gamma & = & \iota_i \dent t \gamma \\
-%   \\
-% \end{array}
-% \begin{array}{lcl}
-%   \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
-%   \dent{\tapp\,t\,u} \gamma & = &
-%      \dent t \gamma\;\tid\;\dent u \gamma \\
-%   \\
-%   \dent{\prj i t}\gamma & = & \pi_i \dent t \gamma \\
-%   \dent{\tcase\;u\;t_1\;t_2}\gamma & = &
-%      \dencase\;\dent u \gamma\;\fdent{t_1}\gamma\;\fdent{t_2}\gamma \\
-%   \dent{\tabort\,u}\gamma & = & \denabort\;\dent u \gamma \\
-% \end{array}
-% \begin{array}{lcl}
-%   \\
-%   \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
-%   \dent{\tabort\,u}\gamma & = & \denabort\;\dent u \gamma \\
-%   \dent{\tcase\;u\;t_1\;t_2}\gamma & = &
-%      \dencase\;\dent u \gamma\;\fdent{t_1}\gamma\;\fdent{t_2}\gamma \\
-%   \dent{\inj i t}\gamma & = & \iota_i \dent t \gamma \\
-% \end{array}
-% \begin{array}{lcl}
-%   \dent{\tabs\,t}\gamma & = & \fdent t \gamma \\
-%   \dent{\tapp\,t\,u} \gamma & = &
-%      \dent t \gamma\;\tid\;\dent u \gamma \\
-%   \dent{\tunit}\gamma & = & \ttempty \\
-%   \dent{\tpair\,t_1\,t_2}\gamma & = &
-%     (\dent{t_1}\gamma,\ \dent{t_2}\gamma) \\
-%   \dent{\prj i t}\gamma & = & \pi_i \dent t \gamma \\
-% \end{array}
-\]
-% \[
-% \begin{array}{lcl}
-%   % \fdent{t : B \from \cext \Gamma A}\gamma & = &
-%   %   \lambda
-%   %     \,(\tau : \Delta \sublist \Phi)
-%   %     \,(a : \den A \Phi)
-%   %     .\
-%   %     \dent t {(\tren\,\tau\,\gamma,\ a)} \\[1ex]
-%   \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
-%   \dent{\tabs\,t}\gamma
-%       & = & \fdent t \gamma \\
-%   % \dent{\tabs\,(t : B \from \cext \Gamma A)}\gamma \;
-%   %    (\tau : \Delta \sublist \Phi)\, (a : \den A \Phi) & = &
-%   %       \dent t {(\tren\,\tau\,\gamma,\ a)} \\
-%   % \dent{\tabs\,(t : B \from \cext \Gamma A)}\gamma & = &
-%   %    \lambda\, (\tau : \Delta \sublist \Phi)\, (a : \den A \Phi). \
-%   %       \dent t {(\tren\,\tau\,\gamma,\ a)} \\
-%   \dent{\tapp\,t\,u} \gamma & = &
-%      \dent t \gamma\;\tid\;\dent u \gamma \\
-%   % \dent{\tapp\,(t : A \To B \from \Gamma)\,u} \gamma & = &
-%   %    \dent t \gamma\;\tid\;\dent u \gamma \\
-%   \dent{\tunit}\gamma & = & \ttempty \\
-%   \dent{\tpair\,t_1\,t_2}\gamma & = &
-%     (\dent{t_1}\gamma,\ \dent{t_2}\gamma) \\
-%   \dent{\prj i t}\gamma & = & \pi_i \dent t \gamma \\
-%   \dent{\inj i t}\gamma & = & \iota_i \dent t \gamma \\
-%   \dent{\tabort\,u}\gamma & = & \denabort\;\dent u \gamma \\
-%   \dent{\tcase\;u\;t_1\;t_2}\gamma & = &
-%      \dencase\;\dent u \gamma\;\fdent{t_1}\gamma\;\fdent{t_2}\gamma \\
-%   % \dent{\tcase\,(u : A_1 + A_2 \from \Gamma)\;t_1\;t_2}\gamma & = &
-%     %  \dencase\;\dent u \gamma\;f_1\;f_2
-%     % \ \bwhere \\
-%     % \multicolumn 1 r {f_i\; (\tau : \Delta \sublist \Phi)\,(a_i : \den {A_i} \Phi)}
-%     %     & = &
-%     %     \dent {t_i} {(\tren\;\tau\;\gamma,\ {a_i})}
-%    % \
-%    %  (\lambda\; (\tau : \Delta \sublist \Phi)\,(b : \den A \Phi).\
-%    %      \dent v {(\tren\;\tau\;\gamma,\ b)})
-%     \\
-% \end{array}
-% \]
+\]
 For the interpretation of the binders $\tabs$ and $\tcase$ we use the
@@
 \[
-\begin{array}{lcl}
+\begin{array}{l}
   \fden{t : B \from \cext \Gamma A}
-    & : & \den \Gamma \todot \den{A \To B} \\
-    & = &
+    : \den \Gamma \todot \den{A \To B} \\
+    \quad =
       \lambda\, (\gamma : \den \Gamma \Delta)
@@
 \[
-\begin{array}{lcl}
+\begin{array}{l@{~}c@{~}l}
   \denabort^B & : & \den \tyzero \todot \den B \\
@@
     \todot \den{B} \\
-  \dencase^B\;c\;f_1\;f_2 & = &
+  \dencase^B & = & \lambda\,c\;f_1\;f_2.\
     \trun^B (\stmap[\C]\;(\lambda\,\tau.\ [f_1\,\tau,\ f_2\,\tau])\;c) \\
@@
   \tcase^{\CC}
-     \; (u : \NeP {A_1 + A_2} \Gamma)
+     \; (u : \NeP {A_1 + A_2} \Gamma) \\ \qquad \quad
      \; (c_1 : \CC\,\J\,(\cext \Gamma {A_1}))
@@
     \; (k : (\J \hatto \tNf\,B)\Delta)
-     ~=~
+     ~=~ \\  \qquad
     \tcase\;(\tren\,\tau\,u)\;n_1\;n_2
-    \bwhere \\
+    \\ \qquad \bwhere \\
 \qquad
@@
             : \cext \Gamma {A_i} \sublist \cext \Delta {A_i})
+\\ &&
       \; \left(\lambda
@@
 \[
-\begin{array}{lllrl@{\qquad}l}
+\begin{array}{l@{~~}l@{~~}l@{~~}r@{~~}l}%@{\qquad}l}
   \PTy & \ni & P,Q & ::= & \patom
@@
     \mid \Thunk N
-  & \mbox{positive type}
+%  & \mbox{positive type}
 \\
@@
     \mid \Comp P
-  & \mbox{negative type}
+%  & \mbox{negative type}
 \\
@@
 \end{array}
-\qquad
+%\qquad
+\]
+\[
 \begin{array}{lcl}
@@
 
-\bibliography{auto-fscd19}
+\begin{acks}
+The first author took initial inspiration from some
+  unpublished notes by Thorsten Altenkirch titled
+\emph{Another topological completeness proof for intuitionistic logic}
+received by email on 16th March 2000.
+Thorsten in turn credits his inspiration to an ALF proof by Thierry
+Coquand.  Thanks to Hugo Herbelin for feedback and spotting some typos.
+\end{acks}
+
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{auto-ppdp19}
 
 %\clearpage
+
+\end{document}
+
 \appendix
@@
 trees is sufficient.
-
-\end{document}
-
-\clearpage
-
-\subsection{Old, explicit syntax}
-
-In the following, we mutually define the inductive families of
-presheaves $\tVal\,P$,
-$\tNe\,N$, $\Cov\,\J$,
-and $\tNf\,N$.
-
-Values \fbox{$v : \PVal P \Gamma$}
-of positive type $P$ are constructed by the
-following rules:
-\begin{gather*}
-  \nru{\pvar}{P \in \Gamma}{\PVal P \Gamma}
-\qquad
-  \nru{\tthunk}{\Nf N \Gamma}{\PValP{\Thunk N}\Gamma}
-\\[2ex]
-  \nru{\punit}{}{\PVal{\tyone}\Gamma}
-\qquad
-  \nru{\ppair}
-      {\PVal{P_1}\Gamma \qquad \PVal{P_2}\Gamma}
-      {\PValP{P_1 \times P_2}\Gamma}
-\qquad
-  \nru{\tinj_i}{\PVal{P_i}\Gamma}{\PValP{P_1 + P_2}\Gamma}
-\end{gather*}
-Negative hypotheses can be eliminated by projections and application
-to values, as described by the following inductive family
-\fbox{$\Ne N \Gamma$}:
-\begin{gather*}
-  \nru{\nvar}{N \in \Gamma}{\Ne N \Gamma}
-\qquad
-  \nru{\tprj_i}
-      {\NeP {N_1 \ntimes N_2} \Gamma}
-      {\Ne {N_i} \Gamma}
-\qquad
-  \nru{\tapp}
-      {\NeP {P \To N} \Gamma \qquad \PVal P \Gamma}
-      {\Ne N \Gamma}
-\end{gather*}
-An elimination may either result in a negative atom $\natom$ or a
-positive type $P$, which can be further taken apart by pattern
-matching.  Pattern matching on neutrals is organized in the
-inductively defined strong monad \fbox{$\Cov$}.
-\begin{gather*}
-  \nru{\treturn}{\J\,\Gamma}{\Cover \J \Gamma}
-\qquad
-  \nru{\tmatch}
-      {\NeP {\Comp P} \Gamma \qquad \AddHypP P {\Cov\,\J} \Gamma}
-      {\Cover \J \Gamma}
-\end{gather*}
-Finally, normal forms of negative types are defined as inductive
-family \fbox{$\Nf N \Gamma$}.
-\begin{gather*}
-  \nru{\tne}
-      {\CoverP {\tNe\,\natom} \Gamma}
-      {\Nf {\natom} \Gamma}
-\quad
-  \nru{\tval}
-      {\CoverP {\tVal\,P}  \Gamma}
-      {\NfP {\Comp P} \Gamma}
-\quad
-  \nru{\npair}
-      {\Nf {N_1} \Gamma \qquad \Nf {N_2} \Gamma}
-      {\NfP {N_1 \ntimes N_2} \Gamma}
-\quad
-  \nru{\tabs}
-      {\AddHypP P {\tNf\, N} \Gamma}
-      {\NfP {P \To N} \Gamma}
-\end{gather*}
-All these inductive families are presheaves, due to factored
-presentation using $\Add P$ and $\Cov$ the proof is not a simple
-mutual induction.  Yet, in Agda, the generic proof goes through using a sized
-typing for these inductive families.  Similarly, defining the
-$\tjoin$ for monad $\Cov$ relies on sized typing.
-\[
-\begin{array}{lll}
-  \tjoin & : & \forall i.\
-    \Cov[i]\,(\Cov[\infty]\J) \todot \Cov[\infty]\J  \\
-  \tjoin[i+1]\,(\treturn[i]\;c) & = & c \\
-  \tjoin[i+1]\,(\tmatch^i\;t\;k)
-    & = & \tmatch^\infty\;
-     (t : \Ne P \Gamma)\;
-     \left(\stmap[\Add P]_\Gamma\;\tjoin[i]\;(k : \AddHypP P {\Cov[i]\J} \Gamma)\right) \\
-  % \tjoin^{i+1}\,(\tmatch\,(t : \Ne P \Gamma)\,
-  %   (k : \AddHypP P {\Cov^i\,\J} \Gamma))
-  %   & = & \tmatch\;t\;(\stmap[\Add P]_\Gamma\;\tjoin^i\;k) \\
-\end{array}
-\]
-Herein, we used the sized typing of the constructors of $\Cov$:
-\[
-\begin{array}{lll}
-  \treturn & : & \forall i.\ \J \todot \Cov[i+1] \J \\
-  \tmatch  & : & \forall i.\
-    \tNe\,(\Comp P) \todot \Add P\,(\Cov[i]\J) \todot \Cov[i+1]\J \\
-\end{array}
-\]
-
-\clearpage
-
-\subsection{Parametrized syntax (superseded)}
-
-Assuming a family $\T\,N\,\Gamma$ of terms of negative type $N$ in
-context $\Gamma$, values \fbox{$v : \PValT P \Gamma$}
-of positive type $P$ are constructed by the
-following rules:
-\begin{gather*}
-  \nru{\pvar}{P \in \Gamma}{\PValT P \Gamma}
-\qquad
-  \nru{\tthunk}{\T\,N\,\Gamma}{\PValTP{\Thunk N}\Gamma}
-\\[2ex]
-  \nru{\punit}{}{\PValT{\tyone}\Gamma}
-\qquad
-  \nru{\ppair}
-      {\PValT{P_1}\Gamma \qquad \PValT{P_2}\Gamma}
-      {\PValTP{P_1 \times P_2}\Gamma}
-\qquad
-  \nru{\tinj_i}{\PValT{P_i}\Gamma}{\PValTP{P_1 + P_2}\Gamma}
-\end{gather*}
-Negative hypotheses can be eliminated by projections and application
-to values, as described by the following inductive family
-\fbox{$\NeT N \Gamma$}:
-\begin{gather*}
-  \nru{\nvar}{N \in \Gamma}{\NeT N \Gamma}
-\qquad
-  \nru{\tprj_i}
-      {\NeTP {N_1 \ntimes N_2} \Gamma}
-      {\NeT {N_i} \Gamma}
-\qquad
-  \nru{\tapp}
-      {\NeTP {P \To N} \Gamma \qquad \PValT P \Gamma}
-      {\NeT N \Gamma}
-\end{gather*}
-% Negative hypotheses can be eliminated by projections and application
-% to values $v : \V\,P\,\Gamma$ which we leave abstract in the following
-% inductive definition of \fbox{$\Ne[\V] N \Gamma$}:
-% \begin{gather*}
-%   \nru{\nvar}{N \in \Gamma}{\Ne[\V] N \Gamma}
-% \qquad
-%   \nru{\tprj_i}
-%       {\NeP[\V] {N_1 \ntimes N_2} \Gamma}
-%       {\Ne[\V] {N_i} \Gamma}
-% \qquad
-%   \nru{\tapp}
-%       {\NeP[\V] {P \To N} \Gamma \qquad \V\,P\,\Gamma}
-%       {\Ne[\V] N \Gamma}
-% \end{gather*}
-
-% decomposing
-% decomposes
-% \emph{values}, meaning inhabitants of positive types.
-An elimination may either result in a negative atom $\natom$ or a
-positive type $P$, which can be further taken apart by pattern
-matching.  Pattern matching on neutrals is organized in the
-inductively defined cover monad \fbox{$\Cov[\T]$}.
-\begin{gather*}
-  \nru{\treturn}{\J\,\Gamma}{\CoverT \J \Gamma}
-\qquad
-  \nru{\tmatch}
-      {\NeTP {\Comp P} \Gamma \qquad \AddHypP P {\Cov[\T]\J} \Gamma}
-      {\CoverT \J \Gamma}
-\end{gather*}
-
-Finally, normal forms of negative types are defined as inductive
-family \fbox{$\Nf N \Gamma$}.  It relies on the previous definitions
-with instance $\T = \tNf$.
-\begin{gather*}
-  \nru{\tne}
-      {\Cov[\tNf](\tNe[\tNf]\,\natom)\,\Gamma}
-      {\Nf {\natom} \Gamma}
-\qquad
-  \nru{\tval}
-      {\Cov[\tNf](\tVNf\,P)\,\Gamma}
-      {\NfP {\Comp P} \Gamma}
-\\[2ex]
-  \nru{\npair}
-      {\Nf {N_1} \Gamma \qquad \Nf {N_2} \Gamma}
-      {\NfP {N_1 \ntimes N_2} \Gamma}
-\qquad
-  \nru{\tabs}
-      {\AddHypP P {\tNf\, N} \Gamma}
-      {\NfP {P \To N} \Gamma}
-\end{gather*}

### 2019-05-07 ppdp19/ppdp19.tex
[ ppdp19 ] Crude 2-column formatting done

@@
 % The first command in your LaTeX source must be the \documentclass command.
-\documentclass[sigplan,screen]{acmart}
+\documentclass[sigplan,screen,fleqn]{acmart}
 
@@
 \usepackage{bm} % bold math
+
+%%\usepackage{nccmath} %ceqn  %% NOT COMPATIBLE
 
@@
 % the next command will enable that style.
-\citestyle{acmauthoryear}
+% \citestyle{acmauthoryear}
 
@@
 
+\setlength{\mathindent}{\parindent}
+
 \section{Introduction}
@@
 We can define
-\fbox{$\tid : \Gamma \sublist \Gamma$} and \fbox{$\_{\rcomp}\_ : \Gamma
-\sublist \Delta \to \Delta \sublist \Phi \to \Gamma \sublist \Phi$}
+\[
+\begin{array}{lcl}
+\tid & : & \Gamma \sublist \Gamma
+\\
+\_{\rcomp}\_ & : & \Gamma
+\sublist \Delta \to \Delta \sublist \Phi \to \Gamma \sublist \Phi
+\end{array}
+\]
+% \fbox{$\tid : \Gamma \sublist \Gamma$} and \fbox{$\_{\rcomp}\_ : \Gamma
+% \sublist \Delta \to \Delta \sublist \Phi \to \Gamma \sublist \Phi$}
 by recursion, meaning that the (proof-relevant)
@@
 bit clumsily, be also abbreviated to $\tren$.}
-\fbox{$\tren : \Gamma \sublist \Delta \to A \from \Gamma \to A \from \Delta$}
+%\fbox{$
+\[
+\tren : \Gamma \sublist \Delta \to A \from \Gamma \to A \from \Delta
+\]
+%$}
 defined by recursion over $t : A \from \Gamma$,
@@
 From a normal form we can extract the term via an overloaded function
-\fbox{$\tm\_ : \Nf A \Gamma \to A \from \Gamma$} and
-\fbox{$\tm\_ : \Ne A \Gamma \to A \from \Gamma$}
+\[
+\begin{array}{lcl}
+\tm\_ & : & \Nf A \Gamma \to A \from \Gamma \\
+\tm\_ & : & \Ne A \Gamma \to A \from \Gamma \\
+\end{array}
+\]
+% \fbox{$\tm\_ : \Nf A \Gamma \to A \from \Gamma$} and
+% \fbox{$\tm\_ : \Ne A \Gamma \to A \from \Gamma$}
 that discards constructor $\tne$ but keeps all other constructors.
@@
 \tabort^\C & : & \tNe\, \tyzero \todot \C\,\B
-\\ &
-  & \mbox{case on absurd neutral} \\
+%\\ & & \mbox{case on absurd neutral}
+\\
 \tcase^\C_\Gamma & : & \Ne {(A_1 + A_2)} \Gamma
 \\
-  &\to& \C\,\B\,(\cext \Gamma {A_1})
+  && {} \to \C\,\B\,(\cext \Gamma {A_1})
   \to \C\,\B\,(\cext \Gamma {A_2})
   \to \C\,\B\,\Gamma
-\\ &
-  & \mbox{case on neutral} \\
+%\\ & & \mbox{case on neutral}
+\\
 \trunNf[\C] & : & \C\,(\tNf A) \todot \tNf A
-\\ &
-  & \mbox{run the monad ($\tNf$ only)} \\
+%\\ & & \mbox{run the monad ($\tNf$ only)}
+\\
 \end{array}
@@
       {\PValTP{P_1 \times P_2}\Gamma}
-\qquad
+\\[2ex]
   \nru{\tinj_i}{\PValT{P_i}\Gamma}{\PValTP{P_1 + P_2}\Gamma}
@@
 sum types.
+\begin{figure}[htbp]
+%\begin{ceqn}
+\hrulefill
 \begin{gather*}
@@
       {\TmP {P \To N} \Gamma}
+\\[2ex]
+  \nru{\nunit}
+      {}
+      {\Tm \top \Gamma}
 \quad
@@
       {\TmP {N_1 \ntimes N_2} \Gamma}
-\quad
-  \nru{\nunit}
-      {}
-      {\Tm \top \Gamma}
 \\[2ex]
@@
       {\Tm N \Gamma}
-\qquad
+\\[2ex]
   \nru{\tprj_i}
@@
       {\Tm N \Gamma}
-\qquad
+\\[2ex]
   \nru{\tsplit}
@@
     }{\Tm N \Gamma}
-\qquad
+\\[2ex]
   \nru{\tabort}
@@
 \end{gather*}
+%\end{ceqn}
+\hrulefill
+  \caption{Terms of CBPV}
+  \label{fig:tmcbpv}
+\end{figure}
 
@@
   \dent{\tthunk\,t}\gamma & = & \dent t \gamma \\
-\end{array}
-\quad
-\begin{array}{lcl}
+% \end{array}
+% \quad
+% \begin{array}{lcl}
   \dent{\tabs\,t}\gamma
@@
       {\PValNfP{P_1 \times P_2}\Gamma}
-\qquad
+\\[2ex]
   \nru{\tinj_i}{\PValNf{P_i}\Gamma}{\PValNfP{P_1 + P_2}\Gamma}
@@
       {\Ne {N_i} \Gamma}
-\qquad
+\\[2ex]
   \nru{\tapp}
@@
       {\Cover \J \Gamma}
-\qquad
+\\[2ex]
   \nru{\tabort}
@@
       {\NfP {N_1 \ntimes N_2} \Gamma}
-\qquad
+\\[2ex]
   \nru{\tabs}
@@
     \\
-  \upG{\Thunk N} & x & = & \treturn\,(\upG N (\tforce\,x)) \\
-\end{array}
-\,
-\begin{array}{lcl}
-  \down P & : & \den P \todot \tVNf\,P \\
-  \down {\patom} & = & \tvar \\
-  \downG {\tyone} () & = & \punit \\
-  \downG {P_1 \times P_2}(a_1,a_2) & = & \ppair
+  \upG{\Thunk N} & x & = & \treturn\,(\upG N (\tforce\,x))
+\\[2ex]
+% \end{array}
+% \,
+% \begin{array}{lcl}
+    \multicolumn 2 l {\down P} & : & \den P \todot \tVNf\,P \\
+  \down {\patom} & & = & \tvar \\
+  \downG {\tyone} & () & = & \punit \\
+  \downG {P_1 \times P_2} & (a_1,a_2) & = & \ppair
     \,(\downG {P_1} a_1)
@@
     \\
-  \down {\tyzero} & = & \tmagic \\
-  \down {P_1 + P_2} & = &
+  \down {\tyzero} & & = & \tmagic \\
+  \down {P_1 + P_2} & & = &
     [  \tinj_1 \comp \down{P_1}
@@
     ] \\
-  \down {\Thunk N} & = & \tthunk \circ \down N \\
+  \down {\Thunk N} & & = & \tthunk \circ \down N \\
 \end{array}
@@
 \[
+\renewcommand{\left}{}
+\renewcommand{\right}{}
 \def\arraystretch{1.3}
-\begin{array}{l@{\,}l@{~~}c@{~~}l}
+\begin{array}{llcl} %{l@{\,}l@{~~}c@{~~}l}
   \multicolumn 2 l {\up N} & : & \tNe\,N \todot \den N \\
@@
     \upG {N_2} \; (\prj 2 u)
-    \right) \\
-\end{array}
-\
-\begin{array}{l@{\,}l@{~~}c@{~~}l}
+    \right)
+\\[2ex]
+% \end{array}
+% \
+% \begin{array}{l@{\,}l@{~~}c@{~~}l}
   \multicolumn 2 l {\down N} & : & \den N \todot \tNf\;N \\
@@
 \def\arraystretch{1.3}
+\renewcommand{\left}{\,}
+\renewcommand{\right}{}
 \begin{array}{l@{\,}l@{~~}c@{~~}l}
   \upG {P \To N} & u & = & \lambda
-    \;(\tau : \Gamma \sublist \Delta)
-    \;(a : \den P \Delta)
+    \; \tau \; a
+    % \;(\tau : \Gamma \sublist \Delta)
+    % \;(a : \den P \Delta)
     . \
@@
         \lambda
-          \;(\tau : (\cext \Gamma P) \sublist \Delta)
-          \;(a : \den P \Delta)
+          \; \tau \; a
+          % \;(\tau : (\cext \Gamma P) \sublist \Delta)
+          % \;(a : \den P \Delta)
           . \
@@
         \right)
-      \,\tfresh^P_\Gamma
+      \;\tfresh^P_\Gamma
     \right) \right)
@@
 \qquad
+  \nru{\nhyp}
+      {\J\,(\cext \Gamma N)}
+      {\AddHyp {\Thunk N} \J \Gamma}
+\\[2ex]
   \nru{\nbranch}
@@
       {\AddHyp {\tyzero} \J \Gamma}
-\qquad
+\quad
   \nru{\bbranch}
@@
 \\[2ex]
-  \nru{\nhyp}
-      {\J\,(\cext \Gamma N)}
-      {\AddHyp {\Thunk N} \J \Gamma}
-\qquad
   \nru{\nsplit}
@@
   \nru{\treturn}{\J\,\Gamma}{\Cover \J \Gamma}
-\qquad
+\quad
   \nru{\tbind}
-      {\NeP {\Comp P} \Gamma \qquad \AddHypP P {\Cov\,\J} \Gamma}
+      {\NeP {\Comp P} \Gamma \quad \AddHypP P {\Cov\,\J} \Gamma}
       {\Cover \J \Gamma}
@@
 \[
-\begin{array}{lll}
+\renewcommand{\left}{}
+\renewcommand{\right}{}
+\begin{array}{lcl}
   \tjoin & : & \forall i.\
@@
      (t : \Ne P \Gamma)\;
+\\ \multicolumn 3 {@{\qquad}l} {
      \left(\stmap[\Add P]_\Gamma\;\tjoin[i]\;
@@
      \right)
+}
 \end{array}
@@
 \[
-\begin{array}{lll}
+\begin{array}{l@{~~}l@{~~}l}
   \treturn & : & \forall i.\ \J \todot \Cov[i+1] \J \\
@@
 \[
+\renewcommand{\left}{}
+\renewcommand{\right}{}
 \begin{array}{l@{~}lcl}
@@
     \,\left( \treflect^{P_1}\,(\lambda\,\tau. \ k\,\tau \comp \iota_1) \right)
+\\ &&& \qquad \quad \ \,
     \,\left( \treflect^{P_2}\,(\lambda\,\tau. \ k\,\tau \comp \iota_2) \right)
@@
       \treflect^{P_1} \left( \lambda\,\tau_1\,a_1. \
+\\ &&& \qquad \ \
       \treflect^{P_2} \left( \lambda\,\tau_2\,a_2. \
+\\ &&& \qquad \ \
         k\, (\tau_1 \rcomp \tau_2)\, (\tren\,\tau_2\,a_1, \ a_2)
@@
       {\TmP {P \To N} \Gamma}
-\qquad
+\\[2ex]
   \nru{\tbind}
@@
 shall be as for CBPV except that we need to exchange the
-interpretation function for binders \fbox{$\fden\_ : \Tm N {(\cext
-    \Gamma P)} \to \den \Gamma \todot \den{P \To N}$}.  Since a binder
+interpretation function for binders
+%\fbox{$
+\[
+\fden\_ : \Tm N {(\cext
+    \Gamma P)} \to \den \Gamma \todot \den{P \To N}
+.\]
+%$}.
+Since a binder
 for $P$ performs a maximal splitting on $P$ and takes the form of a
@@
   \tmatch & b & (\nhyp\,e) & \gamma & = & e\,(\gamma,\; b) \\
-  \tmatch & \ttempty & (\nsplit\,e) & \gamma & = & e\,\gamma \\
-  \tmatch & (a_1,a_2) & (\bsplit\,e) & \gamma & = &
-    \tmatch\;a_1\;(\tmap[\Add P]\,(\tmatch\;a_2)\;e)\;\gamma \\
   \tmatch & a & \nbranch & \gamma & = & \tmagic\;a \\
@@
     \tmatch\;a_2\;e_2\;\gamma \\
+  \tmatch & \ttempty & (\nsplit\,e) & \gamma & = & e\,\gamma \\
+  \tmatch & (a_1,a_2) & (\bsplit\,e) & \gamma & = &
+\\ \multicolumn 6 {@{\qquad}l}{
+    \tmatch\;a_1\;(\tmap[\Add P]\,(\tmatch\;a_2)\;e)\;\gamma
+}
+\\
 \end{array}
@@
 is defined as follows:
-\[
+\begin{multline*}
   \fdent{t}
@@
     ~~=~~
+\\ \qquad
     \tmatch
@@
       \; (\tren\,\tau\,\gamma)
-\]
+\end{multline*}
 This completes the definition of the normalization function

### 2019-05-07 ppdp19/ppdp19.tex
[ ppdp19 ] draft more for hyperref no longer needed
[ ppdp19 ] proof reading until half of 2.3

@@
 
-\hypersetup{draft}
+% \hypersetup{draft}  %% debugging url pagebreaks
 
@@
 %\authornote{Both authors contributed equally to this research.}
-\email{andreas.abel@gu.se}
+%\email{andreas.abel@gu.se}
 \orcid{0000-0003-0420-4492}
@@
 }
-\email{christian.sattler@gmail.com}
+%\email{christian.sattler@gmail.com}
 
@@
 by modelling uninterpreted base types as sets of neutral (aka atomic)
-terms, or more precisely, as presheaves or sets of neutral term
-families, in order to facilitate fresh bound variable generation during
+terms.   More precisely, base types are interpreted
+as presheaves or sets of neutral term
+families, in order to facilitate the generation of
+fresh variables of base type during
 reification of functions to lambdas.
 Thanks to $\eta$-equality at function types, free variables of
-function type can be reflected into the semantics as functions
+\emph{function} type can be reflected into the semantics as functions
 applying the variable to their reified argument,
 forming a neutral term.  This mechanism provides us with unknowns of
-function type which can be faithfully reified to normal forms.
+base and function type which can be faithfully reified to normal forms.
 % Since neutrals are closed under application to normal arguments, they
@@
 
-For the extension to sum types (logically, disjunctions, and categorically,
+For the extension to \emph{sum} types (logically, disjunctions, and categorically,
 weak coproducts), this reflection trick does not work anymore.  A
@@
 In this article we only consider weak sum types, producing non-unique
-normal forms, where it does not
-matter in which order the questions are asked (commuting case splits),
-and whether the same
-question is asked several times (redundant case splits).
-The model would need refinement for strong, extensional sums
+normal forms.  In particular, case trees are not normalized, allowing,
+\eg, redundant case splits (\ie, asking the same question twice).
+Further, the order of case splits is not normalized: changing the
+order in which the questions are asked (commuting case splits),
+alters the normal form.
+The model would need refinement for extensional sums (strong
+coproducts) with unique normal forms
 \cite{altenkirchDybjerHofmannScott:lics01,altenkirchUustalu:flops04,balatDiCosmoFiore:popl04,barral:PhD,scherer:popl17}.
@@
 In this work, we formulate NbE for CBPV
-(Section~\ref{sec:cbpv}), with the aim to investigate
-later whether CBN and CBV NbE can be recovered from CBPV NbE via the
-standard translations of the CBN and CBV calculi into CBPV.
-
+(Section~\ref{sec:cbpv}).
+% , with the aim to investigate
+% later whether CBN and CBV NbE can be recovered from CBPV NbE via the
+% standard translations of the CBN and CBV calculi into CBPV.
+%
 In contrast to the normal forms of CBN NbE, which is the algorithmic
@@
 
-By a \emph{presheaf} we will always mean a presheaf over the category $\Cxt\op$,
-i.e., a functor from $\Cxt$ to $\Set$.\footnote{With respect to the terms
-defined later, $\Cxt\op$ is to be seen as a category of contexts and those
-substitutions given by weakening.}  For example, given any type $A$, we can
-consider $A{\in}\_$ as such a presheaf, with action on objects mapping $\Gamma$
-to the set $A \in \Gamma$ of indices of $A$ in $\Gamma$ and action on morphisms
-witnessed by $\treindex : \Gamma \sublist \Delta \to A \in \Gamma \to A \in
+In category theory, a presheaf $\A$ over category $\bC$ is a contravariant
+functor from $\bC$ into the category $\Set$ of sets and functions, \ie, a
+(covariant) functor from the category $\bC\op$ with flipped morphisms
+to $\Set$.
+In this paper, we are only interested in (covariant) functors from $\Cxt$ to
+$\Set$, thus,
+by a \emph{presheaf} we will always mean a presheaf over the category $\Cxt\op$.
+% i.e., a functor from $\Cxt$ to $\Set$.
+% \footnote{With respect to the terms
+% defined later, $\Cxt\op$ is to be seen as a category of contexts and those
+% substitutions given by weakening.}
+For example, given any type $A$, we can
+consider $A{\in}\_$ as such a presheaf, with the action on objects mapping $\Gamma$
+to the set $A \in \Gamma$ of indices of $A$ in $\Gamma$, and the action on morphisms
+% witnessed by
+being
+$\treindex : \Gamma \sublist \Delta \to A \in \Gamma \to A \in
 \Delta$.  The associated functor laws $\treindex\,\tid\,x \equiv x$ and
@@
 Simple types shall be distinguished into positive types $P$ and
-negative types $N$, depending on their root type former.  Function
+negative types $N$, depending on their root type former (for now).  Function
 ($\To$) and product types ($\times$ and $\tyone$) are negative,
@@
     }{\Nf P \Gamma}
-\\[2ex]
+\\[1ex]
   \nru{\tabort}
@@
 
+We use the \fbox{$\forall$} quantifier as implicit dependent function
+type former in our meta-language,
+similar to the polymorphic quantifier in functional
+programming languages such as Haskell.  We may instantiate a
+polymorphic function such as $f : \A \todot \B$ silently, \eg, when $a
+: \A\,\Gamma$ then $f\,a : \B\,\Gamma$.  For clarity, we may sometimes
+provide the instantiation explicitely via subscript, \eg, $f_\Gamma :
+\A\,\Gamma \to \B\,\Gamma$ and $f_\Gamma\,a : B\,\Gamma$.  Similarly,
+to introduce a polymorphic function we may provide the implicit
+abstraction as subscript, such as in
+$\lambda\,{}_\Gamma\, a.\ b : \A \todot \B$ or
+$\lambda\,{}_\Gamma\, (a : \A\,\Gamma).\ b : \A \todot \B$, or
+omit it, \eg,
+$\lambda\,(a : \A\,\Gamma).\ b : \A \todot \B$.
+
 \begin{remark}
@@
 %   \\
-  \tabort^{A \times B} & u & = & \tpair\,(\tabort^A\,u)\,(\tabort^B\,u) \\
+  \tabort^{A_1 \times A_2} & u & = & \tpair\,(\tabort^{A_1}\,u)\,(\tabort^{A_2}\,u) \\
   \tabort^{A \To B}    & u & = & \tabs\,(\tabort^B\,(\tren\,\twk^A\,u)) \\
@@
 \ie, $\tm n \cong t$
-with respect to an equational theory $\cong$ on terms
-(see \cref{sec:eq}).
+with respect to an equational theory $\cong$ on terms%
+. %(see \cref{sec:eq}).
 Further, normalization should decide $\cong$, \ie,
@@
 \emph{evaluation}
-\nofbox{$\denty\_ : (t : A \from \Gamma) \to \den \Gamma \todot
-  \den A$}
-in the identity environment \nofbox{$\idenv\Gamma : \den \Gamma \Gamma$}
+\[%\nofbox{$
+\denty\_ : (t : A \from \Gamma) \to \den \Gamma \todot
+  \den A
+\]%$}
+in the identity environment
+\[%\nofbox{$
+\idenv\Gamma : \den \Gamma \Gamma
+\]%$}
 followed by
 \emph{reification}
-\nofbox{$\down A : \den A \todot \tNf A$}
+\[%\nofbox{$
+\down A : \den A \todot \tNf A
+\]%$}
 (aka quoting).
@@
 
-Presheaves (co)products $\hat0$, $\hat1$, $\hatplus$, and $\hattimes$ are
+Presheaf (co)products $\hat0$, $\hat1$, $\hatplus$, and $\hattimes$ are
 constructed pointwise, \eg, $\hat0\,\Gamma = \bzero$, and given two
@@
 reflection $\up A : \tNe\, A \todot \den A$
-and reification $\down A : \den A \todot \tNf A$ along the way.
+and reification $\down A : \den A \todot \tNf\, A$ along the way.
 \[
@@
   \upG   \tyone \; u        & = & \ttempty \\
-  \downG \tyone \; \ttempty & = & \tunit \\[1ex]
+  \downG \tyone \; \ttempty & = & \tunit \\[2ex]
 % \end{array}
@@
   \upG   {A \times B} \; u  & = & (\upG A (\prj1 u),\;\upG B (\prj2 u)) \\
-  \downG {A \times B} (a,b) & = & \tpair\,(\downG A a)\,(\downG B b) \\[1ex]
+  \downG {A \times B} (a,b) & = & \tpair\,(\downG A a)\,(\downG B b) \\[2ex]
 % \end{array}
@@
   \den {A \To B}    & = & \den A \hatto \den B \\
-  \upG   {A \To B}\; u & = & \lambda \, (\tau : \Gamma \sublist \Delta) \; a %(a : \den A \Delta)
+  \upG   {A \To B}\; u & = & \lambda \, {}_\Delta \, (\tau : \Gamma \sublist \Delta) \, a %(a : \den A \Delta)
      .\ \up[\Delta] B (\tapp\;(\tren\,\tau\, u)\;(\down[\Delta] A a)) \\

### 2019-05-09 ppdp19/ppdp19.tex
[ ppdp19 ] proof-read rest of 2.3 and 2.4
- added informal explanation of the cover monad, linking to oracle
- added implicit context bindings as subscripts
- runNf can be restricted to positive types P
- added conclusion to 2.3 and motivation of 2.4

@@
 
-\begin{remark}
+\begin{remark}[Positive eliminations into negative types]
+\label{rem:abort}
 While the coproduct eliminations $\tcase$ and $\tabort$ are limited to
@@
 $\down[\Delta]A a$
-reified from the semantic value $a$.
+reified from the semantic value $a : \den A \Delta$.
 
@@
 $\tabs\,(\tcase\,\vv0\,\vv0\,\vv0) :
-(o+o) \To o$ in our semantics, as $\Ne o (o+o)$ is empty.
+(o+o) \To o$ in our semantics, because $\Ne o (o+o)$ is empty.
 What is needed are case distinctions on neutrals in the semantics,
@@
 
-We are looking for a cover monad $\C$ that
-offers us these services:
+We are looking for a monad $\C$, called a \emph{cover monad}, that
+takes the role of the \emph{oracle} alluded to in the introduction,
+and offers us these services:
 % \begin{enumerate}
@@
 \begin{array}{l@{~}c@{~}l@{\quad}l}
+\trunNf[\C] & : & \C\,(\tNf\, P) \todot \tNf\, P
+%\\ & & \mbox{run the monad ($\tNf$ only)}
+\\
 \tabort^\C & : & \tNe\, \tyzero \todot \C\,\B
@@
 \\
-\trunNf[\C] & : & \C\,(\tNf A) \todot \tNf A
-%\\ & & \mbox{run the monad ($\tNf$ only)}
-\\
-\end{array}
-\]
-To make things concrete,
-we shall immediately construct an instance of such a cover monad:
+\end{array}
+\]
+Method $\trunNf[\C]$ enables us to \emph{run} a monadic computation of
+a normal form, $\C\,(\tNf\,P)\,\Gamma$, and delivers a normal form
+$\Nf P \Gamma$.  However, $\trunNf[\C]$ is not just the inverse of
+$\treturn[\C]$; instead, the information contained in $\C$ is reified and
+becomes part of the normal form delivered by $\trunNf[\C]$.
+
+Method $\tabort^\C_\Gamma$ allows us to end a computation by exhibiting an
+inconsistency witnessed by a neutral term $u : \Nf A \Gamma$.
+Remember that, ultimately, computations in $\C$ produce a normal form
+extracted by $\trunNf[\C]$.  With $\tabort^\C$ we notify the oracle
+that we are in an absurd case and the desired normal form can be
+trivially constructed by the $\tabort$-function of \Cref{rem:abort}.
+
+Method $\tcase^\C_\Gamma$ allows us, in a computation
+$\C\,\B\,\Gamma$,
+to ask the oracle about a neutral
+term $\NeP{A_1 + A_2}\Gamma$ of sum type.  We have to supply
+handlers for both possible answers:
+A computation $\C\,\B\,(\cext\Gamma{A_1})$ that can utilize an
+additional hypothesis of type $A_1$ in the case of a left injection,
+and analogously a computation $\C\,\B\,(\cext\Gamma{A_2})$ for the
+case of a right injection.
+
+The similarity of the monad services $\tabort^\C$ and $\tcase^\C$ to
+the corresponding constructors for normal forms, or their
+generalization of \Cref{rem:abort}, is hard to miss.
+Unsurprisingly, just \emph{case trees} are a first instance of a cover monad,
+% To make things concrete,
+% we shall immediately construct an instance of such a cover monad:
 the free cover monad
@@
 $\tcase^{\Cov}$.  One can visualize an element $c : \Cov\,\A\,\Gamma$ as
-binary case tree whose inner nodes ($\tcase$) are labeled by a neutral term
+binary case tree whose inner nodes ($\tcase$) are labeled by neutral terms
 of sum type $A_1+A_2$ and its two branches by the context extensions $A_1$
@@
 \[
-\begin{array}{ll@{~~}cl}%{l@{~}l@{~~}c@{~~}l}
+\begin{array}{l@{~~}l@{~~}c@{~~}l}%{l@{~}l@{~~}c@{~~}l}
   \multicolumn 4 l {\trun^A ~:~ \C \den A \todot \den A} \\
+  \trun^P & c & = & \tjoin[\C]\;c \\
   \trun^{\tyone} & c & = & \ttempty
@@
   \\
-  \trun^{A \To B} & c  & = & \lambda \,\tau \; a. \ \trun^B\,
-      (\stmap[\C]\;(\lambda\;\tau'\,f.\;f\;\tid\;(\tren\;\tau'\;a))
+  \trun^{A \To B} & c  & = & \lambda\,_\Delta \,\tau \; a. \ \trun^B\,
+      (\stmap[\C]\;(\lambda\,_\Phi\;\tau'\,f.\;f\;\tid\;(\tren\;\tau'\;a))
   \\ &&& \qquad \qquad (\tren\;\tau\;c))
   \\
-  \trun^{o}      & & = & \tjoin[\C]
-  \\
-  \trun^{\tyzero} & & = & \tjoin[\C]
-  \\
-  \trun^{A + B}   & & = & \tjoin[\C]
-  \\
+  % \trun^{o}      & & = & \tjoin[\C]
+  % \\
+  % \trun^{\tyzero} & & = & \tjoin[\C]
+  % \\
+  % \trun^{A + B}   & & = & \tjoin[\C]
+  % \\
 \end{array}
@@
 since we can transport $a$ to such a $\Phi$ via
-$a' := \tren\; \tau'\; a : \den A \Phi$ but not to a context unrelated to
+$a' := \tren\; \tau'\; a : \den A \Phi$, but not to a context unrelated to
 $\Delta$.  We obtain our input to $\trun^B$ of type $\C \den B \Gamma$
 as an instance of $\stmap[\C]_\Gamma$ applied to the local presheaf morphism
-$(\lambda \; \tau'\, f. \  f\;\tid\;a') :
+$(\lambda\,_\Phi \; \tau'\, f. \  f\;\tid\;a') :
 \Delta \sublist \Phi \to  \den{A \To B}\Phi \to \den B \Phi$
@@
     \quad =
-      \lambda\, (\gamma : \den \Gamma \Delta)
+      \lambda\,_\Delta\, (\gamma : \den \Gamma \Delta)
+      \,_\Phi
       \,(\tau : \Delta \sublist \Phi)
@@
     \todot \den{B} \\
-  \dencase^B & = & \lambda\,c\;f_1\;f_2.\
+  \dencase^B & = & \lambda\,c\,f_1\,f_2.\
     \trun^B (\stmap[\C]\;(\lambda\,\tau.\ [f_1\,\tau,\ f_2\,\tau])\;c) \\
@@
 \]
+We have obtained a computable function $\tnorm$ that yields for any
+derivation $A \from \Gamma$ a normal derivation $\Nf A \Gamma$.
+This amounts to proving
+consistency and canonicity of intuitionistic propositional logic, the
+Curry-Howard counterpart of our STLC with products and sums.
+The normalization function is parametrized by an arbitrary cover monad
+$\C$, which can be implemented via case trees ($\Cov$).  However,
+different implementations are possible.  In the next section, we exhibit
+another canonical choice for $\C$: the parameterized continuation monad on presheafs.
 
@@
 normalization by evaluation.
-In our setting, we use a continuation monad $\CC$ on presheaves
-defined as
+In our setting, we use a continuation monad $\CC$ on \emph{presheaves}
+$\B$ defined by
 % An alternative implementation of $\C$ is the continuation monad $\CC$
@@
 \[
-  \CC\;\J = \forall A.\ (\J \hatto \tNf\,A) \hatto \tNf\,A
+  \CC\;\B = \forall P.\ (\B \hatto \tNf\,P) \hatto \tNf\,P
   .
@@
 The answer type of this continuation monad is always $\tNf$, however,
-we are polymorphic in the simple type $A$ of normal forms we produce.%
+we are polymorphic in the positive type $P$ of normal forms we produce.%
 
@@
 straightforward evidence that $\CC$ is a strong monad.
-The method $\trunNf[\CC] : \CC\,(\tNf\,A) \todot \tNf\,A$
-exists by definition, using the identity continuation $\tNf\,A \hatto \tNf\,A$.
+The method $\trunNf[\CC] : \CC\,(\tNf\,P) \todot \tNf\,P$
+exists by definition, using the identity continuation $\tNf\,P \hatto \tNf\,P$.
 In the following, we demonstrate that $\CC$
@@
 \[
+\renewcommand{\left}{}
+\renewcommand{\right}{}
 \begin{array}{l}
-  \tabort^{\CC} \, (u : \Ne \tyzero \Gamma) ~:~  \CC\,\J\,\Gamma \\
-  \tabort^{\CC}
+  \tabort^{\CC}_\Gamma \, (u : \Ne \tyzero \Gamma) ~:~  \CC\,\B\,\Gamma \\[0.2ex]
+  \tabort^{\CC}_\Gamma
     \; u
-    \; (\tau : \Gamma \sublist \Delta)
-    \; (k : (\J \hatto \tNf\,B)\Delta)
-     ~=~
+    \, _\Delta
+    \, (\tau : \Gamma \sublist \Delta)
+    \; (k : (\B \hatto \tNf\,P)\,\Delta)
+     ~=~ \\  \qquad
     \tabort\,(\tren\,\tau\,u)
 \\[2ex]
-  \tcase^{\CC}
-     \; (u : \NeP {A_1 + A_2} \Gamma) \\ \qquad \quad
-     \; (c_1 : \CC\,\J\,(\cext \Gamma {A_1}))
-     \; (c_2 : \CC\,\J\,(\cext \Gamma {A_2}))
-    ~:~ \CC\,\J\,\Gamma \\
-  \tcase^{\CC}
+  \tcase^{\CC}_\Gamma
+     \; (u : \NeP {A_1 + A_2} \Gamma) \, _\Delta \\ \qquad \quad
+     \; (c_1 : \CC\,\B\,(\cext \Gamma {A_1}))
+     \; (c_2 : \CC\,\B\,(\cext \Gamma {A_2}))
+    ~:~ \CC\,\B\,\Gamma \\
+  \tcase^{\CC}_\Gamma
     \; u
@@
     \; (\tau : \Gamma \sublist \Delta)
-    \; (k : (\J \hatto \tNf\,B)\Delta)
+    \; (k : (\B \hatto \tNf\,P)\,\Delta)
      ~=~ \\  \qquad
@@
   \begin{array}{l@{~~}c@{~~}l}
-     n_i & : & \Nf B {(\cext \Delta {A_i})} \\
+     n_i & : & \Nf P {(\cext \Delta {A_i})} \\
      n_i & = & c_i
@@
       \; \left(\lambda
-            \; (\tau' : \cext \Delta {A_i} \sublist \Phi)
-            \; (j :  \J\,\Phi) .
+            \, _\Phi
+            \, (\tau' : \cext \Delta {A_i} \sublist \Phi)
+            \; (j :  \B\,\Phi) .
             \ k
@@
 partial evaluation \cite[Figure~8]{danvy:popl96}.  However, he uses
-shift-reset style continuations which can be programmed in the
+shift-reset style continuations, which can be programmed in the
 continuation monad, and relies on Scheme's \texttt{gensym} to produce
-fresh variables names rather than using Kripke function space / presheaves.
+fresh variables names rather than using the Kripke function space / presheaves.
 

### 2019-05-09 ppdp19/ppdp19.tex
[ ppdp19 ] read 3.1

@@
 However, we shall use $\tThunk$ for switch $\downarrow$ and $\tComp$ for
-switch $\uparrow$.
+switch $\uparrow$, to avoid confusion with our notation for reflection
+and reification.
 \[
 \begin{array}{l@{~~}l@{~~}l@{~~}r@{~~}l}%@{\qquad}l}
-  \PTy & \ni & P,Q & ::= & \patom
+  \PTy & \ni & P & ::= & \patom
     \mid \tyone \mid P_1 \times P_2
@@
 \\
-  \NTy & \ni & N,M & ::= & \natom
+  \NTy & \ni & N & ::= & \natom
     \mid \top \mid N_1 \ntimes N_2
@@
 these names uninspiring unless you have good knowledge of the intended
-model.  Further, CBPV employs labeled sums $\Sigma_I (P_i)_{i : I}$
+model.  Further, CBPV \cite{levy:hosc06}
+employs labeled sums $\Sigma_I (P_i)_{i : I}$
 and labeled records $\Pi_I (P_i)_{i : I}$ for up to countably infinite
@@
 switch from the negative to the positive type interpretation, places
-the cover monad.  Positive atoms, standing for value types without
+the cover monad.  Positive atoms $\patom$, standing for value types without
 constructors, are only inhabited by variables $x : \patom \in \Gamma$.
-Negative atoms stand for computation types without own eliminations,
-thus, their inhabitants stem only from eliminations of more complex
-types, made from positive eliminations captured in $\C$ and negative
+Negative atoms $\natom$ stand for computation types without own eliminations.
+Thus, their inhabitants stem only from eliminations of more complex
+types.  They are built from positive eliminations captured in $\C$ and negative
 eliminations chained together as neutral $\tNe\,\natom$, which we
@@
 $\Thunk N$ are embedded via $\tforce$.
-Further, values of type $P$ can embedded
+Further, values of type $P$ can be embedded
 via $\tret$, producing a term of type $\Comp P$.  Such terms are
@@
 % \begin{array}{lllrl@{\qquad}l}
-%   \PTy & \ni & P,Q & ::= & \patom
+%   \PTy & \ni & P & ::= & \patom
 %     \mid \tyzero \mid P_1 + P_2
@@
 % \\
-%   \NTy & \ni & N,M & ::= & \natom
+%   \NTy & \ni & N & ::= & \natom
 %     \mid \top \mid N_1 \ntimes N_2

### 2019-05-10 ppdp19/ppdp19.tex
[ ppdp19 ] run is independent of runNf!  CBPV has no runNf
Fixed implementation of reification, using run instead of runNf.
Stressed that run is definable from join thanks to the definition of the semantics.

@@
 
-\usepackage{cleveref}
+\usepackage{cleveref} % \cref \Cref
+%\usepackage{dashrule}
 
@@
 \subsection{STLC and its normal forms}
+\label{sec:stlc}
 
@@
 \subsection{Normalization}
+\label{sec:normstlc}
 
@@
 
-The unwinding of the monad $\C$ via $\trunNf$ can be extended to a
+Since positive types $P$ have a monadic interpretation, there is a
+natural transformation $\C\,\den P \todot \den P$ which is simply
+$\tjoin[\C]$.  It
+can be extended to a
 natural transformation $\trun^A : \C\,\den A \todot \den A$ for any simple
@@
 We proceed by induction on $A$.
-Positive semantic types $\den P$ are already monadic,
-and $\trun^P$ is simply the $\tjoin$ of the monad $\C$.
 At negative types we can recurse pointwise at a smaller type,
@@
 \]
+It is noteworthy that the interpreter $\denty{\_}$ does not use any of
+the services of the cover monad; in fact, it only requires the monad
+$\C$ to be strong.  The interpreter is completely \emph{generic} for
+the CBN monadic semantics of types.  This restores the spirit of STLC
+NbE for sum types:  Take an \emph{off-the-shelf} interpreter for terms,
+implement reflection and reification after interpreting base types as sets
+(presheaves) of neutrals, plug them together with the interpreter, and
+voila!, there is your normalizer!
+Our insight here is that \emph{off-the-shelf interpreter} means
+\emph{monadic} in the presence of positive types.
+
 To complete the normalization function
@@
 mutually defined interpretations
-$\den P$ and $\den N$ of positive and negative types as presheaves.
-\[
+$\den P$ and $\den N$ of positive and negative types as presheaves
+(see \Cref{fig:tyint}).
+This interpretation is parametrized by a strong monad $\C$ on
+presheaves.
+\begin{figure}[htbp]
+\centering
+%\hrule width0.3\textwidth
+\hrulefill
+$$
 \begin{array}{lcl}
@@
   \den{\patom} & = & \patom{\in}\_ \\
-\end{array}
-%\qquad
-\]
-\[
-\begin{array}{lcl}
+\\
+% \end{array}
+% %\qquad
+% \begin{array}{lcl}
   \den{\top} & = & \hatone \\
   \den{N_1 \ntimes N_2} & = & \den{N_1} \hattimes \den{N_2} \\
-  \\
   \den{P \To N} & = & \den P \hatto \den N \\
@@
 \end{array}
-\]
+$$
+\hrulefill
+  \caption{Interpretation of polarized types}
+  \label{fig:tyint}
+\end{figure}
 Semantically, we do not distinguish between positive and negative
@@
 switch from the negative to the positive type interpretation, places
-the cover monad.  Positive atoms $\patom$, standing for value types without
+the monad $\C$.  Positive atoms $\patom$, standing for value types without
 constructors, are only inhabited by variables $x : \patom \in \Gamma$.
@@
 shall define below.
-The method $\trunNf[\C]$ of cover monad $\C$ can be extended to
-$\trun^N : \C \den N \todot \den N$ for negative types $N$, by recursion
-on $N$.
+The multiplication $\tjoin[\C]$ of the strong monad $\C$ can be
+extended to a natural transformation
+$\trun^N : \C \den N \todot \den N$ for negative types $N$.
+The construction proceeds by recursion on $N$, exactly as in \Cref{sec:normstlc}.
 This makes the effects of $\C$ available at any negative type,
@@
 
-Assuming a family $\Tm N \Gamma$ of terms of negative type $N$ in
-context $\Gamma$, values \fbox{$v : \PValT P \Gamma$}
-of positive type $P$ shall be constructed by the
-following rules:
-\begin{gather*}
-  \nru{\tvar}{P \in \Gamma}{\PValT P \Gamma}
-  % \nru{\pvar}{P \in \Gamma}{\PValT P \Gamma}
-\qquad
-  \nru{\tthunk}{\Tm N \Gamma}{\PValTP{\Thunk N}\Gamma}
-\\[2ex]
-  \nru{\punit}{}{\PValT{\tyone}\Gamma}
-\qquad
-  \nru{\ppair}
-      {\PValT{P_1}\Gamma \qquad \PValT{P_2}\Gamma}
-      {\PValTP{P_1 \times P_2}\Gamma}
-\\[2ex]
-  \nru{\tinj_i}{\PValT{P_i}\Gamma}{\PValTP{P_1 + P_2}\Gamma}
-\end{gather*}
-
-The terms of pure CBPV are given by the inductive family \fbox{$\Tm N \Gamma$}.
-It repeats the introductions and eliminations of negative types,
-except that application is restricted to values.  Values of type
+Value terms \fbox{$v : \PValT P \Gamma$}, or short, \emph{values},
+and computation terms \fbox{$t : \Tm N \Gamma$}, or short,
+\emph{terms}, are defined mutually by the rules in \Cref{fig:tmcbpv}.
+Values inhabit positive types $P$ and are given by introduction rules
+only, whereas terms inhabit negative types $N$ and comprise both
+introduction rules for negative types as well as elimination rules for
+both positive ($\tsplit$, $\tcase$, $\tabort$)
+and negative types.
+%
+% Assuming a family $\Tm N \Gamma$ of terms of negative type $N$ in
+% context $\Gamma$, values \fbox{$v : \PValT P \Gamma$}
+% of positive type $P$ shall be constructed by the
+% following rules:
+%
+% The terms of pure CBPV are given by the inductive family \fbox{$\Tm N \Gamma$}.
+% It repeats the introductions and eliminations of negative types,
+% except
+Note
+that application is restricted to values.  Values of type
 $\Thunk N$ are embedded via $\tforce$.
@@
 $N$.  This is justified by the monadic character of negative
-types, by virtue of $\trun^N$.  Finally, there are eliminators
-($\tsplit$, $\tcase$, $\tabort$) for values of positive product and
-sum types.
+types, by virtue of $\trun^N$.
+% Finally, there are eliminators
+% ($\tsplit$, $\tcase$, $\tabort$) for values of positive product and
+% sum types.
 \begin{figure}[htbp]
@@
 \hrulefill
+\begin{gather*}
+  \nru{\tvar}{P \in \Gamma}{\PValT P \Gamma}
+  % \nru{\pvar}{P \in \Gamma}{\PValT P \Gamma}
+\qquad
+  \nru{\tthunk}{\Tm N \Gamma}{\PValTP{\Thunk N}\Gamma}
+\\[2ex]
+  \nru{\punit}{}{\PValT{\tyone}\Gamma}
+\qquad
+  \nru{\ppair}
+      {\PValT{P_1}\Gamma \qquad \PValT{P_2}\Gamma}
+      {\PValTP{P_1 \times P_2}\Gamma}
+\\[2ex]
+  \nru{\tinj_i}{\PValT{P_i}\Gamma}{\PValTP{P_1 + P_2}\Gamma}
+\end{gather*}
+\dotfill
 \begin{gather*}
@@
       {\TmP {N_1 \ntimes N_2} \Gamma}
-\\[2ex]
+\end{gather*}
+\dotfill
+\begin{gather*}
+%\\[2ex]
   \nru{\tforce}
@@
       {\Tm N \Gamma}
-\qquad
+\\[2ex]
   \nru{\tapp}
@@
       {\Tm N \Gamma}
-\\[2ex]
+\qquad
   \nru{\tprj_i}
@@
       {\Tm N \Gamma}
-\\[2ex]
+\end{gather*}
+\dotfill
+\begin{gather*}
+%\\[2ex]
   \nru{\tsplit}
@@
     }{\Tm N \Gamma}
+\quad
+  \nru{\tabort}
+    {\PValTm \tyzero \Gamma}
+    {\Tm N \Gamma}
 \\[2ex]
@@
     }{\Tm N \Gamma}
-\\[2ex]
-  \nru{\tabort}
-    {\PValTm \tyzero \Gamma}
-    {\Tm N \Gamma}
+%\\[2ex]
 \end{gather*}
@@
 \hrulefill
-  \caption{Terms of CBPV}
+  \caption{Value and computation terms (CBPV)}%Terms of CBPV}
   \label{fig:tmcbpv}
@@
 is straightforward, thanks to the pioneering work of
-Moggi \cite{moggi:infcomp91} and
+Moggi \cite{moggi:infcomp91} and the work that
 Levy \cite{levy:hosc06}
@@
 
+The design of normal forms for pure CBPV follows the same principles
+as for the STLC with sums in \Cref{sec:stlc}.  We prevent
+$\beta$-redexes, i.e., eliminations of a type immediately following
+its introduction, by restricting the terms in elimination positions to
+neutrals.  Due to the focused nature of CBPV, neutrals of positive
+type are just variables.  We achieve $\eta$-long forms by only
+embedding neutrals of negative base type $\natom$ into normal forms.
+Maximal focusing for positive types is achieved by only admitting
+variables of base type $\patom$ in values.
+% and only variables of positive base type $\patom$ into.  For CBPV, this
+% concerns only negative atoms $\natom$, as positive atoms $\patom$ are
+% only inhabited by variables.
+In the following, we present the
+grammar for normal forms in detail.
+
 Positive normal forms are values $v : \VNf P \Gamma$
@@
 Variables are originally introduced by either $\tabs$ or
-the $\tbind$ing of a neutral of type $\Comp P$ to a new variable of type
+the $\tbind$\-ing of a neutral of type $\Comp P$ to a new variable of type
 $P$.  Variables of composite value type can be broken down by pattern
@@
   \nru{\treturn}{\J\,\Gamma}{\Cover \J \Gamma}
-\qquad
+\quad
   \nru{\tbind}
@@
       {\Cover \J \Gamma}
+%\\[1ex]
+\quad
+  \nru{\tabort}
+      {\tyempty \in \Gamma}
+      {\Cover \J \Gamma}
 \\[2ex]
@@
       }
-      {\Cover \J \Gamma}
-\\[2ex]
-  \nru{\tabort}
-      {\tyempty \in \Gamma}
       {\Cover \J \Gamma}
@@
 \end{gather*}
-We again can run the cover monad on normal forms, \ie, have
-$\trunNf : \Cov\,(\tNf N) \todot \tNf N$, which extends to
-negative semantic values
-$\trun : \Cov \den N \todot \den N$.
+Note that we do not have a function
+$\trunNf[\Cov] : \Cov\,(\tNf\,N) \todot \tNf\,N$ this time, instead, we
+directly employ $\Cov$ in the definition of $\tNf$ at the base case
+$\natom$ and $\Comp P$.
 
@@
 positive reflection $\upG P : P \in \Gamma \to \Cover {\den P} \Gamma$
-needs to run in the monad.  Luckily, as $\down N$ produces a normal
-form, monadic intermediate computations are permitted under a final
-$\trunNf$.  Negative reflection $\up N : \tNe\,N \todot \den N$ is as
+needs to run in the monad.
+Luckily, we can unwind the monad using $\trun^N$ before we recursively
+reify at type $N$.
+% %% TODO
+% Luckily, as $\down N$ produces a normal
+% form, monadic intermediate computations are permitted under a final
+% $\trunNf$.
+Negative reflection $\up N : \tNe\,N \todot \den N$ is as
 before generalized from variables to neutrals, to handle the breaking
@@
 \def\arraystretch{1.3}
-\begin{array}{l@{\,}l@{~~}c@{~~}l}
+\begin{array}{l@{\,}lcl}%@{~~~}c@{~~~}l}
   \multicolumn 2 l {\upG P} & : & P \in \Gamma \to \Cov\,\den P \, \Gamma\\
@@
 \\[2ex]
-% \end{array}
-% \,
-% \begin{array}{lcl}
+\end{array}
+\]
+\[
+\def\arraystretch{1.3}
+\begin{array}{l@{\,}lcl}
     \multicolumn 2 l {\down P} & : & \den P \todot \tVNf\,P \\
@@
   \upG {P \To N} & u & = & \lambda
-    \; \tau \; a
+    \, _\Delta
+    \, \tau \; a
     % \;(\tau : \Gamma \sublist \Delta)
@@
     \\
-  \downG {P \To N} & f & = & \tabs \left( \trunNf \left(
-    \stmap
+  \downG {P \To N} & f & = &
+    \left( \tabs
+     \comp \down[\cext \Gamma P] N
+     \comp \trun^N_{\cext \Gamma P}
+\\ &&& \quad {}
+     \comp \stmap
       \left(
         \lambda
-          \; \tau \; a
-          % \;(\tau : (\cext \Gamma P) \sublist \Delta)
+          \,_\Delta
+          \, \tau
+          % \,(\tau : (\cext \Gamma P) \sublist \Delta)
+          \; a
           % \;(a : \den P \Delta)
@@
         \right)
+    \right)
     \;\tfresh^P_\Gamma
-    \right) \right)
+    % \tabs \left( \down[\cext \Gamma P] N
+    % \left( \trun^N_{\cext \Gamma P} \left(
+    % \stmap
+    %   \left(
+    %     \lambda
+    %       \,_\Delta
+    %       \, \tau \; a
+    %       % \;(\tau : (\cext \Gamma P) \sublist \Delta)
+    %       % \;(a : \den P \Delta)
+    %       . \
+    %       f\; (\twk^P \rcomp \tau)\; a
+    %     \right)
+    %   \;\tfresh^P_\Gamma
+    % \right) \right) \right)
     \\
@@
 \[
-  \tnorm\,(t : \Tm N \Gamma) = \trunNf\,
-    \left(
-      \tmap\,\left(\down N \comp \denty t\right)\,\idenv\Gamma
-    \right)
+  \tnorm\,(t : \Tm N \Gamma) = \left (\downG N
+    \comp \trun^N_\Gamma
+    \comp \tmap\,\denty t\right) \idenv\Gamma
+  % \tnorm\,(t : \Tm N \Gamma) = \downG N
+  %   \left(
+  %     \trun^N_\Gamma\,(\tmap\,\denty t\,\idenv\Gamma)
+  %   \right)
 .
@@
 introduce positive types ($\tVnf$).  However, the elimination of
-positive types is still rather non-deterministic.
+positive types is still rather non-deterministic:
 It is possible to only partially break up a composite positive type
@@
 pattern matching.
-The last refinement, chaining also the positive eliminations, will be
+The last refinement of normal forms,
+chaining also the positive eliminations, will be
 discussed in the following section.

### 2019-05-10 ppdp19/ppdp19.tex
[ ppdp19 ] quick reading of Section 4
[ ppdp19 ] direct links to Agda HTML for sections 2 and 4

@@
 % The first command in your LaTeX source must be the \documentclass command.
-\documentclass[sigplan,screen,fleqn]{acmart}
+\documentclass[sigplan,screen,fleqn,review]{acmart} % review for draft
+%% DRAFT only:
+\settopmatter{printacmref=false} % Removes citation information below abstract
+\renewcommand\footnotetextcopyrightpermission[1]{} % removes footnote with conference information in first column
+\sloppy
 
@@
 Our design choices were guided by an Agda formalization of sections 2
-(complete) and 4 (partial), available at \url{https://github.com/andreasabel/ipl}.
+(complete, see \url{https://andreasabel.github.io/ipl/html/NfModelMonad.html})
+and 4 (partial, see \url{https://andreasabel.github.io/ipl/html-focusing/Polarized.html}).
 Agda was particularly helpful to correctly handle the renamings
@@
 continuation is necessary since we can conclude with
-\emph{ex falsum quod libet}.
+\emph{ex falso quodlibet}.
 At type $\tyone$, there is no need to add a new variable, as values
@@
 \[
-\begin{array}{l@{~}lcl}
+\begin{array}{l@{~}l@{~~}c@{~~}l}
   \downG {P \To N} & f & = & \tabs
@@
         \left(
-          \lambda \, (\tau : \Gamma \sublist \Delta) \; a .\
+          \lambda \, _\Delta \, (\tau : \Gamma \sublist \Delta) \; a .\
           \down[\Delta] P (f \; \tau \; a)
@@
 
-We have implemented NbE for CBPV and polarized lambda calculus
+We have defined NbE for CBPV and polarized lambda calculus,
 formulated with intrinsically well-typed syntax and presheaf

### 2019-05-10 ppdp19/ppdp19.tex
[ ppdp19 ] link to Agda soundness of NbE for IPL in footnote

@@
 terms $t,t'$ with $t \cong t'$ should have the same normal form $n$.
-In this article, we implement only the normalization function
+In this article, we present only the normalization function
 \fbox{$\tnorm : A \from \Gamma \to \Nf A \Gamma$}
-with proving its soundness and completeness.
+without proving its soundness and completeness.%
+\footnote{An Agda formalization of NbE for IPL with
+soundness-by-construction is listed at
+  \url{https://andreasabel.github.io/ipl/html/NbeModel.html}.
+It is demonstrated that normalization does not change the
+interpretation of a term as function in the meta-language.
+Instead of presheaves, we work with Kripke predicates there.
+The main contribution of our work, the utilization of a cover monad
+for NbE, is however more clearly presented by just constructing the
+normalization function, as we do in this article.
+}
 From a logical perspective, we will compute for each derivation of $A

### 2019-05-10 ppdp19/ppdp19.tex
Some typos and a comment.

@@
 
-In category theory, a presheaf $\A$ over category $\bC$ is a contravariant
+In category theory, a presheaf $\A$ over a category $\bC$ is a contravariant
 functor from $\bC$ into the category $\Set$ of sets and functions, \ie, a
@@
 : \A\,\Gamma$ then $f\,a : \B\,\Gamma$.  For clarity, we may sometimes
-provide the instantiation explicitely via subscript, \eg, $f_\Gamma :
+provide the instantiation explicitly via subscript, \eg, $f_\Gamma :
 \A\,\Gamma \to \B\,\Gamma$ and $f_\Gamma\,a : B\,\Gamma$.  Similarly,
@@
 % $\up[\tyone + \tyone]^{\tyone + \tyone} \vv0 : \bone \bplus \bone$
-$\tfresh^{A + B}_{\cempty} : \den A (A{+}B) \bplus \den B (A{+}B)$
+$\tfresh^{A + B}_{\cempty} : \den A (A{+}B) \bplus \den B (A{+}B)$,
 % $\up[A + B]{A + B} \vv0 : \den A (A{+}B) \bplus \den B (A{+}B)$
@@
 handlers for both possible answers:
-A computation $\C\,\B\,(\cext\Gamma{A_1})$ that can utilize an
+a computation $\C\,\B\,(\cext\Gamma{A_1})$ that can utilize an
 additional hypothesis of type $A_1$ in the case of a left injection,
@@
 $\treturn[\Cov]$, $\tabort^{\Cov}$, and
-$\tcase^{\Cov}$.  One can visualize an element $c : \Cov\,\A\,\Gamma$ as
+$\tcase^{\Cov}$.
+% TODO: Shouldn't this better be trees with nullary and binary nodes
+% and leaves always an element of A?
+One can visualize an element $c : \Cov\,\A\,\Gamma$ as a
 binary case tree whose inner nodes ($\tcase$) are labeled by neutral terms
@@
 (aka Kleisli extension) replaces
-these leaves by further case trees.  Bind is realized via $\tjoin[\Cov]$
+these leaves by further case trees.  Bind is realized via $\tjoin[\Cov]$,
 which flattens a 2-level case tree, \ie, a case tree with case trees as
@@
 Since positive types $P$ have a monadic interpretation, there is a
-natural transformation $\C\,\den P \todot \den P$ which is simply
+natural transformation $\C\,\den P \todot \den P$, which is simply
 $\tjoin[\C]$.  It
@@
 $\den \cempty = \hatone$ and
-$\den{\cext \Gamma A} = \den \Gamma \hattimes \den A$
+$\den{\cext \Gamma A} = \den \Gamma \hattimes \den A$,
 and obtain a natural projection function

### 2019-05-10 ppdp19/ppdp19.tex
Typos.

@@
 \affiliation{%
-  \institution{Nottingham University}
+  \institution{School for Computer Science, University of Nottingham}
   \streetaddress{}
   \city{Nottingham}
-  \country{UK}
+  \country{United Kingdom}
 }
@@
 $a : \den A \Delta$, and now we want to apply every function
-$f : \den{A \To B}$ in the cover $c$ to argument $a$.  Clearly,
+$f : \den{A \To B}$ % TODO: this doesn't really typecheck...
+in the cover $c$ to argument $a$.  Clearly,
 $\tmap[\C]$ is not applicable since it would expect a
@@
 our treatment extends directly to the infinite case, since we are
-working in type theory which allows infinitely branching inductive
+working in type theory, which allows infinitely branching inductive
 types.   As a last difference, CBPV does not consider base types; in
@@
 $\C\,\A_1 \todot \C\,\A_2 \todot \C\,(\A_1 \hattimes \A_2)$
-called $\star$ by McBride and Paterson \cite[Section 7]{mcbridePaterson:applicative}
+called $\star$ by McBride and Paterson \cite[Section 7]{mcbridePaterson:applicative},
 which for monads $\C$ can be defined by
@@
 In the first branch, we recursively reflect at $P_1$.  Its
-continuation will receive a semantic value in $\den{P_1}$ which we
+continuation will receive a semantic value in $\den{P_1}$, which we
 inject via $\iota_1$ into $\den{P_1 + P_2}$ to pass it to $k$.  The
@@
 $\tau_1$ and still needs
-to be transported with $\tau_2$ before paired up with $a_2$.
+to be transported with $\tau_2$ before being paired up with $a_2$.
 
@@
 ones of CBPV minus the positive eliminations ($\tsplit$, $\tcase$,
-$\tabort$), the added negative variable rule ($\nvar$) and the
+$\tabort$), the added negative variable rule ($\nvar$), and the
 necessary changes to the binders $\tabs$ and $\tbind$.
@@
 applying it to a value $v$ of type $P$ amounts to a complete
-matching of $v$ against the case tree, and bind the remaining
+matching of $v$ against the case tree and binding the remaining
 atomic and negative crumbs.  This matching can be defined for a

### 2019-05-10 ppdp19/ppdp19.tex
Added the two references from FSCD review #2. (Warning: appended to auto-ppdp19.bib)
[ ppdp19 ] contributions, acks, conclusion (citing Alechina, Mendler... CSL01)

--- 2019-05-10 ppdp19/ppdp19.tex
@@
 abundantly present when working with presheaves.
+
+To summarize, in this article we make the following contributions:
+\begin{itemize}
+\item We retell the story of normalization by evaluation (NbE) for intrinsically
+  typed lambda terms with weak sum through a generic monadic
+  interpreter.  We isolate the services the monad needs to offer in
+  order to facilitate reification.
+
+\item We observe that the placement of the monad in the process of the
+  type interpretation can be determined by polarization, \ie,
+  separating types into positive and negative types.
+  The most economic placement of the monad is achieved with deep
+  polarization such as in focused calculi.
+\item As a first focused calculus, we consider call-by-push-value
+  (CBPV) which identifies positive types with value types and negative
+  types with computation types.  To our best knowledge, we are the
+  first to study NbE for pure, effect-free CBPV.  Therein, we observe
+  that the structures of CBPV to control effects naturally structure
+  the NbE process as well.
+\item Finally, we define NbE for a fully focused calculus, the
+  polarized lambda calculus.
+  We use a factored presentation of terms and normal forms,
+  which uses advanced features of Agda (nested and sized types)
+  in our partial formalization.
+  % Again, we seem to be the first to define
+  % NbE for the polarized lambda calculus, in spite of the substantial
+  % proof-theoretic studies it has been subjected to.
+\end{itemize}
 
@@
 
-\section{Conclusion}
+\section{Conclusion and Further Work}
 \label{sec:concl}
@@
 formulated with intrinsically well-typed syntax and presheaf
-semantics.  As a side result, we have proven semantically
+semantics.  At the heart of our development stands the notion of a
+cover monad, as alternative to the more common sheaf semantics, to
+handle sum types.
+
+As a side result, we have proven semantically
 that the normal forms of both systems are logically complete, \ie,
@@
 normal form should agree, and normalization should decide a suitable
-equational theory on terms.
+equational theory on terms.  To this end, we may switch from the
+category of presheaves to the category of Kripke predicates, as
+investigated through an Agda formalization.%
+\footnote{\url{https://andreasabel.github.io/ipl/html/NbeModel.html}}
 
@@
 the one inserting a minimal amount of $\tThunk$ and $\tComp$ transitions.
+Finally, we might revisit our interpretation of $\tThunk$ as a monoidal
+comonad as suggested by the categorical models of modal logic S4
+\cite{alechinaMendlerDePaivaRitter:csl01}.
 
@@
 Thorsten in turn credits his inspiration to an ALF proof by Thierry
-Coquand.  Thanks to Hugo Herbelin for feedback and spotting some typos.
+Coquand.  Thanks to Hugo Herbelin for encouraging feedback and spotting some
+typos.  Thanks to the anonymous referees of a previous version of this
+paper for the valuable feedback and remarks that helped improving the
+text of this article.
+
+The first author acknowledges financial support from
+VR Grant 2014-04864 \emph{Termination
+      Certificates for Dependently-Typed Programs and Proofs via
+      Refinement Types}
+and the EU Cost Action CA15123 EUTYPES \emph{Types for Programming and Verification}.
 \end{acks}

### 2019-05-10 ppdp19/ppdp19.tex

--- 2019-05-10 ppdp19/ppdp19.tex
@@
 who designed the call-by-push-value (CBPV) lambda-calculus to embed
-both the CBN and CBV lambda calculus.
+both the CBN and CBV lambda calculus.  Equational theories and
+reduction-based normalization for CBPV were recently studied and
+formalized in Coq~\cite{rizkallah2018formal,forster2019call}.
+% TODO: improve last sentence.
 

### 2019-05-10 ppdp19/ppdp19.tex
[ ppdp19 ] DBLP entries for new citations

@@
 reduction-based normalization for CBPV were recently studied and
-formalized in Coq~\cite{rizkallah2018formal,forster2019call}.
+formalized in Coq~\cite{rizkallahGarbuzovZdancewic:itp18,forsterSchaeferSpiesStark:cpp19}
 % TODO: improve last sentence.
@@
 \bibliographystyle{ACM-Reference-Format}
-\bibliography{auto-ppdp19}
+\bibliography{auto-ppdp19,byhand}
 

### 2019-06-12 ppdp19/ppdp19.tex
[ ppdp19 ] fixed issues found by reviewer 1

@@
 \emph{subformula} property \cite{prawitz:natded,matthes:shortproofs}.
-Normal forms \fbox{$n : \tNf A\, \Gamma$} are mutually defined with
+Normal forms \fbox{$n : \Nf A \Gamma$} are mutually defined with
 \emph{neutral} normal forms \fbox{$u : \Ne A \Gamma$}.
@@
 Method $\tabort^\C_\Gamma$ allows us to end a computation by exhibiting an
-inconsistency witnessed by a neutral term $u : \Nf A \Gamma$.
+inconsistency witnessed by a neutral term $u : \Ne 0 \Gamma$.
 Remember that, ultimately, computations in $\C$ produce a normal form
@@
 Since $\tThunk$ serves only as an embedding of negative into positive
-types and has no semantic effect,
+types in our semantics,
 we interpret thunking and forcing by the identity.

### 2019-07-11 ppdp19/ppdp19.tex
[ ppdp19 ] switched to proceedings version

@@
 % The first command in your LaTeX source must be the \documentclass command.
-\documentclass[sigplan,screen,fleqn,review]{acmart} % review for draft
+\documentclass[sigconf,screen,fleqn]{acmart} % review for draft
 %% DRAFT only:
-\settopmatter{printacmref=false} % Removes citation information below abstract
-\renewcommand\footnotetextcopyrightpermission[1]{} % removes footnote with conference information in first column
+%\settopmatter{printacmref=false} % Removes citation information below abstract
+%\renewcommand\footnotetextcopyrightpermission[1]{} % removes footnote with conference information in first column
 \sloppy
@@
 
-The terms \fbox{$\Tm N \Gamma$} of the focused lambda calculus are the
+The terms \fbox{$\Tm N \Gamma$} of the polarized lambda calculus are the
 ones of CBPV minus the positive eliminations ($\tsplit$, $\tcase$,

### 2019-07-12 ppdp19/ppdp19.tex
[ ppdp19 ] proof reading for final version

@@
 \sloppy
-
+%\usepackage{array}
 \newcommand\hmmax{0}
@@
 The idea behind \emph{normalization by evaluation} (NbE) is to utilize a
-standard interpreter, usually evaluating closed terms, to compute the
-normal form of an open term.
+standard interpreter that evaluates closed terms to compute the
+normal form of an open term, \ie, a term that may contain free variables.
 The normal form is obtained by a type-directed \emph{reification}
@@
 terms.   More precisely, base types are interpreted
-as presheaves or sets of neutral term
+as presheaves, or sets of neutral term
 families, in order to facilitate the generation of
@@
 For the extension to \emph{sum} types (logically, disjunctions, and categorically,
-weak coproducts), this reflection trick does not work anymore.  A
+weak coproducts), this reflection trick does no longer work.  A
 semantic value of binary sum type is either a left or a right
@@
 interpreter for each of the possible answers.
-A monadic semantic value is thus a \emph{case tree} where the leaves
-are occupied by non-monadic values \cite{altenkirchUustalu:flops04}.
-In this article we only consider weak sum types, producing non-unique
-normal forms.  In particular, case trees are not normalized, allowing,
+A monadic semantic value is thus a \emph{case tree} where the inner
+nodes are labeled with unknowns and the leaves
+with non-monadic values \cite{altenkirchUustalu:flops04}.
+
+In this article, we only consider weak sum types, lacking the universal
+property of coproducts.  As a consequence, terms with the same
+denotation may have different normal forms under NbE.
+In particular, case trees are not normalized, allowing,
 \eg, redundant case splits (\ie, asking the same question twice).
@@
 coproducts) with unique normal forms
-\cite{altenkirchDybjerHofmannScott:lics01,altenkirchUustalu:flops04,balatDiCosmoFiore:popl04,barral:PhD,scherer:popl17}.
+\cite{altenkirchDybjerHofmannScott:lics01,altenkirchUustalu:flops04,balatDiCosmoFiore:popl04,barral:PhD,scherer:popl17},
+\ie, where NbE decides equality of denotations.
 
@@
 %
-In contrast to the normal forms of CBN NbE, which is the algorithmic
+In contrast to the normal forms of CBN NbE---which is the algorithmic
 counterpart of the completeness proof for intuitionistic propositional
-logic (IPL) using Beth models, CBPV NbE gives us more restrained
+logic (IPL) using Beth models---CBPV NbE gives us more restrained
 normal forms, where the production of a value via injections cannot be
@@
 In our presentation of the various lambda calculi we ignore the
-concrete syntax, only consider the abstract syntax obtained by the
+concrete syntax, only considering the abstract syntax obtained by the
 Curry-Howard-Isomorphism.  A term is simply a derivation tree whose
@@
 Our design choices were guided by an Agda formalization of sections 2
-(complete, see \url{https://andreasabel.github.io/ipl/html/NfModelMonad.html})
+(complete, see
+\url{https://andreasabel.github.io/ipl/html/NfModelMonad.html}),
+3 (in a variantion, see \url{https://andreasabel.github.io/ipl/html-cbpv/NfCBPV.html})
 and 4 (partial, see \url{https://andreasabel.github.io/ipl/html-focusing/Polarized.html}).
@@
 as we shall not be concerned with considering
-concrete terms in this article.
+\emph{concrete} terms in this article.
 
@@
 % \nofbox{$\downG A : \Den A \Gamma \to \Nf A \Gamma$}.
-\emph{Normalization by evaluation} (NbE)
+The method
+\emph{normalization by evaluation} (NbE)
 %\fbox{\(
@@
 takes the role of the \emph{oracle} alluded to in the introduction,
-and offers us these services:
+and offers us the following services:
 % \begin{enumerate}
@@
 a normal form, $\C\,(\tNf\,P)\,\Gamma$, and delivers a normal form
-$\Nf P \Gamma$.  However, $\trunNf[\C]$ is not just the inverse of
-$\treturn[\C]$; instead, the information contained in $\C$ is reified and
+$\Nf P \Gamma$.
+% However, $\trunNf[\C]$ is not just the inverse of
+% $\treturn[\C]$; instead, the information contained in $\C$ is reified and
+% becomes part of the normal form delivered by $\trunNf[\C]$.
+While $\trunNf[\C]$ is a left inverse of
+$\treturn[\C]$, the converse is not true.
+Instead, the information contained in $\C$ is reified and
 becomes part of the normal form delivered by $\trunNf[\C]$.
+Technically, $(\tNf\,P, \trunNf[\C])$ is a
+\emph{monad algebra} for monad $\C$.
+Besides $\trunNf[\C] \circ \treturn[\C] = \tid_{\tNf\,P}$, the square
+$\trunNf[\C] \circ \tjoin[\C] = \trunNf[\C] \circ
+ \tmap[\C]\trunNf[\C]$ stands.  This means that given a case tree whose
+leaves are again case trees ending in normal forms, it does not
+matter whether we first join the nested case trees and then assemble
+the normal form, or whether we first turn the inner case trees into
+normal forms via $\tmap[\C]\trunNf[\C]$, and then the outer one.
 
@@
 case of a right injection.
-
-The similarity of the monad services $\tabort^\C$ and $\tcase^\C$ to
-the corresponding constructors for normal forms, or their
-generalization of \Cref{rem:abort}, is hard to miss.
-Unsurprisingly, just \emph{case trees} are a first instance of a cover monad,
-% To make things concrete,
-% we shall immediately construct an instance of such a cover monad:
-the free cover monad
-$\Cov$ defined
-%inductively by
-as an inductive family with constructors
-$\treturn[\Cov]$, $\tabort^{\Cov}$, and
-$\tcase^{\Cov}$.
-% TODO: Shouldn't this better be trees with nullary and binary nodes
-% and leaves always an element of A?
-One can visualize an element $c : \Cov\,\A\,\Gamma$ as a
-binary case tree whose inner nodes ($\tcase$) are labeled by neutral terms
-of sum type $A_1+A_2$ and its two branches by the context extensions $A_1$
-and $A_2$, resp.  Leaves are either
-labeled by a neutral term of empty type $\tyzero$ (see $\tabort$),
-or by an element of $\A$ (see $\treturn$).  Functoriality amounts to
-replacing the labels of the $\treturn$-leaves, and the monadic bind
-(aka Kleisli extension) replaces
-these leaves by further case trees.  Bind is realized via $\tjoin[\Cov]$,
-which flattens a 2-level case tree, \ie, a case tree with case trees as
-leaves, into a single one.
-Finally $\trunNf[\Cov]$ is a simple recursion on the tree, replacing $\tcase^{\Cov}$
-and $\tabort^{\Cov}$
-by the $\tcase$ and $\tabort$ constructions on normal forms, and
-$\treturn[\Cov]$ by the identity.
-
-% the following constructions:
-% \begin{gather*}
-%   \nru{\treturn[\Cov]}{\A\,\Gamma}{\Cov\,\A\,\Gamma}
-% \end{gather*}
-
-Using the services of a generic cover monad $\C$, we can complete our semantics:
-%\begin{gather*}
+\newcommand{\figdenpos}{
+\begin{figure}[bthp]
+\hrulefill
 \[
@@
 \]
+\hrulefill
+  \caption{Interpretation of positive types.}
+  \label{fig:denpos}
+\end{figure}
 % \[
@@
 % \]
-%\end{gather*}
+} % END \newcommand
+\figdenpos
+
+The similarity of the monad services $\tabort^\C$ and $\tcase^\C$ to
+the corresponding constructors for normal forms, or their
+generalization of \Cref{rem:abort}, is hard to miss.
+Unsurprisingly, just \emph{case trees} are a first instance of a cover monad:
+the free cover monad
+$\Cov$ defined
+as an inductive family with constructors
+$\treturn[\Cov]$, $\tabort^{\Cov}$, and
+$\tcase^{\Cov}$.
+% TODO: Shouldn't this better be trees with nullary and binary nodes
+% and leaves always an element of A?
+One can visualize an element $c : \Cov\,\A\,\Gamma$ as a
+binary case tree whose inner nodes ($\tcase$) are labeled by neutral terms
+of sum type $A_1+A_2$ and its two branches by the context extensions $A_1$
+and $A_2$, resp.  Leaves are either
+labeled by a neutral term of empty type $\tyzero$ (see $\tabort$),
+or by an element of $\A$ (see $\treturn$).  Functoriality amounts to
+replacing the labels of the $\treturn$-leaves, and the monadic bind
+(aka Kleisli extension) replaces
+these leaves by further case trees.  Bind is realized via $\tjoin[\Cov]$,
+which flattens a 2-level case tree, \ie, a case tree with case trees as
+leaves, into a single one.
+Finally, $\trunNf[\Cov]$ is a simple recursion on the tree, replacing $\tcase^{\Cov}$
+and $\tabort^{\Cov}$
+by the $\tcase$ and $\tabort$ constructions on normal forms, and
+$\treturn[\Cov]$ by the identity.
+
+Using the services of a generic cover monad $\C$, we can complete our
+semantics, see \Cref{fig:denpos}.
 
 Since positive types $P$ have a monadic interpretation, there is a
-natural transformation $\C\,\den P \todot \den P$, which is simply
+monad algebra %natural transformation
+$\C\,\den P \todot \den P$, which is simply
 $\tjoin[\C]$.  It
 can be extended to a
-natural transformation $\trun^A : \C\,\den A \todot \den A$ for any simple
+monad algebra %natural transformation
+$\trun^A : \C\,\den A \todot \den A$ for any simple
 type $A$, meaning we can \emph{run} the monad,
-pushing its effects into $\den A$.%
 \footnote{In categorical terminology, the existence of
@@
 aka \emph{weak pasting}.}
-% All semantic types fulfill the weak sheaf condition aka weak pasting,
-% meaning there is
-% a natural transformation $\trun^A : \C\,\den A \todot \den A$ for any simple
-% type $A$.  In other words, we can \emph{run} the monad, pushing its
-% effects into $\den A$.
+pushing its effects into $\den A$.%
 We proceed by induction on $A$.
@@
 % \fbox{$\dent {t : A \from \Gamma} \_ : \den \Gamma \todot \den A$}
-can now be defined by recursion on $t$.
+can now be defined by recursion on $t$, see \Cref{fig:eval}.
 % using an auxiliary definition
@@
 Herein, the environment $\gamma$ lives in $\den \Gamma \Delta$, thus, $\dent t \gamma : \den A \Delta$.
-% Herein, $\gamma : \den \Gamma \Delta$, thus, $\dent t \gamma : \den A \Delta$.
+% Herein, $\gamma : \den \Gamma \Delta$, thus, $\dent t \gamma : \den
+% A \Delta$.
+\begin{figure}[htbp]
+\hrulefill
 \[
 \begin{array}{lcl}
+  \denty {t : A \from \Gamma}  & : & \den \Gamma \todot \den A \\
   \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
@@
 \]
-For the interpretation of the binders $\tabs$ and $\tcase$ we use the
-mutually defined $\fden{\_}$.
+\dotfill
 \[
@@
 \]
-The coproduct eliminations $\denabort$ and $\dencase$
-targeting an arbitrary semantic type $\den B$ are definable thanks to
-the weak sheaf property, \ie, the presence of pasting via $\trun^B$
-for any type $B$, and strong functoriality of $\C$.
+\dotfill
 \[
@@
 \]
+\hrulefill
+  \caption{Evaluation of terms.}
+  \label{fig:eval}
+\end{figure}
+For the interpretation of the binders $\tabs$ and $\tcase$ we use the
+mutually defined $\fden{\_}$.
+The coproduct eliminations $\denabort$ and $\dencase$
+targeting an arbitrary semantic type $\den B$ are definable thanks to
+the weak sheaf property, \ie, the presence of pasting via $\trun^B$
+for any type $B$, and strong functoriality of $\C$.
 It is noteworthy that the interpreter $\denty{\_}$ does not use any of
@@
 (presheaves) of neutrals, plug them together with the interpreter, and
-voila!, there is your normalizer!
+\emph{voila!}, there is your normalizer!
 Our insight here is that \emph{off-the-shelf interpreter} means
@@
 The answer type of this continuation monad is always $\tNf$, however,
-we are polymorphic in the positive type $P$ of normal forms we produce.%
+we are polymorphic in the positive type $P$ of normal forms we emit in
+the end.%
 
@@
 \begin{array}{l}
-  \tabort^{\CC}_\Gamma \, (u : \Ne \tyzero \Gamma) ~:~  \CC\,\B\,\Gamma \\[0.2ex]
+  \tabort^{\CC}_\Gamma \, (u : \Ne \tyzero \Gamma) ~:~  \CC\,\B\,\Gamma \\[0.4ex]
   \tabort^{\CC}_\Gamma
@@
 \]
+It is noteworthy (yet unsurprising) that $\tabort^\CC$ discards continuation $k$, while
+$\tcase^\CC$ uses it twice, once in each case.  Thus, normal forms can
+be of exponential size; for example, the normal form of the identity
+function over $n$-tuples of booleans has a case tree of height $n$,
+hence, size $\Theta(2^n)$.
+
 The NbE algorithm using $\CC$ is comparable to Danvy's type-directed
-partial evaluation \cite[Figure~8]{danvy:popl96}.  However, he uses
-shift-reset style continuations, which can be programmed in the
+partial evaluation \cite[Figure~8]{danvy:popl96}.  However, Danvy uses
+shift-reset style continuations, which can be expressed via the
 continuation monad, and relies on Scheme's \texttt{gensym} to produce
-fresh variables names rather than using the Kripke function space / presheaves.
+fresh variables names rather than presheaves and the Kripke function space.
 
@@
 working in type theory, which allows infinitely branching inductive
-types.   As a last difference, CBPV does not consider base types; in
+types.
+As a last difference, CBPV does not consider uninterpreted base types; in
 anticipation of the next section, we add them as both positive atoms
@@
 The multiplication $\tjoin[\C]$ of the strong monad $\C$ can be
-extended to a natural transformation
+extended to a monad algebra % natural transformation
 $\trun^N : \C \den N \todot \den N$ for negative types $N$.
@@
 % except
-Note
-that application is restricted to values.  Values of type
-$\Thunk N$ are embedded via $\tforce$.
+Note that function application is restricted to value arguments.
+Values of type $\Thunk N$ are embedded by $\tforce$.
 Further, values of type $P$ can be embedded
@@
 
-Interpretation of values
+\Cref{fig:evalcbpv} defines
+interpretation of values
 $\denty{ v : \PValTm P \Gamma } : \den \Gamma \todot \den P$
 and terms
-$\denty{ t : \Tm N \Gamma } : \den \Gamma \todot \den N$
-is straightforward, thanks to the pioneering work of
-Moggi \cite{moggi:infcomp91} and the work that
-Levy \cite{levy:hosc06}
-put into the design of CBPV.
-\[
-\begin{array}{lcl}
-  \\
+$\denty{ t : \Tm N \Gamma } : \den \Gamma \todot \den N$.
+% Interpretation of values
+% $\denty{ v : \PValTm P \Gamma } : \den \Gamma \todot \den P$
+% and terms
+% $\denty{ t : \Tm N \Gamma } : \den \Gamma \todot \den N$
+% is displayed in \Cref{fig:evalcbpv}.
+It is straightforward, thanks to the pioneering work of
+Moggi \cite{moggi:infcomp91} and the
+design of CBPV by Levy \cite{levy:hosc06}.
+\begin{figure}[htbp]
+\hrulefill
+\[
+\begin{array}{p{13ex}cl}
   \dent{\tvar\,x}\gamma & = & \tlookup\,x\,\gamma \\
@@
   \dent{\tthunk\,t}\gamma & = & \dent t \gamma \\
-% \end{array}
-% \quad
-% \begin{array}{lcl}
+\end{array}
+\]
+\dotfill
+\[
+\begin{array}{p{13ex}cl}
   \dent{\tabs\,t}\gamma
       & = & \fdent t \gamma \\
-  \dent{\tapp\,t\,v} \gamma & = &
-     \dent t \gamma\;\tid\;\dent v \gamma \\
   \dent{\nunit}\gamma & = & \ttempty \\
@@
     (\dent{t_1}\gamma,\ \dent{t_2}\gamma) \\
+\end{array}
+\]
+\dotfill
+\[
+\begin{array}{p{13ex}cl}
+  \dent{\tapp\,t\,v} \gamma & = &
+     \dent t \gamma\;\tid\;\dent v \gamma \\
   \dent{\prj i t}\gamma & = & \pi_i \dent t \gamma \\
@@
 \]
-Since $\tThunk$ serves only as an embedding of negative into positive
-types in our semantics,
-we interpret thunking and forcing by the identity.
-The eliminations for positive types deal now only with values, thus,
-need not reference the monad operations.
-\[
-\begin{array}{lcl}
+\dotfill
+\[
+\begin{array}{p{13ex}cl}
   \dent{\tsplit\,v\,t}\gamma & = & (\lambda\,(a_1,a_2).\
@@
 \]
+\dotfill
+\[
+\begin{array}{p{13ex}cl}
+  \dent{\tret\,v}\gamma & = & \treturn[\C] \; \dent v \gamma \\
+  \dent{\tbind\,u\,t}\gamma & = & \trun^\C\,
+    (\stmap[\C] \; \fdent t \gamma\; \dent u \gamma) \\
+\end{array}
+\]
+\hrulefill
+  \caption{Evaluation (CBPV).}
+  \label{fig:evalcbpv}
+\end{figure}
+Since $\tThunk$ serves only as an embedding of negative into positive
+types in our semantics,
+we interpret thunking and forcing by the identity.
+The eliminations $\tsplit$, $\tcase$ and $\tabort$
+for positive types deal now only with values, thus,
+need not reference the monad operations.
 The use of the monad is confined to $\tret$ and $\tbind$.
@@
 at any negative type $N$ for the interpretation of $\tbind$.
-\[
-\begin{array}{lcl}
-  \dent{\tret\,v}\gamma & = & \treturn[\C] \; \dent v \gamma \\
-  \dent{\tbind\,u\,t}\gamma & = & \trun^\C\,
-    (\stmap[\C] \; \fdent t \gamma\; \dent u \gamma) \\
-\end{array}
-\]
 
@@
 \end{gather*}
+% The analogue of $\trunNf$ from \Cref{sec:normstlc} would be a map
+% $\C\,(\tVNf\,P) \todot \tNf\,(\Comp P)$
 Note that we do not have a function
-$\trunNf[\Cov] : \Cov\,(\tNf\,N) \todot \tNf\,N$ this time, instead, we
-directly employ $\Cov$ in the definition of $\tNf$ at the base case
+$\trunNf[\Cov]$
+% $\trunNf[\Cov] : \Cov\,(\tNf\,N) \todot \tNf\,N$
+this time,\footnote{%
+A trivial $\trunNf[\Cov] : \Cov\,(\tNf\,N_0) \todot \tNf\,N_0$
+for $N_0 ::= \natom \mid \Comp P$ could be given by induction on the
+cover, essentially being a $\tjoin$.  However its extension to
+arbitrary negative types $N$ would fail for the case of pushing a
+binder ($\tbind$, $\tsplit$ or $\tcase$) under an $\tabs$traction.
+This is because
+our base category $\Cxt$ of order-preserving embeddings
+does not permit swapping of variables in the context.
+}
+instead, we
+directly employ $\Cov$ in the definition of $\tNf$ at the base cases
 $\natom$ and $\Comp P$.
@@
 % $\trunNf$.
-Negative reflection $\up N : \tNe\,N \todot \den N$ is as
-before generalized from variables to neutrals, to handle the breaking
+Negative reflection $\up N : \tNe\,N \todot \den N$ is, as
+before, generalized from variables to neutrals, to handle the breaking
 down of $N$ via eliminations.
@@
 called $\star$ by McBride and Paterson \cite[Section 7]{mcbridePaterson:applicative},
-which for monads $\C$ can be defined by
+which in our case can be defined by \eg\
 $c_1 \star c_2 =
-   \tjoin \; (\tmap\;(\lambda\,a_1.\
-   \tmap\;(\lambda\,a_2.\ (a_1,a_2))
-    \;c_2)
+   \tjoin \; (\stmap\;(\lambda\,\tau_1\,a_1.\
+   \stmap\;(\lambda\,\tau_2\,a_2.\ (\tren\;\tau_2\;a_1,\;a_2))
+    \;(\tren\;\tau_1\;c_2))
     \;c_1)$.
+
+% which for monads $\C$ can be defined by
 % $c_1 \star c_2 =
 %    \tjoin \; (\tmap\;(\lambda\,a_1.\
-%    \tjoin \; (\tmap\;(\lambda\,a_2.\ \treturn\,(a_1,a_2))
-%     \;c_2))
+%    \tmap\;(\lambda\,a_2.\ (a_1,a_2))
+%     \;c_2)
 %     \;c_1)$.
-% $c \star d = \tjoin \; (\tmap\;(\lambda\,b.\ (c,\, \treturn\,b))\;d$.
-% $\tzip\;c = \tjoin \comp \tmap\,(\lambda\,a.\ (c,\, \treturn\,a)$.
+% % $c_1 \star c_2 =
+% %    \tjoin \; (\tmap\;(\lambda\,a_1.\
+% %    \tjoin \; (\tmap\;(\lambda\,a_2.\ \treturn\,(a_1,a_2))
+% %     \;c_2))
+% %     \;c_1)$.
+% % $c \star d = \tjoin \; (\tmap\;(\lambda\,b.\ (c,\, \treturn\,b))\;d$.
+% % $\tzip\;c = \tjoin \comp \tmap\,(\lambda\,a.\ (c,\, \treturn\,a)$.
 
@@
         \lambda
-          \,_\Delta
+          % \,_\Delta
           \, \tau
@@
 \]
-The identity environment $\idenv \Gamma : \Cover {\den \Gamma} \Gamma$
+For reification of a (Kripke) function $f : \den{P \To N} \Gamma$ we extend context
+$\Gamma$ to $\cext \Gamma P$ and create a case tree
+$\tfresh^P_\Gamma : \Cover {\den P} {(\cext \Gamma P)}$ representing the new
+variable.  Using strong functoriality $\stmap$,
+we then apply $f$ to all leaves $a$ of that case tree,
+which may live in further extended contexts $\Delta$ reachable by
+$\tau : \cext \Gamma P \sublist \Delta$.
+The resulting case tree
+$\Cover {\den N} {(\cext \Gamma P)}$ is then $\trun$ giving us a value
+in $\den N {(\cext \Gamma P)}$, which can be reified to a
+$\Nf N {(\cext \Gamma P)}$.  Finally, abstraction gives the desired
+normal form in $\NfP {P \To N} \Gamma$.
+
+This time, the identity environment $\idenv \Gamma : \Cover {\den \Gamma} \Gamma$
 can only be generated in the monad, due to monadic positive
@@
 elimination rules for positive types.  As a consequence of the latter,
-variables are either of atomic or negative type $H ::= \patom \mid N$.
-%($\HTy$).
-Contexts $\Gamma,\Delta$ are lists of $H$s.
-%$\HTy$s.
-%\fbox{$X ::= \patom \mid P$}.
+variables are either of atomic or negative type.
+%  $H ::= \patom \mid N$.
+% %($\HTy$).
+% Contexts $\Gamma,\Delta$ are lists of $H$s.
+% %$\HTy$s.
+% %\fbox{$X ::= \patom \mid P$}.
+Types in contexts $\Gamma,\Delta$ are of one of the forms $\patom$ or $N$.
 
@@
 however, this coincidence shall not matter for our further considerations.
+
+\begin{remark}
+The notation $\Add P$ is chosen in analogy of the next-time operator
+of dynamic logic \citep{pratt:focs76}.
+The deeper reason is that the category $\Cxt$ can be seen as
+branching time where a context represents a point in time and an
+extension a possible future.  The modality $\Add P$ picks a certain
+future, the ``proposition'' $\Add P\,\J$ states that
+$\J$ should hold in the future determined by $P$.
+
+While in dynamic logic $P$ would be a regular expression, our positive
+types represent the fragment featuring choice $(\tyzero, +)$ and
+sequence $(1, \times)$.  The introduction rules for $\Add P\,\J$
+resemble the axioms of choice $(\nbranch, \bbranch)$
+and sequence $(\nsplit,\bsplit)$ in dynamic logic.
+\end{remark}
 
@@
 for every branch.  We implement
-\fbox{$\reflect P : (\den P \hatto \J) \todot \Add P \J$}
+$\reflect P : (\den P \hatto \J) \todot \Add P \J$
+%\fbox{$\reflect P : (\den P \hatto \J) \todot \Add P \J$}
 as a higher-order function expecting a continuation $k$ which is
 invoked for each generated branch with the semantic value of type $P$
-constructed for this branch.
+constructed for this branch, see \Cref{fig:posrefl}.
+\begin{figure}[htbp]
+\hrulefill
 \[
@@
 \begin{array}{l@{~}lcl}
+  \multicolumn 2 l {\reflect P} & : & (\den P \hatto \J) \todot \Add P \J \\
   \reflect {\patom} & k & = &
@@
 \]
+\hrulefill
+  \caption{Positive reflection (polarized lambda calculus).}
+  \label{fig:posrefl}
+\end{figure}
+
 Reflecting at a positive atomic type $\patom$ is the regular ending of a
@@
 Due to the absence of composite positive types in contexts,
-the identity environment $\idenv \Gamma$ can be built
+the identity environment
+$\idenv \Gamma$
+%$\idenv \Gamma : \den \Gamma \Gamma$
+can be built
 straightforwardly using negative reflection.
@@
 \begin{array}{lcl}
+  \idenv \Gamma & : & \den \Gamma \Gamma \\
   \idenv{\cempty} & = & \ttempty \\
@@
 Coquand.  Thanks to Hugo Herbelin for encouraging feedback and spotting some
-typos.  Thanks to the anonymous referees of a previous version of this
+typos.  Thanks to the anonymous referees of previous versions of this
 paper for the valuable feedback and remarks that helped improving the

### 2019-07-12 ppdp19/rebuttal-ppdp19.txt
[ ppdp19 ] rebuttal

First, we'd like to thank the reviewers for carefully reading our submission, spotting mistakes and suggesting improvements, and stimulating a discussion how to complete the picture.

We'd like to briefly answer some of the questions raised, to the extent that the rebuttal format permits it.

> ----------------------- REVIEW 1 ---------------------

> I would like to find a simple sentence that the Levy's  "A value is, a computation does"-mantra does not play a role in the NbE proof, because NbE naturally considers a more logical form of normalisation which also reduces inside abstraction or other normally reduction-blocking constructs (like thunks in case of CBPV).

Might add just your sentence, if you don't mind.

> The paper comes with a partial Agda formalization (fully done for simply-typed, not done for CBPV, partially done for polarized).

A formalization of NbE for pure CBPV (with Levy's original syntax) is now done, albeit in some variation to the development of section 3:

  https://andreasabel.github.io/ipl/html-cbpv/NfCBPV.html

> I would like to see a sentence or two what is exactly missing in the polarized formalization.

The current formalization covers normal forms, semantics, reflection and reification, but still lacks terms and evaluation.  We shall complete the formalization for the final version.

> 283 The spacing for Nf A Γ and Ne A Γ is chosen differently, which reads strange.
> 485 Shouldn't it be Ne 0 Γ here?

Thanks for spotting; fixed these.

> 840 - "Thunk has no semantic effect" reads a bit strange.

You are right; deleted this phrase.

> ----------------------- REVIEW 2 ---------------------

> * What are the relationships between these NbE algorithms? Each of these algorithms has a different notion of normal form, a different notion of context, and often a different source (either terms or cbpv). Is there some way to translate one reflect/reify algorithm into another?

Very good point, but we don't have the answers yet.  It seems like these algorithms are refinements of each other, for instance, there should be a translation from STLC (Λ⇒×+01) to CBPV and a reverse translation of their normal forms such that we get finer Nfs for STLC by going through CBPV, and similarly for CBPV and polarized Λ.  However, this is just speculation for now.

> I was especially curious about the different notions of context in play -- how do they relate? In particular, I've always wondered what exactly is the relationship between CBPV and focalization? The type structure is almost identical but the treatment of variables is almost exactly opposite.

If you decorate each negative hypothesis with a "Thunk" coercion, then each focalized context is also a CBPV context (but not vice versa).  The point we are trying to make is that focalization is a refinement of CBPV by also requiring eager splitting of non-atomic positive hypotheses.

> Could the case monad be decomposed further into a pair of adjoints?

Shall look into this.

> * I was not sure what was a theorem and what wasn't. The paper remarks that section 4 has not yet been formalized in Agda, but has it been proved on paper?

Yes, if you accept section 4 as a paper proof.

> * Some comment about how much bureaucracy there was in the Agda development would be useful. Presheaves on paper are pretty miserable -- does Agda help with this?

Yes, but notice that since we only define a function from terms to normal forms without soundness, i.e., a proof that term and normal form have the same denotation, we are not needing the category laws for the context category and the functor laws for our presheaves a lot.  A more serious test is version including soundness as formalized in

  https://andreasabel.github.io/ipl/html/NbeModel.html

Here we profit from Agda's REWRITE extension that allows us to turn e.g. the category laws into definitional equalities automatically handled by Agda.

> ----------------------- REVIEW 3 ---------------------

> * In section 3, a specific monad is constructed. Would an axiomatic approach
>    be possible to cover a range of models?

The specific monad here should be generalizable just as in section 2, to allow, for instance its implementation by the continuation monad.  Of course, we would still need this specific monad (or an inlining of it) for the definition of normal forms of CBPV.

> * L.969-972: This should be given by any strong monad

Well spotted, thank you!  Indeed, we need strong functoriality to define (a useful version of) it.
DONE

> * L.1169: Focused -> polarized

Fixed.

### 2020-07-02 ppdp19/ppdp19.tex
[ ppdp19 ] DOI etc.

@@
 %\usepackage{array}
-\newcommand\hmmax{0}
-\newcommand\bmmax{0}
+% The following commands are for package bm to shut off error "Too many alphabets"
+\newcommand\hmmax{0} % default 3
+\newcommand\bmmax{0} % default 4
 \usepackage{bm} % bold math
@@
 % These commands are for a PROCEEDINGS abstract or paper.
+%\setcopyright{acmlicensed}
+% \acmConference[PPDP 2019]{PPDP 2019: 21st International Symposium on Principles
+%   and Practice of Declarative Programming}{7-9 October 2019}{Porto, Portugal}
+% \acmBooktitle{PPDP 2019: 21st International Symposium on Principles
+%   and Practice of Declarative Programming, 7-9 October 2019, Porto, Portugal}
+
 \copyrightyear{2019}
 \acmYear{2019}
-\setcopyright{acmlicensed}
-\acmConference[PPDP 2019]{PPDP 2019: 21st International Symposium on Principles
-  and Practice of Declarative Programming}{7-9 October 2019}{Porto, Portugal}
-\acmBooktitle{PPDP 2019: 21st International Symposium on Principles
-  and Practice of Declarative Programming, 7-9 October 2019, Porto, Portugal}
-\acmPrice{0.00}
-\acmDOI{10.1145/1122445.1122456}
-\acmISBN{978-1-4503-9999-9/18/06}
-
-%
+\acmConference[PPDP '19]{Principles and Practice of Programming Languages 2019}{October 7--9, 2019}{Porto, Portugal}
+\acmBooktitle{Principles and Practice of Programming Languages 2019 (PPDP '19), October 7--9, 2019, Porto, Portugal}
+\acmPrice{15.00}
+\acmDOI{10.1145/3354166.3354168}
+\acmISBN{978-1-4503-7249-7/19/10}
+
 % These commands are for a JOURNAL article.

### 2020-07-02 popl21/popl21.tex
[ popl21 ] effect-graded CBPV: typing rules

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{5}
\acmNumber{POPL} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2021}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Note: Authors migrating a paper from PACMPL format to traditional
%% SIGPLAN proceedings format must update the '\documentclass' and
%% topmatter commands above; see 'acmart-sigplanproc-template.tex'.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

\usepackage{stmaryrd}
\usepackage{xspace}
\usepackage{xypic}

\usepackage{dashrule}
\newcommand{\dashruler}{\hdashrule[0.5ex]{\textwidth}{0.2pt}{1ex}}
\newcommand{\ruler}{\rule{\textwidth}{0.2pt}}

% % https://tex.stackexchange.com/questions/46828/how-to-highlight-important-parts-with-a-gray-background
% \usepackage[breakable, theorems, skins]{tcolorbox}
% \tcbset{enhanced}

% \DeclareRobustCommand{\graytextbox}[2][gray!20]{%
% \begin{tcolorbox}[   %% Adjust the following parameters at will.
%         breakable,
%         left=0pt,
%         right=0pt,
%         top=0pt,
%         bottom=0pt,
%         colback=#1,
%         colframe=#1,
%         width=\dimexpr\textwidth\relax,
%         enlarge left by=0mm,
%         boxsep=5pt,
%         arc=0pt,outer arc=0pt,
%         ]
%         #2
% \end{tcolorbox}
% }

% https://tex.stackexchange.com/questions/502652/define-tcolorbox-in-math-mode
\usepackage{tcolorbox}
\newtcbox{\grayboxtext}{on line,colback=gray!20,colframe=gray!20,size=fbox,arc=0pt,boxrule=0pt}
\newcommand{\graybox}[1]{\grayboxtext{$#1$}}

\newtheorem{remark}{Remark}

\begin{document}

%% Title information
\title[Graded CBPV]{Graded Call-By-Push-Value}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%% Author with single affiliation.
\author{Andreas Abel}
%\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{0000-0003-0420-4492}             %% \orcid is optional
\affiliation{
  \department{Department of Computer Science and Engineering}
  \institution{Gothenburg University}
  \streetaddress{Rännvägen 6b}
  \city{Göteborg}
%  \state{State1}
  \postcode{41296}
  \country{Sweden}
}
\email{andreas.abel@@gu.se}          %% \email is recommended

% %% Author with two affiliations and emails.
% \author{First2 Last2}
% \authornote{with author2 note}          %% \authornote is optional;
%                                         %% can be repeated if necessary
% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
% \affiliation{
%   \position{Position2a}
%   \department{Department2a}             %% \department is recommended
%   \institution{Institution2a}           %% \institution is required
%   \streetaddress{Street2a Address2a}
%   \city{City2a}
%   \state{State2a}
%   \postcode{Post-Code2a}
%   \country{Country2a}                   %% \country is recommended
% }
% \email{first2.last2@inst2a.com}         %% \email is recommended
% \affiliation{
%   \position{Position2b}
%   \department{Department2b}             %% \department is recommended
%   \institution{Institution2b}           %% \institution is required
%   \streetaddress{Street3b Address2b}
%   \city{City2b}
%   \state{State2b}
%   \postcode{Post-Code2b}
%   \country{Country2b}                   %% \country is recommended
% }
% \email{first2.last2@inst2b.org}         %% \email is recommended

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Call-by-push-value (CBPV) is a simply typed lambda calculus that polarizes
types into value and computation types and can thus express both
call-by-name and call-by-value evaluation in the presence of effects.
Semantically, effects are modeled by a monad, and computation types as
algebras over this monad.
Effect type systems usually express more information than the presence
of an effect; typically effects are structured as pre-ordered monoid
where the monoid operation represents accumulation of effects and the
order subsumption, in analogy to subtyping.
In this work, a \emph{graded} version of is presented where the typing
of computations likens effect typing.  Semantically, computation types
are then represented as graded monad algebras.
Further we present a version of CBPV which additionally has graded coeffects.
\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
 \begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003752.10003790.10011740</concept_id>
<concept_desc>Theory of computation~Type theory</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003752.10010124.10010125.10010130</concept_id>
<concept_desc>Theory of computation~Type structures</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003752.10010124.10010138.10010142</concept_id>
<concept_desc>Theory of computation~Program verification</concept_desc>
<concept_significance>300</concept_significance>
</concept>
<concept>
<concept_id>10003752.10010124.10010131.10010134</concept_id>
<concept_desc>Theory of computation~Operational semantics</concept_desc>
<concept_significance>100</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Type theory}
\ccsdesc[500]{Theory of computation~Type structures}
\ccsdesc[300]{Theory of computation~Program verification}
\ccsdesc[100]{Theory of computation~Operational semantics}
%% End of generated code

%% Keywords
%% comma separated list
\keywords{effects, coeffects, call-by-name, call-by-value, linear types}  %% \keywords are mandatory in final camera-ready submission

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle

\section{Introduction}

Call-by-push-value (CBPV) is a simply typed lambda calculus that polarizes
types into value and computation types and can thus express both
call-by-name and call-by-value evaluation in the presence of effects.
Semantically, effects are modeled by a monad, and computation types as
algebras over this monad.
Effect type systems usually express more information than the presence
of an effect; typically effects are structured as pre-ordered monoid
where the monoid operation represents accumulation of effects and the
order subsumption, in analogy to subtyping.
In this work, a \emph{graded} version of is presented where the typing
of computations likens effect typing.  Semantically, computation types
are then represented as graded monad algebras.
Further we present a version of CBPV which additionally has graded coeffects.

\section{An effect-graded version of CBPV}

%\newcommand{}{\ensuremath{}}
\newcommand{\ie}{\emph{i.e.}}
\newcommand{\eg}{\emph{e.g.}}
\newcommand{\Eg}{\emph{E.g.}}
\newcommand{\loccit}{\emph{loc.\,cit.}}
\newcommand{\den}[1]{\llbracket#1\rrbracket}
\newcommand{\dent}[1]{\llparenthesis#1\rrparenthesis}
\newcommand{\To}{\ensuremath{\Rightarrow}}
\newcommand{\bN}{\ensuremath{\mathbb{N}}}
\newcommand{\tEff}{\ensuremath{\mathsf{Eff}}}
\newcommand{\Eff}{\ensuremath{\mathsf{Eff}}}
\newcommand{\bu}{\ensuremath{\bullet}}
\newcommand{\Ge}{\ensuremath{\varepsilon}}
\newcommand{\GG}{\ensuremath{\Gamma}}
\newcommand{\GS}{\ensuremath{\Sigma}}
\newcommand{\CC}{\ensuremath{\mathcal{C}}}
\newcommand{\SET}{\mathsf{SET}}
\newcommand{\CPO}{\mathsf{CPO}}
\newcommand{\T}[1][]{\mathsf{T}_{#1}}
\newcommand{\trun}{\mathsf{run}}
\newcommand{\run}[1][]{\mathsf{run}_{#1}}
\newcommand{\tA}{\mathsf{A}}
\newcommand{\tB}{\mathsf{B}}
\newcommand{\A}{\mathsf{A}}
\newcommand{\B}{\mathsf{B}}
\newcommand{\tExc}{\mathsf{Exc}}
\newcommand{\Exc}{\mathsf{Exc}}
\newcommand{\String}{\mathsf{String}}
\newcommand{\StringLe}[1]{\mathsf{String}{\leq}\,#1}
\newcommand{\tfmap}{\mathsf{fmap}}
\newcommand{\treturn}{\mathsf{return}}
\newcommand{\tjoin}{\mathsf{join}}
\newcommand{\join}{\mathsf{join}}
\newcommand{\append}{\ensuremath{\mathbin{\mathtt{++}}}}
\newcommand{\tcast}{\mathsf{cast}}
\newcommand{\tid}{\mathsf{id}}

% Types
\newcommand{\tTy}{\mathsf{Ty}}
\newcommand{\Ty}[1]{\tTy^{#1}}
\newcommand{\PTy}{\Ty+}
\newcommand{\NTy}{\Ty-}
\newcommand{\extr}[2]{#1.#2}
\newcommand{\ext}[3]{#1.#2{:}#3}

\newcommand{\sumt}   [2]{\ensuremath{\Sigma_{#1} #2}}
\newcommand{\tupt}   [2]{\ensuremath{\otimes_{#1} #2}}
\newcommand{\rect}   [2]{\ensuremath{\Pi_{#1} #2}}
\newcommand{\sumty}  [3]{\ensuremath{\Sigma_{#1:#2} #3_{#1}}}
\newcommand{\tupty}  [3]{\ensuremath{\otimes_{#1:#2} #3_{#1}}}
\newcommand{\recty}  [3]{\ensuremath{\Pi_{#1:#2} #3_{#1}}}
\newcommand{\thunkty}[2]{\ensuremath{[#1]#2}}
\newcommand{\compty} [1]{\ensuremath{\diamond#1}}

% Terms
\newcommand{\tTm}{\mathsf{Tm}}
\newcommand{\Tm}[1]{\tTm^{#1}}
\newcommand{\PTm}{\Tm+}
\newcommand{\NTm}{\Tm-}
\newcommand{\tthunk}{\mathsf{thunk}}
\newcommand{\thunk}[1]{\tthunk\,#1}
\newcommand{\tforce}{\mathsf{force}}
\newcommand{\force}[1]{\tforce\,#1}
\newcommand{\tin}[1]{\mathsf{in}_{#1}}
\newcommand{\inj}[2]{\tin{#1}\,#2}
\newcommand{\ttup}[1]{\mathsf{tup}^{#1}}
\newcommand{\tup}[2]{\ttup{#1}\,{#2}}
\newcommand{\ptup}[1]{\ttup\,{#1}}
% \newcommand{\ptup}[1]{\ttup{+}\,{#1}}
\newcommand{\ntup}[1]{\ttup{-}\,{#1}}
\newcommand{\tret}{\mathsf{ret}}
%\newcommand{\vlet}[3]{#1\,{=}\,#2;\,#3}
\newcommand{\ret}[1]{\tret\,#1}
\newcommand{\vlet}[3]{#2\,{\mathsf{be}}\,#1.\,#3}
% \newcommand{\let}[3]{#1\,{\leftarrow}\,#2;\,#3}
\newcommand{\bind}[3]{#2\,{\mathsf{to}}\,#1.\,#3}
% \newcommand{\bind}[3]{\ret #1\,{\leftarrow}\,#2;\,#3}
\newcommand{\tsplit}{\mathsf{split}}
\newcommand{\splits}[3]{#1\,{\tsplit}\,#2.\,#3}
\newcommand{\tcases}{\mathsf{cases}}
\newcommand{\case}[2]{#1\,{\tcases}\,\{#2\}}
\newcommand{\caser}[2]{#1\,{\tcases}\,#2}
% \newcommand{\split}[3]{\ptup #1\,{\leftarrow}\,#2;\,#3}
\newcommand{\lam}[2]{\lambda#1.\,#2}
\newcommand{\app}[2]{#1\,#2}
\newcommand{\trecord}{\mathsf{record}}
\newcommand{\recrd}[2]{\trecord_{#1}\,#2}
\newcommand{\record}[2]{\trecord_{#1}\{#2\}}
\newcommand{\tproj}{\mathsf{proj}}
\newcommand{\proj}[2]{\tproj_{#1}\,{#2}}

% Judgements
\newcommand{\ValTy}[3]{#1 \vdash #2 : #3}
\newcommand{\CompTy}[4]{#1 \vdash #2 : #4 \mid #3}
% \newcommand{\CompTy}[4]{#1 \vdash #2 \mathrel{{:}\langle #3 \rangle} #4}
\newcommand{\ru}{\dfrac}
\newcommand{\nru}[3]{#1\;\dfrac{#2}{#3}}
\newcommand{\nrux}[4]{#1\;\dfrac{#2}{#3}\;#4}

\newcommand{\rulename}[1]{\ensuremath{\mbox{\textsc{#1}}}\xspace}
\newcommand{\rintro}[1]{\ensuremath{\mathord{#1}\mbox{-\rulename{intro}}}}
\newcommand{\relim}[1]{\ensuremath{\mathord{#1}\mbox{-\rulename{elim}}}}
\newcommand{\rvar}{\rulename{var}}
\newcommand{\rlet}{\rulename{let}}
\newcommand{\rsub}{\rulename{sub}}

As effect algebra, we assume a pre-ordered monoid
$(\Eff,\bu,\Ge,\leq)$ such that $\_{\bu}\_$ is monotone wrt.\ the
preorder $\leq$ in both arguments.  The unit $\Ge$ shall mean \emph{no
  effect} and the operation $\_{\bu}\_$ serves to accumulate effects,
possibly in a sequential order---unless the monoid is commutative.  The
preorder represents effect subsumption, i.e., loss in precision of the
effect analysis.  Note that the unit $\Ge$ is not necessarily the least element
wrt.\ $\leq$.

\subsection{Recapitulation: modelling effects via monads}

Consider a categorical model $\CC$ of the simply-typed lambda calculus
(STLC), i.e., where types $\tau$ and contexts $\GG$ are interpreted as
objects $\den \tau$ and $\den \Gamma$ of $\CC$ and terms
$\GG \vdash t : \tau$ as morphisms $\dent t \in \CC(\den\GG,\den\tau)$.
Such a category could be $\SET$, interpreting types as sets and terms
as functions, mapping the valuation of their free variables to their
value, or $\CPO$, interpreting types as complete partial orders and
terms as monotone functions, or another model of the STLC.
Typically $\CC$ is a cartesian-closed category, i.e., has products
$A_1 \times A_2$ of objects modelling product types and exponentials $A \To
B$ modeling function types.  Further $\CC$ maybe be distributive,
\ie, have coproducts $A_1 + A_2$ that distribute over products, to
model variants aka disjoint sum types.

Recall that effects are modeled by a suitable monad $\T$ in
$\CC$.  An example would be the \emph{writer monad} $\T\,A = \String \times
A$ that models the effect \emph{output}.
Another example would be the \emph{exception monad} $\Exc + \_$ where
the result of a computation is either an exception $e : \Exc$ or a
regular result.
The monad operations, with
their implementation for writer in $\SET$, are the following:
\[
\begin{array}{lcl}
\tfmap & : & (A \to B) \to \T\,A \to \T\,B \\
\tfmap\, f\, (s,\; a) & = & (s,\, f\, b)
\\[1ex]
\treturn & : & A \to \T\,A \\
\treturn\, a & = & (\verb|""|, a)
\\[1ex]
\tjoin & : & \T\,(\T\,A) \to \T\,A \\
\tjoin\,(s_1,(s_2,a)) & = & (s_1 \append s_2,\; a) \\
\end{array}
\]
Herein, \verb|""| shall denote the empty string and $\append$ string
concatenation.

We assume familiarity with the functor laws $\tfmap\,\tid = \tid$ and
$\tfmap\,(f \circ g) = \tfmap\,f \circ \tfmap\,g$.
The three equational
laws of $\tjoin$ can be visualized compactly in the following
commutative diagram.
\[
\xymatrix@C=12ex{
  \T\,A     \ar[r]^{\treturn} \ar@{=}[dr]^{\tid}
& \T\,(\T\,A) \ar[d]^{\tjoin}
& \T\,(\T\,(\T\,A)) \ar[l]_{\tjoin} \ar[d]^{\tfmap\,\tjoin}
\\
& \T\,A       \ar@{=}[dr]^{\tid}
& \T\,(\T\,A) \ar[l]_{\tjoin}
\\
&
& \T\,A \ar[u]_{\tfmap\,\treturn}
}
\]

\subsection{Effect algebras and graded monads}

We can obtain more information about effects using a monad $\T[e]$
\emph{graded} over elements $e$ of a suitable effect algebra $\Eff$.
Say we want to track an upper bound on the length of the output
produced by a program.  To this end, we can use the preordered monoid
$\Eff = \bN \cup \{\infty\}$ under addition $\_{\bu}\_ = \_{+}\_$
with unit $\Ge = 0$ and the
natural order $\leq$.  The effect $\infty$ then denotes unbounded output, or
output whose length we cannot track in the type system (\eg, when it
depends on some variable).  Increasing the upper bound along $\leq$
means loss of precision of our analysis, with $\infty$ the least
precise information, meaning no upper bound.  The corresponding graded
writer monad is $\T[e]\,A = (\StringLe e) \times A$ where the output is an
element of $\StringLe e$, a string of length at most $e \in \bN \cup
\{\infty\}$.  The operations of a graded monad are, again given with
their $\SET$-implementation for the graded writer:
\[
\begin{array}{lcl}
\tfmap & : & (A \to B) \to \T[e]\,A \to \T[e]\,B \\
\tfmap\, f\, (s,\; a) & = & (s,\, f\, b)
\\[1ex]
\treturn & : & A \to \T[\Ge]\,A \\
\treturn\, a & = & (\verb|""|, a)
\\[1ex]
\tjoin & : & \T[e_1]\,(\T[e_2]\,A) \to \T[e_1 \bu e_2]\,A \\
\tjoin\,(s_1,(s_2,a)) & = & (s_1 \append s_2,\; a)
\\[1ex]
\tcast & : & \T[e] \to \T[e'] \mbox{ for } e \leq e'\\
\tcast & = & \tid \\
\end{array}
\]
A graded version of the exception monad would use effect algebra
$\Eff = \P(\Exc)$ under union and subset; an effect $e$ is a set of
possible exceptions thrown by a computation.\footnote{An example of a
  exception-tracing type system is the Java language.}

The interesting laws for graded monads are given by the following
commutative diagrams.
\[
\xymatrix@C=8ex{
  \T[e]\,A     \ar[r]^{\treturn} \ar@{=}[dr]^{\tid}
& \T[\Ge]\,(\T[e]\,A) \ar[d]^{\tjoin}
\\
& \T[e]\,A
}
\quad
\xymatrix@C=10ex{
  \T[e]\,A     \ar[r]^{\tfmap\,\treturn} \ar@{=}[dr]^{\tid}
& \T[e]\,(\T[\Ge]\,A) \ar[d]^{\tjoin}
\\
& \T[e]\,A
}
\quad
\xymatrix@C=8ex{
  \T[e_1 \bu e_2]\,(\T[e_3]\,A) \ar[d]^{\tjoin}
& \T[e_1]\,(\T[e_2]\,(\T[e_3]\,A)) \ar[l]_{\tjoin} \ar[d]^{\tfmap\,\tjoin}
\\
  \T[e_1 \bu e_2 \bu e_3]\,A
& \T[e_1]\,(\T[e_2 \bu e_3]\,A) \ar[l]_{\tjoin}
}
\]
Further, $\tcast$ commutes with $\tfmap$, namely $\tfmap\,f \circ
\tcast = \tcast \circ \tfmap\,f$, and $\tjoin$, in two ways with
the latter: $\tjoin \circ \tcast = \tcast \circ \tjoin$ and $\tjoin
\circ \tfmap\,\tcast = \tcast \circ \tjoin$.

\subsection{CBPV, monad algebras, and their graded version}

Call-by-push value \cite{levy:hosc06} is a refinement of Moggi's
computational lambda-calculus \cite{moggi:infcomp91} that allows effects
not only in monadic types, \ie, in objects $\T\,A$, but more generally
in computation types.  These correspond to monad algebras for $\T$ in $\CC$,
aka $\T$-algebras.
Those are objects $B$ together with a morphism $\run[B] : \T\,B
\to B$ that allows to formally \emph{run} the monad, ``merging'' its
effects into $B$.
The prime example of a monad algebra is a monadic
type, because $\run[\T\,A] : \T\,(\T\,A) \to \T\,A$ is simply
$\tjoin$.  Levy \citeyearpar{levy:hosc06} shows that monad algebras are
closed under products and exponentiation with arbitrary objects.  \Eg,
in $\SET$ we can define:
\[
\begin{array}{lcl}
%   \run[B_1 \times B_2] & : & \T\,(B_1 \times B_2) \to B_1 \times B_2 \\
%   \run[B_1 \times B_2]\, (s,(b_1,b_2)) & = & (\run[B_1] (s,b_1),\
%                                            \run[B_2] (s,b_2))
% \\[1ex]
  \run[B \times B'] & : & \T\,(B \times B') \to B \times B' \\
  \run[B \times B']\, (s,(b,b')) & = & (\run[B] (s,b),\
                                           \run[B'] (s,b'))
\\[1ex]
  \run[A \To B] & : & \T\,(A \To B) \to (A \To B) \\
  \run[A \To B]\, (s,f) & = & \lambda a.\; \run[B]\,(s,\,f\,a)
\\
\end{array}
\]
These definition implement \emph{lazy} effects that cannot be observed
at computation types such as $A \To B$, but only at value types; the
$\run[]$ of the monad algebra pushes the effects towards result types
that are eventually types of observable objects (values).

With $\run[]$ being a generalization of $\tjoin$, the laws for $\run[]$
are in analogy of those for $\tjoin$:
\[
\xymatrix@C=10ex{
  B     \ar[r]^{\treturn} \ar@{=}[dr]^{\tid}
& \T\,B \ar[d]^{\run[]}
& \T\,(\T\,B) \ar[l]_{\tjoin} \ar[d]^{\tfmap\,\run[]}
\\
& B
& \T\,B \ar[l]_{\run[]}
}
\]

In graded CBPV, monad algebras get replaced by \emph{graded monad
  algebras}.  Given a graded monad $\T$, a $\T$-algebra is a family of
objects $\left(B_e\right)_{e:\Eff}$ and morphisms
$\run[B] : \T[e_1]\,B_{e_2} \to B_{e_1 \bu e_2}$ satisfying these
laws:
\[
\xymatrix@C=10ex{
  B_e     \ar[r]^{\treturn} \ar@{=}[dr]^{\tid}
& \T[\Ge]\,B_e \ar[d]^{\run[]}
\\
& B_e
}
\qquad
\xymatrix@C=10ex{
  \T[e_1 \bu e_2]\,B_{e_3} \ar[d]^{\run[]}
& \T[e_1]\,(\T[e_2]\,B_{e_3}) \ar[l]_{\tjoin} \ar[d]^{\tfmap\,\run[]}
\\
  B_{e_1 \bu e_2 \bu e_3}
& \T[e_1]\,B_{e_2 \bu e_3} \ar[l]_{\run[]}
}
\]
Further the family $B$ should be functorial in the sense that there is
a family of coercion morphisms $B_{e_1 \leq e_2} : B_{e_1} \to B_{e_2}$ with
$B_{e \leq e} = \tid$ and $B_{e_2 \leq e_3} \circ B_{e_1 \leq e_2} =
B_{e_1 \leq e_3}$.

Graded monad algebras are as well closed under pointwise products and
exponentiation with objects, here the graded writer example:
\[
\begin{array}{lcl}
  \run[B \times B'] & : & \T[e_1]\,(B \times B')_{e_2} \to (B \times
                          B')_{e_1 \bu e_2} \\
  \run[B \times B']\, (s,(b,b')) & = & (\run[B] (s,b),\
                                           \run[B'] (s,b'))
\\[1ex]
  \run[A \To B] & : & \T[e_1]\,(A \To B_{e_2}) \to (A \To B_{e_1 \bu e_2}) \\
  \run[A \To B]\, (s,f) & = & \lambda a.\; \run[B]\,(s,\,f\,a)
\\
\end{array}
\]

\begin{figure}[htbp]
\flushleft
\ruler{}
Types.
\[
\begin{array}{lllrl@{\qquad}l}
\PTy & \ni & P
  & ::= & \graybox{\thunkty e N}
     \mid o \mid \sumty i I P \mid \tupty i I P
  & \mbox{Value types (positive types)} \\
\NTy & \ni & N
  & ::= & \compty P
     \mid P \To N \mid \recty i I N
  & \mbox{Computation types (negative types)} \\
\end{array}
\]
\dashruler{}
Terms.
\[
\begin{array}{lllrl@{\qquad}l}
\PTm & \ni & v,w
  & ::= & x
     \mid \thunk t
     \mid \inj i v
     \mid \ptup{\bar v}
  & \mbox{Values (positive terms)} \\
\NTm & \ni & t,u
  & ::= &
         \vlet x v t
  & \mbox{Computations (negative terms):}
\\ &&& \mid &
        \force v
    \mid \case v {\overline{x.\,t}}
    \mid \splits v {\bar x} t
  & \mbox{value eliminations}
\\ &&& \mid &
         \ret v        \mid \bind x u t
  & \mbox{monad operations}
\\ &&& \mid &
         \lam x t      \mid \app t v
  & \mbox{functions}
\\ &&& \mid &
         \trecord\{\overline{i:t}\} % \ntup{\bar t}
       \mid \proj i t
  & \mbox{lazy tuples (records)}
\end{array}
\]
\dashruler{}
Value typing \fbox{$\ValTy \Gamma v P$}\,.
\begin{gather*}
 \nru{\rvar}
     {x{:}P \in \Gamma
    }{\ValTy \Gamma x P}
\qquad
 \nru{\rintro\Box}
     {\CompTy \Gamma t {\graybox e} N
    }{\ValTy \Gamma {\thunk t} {\thunkty {\graybox e} N}}
\\[1.5ex]
 \nru{\rintro\GS}
     {\ValTy \Gamma v {P_i}
    }{\ValTy \Gamma {\inj i v} {\sumt I P}}
\qquad
 \nru{\rintro\otimes}
     {\forall i{:}I, \ \ValTy \Gamma {v_i} {P_i}
    }{\ValTy \Gamma {\ptup v} {\tupt I P}}
    % }{\ValTy \Gamma {\ptup{\overline{v_i}^{i:I}}} {\tupt I P}}
%    }{\ValTy \Gamma {\ptup (v_i)_{i:I}} {\tupt I P}}
\end{gather*}
\dashruler{}
Computation typing \fbox{$\CompTy \Gamma t {\graybox{e}} N$}\,.
\begin{gather*}
 \nru{\rlet}
     {\ValTy \Gamma v P \qquad
      \CompTy {\ext \Gamma x P} t e N
    }{\CompTy \Gamma {\vlet x v t} e N}
\qquad
 \nru{\relim\Box}
     {\ValTy \Gamma v {\thunkty {\graybox e} N}
    }{\CompTy \Gamma {\force v} {\graybox e} N}
\\[1.5ex]
 \nru{\relim\GS}
     {\ValTy \Gamma v {\sumt I P} \qquad
      \forall i{:}I,\ \CompTy {\ext \Gamma {x_i} {P_i}} {t_i} e N
    }{\CompTy \Gamma {\caser v {\{x_i.\, t_i\}_{i:I}}} e N}
\qquad
 \nru{\relim\otimes}
     {\ValTy \Gamma v {\tupt I P} \qquad
      \CompTy {\extr \Gamma {\overline{x_i{:}P_i}^{i:I}}} t e N
    }{\CompTy \Gamma {\splits v {\bar x} t} e N}
\\[1.5ex]
 \nru{\rintro\diamond}
     {\ValTy \Gamma v P
    }{\CompTy \Gamma {\ret v} {\graybox \Ge} {\compty P}}
\qquad
 \nru{\relim\diamond}
     {\CompTy \Gamma u {\graybox{e_1}} {\compty P} \qquad
      \CompTy {\ext \Gamma x P} t {\graybox{e_2}} N
    }{\CompTy \Gamma {\bind x u t} {\graybox{e_1 \bu e_2}} N}
\\[1.5ex]
 \nru{\rintro\To}
     {\CompTy {\ext \Gamma x P} t e N
    }{\CompTy \Gamma {\lam x t} e {P \To N}}
\qquad
 \nru{\relim\To}
     {\CompTy \Gamma t e {P \To N} \qquad
      \ValTy \Gamma v P
    }{\CompTy \Gamma {\app t v} e N}
\\[1.5ex]
 \nru{\rintro\Pi}
     {\forall i{:}I,\ \CompTy \Gamma {t_i} e {N_i}
    }{\CompTy \Gamma {\trecord {\{i : t_i\}_{i:I}}} e {\rect I N}}
    % }{\CompTy \Gamma {\ntup{\overline{t_i}^{i:I}}} e {\rect I N}}
\qquad
 \nru{\relim\Pi}
     {\CompTy \Gamma t e {\rect I N}
    }{\CompTy \Gamma {\proj i t} e  {N_i}}
\\[1.5ex]
\graybox{
 \nrux{\rsub}
      {\CompTy \Gamma t e N}
      {\CompTy \Gamma t {e'} N}
      {e \leq e'}
}
\end{gather*}
\rule{\textwidth}{0.2pt}
  \caption{Effect-graded call-by-push-value.}
  \label{fig:cbpv}
\end{figure}

\subsection{Effect-graded CBPV: syntax and typing}

The syntax and typing rules for effect-graded CBPV are given in
Fig.~\ref{fig:cbpv}.  The differences to pure CBPV are in
\graybox{\mbox{gray boxes}}.

Types are classified into value types $P \in \PTy$ ($A$ in
\citet{levy:hosc06}) and computation types ($\underline B$ in
\loccit).  These are positive and negative types in the terminology of
focusing.  Positive types are generated from base types $o$ via
disjoint sums $\sumty i I P$ with tag set $I$, eager products $\tupty
i I P$ of arity $I$ (composed
from $1$ and $A_1 \times A_2$ in \loccit) and thunking $\thunkty e N$
(written $U\,\underline B$ in \loccit).  In contrast to pure CBPV,
thunk types $\thunkty e N$ are annotated with an effect $e$ that can
be triggered when the thunk is forced.  Negative types are just as in
pure CBPV: monadic types $\compty P$ (written $F\,A$ in \loccit),
function types $P \To N$ (written $A \to \underline B$) in \loccit)
and record types $\recty i I N$ with label set $I$.
Records are \emph{lazy} tuples whose components are only computed by demand.
We abbreviate $\sumty i I P$ by its ``meta-level $\eta$-contraction''
$\sumt I P$; notations $\tupt I P$ and
$\rect I N$ are understood analogously.

Terms are separated into values $v \in \PTm$ and computations $t \in
\NTm$ and identical to pure CBPV, modulo changes in the concrete
syntax.  Values introduce positive types,
computations introduce
negative types and
eliminate both positive and negative types.
We use bars to indicate sequences, e.g., $\bar v$ for a
sequence of values, but drop the bar when the context of discourse
makes clear that we are dealing with sequences rather than single
objects.  For instance, ``$v_i$'' (where $i : I$) in the premise of
\rintro\otimes indicates that $v$ is a sequence of values with
elements $(v_i)_{i:I}$.  We may abbreviate $\trecord \{i :
t_i\}_{i:I}$ by $\recrd I t$ where $I$ is the label set and $t$ a
mapping from labels $i:I$ to terms $t_i$.

The meaning of the term constructors is best understood via their
typing.  Typing contexts $\Gamma$ are finite maps from variables $x$
to value types $P$, with $\ext \Gamma x P$ denoting the update of the
finite map $\Gamma$ at key $x$ with value $P$.

Value typing $\ValTy \Gamma v P$ is just as in pure CBPV, however,
computation typing $\CompTy \Gamma t e N$ also records effects $e :
\Eff$ potentially produced at runtime by computation $t$.  Thunking a
computation (rule \rintro\Box) stores the inferred effect classifier
$e$ in the thunk type $\thunkty e N$.

Effects are accumulated via the introduction and elimination rule for
the graded monad.  The unit $\ret v$ of the monad is effect-free
(\rintro\diamond), running this computation just produces the pure
value $v$.  Sequencing computations $u$ and $t$ via the \emph{bind}
construct $\bind x u t$ composes the effects $e_1$ of $u$ with the
effects $e_2$ of $t$ in that order.  The intuition is that first $u$
is run, producing effects classified by $e_1$, and its result is bound
to $x$ to run $t$, producing effects classified by $e_2$.  The sum of
the effects is classified by $e_1 \bu e_2$.

The other introduction and elimination rules are just as in pure
CBPV, except that they propagate the effect classifier $e$ from
hypothesis to conclusion.  Note that in case distinction (\relim\GS)
and record construction (\rintro\Pi) all subterms $t_i$ are required
to produce effects classified by the same $e$.  However, in reality,
different branches of a case distinction, for instance, may produce
very different effects.  To end up with a unique classifier $e$, the
branches may have to be typed using effect subsumption (\rsub).  In
fact, the uses of \rsub can be confined to the hypotheses of \relim\GS
and \rintro\Pi, except for a final invokation of \rsub at the very end
of the typing derivation.  Alternatively, we could have introduced
effect algebras with suprema $\sup_{i:I} e_i$ instead of a preorder $e
\leq e'$.  However, suprema might not always exist; by using
subsumption \rsub, we delegate the problem of partiality to the
construction of a typing derivation.

\begin{remark}\label{rem:cbpv}
  We recover pure CBPV from graded CBPV using the trivial effect
  algebra $\Eff = \{\Ge\}$.
\end{remark}

\begin{example}

\end{example}

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.

  This material is based upon work supported by the
  Swedish Research Council (Vetenskapsrådet)
  under Grant
  No.~2019-04216 \emph{Modal Dependent Type Theory}.

  % This material is based upon work supported by the
  % \grantsponsor{GS100000001}{National Science
  %   Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
  % No.~\grantnum{GS100000001}{nnnnnnn} and Grant
  % No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
  % conclusions or recommendations expressed in this material are those
  % of the author and do not necessarily reflect the views of the
  % National Science Foundation.
\end{acks}

%% Bibliography
\bibliography{medium}

%% Appendix
\appendix
\section{Appendix}

Text of appendix \ldots

\end{document}

### 2020-07-07 popl21/popl21.tex
[ popl21 ] graded comonads

@@
 \usepackage{tcolorbox}
-\newtcbox{\grayboxtext}{on line,colback=gray!20,colframe=gray!20,size=fbox,arc=0pt,boxrule=0pt}
+\newtcbox{\grayboxtext}{on line,colback=gray!20,colframe=gray!20,size=fbox,arc=0pt,boxrule=0pt}%,beforeafter skip=0pt,leftright skip=0pt}
 \newcommand{\graybox}[1]{\grayboxtext{$#1$}}
@@
 Effect type systems usually express more information than the presence
-of an effect; typically effects are structured as pre-ordered monoid
+of an effect; often effects are categorized by pre-ordered monoid
 where the monoid operation represents accumulation of effects and the
-order subsumption, in analogy to subtyping.
-In this work, a \emph{graded} version of is presented where the typing
+order expresses effect subsumption, in analogy to subtyping.
+In this work, a \emph{graded} version of CBPV is presented where the typing
 of computations likens effect typing.  Semantically, computation types
@@
 \newcommand{\loccit}{\emph{loc.\,cit.}}
-\newcommand{\den}[1]{\llbracket#1\rrbracket}
+\newcommand{\den}[2][]{\llbracket#2\rrbracket^{#1}}
 \newcommand{\dent}[1]{\llparenthesis#1\rrparenthesis}
 \newcommand{\To}{\ensuremath{\Rightarrow}}
+%\newcommand{\todot}{\stackrel\cdot\to}
+\newcommand{\todot}{\mathbin{\dot{\to}}}
 \newcommand{\bN}{\ensuremath{\mathbb{N}}}
+\newcommand{\dom}{\mathop{\mathrm{dom}}\nolimits}
+\newcommand{\Pot}[1]{\mathcal{P}\,#1}
+%\newcommand{\defiff}{:\iff}
+\newcommand{\defiff}{\mathrel{{{:}{\Longleftrightarrow}}}}
+
+\newcommand{\denp}{\den[+]}
+\newcommand{\denn}{\den[-]}
+\newcommand{\Denn}[2]{\denn{#1}_{#2}}
+
 \newcommand{\tEff}{\ensuremath{\mathsf{Eff}}}
@@
 \newcommand{\Ge}{\ensuremath{\varepsilon}}
+\newcommand{\Ga}{\ensuremath{\alpha}}
+\newcommand{\Gl}{\ensuremath{\lambda}}
+\newcommand{\Gr}{\ensuremath{\rho}}
 \newcommand{\GG}{\ensuremath{\Gamma}}
 \newcommand{\GS}{\ensuremath{\Sigma}}
+\newcommand{\R}{\mathsf{R}}
+\newcommand{\I}{\mathsf{I}}
+\newcommand{\LL}{\mathcal{L}}
+\newcommand{\W}{\mathsf{W}}
+\newcommand{\M}{\mathsf{M}}
+\newcommand{\wleq}{\sqsubseteq}
+\newcommand{\winf}{\sqcap}
+\newcommand{\wsup}{\sqcup}
+%\newcommand{\winf}{\bigsqcap}
+\newcommand{\emp}{0_{\W}}
+% \newcommand{\emp}{\varepsilon}
+% \newcommand{\emp}{\emptyset}
+
 \newcommand{\CC}{\ensuremath{\mathcal{C}}}
+\newcommand{\Hom}[2]{\CC\,(#1,\,#2)}
 \newcommand{\SET}{\mathsf{SET}}
@@
 \newcommand{\T}[1][]{\mathsf{T}_{#1}}
+\newcommand{\D}[1][]{\mathsf{D}_{#1}}
 \newcommand{\trun}{\mathsf{run}}
 \newcommand{\run}[1][]{\mathsf{run}_{#1}}
+\newcommand{\texpose}{\mathsf{expose}}
+\newcommand{\expose}[1][]{\texpose_{#1}}
 \newcommand{\tA}{\mathsf{A}}
@@
 \newcommand{\tfmap}{\mathsf{fmap}}
+\newcommand{\fmap}[2][]{\tfmap_{#1}\,#2}
 \newcommand{\treturn}{\mathsf{return}}
@@
 \newcommand{\tcast}{\mathsf{cast}}
+\newcommand{\Id}{\mathsf{Id}}
 \newcommand{\tid}{\mathsf{id}}
+\newcommand{\teval}{\mathsf{eval}}
+\newcommand{\tstrength}{\mathsf{strength}}
+\newcommand{\strengthl}{\tstrength^{\tl}}
+\newcommand{\strengthr}{\tstrength^{\tr}}
+\newcommand{\tl}{\mathsf{l}}
+\newcommand{\tr}{\mathsf{r}}
+\newcommand{\tcurry}{\mathsf{curry}}
+\newcommand{\comp}{\circ}
+\newcommand{\tweaken}{\mathsf{weaken}}
+\newcommand{\tcontract}{\mathsf{contract}}
 
@@
 \newcommand{\ext}[3]{#1.#2{:}#3}
+\newcommand{\qext}[5]{#1#2.#3#4{:}#5}
+\newcommand{\gqext}[5]{\graybox{#1}#2.\graybox{#3}#4{:}#5}
 
@@
 \newcommand{\thunkty}[2]{\ensuremath{[#1]#2}}
-\newcommand{\compty} [1]{\ensuremath{\diamond#1}}
+\newcommand{\compty} [2]{\ensuremath{\langle#1\rangle#2}}
 
@@
 \newcommand{\thunk}[1]{\tthunk\,#1}
+\newcommand{\tforced}{\mathsf{forced}}
 \newcommand{\tforce}{\mathsf{force}}
@@
 \newcommand{\caser}[2]{#1\,{\tcases}\,#2}
+\newcommand{\qlet}[4]{#3\,{\mathsf{be}}\,#1\,#2.\,#4}
+\newcommand{\qbind}[4]{#3\,{\mathsf{to}}\,#1\,#2.\,#4}
+\newcommand{\qsplits}[4]{#2\,{\tsplit}\,#1\,#3.\,#4}
+\newcommand{\qforced}[4]{#2\,{\tforced}\,#1\,#3.\,#4}
+\newcommand{\qcase}[3]{#2\,{\tcases}\,\{#1\,#3\}}
+\newcommand{\gqlet}[4]{#3\,{\mathsf{be}}\,\graybox{#1}#2.\,#4}
+\newcommand{\gqbind}[4]{#3\,{\mathsf{to}}\,\graybox{#1}#2.\,#4}
+\newcommand{\gqsplits}[4]{#2\,{\tsplit}\,\graybox{#1}#3.\,#4}
+\newcommand{\gqforced}[4]{#2\,{\tforced}\,\graybox{#1}#3.\,#4}
+\newcommand{\gqcase}[3]{#2\,{\tcases}\,\{\graybox{#1}#3\}}
+% \newcommand{\qlet}[4]{#1\,#3\,{\mathsf{be}}\,#2.\,#4}
+% \newcommand{\qbind}[4]{#1\,#3\,{\mathsf{to}}\,#2.\,#4}
+% \newcommand{\qsplits}[4]{#1\,#2\,{\tsplit}\,#3.\,#4}
+% \newcommand{\qforced}[4]{#1\,#2\,{\tforced}\,#3.\,#4}
+% \newcommand{\qcase}[3]{#1\,#2\,{\tcases}\,\{#3\}}
+% \newcommand{\qcaser}[3]{#1\,#2\,{\tcases}\,#3}
+% \newcommand{\gqlet}[4]{\graybox{#1}\,#3\,{\mathsf{be}}\,#2.\,#4}
+% \newcommand{\gqbind}[4]{\graybox{#1}\,#3\,{\mathsf{to}}\,#2.\,#4}
+% \newcommand{\gqsplits}[4]{\graybox{#1}\,#2\,{\tsplit}\,#3.\,#4}
+% \newcommand{\gqforced}[4]{\graybox{#1}\,#2\,{\tforced}\,#3.\,#4}
+% \newcommand{\gqcase}[3]{\graybox{#1}\,#2\,{\tcases}\,\{#3\}}
+% \newcommand{\gqcaser}[3]{\graybox{#1}\,#2\,{\tcases}\,#3}
 % \newcommand{\split}[3]{\ptup #1\,{\leftarrow}\,#2;\,#3}
@@
 \newcommand{\ValTy}[3]{#1 \vdash #2 : #3}
+\newcommand{\oCompTy}{\ValTy}
 \newcommand{\CompTy}[4]{#1 \vdash #2 : #4 \mid #3}
-% \newcommand{\CompTy}[4]{#1 \vdash #2 \mathrel{{:}\langle #3 \rangle} #4}
+\newcommand{\gpValTy}[4]{\graybox{(#1)}#2 \vdash #3 : #4}
+\newcommand{\gqValTy}[4]{\graybox{#1}#2 \vdash #3 : #4}
+\newcommand{\gpCompTy}[4]{\graybox{(#1)}#2 \vdash #3 : #4}
+\newcommand{\gqCompTy}[4]{\graybox{#1}#2 \vdash #3 : #4}
+\newcommand{\pValTy}[4]{(#1)#2 \vdash #3 : #4}
+\newcommand{\qValTy}[4]{#1#2 \vdash #3 : #4}
+\newcommand{\pCompTy}[4]{(#1)#2 \vdash #3 : #4}
+\newcommand{\qCompTy}[4]{#1#2 \vdash #3 : #4}
+%\newcommand{\CompTy}[4]{#1 \vdash #2 \mathrel{{:}\langle #3 \rangle} #4}
 \newcommand{\ru}{\dfrac}
@@
 \newcommand{\rulename}[1]{\ensuremath{\mbox{\textsc{#1}}}\xspace}
-\newcommand{\rintro}[1]{\ensuremath{\mathord{#1}\mbox{-\rulename{intro}}}}
-\newcommand{\relim}[1]{\ensuremath{\mathord{#1}\mbox{-\rulename{elim}}}}
+\newcommand{\rintro}[1]{\ensuremath{\mathord{#1}\mbox{-\rulename{intro}}}\xspace}
+\newcommand{\relim}[1]{\ensuremath{\mathord{#1}\mbox{-\rulename{elim}}}\xspace}
 \newcommand{\rvar}{\rulename{var}}
@@
 \newcommand{\rsub}{\rulename{sub}}
-
-As effect algebra, we assume a pre-ordered monoid
-$(\Eff,\bu,\Ge,\leq)$ such that $\_{\bu}\_$ is monotone wrt.\ the
-preorder $\leq$ in both arguments.  The unit $\Ge$ shall mean \emph{no
-  effect} and the operation $\_{\bu}\_$ serves to accumulate effects,
-possibly in a sequential order---unless the monoid is commutative.  The
-preorder represents effect subsumption, i.e., loss in precision of the
-effect analysis.  Note that the unit $\Ge$ is not necessarily the least element
-wrt.\ $\leq$.
+\newcommand{\rweak}{\rulename{weak}}
+
+% As effect algebra, we assume a pre-ordered monoid
+% $(\Eff,\bu,\Ge,\leq)$ such that $\_{\bu}\_$ is monotone wrt.\ the
+% preorder $\leq$ in both arguments.  The unit $\Ge$ shall mean \emph{no
+%   effect} and the operation $\_{\bu}\_$ serves to accumulate effects,
+% possibly in a sequential order---unless the monoid is commutative.  The
+% preorder represents effect subsumption, i.e., loss in precision of the
+% effect analysis.  Note that the unit $\Ge$ is not necessarily the least element
+% wrt.\ $\leq$.
 
 \subsection{Recapitulation: modelling effects via monads}
+\label{sec:ccc}
 
@@
 value, or $\CPO$, interpreting types as complete partial orders and
-terms as monotone functions, or another model of the STLC.
-Typically $\CC$ is a cartesian-closed category, i.e., has products
-$A_1 \times A_2$ of objects modelling product types and exponentials $A \To
+terms as monotone functions, or a presheaf model of the STLC etc.
+Typically, $\CC$ is a cartesian-closed category, i.e., has products
+$A_1 \times A_2$ of objects modelling product types and contexts,
+and exponentials $A \To
 B$ modeling function types.  Further $\CC$ maybe be distributive,
@@
 regular result.
-The monad operations, with
-their implementation for writer in $\SET$, are the following:
+The monad operations, together with
+their implementation for \emph{writer} in $\SET$, are the following:
 \[
@@
 \subsection{Effect algebras and graded monads}
-
-We can obtain more information about effects using a monad $\T[e]$
+\label{sec:graded_monad}
+
+Following \citet{katsumata:popl14}
+we can obtain more information about effects using a monad $\T[e]$
 \emph{graded} over elements $e$ of a suitable effect algebra $\Eff$.
+An effect algebra be a pre-ordered monoid
+$(\Eff,\bu,\Ge,\leq)$ such that $\_{\bu}\_$ is monotone wrt.\ the
+preorder $\leq$ in both arguments.  The unit $\Ge$ shall mean \emph{no
+  effect} and the operation $\_{\bu}\_$ serves to accumulate effects,
+possibly in a sequential order---unless the monoid is commutative.  The
+preorder represents effect subsumption, i.e., loss in precision of the
+effect analysis.  Note that the unit $\Ge$ is not necessarily the least element
+wrt.\ $\leq$.
+
 Say we want to track an upper bound on the length of the output
@@
 A graded version of the exception monad would use effect algebra
-$\Eff = \P(\Exc)$ under union and subset; an effect $e$ is a set of
+$\Eff = \Pot\Exc$ under union and subset; an effect $e$ is a set of
 possible exceptions thrown by a computation.\footnote{An example of a
@@
 Further, $\tcast$ commutes with $\tfmap$, namely $\tfmap\,f \circ
-\tcast = \tcast \circ \tfmap\,f$, and $\tjoin$, in two ways with
-the latter: $\tjoin \circ \tcast = \tcast \circ \tjoin$ and $\tjoin
+\tcast = \tcast \circ \tfmap\,f$, and in two ways with $\tjoin$, namely
+$\tjoin \circ \tcast = \tcast \circ \tjoin$ and $\tjoin
 \circ \tfmap\,\tcast = \tcast \circ \tjoin$.
@@
 \subsection{CBPV, monad algebras, and their graded version}
+\label{sec:graded_algebra}
 
@@
 effects into $B$.
-The prime example of a monad algebra is a monadic
-type, because $\run[\T\,A] : \T\,(\T\,A) \to \T\,A$ is simply
+The prime example of a monad algebra is simply a monadic
+type, because $\run[\T\,A] : \T\,(\T\,A) \to \T\,A$ is just
 $\tjoin$.  Levy \citeyearpar{levy:hosc06} shows that monad algebras are
 closed under products and exponentiation with arbitrary objects.  \Eg,
-in $\SET$ we can define:
+in $\SET$ we can define, for \emph{writer} algebras:
 \[
@@
 With $\run[]$ being a generalization of $\tjoin$, the laws for $\run[]$
-are in analogy of those for $\tjoin$:
+are in analogy of those for $\tjoin$ (if $B$ were $\T\,A$):
 \[
@@
 
-Graded monad algebras are as well closed under pointwise products and
-exponentiation with objects, here the graded writer example:
+Graded monad algebras are as well closed under
+pointwise products $(B \times B')_e = B_e \times B'_e$ and
+exponentiation with objects $(A \To B)_e = A \to B_e$;
+here the graded writer example:
 \[
@@
 \\[1ex]
-  \run[A \To B] & : & \T[e_1]\,(A \To B_{e_2}) \to (A \To B_{e_1 \bu e_2}) \\
+  \run[A \To B] & : & \T[e_1]\,(A \To B)_{e_2} \to (A \To B)_{e_1 \bu e_2} \\
+  % \run[A \To B] & : & \T[e_1]\,(A \To B_{e_2}) \to (A \To B_{e_1 \bu e_2}) \\
   \run[A \To B]\, (s,f) & = & \lambda a.\; \run[B]\,(s,\,f\,a)
@@
 \NTy & \ni & N
-  & ::= & \compty P
+  & ::= & \diamond P
      \mid P \To N \mid \recty i I N
@@
      {\ValTy \Gamma v P
-    }{\CompTy \Gamma {\ret v} {\graybox \Ge} {\compty P}}
+    }{\CompTy \Gamma {\ret v} {\graybox \Ge} {\diamond P}}
 \qquad
  \nru{\relim\diamond}
-     {\CompTy \Gamma u {\graybox{e_1}} {\compty P} \qquad
+     {\CompTy \Gamma u {\graybox{e_1}} {\diamond P} \qquad
       \CompTy {\ext \Gamma x P} t {\graybox{e_2}} N
@@
 
-Types are classified into value types $P \in \PTy$ ($A$ in
-\citet{levy:hosc06}) and computation types ($\underline B$ in
+Types are classified into value types $P \in \PTy$ (written $A$ in
+\citet{levy:hosc06}) and computation types $N \in \NTy$
+(written $\underline B$ in
 \loccit).  These are positive and negative types in the terminology of
-focusing.  Positive types are generated from base types $o$ via
+focusing \cite{zeilberger:PhD}.
+Positive types are generated from base types $o$ via
 disjoint sums $\sumty i I P$ with tag set $I$, eager products $\tupty
@@
 be triggered when the thunk is forced.  Negative types are just as in
-pure CBPV: monadic types $\compty P$ (written $F\,A$ in \loccit),
+pure CBPV: monadic types $\diamond P$ (written $F\,A$ in \loccit),
 function types $P \To N$ (written $A \to \underline B$) in \loccit)
@@
 
-\begin{example}
-
-\end{example}
+\subsection{Effect graded CBPV: denotational semantics}
+
+The denotational semantics of graded CBPV has been informally explained in
+sections \ref{sec:ccc} to \ref{sec:graded_algebra} already; in the following,
+we spell out the details.  We assume a distributive cartesian-closed
+category $\CC$ with a strong graded monad $\T : \Eff \to \CC \to \CC$.
+Let us agree on some notation for the constructions on objects and
+morphisms:
+\begin{itemize}
+\item Product $\prod_{i:I} A_i$ with projections $\pi_i : \CC(\prod_I
+  A, A_i)$ and tupling $\langle f_i
+  \rangle_{i:I} : \CC(C,\prod_i A)$ for $f_i : \CC(C,A_i)$.  Binary
+  products $\prod_{\{1,2\}} A$ are written $A_1 \times A_2$.
+\item Coproduct $\coprod_{i:I} A_i$ with injections $\iota_i :
+  \CC(A_i,\coprod_I A)$ and cotupling $[f_i]_{i:I} : \CC(\coprod_I A,
+  B)$ for $f_i : \CC(A_i,B)$.
+\item Exponential $A \To B$ with $\Lambda : \CC(C \times A, B) \to
+  \CC(C, A \To B)$ and $\teval : \CC((A \To B) \times A, B)$.
+\item Graded monad $\T$ with functoriality $\fmap[\T] f :
+  \CC(\T[e]A,\T[e]B)$ for $f : \CC(A,B)$,
+  unit $\treturn : \CC(A,\T[\Ge]A)$,
+  multiplication $\tjoin : \CC(\T[e_1](\T[e_2]A),\T[e_1 \bu e_2]A)$,
+  strength $\strengthl : \CC(A \times \T[e]B, \T[e](A \times B))$ and
+  coercion $\tcast : \CC(\T[e]A,\T[e']A)$ for $e \leq e'$.
+  Costrength $\strengthr : \CC(\T[e]A \times B, \T[e](A \times B))$
+  is derivable in the standard way.
+\end{itemize}
+
+\subsubsection{Interpretation of types}
+
+Positive types $P$ are interpreted as objects $\denp P$ of $\CC$,
+where a denotation $\denp o : \CC$ of base types $o$ is assumed.
+This interpretation lifts to contexts $\Gamma$ via $\denp \Gamma =
+\prod_{x:\dom(\Gamma)} \denp{\Gamma(x)}$.
+
+Negative types $N$ are interpreted
+as functors $\denn N : (\Eff,\leq) \to \CC$ mapping
+effect classifiers $e$ to objects $\Denn N e$ and effect subsumption $e
+\leq e'$ to morphisms $\Denn N {e \leq e'} : \Denn N e \to \Denn N {e'}$.
+\[
+\begin{array}{lcl}
+  \denp\_ & : & \PTy \to \CC \\
+  \denp{\thunkty e N} & = & \Denn N e \\
+  \denp{\sumt I P} & = & \coprod_{i:I} \denp{P_i} \\
+  \denp{\tupt I P} & = & \prod_{i:I} \denp{P_i} \\
+\end{array}
+\qquad\qquad
+\begin{array}{lcl}
+  \Denn\_\_ & : & \NTy \to \Eff \to \CC \\
+  \Denn{\diamond P} e & = & \T[e] \denp P \\
+  \Denn{P \To N}   e & = & \denp P \To \Denn N e \\
+  \Denn{\rect I N} e & = & \prod_{i:I} \Denn{N_i}e \\
+  % \denn{P \To N}   e & = & \denp P \To (\denn N e) \\
+  % \denn{\rect I N} e & = & \prod_{i:I} \left(\denn{N_i}e\right) \\
+\end{array}
+\]
+The graded $\T$-algebra structure $\run{\denn N}$ is constructed by
+induction on $N$:
+\[
+\begin{array}{lcl}
+  \run{\denn N} & : & \Hom {\T[e_1](\Denn N {e_2})} {\Denn N {e_1 \bu e_2}} \\
+  \run{\denn {\diamond P}} & = & \tjoin \\
+  \run{\denn {P \To N}}   & = & \Lambda(\run{\denn N} \circ
+                                \fmap[\T]\teval \circ \strengthr) \\
+  \run{\denn {\rect I N}} & = & \langle \run{\denn{N_i}} \comp
+                                \fmap[\T]{\pi_i} \rangle_{i:I} \\
+\end{array}
+\]
+This construction is the same as \citet{levy:hosc06}, modulo grading.
+The algebra laws ensue.
+
+The coercion morphisms $\Denn N {e_1 \leq e_2}$ are obtained
+analogously:
+\[
+\begin{array}{rcl}
+  \Denn N {e_1 \leq e_2} & : & \Hom {\Denn N {e_1}} {\Denn N {e_2}} \\
+  \Denn {\diamond P} {e_1 \leq e_2} & = & \tcast \\
+  \Denn {P \To N}   {e_1 \leq e_2} & = &
+    \Lambda(\Denn N {e_1 \leq e_2} \comp \teval) \\
+  \Denn {\rect I N} {e_1 \leq e_2} & = &
+    \langle \Denn N {e_1 \leq e_2} \comp \pi_i {} \rangle_{i:I} \\
+\end{array}
+\]
+That these coercions satify the identity and composition---the functor
+laws for $\denn N$---is easy to verify.
+
+\subsubsection{Interpretation of terms}
+
+Values $\ValTy \Gamma v P$ are interpreted as morphisms $\dent v :
+\Hom {\denp \Gamma} {\denp P}$ just as in pure CBPV.  Computations
+$\CompTy \Gamma t e N$ are interpreted as morphisms $\dent t : \Hom
+{\denp \Gamma} {\Denn N e}$.  Most of the cases are straightforward
+and in analogy to CBPV, let us thus focus on the
+\graybox{\mbox{modified}} rules where grading comes into play.  A
+subtlety is that we interpret \emph{typing derivations} rather than terms,
+because rule \rsub is a silent construction on raw terms but becomes a
+coercion in the denotational semantics.  The most natural way to make
+this precise is to use intrinsically well-typed syntax; from typing
+rules to such well-typed syntax is a routine transformation which we
+do not spell out here.  The reader interested in well-typed syntax for
+CBPV is referred to \citet{abelSattler:ppdp19}.
+The other subtlety, that we ignore variable names in the
+interpretation, can also be made precise by well-typed syntax which
+uses de Bruijn indices.
+
+As in Levy's algebra semantics \citeyearpar{levy:hosc06}, creation and
+forcing of thunks is invisible in the model.  The bind operation
+(\relim\diamond) utilizes the generalization of $\tjoin$ to $\trun$.
+Subsumption is interpreted by functoriality in $\Eff$.
+\[
+\begin{array}{llcl}
+\rintro\Box & \dent{\thunk t} & = & \dent t \\
+\relim\Box  & \dent{\force v} & = & \dent v
+\\[1ex]
+\rintro\diamond & \dent{\ret v} & = & \treturn \comp \dent v \\
+\relim\diamond  & \dent{\bind x u t} & = &
+  \run{\denn N} \comp \fmap[\T]\teval \comp \strengthl \comp
+  \langle \Lambda\dent t,\,\dent u \rangle
+\\[1ex]
+\rsub & \dent t & = & \Denn N {e \leq e'} \comp \dent t
+\\
+\end{array}
+\]
+These definitions are understood in the context of the given typing rules.
+
+\subsection{Grading via a partial monoid}
+
+Not all sequences of effects are always meaningful: For instance,
+reading from a file before it was opened is impossible and could be
+prevented statically by graded effect typing.  This could be modelled
+by adding a maximal element $\top : \Eff$---with $e \leq top$ for all
+$e : \Eff$ that signifies an inconsistent state.  This error element
+would also be dominant in sequences, \ie,
+$\top \bu e = e \bu \top = \top$.
+A program $\CompTy \Gamma t e N$ would only be accepted if $e \not= \top$.
+
+Alternatively, we could work with a partial monoid, \ie, a carrier $\Eff$ with a predicate $\Ge{\leq}\_$ and a ternary relation $\_{\bu}\_{\leq}\_$ such that
+the following laws hold:
+\begin{enumerate}
+\item Unit: $e \leq e' \defiff \exists e_0.\, \Ge \leq e_0 \land e_0 \bu e \leq e'$ iff $\exists e_0.\, \Ge \leq e_0 \land e \bu e_0 \leq e'$.
+% \item Unit: $e \leq e' \defiff (\exists e_0.\, \Ge \leq e_0 \land e_0 \bu e \leq e')$ iff $(\exists e_0.\, \Ge \leq e_0 \land e \bu e_0 \leq e')$.
+\item Associativity: $\exists e_{12}.\ e_1 \bu e_2 \leq e_{12} \land e_{12} \bu e_3 \leq e_{123}$ iff $\exists e_{23}.\ e_1 \bu e_{23} \leq e_{123} \land e_2 \bu e_3 \leq e_{23}$.
+\item Monotonicity of $\Ge{\leq}\_$: If $\Ge \leq e$ and $e \leq e'$ then $\Ge \leq e'$.
+\item Monotonicity of $\_{\bu}\_{\leq}\_$:  If $e_1' \leq e_1$ and $e_2' \leq e_2$ and $e \leq e'$ and $e_1 \bu e_2 \leq e$ then $e_1' \bu e_2' \leq e'$.
+\item Reflexivity and transitivity of $\leq$.
+\end{enumerate}
+The typing rules for $\diamond$ would change accordingly.
+\begin{gather*}
+ \nrux{\rintro\diamond}
+     {\ValTy \Gamma v P
+    }{\CompTy \Gamma {\ret v} e {\diamond P}
+    }{\graybox{\Ge \leq e}}
+\qquad
+ \nrux{\relim\diamond}
+     {\CompTy \Gamma u {e_1} {\diamond P} \qquad
+      \CompTy {\ext \Gamma x P} t {e_2} N
+    }{\CompTy \Gamma {\bind x u t} e N
+    }{\graybox{e_1 \bu e_2 \leq e}}
+\end{gather*}
+Rule $\rsub$ would be admissible.
+
+A monad $\T : \Eff \to [\CC \to \CC]$ graded by a partial monoid
+$\Eff$ has natural transformations $\Id \todot \T[e]$ for $\Ge \leq e$
+and $\T[e_1] \circ \T[e_2] \todot \T[e]$ for $e_1 \bu e_2 \leq e$.  A
+$\T$-algebra $B$ has morphism $\run[B] : \T[e_1]B_{e_2} \to B_{e}$
+for $e_1 \bu e_2 \leq e$.
+
+\section{Coeffect-graded CBPV}
+\label{sec:coeff}
+
+\newcommand{\lolli}{\multimap}
+\newcommand{\qfun}[3]{#1#2 \lolli #3}
+\newcommand{\gqfun}[3]{\graybox{#1#2 \lolli #3}}
+\renewcommand{\CompTy}[3]{#1 \vdash #2 : #3}
+\newcommand{\nbox}[1]{}
+\newcommand{\tzip}{\mathsf{zip}}
+\newcommand{\textract}{\mathsf{extract}}
+\newcommand{\extract}[1][]{\textract_{#1}}
+\newcommand{\tduplicate}{\mathsf{duplicate}}
+\newcommand{\duplicate}[1][]{\tduplicate_{#1}}
+\newcommand{\tcobind}{\mathsf{cobind}}
+\newcommand{\tbind}{\mathsf{bind}}
+\newcommand{\tget}{\mathsf{get}}
+\newcommand{\tmodify}{\mathsf{modify}}
+\newcommand{\thead}{\mathsf{head}}
+\newcommand{\ttail}{\mathsf{tail}}
+\newcommand{\suc}{(\_{+}1)}
+\newcommand{\Float}{\mathsf{Float}}
+
+CBPV places the monad $\diamond P$ at the transition from positive
+types to negative types.  Dually, the transition $\Box N$ from
+negative types to positive types is a vessel that can be filled with a
+comonad.  Just like negative types $N$ are monad algebras, positive
+types $P$ can be \emph{comonad coalgebras}.
+
+\subsection{Comonadic CBPV and their comonad coalgebras}
+
+% Let us consider the \emph{context comonad} $\D\,B = S \times B$ that
+% gives us access to a value in $S$.
+% \footnote{\url{https://bartoszmilewski.com/2017/01/02/comonads/}}
+% Besides functoriality, a comonad
+% has the natural transformations $\textract$ and $\duplicate$, dual to
+% $\treturn$ and $\tjoin$.  The implementation of the context comonad is
+% as follows:
+% \[
+% \begin{array}{lcl}
+%   \textract & : & \D\,B \to B \\
+%   \textract\,(s,b) & = & b
+% \\[1.5ex]
+%   \tduplicate & : & \D,\B \to \D\,(\D\,B) \\
+%   \tduplicate\,(s,b) & = & (s, (s, b))
+% \\[1.5ex]
+%   \tget & : & \D\,B \to S \\
+%   \tget\,(s,b) & = & s
+% \\[1.5ex]
+%   \tmodify
+% \end{array}
+% \]
+% The context comonad is not monoidal.
+
+Let us consider the \emph{stream comonad}%
+\footnote{\url{https://bartoszmilewski.com/2017/01/02/comonads/}}
+$\D\,B = \bN \to B$,
+with $\thead\,s = s\,0$ and $\ttail\,s = s \comp \suc$.
+The stream comonad lets us work in a setting where values are not
+single data points, but streams of data.
+Besides functoriality, a
+comonad has the natural transformations $\textract$ and $\duplicate$,
+dual to $\treturn$ and $\tjoin$.  The implementation of the stream
+comonad in $\SET$ is as follows:
+\[
+\begin{array}{lcl}
+  \textract & : & \D\,B \to B \\
+  \textract\,s & = & \thead\,s
+\\[1.5ex]
+  \tduplicate & : & \D\,B \to \D\,(\D\,B) \\
+  \thead\,(\tduplicate\,s) & = & s \\
+  \ttail\,(\tduplicate\,s) & = & \tduplicate\,(\ttail\,s) \\
+\end{array}
+\]
+The comonad laws are a simple dualization of the monad laws:
+\[
+\xymatrix@C=10ex{
+\D\,B  \ar@{=}[dr]
+  & \D\,(\D\,B) \ar[l]_{\textract} \ar[r]^{\tduplicate}
+  & \D\,(\D\,(\D\,B))
+\\
+  & \D\,B \ar[u]_{\tduplicate} \ar[r]^{\tduplicate} \ar@{=}[dr]
+  & \D\,(\D\,B) \ar[u]_{\tfmap\,\tduplicate} \ar[d]^{\tfmap\,\textract}
+\\
+  &
+  & \D\,B
+}
+\]
+
+A \emph{monoidal} comonad implements an morphism $\tzip$ that combines
+a tuples of comonadic values into one comonadic tuple:
+\[
+\begin{array}{lcl}
+  \tzip & : & \tupt {i:I} {(\D\,B_i)} \to \D(\tupt I B) \\
+  \tzip\,(s_i)_{i:I}\,n & = & (s_i\,n)_{i:I} \\
+\end{array}
+\]
+Many other comonads are monoidal,
+like the store comonad $(S \To \_) \times S$,
+though not all, \eg, the context comonad $S{\times}\_$.
+In this article, we consider only monoidal comonads.
+
+If type constructor $\Box$ is interpreted as a monoidal comonad $\D$,
+then all positive types $P$ can be intepreted as $\D$-coalgebras
+$A = \den P$, \ie, implement a morphism $\expose[A] : A \to \D\,A$
+satisfying the laws of a comonad coalgebra:
+\[
+\xymatrix@C=10ex{
+A  \ar@{=}[dr]
+  & \D\,A \ar[l]_{\textract} \ar[r]^{\tduplicate}
+  & \D\,(\D\,A)
+\\
+  & A \ar[u]_{\texpose} \ar[r]^{\texpose}
+  & \D\,A \ar[u]_{\tfmap\,\texpose}
+}
+\]
+The implementation of $\expose{\den P}$ proceeds by induction on $P$.
+\[
+\begin{array}{lcl}
+  \expose{\den{\Box N}} & : & {\D\,\den N} \to {\D\,(\D\,\den N)} \\
+  \expose{\den{\Box N}} & = & \duplicate
+\\[1ex]
+  \expose{\den{\tupt I P}} & : & \tupt {i:I} {\den{P_i}} \to \D\,(\tupt {i:I} {\den{P_i}}) \\
+  \expose{\den{\tupt I P}} & = & \tzip \comp \tupt {i:I} \expose{\den{P_i}}
+\\[1ex]
+  \expose{\den{\sumt I P}} & : & \sumt {i:I} {\den{P_i}} \to \D\,(\sumt {i:I} {\den{P_i}}) \\
+  \expose{\den{\sumt I P}} & = & [\tfmap\,\iota_i \comp \expose{\den{P_i}}]_{i:I} \\
+\end{array}
+\]
+Positive base types $o \in \PTy$ are required to be $\D$-coalgebras.
+For instance, elements of a base type like
+$\Float \in \PTy$ could represent streams of floating point numbers,
+\eg, continuous measurements from a sensor.
+
+Because comonad coalgebras are closed under products for monoidal
+coalgebras, contexts $\Gamma$ are interpreted as comonad coalgebras
+$\den\Gamma$.  In the presence of a comonad interpretation of $\Box$,
+the operations $\tthunk$ and $\tforce$ are no longer the identity, but
+``generalized cobind'' and $\extract$:
+\[
+\begin{array}{lllcl}
+ \rintro\Box
+  & \ru{\oCompTy \Gamma t N
+      }{\ValTy \Gamma {\thunk t} {\Box N}}
+  & \dent{\thunk t} & = & \fmap[\D]{\dent t} \comp \expose[\den\Gamma]
+\\[3ex]
+ \relim\Box
+  & \ru{\ValTy \Gamma v {\Box N}
+      }{\oCompTy \Gamma {\force v} N}
+  & \dent{\force v} & = & \textract \comp \dent v
+\end{array}
+\]
+The map $\dent\tthunk : \Hom A B \to \Hom A {\D\,B}$ generalizes
+$\tcobind : \Hom{\D\,A} B \to \Hom{\D\,A}{\D\,B}$ to $\D$-coalgebras
+$A$ in the same way that the monadic
+$\tbind : \Hom{A}{\T\,B} \to \Hom{\T\,A}{\T\,B}$ is generalized to
+$\Hom{A}{B} \to \Hom{\T\,A}{B}$ for $\T$-algebras $B$ in CBPV.
+
+% In a stream-valued instance of CBPV, elements of a base type like
+% $\Float \in \PTy$ would represent streams of floating point numbers,
+% \eg, continuous measurements from a sensor.
+% However, a single comonad $\D$ may not sufficient to express what we
+% want.  In the next section, we look at \emph{graded comonads}.
+
+\subsection{Graded comonads and their coalgebras}
+
+Graded comonads have been utilized to give semantics to
+context-dependent computation \citep{orchard:icfp14}.  For grading,
+loccit uses a \emph{resource algebra} in form of a pre-ordered
+semiring
+$(\R,{+},0,{\cdot},1,\leq)$.
+% $(\R,\_{+}\_,0,\_{\cdot}\_,1,\leq)$.
+%
+The semantics is \emph{resource aware} and thus not constructed in a
+cartesian-closed category but in a symmetric monoidal closed category
+$(\CC,\otimes,\I,\lolli)$.
+%
+As a running example, we use the semiring $\R = \Pot\bN \setminus \emptyset$ of
+\emph{multiplicities} ordered by $\supseteq$ with pointwise sum and
+product, \eg, $r + s = \{ n + m \mid n \in r, m \in s \}$.
+Subsemirings of $\R$ have been used for quantitative typing
+\cite{sergeyVytiniotisPeytonJones:popl14,mcBride:wadler60,atkey:lics08}.
+The order expresses precision of the quantities, \eg,
+$\{1\} \geq \{0,1\} \geq \bN$ states that linear use of a variable is more informative than affine use than unrestricted use.
+
+A matching symmetric monoidal closed category $\CC$ can be obtained as
+follows:
+% Assume a preordered commutative monoid $(\W,\oplus,\emp,\wleq,\winf)$
+% that is also a $\winf$-semilattice with distribution law $\winf_{i:I} (w_i \oplus w'_i) \wleq \winf_I w \oplus \winf_I w'$.  The elements (``worlds'') can be
+% seen as collection of available resources under uncertainty.  The
+% order $w \wleq w'$ again expresses precision of information, \ie, if
+% we can construct something from resources $w$ we can also construct it
+% from $w'$.  The infimum $\winf_{i:I}w_i$ expresses we can build the
+% ``thing'' from any of the $w_i$.  Multiplication $n \cdot w$ is
+% understood as $w \oplus \dots \oplus w$ with $n$ summands.
+% %
+% Given a commutative monoid $(\M,\uplus,\emptyset)$ whose elements can
+% be thought of as bags of atomic resources, an instance for $\W$ would
+% be $\W = \Pot \M$ under ${\wleq} = {\subseteq}$ and pointwise union
+% $w \oplus w' = \{ m \uplus m' \mid m \in w, m' \in w' \}$.
+%
+% Resource qualifiers $r \in \R$ operate on worlds via $r \cdot w = \winf_{n \in r} (n \cdot w)$.  It is routine to verify that $\W$ is almost a left semimodule to $\R$, \ie, the following laws hold:
+% \[
+% \begin{array}{rclclcl}
+%   1 \cdot w
+%     & = & \winf_{n \in \{1\}} nw
+%     & = & 1w
+%     & = & w
+% \\
+%   (rs) \cdot w
+%     & = & \winf_{n \in r, m \in s} (nm)w
+%     & = & \winf_{n \in r} n(\winf_{m \in s} mw)
+%     & = & r \cdot (s \cdot w)
+% \\
+%   0 \cdot w
+%     & = & \winf_{n \in \{0\}} nw
+%     & = & 0w
+%     & = & \emp
+% \\
+%   (r + s) \cdot w
+%     & = & \winf_{n \in r, m \in s} (n+m)w
+%     & = & \winf_{n \in r} \winf_{m \in s} (nw \oplus mw)
+%     & = & (r \cdot w) \oplus (s \cdot w)
+% \\
+%   r \cdot \emp
+%     & = & \winf_{n \in r} n\emp
+%     & = & \winf_{n \in r} \emp
+%     & = & \emp
+% \\
+%   r \cdot (w \oplus w')
+%     & = & \winf_{n \in r} n(w \oplus w')
+%     & \wleq & \winf_{n \in r} nw \oplus \winf_{n \in r} nw'
+%     & = & (r \cdot w) \oplus (r \cdot w')
+% \\
+% \end{array}
+% \]
+Assume a preordered commutative monoid $(\W,\oplus,\emp,\wleq,\wsup)$
+that is also a $\wsup$-semilattice with distribution law $\wsup_{i:I} (w_i \oplus w'_i) \wleq \wsup_I w \oplus \wsup_I w'$.  The elements (``worlds'') can be
+seen as collection of available resources under choice.  The
+order $w \wleq w'$ again expresses precision of information, \ie, if
+we can construct something from resources $w$ we can also construct it
+from $w'$ where we have additional choices.
+The supremum $\wsup_{i:I}w_i$ expresses we can build the
+``thing'' from any of the $w_i$.  Multiplication $n \cdot w$ is
+understood as $w \oplus \dots \oplus w$ with $n$ summands.
+%
+Given a commutative monoid $(\M,\uplus,\emptyset)$ whose elements can
+be thought of as bags of atomic resources, an instance for $\W$ would
+be $\W = \Pot \M$ under ${\wleq} = {\subseteq}$ and pointwise union
+$w \oplus w' = \{ m \uplus m' \mid m \in w, m' \in w' \}$.
+
+Resource qualifiers $r \in \R$ operate on worlds via
+$r \cdot w = \wsup_{n \in r} (n \cdot w)$.
+%
+For instance the \emph{affine} qualifier $\{0,1\}w = \emp \oplus w$
+gives us the choice of using resources $w$ or not ($\emp$).
+%
+It is routine to verify that $\W$ is almost a left semimodule to $\R$,
+\ie, the following laws hold:
+\[
+\begin{array}{rclclcl}
+  1 \cdot w
+    & = & \wsup_{n \in \{1\}} nw
+    & = & 1w
+    & = & w
+\\
+  (rs) \cdot w
+    & = & \wsup_{n \in r, m \in s} (nm)w
+    & = & \wsup_{n \in r} n(\wsup_{m \in s} mw)
+    & = & r \cdot (s \cdot w)
+\\
+  0 \cdot w
+    & = & \wsup_{n \in \{0\}} nw
+    & = & 0w
+    & = & \emp
+\\
+  (r + s) \cdot w
+    & = & \wsup_{n \in r, m \in s} (n+m)w
+    & = & \wsup_{n \in r} \wsup_{m \in s} (nw \oplus mw)
+    & = & (r \cdot w) \oplus (s \cdot w)
+\\
+  r \cdot \emp
+    & = & \wsup_{n \in r} n\emp
+    & = & \wsup_{n \in r} \emp
+    & = & \emp
+\\
+  r \cdot (w \oplus w')
+    & = & \wsup_{n \in r} n(w \oplus w')
+    & \wleq & \wsup_{n \in r} nw \oplus \wsup_{n \in r} nw'
+    & = & (r \cdot w) \oplus (r \cdot w')
+\\
+\end{array}
+\]
+Because the last law is not an equality but just the inequality
+$r (w \oplus w') \wleq rw \oplus rw'$, we have ``almost'' a
+semimodule.
+
+Objects of $\CC$ are functors $A : (\W,\wleq) \to \SET$ and morphisms
+$f : \Hom A B$ are natural transformations
+$(f_w : A_w \to B_w)_{w:W}$, \ie,
+$B_{w \wleq w'} \comp f_w = f_{w'} \comp A_{w \wleq w'}$.  The tensor
+$A \otimes B$ is Day's convolution
+$(A \otimes B)_w = \bigcup_{w_1 \oplus w_2 \wleq w} (A_{w_1} \times B_{w_2})$
+% $(A \otimes B)_w = \bigcup_{w_1,w_2 \mid w \geq w_1 \oplus w_2} A_{w_1} \times A_{w_2}$
+with unit $\I_w = \bigcup_{\emp \wleq w} 1$.  The unit $\I$ is
+constructible at world $w$ from nothing (set $1$) if $\emp \wleq w$,
+\ie, if the world $w$ includes the choice of using no resources.  A
+tensor $A \otimes B$ is constructible at world $w$ if $w$ includes the
+choice to split the resources into $w_1$ and $w_2$ to construct $A$
+and $B$, resp.  The exponential $A \lolli B$ is determined by currying
+$C \otimes A \lolli B \cong C \lolli (A \lolli B)$ and given by
+$(A \lolli B)_w = \bigcap_{w_1 \oplus w \wleq w_2} (A_{w_1} \to
+B_{w_2})$.
+
+The thus constructed symmetric monoidal closed category $\CC$ has a
+$\R$-graded comonad
+\[
+\begin{array}{lcl}
+  \D & : & (\R,\leq) \to [\CC \to \CC] \\
+(\D[r] A)_w & = &
+  \bigcup_{rw' \wleq w} A_{w'}
+  % \bigcup_{w' \mid w \wleq rw'} A_{w'}
+\end{array}
+\]
+implementing trivially the following natural transformations.  Herein, we use the
+pointwise tensor product of functors
+$(\dot\otimes_{i:I}F_i)\,A = \otimes_{i:I}(F_i\,A)$ in
+$[\CC \to \CC]$, in particular $\dot\I\,A = \I$.
+\[
+\begin{array}{lcl@{~}c@{~}l}
+  \textract   & : & \D[1]   & \todot & \Id \\
+  \tduplicate & : & \D[rs]  & \todot & \D[r] \comp \D[s] \\
+  \tweaken    & : & \D[0]   & \todot & \dot{\I} \\
+  \tcontract  & : & \D[r+s] & \todot & \D[r] \mathbin{\dot\otimes} \D[s] \\
+  % \tweaken    & : & \Hom {\D[0]\,A} A \\
+  % \tcontract  & : & \Hom {\D[r+s]\,A} {\D[r]\,\A \otimes \D[s]\,A} \\
+\end{array}
+\]
+For example, $\tcontract$ takes
+$a \in \bigcup_{rw' \oplus sw' \wleq w} A_{w'}$ to
+$(a,a) \in \bigcup_{w_1 \oplus w_2 \wleq w} \bigcup_{rw_1' \leq w_1}
+\bigcup_{sw_2' \leq w_2} A_{w'_1} \times A_{w'_2}$.
+The intermediate worlds are $w_1' = w_2' = w'$ and $w_1 = rw'$ and $w_2 = sw'$.
+
+The monoidal character of $\D$ is witnessed by the trivial morphism $\tzip$:
+\[
+%\newcommand{\Aiw}{(A_i)_{w'_i}}
+\newcommand{\Aiw}{A_{iw'_i}}
+\begin{array}{lcl}
+  \tzip       & : & \Hom {\tupt {i:I} {(\D[r]\,A_i)}} {\D[r]\,(\tupt I A)} \\
+  \tzip_w     & : & \bigcup_{\oplus_{i:I}w_i \wleq w} \Pi_{i:I} (\D[r]\,A_i)_{w_i} \to \bigcup_{rw' \wleq w} (\tupt I A)_{w'} \\
+  \tzip_w     & : & \bigcup_{\oplus_{i:I} w_i \wleq w} \Pi_{i:I} (\bigcup_{rw'_i \wleq w_i} \Aiw) \to \bigcup_{rw' \wleq w} \bigcup_{\oplus_{i:I} w'_i \wleq w'} \Pi_{i:I} \Aiw \\
+   \tzip_w\,(a_i)_{i:I} & = & (a_i)_{i:I} \\
+\end{array}
+\]
+
+The comonad laws generalize to graded comonads as follows:
+\[
+\xymatrix@C=10ex{
+\D[r]\,B  \ar@{=}[dr]
+  & \D[1]\,(\D[r]\,B) \ar[l]_{\textract}
+\\
+  & \D[r]\,B \ar[u]_{\tduplicate}
+}
+\qquad
+\xymatrix@C=10ex{
+\D[r]\,B  \ar[r]^{\tduplicate} \ar@{=}[dr]
+  & \D[r]\,(\D[1]\,B)  \ar[d]^{\tfmap\,\textract}
+\\
+  & \D[r]\,B
+}
+\]
+\[
+\xymatrix@C=10ex{
+\D[q]\,(\D[rs]\,B) \ar[r]^{\tduplicate}
+  & \D[q]\,(\D[r]\,(\D[s]\,B))
+\\
+\D[qrs]\,B \ar[u]_{\tduplicate} \ar[r]^{\tduplicate}
+  & \D[qr]\,(\D[s]\,B) \ar[u]_{\tfmap\,\tduplicate}
+}
+\]
+These laws reflect that $\D$ maps the multiplicative monoid
+$(\R,{\cdot},1)$ to the monoid structure $([\CC \to \CC],{\circ},\Id)$
+of composition in the functor category $[\CC \to \CC]$.  Similar laws
+need to hold for the additive monoid and distributivity.
+\[
+\xymatrix@C=12ex{
+\D[0]\,B \otimes \D[r]\,B \ar[d]_{\tweaken \otimes \tid}
+& \D[r]\,B \ar[l]_{\tcontract} \ar[r]^{\tcontract} \ar@{=}[d]
+& \D[r]\,B \otimes \D[0]\,B \ar[d]^{\tid \otimes \tweaken}
+\\
+\I \otimes \D[r]\,B \ar[r]^{\Gl}
+& \D[r]\,B
+& \D[r]\,B \otimes \I \ar[l]_{\Gr}
+}
+\]
+\[
+\xymatrix@C=10ex{
+\D[q]\,B \otimes \D[r+s]\,B  \ar[d]_{\tid \otimes \tcontract}
+& \D[q+r+s]\,B \ar[l]_{\tcontract} \ar[r]^{\tcontract}
+& \D[q+r]\,B \otimes \D[s]\,B \ar[d]^{\tcontract \otimes \tid}
+\\
+\D[q]\,B \otimes (\D[r]\,B \otimes \D[s]\,B) \ar[rr]^{\Ga}
+& & (\D[q]\,B \otimes \D[r]\,B) \otimes \D[s]\,B
+}
+\]
+\[
+\xymatrix{
+\D[qr]\,B \otimes \D[qs]\,B \ar[d]_{\duplicate \otimes \duplicate}
+  & \D[q(r+s)]\,B \ar[l]_{\tcontract} \ar[r]^{\duplicate}
+  & \D[q](\D[r+s]\,B) \ar[d]^{\fmap\tcontract}
+\\
+\D[q](\D[r]\,B) \otimes \D[q](\D[s]\,B) \ar[rr]^{\tzip}
+&&
+\D[q](\D[r]\,B \otimes \D[s]\,B)
+}
+\]
+\[
+\xymatrix@C=22ex{
+\D[(q+r)s]\,B \ar[r]^{\tduplicate} \ar[d]_{\tcontract}
+& \D[q+r]\,(\D[s]\,B) \ar[d]^{\tcontract}
+\\
+\D[qs]\,B \otimes \D[rs]\,B \ar[r]^{\tduplicate \otimes \tduplicate}
+& \D[q](\D[s]\,B) \otimes \D[r](\D[s]\,B)
+}
+\]
+
+\begin{figure}[htbp]
+\flushleft
+\ruler{}
+Types.
+\[
+\begin{array}{lllrl@{\qquad}l}
+\PTy & \ni & P
+  & ::= & \Box N
+     \mid o \mid \sumty i I P \mid \tupty i I P
+  & \nbox{Value types (positive types)} \\
+\NTy & \ni & N
+  & ::= & \diamond P % \graybox{\compty r P}
+     \mid \gqfun r P N \mid \recty i I N
+  & \nbox{Computation types (negative types)} \\
+\end{array}
+\]
+\dashruler{}
+Terms.
+\[
+\begin{array}{lllrl@{\qquad}l}
+\PTm & \ni & v,w
+  & ::= & x
+     \mid \thunk t
+     \mid \inj i v
+     \mid \ptup{\bar v}
+  & \nbox{Values (positive terms)} \\
+\NTm & \ni & t,u
+  & ::= & \graybox{y} \mid \graybox{\qlet r y u t}
+     \mid \qlet {\graybox r} x v t
+  & \nbox{Computations (negative terms):}
+\\ &&& \mid &
+         \graybox{\qforced r v y t}
+    \mid \qcase {\graybox r} v {\overline{x.\,t}}
+    \mid \qsplits {\graybox r} v {\bar x} t
+  & \nbox{value eliminations}
+\\ &&& \mid &
+         \ret v        \mid \qbind {\graybox r} x u t
+  & \nbox{monad operations}
+\\ &&& \mid &
+         \lam x t      \mid \app t v
+  & \nbox{functions}
+\\ &&& \mid &
+         \trecord\{\overline{i:t}\} % \ntup{\bar t}
+       \mid \proj i t
+  & \nbox{lazy tuples (records)}
+\end{array}
+\]
+\dashruler{}
+Value typing \fbox{$\gqValTy \gamma \Gamma v P$}\,.
+\begin{gather*}
+ \nru{\rvar^+}{}
+     {\ValTy {\gqext 0\Gamma 1x P} x P}
+\qquad
+ \nru{\rintro\Box}
+     {\qCompTy \gamma \Gamma t N
+    }{\qValTy \gamma \Gamma {\thunk t} {\Box N}}
+\\[1.5ex]
+ \nru{\rintro\GS}
+     {\qValTy \gamma \Gamma v {P_i}
+    }{\qValTy \gamma \Gamma {\inj i v} {\sumt I P}}
+\qquad
+ \nru{\rintro\otimes}
+     {\forall i{:}I, \ \gqValTy {\gamma_i} \Gamma {v_i} {P_i}
+    }{\gpValTy {\GS_{i:I}\gamma_i} \Gamma {\ptup v} {\tupt I P}}
+\end{gather*}
+\dashruler{}
+Computation typing \fbox{$\gqCompTy \gamma \Gamma t N$}\,.
+\vspace{-2ex}
+\begin{gather*}
+\graybox{
+ \nru{\rvar^-}{}
+     {\ValTy {\gqext 0 \Gamma 1 y N} y N}
+}
+\qquad
+\graybox{
+ \nru{\rlet^-}
+     {\qCompTy \gamma \Gamma u N \qquad
+      \CompTy {\qext \delta \Gamma r y N} t {N'}
+    }{\gpCompTy {r\gamma + \delta} \Gamma {\qlet r y u t} {N'}}
+}
+\\[1.5ex]
+ \nru{\rlet^+}
+     {\gqValTy \gamma \Gamma v P \qquad
+      \CompTy {\gqext \delta \Gamma r x P} t N
+    }{\gpCompTy {r\gamma + \delta} \Gamma {\gqlet r x v t} N}
+\qquad
+ \nru{\relim\Box}
+     {\gqValTy \gamma \Gamma v {\Box N} \qquad
+      \CompTy {\gqext \delta \Gamma r y N} t {N'}
+    }{\gpCompTy {r\gamma + \delta} \Gamma {\gqforced r v y t} {N'}}
+\\[1.5ex]
+ \nru{\relim\GS}
+     {\gqValTy \gamma \Gamma v {\sumt I P} \qquad
+      \forall i{:}I,\ \CompTy {\gqext \delta \Gamma r {x_i} {P_i}} {t_i} N
+    }{\gpCompTy {r\gamma + \delta} \Gamma {\caser v {\{\graybox{r} x_i.\, t_i\}_{i:I}}} N}
+\qquad
+ \nru{\relim\otimes}
+     {\gqValTy \gamma \Gamma v {\tupt I P} \qquad
+      \gqCompTy \delta {\extr \Gamma {\overline{\graybox{r}x_i{:}P_i}^{i:I}}} t N
+    }{\gpCompTy {r\gamma + \delta} \Gamma {\gqsplits r v {\bar x} t} N}
+\\[1.5ex]
+ \nru{\rintro\diamond}
+     {\qValTy \gamma \Gamma v P
+    }{\qCompTy \gamma \Gamma {\ret v} {\diamond P}}
+\qquad
+ \nru{\relim\diamond}
+     {\gqCompTy \gamma \Gamma u {\diamond P} \qquad
+      \CompTy {\gqext \delta \Gamma r x P} t  N
+    }{\gpCompTy {r\gamma + \delta} \Gamma {\gqbind r x u t} N}
+\\[1.5ex]
+ \nru{\rintro\lolli}
+     {\CompTy {\gqext \gamma \Gamma r x P} t N
+    }{\gqCompTy \gamma \Gamma {\lam x t} {\gqfun r P N}}
+\qquad
+ \nru{\relim\lolli}
+     {\gqCompTy \gamma \Gamma t {\gqfun r P N} \qquad
+      \gqValTy \delta \Gamma v P
+    }{\gpCompTy {\gamma + r\delta} \Gamma {\app t v} N}
+\\[1.5ex]
+ \nru{\rintro\Pi}
+     {\forall i{:}I,\ \qCompTy \gamma \Gamma {t_i} {N_i}
+    }{\qCompTy \gamma \Gamma {\trecord {\{i : t_i\}_{i:I}}} {\rect I N}}
+    % }{\qCompTy \gamma \Gamma {\ntup{\overline{t_i}^{i:I}}} {\rect I N}}
+\qquad
+ \nru{\relim\Pi}
+     {\qCompTy \gamma \Gamma t {\rect I N}
+    }{\qCompTy \gamma \Gamma {\proj i t}  {N_i}}
+\\[1.5ex]
+\graybox{
+ \nrux{\rweak}
+      {\gqCompTy {\gamma'} \Gamma t N}
+      {\gqCompTy \gamma \Gamma t N}
+      {\gamma \leq \gamma'}
+}
+\end{gather*}
+\rule{\textwidth}{0.2pt}
+  \caption{Coeffect-graded call-by-push-value.}
+  \label{fig:cbpv}
+\end{figure}
 

### 2020-07-07 popl21/popl21.tex
[ popl21 ] old coeffect grading (<> instead of < r >)

@@
+\nonstopmode
 %% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
@@
 \newcommand{\Denn}[2]{\denn{#1}_{#2}}
+\newcommand{\Den}[2]{\den{#1}_{#2}}
 
@@
 \newcommand{\Ga}{\ensuremath{\alpha}}
+\newcommand{\Gd}{\ensuremath{\delta}}
 \newcommand{\Gl}{\ensuremath{\lambda}}
@@
 \newcommand{\tweaken}{\mathsf{weaken}}
+\newcommand{\tdrop}[1][]{\mathsf{drop}_{#1}}
+%\newcommand{\drop}[1][]{\tdrop_{#1}} % Name clash
 \newcommand{\tcontract}{\mathsf{contract}}
+\newcommand{\contract}[1][]{\tcontract_{#1}}
 
@@
 \newcommand{\NTy}{\Ty-}
+\newcommand{\Cxt}{\mathsf{Cxt}}
+\newcommand{\RCxt}{\mathsf{RCxt}}
+\newcommand{\cempty}{\emptyset}
 \newcommand{\extr}[2]{#1.#2}
@@
   & \mbox{Computation types (negative types)} \\
+\Cxt & \ni & \Gamma
+  & ::= & \cempty \mid \ext \Gamma x P
+  & \mbox{Typing context}
+\\
 \end{array}
@@
 \newcommand{\extract}[1][]{\textract_{#1}}
+\newcommand{\tdisplay}{\mathsf{display}}
+\newcommand{\display}[1][]{\tdisplay_{#1}}
+% \newcommand{\tduplicate}{\mathsf{display}}
+% \newcommand{\duplicate}[1][]{\tdisplay_{#1}}
 \newcommand{\tduplicate}{\mathsf{duplicate}}
@@
 \newcommand{\Float}{\mathsf{Float}}
+\newcommand{\tiso}{\mathsf{iso}}
+\newcommand{\isot}{\tiso^\otimes}
 
@@
 % Besides functoriality, a comonad
-% has the natural transformations $\textract$ and $\duplicate$, dual to
+% has the natural transformations $\textract$ and $\display$, dual to
 % $\treturn$ and $\tjoin$.  The implementation of the context comonad is
@@
 % \\[1.5ex]
-%   \tduplicate & : & \D,\B \to \D\,(\D\,B) \\
-%   \tduplicate\,(s,b) & = & (s, (s, b))
+%   \tdisplay & : & \D,\B \to \D\,(\D\,B) \\
+%   \tdisplay\,(s,b) & = & (s, (s, b))
 % \\[1.5ex]
@@
 Besides functoriality, a
-comonad has the natural transformations $\textract$ and $\duplicate$,
+comonad has the natural transformations $\textract$ and $\display$
+(in category theory called $\Ge$ and $\Gd$),
 dual to $\treturn$ and $\tjoin$.  The implementation of the stream
@@
 \\[1.5ex]
-  \tduplicate & : & \D\,B \to \D\,(\D\,B) \\
-  \thead\,(\tduplicate\,s) & = & s \\
-  \ttail\,(\tduplicate\,s) & = & \tduplicate\,(\ttail\,s) \\
-\end{array}
-\]
+  \tdisplay & : & \D\,B \to \D\,(\D\,B) \\
+  \thead\,(\tdisplay\,s) & = & s \\
+  \ttail\,(\tdisplay\,s) & = & \tdisplay\,(\ttail\,s) \\
+\end{array}
+\]
+The generic operations of a comonad allow us to \emph{extract} the
+value wrapped in a comonadic structure and to \emph{display} another
+layer of that structure.
+
 The comonad laws are a simple dualization of the monad laws:
@@
 \D\,B  \ar@{=}[dr]
-  & \D\,(\D\,B) \ar[l]_{\textract} \ar[r]^{\tduplicate}
+  & \D\,(\D\,B) \ar[l]_{\textract} \ar[r]^{\tdisplay}
   & \D\,(\D\,(\D\,B))
 \\
-  & \D\,B \ar[u]_{\tduplicate} \ar[r]^{\tduplicate} \ar@{=}[dr]
-  & \D\,(\D\,B) \ar[u]_{\tfmap\,\tduplicate} \ar[d]^{\tfmap\,\textract}
+  & \D\,B \ar[u]_{\tdisplay} \ar[r]^{\tdisplay} \ar@{=}[dr]
+  & \D\,(\D\,B) \ar[u]_{\tfmap\,\tdisplay} \ar[d]^{\tfmap\,\textract}
 \\
@@
 A  \ar@{=}[dr]
-  & \D\,A \ar[l]_{\textract} \ar[r]^{\tduplicate}
+  & \D\,A \ar[l]_{\textract} \ar[r]^{\tdisplay}
   & \D\,(\D\,A)
@@
 \]
-The implementation of $\expose{\den P}$ proceeds by induction on $P$.
+The implementation of $\expose[\den P]$ proceeds by induction on $P$.
 \[
 \begin{array}{lcl}
-  \expose{\den{\Box N}} & : & {\D\,\den N} \to {\D\,(\D\,\den N)} \\
-  \expose{\den{\Box N}} & = & \duplicate
+  \expose[\den{\Box N}] & : & {\D\,\den N} \to {\D\,(\D\,\den N)} \\
+  \expose[\den{\Box N}] & = & \display
 \\[1ex]
-  \expose{\den{\tupt I P}} & : & \tupt {i:I} {\den{P_i}} \to \D\,(\tupt {i:I} {\den{P_i}}) \\
-  \expose{\den{\tupt I P}} & = & \tzip \comp \tupt {i:I} \expose{\den{P_i}}
+  \expose[\den{\tupt I P}] & : & \tupt {i:I} {\den{P_i}} \to \D\,(\tupt {i:I} {\den{P_i}}) \\
+  \expose[\den{\tupt I P}] & = & \tzip \comp \tupt {i:I} \expose[\den{P_i}]
 \\[1ex]
-  \expose{\den{\sumt I P}} & : & \sumt {i:I} {\den{P_i}} \to \D\,(\sumt {i:I} {\den{P_i}}) \\
-  \expose{\den{\sumt I P}} & = & [\tfmap\,\iota_i \comp \expose{\den{P_i}}]_{i:I} \\
+  \expose[\den{\sumt I P}] & : & \sumt {i:I} {\den{P_i}} \to \D\,(\sumt {i:I} {\den{P_i}}) \\
+  \expose[\den{\sumt I P}] & = & \left[\tfmap\,\iota_i \comp \expose[\den{P_i}]\right]_{i:I} \\
 \end{array}
@@
 cartesian-closed category but in a symmetric monoidal closed category
-$(\CC,\otimes,\I,\lolli)$.
-%
-As a running example, we use the semiring $\R = \Pot\bN \setminus \emptyset$ of
+$(\CC,\otimes,\I,\lolli)$.  We overload the operators on
+exponentials, $\Lambda : \Hom {C \otimes A} B \to \Hom C {A \lolli B}$
+and $\teval : \Hom{(A \lolli B) \otimes A} B$.  The tensor product
+$\otimes$ is a bifunctor and it is customary to overload $\otimes$ for
+the functorial action
+$f_1 \otimes f_2 : \Hom{A_1 \otimes A_2}{B_1 \otimes B_2}$ where
+$f_i : \Hom{A_i}{B_i}$.  We shall also use $I$-ary products
+$\tupt {i:I} {A_i}$ and their functorial action $\tupt {i:I} {f_i}$.
+Further, the symmetric monoidal structure is usually witnessed by
+isomorphisms for left $\lambda_A : \Hom {\I \otimes A} A$ and right
+unit $\rho_A : \Hom {A \otimes I} A$, associativity
+$\Ga_{A,B,C} : \Hom{A \otimes (B \otimes C)} {(A \otimes B) \otimes
+  C}$
+and swap $\sigma_{A,B} : \Hom {A \otimes B} {B \otimes A}$.  We shall
+summarize combinations of these isomorphisms under the name
+$\isot : \Hom {\tupt I A} {\tupt J B}$ when the multisets
+$\{A_i\}_{i:I}$ and $\{B_j\}_{j:J}$ coincide modulo addition and
+deletion of $\I$s.
+
+As a running example for resource accounting,
+we use the semiring $\R = \Pot\bN \setminus \emptyset$ of
 \emph{multiplicities} ordered by $\supseteq$ with pointwise sum and
@@
   \textract   & : & \D[1]   & \todot & \Id \\
-  \tduplicate & : & \D[rs]  & \todot & \D[r] \comp \D[s] \\
-  \tweaken    & : & \D[0]   & \todot & \dot{\I} \\
-  \tcontract  & : & \D[r+s] & \todot & \D[r] \mathbin{\dot\otimes} \D[s] \\
-  % \tweaken    & : & \Hom {\D[0]\,A} A \\
-  % \tcontract  & : & \Hom {\D[r+s]\,A} {\D[r]\,\A \otimes \D[s]\,A} \\
-\end{array}
-\]
-For example, $\tcontract$ takes
+  \tdisplay & : & \D[rs]  & \todot & \D[r] \comp \D[s] \\
+  \tdrop    & : & \D[0]   & \todot & \dot{\I} \\
+  \tduplicate  & : & \D[r+s] & \todot & \D[r] \mathbin{\dot\otimes} \D[s] \\
+  % \tdrop    & : & \Hom {\D[0]\,A} A \\
+  % \tduplicate  & : & \Hom {\D[r+s]\,A} {\D[r]\,\A \otimes \D[s]\,A} \\
+\end{array}
+\]
+For example, $\tduplicate$ takes
 $a \in \bigcup_{rw' \oplus sw' \wleq w} A_{w'}$ to
@@
 \xymatrix@C=10ex{
-\D[r]\,B  \ar@{=}[dr]
-  & \D[1]\,(\D[r]\,B) \ar[l]_{\textract}
-\\
-  & \D[r]\,B \ar[u]_{\tduplicate}
+\D[r]\,B  \ar[d]_{\tdisplay} \ar[r]^{\tdisplay} \ar@{=}[dr]
+  & \D[r]\,(\D[1]\,B)  \ar[d]^{\tfmap\,\textract}
+\\
+\D[1]\,(\D[r]\,B) \ar[r]^{\textract}
+  & \D[r]\,B
 }
 \qquad
-\xymatrix@C=10ex{
-\D[r]\,B  \ar[r]^{\tduplicate} \ar@{=}[dr]
-  & \D[r]\,(\D[1]\,B)  \ar[d]^{\tfmap\,\textract}
-\\
-  & \D[r]\,B
-}
-\]
-\[
-\xymatrix@C=10ex{
-\D[q]\,(\D[rs]\,B) \ar[r]^{\tduplicate}
+\xymatrix@C=12ex{
+\D[qrs]\,B \ar[d]_{\tdisplay} \ar[r]^{\tdisplay}
+  & \D[qr]\,(\D[s]\,B) \ar[d]^{\tdisplay}
+\\
+\D[q]\,(\D[rs]\,B) \ar[r]^{\tfmap\,\tdisplay}
   & \D[q]\,(\D[r]\,(\D[s]\,B))
-\\
-\D[qrs]\,B \ar[u]_{\tduplicate} \ar[r]^{\tduplicate}
-  & \D[qr]\,(\D[s]\,B) \ar[u]_{\tfmap\,\tduplicate}
-}
-\]
+}
+\]
+% \[
+% \xymatrix@C=10ex{
+% \D[r]\,B  \ar@{=}[dr]
+%   & \D[1]\,(\D[r]\,B) \ar[l]_{\textract}
+% \\
+%   & \D[r]\,B \ar[u]_{\tdisplay}
+% }
+% \qquad
+% \xymatrix@C=10ex{
+% \D[r]\,B  \ar[r]^{\tdisplay} \ar@{=}[dr]
+%   & \D[r]\,(\D[1]\,B)  \ar[d]^{\tfmap\,\textract}
+% \\
+%   & \D[r]\,B
+% }
+% \]
+% \[
+% \xymatrix@C=10ex{
+% \D[q]\,(\D[rs]\,B) \ar[r]^{\tdisplay}
+%   & \D[q]\,(\D[r]\,(\D[s]\,B))
+% \\
+% \D[qrs]\,B \ar[u]_{\tdisplay} \ar[r]^{\tdisplay}
+%   & \D[qr]\,(\D[s]\,B) \ar[u]_{\tfmap\,\tdisplay}
+% }
+% \]
 These laws reflect that $\D$ maps the multiplicative monoid
@@
 \xymatrix@C=12ex{
-\D[0]\,B \otimes \D[r]\,B \ar[d]_{\tweaken \otimes \tid}
-& \D[r]\,B \ar[l]_{\tcontract} \ar[r]^{\tcontract} \ar@{=}[d]
-& \D[r]\,B \otimes \D[0]\,B \ar[d]^{\tid \otimes \tweaken}
+\D[0]\,B \otimes \D[r]\,B \ar[d]_{\tdrop \otimes \tid}
+& \D[r]\,B \ar[l]_{\tduplicate} \ar[r]^{\tduplicate} \ar@{=}[d]
+& \D[r]\,B \otimes \D[0]\,B \ar[d]^{\tid \otimes \tdrop}
 \\
@@
 \xymatrix@C=10ex{
-\D[q]\,B \otimes \D[r+s]\,B  \ar[d]_{\tid \otimes \tcontract}
-& \D[q+r+s]\,B \ar[l]_{\tcontract} \ar[r]^{\tcontract}
-& \D[q+r]\,B \otimes \D[s]\,B \ar[d]^{\tcontract \otimes \tid}
+\D[q]\,B \otimes \D[r+s]\,B  \ar[d]_{\tid \otimes \tduplicate}
+& \D[q+r+s]\,B \ar[l]_{\tduplicate} \ar[r]^{\tduplicate}
+& \D[q+r]\,B \otimes \D[s]\,B \ar[d]^{\tduplicate \otimes \tid}
 \\
@@
 \[
+\xymatrix@C=22ex{
+\D[(q+r)s]\,B \ar[r]^{\tdisplay} \ar[d]_{\tduplicate}
+& \D[q+r]\,(\D[s]\,B) \ar[d]^{\tduplicate}
+\\
+\D[qs]\,B \otimes \D[rs]\,B \ar[r]^{\tdisplay \otimes \tdisplay}
+& \D[q](\D[s]\,B) \otimes \D[r](\D[s]\,B)
+}
+\]
+\[
 \xymatrix{
-\D[qr]\,B \otimes \D[qs]\,B \ar[d]_{\duplicate \otimes \duplicate}
-  & \D[q(r+s)]\,B \ar[l]_{\tcontract} \ar[r]^{\duplicate}
-  & \D[q](\D[r+s]\,B) \ar[d]^{\fmap\tcontract}
+\D[qr]\,B \otimes \D[qs]\,B \ar[d]_{\display \otimes \display}
+  & \D[q(r+s)]\,B \ar[l]_{\tduplicate} \ar[r]^{\display}
+  & \D[q](\D[r+s]\,B) \ar[d]^{\fmap\tduplicate}
 \\
@@
 \]
+
+A $\D$-coalgebra is a functor $A : (\R,\leq) \to \CC$ with family of
+morphisms $\expose[r,s] : A_{rs} \to \D[r]\,A_s$ natural in $r$ and
+$s$.  For the $\D$-coalgebra $\D[\_]\,B$ the family $\expose$ is just
+$\tdisplay$, and the suitably generalizable laws for $\tdisplay$ are
+required to hold for $\texpose$:
+\[
+\xymatrix@C=10ex{
+A_r  \ar[d]_{\texpose} \ar@{=}[dr]
+\\
+\D[1]\,A_r \ar[r]^{\textract}
+  & A_r
+}
+\qquad
+\xymatrix@C=12ex{
+A_{qrs} \ar[d]_{\texpose} \ar[r]^{\texpose}
+  & \D[qr]\,A_s \ar[d]^{\tdisplay}
+\\
+\D[q]\,A_{rs} \ar[r]^{\tfmap\,\texpose}
+  & \D[q]\,(\D[r]\,A_s)
+}
+\]
+Further, $A$ needs to map the additive monoidal structure $(\R,+,0)$
+to the monoidal structure $(\CC,\otimes,\I)$ associated with the
+tensor product in $\CC$.  We choose to overload the names
+$\tduplicate$ and $\tdrop$ here to accommodate for the generalization
+from $\R$-comonads to $\R$-comonad algebras:
+\[
+\begin{array}{lcl}
+  \tduplicate & : & \Hom {A_{r+s}} {A_r \otimes A_s} \\
+  \tdrop      & : & \Hom {A_0} {\I} \\
+\end{array}
+\]
+The laws of $\tduplicate$ and $\tdrop$ for comonadic objects
+$\D[r]\,B$ immediately generalize to comonad algebras $A_r$:
+\[
+\xymatrix@C=12ex{
+A_{0} \otimes A_r \ar[d]_{\tdrop \otimes \tid}
+& A_r \ar[l]_{\tduplicate} \ar[r]^{\tduplicate} \ar@{=}[d]
+& A_r \otimes A_{0} \ar[d]^{\tid \otimes \tdrop}
+\\
+\I \otimes A_r \ar[r]^{\Gl}
+& A_r
+& A_r \otimes \I \ar[l]_{\Gr}
+}
+\]
+\[
+\xymatrix@C=10ex{
+A_q \otimes A_{r+s}  \ar[d]_{\tid \otimes \tduplicate}
+& A_{q+r+s} \ar[l]_{\tduplicate} \ar[r]^{\tduplicate}
+& A_{q+r} \otimes A_s \ar[d]^{\tduplicate \otimes \tid}
+\\
+A_q \otimes (A_r \otimes A_s) \ar[rr]^{\Ga}
+& & (A_q \otimes A_r) \otimes A_s
+}
+\]
 \[
 \xymatrix@C=22ex{
-\D[(q+r)s]\,B \ar[r]^{\tduplicate} \ar[d]_{\tcontract}
-& \D[q+r]\,(\D[s]\,B) \ar[d]^{\tcontract}
-\\
-\D[qs]\,B \otimes \D[rs]\,B \ar[r]^{\tduplicate \otimes \tduplicate}
-& \D[q](\D[s]\,B) \otimes \D[r](\D[s]\,B)
-}
-\]
+A_{(q+r)s} \ar[r]^{\texpose} \ar[d]_{\tduplicate}
+& \D[q+r]\,A_s \ar[d]^{\tduplicate}
+\\
+A_{qs} \otimes A_{rs} \ar[r]^{\texpose \otimes \texpose}
+& \D[q]A_s \otimes \D[r]A_s
+}
+\]
+\[
+\xymatrix{
+A_{qr} \otimes A_{qs} \ar[d]_{\expose \otimes \expose}
+  & A_{q(r+s)} \ar[l]_{\tduplicate} \ar[r]^{\expose}
+  & \D[q]\,A_{r+s} \ar[d]^{\fmap\tduplicate}
+\\
+\D[q]A_r \otimes \D[q]A_s \ar[rr]^{\tzip}
+&&
+\D[q](A_r \otimes A_s)
+}
+\]
+
+Graded comonad algebras are closed under pointwise sums $(\sumt I A)_r = \sumt {i:I} {A_{i,r}}$ and products $(\tupt I A)_r = \tupt {i:I} {A_{i,r}}$.
+\[
+\begin{array}{lcl}
+  \expose[\sumt I A] & : & \Hom {(\sumt I A)_{rs}} {\D[r](\sumt I A)_s} \\
+  \expose[\sumt I A] & = & \left[  \fmap{\iota_i} \comp \expose[A_i] \right]_{i:I}
+\\[1ex]
+  \tdrop[\sumt I A] & : & \Hom {(\sumt I A)_0} \I \\
+  \tdrop[\sumt I A] & = & \left[ \tdrop[A_i] \right]_{i:I}
+\\[1ex]
+  \duplicate[\sumt I A] & : & \Hom {(\sumt I A)_{r+s}} {(\sumt I A)_r \otimes (\sumt I A)_s} \\
+  \duplicate[\sumt I A] & = & \left[ (\iota_i \otimes \iota_i) \comp \duplicate[A_i] \right]_{i:I}
+\\[2ex]
+  \expose[\tupt I A] & : & \Hom {(\tupt I A)_{rs}} {\D[r](\tupt I A)_s} \\
+  \expose[\tupt I A] & = & \tzip \comp \tupt {i:I} {\expose[A_i]}
+\\[1ex]
+  \tdrop[\tupt I A] & : & \Hom {(\tupt I A)_0} \I \\
+  \tdrop[\tupt I A] & = & \isot \comp \tupt {i:I} {\tdrop[A_i]}
+\\[1ex]
+  \duplicate[\tupt I A] & : & \Hom {(\tupt I A)_{r+s}} {(\tupt I A)_r \otimes (\tupt I A)_s} \\
+  \duplicate[\tupt I A] & = & \isot \comp \tupt {i:I} {\duplicate[A_i]}
+\end{array}
+\]
+This enables us to interpret all positive types of CBPV as graded
+comonad coalgebras.
+
+\subsection{Structured graded comonads and their algebras}
+
+Coeffect, quantitative, and many modal type systems maintain a typing
+context where each variable $x$ is annotated by a resource qualifier
+$r$ in addition to its type $\tau$.  Such a typing context
+$(r_1x_1{:}\tau_1, \dots, r_nx_n{:}\tau_n) =: \gamma\Gamma$ can be
+split into a pure typing context
+$\Gamma = x_1{:}\tau_1, \dots, x_n{:}\tau_n$ and a resource context
+$\gamma = x_1{:}r_1, \dots, x_n{:}r_n$ such that
+$\dom \gamma = \dom \Gamma$.  We shall freely mix the two notations as
+it suits our purpose.
+
+In coeffect-graded CBPV, we wish to interpret each type in the context
+as a $\D$-coalgebra for a fixed graded comonad $\D$.  The
+interpretation $\den \Gamma$ of the context $\Gamma$ should then be a
+comonad coalgebra over the grading $\gamma$ such that judgements
+$\qCompTy \gamma \Gamma t N$ can be interpreted as morphisms
+$\dent t : \Hom {\Den\Gamma \gamma} {\denn N}$.  Since each variable
+comes with its own resource qualifier, we cannot simply model the
+context as a tensor product $\tupt {z:\dom\Gamma} \den{\Gamma(z)}$ of
+$\D$-coalgebras since this product would indexed by a single resource
+qualifier $r \in \R$ rather than a resource context
+$\gamma : \dom\Gamma \to \R$.  The solution offered by
+\citet{orchard:icfp14} are \emph{structured indexed comonads}.  We
+shall generalize this to comonad coalgebras to the extend needed for
+interpreting contexts.
+
+\citet{mcBride:wadler60} observed that resource contexts
+$\gamma : \R^I$ form a left $\R$-semimodule under pointwise addition
+$(\gamma + \delta)(i) = \gamma(i) + \delta(i)$ and scaling
+$(r \cdot \gamma)(i) = r \cdot \gamma(i)$.  Contexts $\Gamma$ can thus
+be interpreted as $R^{\dom \Gamma}$-graded $\D$-coalgebras $C$ with
+the following operations:
+\[
+\begin{array}{lcl}
+  \tdrop[C]     & : & \Hom {C_0} {\I} \\
+  \duplicate[C] & : & \Hom {C_{\gamma+\delta}} {C_\gamma \otimes C_\delta} \\
+  \duplicate[C] & : & \Hom {C_{\sumt I \gamma}} {\tupt {i:I} C_{\gamma_i}} \\
+  \expose[C]    & : & \Hom {C_{r\gamma}} {\D[r]\,C_\gamma} \\
+\end{array}
+\]
+The second, generalized form of $\tduplicate$ will be used to split
+resources accumulated from $I$ parties.
+
+We will also need to interpret context extension.  To this end, we
+shall employ a structured product
+$(C \boxtimes D)_{\gamma.\delta} = C_\gamma \otimes D_\delta$ where
+$\gamma : \R^I$ and $\delta : \R^J$ and thus $C \boxtimes D$ is a
+$\R^{I+J}$-graded $\D$-coalgebra.  Further, we implicitely use the
+isomorphism between $1$-structured $\D$-coalgebras $C : \R^1 \to \CC$
+and $\R$-graded $\D$-coalgebras $A : \R \to \CC$ and define
+\[
+\den{\ext \Gamma x P}
+  = \den \Gamma \boxtimes \denp P
+\]
+sweeping name issues under the carpet (to be handled by de Bruijn indices).
 
@@
   & \nbox{Computation types (negative types)} \\
+\Cxt & \ni & \Gamma
+  & ::= & \cempty \mid \ext \Gamma x P \mid \graybox{\ext \Gamma y N}
+  & \nbox{Typing context}
+\\
+\R & \ni & \graybox{r,s}
+   & ::= & 0 \mid 1 \mid r + s \mid rs
+   & \mbox{Resource qualifiers}
+\\
+\RCxt & \ni & \graybox{\gamma,\delta}
+  & ::= & \cempty \mid \ext \gamma x r \mid \ext \gamma y r
+  & \mbox{Resource context}
 \end{array}
@@
     }{\qCompTy \gamma \Gamma {\proj i t}  {N_i}}
-\\[1.5ex]
+%\\[1.5ex]
+\qquad
 \graybox{
@@
   \caption{Coeffect-graded call-by-push-value.}
-  \label{fig:cbpv}
+  \label{fig:coeff-cbpv}
 \end{figure}
+
+\subsection{Coeffect-graded CBPV: typing and semantics}
+
+We now have the mathematical structures in place to define a
+coeffect-graded variant of CBPV (see Fig.~\ref{coeff-cbpv}).
+The difference to pure CBPV is laid out in \emph{gray boxes}.
+
+The fundamental novelty is that the typing judgements
+$\qValTy \gamma \Gamma v P$ and $\qCompTy \gamma \Gamma t N$ are
+equipped with resource contexts $\gamma$ matching the typing contexts
+$\Gamma$.  A common pattern in the rules is that resource requirements
+of the subterms are added when both subterms are or maybe evaluated at
+runtime (rules \rintro\otimes, $\rlet^{+/-}$,
+\relim{\Box/\GS/\otimes/\diamond/\lolli}).  Note that in \relim\GS,
+the branches $t_i$ of the case statement (\relim\GS) share a resource
+context $\delta$, since only one of the branches is executed at
+runtime.  The different resource requirements of the branches can be
+subsumed under their maximum using rule \rweak.  Similarly, the
+components $t_i$ of a record (\rintro\Pi) share a resource context,
+since projection retrieves a single component (\relim\Pi).  This is
+the opposite of eager tuples (\rintro\otimes) where their eliminations
+makes all components available at the same time (\relim\otimes), thus,
+all have to be evaluated at runtime.
+
+Function types $\qfun r P N$ are now graded by a resource qualifier
+$r$ that specifies how the function argument is to be used in the
+function body.  In quantitative type with $\R = \Pot \bN$, qualifier
+$r$ gives the possible usage quantities of the function argument,
+e.g. $\{1\}$ for exactly one use (linear), $\{0,1\}$ for at most one
+use (affine), $\{0\}$ for no use (constant), and $\bN$ for arbitrary
+use (unrestricted).  A resource qualifier could also be a security
+level (public or private), and many other instances of $\R$ exist.  If
+a lambda abstraction $\lam x t$ is typed with $\qfun r P N$, qualifier
+$r$ is attached to variable $x$ in the resource context (rule
+\rintro\lolli).  If a function $t : \qfun r P N$ is applied (rule
+\relim\lolli), we need an $r$-qualified argument $v : rP$.  We could
+have given a qualified value typing judgement
+$\qValTy \gamma \Gamma v {rP}$ in the style of
+\citet{mcBride:wadler60} that---in the quantitative
+interpretation---provides $r$ copies of $v$ to be consumed by the
+function.  Such a judgement could come with a scaling rule
+\[
+\ru{\qValTy \gamma \Gamma v {sP}
+  }{\qValTy{r\gamma}\Gamma v {(rs)P}}
+\]
+that allow to scale the production of $v$ by $r$ if the resources are
+scaled accordingly (from $\gamma$ to $r\gamma$).  We bake scaling into
+the elimination rules instead.  Thus, $\relim\lolli$ receives an
+argument $\qValTy \delta \Gamma v P$, and to satisfy the demands of
+the function $\qCompTy \gamma \Gamma t {\qfun r P N}$, the resources
+$\delta$ for the argument are scaled by $r$, summing the resource
+requirements for the application to $\gamma + r\delta$.
+%
+Analogous scaling of the eliminatee is baked into the other value
+eliminators (\relim\Box, \relim\GS, \relim\otimes) and into the
+\rlet-rules.
+
+As a consequence of eliminatee scaling, $\relim\Box$ is no longer
+``natural-deduction'' style like projection, but ``sequent-calculus''
+style like $\relim\diamond$.  The original construct $\force v$ could
+eliminate an unscaled thunk $\qValTy \gamma \Gamma v {\Box N}$,
+however, computation typing $\qCompTy \gamma \Gamma {\force v} N$ has
+no scaling option, since negative types $N$ will not be interpreted as
+$\D$-coalgebras.  Scaling is semantically restricted to positive
+types, and in our type system, to context entries.  Thus, we extend
+CBPV to negative variables $y : N$ that come with a resource qualifier
+$r$ in the context and can there be interpreted as elements of
+$\D[r]\,\denn N$.  This is a modification also taken in call-by-need
+CBPV \citep{mcDermottMycroft:esop19}; this compromise allows us to
+keep CBPV's polarization of types where positive types stand for
+comonad coalgebras and negative types for monad algebras.
 

### 2020-07-07 popl21/popl21.tex
[ popl21 ] denotational coeffect grading

@@
 \newcommand{\denn}{\den[-]}
+\newcommand{\Denp}[2]{\denp{#1}_{#2}}
 \newcommand{\Denn}[2]{\denn{#1}_{#2}}
 \newcommand{\Den}[2]{\den{#1}_{#2}}
+\newcommand{\denv}[2]{\dent{#1}_{#2}}
 
@@
 \newcommand{\Gd}{\ensuremath{\delta}}
+\newcommand{\Gg}{\ensuremath{\gamma}}
 \newcommand{\Gl}{\ensuremath{\lambda}}
@@
 \newcommand{\fmap}[2][]{\tfmap_{#1}\,#2}
+\newcommand{\fmapt}[2][]{\tfmap_{#1}#2}
 \newcommand{\treturn}{\mathsf{return}}
@@
 \newcommand{\tcast}{\mathsf{cast}}
+\newcommand{\tdist}{\mathsf{dist}}
 \newcommand{\Id}{\mathsf{Id}}
-\newcommand{\tid}{\mathsf{id}}
+\newcommand{\tid}[1][]{\mathsf{id}_{#1}}
 \newcommand{\teval}{\mathsf{eval}}
@@
 \newcommand{\sumt}   [2]{\ensuremath{\Sigma_{#1} #2}}
-\newcommand{\tupt}   [2]{\ensuremath{\otimes_{#1} #2}}
+\newcommand{\tupt}   [2]{\ensuremath{\mathop\otimes_{#1} #2}}
 \newcommand{\rect}   [2]{\ensuremath{\Pi_{#1} #2}}
 \newcommand{\sumty}  [3]{\ensuremath{\Sigma_{#1:#2} #3_{#1}}}
-\newcommand{\tupty}  [3]{\ensuremath{\otimes_{#1:#2} #3_{#1}}}
+\newcommand{\tupty}  [3]{\ensuremath{\mathop\otimes_{#1:#2} #3_{#1}}}
 \newcommand{\recty}  [3]{\ensuremath{\Pi_{#1:#2} #3_{#1}}}
@@
 \subsection{Effect-graded CBPV: syntax and typing}
+\label{sec:effect-cbpv}
 
@@
 $(r \cdot \gamma)(i) = r \cdot \gamma(i)$.  Contexts $\Gamma$ can thus
-be interpreted as $R^{\dom \Gamma}$-graded $\D$-coalgebras $C$ with
+be interpreted as \emph{$R^{\dom \Gamma}$-graded}
+(or \emph{structured}) $\D$-coalgebras $C$ with
 the following operations:
@@
 The second, generalized form of $\tduplicate$ will be used to split
-resources accumulated from $I$ parties.
+resources accumulated from $I$ parties.  Note that
+$A_r := C_{r\gamma}$ flattens a structured $\D$-coalgebra $C$ into an
+ordinary one ($A$).
 
@@
   = \den \Gamma \boxtimes \denp P
+  ,
 \]
@@
 \NTy & \ni & N
-  & ::= & \diamond P % \graybox{\compty r P}
+  & ::= & %\diamond P %
+          \graybox{\compty r P}
      \mid \gqfun r P N \mid \recty i I N
@@
 \\ &&& \mid &
-         \ret v        \mid \qbind {\graybox r} x u t
+         \ret v
+    \mid \bind x u t
+    % \mid \qbind {\graybox r} x u t
   & \nbox{monad operations}
@@
  \nru{\rlet^-}
-     {\qCompTy \gamma \Gamma u N \qquad
-      \CompTy {\qext \delta \Gamma r y N} t {N'}
-    }{\gpCompTy {r\gamma + \delta} \Gamma {\qlet r y u t} {N'}}
+     {\qCompTy \delta \Gamma u N \qquad
+      \CompTy {\qext \gamma \Gamma r y N} t {N'}
+    }{\gpCompTy {\gamma + r\delta} \Gamma {\qlet r y u t} {N'}}
 }
@@
  \nru{\rlet^+}
-     {\gqValTy \gamma \Gamma v P \qquad
-      \CompTy {\gqext \delta \Gamma r x P} t N
-    }{\gpCompTy {r\gamma + \delta} \Gamma {\gqlet r x v t} N}
+     {\gqValTy \delta \Gamma v P \qquad
+      \CompTy {\gqext \gamma \Gamma r x P} t N
+    }{\gpCompTy {\gamma + r\delta} \Gamma {\gqlet r x v t} N}
 \qquad
  \nru{\relim\Box}
-     {\gqValTy \gamma \Gamma v {\Box N} \qquad
-      \CompTy {\gqext \delta \Gamma r y N} t {N'}
-    }{\gpCompTy {r\gamma + \delta} \Gamma {\gqforced r v y t} {N'}}
+     {\gqValTy \delta \Gamma v {\Box N} \qquad
+      \CompTy {\gqext \gamma \Gamma r y N} t {N'}
+    }{\gpCompTy {\gamma + r\delta} \Gamma {\gqforced r v y t} {N'}}
 \\[1.5ex]
  \nru{\relim\GS}
-     {\gqValTy \gamma \Gamma v {\sumt I P} \qquad
-      \forall i{:}I,\ \CompTy {\gqext \delta \Gamma r {x_i} {P_i}} {t_i} N
-    }{\gpCompTy {r\gamma + \delta} \Gamma {\caser v {\{\graybox{r} x_i.\, t_i\}_{i:I}}} N}
+     {\gqValTy \delta \Gamma v {\sumt I P} \qquad
+      \forall i{:}I,\ \CompTy {\gqext \gamma \Gamma r {x_i} {P_i}} {t_i} N
+    }{\gpCompTy {\gamma + r\delta} \Gamma {\caser v {\{\graybox{r} x_i.\, t_i\}_{i:I}}} N}
 \qquad
  \nru{\relim\otimes}
-     {\gqValTy \gamma \Gamma v {\tupt I P} \qquad
+     {\gqValTy \delta \Gamma v {\tupt I P} \qquad
       \gqCompTy \delta {\extr \Gamma {\overline{\graybox{r}x_i{:}P_i}^{i:I}}} t N
-    }{\gpCompTy {r\gamma + \delta} \Gamma {\gqsplits r v {\bar x} t} N}
+    }{\gpCompTy {\gamma + r\delta} \Gamma {\gqsplits r v {\bar x} t} N}
 \\[1.5ex]
@@
      {\qValTy \gamma \Gamma v P
-    }{\qCompTy \gamma \Gamma {\ret v} {\diamond P}}
+    }{\qCompTy {\graybox r\gamma} \Gamma {\ret v} {\compty {\graybox r} P}}
 \qquad
  \nru{\relim\diamond}
-     {\gqCompTy \gamma \Gamma u {\diamond P} \qquad
-      \CompTy {\gqext \delta \Gamma r x P} t  N
-    }{\gpCompTy {r\gamma + \delta} \Gamma {\gqbind r x u t} N}
+     {\gqCompTy \delta \Gamma u {\compty {\graybox r} P} \qquad
+      \CompTy {\gqext \gamma \Gamma r x P} t  N
+    }{\gpCompTy {\gamma + \delta} \Gamma {\bind x u t} N}
+%  \nru{\rintro\diamond}
+%      {\qValTy \gamma \Gamma v P
+%     }{\qCompTy \gamma \Gamma {\ret v} {\diamond P}}
+% \qquad
+%  \nru{\relim\diamond}
+%      {\gqCompTy \gamma \Gamma u {\diamond P} \qquad
+%       \CompTy {\gqext \delta \Gamma r x P} t  N
+%     }{\gpCompTy {r\gamma + \delta} \Gamma {\gqbind r x u t} N}
 \\[1.5ex]
@@
 
-\subsection{Coeffect-graded CBPV: typing and semantics}
+\subsection{Coeffect-graded CBPV: syntax and typing}
 
 We now have the mathematical structures in place to define a
-coeffect-graded variant of CBPV (see Fig.~\ref{coeff-cbpv}).
+coeffect-graded variant of CBPV (see Fig.~\ref{fig:coeff-cbpv}).
 The difference to pure CBPV is laid out in \emph{gray boxes}.
@@
 the branches $t_i$ of the case statement (\relim\GS) share a resource
-context $\delta$, since only one of the branches is executed at
-runtime.  The different resource requirements of the branches can be
-subsumed under their maximum using rule \rweak.  Similarly, the
-components $t_i$ of a record (\rintro\Pi) share a resource context,
-since projection retrieves a single component (\relim\Pi).  This is
-the opposite of eager tuples (\rintro\otimes) where their eliminations
+context $\gamma$, since only one of the branches is executed at
+runtime.  Via rule $\rweak$,
+the different resource requirements of the branches can be
+subsumed under their maximum wrt.\ $\leq_\R$.  Similarly, the
+components $t_i$ of a record (\rintro\Pi) share a resource context
+since projection only retrieves one of the components (\relim\Pi).  This is
+the opposite of eager tuples (\rintro\otimes) where the elimination
 makes all components available at the same time (\relim\otimes), thus,
-all have to be evaluated at runtime.
+all components have to be evaluated at runtime.
 
@@
 $r$ that specifies how the function argument is to be used in the
-function body.  In quantitative type with $\R = \Pot \bN$, qualifier
-$r$ gives the possible usage quantities of the function argument,
-e.g. $\{1\}$ for exactly one use (linear), $\{0,1\}$ for at most one
-use (affine), $\{0\}$ for no use (constant), and $\bN$ for arbitrary
-use (unrestricted).  A resource qualifier could also be a security
-level (public or private), and many other instances of $\R$ exist.  If
-a lambda abstraction $\lam x t$ is typed with $\qfun r P N$, qualifier
-$r$ is attached to variable $x$ in the resource context (rule
-\rintro\lolli).  If a function $t : \qfun r P N$ is applied (rule
-\relim\lolli), we need an $r$-qualified argument $v : rP$.  We could
-have given a qualified value typing judgement
-$\qValTy \gamma \Gamma v {rP}$ in the style of
+function body.  (This type is often written $!_{r}P \lolli N$.)  In
+quantitative typing with $\R = \Pot \bN$, qualifier $r$ gives the
+possible usage quantities of the function argument, e.g. $\{1\}$ for
+exactly one use (linear), $\{0,1\}$ for at most one use (affine),
+$\{0\}$ for no use (constant), and $\bN$ for arbitrary use
+(unrestricted).  A resource qualifier could also be a security level
+(public or private), or a sensitivity level (a non-negative real)
+\citep{reedPeirce:icfp10}.  If a lambda abstraction $\lam x t$ is
+typed with $\qfun r P N$, qualifier $r$ is attached to variable $x$ in
+the resource context (rule \rintro\lolli).  If a function
+$t : \qfun r P N$ is applied (rule \relim\lolli), we need an
+$r$-qualified argument $v : rP$.  We could have given a qualified
+value typing judgement $\qValTy \gamma \Gamma v {rP}$ in the style of
 \citet{mcBride:wadler60} that---in the quantitative
@@
 \]
-that allow to scale the production of $v$ by $r$ if the resources are
-scaled accordingly (from $\gamma$ to $r\gamma$).  We bake scaling into
-the elimination rules instead.  Thus, $\relim\lolli$ receives an
+that allows to scale the production of $v$ by $r$ if the resources are
+scaled accordingly (from $\gamma$ to $r\gamma$).  We bake scaling instead
+into the transition from values to computations.
+Thus, $\relim\lolli$ receives an
 argument $\qValTy \delta \Gamma v P$, and to satisfy the demands of
@@
 Analogous scaling of the eliminatee is baked into the other value
-eliminators (\relim\Box, \relim\GS, \relim\otimes) and into the
-\rlet-rules.
+eliminators (\relim\Box, \relim\GS, \relim\otimes), into the
+\rlet-rules, and into \rintro\diamond.
+
+The monadic type $\compty r P$ records the multiplicity $r$ of the
+value of type $P$ resulting from the computation.  This construction
+is dual to the comonadic type $\thunkty e N$ from effect-graded CBPV
+(Section~\ref{sec:effect-cbpv}).  Since negative types are not
+$\D$-coalgebras and do not support scaling, the scaling in
+$\rintro\diamond$ is the last opportunity for scaling before entering
+the monad.  Typing of bind ($\bind x u t$) attaches the resource qualifier $r$ store in $\compty r P$ to the variable $x$ (\relim\diamond).
 
@@
 
+\subsection{Coeffect-graded CBPV: denotational semantics}
+
+%\subsubsection{Interpretation of types}
+
+Positive types $P$ are interpreted as $\D$-algebras
+$\denp P : (\R,\leq) \to \CC$ and negative types $N$ as objects
+$\denn N : \CC$, except for in contexts, where they are interpreted as
+$\D$-algebras $\Den N r = \D[r]\,\denn N$.  With $\den P = \denp P$,
+the interpretation of contexts is thus
+$\den \Gamma = \boxtimes_{x:\dom\Gamma} \den{\Gamma(x)}$.
+\[
+\begin{array}{lcl}
+  \Denp\_\_ & : & \PTy \to [(\R,\leq) \to \CC] \\
+  \Denp{\Box N}    r & = & \D[r]\, \denn N \\
+  \Denp{\sumt I P} r & = & \coprod_{i:I} \Denp{P_i} r \\
+  \Denp{\tupt I P} r & = & \otimes_{i:I} \Denp{P_i} r \\
+\end{array}
+\qquad\qquad
+\begin{array}{lcl}
+  \denn\_\_ & : & \NTy \to \CC \\
+  \denn{\compty r P} & = & \T \Denp P r \\
+  \denn{\qfun r P N} & = & \Denp P r \lolli \denn N \\
+  \denn{\rect I N}   & = & \prod_{i:I} \denn{N_i} \\
+\end{array}
+\]
+The symmetric monoidal category $\CC$ needs to come with cartesian
+products $\rect I B$ as well as with distributive coproducts
+$\coprod_I A$ with
+$\tdist : \Hom {C \otimes \coprod_I A} {\coprod_{i:I} {(C \otimes A_i)}}$.
+
+Computations $\qCompTy \gamma \Gamma t N$ are interpreted as morphisms
+$\dent t : \Hom {\Den\Gamma\gamma} {\denn N}$ and values
+$\qValTy \gamma \Gamma v P$ as families
+$\denv v r : \Hom {\Den\Gamma{r\gamma}} {\Denp P r}$ natural in $r$.
+Naturality here means that
+$\Denp P {r \leq s} \comp \denv v r = \denv v s \comp \left(\Den \Gamma {(\_
+    \cdot \gamma)}\right)_{r \leq s}$.
+Again, because of $\rweak$, we interpret typing derivations rather than terms.
+The interpretation is now rather straightforward, but we spell it out for reference.
+\[
+\begin{array}{llcl}
+\rvar^+
+  & \denv {\ValTy{\qext 0 \Gamma 1 x P} x P} r
+  & = & \Gr \comp (\tdrop[\den\Gamma] \otimes \tid[\Denp P r])
+\\
+\rintro\Box
+  & \denv {\qValTy \gamma \Gamma {\thunk t} {\Box N}} r
+  & = & \fmapt[{\D[r]}]{\dent t} \comp \expose{\den \Gamma}
+\\
+\rintro\GS
+  & \denv{\qValTy \gamma \Gamma {\inj i v} {\sumt I P}} r
+  & = & \iota_i \comp \denv v r
+\\
+\rintro\otimes
+  & \denv{\pValTy {\sumt I \gamma} \Gamma {\tup v} {\tupt I P}} r
+  & = & \tupt {i:I} \denv {v_i}r \comp \duplicate
+\\
+\end{array}
+\]
+\[
+\begin{array}{llcl}
+\rvar^-
+  & \dent {\oCompTy{\qext 0 \Gamma 1 y N} y N}
+  & = & \Gr \comp (\tdrop[\den\Gamma] \otimes \extract[\denn N])
+\\
+\rlet^{-}
+  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qlet r y u t} {N'}}
+  & = & \dent t
+  \comp (\tid[\Den \Gamma \gamma] \otimes
+           (\fmapt[{\D[r]}] \dent u \comp \expose[\den \Gamma]))
+\\ &&& {~}
+  \comp \duplicate
+\\
+\rlet^{+}
+  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qlet r x v t} {N}}
+  & = & \dent t
+  \comp (\tid[\Den \Gamma \gamma] \otimes
+           (\expose[\denp P] \comp \denv v r))
+  \comp \duplicate
+\\
+\relim\Box
+  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qforced r v y t} {N'}}
+  & = & \dent t
+  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
+  \comp \duplicate
+\\
+\relim\GS
+  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\caser v \{r x_i.\, t_i\}_{i:I}} {N}}
+  & = & [\dent {t_i}]_{i:I} \comp \tdist
+  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
+  \comp \duplicate
+\\
+\relim\otimes
+  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qsplits r v {\bar x} t} {N}}
+  & = & \dent t \comp \isot
+  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
+  \comp \duplicate
+\\
+\rintro\diamond
+  & \dent {\pCompTy {r\gamma} \Gamma {\ret v} {\compty r P}}
+  & = & \treturn \comp \denv v r
+\\
+\relim\diamond
+  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qbind r x u t} {N}}
+  & = & \run[\den N] \comp \fmapt[\T]{\dent t} \comp \strengthl
+\\ &&& {~}
+  \comp (\tid[\Den \Gamma \gamma] \otimes \dent u)
+  \comp \duplicate
+\\
+\rintro\lolli
+  & \dent {\qCompTy \gamma \Gamma {\lam x t} {\qfun r P N}}
+  & = & \Lambda \dent t
+\\
+\relim\lolli
+  & \dent{\pCompTy {\gamma + r\delta} \Gamma {\app t v} N}
+  & = & \teval \comp (\dent t \otimes \denv v r) \comp \duplicate
+\\
+\rintro\Pi
+  & \dent {\qCompTy \gamma \Gamma {\recrd I t} {\rect I N}}
+  & = & \langle \dent {t_i} \rangle_{i:I}
+\\
+\relim\Pi
+  & \dent {\qCompTy \gamma \Gamma {\proj i t} {N_i}}
+  & = & \pi_i \comp \dent t
+\\
+\rweak
+  & \dent {\qCompTy \gamma \Gamma t N}
+  & = & \dent t \comp \Den \Gamma {\gamma \leq \gamma'}
+\\
+\end{array}
+\]
+
 %% Acknowledgments
@@
 
-%% Appendix
-\appendix
-\section{Appendix}
-
-Text of appendix \ldots
+% %% Appendix
+% \appendix
+% \section{Appendix}
+
+% Text of appendix \ldots
 

### 2020-07-08 popl21/popl21.tex
[ popl21 ] substitution

@@
 \usepackage{tcolorbox}
-\newtcbox{\grayboxtext}{on line,colback=gray!20,colframe=gray!20,size=fbox,arc=0pt,boxrule=0pt}%,beforeafter skip=0pt,leftright skip=0pt}
+\newtcbox{\grayboxtext}{on line,colback=gray!50,colframe=gray!50,size=fbox,arc=0pt,boxrule=0pt}%,beforeafter skip=0pt,leftright skip=0pt}
 \newcommand{\graybox}[1]{\grayboxtext{$#1$}}
@@
 \newcommand{\loccit}{\emph{loc.\,cit.}}
+\newcommand{\cf}{cf.\ }
 \newcommand{\den}[2][]{\llbracket#2\rrbracket^{#1}}
@@
 \newcommand{\defiff}{\mathrel{{{:}{\Longleftrightarrow}}}}
+\newcommand{\subst}[3]{#3[#1/#2]}
+\newcommand{\qsubst}[4]{#4[#1/#2#3]}
 
@@
 \newcommand{\cempty}{\emptyset}
+\newcommand{\sempty}{\emptyset}
+\newcommand{\sext}[3]{#1,#3/#2}
 \newcommand{\extr}[2]{#1.#2}
@@
 % Judgements
+\newcommand{\SubstTy}[4]{#1#2 \vdash #3 : #4}
+\newcommand{\pSubstTy}[4]{(#1)#2 \vdash (#3) : (#4)}
 \newcommand{\ValTy}[3]{#1 \vdash #2 : #3}
@@
 \newcommand{\rulename}[1]{\ensuremath{\mbox{\textsc{#1}}}\xspace}
+\newcommand{\rbeta}[1]{\ensuremath{\beta\mbox{-}\mathord{#1}}\xspace}
+\newcommand{\reta}[1]{\ensuremath{\eta\mbox{-}\mathord{#1}}\xspace}
+\newcommand{\rpi}[1]{\ensuremath{\pi\mbox{-}\mathord{#1}}\xspace}
 \newcommand{\rintro}[1]{\ensuremath{\mathord{#1}\mbox{-\rulename{intro}}}\xspace}
@@
 \]
-The graded $\T$-algebra structure $\run{\denn N}$ is constructed by
+The graded $\T$-algebra structure $\run[\denn N]$ is constructed by
 induction on $N$:
@@
 \begin{array}{lcl}
-  \run{\denn N} & : & \Hom {\T[e_1](\Denn N {e_2})} {\Denn N {e_1 \bu e_2}} \\
-  \run{\denn {\diamond P}} & = & \tjoin \\
-  \run{\denn {P \To N}}   & = & \Lambda(\run{\denn N} \circ
+  \run[\denn N] & : & \Hom {\T[e_1](\Denn N {e_2})} {\Denn N {e_1 \bu e_2}} \\
+  \run[\denn {\diamond P}] & = & \tjoin \\
+  \run[\denn {P \To N}]   & = & \Lambda(\run[\denn N] \circ
                                 \fmap[\T]\teval \circ \strengthr) \\
-  \run{\denn {\rect I N}} & = & \langle \run{\denn{N_i}} \comp
+  \run[\denn {\rect I N}] & = & \langle \run[\denn{N_i}] \comp
                                 \fmap[\T]{\pi_i} \rangle_{i:I} \\
@@
 \relim\diamond  & \dent{\bind x u t} & = &
-  \run{\denn N} \comp \fmap[\T]\teval \comp \strengthl \comp
+  \run[\denn N] \comp \fmap[\T]\teval \comp \strengthl \comp
   \langle \Lambda\dent t,\,\dent u \rangle
@@
 \newcommand{\gqfun}[3]{\graybox{#1#2 \lolli #3}}
-\renewcommand{\CompTy}[3]{#1 \vdash #2 : #3}
+%\renewcommand{\CompTy}[3]{#1 \vdash #2 : #3}
 \newcommand{\nbox}[1]{}
@@
      {\qCompTy \delta \Gamma u N \qquad
-      \CompTy {\qext \gamma \Gamma r y N} t {N'}
+      \oCompTy {\qext \gamma \Gamma r y N} t {N'}
     }{\gpCompTy {\gamma + r\delta} \Gamma {\qlet r y u t} {N'}}
@@
      {\gqValTy \delta \Gamma v P \qquad
-      \CompTy {\gqext \gamma \Gamma r x P} t N
+      \oCompTy {\gqext \gamma \Gamma r x P} t N
     }{\gpCompTy {\gamma + r\delta} \Gamma {\gqlet r x v t} N}
@@
      {\gqValTy \delta \Gamma v {\Box N} \qquad
-      \CompTy {\gqext \gamma \Gamma r y N} t {N'}
+      \oCompTy {\gqext \gamma \Gamma r y N} t {N'}
     }{\gpCompTy {\gamma + r\delta} \Gamma {\gqforced r v y t} {N'}}
@@
  \nru{\relim\GS}
-     {\gqValTy \delta \Gamma v {\sumt I P} \qquad
-      \forall i{:}I,\ \CompTy {\gqext \gamma \Gamma r {x_i} {P_i}} {t_i} N
+     {\gqValTy \delta \Gamma v {\sumt I P} \quad
+      \forall i{:}I,\ \oCompTy {\gqext \gamma \Gamma r {x_i} {P_i}} {t_i} N
     }{\gpCompTy {\gamma + r\delta} \Gamma {\caser v {\{\graybox{r} x_i.\, t_i\}_{i:I}}} N}
-\qquad
+\quad
  \nru{\relim\otimes}
-     {\gqValTy \delta \Gamma v {\tupt I P} \qquad
-      \gqCompTy \delta {\extr \Gamma {\overline{\graybox{r}x_i{:}P_i}^{i:I}}} t N
+     {\gqValTy \delta \Gamma v {\tupt I P} \quad
+      \gqCompTy \gamma {\extr \Gamma {\overline{\graybox{r}x_i{:}P_i}^{i:I}}} t N
     }{\gpCompTy {\gamma + r\delta} \Gamma {\gqsplits r v {\bar x} t} N}
@@
      {\gqCompTy \delta \Gamma u {\compty {\graybox r} P} \qquad
-      \CompTy {\gqext \gamma \Gamma r x P} t  N
+      \oCompTy {\gqext \gamma \Gamma r x P} t  N
     }{\gpCompTy {\gamma + \delta} \Gamma {\bind x u t} N}
@@
  \nru{\rintro\lolli}
-     {\CompTy {\gqext \gamma \Gamma r x P} t N
+     {\oCompTy {\gqext \gamma \Gamma r x P} t N
     }{\gqCompTy \gamma \Gamma {\lam x t} {\gqfun r P N}}
@@
   & \denv {\qValTy \gamma \Gamma {\thunk t} {\Box N}} r
-  & = & \fmapt[{\D[r]}]{\dent t} \comp \expose{\den \Gamma}
+  & = & \fmapt[{\D[r]}]{\dent t} \comp \expose[\den \Gamma]
 \\
@@
   & = & \dent t
-  \comp (\tid[\Den \Gamma \gamma] \otimes
-           (\expose[\denp P] \comp \denv v r))
+  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
   \comp \duplicate
@@
 \]
+
+It is remarkable that coeffect-graded CBPV works \emph{without any
+  distributive law} \cite{orchard:icfp16}
+for the monad $\T$ and the graded comonad $\D$.
+%
+This was a crucial design criterion, leading to resource qualifiers in
+the monadic type $\compty r P$.
+%
+A distributive law would be required if we allowed scaling of
+computations, not just of values.
+
+\subsection{Coeffect-graded CBPV: equational theory and operational semantics}
+
+The equational theory of pure CBPV comprises $\beta$ and $\eta$ laws,
+and \emph{sequencing laws} that syntactically express that negative
+types are monad algebras \citep[Fig.~11]{levi:hosc06}.  The sequencing
+laws contain a generalization of the associativity law for monads and
+allow to permute a bind under a $\lambda$ or record construction.  In
+the presence of $\eta$ for functions and records, they are
+interderivable with the following permutation laws:
+\[
+\begin{array}{llcl}
+\rpi\diamond
+  & \bind {x_2} {(\bind {x_1} {t_1} {t_2})} {t_3}
+  & = & \bind {x_1} {t_1} {(\bind {x_2} {t_2} {t_3})}
+\\
+\rpi\lolli
+  & \app {(\bind x u t)} v
+  & = & \bind x u {\app t v}
+\\
+\rpi\Pi
+  & \proj i {(\bind x  u t)}
+  & = & \bind x u {\proj i t}
+\end{array}
+\]
+Note that permutations for let-bindings like
+$\app{(\qlet r x v t)}w = \qlet r x v {\app t w}$ are instances of the
+$\beta$-law $\qlet r x v t = \qsubst v r x t$.
+%
+Herein, \fbox{$\qsubst v r x t$} shall the denote the
+(capture-avoiding) replacement $\subst v x t$ of a positive variable
+$x$ annotated by $r$ by value $v$ in term $t$.
+%
+Permutations for value-eliminations like
+$\app {(\qsplits r v {\bar x} t)} w = \qsplits r v {\bar x} {\app t
+  w}$
+are derivable (using the $\beta$-laws) from the $\eta$-laws, like
+$\qsubst{v}{r}{z}{t'} = \qsplits r {v} {\bar x} {\qsubst {\ptup {\bar
+      x}} r z {t'}}$ for $\otimes$.
+
+Our extension of CBPV features negative variables $y$ standing for
+computations $u$, thus, we require a notion of computation
+substitution $\qsubst u r t y$ that should have the following semantics.
+\begin{align*}
+&
+  \ru{\qCompTy \delta \Gamma u N \qquad
+      \oCompTy {\qext \gamma \Gamma r y N} t {N'}
+    }{\pCompTy {\gamma + r\delta} \Gamma {\qsubst u r t y} {N'}}
+\\ &
+  \dent{\qsubst u r t y} = \dent t
+    \comp (\tid[\den\Gamma \gamma]
+             \otimes (\fmapt[{\D[r]}] \dent u \comp \expose[\den\Gamma]))
+    \comp \duplicate[\den\Gamma]
+\end{align*}
+
+Further, we add $\beta$-laws for the negative
+let-binding and $\relim\Box$, since $\tforce$ now has let-style elimination:
+\[
+\begin{array}{llcl}
+\rbeta{\rlet^{-}}
+  & \qlet r y u t
+  & = & \qsubst u r y t
+\\
+\rbeta\Box
+  & \qforced r {\thunk u} y t
+  & = & \qsubst u r y t
+\\
+\end{array}
+\]
+These laws are immediately sound, since left and right hand sides of
+these equations have the same denotation.
+%
+Read as reduction rules, these laws are added to the operational
+semantics of CBPV, replacing $\beta$-law $\force{(\thunk u)} = u$.
+
+Substitution for coeffect-graded type system has been worked
+out in detail by \citet{atkeyWood:types19}.  It straightforwardly extends to
+coeffect-graded CBPV.
+%
+A substitution $\sigma$ is a finite map from variable to terms,
+mapping positive variables $x$ to values $v$ and negative variables
+$y$ to computations $t$.  Substitution typing
+\fbox{$\SubstTy \Psi \Gamma \sigma \Delta$} is equipped with a matrix
+$\Psi \in \R^{\dom\Gamma \times \dom\Delta}$ recording the usage
+vector $\Psi(z) \in \R^{\dom\Gamma}$ for each variable
+$z \in \dom\Delta = \dom \sigma$ in the domain of the substitution.
+\[
+  \ru{
+    }{\SubstTy {0} \Gamma \sempty \cempty}
+\qquad
+  \ru{\SubstTy \Psi \Gamma \sigma \Delta \qquad
+      \qValTy \gamma \Gamma v P
+    }{\pSubstTy {\Psi,\gamma} \Gamma {\sext \sigma x v} {\ext \Delta x P}}
+\qquad
+  \ru{\SubstTy \Psi \Gamma \sigma \Delta \qquad
+      \qCompTy \gamma \Gamma t N
+    }{\pSubstTy {\Psi,\gamma} \Gamma {\sext \sigma y t} {\ext \Delta y N}}
+\]
+Capture-avoiding parallel substitution into values \fbox{$v \sigma$} and
+computations \fbox{$t \sigma$} is defined as usual by recursion on the term.
+
+The matrix $\Psi : \R^{\dom \Gamma \times \dom \Delta}$ acts as a
+linear map $\R^{\dom \Delta} \to \R^{\dom \Gamma}$ and allows us to
+state the substitution theorem.
+\begin{theorem}[Substitution preserves typing]
+  Let $\SubstTy \Psi \Gamma \sigma \Delta$.
+  \begin{enumerate}
+  \item If $\qValTy \delta \Delta v P$ then $\pValTy {\Psi\delta} \Gamma {v\sigma} P$.
+  \item If $\qCompTy \delta \Delta t N$ then
+    $\pCompTy {\Psi\delta} \Gamma {t\sigma} N$.
+  \end{enumerate}
+\end{theorem}
+\begin{proof}
+  Straightforward adaptation of \citet{atkeyWood:types19}.
+\end{proof}
+\begin{corollary}[Subject reduction]
+  $\beta$-reduction of coeffect-graded CBPV preserves types.
+\end{corollary}
+In the pure (effect-free) version, every computation at value type
+returns a value.
+\begin{conjecture}[Canonicity of pure coeffect-graded CBPV]
+  If $\vdash t : \diamond P$ then $t =_{\beta} \ret v$ for some $\vdash v : P$.
+\end{conjecture}
+This normalization result can be proved with standard techniques such
+as reducibility candidates or normalization by evaluation (\cf
+\citet{abelSattler:ppdp19}).
+
+The denotation of a substitution $\SubstTy \Psi \Gamma \sigma \Delta$
+is a natural transformation
+$\denv \sigma \delta : \Hom {\Den \Gamma {\Psi\delta}} {\Den \Delta
+  \delta}$
+such that
+$
+\Den \Delta {\delta \leq \delta'}
+\comp
+\denv \sigma \delta
+=
+\denv \sigma {\delta'}
+\comp
+\Den \Gamma {\Psi\delta \leq \Psi\delta'}
+$.
+It is defined by recursion as follows:
+\begin{align*}
+& \ru{
+    }{\SubstTy {0} \Gamma \sempty \cempty}
+& \dent \sempty
+& : \Hom {\Den\Gamma 0} \I
+& & = \tdrop[\den\Gamma]
+\\[2ex]
+& \ru{\SubstTy \Psi \Gamma \sigma \Delta \qquad
+      \qValTy \gamma \Gamma v P
+    }{\pSubstTy {\Psi,\gamma} \Gamma {\sext \sigma x v} {\ext \Delta x P}}
+& \dent{\sext \sigma x v}
+& : \Hom  {\Den\Gamma{\Psi\delta,r\gamma}} {\Den \Delta \delta \otimes \Denp P r}
+& & = \denv \sigma \delta \otimes \denv v r
+\\[2ex]
+& \ru{\SubstTy \Psi \Gamma \sigma \Delta \qquad
+      \qCompTy \gamma \Gamma t N
+    }{\pSubstTy {\Psi,\gamma} \Gamma {\sext \sigma y t} {\ext \Delta y N}}
+\end{align*}
+
+\section{Fully graded CBPV}
+\label{sec:full-cbpv}
+
+\newcommand{\peCompTy}[5]{(#1)#2 \vdash #3 : #5 \mid #4}
+\newcommand{\qeCompTy}[5]{#1#2 \vdash #3 : #5 \mid #4}
+% \newcommand{\eext}[4]{#1.#2{:}#4#3}
+% \newcommand{\qeext}[6]{#1#2.#3#4{:}#6#5}
+\newcommand{\eext}[4]{#1\mathbin.#2:#4|#3}
+\newcommand{\qeext}[6]{#1#2\mathbin.#3#4:#6|#5}
+
+\newcommand{\gCompTy}[4]{\CompTy{#1}{#2}{\graybox{#3}}{#4}}
+\newcommand{\gpeCompTy}[5]{(#1)#2 \vdash #3 : #5 \mid {\graybox{#4}}}
+\newcommand{\gqeCompTy}[5]{#1#2 \vdash #3 : #5 \mid {\graybox{#4}}}
+\newcommand{\gthunkty}[2]{\thunkty{\graybox{#1}}{#2}}
+\newcommand{\geext}[4]{\eext{#1}{#2}{\graybox{#3}}{#4}}
+\newcommand{\gqeext}[6]{\qeext{#1}{#2}{#3}{#4}{\graybox{#5}}{#6}}
+
+In this section we finally present a \emph{fully graded} version of
+CBPV where both effects and coeffects are graded.
+Fig.~\ref{fig:full-cbpv} presents its types and typing rules where the
+differences to the coeffect-graded version are
+\graybox{\mbox{highlighted}}.
+%
+We reintroduced effect grading into the computation typing
+$\qeCompTy \gamma \Gamma t e N$.  As in the effect-graded version,
+effects are accumulated in a single place, the monadic bind construct
+(\relim\diamond).  The negative variables $y : N$ we introduced for
+coeffect grading are now annotated with an effect qualifier $e$ in the
+typing context $\Gamma$.  The full information about a negative
+variable $y$ in a graded context $\gamma\Gamma$ is type $N$, effect
+$e$ and coeffect $r$.  This combination is interpreted as
+$\D[r]\,\denn N e$ where the resource qualifier acts as ``multiplier''
+for the effect qualifier $e$.  However, we never have to carry out
+this ``multiplication'', thus, do not require a law how the comonad
+$\D$ distributes over the monad algebra $N$.  The reason is that only
+variables $y$ with $r=1$ can be used from the context $(\rvar^{-})$.
+In a well-formed derivation, other $r$s for variable are broken down
+in the course of the derivation until at the leaves (\rvar), only a
+single resource qualifier $r=1$ is left.
+%
+The crucial insight concerning the absence of distributive laws is the
+laziness of effects in CBPV.  They are only observed at positive
+types \citep{levy:hosc06},
+thus, using a negative variable $y : N \mid e$ only schedules
+the effect $e$, and likewise, in a negative let-binding ($\rlet^{-}$),
+the effects $e$ associated with the computation $u$ bound to $y$ are
+not run but ``bottled up'' with the variable in the context.  The same
+happens for thunk elimination ($\relim\Box$): the effects of the
+forced value $v : \thunkty e N$ are stored with the variable $y$
+holding the contents of the thunk $v$.
+
+\begin{figure}[htbp]
+\flushleft
+\ruler{}
+Types.
+\[
+\begin{array}{lllrl@{\qquad}l}
+\PTy & \ni & P
+  & ::= & \gthunkty e N
+     \mid o \mid \sumty i I P \mid \tupty i I P
+  & \nbox{Value types (positive types)} \\
+\NTy & \ni & N
+  & ::= & \compty r P
+     \mid \qfun r P N \mid \recty i I N
+  & \nbox{Computation types (negative types)} \\
+\Cxt & \ni & \Gamma
+  & ::= & \cempty \mid \ext \Gamma x P \mid \geext \Gamma y e N
+  & \nbox{Typing context}
+% \\
+% \R & \ni & \graybox{r,s}
+%    & ::= & 0 \mid 1 \mid r + s \mid rs
+%    & \mbox{Resource qualifiers}
+% \\
+% \RCxt & \ni & \graybox{\gamma,\delta}
+%   & ::= & \cempty \mid \ext \gamma x r \mid \ext \gamma y r
+%   & \mbox{Resource context}
+\end{array}
+\]
+% \dashruler{}
+% Terms.
+% \[
+% \begin{array}{lllrl@{\qquad}l}
+% \PTm & \ni & v,w
+%   & ::= & x
+%      \mid \thunk t
+%      \mid \inj i v
+%      \mid \ptup{\bar v}
+%   & \nbox{Values (positive terms)} \\
+% \NTm & \ni & t,u
+%   & ::= & y \mid \qlet r y u t
+%      \mid \qlet r x v t
+%   & \nbox{Computations (negative terms):}
+% \\ &&& \mid &
+%          \qforced r v y t
+%     \mid \qcase r v {\overline{x.\,t}}
+%     \mid \qsplits r v {\bar x} t
+%   & \nbox{value eliminations}
+% \\ &&& \mid &
+%          \ret v
+%     \mid \bind x u t
+%   & \nbox{monad operations}
+% \\ &&& \mid &
+%          \lam x t      \mid \app t v
+%   & \nbox{functions}
+% \\ &&& \mid &
+%          \trecord\{\overline{i:t}\}
+%        \mid \proj i t
+%   & \nbox{lazy tuples (records)}
+% \end{array}
+% \]
+\dashruler{}
+Value typing \fbox{$\qValTy \gamma \Gamma v P$}\,.
+\begin{gather*}
+ \nru{\rvar^+}{}
+     {\ValTy {\qext 0\Gamma 1x P} x P}
+\qquad
+ \nru{\rintro\Box}
+     {\gqeCompTy \gamma \Gamma t e N
+    }{\qValTy \gamma \Gamma {\thunk t} {\gthunkty e N}}
+\\[1.5ex]
+ \nru{\rintro\GS}
+     {\qValTy \gamma \Gamma v {P_i}
+    }{\qValTy \gamma \Gamma {\inj i v} {\sumt I P}}
+\qquad
+ \nru{\rintro\otimes}
+     {\forall i{:}I, \ \qValTy {\gamma_i} \Gamma {v_i} {P_i}
+    }{\pValTy {\GS_{i:I}\gamma_i} \Gamma {\ptup v} {\tupt I P}}
+\end{gather*}
+\dashruler{}
+Computation typing \fbox{$\gqeCompTy \gamma \Gamma t e N$}\,.
+%\vspace{-2ex}
+\begin{gather*}
+ \nru{\rvar^-}{}
+     {\gCompTy {\gqeext 0 \Gamma 1 y e N} y e N}
+\qquad
+ \nru{\rlet^-}
+     {\gqeCompTy \delta \Gamma u e N \qquad
+      \gCompTy {\gqeext \gamma \Gamma r y e N} t {e'} {N'}
+    }{\gpeCompTy {\gamma + r\delta} \Gamma {\qlet r y u t} {e'} {N'}}
+\\[1.5ex]
+ \nru{\rlet^+}
+     {\qValTy \delta \Gamma v P \qquad
+      \gCompTy {\qext \gamma \Gamma r x P} t e N
+    }{\gpeCompTy {\gamma + r\delta} \Gamma {\qlet r x v t} e N}
+\qquad
+ \nru{\relim\Box}
+     {\qValTy \delta \Gamma v {\gthunkty e N} \qquad
+      \gCompTy {\gqeext \gamma \Gamma r y e N} t {e'} {N'}
+    }{\gpeCompTy {\gamma + r\delta} \Gamma {\qforced r v y t} {e'} {N'}}
+\\[1.5ex]
+ \nru{\relim\GS}
+     {\qValTy \delta \Gamma v {\sumt I P} \quad
+      \forall i{:}I,\ \gCompTy {\qext \gamma \Gamma r {x_i} {P_i}} {t_i} e N
+    }{\gpeCompTy {\gamma + r\delta} \Gamma {\caser v {\{r x_i.\, t_i\}_{i:I}}} e N}
+\quad
+ \nru{\relim\otimes}
+     {\qValTy \delta \Gamma v {\tupt I P} \quad
+      \gqeCompTy \gamma {\extr \Gamma {\overline{r x_i{:}P_i}^{i:I}}} t e N
+    }{\gpeCompTy {\gamma + r\delta} \Gamma {\qsplits r v {\bar x} t} e N}
+\\[1.5ex]
+ \nru{\rintro\diamond}
+     {\qValTy \gamma \Gamma v P
+    }{\gqeCompTy {r\gamma} \Gamma {\ret v} \Ge {\compty r P}}
+\qquad
+ \nru{\relim\diamond}
+     {\gqeCompTy \delta \Gamma u {e_1} {\compty r P} \qquad
+      \gCompTy {\qext \gamma \Gamma r x P} t {e_2} N
+    }{\gpeCompTy {\gamma + \delta} \Gamma {\bind x u t} {e_1 \bu e_2} N}
+\\[1.5ex]
+ \nru{\rintro\lolli}
+     {\gCompTy {\qext \gamma \Gamma r x P} t e N
+    }{\gqeCompTy \gamma \Gamma {\lam x t} e {\qfun r P N}}
+\qquad
+ \nru{\relim\lolli}
+     {\gqeCompTy \gamma \Gamma t e {\qfun r P N} \qquad
+      \qValTy \delta \Gamma v P
+    }{\gpeCompTy {\gamma + r\delta} \Gamma {\app t v} e N}
+\\[1.5ex]
+ \nru{\rintro\Pi}
+     {\forall i{:}I,\ \gqeCompTy \gamma \Gamma {t_i} e {N_i}
+    }{\gqeCompTy \gamma \Gamma {\trecord {\{i : t_i\}_{i:I}}} e {\rect I N}}
+\qquad
+ \nru{\relim\Pi}
+     {\gqeCompTy \gamma \Gamma t e {\rect I N}
+    }{\gqeCompTy \gamma \Gamma {\proj i t} e {N_i}}
+\\[1.5ex]
+ \nrux{\graybox{\rsub}}
+      {\gqeCompTy {\gamma'} \Gamma t e N}
+      {\gqeCompTy \gamma \Gamma t {e'} N}
+      {\gamma \leq \gamma', \graybox{e \leq e'}}
+\end{gather*}
+\rule{\textwidth}{0.2pt}
+  \caption{Fully graded call-by-push-value.}
+  \label{fig:full-cbpv}
+\end{figure}
 

### 2020-07-08 popl21/popl21.tex
[ popl21 ] spell check run

@@
 </concept>
-</ccs2012>
+</Cs2012>
 \end{CCSXML}
@@
 %\newcommand{}{\ensuremath{}}
+\newcommand{\bla}{\ensuremath{\mbox{$$}}}
 \newcommand{\ie}{\emph{i.e.}}
@@
 \newcommand{\Den}[2]{\den{#1}_{#2}}
+\newcommand{\Denpar}[2]{\Den{#1}{(#2)}}
 \newcommand{\denv}[2]{\dent{#1}_{#2}}
+\newcommand{\denvp}[2]{\dent{#1}_{(#2)}}
 
@@
 fact, the uses of \rsub can be confined to the hypotheses of \relim\GS
-and \rintro\Pi, except for a final invokation of \rsub at the very end
+and \rintro\Pi, except for a final invocation of \rsub at the very end
 of the typing derivation.  Alternatively, we could have introduced
@@
 \]
-That these coercions satify the identity and composition---the functor
+That these coercions satisfy the identity and composition---the functor
 laws for $\denn N$---is easy to verify.
@@
 If type constructor $\Box$ is interpreted as a monoidal comonad $\D$,
-then all positive types $P$ can be intepreted as $\D$-coalgebras
+then all positive types $P$ can be interpreted as $\D$-coalgebras
 $A = \den P$, \ie, implement a morphism $\expose[A] : A \to \D\,A$
@@
 context-dependent computation \citep{orchard:icfp14}.  For grading,
-loccit uses a \emph{resource algebra} in form of a pre-ordered
+\loccit uses a \emph{resource algebra} in form of a pre-ordered
 semiring
@@
 $\gamma : \R^I$ and $\delta : \R^J$ and thus $C \boxtimes D$ is a
-$\R^{I+J}$-graded $\D$-coalgebra.  Further, we implicitely use the
+$\R^{I+J}$-graded $\D$-coalgebra.  Further, we implicitly use the
 isomorphism between $1$-structured $\D$-coalgebras $C : \R^1 \to \CC$
@@
 (public or private), or a sensitivity level (a non-negative real)
-\citep{reedPeirce:icfp10}.  If a lambda abstraction $\lam x t$ is
+\citep{reedPierce:icfp10}.  If a lambda abstraction $\lam x t$ is
 typed with $\qfun r P N$, qualifier $r$ is attached to variable $x$ in
@@
 the presence of $\eta$ for functions and records, they are
-interderivable with the following permutation laws:
+inter-derivable with the following permutation laws:
 \[
@@
 computations $u$, thus, we require a notion of computation
-substitution $\qsubst u r t y$ that should have the following semantics.
+substitution $\qsubst u r t y$ that should have the following semantics
+(see corollaries \ref{cor:tysinglesub} and \ref{cor:semsinglesub}).
 \begin{align*}
@@
       \oCompTy {\qext \gamma \Gamma r y N} t {N'}
-    }{\pCompTy {\gamma + r\delta} \Gamma {\qsubst u r t y} {N'}}
+    }{\pCompTy {\gamma + r\delta} \Gamma {\qsubst u r y t} {N'}}
 \\ &
-  \dent{\qsubst u r t y} = \dent t
+  \dent{\qsubst u r y t} = \dent t
     \comp (\tid[\den\Gamma \gamma]
@@
 
+\subsection{Coeffect-graded CBPV: substitution and metatheory}
+
 Substitution for coeffect-graded type system has been worked
-out in detail by \citet{atkeyWood:types19}.  It straightforwardly extends to
-coeffect-graded CBPV.
+out in detail by \citet{atkeyWood:types19,woodAtkey:linearity20}.
+It straightforwardly extends to coeffect-graded CBPV.
 %
@@
 \fbox{$\SubstTy \Psi \Gamma \sigma \Delta$} is equipped with a matrix
-$\Psi \in \R^{\dom\Gamma \times \dom\Delta}$ recording the usage
-vector $\Psi(z) \in \R^{\dom\Gamma}$ for each variable
+$\Psi : \R^{\dom\Gamma \times \dom\Delta}$ recording the usage
+vector $\Psi(z) : \R^{\dom\Gamma}$ for each variable
 $z \in \dom\Delta = \dom \sigma$ in the domain of the substitution.
@@
 The matrix $\Psi : \R^{\dom \Gamma \times \dom \Delta}$ acts as a
-linear map $\R^{\dom \Delta} \to \R^{\dom \Gamma}$ and allows us to
-state the substitution theorem.
+linear map
+$(\delta : \R^{\dom \Delta}) \mapsto (\delta\Psi : \R^{\dom \Gamma})$
+and allows us to state the substitution theorem.
 \begin{theorem}[Substitution preserves typing]
@@
   \begin{enumerate}
-  \item If $\qValTy \delta \Delta v P$ then $\pValTy {\Psi\delta} \Gamma {v\sigma} P$.
+  \item If $\qValTy \delta \Delta v P$ then $\pValTy {\delta\Psi} \Gamma {v\sigma} P$.
   \item If $\qCompTy \delta \Delta t N$ then
-    $\pCompTy {\Psi\delta} \Gamma {t\sigma} N$.
+    $\pCompTy {\delta\Psi} \Gamma {t\sigma} N$.
   \end{enumerate}
@@
 \end{proof}
+\begin{corollary}[Single substitution preserves typing]
+  \label{cor:tysinglesub}
+  \bla
+  \begin{enumerate}
+  \item If $\qValTy \delta \Gamma v P$ and
+      $\oCompTy {\qext \gamma \Gamma r x P} t {N}$
+      then $\pCompTy {\gamma + r\delta} \Gamma {\qsubst v r x t} {N}$.
+  \item If $\qCompTy \delta \Gamma u N$ and
+      $\oCompTy {\qext \gamma \Gamma r y N} t {N'}$
+      then $\pCompTy {\gamma + r\delta} \Gamma {\qsubst u r y t} {N'}$.
+  \end{enumerate}
+\end{corollary}
+%\begin{remark}
+Analogous statements are available for the substitution into values
+$\ValTy {\qext \gamma \Gamma r x P} w {P'}$ and
+$\ValTy {\qext \gamma \Gamma r y N} w {P}$, however, they are not
+invoked in the proof of subject reduction.
+%\end{remark}
 \begin{corollary}[Subject reduction]
@@
 is a natural transformation
-$\denv \sigma \delta : \Hom {\Den \Gamma {\Psi\delta}} {\Den \Delta
+$\denv \sigma \delta : \Hom {\Den \Gamma {\delta\Psi}} {\Den \Delta
   \delta}$
@@
 \comp
-\Den \Gamma {\Psi\delta \leq \Psi\delta'}
+\Den \Gamma {\delta\Psi \leq \delta'\Psi}
 $.
 It is defined by recursion as follows:
-\begin{align*}
-& \ru{
+\[
+\begin{array}{ll@{~}c@{~}l@{~}c@{~}l}
+\ru{
     }{\SubstTy {0} \Gamma \sempty \cempty}
-& \dent \sempty
-& : \Hom {\Den\Gamma 0} \I
-& & = \tdrop[\den\Gamma]
-\\[2ex]
-& \ru{\SubstTy \Psi \Gamma \sigma \Delta \qquad
+& \denv \sempty \delta
+& : & \Hom {\Den\Gamma 0} \I
+\\ &
+& = & \tdrop[\den\Gamma]
+\\[3ex]
+\ru{\SubstTy \Psi \Gamma \sigma \Delta \qquad
       \qValTy \gamma \Gamma v P
     }{\pSubstTy {\Psi,\gamma} \Gamma {\sext \sigma x v} {\ext \Delta x P}}
-& \dent{\sext \sigma x v}
-& : \Hom  {\Den\Gamma{\Psi\delta,r\gamma}} {\Den \Delta \delta \otimes \Denp P r}
-& & = \denv \sigma \delta \otimes \denv v r
-\\[2ex]
-& \ru{\SubstTy \Psi \Gamma \sigma \Delta \qquad
+& \denvp {\sext \sigma x v} {\delta,r}
+& : & \Hom  {\Den\Gamma{(\delta,r)(\Psi,\gamma)}} {\Den \Delta \delta \otimes \Denp P r}
+\\ &
+& = & (\denv \sigma \delta \otimes \denv v r)
+\comp \duplicate[\den\Gamma]
+\\[3ex]
+\ru{\SubstTy \Psi \Gamma \sigma \Delta \qquad
       \qCompTy \gamma \Gamma t N
     }{\pSubstTy {\Psi,\gamma} \Gamma {\sext \sigma y t} {\ext \Delta y N}}
-\end{align*}
+& \denvp {\sext \sigma y t} {\delta,r}
+& : & \Hom  {\Den\Gamma{(\delta,r)(\Psi,\gamma)}} {\Den \Delta \delta \otimes \D[r]\,\denn N}
+\\ &
+& = & (\denv \sigma \delta \otimes (\fmapt[{\D[r]}]{\dent t} \comp \expose[{\den\Gamma}]))
+\comp \duplicate[\den\Gamma]
+\end{array}
+\]
+Observe that $(\delta,r)(\Psi,\gamma) = \delta\Psi + r\gamma$ and,
+consequently,
+$\duplicate[\den\Gamma] : \Hom {\Den\Gamma{(\delta,r)(\Psi,\gamma)}}
+{\Den\Gamma{\delta\Psi} \otimes \Den\Gamma{r\gamma}}$.
+\begin{theorem}[Soundness of substitution]
+  Let $\SubstTy \Psi \Gamma \sigma \Delta$.
+  \begin{enumerate}
+  \item If $\qValTy \delta \Delta v P$ then
+     $\denv{v\sigma}r = \denv v {r} \comp \denv \sigma {r\delta}
+      : \Hom {\Den{\Gamma}{r\delta\Psi}} {\Denp P r}$.
+  \item If $\qCompTy \delta \Delta t N$ then
+     $\dent{t\sigma} = \dent t \comp \denv \sigma \delta
+      : \Hom {\Den\Gamma{\delta\Psi}} {\denn N}$.
+  \end{enumerate}
+\end{theorem}
+\begin{proof}
+  By induction on the typing derivation of the term $v$ and $t$, resp.
+\end{proof}
+\begin{corollary}[Soundness of single substitution]
+  \label{cor:semsinglesub}
+  \bla
+  \begin{enumerate}
+  \item If $\qValTy \delta \Gamma v P$ and
+      $\oCompTy {\qext \gamma \Gamma r x P} t {N}$
+      then
+\[
+  \dent{\qsubst v r x t} = \dent t
+    \comp (\tid[\den\Gamma \gamma] \otimes \denv v r)
+    \comp \duplicate[\den\Gamma]
+  .
+\]
+  \item If $\qCompTy \delta \Gamma u N$ and
+      $\oCompTy {\qext \gamma \Gamma r y N} t {N'}$ then
+\[
+  \dent{\qsubst u r y t} = \dent t
+    \comp (\tid[\den\Gamma \gamma]
+             \otimes (\fmapt[{\D[r]}] \dent u \comp \expose[\den\Gamma]))
+    \comp \duplicate[\den\Gamma]
+  .
+\]
+  \end{enumerate}
+\end{corollary}
+\begin{proof}
+  After comprehending that the semantics of identity substitutions is
+  just $\tid$, we observe that the semantics of single substitutions
+  is $\dent {v/rx} = (\tid \otimes \denv v r) \comp \duplicate$ and
+  $\dent {t/ry} = (\tid \otimes (\fmapt[{\D[r]}]{\dent t} \comp \expose[{\den\Gamma}])) \comp
+  \duplicate$, resp.
+\end{proof}
+
+\subsection{Coeffect-graded CBPV without negative variables}
+
+We introduce negative variables $y : N$ for the sake of grading thunking ($\Box$).  There is an alternative
 
@@
 \graybox{\mbox{highlighted}}.
-%
-We reintroduced effect grading into the computation typing
+
+We reintroduce effect grading into the computation typing
 $\qeCompTy \gamma \Gamma t e N$.  As in the effect-graded version,
 effects are accumulated in a single place, the monadic bind construct
-(\relim\diamond).  The negative variables $y : N$ we introduced for
+(\relim\diamond).  The negative variables $y : N$ we have introduced for
 coeffect grading are now annotated with an effect qualifier $e$ in the
@@
 $e$ and coeffect $r$.  This combination is interpreted as
-$\D[r]\,\denn N e$ where the resource qualifier acts as ``multiplier''
+$\D[r]\,\Denn N e$ where the resource qualifier acts as ``multiplier''
 for the effect qualifier $e$.  However, we never have to carry out
@@
 $\D$ distributes over the monad algebra $N$.  The reason is that only
-variables $y$ with $r=1$ can be used from the context $(\rvar^{-})$.
-In a well-formed derivation, other $r$s for variable are broken down
-in the course of the derivation until at the leaves (\rvar), only a
+variables $y$ with qualifier $r=1$ can be used from the context $(\rvar^{-})$.
+In a well-formed derivation, other resource qualifiers for variables are broken down
+in the course of the derivation until at the leaves (\rvar) only a
 single resource qualifier $r=1$ is left.
-%
+
 The crucial insight concerning the absence of distributive laws is the
@@
 thus, using a negative variable $y : N \mid e$ only schedules
-the effect $e$, and likewise, in a negative let-binding ($\rlet^{-}$),
+the effect $e$.  Likewise, in a negative let-binding ($\rlet^{-}$),
 the effects $e$ associated with the computation $u$ bound to $y$ are

### 2020-07-08 popl21/popl21.tex
[ popl21 ] removed negative variables

@@
 context where each variable $x$ is annotated by a resource qualifier
-$r$ in addition to its type $\tau$.  Such a typing context
-$(r_1x_1{:}\tau_1, \dots, r_nx_n{:}\tau_n) =: \gamma\Gamma$ can be
+$r$ in addition to its type $P$.  Such a typing context
+$(r_1x_1{:}P_1, \dots, r_nx_n{:}P_n) =: \gamma\Gamma$ can be
 split into a pure typing context
-$\Gamma = x_1{:}\tau_1, \dots, x_n{:}\tau_n$ and a resource context
+$\Gamma = x_1{:}P_1, \dots, x_n{:}P_n$ and a resource context
 $\gamma = x_1{:}r_1, \dots, x_n{:}r_n$ such that
@@
 comes with its own resource qualifier, we cannot simply model the
-context as a tensor product $\tupt {z:\dom\Gamma} \den{\Gamma(z)}$ of
+context as a tensor product $\tupt {x:\dom\Gamma} \den{\Gamma(x)}$ of
 $\D$-coalgebras since this product would indexed by a single resource
@@
 \Cxt & \ni & \Gamma
-  & ::= & \cempty \mid \ext \Gamma x P \mid \graybox{\ext \Gamma y N}
+  & ::= & \cempty \mid \ext \Gamma x P % \mid \graybox{\ext \Gamma y N}
   & \nbox{Typing context}
@@
 \RCxt & \ni & \graybox{\gamma,\delta}
-  & ::= & \cempty \mid \ext \gamma x r \mid \ext \gamma y r
+  & ::= & \cempty \mid \ext \gamma x r % \mid \ext \gamma y r
   & \mbox{Resource context}
@@
 \NTm & \ni & t,u
-  & ::= & \graybox{y} \mid \graybox{\qlet r y u t}
-     \mid \qlet {\graybox r} x v t
+  & ::= &
+     %     \graybox{y} \mid \graybox{\qlet r y u t}
+     % \mid
+         \qlet {\graybox r} x v t
   & \nbox{Computations (negative terms):}
 \\ &&& \mid &
-         \graybox{\qforced r v y t}
+         \force v
+         % \graybox{\qforced r v y t}
     \mid \qcase {\graybox r} v {\overline{x.\,t}}
@@
 \begin{gather*}
-\graybox{
- \nru{\rvar^-}{}
-     {\ValTy {\gqext 0 \Gamma 1 y N} y N}
-}
-\qquad
-\graybox{
- \nru{\rlet^-}
-     {\qCompTy \delta \Gamma u N \qquad
-      \oCompTy {\qext \gamma \Gamma r y N} t {N'}
-    }{\gpCompTy {\gamma + r\delta} \Gamma {\qlet r y u t} {N'}}
-}
-\\[1.5ex]
- \nru{\rlet^+}
+% \graybox{
+%  \nru{\rvar^-}{}
+%      {\ValTy {\gqext 0 \Gamma 1 y N} y N}
+% }
+% \qquad
+% \graybox{
+%  \nru{\rlet^-}
+%      {\qCompTy \delta \Gamma u N \qquad
+%       \oCompTy {\qext \gamma \Gamma r y N} t {N'}
+%     }{\gpCompTy {\gamma + r\delta} \Gamma {\qlet r y u t} {N'}}
+% }
+% \\[1.5ex]
+%  \nru{\rlet^+}
+ \nru{\rlet}
      {\gqValTy \delta \Gamma v P \qquad
@@
  \nru{\relim\Box}
-     {\gqValTy \delta \Gamma v {\Box N} \qquad
-      \oCompTy {\gqext \gamma \Gamma r y N} t {N'}
-    }{\gpCompTy {\gamma + r\delta} \Gamma {\gqforced r v y t} {N'}}
+     {\qValTy \gamma \Gamma v {\Box N}
+    }{\qCompTy \gamma \Gamma {\force v} {N}}
+ % \nru{\relim\Box}
+ %     {\gqValTy \delta \Gamma v {\Box N} \qquad
+ %      \oCompTy {\gqext \gamma \Gamma r y N} t {N'}
+ %    }{\gpCompTy {\gamma + r\delta} \Gamma {\gqforced r v y t} {N'}}
 \\[1.5ex]
@@
 of the subterms are added when both subterms are or maybe evaluated at
-runtime (rules \rintro\otimes, $\rlet^{+/-}$,
-\relim{\Box/\GS/\otimes/\diamond/\lolli}).  Note that in \relim\GS,
+runtime (rules \rintro\otimes, \rlet, %$\rlet^{+/-}$,
+\relim{%\Box/
+\GS/\otimes/\diamond/\lolli}).  Note that in \relim\GS,
 the branches $t_i$ of the case statement (\relim\GS) share a resource
@@
 that allows to scale the production of $v$ by $r$ if the resources are
-scaled accordingly (from $\gamma$ to $r\gamma$).  We bake scaling instead
+scaled accordingly (from $\gamma$ to $r\gamma$).
+%
+However, this would have prevented the typing of $\force v$ (rule
+\relim\Box), bcause there is no place for the scaling factor in
+computation typing.  Semantically, $v$ lives in $\D[\_]\denn N$, and
+we can only $\textract$ the computation in $\denn N$ if we can
+instantiate the coeffect qualifier $\_$ to $1$.  Such is not possible
+if scaling already happened (and needs to be respected).
+%
+Instead, we bake scaling
 into the transition from values to computations.
@@
 Analogous scaling of the eliminatee is baked into the other value
-eliminators (\relim\Box, \relim\GS, \relim\otimes), into the
-\rlet-rules, and into \rintro\diamond.
+eliminators (%\relim\Box,
+\relim\GS, \relim\otimes), into %the \rlet-rules,
+ \rlet,
+and into \rintro\diamond.
 
@@
 
-As a consequence of eliminatee scaling, $\relim\Box$ is no longer
-``natural-deduction'' style like projection, but ``sequent-calculus''
-style like $\relim\diamond$.  The original construct $\force v$ could
-eliminate an unscaled thunk $\qValTy \gamma \Gamma v {\Box N}$,
-however, computation typing $\qCompTy \gamma \Gamma {\force v} N$ has
-no scaling option, since negative types $N$ will not be interpreted as
-$\D$-coalgebras.  Scaling is semantically restricted to positive
-types, and in our type system, to context entries.  Thus, we extend
-CBPV to negative variables $y : N$ that come with a resource qualifier
-$r$ in the context and can there be interpreted as elements of
-$\D[r]\,\denn N$.  This is a modification also taken in call-by-need
-CBPV \citep{mcDermottMycroft:esop19}; this compromise allows us to
-keep CBPV's polarization of types where positive types stand for
-comonad coalgebras and negative types for monad algebras.
+% As a consequence of eliminatee scaling, $\relim\Box$ is no longer
+% ``natural-deduction'' style like projection, but ``sequent-calculus''
+% style like $\relim\diamond$.  The original construct $\force v$ could
+% eliminate an unscaled thunk $\qValTy \gamma \Gamma v {\Box N}$,
+% however, computation typing $\qCompTy \gamma \Gamma {\force v} N$ has
+% no scaling option, since negative types $N$ will not be interpreted as
+% $\D$-coalgebras.  Scaling is semantically restricted to positive
+% types, and in our type system, to context entries.  Thus, we extend
+% CBPV to negative variables $y : N$ that come with a resource qualifier
+% $r$ in the context and can there be interpreted as elements of
+% $\D[r]\,\denn N$.  This is a modification also taken in call-by-need
+% CBPV \citep{mcDermottMycroft:esop19}; this compromise allows us to
+% keep CBPV's polarization of types where positive types stand for
+% comonad coalgebras and negative types for monad algebras.
 
@@
 $\denp P : (\R,\leq) \to \CC$ and negative types $N$ as objects
-$\denn N : \CC$, except for in contexts, where they are interpreted as
-$\D$-algebras $\Den N r = \D[r]\,\denn N$.  With $\den P = \denp P$,
-the interpretation of contexts is thus
-$\den \Gamma = \boxtimes_{x:\dom\Gamma} \den{\Gamma(x)}$.
+$\denn N : \CC$.
+% , except for in contexts, where they are interpreted as
+% $\D$-algebras $\Den N r = \D[r]\,\denn N$.  With $\den P = \denp P$,
+% the interpretation of contexts is thus
+% $\den \Gamma = \boxtimes_{x:\dom\Gamma} \den{\Gamma(x)}$.
+The interpretation of contexts is
+$\den \Gamma = \boxtimes_{x:\dom\Gamma} \denp{\Gamma(x)}$.
 \[
@@
 \]
-The symmetric monoidal category $\CC$ needs to come with cartesian
+The symmetric monoidal category $\CC$ needs to be equipped with cartesian
 products $\rect I B$ as well as with distributive coproducts
@@
 \begin{array}{llcl}
-\rvar^-
-  & \dent {\oCompTy{\qext 0 \Gamma 1 y N} y N}
-  & = & \Gr \comp (\tdrop[\den\Gamma] \otimes \extract[\denn N])
-\\
-\rlet^{-}
-  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qlet r y u t} {N'}}
-  & = & \dent t
-  \comp (\tid[\Den \Gamma \gamma] \otimes
-           (\fmapt[{\D[r]}] \dent u \comp \expose[\den \Gamma]))
-\\ &&& {~}
-  \comp \duplicate
-\\
-\rlet^{+}
+% \rvar^-
+%   & \dent {\oCompTy{\qext 0 \Gamma 1 y N} y N}
+%   & = & \Gr \comp (\tdrop[\den\Gamma] \otimes \extract[\denn N])
+% \\
+% \rlet^{-}
+%   & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qlet r y u t} {N'}}
+%   & = & \dent t
+%   \comp (\tid[\Den \Gamma \gamma] \otimes
+%            (\fmapt[{\D[r]}] \dent u \comp \expose[\den \Gamma]))
+% \\ &&& {~}
+%   \comp \duplicate
+% \\
+\rlet %^{+}
   & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qlet r x v t} {N}}
@@
 \relim\Box
-  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qforced r v y t} {N'}}
-  & = & \dent t
-  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
-  \comp \duplicate
+  & \dent {\qCompTy \gamma \Gamma {\force v} {N}}
+  & = & \extract[\D] \comp \denv v 1
+% \relim\Box
+%   & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qforced r v y t} {N'}}
+%   & = & \dent t
+%   \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
+%   \comp \duplicate
 \\
@@
 and \emph{sequencing laws} that syntactically express that negative
-types are monad algebras \citep[Fig.~11]{levi:hosc06}.  The sequencing
+types are monad algebras \citep[Fig.~11]{levy:hosc06}.  The sequencing
 laws contain a generalization of the associativity law for monads and
@@
 
-Our extension of CBPV features negative variables $y$ standing for
-computations $u$, thus, we require a notion of computation
-substitution $\qsubst u r t y$ that should have the following semantics
-(see corollaries \ref{cor:tysinglesub} and \ref{cor:semsinglesub}).
-\begin{align*}
-&
-  \ru{\qCompTy \delta \Gamma u N \qquad
-      \oCompTy {\qext \gamma \Gamma r y N} t {N'}
-    }{\pCompTy {\gamma + r\delta} \Gamma {\qsubst u r y t} {N'}}
-\\ &
-  \dent{\qsubst u r y t} = \dent t
-    \comp (\tid[\den\Gamma \gamma]
-             \otimes (\fmapt[{\D[r]}] \dent u \comp \expose[\den\Gamma]))
-    \comp \duplicate[\den\Gamma]
-\end{align*}
-
-Further, we add $\beta$-laws for the negative
-let-binding and $\relim\Box$, since $\tforce$ now has let-style elimination:
-\[
-\begin{array}{llcl}
-\rbeta{\rlet^{-}}
-  & \qlet r y u t
-  & = & \qsubst u r y t
-\\
-\rbeta\Box
-  & \qforced r {\thunk u} y t
-  & = & \qsubst u r y t
-\\
-\end{array}
-\]
-These laws are immediately sound, since left and right hand sides of
-these equations have the same denotation.
-%
-Read as reduction rules, these laws are added to the operational
-semantics of CBPV, replacing $\beta$-law $\force{(\thunk u)} = u$.
+% Our extension of CBPV features negative variables $y$ standing for
+% computations $u$, thus, we require a notion of computation
+% substitution $\qsubst u r t y$ that should have the following semantics
+% (see corollaries \ref{cor:tysinglesub} and \ref{cor:semsinglesub}).
+% \begin{align*}
+% &
+%   \ru{\qCompTy \delta \Gamma u N \qquad
+%       \oCompTy {\qext \gamma \Gamma r y N} t {N'}
+%     }{\pCompTy {\gamma + r\delta} \Gamma {\qsubst u r y t} {N'}}
+% \\ &
+%   \dent{\qsubst u r y t} = \dent t
+%     \comp (\tid[\den\Gamma \gamma]
+%              \otimes (\fmapt[{\D[r]}] \dent u \comp \expose[\den\Gamma]))
+%     \comp \duplicate[\den\Gamma]
+% \end{align*}
+
+% Further, we add $\beta$-laws for the negative
+% let-binding and $\relim\Box$, since $\tforce$ now has let-style elimination:
+% \[
+% \begin{array}{llcl}
+% \rbeta{\rlet^{-}}
+%   & \qlet r y u t
+%   & = & \qsubst u r y t
+% \\
+% \rbeta\Box
+%   & \qforced r {(\thunk u)} y t
+%   & = & \qsubst u r y t
+% \\
+% \end{array}
+% \]
+% These laws are immediately sound, since left and right hand sides of
+% these equations have the same denotation.
+% %
+% Read as reduction rules, these laws are added to the operational
+% semantics of CBPV, replacing $\beta$-law $\force{(\thunk u)} = u$.
 
@@
     }{\pSubstTy {\Psi,\gamma} \Gamma {\sext \sigma x v} {\ext \Delta x P}}
-\qquad
-  \ru{\SubstTy \Psi \Gamma \sigma \Delta \qquad
-      \qCompTy \gamma \Gamma t N
-    }{\pSubstTy {\Psi,\gamma} \Gamma {\sext \sigma y t} {\ext \Delta y N}}
+% \qquad
+%   \ru{\SubstTy \Psi \Gamma \sigma \Delta \qquad
+%       \qCompTy \gamma \Gamma t N
+%     }{\pSubstTy {\Psi,\gamma} \Gamma {\sext \sigma y t} {\ext \Delta y N}}
 \]
@@
   \item If $\qValTy \delta \Gamma v P$ and
+      $\ValTy {\qext \gamma \Gamma r x P} w {P'}$
+      then $\ValTy {\gamma + r\delta} \Gamma {\qsubst v r x w} {P'}$.
+  \item If $\qValTy \delta \Gamma v P$ and
       $\oCompTy {\qext \gamma \Gamma r x P} t {N}$
       then $\pCompTy {\gamma + r\delta} \Gamma {\qsubst v r x t} {N}$.
-  \item If $\qCompTy \delta \Gamma u N$ and
-      $\oCompTy {\qext \gamma \Gamma r y N} t {N'}$
-      then $\pCompTy {\gamma + r\delta} \Gamma {\qsubst u r y t} {N'}$.
+  % \item If $\qCompTy \delta \Gamma u N$ and
+  %     $\oCompTy {\qext \gamma \Gamma r y N} t {N'}$
+  %     then $\pCompTy {\gamma + r\delta} \Gamma {\qsubst u r y t} {N'}$.
   \end{enumerate}
 \end{corollary}
-%\begin{remark}
-Analogous statements are available for the substitution into values
-$\ValTy {\qext \gamma \Gamma r x P} w {P'}$ and
-$\ValTy {\qext \gamma \Gamma r y N} w {P}$, however, they are not
-invoked in the proof of subject reduction.
-%\end{remark}
+% %\begin{remark}
+% Analogous statements are available for the substitution into values
+% $\ValTy {\qext \gamma \Gamma r x P} w {P'}$ and
+% $\ValTy {\qext \gamma \Gamma r y N} w {P}$, however, they are not
+% invoked in the proof of subject reduction.
+% %\end{remark}
 \begin{corollary}[Subject reduction]
@@
 \comp \duplicate[\den\Gamma]
-\\[3ex]
-\ru{\SubstTy \Psi \Gamma \sigma \Delta \qquad
-      \qCompTy \gamma \Gamma t N
-    }{\pSubstTy {\Psi,\gamma} \Gamma {\sext \sigma y t} {\ext \Delta y N}}
-& \denvp {\sext \sigma y t} {\delta,r}
-& : & \Hom  {\Den\Gamma{(\delta,r)(\Psi,\gamma)}} {\Den \Delta \delta \otimes \D[r]\,\denn N}
-\\ &
-& = & (\denv \sigma \delta \otimes (\fmapt[{\D[r]}]{\dent t} \comp \expose[{\den\Gamma}]))
-\comp \duplicate[\den\Gamma]
+% \\[3ex]
+% \ru{\SubstTy \Psi \Gamma \sigma \Delta \qquad
+%       \qCompTy \gamma \Gamma t N
+%     }{\pSubstTy {\Psi,\gamma} \Gamma {\sext \sigma y t} {\ext \Delta y N}}
+% & \denvp {\sext \sigma y t} {\delta,r}
+% & : & \Hom  {\Den\Gamma{(\delta,r)(\Psi,\gamma)}} {\Den \Delta \delta \otimes \D[r]\,\denn N}
+% \\ &
+% & = & (\denv \sigma \delta \otimes (\fmapt[{\D[r]}]{\dent t} \comp \expose[{\den\Gamma}]))
+% \comp \duplicate[\den\Gamma]
 \end{array}
@@
   \item If $\qValTy \delta \Gamma v P$ and
+      $\oCompTy {\qext \gamma \Gamma r x P} w {P'}$
+      then
+\[
+  \denv {\qsubst v r x w} s = \denv w s
+    \comp (\tid[\Den\Gamma{s\gamma}] \otimes \denv v {sr})
+    \comp \duplicate[\den\Gamma]
+  .
+\]
+  \item If $\qValTy \delta \Gamma v P$ and
       $\oCompTy {\qext \gamma \Gamma r x P} t {N}$
@@
   \dent{\qsubst v r x t} = \dent t
-    \comp (\tid[\den\Gamma \gamma] \otimes \denv v r)
+    \comp (\tid[\Den\Gamma \gamma] \otimes \denv v r)
     \comp \duplicate[\den\Gamma]
@@
 \]
-  \item If $\qCompTy \delta \Gamma u N$ and
-      $\oCompTy {\qext \gamma \Gamma r y N} t {N'}$ then
-\[
-  \dent{\qsubst u r y t} = \dent t
-    \comp (\tid[\den\Gamma \gamma]
-             \otimes (\fmapt[{\D[r]}] \dent u \comp \expose[\den\Gamma]))
-    \comp \duplicate[\den\Gamma]
-  .
-\]
+%   \item If $\qCompTy \delta \Gamma u N$ and
+%       $\oCompTy {\qext \gamma \Gamma r y N} t {N'}$ then
+% \[
+%   \dent{\qsubst u r y t} = \dent t
+%     \comp (\tid[\Den\Gamma \gamma]
+%              \otimes (\fmapt[{\D[r]}] \dent u \comp \expose[\den\Gamma]))
+%     \comp \duplicate[\den\Gamma]
+%   .
+% \]
   \end{enumerate}
@@
   just $\tid$, we observe that the semantics of single substitutions
-  is $\dent {v/rx} = (\tid \otimes \denv v r) \comp \duplicate$ and
-  $\dent {t/ry} = (\tid \otimes (\fmapt[{\D[r]}]{\dent t} \comp \expose[{\den\Gamma}])) \comp
-  \duplicate$, resp.
+  is $\dent {v/rx} = (\tid \otimes \denv v r) \comp \duplicate$.
+% and
+%   $\dent {t/ry} = (\tid \otimes (\fmapt[{\D[r]}]{\dent t} \comp \expose[{\den\Gamma}])) \comp
+%   \duplicate$, resp.
 \end{proof}
 
-\subsection{Coeffect-graded CBPV without negative variables}
-
-We introduce negative variables $y : N$ for the sake of grading thunking ($\Box$).  There is an alternative
+% \subsection{Coeffect-graded CBPV without negative variables}
+
+% We introduce negative variables $y : N$ for the sake of grading thunking ($\Box$).  There is an alternative
 
@@
 
-We reintroduce effect grading into the computation typing
-$\qeCompTy \gamma \Gamma t e N$.  As in the effect-graded version,
-effects are accumulated in a single place, the monadic bind construct
-(\relim\diamond).  The negative variables $y : N$ we have introduced for
-coeffect grading are now annotated with an effect qualifier $e$ in the
-typing context $\Gamma$.  The full information about a negative
-variable $y$ in a graded context $\gamma\Gamma$ is type $N$, effect
-$e$ and coeffect $r$.  This combination is interpreted as
-$\D[r]\,\Denn N e$ where the resource qualifier acts as ``multiplier''
-for the effect qualifier $e$.  However, we never have to carry out
-this ``multiplication'', thus, do not require a law how the comonad
-$\D$ distributes over the monad algebra $N$.  The reason is that only
-variables $y$ with qualifier $r=1$ can be used from the context $(\rvar^{-})$.
-In a well-formed derivation, other resource qualifiers for variables are broken down
-in the course of the derivation until at the leaves (\rvar) only a
-single resource qualifier $r=1$ is left.
-
-The crucial insight concerning the absence of distributive laws is the
-laziness of effects in CBPV.  They are only observed at positive
-types \citep{levy:hosc06},
-thus, using a negative variable $y : N \mid e$ only schedules
-the effect $e$.  Likewise, in a negative let-binding ($\rlet^{-}$),
-the effects $e$ associated with the computation $u$ bound to $y$ are
-not run but ``bottled up'' with the variable in the context.  The same
-happens for thunk elimination ($\relim\Box$): the effects of the
-forced value $v : \thunkty e N$ are stored with the variable $y$
-holding the contents of the thunk $v$.
+% We reintroduce effect grading into the computation typing
+% $\qeCompTy \gamma \Gamma t e N$.  As in the effect-graded version,
+% effects are accumulated in a single place, the monadic bind construct
+% (\relim\diamond).  The negative variables $y : N$ we have introduced for
+% coeffect grading are now annotated with an effect qualifier $e$ in the
+% typing context $\Gamma$.  The full information about a negative
+% variable $y$ in a graded context $\gamma\Gamma$ is type $N$, effect
+% $e$ and coeffect $r$.  This combination is interpreted as
+% $\D[r]\,\Denn N e$ where the resource qualifier acts as ``multiplier''
+% for the effect qualifier $e$.  However, we never have to carry out
+% this ``multiplication'', thus, do not require a law how the comonad
+% $\D$ distributes over the monad algebra $N$.  The reason is that only
+% variables $y$ with qualifier $r=1$ can be used from the context $(\rvar^{-})$.
+% In a well-formed derivation, other resource qualifiers for variables are broken down
+% in the course of the derivation until at the leaves (\rvar) only a
+% single resource qualifier $r=1$ is left.
+
+% The crucial insight concerning the absence of distributive laws is the
+% laziness of effects in CBPV.  They are only observed at positive
+% types \citep{levy:hosc06},
+% thus, using a negative variable $y : N \mid e$ only schedules
+% the effect $e$.  Likewise, in a negative let-binding ($\rlet^{-}$),
+% the effects $e$ associated with the computation $u$ bound to $y$ are
+% not run but ``bottled up'' with the variable in the context.  The same
+% happens for thunk elimination ($\relim\Box$): the effects of the
+% forced value $v : \thunkty e N$ are stored with the variable $y$
+% holding the contents of the thunk $v$.
 
@@
   & \nbox{Computation types (negative types)} \\
-\Cxt & \ni & \Gamma
-  & ::= & \cempty \mid \ext \Gamma x P \mid \geext \Gamma y e N
-  & \nbox{Typing context}
+% \Cxt & \ni & \Gamma
+%   & ::= & \cempty \mid \ext \Gamma x P % \mid \geext \Gamma y e N
+%   & \nbox{Typing context}
 % \\
@@
 \begin{gather*}
- \nru{\rvar^-}{}
-     {\gCompTy {\gqeext 0 \Gamma 1 y e N} y e N}
-\qquad
- \nru{\rlet^-}
-     {\gqeCompTy \delta \Gamma u e N \qquad
-      \gCompTy {\gqeext \gamma \Gamma r y e N} t {e'} {N'}
-    }{\gpeCompTy {\gamma + r\delta} \Gamma {\qlet r y u t} {e'} {N'}}
-\\[1.5ex]
- \nru{\rlet^+}
+%  \nru{\rvar^-}{}
+%      {\gCompTy {\gqeext 0 \Gamma 1 y e N} y e N}
+% \qquad
+%  \nru{\rlet^-}
+%      {\gqeCompTy \delta \Gamma u e N \qquad
+%       \gCompTy {\gqeext \gamma \Gamma r y e N} t {e'} {N'}
+%     }{\gpeCompTy {\gamma + r\delta} \Gamma {\qlet r y u t} {e'} {N'}}
+% \\[1.5ex]
+ \nru{\rlet} %^+}
      {\qValTy \delta \Gamma v P \qquad
@@
  \nru{\relim\Box}
-     {\qValTy \delta \Gamma v {\gthunkty e N} \qquad
-      \gCompTy {\gqeext \gamma \Gamma r y e N} t {e'} {N'}
-    }{\gpeCompTy {\gamma + r\delta} \Gamma {\qforced r v y t} {e'} {N'}}
+     {\qValTy \gamma \Gamma v {\gthunkty e N}
+    }{\gqeCompTy \gamma \Gamma {\force v} e {N}}
+ % \nru{\relim\Box}
+ %     {\qValTy \delta \Gamma v {\gthunkty e N} \qquad
+ %      \gCompTy {\gqeext \gamma \Gamma r y e N} t {e'} {N'}
+ %    }{\gpeCompTy {\gamma + r\delta} \Gamma {\qforced r v y t} {e'} {N'}}
 \\[1.5ex]
@@
 
+We reintroduce effect grading into the computation typing
+$\qeCompTy \gamma \Gamma t e N$.  As in the effect-graded version,
+effects are accumulated in a single place, the monadic bind construct
+(\relim\diamond).
+
+The crucial insight concerning the absence of distributive laws is the
+laziness of effects in CBPV.  They are only observed at positive
+types \citep{levy:hosc06},
+thus, can be ``bottled up'' in $\thunk t : \thunkty e N$ and
+reactivated via forcing in $\force v : N \mid e$.
+
 %% Acknowledgments

### 2020-07-08 popl21/popl21.tex
[ popl21 ] semantics of graded CBPV

@@
 \begin{gather*}
- \nru{\rvar^+}{}
-     {\ValTy {\gqext 0\Gamma 1x P} x P}
+ \nru{\rvar %^+
+    }{
+    }{\ValTy {\gqext 0\Gamma 1x P} x P}
 \qquad
@@
 \begin{array}{llcl}
-\rvar^+
+\rvar %^+
   & \denv {\ValTy{\qext 0 \Gamma 1 x P} x P} r
@@
 \subsection{Coeffect-graded CBPV: equational theory and operational semantics}
+
+The equational theory and operational semantics of coeffect-graded
+CBPV is identical to the one of CBPV \citep[Fig.~11]{levy:hosc06}.  In
+this section, we make just a few remarks on an alternative
+presentation of the permutation laws (called ``sequencing laws'' in
+\loccit).
 
@@
 \subsection{Coeffect-graded CBPV: substitution and metatheory}
+\label{sec:subst}
 
@@
 \begin{gather*}
- \nru{\rvar^+}{}
-     {\ValTy {\qext 0\Gamma 1x P} x P}
+ \nru{\rvar %^+
+    }{
+    }{\ValTy {\qext 0\Gamma 1x P} x P}
 \qquad
@@
      {\qValTy \delta \Gamma v P \qquad
-      \gCompTy {\qext \gamma \Gamma r x P} t e N
-    }{\gpeCompTy {\gamma + r\delta} \Gamma {\qlet r x v t} e N}
+      \CompTy {\qext \gamma \Gamma r x P} t e N
+    }{\peCompTy {\gamma + r\delta} \Gamma {\qlet r x v t} e N}
 \qquad
@@
      {\qValTy \delta \Gamma v {\sumt I P} \quad
-      \forall i{:}I,\ \gCompTy {\qext \gamma \Gamma r {x_i} {P_i}} {t_i} e N
-    }{\gpeCompTy {\gamma + r\delta} \Gamma {\caser v {\{r x_i.\, t_i\}_{i:I}}} e N}
+      \forall i{:}I,\ \CompTy {\qext \gamma \Gamma r {x_i} {P_i}} {t_i} e N
+    }{\peCompTy {\gamma + r\delta} \Gamma {\caser v {\{r x_i.\, t_i\}_{i:I}}} e N}
 \quad
@@
      {\qValTy \delta \Gamma v {\tupt I P} \quad
-      \gqeCompTy \gamma {\extr \Gamma {\overline{r x_i{:}P_i}^{i:I}}} t e N
-    }{\gpeCompTy {\gamma + r\delta} \Gamma {\qsplits r v {\bar x} t} e N}
+      \qeCompTy \gamma {\extr \Gamma {\overline{r x_i{:}P_i}^{i:I}}} t e N
+    }{\peCompTy {\gamma + r\delta} \Gamma {\qsplits r v {\bar x} t} e N}
 \\[1.5ex]
@@
  \nru{\rintro\lolli}
-     {\gCompTy {\qext \gamma \Gamma r x P} t e N
-    }{\gqeCompTy \gamma \Gamma {\lam x t} e {\qfun r P N}}
+     {\CompTy {\qext \gamma \Gamma r x P} t e N
+    }{\qeCompTy \gamma \Gamma {\lam x t} e {\qfun r P N}}
 \qquad
  \nru{\relim\lolli}
-     {\gqeCompTy \gamma \Gamma t e {\qfun r P N} \qquad
+     {\qeCompTy \gamma \Gamma t e {\qfun r P N} \qquad
       \qValTy \delta \Gamma v P
-    }{\gpeCompTy {\gamma + r\delta} \Gamma {\app t v} e N}
+    }{\peCompTy {\gamma + r\delta} \Gamma {\app t v} e N}
 \\[1.5ex]
  \nru{\rintro\Pi}
-     {\forall i{:}I,\ \gqeCompTy \gamma \Gamma {t_i} e {N_i}
-    }{\gqeCompTy \gamma \Gamma {\trecord {\{i : t_i\}_{i:I}}} e {\rect I N}}
+     {\forall i{:}I,\
+      \qeCompTy \gamma \Gamma {t_i} e {N_i}
+    }{\qeCompTy \gamma \Gamma {\trecord {\{i : t_i\}_{i:I}}} e {\rect I N}}
 \qquad
  \nru{\relim\Pi}
-     {\gqeCompTy \gamma \Gamma t e {\rect I N}
-    }{\gqeCompTy \gamma \Gamma {\proj i t} e {N_i}}
+     {\qeCompTy \gamma \Gamma t e {\rect I N}
+    }{\qeCompTy \gamma \Gamma {\proj i t} e {N_i}}
 \\[1.5ex]
@@
 
+\subsection{Graded CBPV: semantics}
+
+As in effect-graded CBPV, negative types $N$ are interpreted as graded
+monad algebras $\denn N : (\Eff,\leq) \to \CC$, and as in
+coeffect-graded CBPV, positive types are interpreted as graded comonad
+coalgebras $\denp P : (\R,\leq) \to \CC$.  Values
+$\qValTy \gamma \Gamma v P$ are interpreted as natural transformations
+$\dent v : \Denpar \Gamma {\_\cdot\gamma} \todot \denp P$ and
+computations $\qeCompTy \gamma \Gamma t e N$ as morphisms
+$\dent t : \Hom {\Den\Gamma \gamma} {\Denn N e}$.  Except for some
+change in typing, the denotation of terms is unchanged from
+coeffect-graded CBPV.  We recapitulate the most interesting cases here
+with the updated typing:
+\[
+\xymatrix@C=10ex@R=3ex{
+\denv {\thunk t} r
+& \Den \Gamma {r\gamma}     \ar[r]^{\extract}
+& \D[r]\,\Den\Gamma\gamma   \ar[r]^{\fmapt[{\D[r]}]{\dent t}}
+& \D[r]\,\Denn N e
+\\
+\dent {\force v}
+& \Den \Gamma \gamma    \ar[r]^{\denv v 1}
+& \D[1]\,\Denn N e      \ar[r]^{\extract}
+& \Denn N e
+\\
+\dent {\ret v}
+& \Den \Gamma {r\gamma}  \ar[r]^{\denv v r}
+& \Denp P r              \ar[r]^{\treturn}
+& \T[\Ge]\,\Denp P r
+\\
+\dent {\bind x u t}
+& \Den \Gamma {\gamma + \delta}                 \ar[r]^{\duplicate}
+& \Den \Gamma \gamma \otimes \Den \Gamma \delta \ar[r]^{\tid \otimes \dent u}
+& \Den \Gamma \gamma \otimes \T[e_1] \Denp P r  \ar[d]^{\strengthl}
+\\
+& \Denn N {e_1 \bu e_2}
+& \T[e_1]\,\Denn N {e_2}                        \ar[l]_{\run[\denn N]}
+& \T[e_1](\Den \Gamma \gamma \otimes \Denp P r) \ar[l]_{\fmapt[{\T[e_1]}] {\dent t}}
+% \dent {\bind x u t}
+% & \Den \Gamma {\gamma + \delta}                 \ar[r]^{\duplicate}
+% & \Den \Gamma \gamma \otimes \Den \Gamma \delta \ar[r]^{\tid \otimes \dent u}
+% & \Den \Gamma \gamma \otimes \T[e_1] \Denp P r  \ar[r]^{\strengthl}
+% & \T[e_1](\Den \Gamma \gamma \otimes \Denp P r) \ar[r]^{\fmap {\dent t}}
+% & \T[e_1]\,N_{e_2}                               \ar[r]^{\trun}
+% & N_{e_1 \bu e_2}
+}
+\]
+Subsumption (\rsub) is interpreted by pre- and post-composition with
+the functorial actions of $\den \Gamma$ and $\denn N$:
+\[
+\xymatrix@C=10ex{
+% \rsub
+% &
+\Den \Gamma {\gamma}   \ar[r]^{\Den\Gamma{\gamma \leq \gamma'}}
+& \Den \Gamma {\gamma'}  \ar[r]^{\dent t}
+& \Denn N e              \ar[r]^{\Denn N {e \leq e'}}
+& \Denn N {e'}
+}
+\]
+
+% \begin{align*}
+% & \denv {\thunk t} r
+% & &
+% \xymatrix{
+%   \Den \Gamma {r\gamma}     \ar[r]^{\extract}
+% & \D[r]\,\Den\Gamma\gamma   \ar[r]^{\fmapt[{\D[r]}]{\dent t}}
+% & \D[r]\,\Denn N e
+% }
+% \\
+% \xymatrix{
+%   \Den \Gamma \gamma    \ar[r]^{\denv v 1}
+% & \D[1]\,\Denn N e      \ar[r]^{\extract}
+% & \Denn N e
+% }
+% \\
+% & \dent {\ret v}
+% & &
+% \xymatrix@C=8ex{
+%   \Den \Gamma {r\gamma}  \ar[r]^{\denv v r}
+% & \Denp P r              \ar[r]^{\treturn}
+% & \T[\Ge]\,\Denp P r
+% }
+% \end{align*}
+
+The operational semantics of graded CBPV is identical to the one of
+CBPV, and the substitution typing of coeffect-graded CBPV
+(Section~\ref{sec:subst}) can be mechanically transferred to fully
+graded CBPV.
+
 %% Acknowledgments

### 2020-07-09 popl21/popl21.tex
[ popl21 ] reread section 2

@@
 
+\theoremstyle{remark}
 \newtheorem{remark}{Remark}
@@
 
-\section{Introduction}
-
-Call-by-push-value (CBPV) is a simply typed lambda calculus that polarizes
-types into value and computation types and can thus express both
-call-by-name and call-by-value evaluation in the presence of effects.
-Semantically, effects are modeled by a monad, and computation types as
-algebras over this monad.
-Effect type systems usually express more information than the presence
-of an effect; typically effects are structured as pre-ordered monoid
-where the monoid operation represents accumulation of effects and the
-order subsumption, in analogy to subtyping.
-In this work, a \emph{graded} version of is presented where the typing
-of computations likens effect typing.  Semantically, computation types
-are then represented as graded monad algebras.
-Further we present a version of CBPV which additionally has graded coeffects.
-
-\section{An effect-graded version of CBPV}
-
 %\newcommand{}{\ensuremath{}}
@@
 \newcommand{\CC}{\ensuremath{\mathcal{C}}}
-\newcommand{\Hom}[2]{\CC\,(#1,\,#2)}
+\newcommand{\EE}{\ensuremath{\mathcal{E}}}
+\newcommand{\Hom}[3][\CC]{#1\,(#2,\,#3)}
 \newcommand{\SET}{\mathsf{SET}}
@@
 \newcommand{\treturn}{\mathsf{return}}
+\newcommand{\return}[1][]{\treturn_{#1}}
 \newcommand{\tjoin}{\mathsf{join}}
-\newcommand{\join}{\mathsf{join}}
+\newcommand{\join}[1][]{\mathsf{join}_{#1}}
 \newcommand{\append}{\ensuremath{\mathbin{\mathtt{++}}}}
@@
 \newcommand{\nru}[3]{#1\;\dfrac{#2}{#3}}
+\newcommand{\rux}[3]{\dfrac{#1}{#2}\;#3}
 \newcommand{\nrux}[4]{#1\;\dfrac{#2}{#3}\;#4}
@@
 
+\section{Introduction}
+
+Call-by-push-value (CBPV) is a simply typed lambda calculus that polarizes
+types into value and computation types and can thus express both
+call-by-name and call-by-value evaluation in the presence of effects.
+Semantically, effects are modeled by a monad, and computation types as
+algebras over this monad.
+Effect type systems usually express more information than the presence
+of an effect; typically effects are structured as pre-ordered monoid
+where the monoid operation represents accumulation of effects and the
+order subsumption, in analogy to subtyping.
+In this work, a \emph{graded} version of is presented where the typing
+of computations likens effect typing.  Semantically, computation types
+are then represented as graded monad algebras.
+Further we present a version of CBPV which additionally has graded coeffects.
+
+\paragraph*{Preliminaries.}  The reader should bring some elementary
+knowledge of category theory, such as the interpretation of
+simply-typed lambda calculus in cartesian-closed categories.  However,
+we try to be gentle with categorical concepts such as monads and
+recapitulate their definition where needed.  In many cases, it is
+sufficient to think in terms of the category $\SET$ where objects are
+sets and morphisms functions between sets, or in the functor category
+$[\CC \to \SET]$ where objects are monotone families of sets
+$(A_i)_{i:\CC}$ indexed by objects $i$ of $\CC$ such that
+$\CC_i \to \CC_j$ in $\SET$ for $i \to j$ in $\CC$ and morphisms are
+natural transformations $(f_i : A_i \to B_i)_{i:\CC}$.
+
+\section{An effect-graded version of CBPV}
+
 % As effect algebra, we assume a pre-ordered monoid
@@
 \label{sec:ccc}
+
+In this section, we recapitulate monads and some essential vocabulary
+of category theory.  There are no surprises, thus, the experienced
+reader is invited to skip this section.
 
@@
 Typically, $\CC$ is a cartesian-closed category, i.e., has products
-$A_1 \times A_2$ of objects modelling product types and contexts,
-and exponentials $A \To
-B$ modeling function types.  Further $\CC$ maybe be distributive,
+$A_1 \times A_2$ of objects to model product types and contexts,
+and exponentials $A \To B$ to model function types.
+Further $\CC$ maybe be distributive,
 \ie, have coproducts $A_1 + A_2$ that distribute over products, to
@@
 
-Recall that effects are modeled by a suitable monad $\T$ in
-$\CC$.  An example would be the \emph{writer monad} $\T\,A = \String \times
-A$ that models the effect \emph{output}.
+Recall that effects are modeled by a suitable monad $\T : \CC \to \CC$
+in $\CC$.  We shall refer to elements of a monadic type $\T\,A$ as
+\emph{computations} when $\CC = \SET$.  For a general $\CC$,
+computations shall be the morphisms $\Hom A {\T\,B}$ called
+\emph{Kleisli arrows}.
+An example module would be the \emph{writer monad}
+$\T\,A = \String \times A$
+that models the effect \emph{output}.
 Another example would be the \emph{exception monad} $\Exc + \_$ where
@@
 \begin{array}{lcl}
-\tfmap & : & (A \to B) \to \T\,A \to \T\,B \\
-\tfmap\, f\, (s,\; a) & = & (s,\, f\, b)
+\fmap[\T] & : & (A \to B) \to \T\,A \to \T\,B \\
+\fmap[\T] f\, (s,\; a) & = & (s,\, f\, b)
 \\[1ex]
-\treturn & : & A \to \T\,A \\
-\treturn\, a & = & (\verb|""|, a)
+\return[\T] & : & A \to \T\,A \\
+\return[\T]\, a & = & (\verb|""|, a)
 \\[1ex]
-\tjoin & : & \T\,(\T\,A) \to \T\,A \\
-\tjoin\,(s_1,(s_2,a)) & = & (s_1 \append s_2,\; a) \\
+\join[\T] & : & \T\,(\T\,A) \to \T\,A \\
+\join[\T]\,(s_1,(s_2,a)) & = & (s_1 \append s_2,\; a) \\
 \end{array}
@@
 
-We assume familiarity with the functor laws $\tfmap\,\tid = \tid$ and
-$\tfmap\,(f \circ g) = \tfmap\,f \circ \tfmap\,g$.
+Monad unit $\treturn$ (written $\eta$ in category speak) turns a value
+into a pure computation and monad multiplication $\tjoin$ (written
+$\mu$ in category speak) combines two effects and thus allows
+sequencing of computations via \emph{Kleisli composition}
+$ (g : \Hom B {\T\,C}) \comp_{\T} (f : \Hom A {\T\,B}) = (\tjoin \comp
+  \fmap g \comp f : \Hom A {\T\,C})$.
+
+The presence of $\tfmap : \Hom A B \to \Hom {\T\,A} {\T\,B}$
+satisfying the functor laws $\tfmap\,\tid = \tid$ and
+$\tfmap\,(f \circ g) = \tfmap\,f \circ \tfmap\,g$ makes $\T$ a
+\emph{(endo)functor}, written $\T : [\CC \to \CC]$.  The
+endofunctors $F : [\CC \to \CC]$ form a category, the
+$\emph{functor category}$, with identity $\Id\,A = A$ and
+composition $(F \comp G)\,A = F\,(G\,A)$.  We shall write
+$f : F \todot G$ for morphisms in the functor category, called
+\emph{natural transformations}.  These are families
+$f_A : \Hom {F\,A} {G\,A}$ of morphisms that commute with the
+functor action $\tfmap$, \ie,
+$\fmap[G]h \comp f_A = f_B \comp \fmap[F]h$ for any $h : \Hom A B$.
+
+Unit $\treturn : \Id \todot \T$ and multiplication
+$\tjoin : \T \comp \T \todot \T$ are \emph{natural transformations}
+which breaks down to $\fmap[\T] h \comp \treturn = \treturn \comp h$
+and $\fmap[\T] h \comp \tjoin = \tjoin \comp \fmap{(\fmap h)}$.
+%
 The three equational
@@
 \emph{graded} over elements $e$ of a suitable effect algebra $\Eff$.
-An effect algebra be a pre-ordered monoid
+An effect algebra be a preordered monoid
 $(\Eff,\bu,\Ge,\leq)$ such that $\_{\bu}\_$ is monotone wrt.\ the
@@
 \\[1ex]
-\tcast & : & \T[e] \to \T[e'] \mbox{ for } e \leq e'\\
+\tcast & : & \T[e]\,A \to \T[e']\,A \mbox{ for } e \leq e'\\
 \tcast & = & \tid \\
@@
 \circ \tfmap\,\tcast = \tcast \circ \tjoin$.
+We may write $\T[e_1 \leq e_2]$ for the natural transformation
+$\tcast : \T[e_1] \todot \T[e_2]$.
+
+\begin{remark}[$\Eff$-graded monad are lax monoidal functors $\T : {[\Eff \to [\CC \to \CC]]}$]
+The concept of a graded monad can be more succinctly expressed by
+using more advanced language of category theory.  (This reformulation
+is not essential for the remainder of the exposition and may be
+skipped on first reading.)
+
+Recall that a \emph{monoidal category} $\EE$ has a designated object
+$\I : \EE$ and an operation $\_{\otimes}\_ : [\EE \to [\EE \to \EE]$,
+the \emph{tensor product} on objects of $\EE$ that is functorial in
+both positions.  Further, there are natural isomorphisms
+$\Gl : (\I \otimes A) \cong A$ and $\Gr : (A \otimes \I) \cong A$
+witnessing the unitality of $\I$ and
+$\alpha : (A \otimes (B \otimes C)) \cong ((A \otimes B) \otimes C)$
+witnessing associativity of $\otimes$.
+
+Any preordered monoid, such as $(\Eff,\_\bu\_,\Ge,\leq)$, makes a
+monoidal category $\EE = \Eff$ with Homset
+$\EE(e,e') = \{ () \mid e \leq e' \}$, tensor $\otimes = \bu$ and unit
+$\I = \Ge$.  Further, this category is \emph{thin}, \ie, there is at
+most one morphism between any two objects $e,e'$.
+
+A graded monad $\T : [\Eff \to [\CC \to \CC]]$ is then a morphism from
+monoidal category $(\Eff,\bu,\Ge)$ to monoidal category
+$([\CC \to \CC],\comp,\Id)$.  The monoidal structure in the latter is
+just functor composition and identity functor.  Operations $\tcast$
+and $\tfmap$ witness the functoriality of $\T$ in its first and second
+argument.  The operations $\treturn$ and $\tjoin$ witness (in a
+directed way) the preservation of unit $\Ge$ and multiplication $\bu$
+by $\T$, making $\T$ a \emph{lax monoidal functor}.
+(End of remark.)
+\end{remark}
 
@@
 aka $\T$-algebras.
-Those are objects $B$ together with a morphism $\run[B] : \T\,B
+Those algebras are objects $B$ together with a morphism $\run[B] : \T\,B
 \to B$ that allows to formally \emph{run} the monad, ``merging'' its
@@
 $\tjoin$.  Levy \citeyearpar{levy:hosc06} shows that monad algebras are
-closed under products and exponentiation with arbitrary objects.  \Eg,
+closed under products ($\times$) and exponentiation ($\To$)
+with arbitrary objects.  \Eg,
 in $\SET$ we can define, for \emph{writer} algebras:
@@
 \]
+
+\begin{remark}[Naturality of $\trun$]
+  The $\T$-algebra $B : [\Eff \to \CC]$ comes with a natural morphism
+  $\run[e_1,e_2] : \Hom {\T[e_1]\,B_{e_2}} {B_{e_1 \bu e_2}}$ in the
+  sense that
+  $B_{(e_1\leq e_1') \bu (e_2 \leq e_2')} \comp {\run[e_1,e_2]} =
+  {\run[e_1',e_2']} \comp \T[e_1 \leq e_1']\,B_{e_2 \leq e_2'}$.
+\end{remark}
 
@@
 
+With the theory of graded monads in place,
+we design a graded version of CBPV.
 The syntax and typing rules for effect-graded CBPV are given in
@@
 the graded monad.  The unit $\ret v$ of the monad is effect-free
-(\rintro\diamond), running this computation just produces the pure
+(\rintro\diamond); running this computation just produces the pure
 value $v$.  Sequencing computations $u$ and $t$ via the \emph{bind}
-construct $\bind x u t$ composes the effects $e_1$ of $u$ with the
+construct ``$\bind x u t$'' composes the effects $e_1$ of $u$ with the
 effects $e_2$ of $t$ in that order.  The intuition is that first $u$
@@
 to produce effects classified by the same $e$.  However, in reality,
-different branches of a case distinction, for instance, may produce
+different branches of \eg a case distinction may produce
 very different effects.  To end up with a unique classifier $e$, the
@@
 
-The denotational semantics of graded CBPV has been informally explained in
+The denotational semantics of the novel parts
+of graded CBPV has been informally explained in
 sections \ref{sec:ccc} to \ref{sec:graded_algebra} already; in the following,
 we spell out the details.  We assume a distributive cartesian-closed
-category $\CC$ with a strong graded monad $\T : \Eff \to \CC \to \CC$.
+category $\CC$ with a strong graded monad $\T : [\Eff \to [\CC \to \CC]]$.
 Let us agree on some notation for the constructions on objects and
@@
   \rangle_{i:I} : \CC(C,\prod_i A)$ for $f_i : \CC(C,A_i)$.  Binary
-  products $\prod_{\{1,2\}} A$ are written $A_1 \times A_2$.
+  products $\prod_{\{1,2\}} A$ are written $A_1 \times A_2$,
+  and the nullary product (terminal object)
+  is written $1$ with nullary tupling $\langle\rangle : \CC(C,1)$.
 \item Coproduct $\coprod_{i:I} A_i$ with injections $\iota_i :
@@
 Negative types $N$ are interpreted
-as functors $\denn N : (\Eff,\leq) \to \CC$ mapping
+as functors
+$\denn N : [\Eff \to \CC]$
+%$\denn N : (\Eff,\leq) \to \CC$
+mapping
 effect classifiers $e$ to objects $\Denn N e$ and effect subsumption $e
@@
 The graded $\T$-algebra structure $\run[\denn N]$ is constructed by
-induction on $N$:
-\[
-\begin{array}{lcl}
+induction on $N$, as well as functoriality $\Denn N {e_1 \leq e_2}$:
+\[
+\begin{array}{l@{~~}c@{\hspace{1ex}}l}
   \run[\denn N] & : & \Hom {\T[e_1](\Denn N {e_2})} {\Denn N {e_1 \bu e_2}} \\
@@
 \end{array}
-\]
-This construction is the same as \citet{levy:hosc06}, modulo grading.
-The algebra laws ensue.
-
-The coercion morphisms $\Denn N {e_1 \leq e_2}$ are obtained
-analogously:
-\[
-\begin{array}{rcl}
+\
+%\qquad
+\begin{array}{r@{\hspace{1ex}}c@{\hspace{1ex}}l}
   \Denn N {e_1 \leq e_2} & : & \Hom {\Denn N {e_1}} {\Denn N {e_2}} \\
@@
 \]
-That these coercions satisfy the identity and composition---the functor
+This construction is the same as \citet{levy:hosc06}, modulo grading.
+The algebra laws ensue.
+That the coercions $\Denn N {e_1 \leq e_2}$
+satisfy identity and composition---the functor
 laws for $\denn N$---is easy to verify.
+
+% The coercion morphisms $\Denn N {e_1 \leq e_2}$ are obtained
+% analogously:
+% \[
+% \begin{array}{rcl}
+%   \Denn N {e_1 \leq e_2} & : & \Hom {\Denn N {e_1}} {\Denn N {e_2}} \\
+%   \Denn {\diamond P} {e_1 \leq e_2} & = & \tcast \\
+%   \Denn {P \To N}   {e_1 \leq e_2} & = &
+%     \Lambda(\Denn N {e_1 \leq e_2} \comp \teval) \\
+%   \Denn {\rect I N} {e_1 \leq e_2} & = &
+%     \langle \Denn N {e_1 \leq e_2} \comp \pi_i {} \rangle_{i:I} \\
+% \end{array}
+% \]
+% That these coercions satisfy the identity and composition---the functor
+% laws for $\denn N$---is easy to verify.
 
@@
 {\denp \Gamma} {\Denn N e}$.  Most of the cases are straightforward
-and in analogy to CBPV, let us thus focus on the
+and in analogy to CBPV, so let us focus on the
 \graybox{\mbox{modified}} rules where grading comes into play.  A
@@
 coercion in the denotational semantics.  The most natural way to make
-this precise is to use intrinsically well-typed syntax; from typing
+this precise is to use intrinsically well-typed syntax.  Going from typing
 rules to such well-typed syntax is a routine transformation which we
@@
 
-\subsection{Grading via a partial monoid}
+\newcommand{\tprint}{\mathsf{print}}
+\newcommand{\print}[2]{\tprint\,#1.\,#2}
+\newcommand{\toutput}{\mathsf{output}}
+%\newcommand{\output}[2][]{\toutput_{#1}\,#2}
+\newcommand{\tthrow}{\mathsf{throw}}
+\newcommand{\throw}[1]{\tthrow\,#1}
+\newcommand{\tcatch}{\mathsf{catch}}
+\newcommand{\catch}[3]{#1\,\tcatch\,#2 \mapsto #3}
+\newcommand{\traise}{\mathsf{raise}}
+%\newcommand{\raise}[1]{\traise\,#1} %clash
+\newcommand{\thandle}{\mathsf{handle}}
+\newcommand{\handle}[4][]{\thandle_{#1}\,#2\,#3\,#4}
+
+\subsection{Example effects}
+
+We replay some of Levy's \citeyearpar{levy:hosc06} effect examples in
+graded CBPV.
+
+\paragraph{Printing}
+
+Outputting a fixed string $s$ before computing $t$ is
+facilitated by ``$\print s t$'' which we type by the following rule:
+\[
+  \rux{\CompTy \Gamma t m N
+     }{\CompTy \Gamma {\print s t} {n + m} N
+     }{s : \StringLe n}
+\]
+Herein, we use effect algebra $(\bN \cup \{\infty\},+,0,\leq)$ and
+graded monad $\T[n]A = \StringLe n \times A$.
+%
+This allows us to implement a family of morphisms
+$\toutput : \StringLe n \to \Hom[\SET] {1} {\T[n]\,1}$ to interpret
+the print statement as
+$\dent {\print s t} = \run[\den N] \comp \fmap[\T]{\pi_2} \comp
+\strengthr \comp \langle \toutput\,s \comp \langle\rangle,\, \dent t
+\rangle$:
+\[
+\xymatrix@C=8ex{
+  \den \Gamma
+     \ar[rr]^{\langle \toutput\,s \comp \langle\rangle,\, \dent t \rangle}
+&& \T[n]\,1 \times \Denn N m
+     \ar[r]^{\strengthr}
+& \T[n](1 \times \Denn N m)
+     \ar[r]^{\fmap[\T]{\pi_2}}
+& \T[n]\,\Denn N m
+     \ar[r]^{\trun}
+& \Denn N {n + m}
+}
+\]
+
+% This allows us to implement a family of natural transformations $\toutput_A : \StringLe n \to \Hom[\SET] {\T[m]\,A} {\T[n+m]\,A}$
+
+\paragraph{Exceptions}
+
+Given a set $\Exc$ of exceptions whose elements we refer to by $e$ and whose subsets by $E$,
+consider the effect algebra $(\Pot\Exc,\cup,\emptyset,\subseteq)$.
+Primitives for throwing and catching exceptions can be added to graded
+CBPV by the following rules:
+\newcommand{\eee}{E_1\setminus\{e\} \cup E_2}
+\[
+  \ru{}{\CompTy \Gamma {\throw e} {\{e\}} N} %{e : \Exc}
+\qquad
+  \ru{\CompTy \Gamma u {E_1} N \qquad
+      \CompTy \Gamma t {E_2} N
+    }{\CompTy \Gamma {\catch u e t} {\eee} N}
+\]
+If $u$ throws exception $e$, then ``$\catch u e t$'' computes $t$, else $u$.
+%
+In $\SET$ we use the graded monad $\T[E]\,A = E + A$ and a family of
+morphism $\traise\,e : \SET(1,\T[\{e\}]A)$ to interpret
+$\dent{\throw e} = \traise\,e \comp \langle\rangle$.  To interpret
+$\tcatch$, define a family
+$\thandle_N\,e : \Hom[\SET] {N_{E_1} \times N_{E_2}} {N_{\eee}}$ by
+induction on $N \in \NTy$:
+\[
+\begin{array}{lcl}
+  \handle[\diamond P] e g h & = & \left\{
+                                  \begin{array}{ll}
+                                    h & \mbox{if } g = \iota_1 e \\
+                                    g & \mbox{otherwise}
+                                  \end{array}
+\right. \\
+  \handle[P \To N] e g h \, a & = & \handle[N] e {(g\,a)} {(h\,a)} \\
+  \handle[\tupt I N] e g h \, i & = & \handle[N_i] e {(g\,i)} {(h\,i)} \\
+\end{array}
+\]
+Catching is then
+$\dent {\catch u e t} = \handle[N] e \comp \langle \dent u, \dent t
+\rangle$.
+Note that $\thandle$ cannot be defined in terms of $\trun$, but we
+have to break $N$ down to monadic type $\diamond P$
+to get access to the exception
+thrown by $u$.
+
+\subsection{Digression: grading via a partial monoid}
 
@@
 prevented statically by graded effect typing.  This could be modelled
-by adding a maximal element $\top : \Eff$---with $e \leq top$ for all
+by adding a maximal element $\top : \Eff$---with $e \leq \top$ for all
 $e : \Eff$ that signifies an inconsistent state.  This error element
@@
 
+\begin{remark}
+Another way of restricting effect composition is to use a 2-category
+$\Eff$ where the objects $i,j,k$ are state types, the morphisms $e$
+effect classifiers and the 2-cells effect subsumption $e \leq e'$.
+The necessary theory has been worked out by
+\citet{orchardWadlerEades:msfp20}.
+%
+An example would be a typed state monad
+$\T[e: i \to j] A = S_i \to (A \times S_j)$ where the state type $S_i$
+is indexed, and effects $e : i \to j$ may only make valid
+modifications to the state.
+%
+For instance, the index could be a set of pointers denoting the
+allocated heap cells and a read/write/deallocate effect would require
+the respective pointer to be a member of this set (and remove it in
+case of deallocate).
+\end{remark}
+
 \section{Coeffect-graded CBPV}
@@
 context-dependent computation \citep{orchard:icfp14}.  For grading,
-\loccit uses a \emph{resource algebra} in form of a pre-ordered
+\loccit uses a \emph{resource algebra} in form of a preordered
 semiring

### 2020-07-09 popl21/popl21.tex
[ popl21 ] introduction

@@
 algebras over this monad.
+%
 Effect type systems usually express more information than the presence
-of an effect; often effects are categorized by pre-ordered monoid
+of an effect; often effects are categorized by preordered monoid
 where the monoid operation represents accumulation of effects and the
@@
 are then represented as graded monad algebras.
-Further we present a version of CBPV which additionally has graded coeffects.
+%
+Observing that the value types of CBPV can be interpreted as comonad coalgebras, we further present a version of CBPV that has coeffects graded by a preordered semiring.
+Value types and contexts are interpreted as graded comonad coalgebras, allowing resource-aware interpretations of CBPV.
+%
+Finally, we combine the two systems into a fully graded version of CBPV where both effects and coeffects are graded.
+Thus turns out to be possible without specifying an interaction between effects and coeffects.
 \end{abstract}
@@
 \newcommand{\Eg}{\emph{E.g.}}
-\newcommand{\loccit}{\emph{loc.\,cit.}}
+\newcommand{\loccit}{\emph{loc.\,cit.}\xspace}
 \newcommand{\cf}{cf.\ }
@@
 
+% Concrete effects
+\newcommand{\tprint}{\mathsf{print}}
+\newcommand{\print}[2]{\tprint\,#1.\,#2}
+\newcommand{\toutput}{\mathsf{output}}
+%\newcommand{\output}[2][]{\toutput_{#1}\,#2}
+\newcommand{\tthrow}{\mathsf{throw}}
+\newcommand{\throw}[1]{\tthrow\,#1}
+\newcommand{\tcatch}{\mathsf{catch}}
+\newcommand{\catch}[3]{#1\,\tcatch\,#2 \mapsto #3}
+\newcommand{\traise}{\mathsf{raise}}
+%\newcommand{\raise}[1]{\traise\,#1} %clash
+\newcommand{\thandle}{\mathsf{handle}}
+\newcommand{\handle}[4][]{\thandle_{#1}\,#2\,#3\,#4}
+
 \section{Introduction}
 
-Call-by-push-value (CBPV) is a simply typed lambda calculus that polarizes
-types into value and computation types and can thus express both
-call-by-name and call-by-value evaluation in the presence of effects.
-Semantically, effects are modeled by a monad, and computation types as
-algebras over this monad.
-Effect type systems usually express more information than the presence
-of an effect; typically effects are structured as pre-ordered monoid
-where the monoid operation represents accumulation of effects and the
-order subsumption, in analogy to subtyping.
-In this work, a \emph{graded} version of is presented where the typing
-of computations likens effect typing.  Semantically, computation types
-are then represented as graded monad algebras.
-Further we present a version of CBPV which additionally has graded coeffects.
-
-\paragraph*{Preliminaries.}  The reader should bring some elementary
+Levy's call-by-push-value calculus (CBPV) \citeyearpar{levy:hosc06}
+is a simply typed lambda
+calculus with disjoint sums and eager and lazy products that allows
+the modelling of ``lazy'' effects.  For instance, the CBPV-program
+\[\print {\mathtt{"function"}} \lam x {\print {\mathtt{"argument"}} x}\]
+of type $P \To \diamond P$, will, unlike the corresponding ML-program,
+not print the word ``function'' and return a function; it will simply
+wait for an input value of type $P$.  Only after an argument has been
+supplied, the program will print the words ``function'' and
+``argument'' and then return the argument.  The reason is that effects
+in CBPV can only be observed at \emph{value types} like $P$, and
+function types are not value types but computation types.  Effectful
+actions (like printing) at computation types $N$ are ``pushed down''
+the type structure until they reach a position at a monadic type
+$\diamond P$ where they can be executed.  Semantically, this is
+facilitated by interpreting computation types as monad algebras with
+an action $\trun : \T\,N \to N$ that allows to formally run effects
+transported by the monad $\T$ at type $N$, even though $N$ is not
+directly a monadic type $\diamond P$.
+
+CBPV polarizes types into value types $P$ and computation types $N$
+where the former embed into the latter as $\diamond P$ by the formal
+``monad''~$\diamond$ and the latter embed into the former as $\Box N$
+by ``thunking''~$\Box$.  In Levy's \emph{behavioral semantics},
+functor $\diamond$ is a left adjoint to functor $\Box$ and need not be
+a monad.  Quite the opposite, when modelling store
+$\diamond P = P \times S$ is a comonad and $\Box N = S \to N$ a monad.
+Contrastingly, in the above explained \emph{algebra semantics},
+$\diamond$ is indeed modeled by a monad $\T$, yet $\Box$ just by the
+identity.  It is however possible to model $\Box$ by a monoidal
+comonad $\D$, and positive types by comonad coalgebras with an action
+$\texpose : P \to \D\,P$ that makes the services of the comonad
+available at all value types, not just at $\Box N$.
+
+Building on these observations, we investigate in this article how
+CBPV %would look like if it would
+can
+keep track of not just the presence
+of an effect or coeffect but also the kind of effect and coeffect.
+Semantically, precise information about effects can be obtained by
+using a family of monads $\T[e]$ graded over effect classifiers $e$
+drawn from a preordered monoid \cite{katsumata:popl14}.  For CBPV, we
+generalize the concept of monad algebras $N$ to \emph{graded} such
+ones, $N_e$.  Syntactically, computation typing $\CompTy \Gamma t e N$
+is extended by an effect classifier $e$
+(as usual in type and effect systems \citep{nielsen:effect99})
+such that a computation can be
+interpreted by a morphism $\Gamma \to N_e$.
+
+Coeffect type systems
+\citep{brunel:esop14,orchard:icfp14,ghicaSmith:esop14} have been
+developed to track resource consumption by attaching usage information
+to each variable in the typing context.  Coeffect typing generalizes
+linear typing \citep{girard:linear} to quantitative typing
+\citep{sergeyVytiniotisPeytonJones:popl14,mcBride:wadler60,atkey:lics18}
+and subsumes sensitivity analysis \citep{reedPierce:icfp10} and static
+information control flow \citep{volpano:jcs96} aka security typing.
+Coeffects can be modeled by a comonad $\D[r]$ graded over resource
+qualifiers $r$ drawn from a preordered semiring.  For the sake of
+coeffect-graded CBPV, we generalize graded comonads to graded comonad
+coalgebras.  Value types $P$ and contexts $\Gamma$ are then
+interpreted as such coalgebras.  Value typing
+$\qValTy \gamma \Gamma v P$ and computation typing
+$\qCompTy \gamma \Gamma t N$ are based on linear typing an come
+equipped with a resource context~$\gamma$.  Semantically, values are
+interpreted as morphisms $\Gamma_{r\gamma} \to P_r$ allowing the
+``multiplication'' or \emph{scaling} of a value by $r$ % natural in $r$
+and computations as morphisms $\Gamma_\gamma \to N$.
+
+Finally, effect and coeffect graded CBPV can be combined into a fully
+graded CBPV calculus, surprisingly without sorting out any interaction
+between effects and coeffects, such as the distributive laws of
+\citet{orchard:icfp16}.  We credit the smoothness of the integration
+to the careful placement of monad and comonad in CBPV's type system,
+and the lack of need for scaling % of not just values but also
+of computations.
+
+\paragraph*{Contributions}
+\begin{enumerate}
+
+\item We introduce graded monad algebras and an effect-graded version
+  of CBPV in Section~\ref{sec:effect}, after recapitulating monads and
+  their algebras and graded monad.
+
+\item We further introduce graded comonad coalgebras and a
+  coeffect-graded version of CBPV in Section~\ref{sec:coeff}.  We give
+  its denotational and operational semantics and adapt Atkey and
+  Wood's substitution theorem \citeyearpar{atkeyWood:types19} to this
+  version of CBPV.
+
+\item We present an effect- and coeffect-graded version of CBPV in Section~\ref{sec:full-cbpv}.
+
+\end{enumerate}
+
+\paragraph*{Preliminaries}  The reader should bring some elementary
 knowledge of category theory, such as the interpretation of
@@
 \section{An effect-graded version of CBPV}
+\label{sec:effect}
 
@@
 
-\newcommand{\tprint}{\mathsf{print}}
-\newcommand{\print}[2]{\tprint\,#1.\,#2}
-\newcommand{\toutput}{\mathsf{output}}
-%\newcommand{\output}[2][]{\toutput_{#1}\,#2}
-\newcommand{\tthrow}{\mathsf{throw}}
-\newcommand{\throw}[1]{\tthrow\,#1}
-\newcommand{\tcatch}{\mathsf{catch}}
-\newcommand{\catch}[3]{#1\,\tcatch\,#2 \mapsto #3}
-\newcommand{\traise}{\mathsf{raise}}
-%\newcommand{\raise}[1]{\traise\,#1} %clash
-\newcommand{\thandle}{\mathsf{handle}}
-\newcommand{\handle}[4][]{\thandle_{#1}\,#2\,#3\,#4}
-
 \subsection{Example effects}
@@
 
-A \emph{monoidal} comonad implements an morphism $\tzip$ that combines
+A \emph{monoidal} comonad implements a morphism $\tzip$ that combines
 a tuples of comonadic values into one comonadic tuple:
@@
 The map $\dent\tthunk : \Hom A B \to \Hom A {\D\,B}$ generalizes
-$\tcobind : \Hom{\D\,A} B \to \Hom{\D\,A}{\D\,B}$ to $\D$-coalgebras
+$\tcobind : \Hom{\D\,C} B \to \Hom{\D\,C}{\D\,B}$ to $\D$-coalgebras
 $A$ in the same way that the monadic
-$\tbind : \Hom{A}{\T\,B} \to \Hom{\T\,A}{\T\,B}$ is generalized to
+$\tbind : \Hom{A}{\T\,D} \to \Hom{\T\,A}{\T\,D}$ is generalized to
 $\Hom{A}{B} \to \Hom{\T\,A}{B}$ for $\T$-algebras $B$ in CBPV.
@@
 
-Graded comonads have been utilized to give semantics to
+\emph{Graded comonads} have been utilized to give semantics to
 context-dependent computation \citep{orchard:icfp14}.  For grading,
@@
 semiring
-$(\R,{+},0,{\cdot},1,\leq)$.
+$(\R,{+},0,{\cdot},1,{\leq})$.
 % $(\R,\_{+}\_,0,\_{\cdot}\_,1,\leq)$.
@@
 cartesian-closed category but in a symmetric monoidal closed category
-$(\CC,\otimes,\I,\lolli)$.  We overload the operators on
-exponentials, $\Lambda : \Hom {C \otimes A} B \to \Hom C {A \lolli B}$
+$(\CC,\otimes,\I,\lolli)$.
+Introduction $\Lambda$ and elimination $\teval$ of
+exponentials are there
+$\Lambda : \Hom {C \otimes A} B \to \Hom C {A \lolli B}$
 and $\teval : \Hom{(A \lolli B) \otimes A} B$.  The tensor product
@@
 Further, the symmetric monoidal structure is usually witnessed by
+natural
 isomorphisms for left $\lambda_A : \Hom {\I \otimes A} A$ and right
@@
 $\{A_i\}_{i:I}$ and $\{B_j\}_{j:J}$ coincide modulo addition and
-deletion of $\I$s.
+deletion of units ($\I$).
 
@@
 Subsemirings of $\R$ have been used for quantitative typing
-\cite{sergeyVytiniotisPeytonJones:popl14,mcBride:wadler60,atkey:lics08}.
+\cite{sergeyVytiniotisPeytonJones:popl14,mcBride:wadler60,atkey:lics18}.
 The order expresses precision of the quantities, \eg,
@@
 
-Objects of $\CC$ are functors $A : (\W,\wleq) \to \SET$ and morphisms
+Objects of $\CC$ are functors $A : [(\W,\wleq) \to \SET]$ and morphisms
 $f : \Hom A B$ are natural transformations
@@
 with unit $\I_w = \bigcup_{\emp \wleq w} 1$.  The unit $\I$ is
-constructible at world $w$ from nothing (set $1$) if $\emp \wleq w$,
+constructible at world $w$ from nothing (the unit set $1$) if $\emp \wleq w$,
 \ie, if the world $w$ includes the choice of using no resources.  A
@@
 \begin{array}{lcl}
-  \D & : & (\R,\leq) \to [\CC \to \CC] \\
+  \D & : & [(\R,\leq) \to [\CC \to \CC]] \\
 (\D[r] A)_w & = &
@@
 \]
-implementing trivially the following natural transformations.  Herein, we use the
+implementing trivially the following natural transformations.
+Herein, we use as second monoidal structure
+on the functor category $[\CC \to CC]$ the
 pointwise tensor product of functors
@@
 
-A $\D$-coalgebra is a functor $A : (\R,\leq) \to \CC$ with family of
+A $\D$-coalgebra is a functor $A : [(\R,\leq) \to \CC]$ with family of
 morphisms $\expose[r,s] : A_{rs} \to \D[r]\,A_s$ natural in $r$ and
@@
 context as a tensor product $\tupt {x:\dom\Gamma} \den{\Gamma(x)}$ of
-$\D$-coalgebras since this product would indexed by a single resource
-qualifier $r \in \R$ rather than a resource context
+$\D$-coalgebras since this product would be indexed by a single resource
+qualifier $r : \R$ rather than a resource context
 $\gamma : \dom\Gamma \to \R$.  The solution offered by
@@
 $(r \cdot \gamma)(i) = r \cdot \gamma(i)$.  Contexts $\Gamma$ can thus
-be interpreted as \emph{$R^{\dom \Gamma}$-graded}
-(or \emph{structured}) $\D$-coalgebras $C$ with
+be interpreted as
+(\emph{structured} or)
+\emph{$\R^{\dom \Gamma}$-graded}
+$\D$-coalgebras $C$ with
 the following operations:
@@
 $A_r := C_{r\gamma}$ flattens a structured $\D$-coalgebra $C$ into an
-ordinary one ($A$).
+ordinary one, $A$.
 
@@
 Computation typing \fbox{$\gqCompTy \gamma \Gamma t N$}\,.
-\vspace{-2ex}
+%\vspace{-2ex}
 \begin{gather*}
@@
 the branches $t_i$ of the case statement (\relim\GS) share a resource
-context $\gamma$, since only one of the branches is executed at
+context $\gamma$ since only one of the branches is executed at
 runtime.  Via rule $\rweak$,
@@
 \relim\Box), bcause there is no place for the scaling factor in
-computation typing.  Semantically, $v$ lives in $\D[\_]\denn N$, and
+computation typing.  Semantically, the $v$ in $\force v$
+lives in $\D[\_]\denn N$, and
 we can only $\textract$ the computation in $\denn N$ if we can
@@
 
-Positive types $P$ are interpreted as $\D$-algebras
-$\denp P : (\R,\leq) \to \CC$ and negative types $N$ as objects
+Positive types $P$ are interpreted as $\D$-coalgebras
+$\denp P : [(\R,\leq) \to \CC]$ and negative types $N$ as objects
 $\denn N : \CC$.
@@
 products $\rect I B$ as well as with distributive coproducts
-$\coprod_I A$ with
+$\coprod_I A$ with distribution morphism
 $\tdist : \Hom {C \otimes \coprod_I A} {\coprod_{i:I} {(C \otimes A_i)}}$.
@@
       $\ValTy {\qext \gamma \Gamma r x P} w {P'}$
-      then $\ValTy {\gamma + r\delta} \Gamma {\qsubst v r x w} {P'}$.
+      then $\pValTy {\gamma + r\delta} \Gamma {\qsubst v r x w} {P'}$.
   \item If $\qValTy \delta \Gamma v P$ and
@@
 As in effect-graded CBPV, negative types $N$ are interpreted as graded
-monad algebras $\denn N : (\Eff,\leq) \to \CC$, and as in
+monad algebras $\denn N : [(\Eff,\leq) \to \CC]$, and as in
 coeffect-graded CBPV, positive types are interpreted as graded comonad
-coalgebras $\denp P : (\R,\leq) \to \CC$.  Values
+coalgebras $\denp P : [(\R,\leq) \to \CC]$.  Values
 $\qValTy \gamma \Gamma v P$ are interpreted as natural transformations
@@
 \denv {\thunk t} r
-& \Den \Gamma {r\gamma}     \ar[r]^{\extract}
+& \Den \Gamma {r\gamma}     \ar[r]^{\expose}
 & \D[r]\,\Den\Gamma\gamma   \ar[r]^{\fmapt[{\D[r]}]{\dent t}}

### 2020-07-10 popl21/popl21.tex
[ popl21 ] related work and conclusions

@@
 and subsumes sensitivity analysis \citep{reedPierce:icfp10} and static
-information control flow \citep{volpano:jcs96} aka security typing.
+information control flow \citep{volpano:jcs96} aka security typing
+\citep{abadiBanerjeeHeintzeRiecke:popl99}.
 Coeffects can be modeled by a comonad $\D[r]$ graded over resource
@@
   its denotational and operational semantics and adapt Atkey and
-  Wood's substitution theorem \citeyearpar{atkeyWood:types19} to this
+  Wood's substitution theorem
+  \citeyearpar{atkeyWood:types19,woodAtkey:linearity20} to this
   version of CBPV.
@@
 \begin{proof}
-  Straightforward adaptation of \citet{atkeyWood:types19}.
+  Straightforward adaptation of \citet{atkeyWood:types19,woodAtkey:linearity20}.
 \end{proof}
@@
 
+\subsection{Discussion}
+
+Graded CBPV provides syntax to work with graded monads $\T[e]$ and
+comonads $\D[r]$.  \emph{A priori}, one could have expected that these
+constructs are directly reflected in the syntax of types, as
+$\compty e P$ and $\thunkty r N$.  Surprisingly, the adaptation of
+graded effect and coeffect typing to CBPV places the qualifiers in the
+opposite way, as $\compty r P$ and $\thunkty e N$.  Given our
+semantics of types as (co)monad (co)algebras, this apparent oddity has
+a natural explanation: the qualifiers do not instantiate the grade in
+the monad or comonad, but in the respective (co)algebra.  The
+alternative concrete syntax $\diamond P_r$ and $\Box N_e$ would maybe
+transport the intended semantics, $\T[-]{\Denp P r}$ and
+$\D[-]{\Denn N e}$, more directly.
+
+How does Graded CBPV work as a programming language?  A program is
+usually a set of definitions and then an entrypoint in form of an
+expression or simply the name of the main procedure.  In Graded CBPV,
+the definitions are given as a sequence of let-bindings whose final
+body serves as the entrypoint.  The entrypoint should be an expression
+of type $\compty 1 P$, producing an observable value and thus, the
+effects leading up to this result.  The expressions bound by the lets
+will often be functions (or records), but wrapped in thunks
+$\thunkty e N$ to satisfy the formal requirement to be of positive
+type.  This thunk is where the effects are declared that a function
+produces.  Thus, in the big picture, the effects of a computation
+\emph{are} stated in the types, not just in the typing judgement
+$\qeCompTy \gamma \Gamma t e N$, even though the effect annotation is
+in a \emph{a priori} unexpected location.
+
+\section{Related Work}
+
+\subsection{Effect tracking in CBPV}
+
+\citet{mcDermottMycroft:esop19} observe that CBPV can represent
+call-by-name and call-by-value evaluation strategies but not
+call-by-need.  As a remedy, they extend CBPV to ECBPV by variables
+$\underline x : \diamond P$ that hold monadic values, plus a
+corresponding let-binding construct.  They further present an
+effect-graded version of ECBPV, drawing effect qualifiers from a
+preordered monoid.  In contrast to our version, effects annotations
+$f$ are stored at monadic types $\compty f P$.  Instead of our
+judgement $\CompTy \Gamma t e N$ that accumulates effect traces $e$
+and a semantic interpretation of computation types $N$ as graded monad
+algebras, they define a syntactic operation $\compty f N$ that pushes
+down effect qualifier $f$ to the monadic types $\compty f' P$ in $N$
+where the effects are combined to $\compty{f \bu f'} P$.  This way of
+handling effect grading seems equivalent to our approach, albeit it is
+syntactic rather than semantic.
+
+In previous work, \citet{mcDermottMycroft:ocs18} use coeffect typing
+to track variable usage in call-by-need simply-typed lambda calculus
+with conditionals.  Compared to the usual resource semiring $\R$
+tailored to call-by-name usage analysis, their coeffect algebra is
+more expressive: it can produce traces of variable uses, including
+bound variables.  They show how graded effect tracking can be
+simulated by the coeffect typing, \eg, for the effect of
+non-determinism.  Naturally, they focus on operational semantics.
+
+% Filinski’s M3L (MultiMonadic MetaLanguage) [11],
+% whose latest version draws heavily on CBPV [12]
+
+To accomodate several sorts of effects in one language (CBPV hosts
+only a single effect type), \citet{kammarPlotkin:popl12} present MAIL,
+the Multi-Adjunctive Intermediate Language, which is a version of CBPV
+\emph{parameterized} by effect qualifiers.  In contrast to our
+effect-graded version of CBPV, effect qualifiers are a parameter to
+the type system as a whole, basically a \emph{mode} attached to the
+typing judgement.  Via effect subsumption, one can switch to a
+``wider'' mode.  However, as effects qualifiers are drawn from a
+preordered set rather than a monoid, more detailed static effect
+traces cannot be captured.  In their own words:
+\begin{quotation}
+  Instead of one kind of computation, for each effectset
+  $\varepsilon \in E$, we have $\varepsilon$-computations
+  $\mathsf{Comp}(\varepsilon)$ that can cause effects in
+  \ensuremath{\varepsilon}. We view MAIL as multiple copies of CBPV,
+  one for each \ensuremath{\varepsilon}, sharing the same values. One
+  can translate between these different CBPVs by means of coercion
+  [...]
+\end{quotation}
+
+% POPL12:
+% Algebraic Foundations for Effect-Dependent Optimisations
+% Ohad Kammar   Gordon D. Plotkin
+% w
+% - MAIL: Multi-Adjunctive Intermediate Language
+% - Both \Box and \diamonad annotated with effects
+% - No effect accumulation
+
+\subsection{Coeffect typing}
+
+Our style of coeffect typing is heavily influenced by
+\citet{mcBride:wadler60} and \citet{atkey:lics18} who separate the
+quantification of variable \emph{usage} $\gamma$ from the variable
+\emph{declaration} in an ordinary typing context $\Gamma$.  This
+division of labor removed a major obstacle in the integration of
+linear and dependent types: the customary sorting of linear and
+unrestricted variables into separate typing contexts, and the
+dominance of context concatenation operations in the typing of
+multiplicative connectives in linear logic, \eg:
+\[
+  \ru{\Delta_1 \vdash v_1 : A_1 \qquad
+      \Delta_2 \vdash v_2 : A_2
+    }{\Delta_1.\Delta_2 \vdash (v_1, v_2) : A_1 \otimes A_2}
+\]
+In contrast, the coeffect type systems of
+\citet{reedPierce:icfp10},
+\citet{orchard:icfp14},
+\citet{brunel:esop14},
+\citet{ghicaSmith:esop14}, and
+\citet{orchard:icfp19}
+keep context concatenation which does no harm in simply-typed settings
+but seems to duplicate a mechanism already present in form of the
+resource qualifier $0$.
+
+Our treatment of substitution in coeffect-graded CBPV is based on
+\citet{woodAtkey:linearity20}, who formalize (in Agda) substitution
+for intuitionistic linear logic with a subexponential representing a
+graded comonad.  They observe that instead of a preordered semiring, a
+left skew semiring can be used as resource algebra.  In a skew
+semiring, the laws involving multiplication only hold as inequalities,
+not equalities.
+
+\subsection{Interaction of effects and coeffects}
+
+\citet{orchard:icfp16} investigate systematically the possible
+interactions between semiring-graded coeffects and monoid-graded
+effects, in form of distributive laws between the graded comonad and
+monad.  We do not rely on such distributive laws in Graded CBPV.  The
+strict separation of value and computation types and the placement of
+the monad and comonad at the transition points makes interaction
+optional.  How to integrate distributive laws into Graded CBPV is left
+for future research.
+
+\section{Conclusions}
+
+In this article, we have demonstrated that CBPV can accommodate graded
+effects and coeffects in a smooth way, keeping the term grammar
+virtually unchanged and only adding effect and coeffect annotations at
+the transition points between value and computation types.  The
+surprising simplicity of our solution speaks for the design quality of
+Levy's CBPV calculus \cite{levy:hosc06}.  For instance, the analogy to
+intuitionistic linear logic---that lies at the heart of coeffect
+typing---is already implicitly present in the separation of value
+$(\otimes)$ and computation type products ($\Pi$).
+
+For the semantics of Graded CBPV, we have evolved the concepts of
+monad algebra and comonad coalgebra to their graded versions.
+
+In future work, we would like to investigate whether the distributive
+laws of \citet{orchard:icfp16} carry over to Graded CBPV, \eg, whether
+and how a graded comonad can be distributed over a graded monad
+algebra or, in the symmetric case, a graded monad distributes over a
+graded comonad coalgebra.
+
 %% Acknowledgments

### 2020-07-10 popl21/popl21.tex
[ popl21 ] submission

@@
 \begin{abstract}
-Call-by-push-value (CBPV) is a simply typed lambda calculus that polarizes
-types into value and computation types and can thus express both
-call-by-name and call-by-value evaluation in the presence of effects.
-Semantically, effects are modeled by a monad, and computation types as
-algebras over this monad.
-%
-Effect type systems usually express more information than the presence
-of an effect; often effects are categorized by preordered monoid
-where the monoid operation represents accumulation of effects and the
-order expresses effect subsumption, in analogy to subtyping.
-In this work, a \emph{graded} version of CBPV is presented where the typing
-of computations likens effect typing.  Semantically, computation types
-are then represented as graded monad algebras.
-%
-Observing that the value types of CBPV can be interpreted as comonad coalgebras, we further present a version of CBPV that has coeffects graded by a preordered semiring.
-Value types and contexts are interpreted as graded comonad coalgebras, allowing resource-aware interpretations of CBPV.
-%
-Finally, we combine the two systems into a fully graded version of CBPV where both effects and coeffects are graded.
-Thus turns out to be possible without specifying an interaction between effects and coeffects.
+Call-by-push-value (CBPV) is a simply typed lambda calculus that
+polarizes types into value and computation types and can thus
+express both call-by-name and call-by-value evaluation in the
+presence of effects.  Semantically, effects are modeled by a monad,
+and computation types as algebras over this monad.
+
+Effect type systems usually express more information than the
+presence of an effect; often effects are categorized by preordered
+monoid where the monoid operation represents accumulation of effects
+and the order expresses effect subsumption, in analogy to subtyping.
+In this work, a \emph{graded} version of CBPV is presented where the
+typing of computations likens effect typing.  Semantically,
+computation types are then represented as graded monad algebras.
+
+Observing that the value types of CBPV can be interpreted as comonad
+coalgebras, we further present a version of CBPV that has coeffects
+graded by a preordered semiring.  Value types and contexts are
+interpreted as graded comonad coalgebras, allowing resource-aware
+interpretations of CBPV.
+
+Finally, we combine the two systems into a fully graded version of
+CBPV where both effects and coeffects are graded.  This turns out to
+be possible without specifying any interaction between effects and
+coeffects.
 \end{abstract}
@@
 functor $\diamond$ is a left adjoint to functor $\Box$ and need not be
-a monad.  Quite the opposite, when modelling store
+a monad.  Quite the opposite: when modelling store,
 $\diamond P = P \times S$ is a comonad and $\Box N = S \to N$ a monad.
@@
 keep track of not just the presence
-of an effect or coeffect but also the kind of effect and coeffect.
+of an effect or coeffect but also the \emph{kind} of effect and coeffect.
 Semantically, precise information about effects can be obtained by
@@
 $\qValTy \gamma \Gamma v P$ and computation typing
-$\qCompTy \gamma \Gamma t N$ are based on linear typing an come
+$\qCompTy \gamma \Gamma t N$ are based on linear typing and come
 equipped with a resource context~$\gamma$.  Semantically, values are
@@
 to the careful placement of monad and comonad in CBPV's type system,
-and the lack of need for scaling % of not just values but also
-of computations.
+so that scaling is restricted to values and does not arise for computations.
+% and the lack of need for scaling % of not just values but also
+% of computations.
 
@@
 
+Our style of effect tracking and its interpretation via graded monads
+is based on \citet{wadlerThiemann:tocl03} and \citet{katsumata:popl14}.
+
 \subsection{Effect tracking in CBPV}
@@
 effect-graded version of ECBPV, drawing effect qualifiers from a
-preordered monoid.  In contrast to our version, effects annotations
+preordered monoid.  In contrast to our version, effect annotations
 $f$ are stored at monadic types $\compty f P$.  Instead of our
@@
 algebras, they define a syntactic operation $\compty f N$ that pushes
-down effect qualifier $f$ to the monadic types $\compty f' P$ in $N$
+down effect qualifier $f$ to the monadic types $\compty {f'} P$ in $N$
 where the effects are combined to $\compty{f \bu f'} P$.  This way of
@@
 tailored to call-by-name usage analysis, their coeffect algebra is
-more expressive: it can produce traces of variable uses, including
-bound variables.  They show how graded effect tracking can be
+more expressive: it can produce traces of variable usage, including
+usage of bound variables.  They show how graded effect tracking can be
 simulated by the coeffect typing, \eg, for the effect of

### 2020-07-10 popl21/popl21.tex
[ popl21 ] typos fixed after submission

@@
 Our style of coeffect typing is heavily influenced by
-\citet{mcBride:wadler60} and \citet{atkey:lics18} who separate the
-quantification of variable \emph{usage} $\gamma$ from the variable
+\citet{mcBride:wadler60} and \citet{atkey:lics18} who keep the
+quantification of variable \emph{usage} in usage context $\gamma$
+separate from the variable
 \emph{declaration} in an ordinary typing context $\Gamma$.  This
-division of labor removed a major obstacle in the integration of
+division of labor has removed two major obstacle in the integration of
 linear and dependent types: the customary sorting of linear and
@@
 dominance of context concatenation operations in the typing of
-multiplicative connectives in linear logic, \eg:
+the multiplicative connectives of linear logic, \eg:
 \[
@@
 \]
-In contrast, the coeffect type systems of
+In contrast to our style, the coeffect type systems of
 \citet{reedPierce:icfp10},
@@
 \citet{orchard:icfp19}
-keep context concatenation which does no harm in simply-typed settings
+keep context concatenation. This does no harm in simply-typed settings,
 but seems to duplicate a mechanism already present in form of the

### 2020-07-22 popl21/popl21.tex
[ popl21 ] more typos fixed after submission

@@
 \newcommand{\bla}{\ensuremath{\mbox{$$}}}
-\newcommand{\ie}{\emph{i.e.}}
-\newcommand{\eg}{\emph{e.g.}}
-\newcommand{\Eg}{\emph{E.g.}}
+\newcommand{\ie}{\emph{i.e.}\xspace}
+\newcommand{\eg}{\emph{e.g.}\xspace}
+\newcommand{\Eg}{\emph{E.g.}\xspace}
 \newcommand{\loccit}{\emph{loc.\,cit.}\xspace}
@@
 %\newcommand{\raise}[1]{\traise\,#1} %clash
-\newcommand{\thandle}{\mathsf{handle}}
-\newcommand{\handle}[4][]{\thandle_{#1}\,#2\,#3\,#4}
+\newcommand{\thandle}[1][]{\mathsf{handle}_{#1}}
+\newcommand{\handle}[4][]{\thandle[{#1}]\,#2\,(#3,#4)}
 
@@
 \emph{Kleisli arrows}.
-An example module would be the \emph{writer monad}
+An example monad would be the \emph{writer monad}
 $\T\,A = \String \times A$
@@
 satisfying the functor laws $\tfmap\,\tid = \tid$ and
-$\tfmap\,(f \circ g) = \tfmap\,f \circ \tfmap\,g$ makes $\T$ a
+$\tfmap\,(f \circ g) = \tfmap\,f \circ \tfmap\,g$ makes $\T$ an
 \emph{(endo)functor}, written $\T : [\CC \to \CC]$.  The
@@
 Unit $\treturn : \Id \todot \T$ and multiplication
-$\tjoin : \T \comp \T \todot \T$ are \emph{natural transformations}
+$\tjoin : \T \comp \T \todot \T$ are \emph{natural transformations}---%
 which breaks down to $\fmap[\T] h \comp \treturn = \treturn \comp h$
@@
 Recall that a \emph{monoidal category} $\EE$ has a designated object
-$\I : \EE$ and an operation $\_{\otimes}\_ : [\EE \to [\EE \to \EE]$,
+$\I : \EE$ and an operation $\_{\otimes}\_ : [\EE \to [\EE \to \EE]]$,
 the \emph{tensor product} on objects of $\EE$ that is functorial in
@@
 Call-by-push value \cite{levy:hosc06} is a refinement of Moggi's
-computational lambda-calculus \cite{moggi:infcomp91} that allows effects
+computational lambda-calculus \citeyearpar{moggi:infcomp91} that allows effects
 not only in monadic types, \ie, in objects $\T\,A$, but more generally
@@
 
-Graded monad algebras are as well closed under
+Graded monad algebras are, like non-graded ones, closed under
 pointwise products $(B \times B')_e = B_e \times B'_e$ and
@@
 \begin{remark}[Naturality of $\trun$]
-  The $\T$-algebra $B : [\Eff \to \CC]$ comes with a natural morphism
+  The $\T$-algebra $B : [\Eff \to \CC]$ comes with a \emph{natural} morphism
   $\run[e_1,e_2] : \Hom {\T[e_1]\,B_{e_2}} {B_{e_1 \bu e_2}}$ in the
@@
 Positive types are generated from base types $o$ via
-disjoint sums $\sumty i I P$ with tag set $I$, eager products $\tupty
-i I P$ of arity $I$ (composed
-from $1$ and $A_1 \times A_2$ in \loccit) and thunking $\thunkty e N$
+disjoint sums $\sumty i I P$ with tag set $I$,
+eager products $\tupty i I P$
+(composed from $1$ and $A_1 \times A_2$ in \loccit)
+of arity $I$
+and thunking $\thunkty e N$
 (written $U\,\underline B$ in \loccit).  In contrast to pure CBPV,
@@
 Terms are separated into values $v \in \PTm$ and computations $t \in
-\NTm$ and identical to pure CBPV, modulo changes in the concrete
+\NTm$ and are identical to pure CBPV, modulo changes in the concrete
 syntax.  Values introduce positive types,
@@
 \item Coproduct $\coprod_{i:I} A_i$ with injections $\iota_i :
-  \CC(A_i,\coprod_I A)$ and cotupling $[f_i]_{i:I} : \CC(\coprod_I A,
-  B)$ for $f_i : \CC(A_i,B)$.
+  \CC(A_i,\coprod_I A)$,
+  cotupling $[f_i]_{i:I} : \CC(\coprod_I A, B)$ for $f_i : \CC(A_i,B)$,
+  and distribution morphism
+  $\tdist : \Hom {C \times \coprod_I A} {\coprod_{i:I} {(C \times A_i)}}$.
 \item Exponential $A \To B$ with $\Lambda : \CC(C \times A, B) \to
@@
 $\tcatch$, define a family
-$\thandle_N\,e : \Hom[\SET] {N_{E_1} \times N_{E_2}} {N_{\eee}}$ by
+$\thandle[N]\,e : \Hom[\SET] {N_{E_1} \times N_{E_2}} {N_{\eee}}$ by
 induction on $N \in \NTy$:
 \[
-\begin{array}{lcl}
-  \handle[\diamond P] e g h & = & \left\{
+\begin{array}{l@{~}lcl}
+  \handle[\diamond P] {&e} g h & = & \left\{
                                   \begin{array}{ll}
@@
 \right. \\
-  \handle[P \To N] e g h \, a & = & \handle[N] e {(g\,a)} {(h\,a)} \\
-  \handle[\tupt I N] e g h \, i & = & \handle[N_i] e {(g\,i)} {(h\,i)} \\
+  \handle[P \To N] {&e} g h \, a & = & \handle[N] e {g\,a} {h\,a} \\
+  \handle[\tupt I N] {&e} g h \, i & = & \handle[N_i] e {g\,i} {h\,i} \\
 \end{array}
@@
 Catching is then
-$\dent {\catch u e t} = \handle[N] e \comp \langle \dent u, \dent t
+$\dent {\catch u e t} = \thandle[N]\,e \comp \langle \dent u, \dent t
 \rangle$.
@@
 prevented statically by graded effect typing.  This could be modelled
-by adding a maximal element $\top : \Eff$---with $e \leq \top$ for all
-$e : \Eff$ that signifies an inconsistent state.  This error element
+by adding a maximal element $\top : \Eff$ (with $e \leq \top$ for all
+$e : \Eff$) that signifies an inconsistent state.  This error element
 would also be dominant in sequences, \ie,
@@
 \newcommand{\tmodify}{\mathsf{modify}}
-\newcommand{\thead}{\mathsf{head}}
-\newcommand{\ttail}{\mathsf{tail}}
+\newcommand{\thead}{\ensuremath{\mathsf{head}}}
+\newcommand{\ttail}{\ensuremath{\mathsf{tail}}}
 \newcommand{\suc}{(\_{+}1)}
@@
   \thead\,(\tdisplay\,s) & = & s \\
-  \ttail\,(\tdisplay\,s) & = & \tdisplay\,(\ttail\,s) \\
+  \makebox[0pt][l]{\ttail}\phantom{\thead}\,(\tdisplay\,s) & = & \tdisplay\,(\ttail\,s) \\
 \end{array}
@@
 A \emph{monoidal} comonad implements a morphism $\tzip$ that combines
-a tuples of comonadic values into one comonadic tuple:
+a tuples of comonadic values into one comonadic tuple---here
+implemented for the stream comonad:
 \[
@@
 like the store comonad $(S \To \_) \times S$,
-though not all, \eg, the context comonad $S{\times}\_$.
+though not all, \eg, the context comonad $S \times \_$.
 In this article, we consider only monoidal comonads.
@@
 $(\CC,\otimes,\I,\lolli)$.
-Introduction $\Lambda$ and elimination $\teval$ of
-exponentials are there
+There,
+introduction $\Lambda$ and elimination $\teval$ of
+exponentials are
 $\Lambda : \Hom {C \otimes A} B \to \Hom C {A \lolli B}$
@@
 Herein, we use as second monoidal structure
-on the functor category $[\CC \to CC]$ the
+on the functor category $[\CC \to \CC]$, the
 pointwise tensor product of functors
@@
 $\Gamma$.  A common pattern in the rules is that resource requirements
-of the subterms are added when both subterms are or maybe evaluated at
+of the subterms are \emph{added}
+when \emph{both} subterms are or may be evaluated at
 runtime (rules \rintro\otimes, \rlet, %$\rlet^{+/-}$,
@@
 $\rintro\diamond$ is the last opportunity for scaling before entering
-the monad.  Typing of bind ($\bind x u t$) attaches the resource qualifier $r$ store in $\compty r P$ to the variable $x$ (\relim\diamond).
+the monad.
+Typing of bind ($\bind x u t$) attaches the resource qualifier $r$
+stored in $\compty r P$ to the variable $x$ (\relim\diamond).
 
@@
 Herein, \fbox{$\qsubst v r x t$} shall the denote the
-(capture-avoiding) replacement $\subst v x t$ of a positive variable
-$x$ annotated by $r$ by value $v$ in term $t$.
+(capture-avoiding) replacement $\subst v x t$ of
+an $r$-annotated positive variable $x$
+by value $v$ in term $t$.
 %
@@
 are derivable (using the $\beta$-laws) from the $\eta$-laws, like
+the $\eta$-$\otimes$-law
 $\qsubst{v}{r}{z}{t'} = \qsplits r {v} {\bar x} {\qsubst {\ptup {\bar
-      x}} r z {t'}}$ for $\otimes$.
+      x}} r z {t'}}$
+(let $t' = \app {(\qsplits r z {\bar x} t)} w$).
 
@@
 A substitution $\sigma$ is a finite map from variable to terms,
-mapping positive variables $x$ to values $v$ and negative variables
-$y$ to computations $t$.  Substitution typing
+mapping positive variables $x$ to values $v$.
+% and negative variables $y$ to computations $t$.
+Substitution typing
 \fbox{$\SubstTy \Psi \Gamma \sigma \Delta$} is equipped with a matrix
@@
 As in effect-graded CBPV, negative types $N$ are interpreted as graded
-monad algebras $\denn N : [(\Eff,\leq) \to \CC]$, and as in
+monad algebras $\denn N : [(\Eff,\leq) \to \CC]$. Dually, as in
 coeffect-graded CBPV, positive types are interpreted as graded comonad
@@
 $f$ are stored at monadic types $\compty f P$.  Instead of our
-judgement $\CompTy \Gamma t e N$ that accumulates effect traces $e$
+judgement $\CompTy \Gamma t e N$ (that accumulates effect traces $e$)
 and a semantic interpretation of computation types $N$ as graded monad
@@
 surprising simplicity of our solution speaks for the design quality of
-Levy's CBPV calculus \cite{levy:hosc06}.  For instance, the analogy to
+Levy's CBPV calculus \citeyearpar{levy:hosc06}.  For instance, the analogy to
 intuitionistic linear logic---that lies at the heart of coeffect
